\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 33: Normal-Form Games and Pure Nash Equilibrium}}
\fancyhead[R]{\thepage}
\title{Lesson 33: Normal-Form Games and Pure Nash Equilibrium}
\author{AI/ML Mathematics}
\date{}

\DeclareMathOperator{\BR}{BR}

\begin{document}
\maketitle
\tableofcontents
\newpage

%% SECTION 1: MOTIVATION AND OVERVIEW

\section{Motivation and Overview}
\label{sec:motivation}

Consider two autonomous ride-sharing companies operating in the same city.
Each company must set a price for its rides---high or low---without knowing
the competitor's choice in advance.  If both set high prices, both earn
comfortable profits.  If one undercuts the other, the cheaper service
captures most of the market.  If both set low prices, both barely break
even.  How should a company reason about pricing?

This scenario is fundamentally different from single-agent optimization.
In a standard optimization problem, there is one decision-maker and the
outcome depends only on that decision.  Here, each company's profit
depends on what the \emph{other} company does.  This interdependence of
decisions is the defining feature of a \emph{game} in the mathematical
sense, and the mathematical framework for analyzing such situations is
\emph{game theory}.

\begin{rlconnection}
In single-agent reinforcement learning, the environment is modeled as a
Markov decision process whose transition kernel does not \emph{adapt} to
the agent's behavior.  Multi-agent RL (MARL) breaks this assumption: each
agent's effective ``environment'' includes the other agents.  Every MARL
algorithm implicitly targets a game-theoretic solution concept:
\begin{itemize}[nosep]
    \item \textbf{Independent Q-learning}: each agent ignores others; if it
    converges, the result is a Nash equilibrium of the stage game.
    \item \textbf{Minimax-Q} \citep{shoham2008}: targets the minimax value in
    zero-sum stochastic games, applying von Neumann's minimax theorem.
    \item \textbf{Nash-Q}: computes Nash equilibria of stage games at each step.
    \item \textbf{MAPPO, QMIX}: cooperative algorithms whose analysis relies on
    common-payoff game structure.
\end{itemize}
Without understanding Nash equilibria and dominance, one cannot state what
these algorithms compute, let alone prove convergence.
\end{rlconnection}

\paragraph{Prerequisites.}
This lesson assumes no prior exposure to game theory.  The mathematical
prerequisites are modest: basic set notation and finite probability spaces
(Lesson~9), familiarity with $\argmax$ and elementary logic (``for all,''
``there exists''), and comfort reading tables of numbers.  The proof of the
IESDS-rationalizability equivalence (Proposition~\ref{prop:iesds_rational})
uses the supporting hyperplane theorem from convex analysis (Lesson~17),
but this result can be taken on faith on a first reading.

\paragraph{Learning objectives.}
After completing this lesson, you will be able to:
\begin{itemize}[nosep]
    \item \textbf{Define} a normal-form game, strategy profile, payoff
    function, and bimatrix representation.
    \item \textbf{Classify} strategies as strictly dominant, strictly
    dominated, or neither, and apply iterated elimination of strictly
    dominated strategies (IESDS).
    \item \textbf{Compute} best response correspondences for any finite
    bimatrix game and use them to find all pure Nash equilibria.
    \item \textbf{Prove} that dominant strategy equilibria are unique Nash
    equilibria and that every Nash equilibrium survives IESDS.
    \item \textbf{Analyze} the Prisoner's Dilemma, Battle of the Sexes,
    Matching Pennies, and Stag Hunt, explaining why each exhibits
    qualitatively different equilibrium behavior.
    \item \textbf{Explain} the relationship between IESDS survival and
    rationalizability in two-player games.
\end{itemize}

%% SECTION 2: REFERENCES

\section{References}
\label{sec:references}

\begin{itemize}[nosep]
    \item \citet[Chapters 1--3]{osborne1994} --- Comprehensive introduction
    to normal-form games, dominance, and Nash equilibrium with careful proofs.
    \item \citet[Chapters 1--2]{fudenberg1991} --- Graduate-level treatment of
    strategic-form games, rationalizability, and iterated dominance.
    \item \citet[Chapters 3--4]{shoham2008} --- Game theory from the
    perspective of multiagent systems, with connections to AI and MARL.
    \item \citet{nash1950} --- Nash's original one-page proof of equilibrium
    existence (uses mixed strategies; covered in Lesson~34).
\end{itemize}

%% SECTION 3: INFORMAL INTRODUCTION

\section{Games as Models of Strategic Interaction}
\label{sec:what_is_game}

Before introducing formal definitions, we describe the three ingredients
that make up every game.

\paragraph{Players.}
A game involves two or more \emph{players} (also called \emph{agents} in
the MARL literature).  Each player is a distinct decision-maker with their
own objectives.  In the ride-sharing example, the two companies are the
players.

\paragraph{Strategies.}
Each player chooses from a set of available actions, called
\emph{strategies}.  Crucially, players choose \emph{simultaneously}: no
one observes the others' choices before committing to their own.  Even if
decisions are not literally simultaneous in clock time, what matters is that
each player acts without knowing the others' choices.

\paragraph{Payoffs.}
Once all players have chosen their strategies, each player receives a
numerical \emph{payoff}.  The payoff depends on \emph{everyone's} choice,
not just one's own.  This is the key difference from single-agent
optimization: your outcome is determined jointly by your action and the
actions of others.

\paragraph{Reading a payoff table.}
For two-player games with finitely many strategies, we organize payoffs
into a table.  Each row corresponds to a strategy of player~1 (the ``row
player'') and each column to a strategy of player~2 (the ``column
player'').  Each cell contains a pair $(a, b)$: the first number~$a$ is
player~1's payoff, and the second~$b$ is player~2's payoff.

For instance, consider a simplified pricing game where each company
chooses \emph{High} ($H$) or \emph{Low} ($L$):
\[
\begin{array}{c|cc}
    & H & L \\ \hline
  H & (5, 5) & (1, 6) \\
  L & (6, 1) & (2, 2)
\end{array}
\]
If company~1 chooses High and company~2 chooses Low, the entry is
$(1, 6)$: company~1 earns~1 (losing most of its customers) while company~2
earns~6 (capturing the market).  Notice that Low gives a higher payoff
regardless of the opponent's choice ($6 > 5$ and $2 > 1$), so Low is the
``obvious'' strategy---a situation we will formalize as \emph{strict
dominance}.

With these three ingredients in mind, we now give precise definitions.

%% SECTION 4: NORMAL-FORM GAMES

\section{Normal-Form Games}
\label{sec:normal_form}

\begin{definition}[Normal-Form Game]
\label{def:normal_form}
A \emph{normal-form game} (or \emph{strategic-form game}) is a tuple
$\Gamma = (N, (S_i)_{i \in N}, (u_i)_{i \in N})$ where:
\begin{enumerate}[nosep,label=(\roman*)]
    \item $N = \{1, 2, \ldots, n\}$ is a finite set of \emph{players}.
    \item For each $i \in N$, $S_i$ is a non-empty set of \emph{strategies}
    (or \emph{actions}) available to player~$i$.  The set of
    \emph{strategy profiles} is $S = \prod_{i \in N} S_i$.
    \item For each $i \in N$, $u_i \colon S \to \R$ is player~$i$'s
    \emph{payoff function} (or \emph{utility function}).
\end{enumerate}
The game is \emph{finite} if every $S_i$ is a finite set.
\end{definition}

\begin{remark}[Notation]
\label{rem:notation}
For a strategy profile $s = (s_1, \ldots, s_n) \in S$, we write
$s_{-i} = (s_1, \ldots, s_{i-1}, s_{i+1}, \ldots, s_n) \in
S_{-i} := \prod_{j \neq i} S_j$
and use $s = (s_i, s_{-i})$ to emphasize player~$i$'s component.  This
notation separates ``my choice'' ($s_i$) from ``everyone else's choices''
($s_{-i}$) and is used throughout game theory and MARL.
\end{remark}

\begin{definition}[Bimatrix Game]
\label{def:bimatrix}
A \emph{bimatrix game} is a two-player finite normal-form game presented as a
pair of matrices $(A, B) \in \R^{m \times k} \times \R^{m \times k}$, where
player~1 has $m$ strategies, player~2 has $k$ strategies, $A_{jl} = u_1(j,l)$,
and $B_{jl} = u_2(j,l)$.  If $B = -A$, the game is \emph{zero-sum}:
whatever one player gains, the other loses.
\end{definition}

We now introduce four classic two-player games that illustrate qualitatively
different strategic situations.  These games will serve as running examples
throughout the lesson.

\begin{example}[Prisoner's Dilemma]
\label{ex:pd}
Two suspects are interrogated separately.  Each can \emph{Cooperate} ($C$,
stay silent) or \emph{Defect} ($D$, betray the other).  The payoff bimatrix
is:
\[
\begin{array}{c|cc}
    & C & D \\ \hline
  C & (3,3) & (0,5) \\
  D & (5,0) & (1,1)
\end{array}
\]
Each entry $(a,b)$ gives (row player's payoff, column player's payoff).
Regardless of the opponent's choice, Defect yields a strictly higher payoff:
if the opponent cooperates, $D$ gives~5 versus $C$'s~3; if the opponent
defects, $D$ gives~1 versus $C$'s~0.  This makes $D$ a
\emph{strictly dominant strategy} for both players---a concept we formalize
in Section~\ref{sec:dominance}.
\end{example}

\begin{example}[Battle of the Sexes]
\label{ex:bos}
Players~1 and~2 prefer different activities---\emph{Opera} ($O$) and
\emph{Football} ($F$)---but both prefer coordination to miscoordination:
\[
\begin{array}{c|cc}
    & O & F \\ \hline
  O & (3,1) & (0,0) \\
  F & (0,0) & (1,3)
\end{array}
\]
Neither player has a dominant strategy: player~1 prefers $O$ if player~2
chooses $O$, but prefers $F$ if player~2 chooses $F$.  The ``right'' choice
depends entirely on what the other player does.  Both $(O,O)$ and $(F,F)$
will turn out to be pure Nash equilibria (Section~\ref{sec:nash}).
\end{example}

\begin{example}[Matching Pennies]
\label{ex:mp}
A zero-sum game: each player shows \emph{Heads} ($H$) or \emph{Tails} ($T$).
Player~1 wins if the choices match; player~2 wins if they differ:
\[
\begin{array}{c|cc}
    & H & T \\ \hline
  H & (1,-1) & (-1,1) \\
  T & (-1,1) & (1,-1)
\end{array}
\]
Player~1 wants to match; player~2 wants to mismatch.  No matter what
profile is played, one player wants to switch.  This game has no pure Nash
equilibrium (Proposition~\ref{prop:mp_no_pure}), motivating the extension
to mixed strategies in Lesson~34.
\end{example}

\begin{example}[Stag Hunt]
\label{ex:stag}
Two hunters choose to hunt a \emph{Stag} ($S$, requires cooperation) or a
\emph{Hare} ($H$, safe but less rewarding):
\[
\begin{array}{c|cc}
    & S & H \\ \hline
  S & (4,4) & (0,3) \\
  H & (3,0) & (3,3)
\end{array}
\]
Both $(S,S)$ and $(H,H)$ are pure Nash equilibria.  The equilibrium $(S,S)$
is \emph{payoff-dominant} (higher payoffs for both players), while $(H,H)$
is \emph{risk-dominant} (safer against uncertainty about the opponent's
choice).  This tension between ambition and safety recurs throughout
cooperative MARL.
\end{example}

\begin{rlconnection}
The Stag Hunt captures a fundamental tension in cooperative MARL: agents can
achieve higher collective reward through coordination, but each agent bears
risk if the others fail to coordinate.  This is precisely the challenge faced
by centralized-training decentralized-execution (CTDE) algorithms such as
QMIX and MAPPO, where the training procedure must guide agents toward the
payoff-dominant equilibrium despite the safety of the risk-dominant one.
\end{rlconnection}

These four games illustrate four qualitatively different situations:
a game with a dominant strategy (Prisoner's Dilemma), a coordination game
with multiple equilibria (Battle of the Sexes), a purely competitive game
with no pure equilibrium (Matching Pennies), and a game where equilibrium
selection is the central challenge (Stag Hunt).  The rest of this lesson
develops the tools to analyze each systematically.

%% SECTION 5: DOMINANCE AND ITERATED ELIMINATION

\section{Dominance and Iterated Elimination}
\label{sec:dominance}

The simplest way to reason about a game is to ask: are there strategies
that a rational player would \emph{never} use?  If one strategy gives
strictly less payoff than another \emph{no matter what the opponents do},
then there is no reason to play it.  This idea leads to the concept of
dominance.

\begin{definition}[Strict and Weak Dominance]
\label{def:dominance}
Let $\Gamma = (N, (S_i), (u_i))$ be a normal-form game and fix player~$i$.
\begin{enumerate}[nosep,label=(\alph*)]
    \item A strategy $s_i \in S_i$ \emph{strictly dominates} $s_i' \in S_i$
    if
    \[
        u_i(s_i, s_{-i}) > u_i(s_i', s_{-i})
        \quad \text{for all } s_{-i} \in S_{-i}.
    \]
    In words: $s_i$ gives a strictly higher payoff than $s_i'$ against
    \emph{every possible} combination of opponents' strategies.
    \item A strategy $s_i$ \emph{weakly dominates} $s_i'$ if
    \[
        u_i(s_i, s_{-i}) \geq u_i(s_i', s_{-i})
        \quad \text{for all } s_{-i} \in S_{-i},
    \]
    with strict inequality for at least one $s_{-i}$.
\end{enumerate}
A strategy is \emph{strictly dominated} if some other strategy strictly
dominates it.  A strategy is a \emph{strictly dominant strategy} if it
strictly dominates every other strategy of that player.
\end{definition}

\begin{example}[Dominance in the Prisoner's Dilemma]
\label{ex:pd_dominance}
In Example~\ref{ex:pd}, consider player~1.  Compare $D$ against $C$ for
each possible choice by player~2:
\begin{itemize}[nosep]
    \item If player~2 plays $C$: $u_1(D, C) = 5 > 3 = u_1(C, C)$.
    \item If player~2 plays $D$: $u_1(D, D) = 1 > 0 = u_1(C, D)$.
\end{itemize}
Since $D$ gives a higher payoff than $C$ in \emph{both} cases, $D$
strictly dominates $C$ for player~1.  By the symmetry of the payoff matrix,
$D$ also strictly dominates $C$ for player~2.  Both players have a strictly
dominant strategy.
\end{example}

\begin{proposition}[Dominated Strategies Are Never Best Responses]
\label{prop:dominated_not_br}
If $s_i'$ is strictly dominated by $s_i$, then $s_i'$ is never a best
response: for every $s_{-i} \in S_{-i}$,
$s_i' \notin \argmax_{t_i \in S_i} u_i(t_i, s_{-i})$.
\end{proposition}

\begin{proof}
Fix any $s_{-i} \in S_{-i}$.  Since $s_i$ strictly dominates $s_i'$, we
have $u_i(s_i, s_{-i}) > u_i(s_i', s_{-i})$.  Therefore $s_i'$ does not
achieve the maximum of $u_i(\cdot, s_{-i})$ over $S_i$, so
$s_i' \notin \argmax_{t_i \in S_i} u_i(t_i, s_{-i})$.
\end{proof}

\begin{warning}
The converse of Proposition~\ref{prop:dominated_not_br} is false in general
for pure strategies: a strategy can fail to be a best response to any pure
opponent profile without being strictly dominated by any single pure
strategy.  However, in two-player games with finite strategy sets, a pure
strategy that is never a best response is strictly dominated by some
\emph{mixed} strategy (a consequence of the supporting hyperplane theorem).
\end{warning}

Since dominated strategies are never rational to play, we can remove them
from consideration.  But removing one player's dominated strategy may cause
a previously non-dominated strategy of another player to become dominated.
This motivates iterating the elimination process.

\begin{definition}[Iterated Elimination of Strictly Dominated Strategies]
\label{def:iesds}
\emph{Iterated elimination of strictly dominated strategies} (IESDS)
proceeds as follows.  Set $S_i^0 = S_i$ for all $i \in N$.  For
$k = 0, 1, 2, \ldots$, define
\[
    S_i^{k+1} = \bigl\{s_i \in S_i^k :
    \text{no } s_i' \in S_i^k \text{ satisfies }
    u_i(s_i', s_{-i}) > u_i(s_i, s_{-i})
    \text{ for all } s_{-i} \in S_{-i}^k\bigr\}.
\]
The set of \emph{surviving strategies} for player~$i$ is
$S_i^\infty = \bigcap_{k=0}^\infty S_i^k$.  In a finite game, this
intersection stabilizes after finitely many rounds.
\end{definition}

\begin{remark}
\label{rem:iesds_order}
Definition~\ref{def:iesds} eliminates \emph{all} strictly dominated
strategies simultaneously at each round.  One could instead eliminate
strategies one at a time, choosing different players and strategies in
different orders.  Theorem~\ref{thm:iesds_order} shows that the final
surviving set is the same regardless of the elimination order.
\end{remark}

\begin{example}[IESDS in the Prisoner's Dilemma]
\label{ex:pd_iesds}
In the Prisoner's Dilemma (Example~\ref{ex:pd}), $C$ is strictly dominated
by $D$ for both players.  Round~0 eliminates $C$ for both, leaving
$S_1^1 = S_2^1 = \{D\}$.  The unique surviving profile is $(D, D)$.
\end{example}

\begin{example}[Multi-Round IESDS]
\label{ex:iesds_multi}
Consider the game:
\[
\begin{array}{c|ccc}
    & L & C & R \\ \hline
  T & (1,2) & (2,1) & (5,0) \\
  B & (4,1) & (3,5) & (6,2)
\end{array}
\]
\emph{Round~0.}  Compare player~1's strategies: $u_1(B, s_2) > u_1(T, s_2)$
for every $s_2$ (specifically $4 > 1$, $3 > 2$, $6 > 5$).  So $T$ is
strictly dominated by $B$.  Eliminate $T$; the reduced game is:
\[
\begin{array}{c|ccc}
    & L & C & R \\ \hline
  B & (4,1) & (3,5) & (6,2)
\end{array}
\]
\emph{Round~1.}  Player~2 now knows player~1 will play $B$.  Compare
player~2's payoffs: $L$ gives~1, $C$ gives~5, $R$ gives~2.  Here $C$
strictly dominates both $L$ and $R$.  Eliminate $L$ and $R$.

The surviving profile is $(B, C)$ with payoffs $(3, 5)$.  Notice that $C$
did \emph{not} dominate $L$ in the original game (against $T$, $L$
gives~2 while $C$ gives~1); the domination only emerged after eliminating
player~1's dominated strategy.
\end{example}

\begin{theorem}[Order Independence of IESDS]
\label{thm:iesds_order}
In a finite normal-form game, the set of strategy profiles surviving IESDS
is independent of the order in which strategies are eliminated.
\end{theorem}

\begin{proof}
We show that any strategy eliminated under one elimination order is
eventually eliminated under every elimination order.  The proof proceeds by
strong induction on the round of elimination.

Let $E$ and $E'$ denote two arbitrary elimination orders.  Write $S_i^k(E)$
for the surviving strategies of player~$i$ after round~$k$ under order~$E$,
and similarly $S_i^k(E')$ for order~$E'$.  Write $S_i^\infty(E)$ and
$S_i^\infty(E')$ for the respective final surviving sets.

We prove by strong induction on $k$ that if a strategy $s_i^*$ is eliminated
at round~$k$ under order~$E$, then $s_i^* \notin S_i^\infty(E')$.

\medskip
\noindent\textbf{Base case ($k = 0$).}
If $s_i^*$ is eliminated at round~$0$ under $E$, then there exists
$s_i' \in S_i$ with $u_i(s_i', s_{-i}) > u_i(s_i^*, s_{-i})$ for all
$s_{-i} \in S_{-i}$.  Since $S_{-i}^\infty(E') \subseteq S_{-i}$, the
domination holds on $S_{-i}^\infty(E')$.  If $s_i' \in S_i^\infty(E')$,
then $s_i^* \notin S_i^\infty(E')$ directly.  Otherwise, following the
chain of dominators from $s_i'$ to some $\hat{s}_i \in S_i^\infty(E')$
(the chain terminates since $S_i$ is finite), we get
$u_i(\hat{s}_i, s_{-i}) > u_i(s_i^*, s_{-i})$ on $S_{-i}^\infty(E')$,
so again $s_i^* \notin S_i^\infty(E')$.

\medskip
\noindent\textbf{Inductive step.}
Assume the claim holds for all strategies eliminated in rounds
$0, 1, \ldots, k-1$ under any elimination order.  Suppose $s_i^*$ is
eliminated at round~$k$ under order~$E$.  Then there exists
$s_i' \in S_i^{k-1}(E)$ such that
\begin{equation}
\label{eq:domination_k}
    u_i(s_i', s_{-i}) > u_i(s_i^*, s_{-i})
    \quad \text{for all } s_{-i} \in S_{-i}^{k-1}(E).
\end{equation}

\emph{Claim:} $S_{-i}^\infty(E') \subseteq S_{-i}^{k-1}(E)$.

To prove the claim, take any $s_j \in S_j^\infty(E')$ for $j \neq i$.
Suppose for contradiction that $s_j \notin S_j^{k-1}(E)$.  Then $s_j$ was
eliminated at some round $l < k$ under order~$E$.  By the inductive
hypothesis, $s_j \notin S_j^\infty(E')$, contradicting
$s_j \in S_j^\infty(E')$.  Hence $s_j \in S_j^{k-1}(E)$, establishing the
claim.

By the claim, the strict inequality~\eqref{eq:domination_k} holds for all
$s_{-i} \in S_{-i}^\infty(E')$.  We now consider two cases.

\emph{Case 1:} $s_i' \in S_i^\infty(E')$.  Then $s_i^*$ is strictly
dominated by $s_i'$ on the surviving set $S_{-i}^\infty(E')$, so
$s_i^* \notin S_i^\infty(E')$.

\emph{Case 2:} $s_i' \notin S_i^\infty(E')$.  Then $s_i'$ was eliminated at
some round~$m$ under $E'$, meaning there exists
$s_i'' \in S_i^{m-1}(E')$ that strictly dominates $s_i'$ on
$S_{-i}^{m-1}(E')$.  Repeating this chain, since $S_i$ is finite, we
eventually reach some $\hat{s}_i \in S_i^\infty(E')$ that strictly dominates
$s_i'$ on $S_{-i}^\infty(E')$.  Combining with~\eqref{eq:domination_k}:
for all $s_{-i} \in S_{-i}^\infty(E')$,
$u_i(\hat{s}_i, s_{-i}) > u_i(s_i', s_{-i}) > u_i(s_i^*, s_{-i})$.
Hence $s_i^*$ is strictly dominated by $\hat{s}_i$ on the surviving set,
so $s_i^* \notin S_i^\infty(E')$.

\medskip
By induction, $S_i^\infty(E) \supseteq S_i^\infty(E')$ for all~$i$.
Swapping the roles of $E$ and $E'$ gives the reverse inclusion, so
$S_i^\infty(E) = S_i^\infty(E')$.
\end{proof}

\begin{keyresult}
The surviving set under IESDS is a canonical object of the game: it does not
depend on procedural choices.  In a game solvable by IESDS (where the
surviving set is a singleton), IESDS yields the unique Nash equilibrium.
\end{keyresult}

%% SECTION 6: BEST RESPONSE AND RATIONALIZABILITY

\section{Best Response and Rationalizability}
\label{sec:rationalizability}

Dominance asks ``what should I never play?''  We now ask the complementary
question: ``what \emph{should} I play?''  The answer depends on what I
expect my opponents to do.  Given a belief about the opponents' strategies,
the rational choice is whichever strategy maximizes my payoff---this is
called a \emph{best response}.

\begin{definition}[Best Response Correspondence]
\label{def:best_response}
Player~$i$'s \emph{best response correspondence} is the set-valued map
$\BR_i \colon S_{-i} \rightrightarrows S_i$ defined by
\[
    \BR_i(s_{-i}) = \argmax_{s_i \in S_i} u_i(s_i, s_{-i}).
\]
In a finite game, $\BR_i(s_{-i})$ is non-empty for every
$s_{-i} \in S_{-i}$, since a finite set always attains its maximum.
\end{definition}

\begin{example}[Best Responses in Battle of the Sexes]
\label{ex:bos_br}
In Example~\ref{ex:bos}, compute player~1's best responses:
\begin{itemize}[nosep]
    \item If player~2 plays $O$: $u_1(O,O) = 3 > 0 = u_1(F,O)$, so
    $\BR_1(O) = \{O\}$.
    \item If player~2 plays $F$: $u_1(F,F) = 1 > 0 = u_1(O,F)$, so
    $\BR_1(F) = \{F\}$.
\end{itemize}
By symmetry (with swapped preferences), $\BR_2(O) = \{O\}$ and
$\BR_2(F) = \{F\}$.  The profiles where both players play a mutual best
response are $(O,O)$ and $(F,F)$---these are exactly the Nash equilibria,
as we will see in Section~\ref{sec:nash}.
\end{example}

\begin{definition}[Rationalizable Strategies]
\label{def:rationalizable}
A strategy $s_i$ is \emph{rationalizable} if it survives iterated
elimination of strategies that are never best responses.  Formally, set
$R_i^0 = S_i$ and define
\[
    R_i^{k+1} = \bigl\{s_i \in R_i^k :
    \exists\, s_{-i} \in R_{-i}^k \text{ such that }
    s_i \in \BR_i(s_{-i})\bigr\}.
\]
The set of \emph{rationalizable strategies} is
$R_i^\infty = \bigcap_{k=0}^\infty R_i^k$.
\end{definition}

\begin{remark}
Rationalizability captures ``common knowledge of rationality'': a strategy
is rationalizable if it is a best response to some belief about the
opponents, where that belief is consistent with the opponents being
rational, and so on ad infinitum.
\end{remark}

\begin{proposition}[IESDS and Rationalizability in Two-Player Games]
\label{prop:iesds_rational}
In a finite two-player game, the set of rationalizable strategies coincides
with the set of strategies surviving IESDS: $R_i^\infty = S_i^\infty$ for
$i = 1, 2$.
\end{proposition}

\begin{proof}
We establish both inclusions.

\medskip
\noindent\textbf{($S_i^\infty \supseteq R_i^\infty$).}
Every strictly dominated strategy is never a best response
(Proposition~\ref{prop:dominated_not_br}).  Hence every strategy eliminated
in IESDS is also eliminated in the rationalizability procedure, giving
$R_i^k \subseteq S_i^k$ for all~$k$ by induction.  Taking intersections,
$R_i^\infty \subseteq S_i^\infty$.

\medskip
\noindent\textbf{($S_i^\infty \subseteq R_i^\infty$).}
We show the contrapositive: if $s_i \notin R_i^\infty$, then
$s_i \notin S_i^\infty$.  Suppose $s_i$ is eliminated at round~$k$ in the
rationalizability procedure, meaning $s_i$ is not a best response to any
$s_{-i} \in R_{-i}^{k-1}$.

In a two-player game, this means: for every $s_j \in R_j^{k-1}$ (where
$j \neq i$), there exists $s_i'(s_j) \in R_i^{k-1}$ with
$u_i(s_i'(s_j), s_j) > u_i(s_i, s_j)$.  We claim that $s_i$ is strictly
dominated on $R_j^{k-1}$ (possibly by a mixed strategy).  Indeed, consider
the payoff vectors $v = (u_i(s_i, s_j))_{s_j \in R_j^{k-1}}
\in \R^{|R_j^{k-1}|}$
and the convex set
$W = \{(u_i(s_i', s_j))_{s_j \in R_j^{k-1}} : s_i' \in R_i^{k-1}\}$.
Since $s_i$ is never a best response on $R_j^{k-1}$, by the supporting
hyperplane theorem applied to the convex hull of payoff vectors, there
exists a convex combination (i.e., a mixed strategy) over $R_i^{k-1}$ that
strictly dominates $s_i$ on $R_j^{k-1}$.

For pure-strategy IESDS where we also allow elimination by mixed-strategy
domination, this gives $s_i \notin S_i^\infty$.  In finite two-player games,
elimination by pure or mixed strict domination yields the same surviving
set \citep{osborne1994}, completing the proof.
\end{proof}

\begin{warning}
For games with three or more players, the equivalence between IESDS and
rationalizability can fail when restricting to pure-strategy domination.
Rationalizability may be strictly coarser than pure-strategy IESDS survival
for $n \geq 3$.  The equivalence is restored when domination by mixed
strategies is permitted \citep{fudenberg1991}.
\end{warning}

%% SECTION 7: NASH EQUILIBRIUM

\section{Pure Strategy Nash Equilibrium}
\label{sec:nash}

We arrive at the central concept of game theory.  Informally, a
\emph{Nash equilibrium} is a strategy profile where every player is
simultaneously doing the best they can, given what everyone else is doing.
If the game were paused and each player were told the others' strategies,
no one would want to change their own.

\begin{definition}[Pure Strategy Nash Equilibrium]
\label{def:nash_pure}
A strategy profile $s^* = (s_1^*, \ldots, s_n^*) \in S$ is a \emph{pure
strategy Nash equilibrium} (NE) if no player can unilaterally improve
their payoff:
\[
    u_i(s_i^*, s_{-i}^*) \geq u_i(s_i, s_{-i}^*)
    \quad \text{for all } s_i \in S_i, \; \text{for all } i \in N.
\]
\end{definition}

\begin{proposition}[Nash Equilibrium as Mutual Best Response]
\label{prop:ne_br}
A strategy profile $s^*$ is a Nash equilibrium if and only if
$s_i^* \in \BR_i(s_{-i}^*)$ for every $i \in N$.
\end{proposition}

\begin{proof}
By Definition~\ref{def:nash_pure}, $s^*$ is a NE if and only if
$u_i(s_i^*, s_{-i}^*) \geq u_i(s_i, s_{-i}^*)$ for all $s_i \in S_i$ and
all $i \in N$.  By Definition~\ref{def:best_response}, this is precisely
$s_i^* \in \argmax_{s_i \in S_i} u_i(s_i, s_{-i}^*) = \BR_i(s_{-i}^*)$
for all~$i$.
\end{proof}

Proposition~\ref{prop:ne_br} gives a practical algorithm for finding all
pure Nash equilibria in a bimatrix game: for each cell in the payoff table,
check whether \emph{both} players are playing a best response.

\begin{example}[The ``Underline Best Responses'' Method]
\label{ex:underline_br}
Consider the $3 \times 2$ game:
\[
\begin{array}{c|cc}
    & L & R \\ \hline
  T & (2,1) & (0,3) \\
  M & (3,0) & (1,2) \\
  B & (1,2) & (2,1)
\end{array}
\]
\emph{Step~1: Find player~1's best responses} (compare payoffs within each
column).
\begin{itemize}[nosep]
    \item Column $L$: payoffs are $2, 3, 1$.  Max is $3$ at $M$, so
    $\BR_1(L) = \{M\}$.
    \item Column $R$: payoffs are $0, 1, 2$.  Max is $2$ at $B$, so
    $\BR_1(R) = \{B\}$.
\end{itemize}
\emph{Step~2: Find player~2's best responses} (compare payoffs within each
row).
\begin{itemize}[nosep]
    \item Row $T$: payoffs are $1, 3$.  Max is $3$ at $R$, so
    $\BR_2(T) = \{R\}$.
    \item Row $M$: payoffs are $0, 2$.  Max is $2$ at $R$, so
    $\BR_2(M) = \{R\}$.
    \item Row $B$: payoffs are $2, 1$.  Max is $2$ at $L$, so
    $\BR_2(B) = \{L\}$.
\end{itemize}
\emph{Step~3: Find mutual best responses.}  A cell is a NE if and only if
both players' entries are best responses.  Here $M$ is player~1's best
response to $L$, but player~2's best response to $M$ is $R$ (not $L$);
and $B$ is player~1's best response to $R$, but player~2's best response
to $B$ is $L$ (not $R$).  No cell has both entries as best responses, so
this game has \emph{no pure Nash equilibrium}.
\end{example}

\begin{keyresult}
The ``underline best responses'' method from
Example~\ref{ex:underline_br}---mark the highest payoff in each column for
player~1 and each row for player~2, then look for cells with both entries
marked---works for any bimatrix game and is the standard way to find pure NE
by hand.
\end{keyresult}

\begin{proposition}[Dominant Strategy Equilibrium Implies Nash Equilibrium]
\label{prop:dominant_nash}
If $s_i^*$ is a strictly dominant strategy for every player $i \in N$, then
$(s_1^*, \ldots, s_n^*)$ is the unique Nash equilibrium.
\end{proposition}

\begin{proof}
Since $s_i^*$ strictly dominates every other strategy of player~$i$, we have
$u_i(s_i^*, s_{-i}) > u_i(s_i, s_{-i})$ for all $s_i \neq s_i^*$ and all
$s_{-i} \in S_{-i}$.  In particular, for $s_{-i} = s_{-i}^*$, this gives
$s_i^* \in \BR_i(s_{-i}^*)$ for all~$i$, so $s^*$ is a NE by
Proposition~\ref{prop:ne_br}.

For uniqueness, suppose $\hat{s} \neq s^*$ is a NE.  Then some
$\hat{s}_i \neq s_i^*$, but $u_i(s_i^*, \hat{s}_{-i}) > u_i(\hat{s}_i,
\hat{s}_{-i})$ by strict dominance, contradicting $\hat{s}_i \in
\BR_i(\hat{s}_{-i})$.
\end{proof}

%% SECTION 8: ANALYSIS OF CLASSIC GAMES

\section{Analysis of Classic Games}
\label{sec:classics}

We systematically analyze the four classic games from
Section~\ref{sec:normal_form}, identifying all pure Nash equilibria and
highlighting the qualitative differences between them.

\subsection{Prisoner's Dilemma}
\label{subsec:pd_analysis}

\begin{proposition}
\label{prop:pd_ne}
The Prisoner's Dilemma (Example~\ref{ex:pd}) has a unique pure Nash
equilibrium at $(D, D)$ with payoffs $(1, 1)$.
\end{proposition}

\begin{proof}
By Example~\ref{ex:pd_dominance}, $D$ is a strictly dominant strategy for
both players.  By Proposition~\ref{prop:dominant_nash}, $(D, D)$ is the
unique Nash equilibrium.
\end{proof}

\begin{remark}[Individual vs.\ Collective Rationality]
\label{rem:pd_paradox}
The profile $(C,C)$ yields payoffs $(3,3)$, which both players prefer to
the NE payoff of $(1,1)$.  Yet rational players cannot sustain cooperation
in a one-shot game.  This tension between individual and collective
rationality is central to MARL: independent learners converge to the
suboptimal NE unless the interaction is repeated (Lesson~36, folk theorems)
or agents can commit to binding agreements.
\end{remark}

\subsection{Battle of the Sexes}
\label{subsec:bos_analysis}

\begin{proposition}
\label{prop:bos_ne}
The Battle of the Sexes (Example~\ref{ex:bos}) has exactly two pure Nash
equilibria: $(O, O)$ with payoffs $(3, 1)$ and $(F, F)$ with payoffs
$(1, 3)$.
\end{proposition}

\begin{proof}
From Example~\ref{ex:bos_br}, $\BR_1(O) = \{O\}$, $\BR_1(F) = \{F\}$,
$\BR_2(O) = \{O\}$, and $\BR_2(F) = \{F\}$.  Of the four profiles, only
$(O,O)$ and $(F,F)$ satisfy $s_i^* \in \BR_i(s_{-i}^*)$ for both players.
At $(O,F)$, player~1 deviates ($O \notin \BR_1(F)$); at $(F,O)$, likewise.
\end{proof}

\begin{remark}[Equilibrium Selection]
The Battle of the Sexes illustrates the \emph{equilibrium selection
problem}: the theory predicts multiple equilibria but does not say which
will be played.  In MARL, independent learners may converge to different
equilibria or cycle.  Correlated equilibrium (Lesson~34) offers one way to
resolve this ambiguity.
\end{remark}

\subsection{Matching Pennies}
\label{subsec:mp_analysis}

\begin{proposition}[Matching Pennies Has No Pure NE]
\label{prop:mp_no_pure}
The Matching Pennies game (Example~\ref{ex:mp}) has no pure strategy Nash
equilibrium.
\end{proposition}

\begin{proof}
Player~1 wants to match and player~2 wants to mismatch, so their best
responses oppose each other: $\BR_1(H) = \{H\}$, $\BR_1(T) = \{T\}$,
$\BR_2(H) = \{T\}$, $\BR_2(T) = \{H\}$.  At every profile, exactly one
player has a profitable deviation:
\begin{itemize}[nosep]
    \item $(H,H)$: player~2 deviates ($H \notin \BR_2(H)$).
    \item $(H,T)$: player~1 deviates ($H \notin \BR_1(T)$).
    \item $(T,H)$: player~1 deviates ($T \notin \BR_1(H)$).
    \item $(T,T)$: player~2 deviates ($T \notin \BR_2(T)$).
\end{itemize}
No profile is a mutual best response, so no pure NE exists.
\end{proof}

\begin{remark}
\label{rem:mp_mixed}
The non-existence of pure NE reflects the adversarial structure of zero-sum
games.  Nash's theorem \citep{nash1950} guarantees a \emph{mixed} NE (each
player randomizes uniformly), which is the subject of Lesson~34.  In
competitive MARL, the best-response cycling $H \to T \to H \to \cdots$ is
a simple instance of the oscillation that plagues naive independent learning;
algorithms like minimax-Q and self-play converge to mixed equilibria instead.
\end{remark}

\subsection{Stag Hunt}
\label{subsec:stag_analysis}

\begin{proposition}
\label{prop:stag_ne}
The Stag Hunt (Example~\ref{ex:stag}) has exactly two pure Nash equilibria:
$(S, S)$ with payoffs $(4, 4)$ and $(H, H)$ with payoffs $(3, 3)$.
\end{proposition}

\begin{proof}
Best responses: $\BR_1(S) = \{S\}$ (since $4 > 3$), $\BR_1(H) = \{H\}$
(since $3 > 0$), and by symmetry $\BR_2(S) = \{S\}$, $\BR_2(H) = \{H\}$.
Only the diagonal profiles $(S,S)$ and $(H,H)$ satisfy mutual best response;
at $(S,H)$ and $(H,S)$, the Stag-hunter deviates.
\end{proof}

\begin{remark}[Payoff Dominance vs.\ Risk Dominance]
\label{rem:stag_selection}
The equilibrium $(S,S)$ is \emph{payoff-dominant} (both players prefer it),
while $(H,H)$ is \emph{risk-dominant} \citep{fudenberg1991}: $S$ is a best
response only when the believed probability $p$ that the opponent plays $S$
satisfies $4p > 3p + 3(1-p)$, i.e., $p > 3/4$.  Unless a player is quite
confident the other will cooperate, the safe choice is $H$.  In MARL, this
tension determines whether independent learners converge to the
Pareto-optimal or the risk-dominant equilibrium.
\end{remark}

%% SECTION 9: EXERCISES

\section{Exercises}
\label{sec:exercises}

\begin{exercise}[Three-Player Dominance]
\label{ex:three_player}
\textnormal{(\(*\))} \\
Consider a three-player game where each player chooses from $\{A, B\}$.
Player~1's payoff is $u_1(s_1, s_2, s_3) = \indicator[s_1 = A] \cdot
(1 + \indicator[s_2 = A] + \indicator[s_3 = A])$.  That is, player~1
gets $0$ from $B$ and gets $1$, $2$, or $3$ from $A$ depending on how
many others play $A$.
\begin{enumerate}[label=(\alph*)]
    \item Show that $B$ is strictly dominated by $A$ for player~1.
    \item Suppose $u_2$ and $u_3$ are defined analogously.  Find all pure
    Nash equilibria.
    \item Apply IESDS and verify that the surviving set equals the set of
    Nash equilibria.
\end{enumerate}
\end{exercise}

\begin{exercise}[Best Response Analysis]
\label{ex:br_analysis}
\textnormal{(\(*\))} \\
Consider the bimatrix game:
\[
\begin{array}{c|ccc}
    & L & C & R \\ \hline
  T & (2,1) & (0,2) & (1,0) \\
  M & (1,2) & (3,1) & (0,0) \\
  B & (0,0) & (1,3) & (2,1)
\end{array}
\]
\begin{enumerate}[label=(\alph*)]
    \item Compute $\BR_1(s_2)$ for each $s_2 \in \{L, C, R\}$ and
    $\BR_2(s_1)$ for each $s_1 \in \{T, M, B\}$.
    \item Identify all pure Nash equilibria using the mutual best response
    characterization (Proposition~\ref{prop:ne_br}).
    \item Apply IESDS.  Does IESDS alone determine the equilibrium?
\end{enumerate}
\end{exercise}

\begin{exercise}[IESDS Order Independence]
\label{ex:iesds_order}
\textnormal{(\(**\))} \\
Consider the following $3 \times 3$ bimatrix game:
\[
\begin{array}{c|ccc}
    & L & C & R \\ \hline
  T & (3,1) & (0,2) & (1,0) \\
  M & (1,0) & (2,3) & (0,1) \\
  B & (0,2) & (1,1) & (3,0)
\end{array}
\]
\begin{enumerate}[label=(\alph*)]
    \item Perform IESDS by first eliminating dominated strategies for
    player~1, then player~2.
    \item Perform IESDS in the reverse order: first player~2, then
    player~1.
    \item Verify that both orders yield the same surviving set, confirming
    Theorem~\ref{thm:iesds_order}.
\end{enumerate}
\end{exercise}

\begin{exercise}[Coordination Game Design]
\label{ex:coordination}
\textnormal{(\(**\))} \\
Design a symmetric $3 \times 3$ bimatrix game (i.e., $B = A^T$) with exactly
three pure NE (all on the diagonal) and no strictly dominated strategies.
Prove both properties hold.
\end{exercise}

\begin{exercise}[Nash Equilibrium and IESDS Survival]
\label{ex:ne_iesds}
\textnormal{(\(**\))} \\
Prove that every Nash equilibrium of a finite game survives IESDS: if $s^*$
is a pure NE, then $s_i^* \in S_i^\infty$ for all $i \in N$.

\textit{Hint:} Induction on the elimination round.
\end{exercise}

\begin{exercise}[Equilibrium Selection in Multi-Agent Training]
\label{ex:marl_selection}
\textnormal{(\(***\))} \\
Consider two Q-learning agents playing the Stag Hunt (Example~\ref{ex:stag})
repeatedly with learning rate $\alpha_t$ and $\varepsilon$-greedy
exploration.  At each round, both agents observe the joint action and update
their Q-tables independently.
\begin{enumerate}[label=(\alph*)]
    \item Suppose both agents initialize $Q(S) = Q(H) = 0$.  If the first
    outcome is $(H, H)$ with reward~3, show that both agents increase
    $Q(H)$ and are subsequently more likely to play $H$, making $(H, H)$
    self-reinforcing.
    \item Argue informally that for sufficiently small $\varepsilon$, the
    risk-dominant equilibrium $(H, H)$ is more likely to emerge than the
    payoff-dominant $(S, S)$.  Why does the basin of attraction of $(H, H)$
    matter here?
    \item Suggest a modification to the training procedure (e.g., optimistic
    initialization, reward shaping, or communication) that could steer
    convergence toward the payoff-dominant equilibrium.  Justify your
    proposal using the best-response structure of the game.
\end{enumerate}
\end{exercise}

%% SECTION 10: SUMMARY

\section{Summary}
\label{sec:summary}

\begin{itemize}
    \item A \textbf{normal-form game} $(N, (S_i), (u_i))$ models
    simultaneous strategic interaction: each player's payoff depends on
    everyone's choice, not just their own.
    \item A strategy is \textbf{strictly dominated} if another strategy
    gives a strictly higher payoff against every possible opponent
    behavior; rational players never use dominated strategies.
    \item \textbf{IESDS} iteratively removes dominated strategies.  The
    surviving set is independent of elimination order
    (Theorem~\ref{thm:iesds_order}) and is a canonical property of the game.
    \item A \textbf{best response} is a strategy that maximizes payoff given
    the opponents' choices.  The best response correspondence $\BR_i$ maps
    each opponent profile to the set of optimal replies.
    \item A \textbf{pure Nash equilibrium} is a profile where every player
    simultaneously plays a best response
    (Proposition~\ref{prop:ne_br}); no one can unilaterally improve.
    \item If every player has a strictly dominant strategy, the resulting
    profile is the unique NE (Proposition~\ref{prop:dominant_nash}).
    \item Not every game has a pure NE (Matching Pennies); some have
    multiple NE (Battle of the Sexes, Stag Hunt), raising the
    \textbf{equilibrium selection problem}.
    \item In two-player games, the set of rationalizable strategies equals
    the IESDS surviving set (Proposition~\ref{prop:iesds_rational}).
    \item These concepts underpin every MARL algorithm: convergence targets,
    solution quality, and equilibrium selection all rely on game-theoretic
    reasoning developed here.
\end{itemize}

%% SECTION 11: FURTHER READING

\section{Further Reading}
\label{sec:further_reading}

\begin{itemize}[nosep]
    \item \textbf{Lesson~34} extends the analysis to \emph{mixed strategies}
    and proves Nash's existence theorem: every finite game has at least one
    (possibly mixed) Nash equilibrium.  The minimax theorem for zero-sum
    games is also covered there.
    \item \textbf{Lesson~35} introduces \emph{extensive-form games}
    (sequential decision trees), perfect information, and subgame perfect
    equilibrium---the appropriate solution concept when players move in
    sequence rather than simultaneously.
    \item \textbf{Lesson~36} treats \emph{repeated games} and folk theorems,
    explaining how cooperation can emerge when the Prisoner's Dilemma is
    played repeatedly with patient agents.
    \item For a deeper treatment of rationalizability and epistemic
    foundations, see \citet[Chapters 4--5]{osborne1994} and
    \citet[Chapter 2]{fudenberg1991}.
    \item \citet{vonneumann1944} is the foundational work that launched game
    theory as a mathematical discipline; \citet{nash1951} introduced the
    equilibrium concept that bears his name.
\end{itemize}

\bigskip
\bibliography{game-theory-refs}

\end{document}
