\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 36: Repeated Games and Folk Theorems}}
\fancyhead[R]{\thepage}
\title{Lesson 36: Repeated Games and Folk Theorems}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% SECTION 1: MOTIVATION

\section{Motivation}
\label{sec:motivation}

In one-shot games, players interact once and the outcome is determined by
the simultaneous choice of strategies. The Prisoner's Dilemma, for
instance, has a unique Nash equilibrium at mutual defection---even though
both players would prefer mutual cooperation. This pessimistic conclusion
changes dramatically when the game is played \emph{repeatedly}.

Repeated interaction introduces two mechanisms absent from one-shot play:
\begin{enumerate}[label=(\roman*)]
    \item \textbf{History-dependent strategies.} Players can condition
    current actions on the entire history of past play, enabling reward
    and punishment schemes.
    \item \textbf{Intertemporal incentives.} A short-term gain from
    defection can be outweighed by the long-term cost of triggering
    punishment.
\end{enumerate}

The central results of this lesson---the \emph{folk theorems}---characterize
precisely which payoff vectors are sustainable as equilibria in infinitely
repeated games. Informally, any feasible payoff that gives each player
more than their worst-case guarantee can be achieved, provided players are
sufficiently patient.

\begin{rlconnection}[title={MARL Training as a Repeated Game}]
In multi-agent reinforcement learning, agents interact over many episodes
and update their policies based on observed outcomes. This is exactly the
structure of a repeated game. The folk theorems explain why independent
learners can converge to cooperative outcomes (or to complex punishment
schemes) that would never arise in single-episode analysis. Understanding
these results is essential for predicting and shaping emergent behavior in
self-play training.
\end{rlconnection}

\noindent\textbf{Roadmap.}
\begin{itemize}
    \item Section~\ref{sec:finitely-repeated}: Finitely repeated games and
    the unraveling theorem.
    \item Section~\ref{sec:multiple-ne}: New equilibria when the stage game
    has multiple Nash equilibria.
    \item Section~\ref{sec:infinitely-repeated}: Infinitely repeated games
    with discounting.
    \item Section~\ref{sec:feasible-ir}: Feasible and individually rational
    payoffs.
    \item Section~\ref{sec:cooperation-pd}: Cooperation in the repeated
    Prisoner's Dilemma via trigger strategies.
    \item Section~\ref{sec:tit-for-tat}: The tit-for-tat strategy.
    \item Section~\ref{sec:nash-folk}: The Nash folk theorem with full proof.
    \item Section~\ref{sec:spe-folk}: The subgame perfect folk theorem
    (Fudenberg--Maskin) with full proof.
    \item Section~\ref{sec:rl-connections}: Connections to reinforcement
    learning.
    \item Section~\ref{sec:exercises}: Exercises.
\end{itemize}

%% SECTION 2: FINITELY REPEATED GAMES

\section{Finitely Repeated Games}
\label{sec:finitely-repeated}

We begin with a \emph{stage game}
$G = (N, (A_i)_{i \in N}, (g_i)_{i \in N})$, where $N = \{1, \ldots, n\}$
is the player set, $A_i$ is the finite action set of player $i$, and
$g_i : A \to \R$ is the stage-game payoff function, with
$A = \prod_{i \in N} A_i$.

\begin{definition}[Finitely Repeated Game]
\label{def:finitely-repeated}
The \emph{$T$-fold repetition} of $G$, denoted $G^T$, is defined as
follows:
\begin{enumerate}[label=(\alph*)]
    \item In each period $t \in \{1, 2, \ldots, T\}$, all players
    simultaneously choose actions $a_i^t \in A_i$.
    \item At the end of period $t$, all players observe the realized action
    profile $a^t = (a_1^t, \ldots, a_n^t)$.
    \item The \emph{history} at period $t$ is
    $h^t = (a^1, a^2, \ldots, a^{t-1}) \in A^{t-1}$. The initial history
    is $h^1 = \emptyset$.
    \item A \emph{(pure) strategy} for player $i$ is a sequence of functions
    $s_i = (s_i^1, s_i^2, \ldots, s_i^T)$ where $s_i^t : A^{t-1} \to A_i$
    maps histories to actions.
    \item The payoff to player $i$ under strategy profile
    $s = (s_1, \ldots, s_n)$ is
    \[
        u_i(s) = \frac{1}{T} \sum_{t=1}^T g_i\bigl(a^t(s, h^t)\bigr),
    \]
    where $a^t(s, h^t) = \bigl(s_1^t(h^t), \ldots, s_n^t(h^t)\bigr)$ and
    histories are generated by the strategy profile.
\end{enumerate}
\end{definition}

\begin{remark}
We use average payoffs $\frac{1}{T}\sum_t g_i(a^t)$ rather than total
payoffs to keep repeated-game payoffs on the same scale as stage-game
payoffs, facilitating comparison with the infinite-horizon discounted case.
\end{remark}

\begin{definition}[SPE for Repeated Games]
\label{def:spe-repeated}
A strategy profile $s^*$ in $G^T$ is a \emph{subgame perfect equilibrium}
(SPE) if, for every period $t \in \{1, \ldots, T\}$ and every history
$h^t \in A^{t-1}$, the continuation strategy profile
$(s_i^{*,\tau}(h^t, \cdot))_{\tau \geq t}$ is a Nash equilibrium of the
subgame starting from history $h^t$.
\end{definition}

\subsection{The Unraveling Theorem}

The fundamental result for finitely repeated games with a unique stage-game
equilibrium is that repetition adds nothing: the unique SPE plays the
stage-game NE in every period.

\begin{keyresult}[title={Unraveling Theorem}]
\begin{theorem}[Unique Stage-Game NE Implies Unique SPE]
\label{thm:unraveling}
If the stage game $G$ has a unique Nash equilibrium
$a^* = (a_1^*, \ldots, a_n^*)$, then the $T$-fold repeated game $G^T$ has
a unique subgame perfect equilibrium in which $a^t = a^*$ for all
$t \in \{1, \ldots, T\}$.
\end{theorem}
\end{keyresult}

\begin{proof}
We proceed by backward induction on $t$.

\textbf{Base case ($t = T$).} In the final period, regardless of the
history $h^T$, the subgame consists of a single simultaneous play of the
stage game $G$. Since $a^*$ is the unique NE of $G$, every SPE must
prescribe $a^*$ in period $T$ after every history.

\textbf{Inductive step.} Suppose that in every SPE, $a^{\tau} = a^*$ for
all $\tau \in \{t+1, \ldots, T\}$, regardless of the history at period
$t+1$. Consider the subgame beginning at an arbitrary history $h^t$. The
continuation payoffs from period $t+1$ onward are fixed at
\[
    \frac{1}{T}\sum_{\tau=t+1}^{T} g_i(a^*)
    = \frac{T-t}{T}\, g_i(a^*),
\]
which is independent of the action chosen at period $t$, since all future
play is $a^*$ regardless of history. Therefore, player $i$'s total payoff
as a function of the period-$t$ action profile $a^t$ is
\[
    \frac{1}{T}\, g_i(a^t) + \frac{T-t}{T}\, g_i(a^*).
\]
Maximizing this over $a_i^t$ is equivalent to maximizing $g_i(a_i^t,
a_{-i}^t)$ in the stage game. Since $a^*$ is the unique NE of $G$, the
unique best response profile at period $t$ is $a^t = a^*$ in every SPE.

By induction, $a^t = a^*$ for all $t \in \{1, \ldots, T\}$. Since the
action in every period after every history is uniquely determined, the SPE
is unique.
\end{proof}

\begin{warning}[title={The Unraveling Trap}]
Theorem~\ref{thm:unraveling} shows that finite repetition of the Prisoner's
Dilemma---which has a unique NE at mutual defection---cannot sustain
cooperation under subgame perfection. The ``last period'' argument
propagates backward through every period. This motivates the study of
infinitely repeated games, where no last period exists.
\end{warning}

%% SECTION 3: MULTIPLE STAGE-GAME EQUILIBRIA

\section{New Equilibria with Multiple Stage-Game NE}
\label{sec:multiple-ne}

When the stage game has multiple Nash equilibria, repetition \emph{can}
create new equilibrium outcomes that are not stage-game NE. The key idea is
that different continuation equilibria can serve as rewards and punishments.

\begin{proposition}[Repetition Enables New Equilibria]
\label{prop:multiple-ne-repetition}
If the stage game $G$ has at least two Nash equilibria yielding different
payoffs for some player, then the two-fold repeated game $G^2$ may have SPE
outcomes in which first-period play is not a stage-game NE.
\end{proposition}

\begin{proof}
We give a constructive proof. Let $G$ have two Nash equilibria $a^*$ and
$\hat{a}$ with $g_i(a^*) > g_i(\hat{a})$ for all $i$ (i.e., $a^*$
Pareto-dominates $\hat{a}$). Suppose there exists an action profile
$\tilde{a} \notin \mathrm{NE}(G)$ with $g_i(\tilde{a}) > g_i(a^*)$ for
all $i$, and suppose the maximum unilateral gain from deviation satisfies
\begin{equation}\label{eq:deviation-bound}
    \max_{a_i' \in A_i} g_i(a_i', \tilde{a}_{-i}) - g_i(\tilde{a})
    < g_i(a^*) - g_i(\hat{a}).
\end{equation}
Consider the following strategy for each player $i$ in $G^2$:
\begin{itemize}
    \item \textbf{Period 1:} Play $\tilde{a}_i$.
    \item \textbf{Period 2:} If $a^1 = \tilde{a}$, play $a_i^*$.
    Otherwise, play $\hat{a}_i$.
\end{itemize}

In period 2, the prescribed play is always a stage-game NE ($a^*$ or
$\hat{a}$), so the strategy is sequentially rational in every period-2
subgame.

In period 1, if player $i$ deviates to $a_i'$ while others play
$\tilde{a}_{-i}$, the payoff change is:
\begin{align*}
    &\frac{1}{2}\bigl[g_i(a_i', \tilde{a}_{-i}) + g_i(\hat{a})\bigr]
    - \frac{1}{2}\bigl[g_i(\tilde{a}) + g_i(a^*)\bigr] \\
    &= \frac{1}{2}\bigl[g_i(a_i', \tilde{a}_{-i}) - g_i(\tilde{a})\bigr]
    - \frac{1}{2}\bigl[g_i(a^*) - g_i(\hat{a})\bigr] < 0,
\end{align*}
where the inequality follows from~\eqref{eq:deviation-bound}. Hence no
player has a profitable deviation, and the profile is an SPE in which
first-period play $\tilde{a}$ is not a stage-game NE.
\end{proof}

\begin{example}[Coordination Game]
\label{ex:coordination-repeated}
Consider the two-player coordination game with payoffs:
\[
\begin{array}{c|cc}
  & L & R \\ \hline
T & 3,3 & 0,0 \\
B & 0,0 & 1,1
\end{array}
\]
The game has two pure NE: $(T,L)$ with payoffs $(3,3)$ and $(B,R)$ with
payoffs $(1,1)$. Suppose we want to sustain the non-NE outcome $(T,R)$ with
payoffs $(0,0)$ in the first period. This fails because the ``reward''
payoff difference $g_i(T,L) - g_i(B,R) = 2$ would need to exceed the
deviation gain, but $g_i(T,R) = 0$ is already the minimum, so no player
gains from unilateral deviation at $(T,R)$... except player 1 prefers $B$
giving $0$ anyway. The construction requires profiles where deviation
is \emph{tempting}, which depends on the specific game structure.

A more instructive application: suppose we add an action $M$ for each
player with payoffs $g(M,M) = (2.5, 2.5)$, $g(M,L) = g(T,M) = (-1,-1)$,
and other entries as before. Then $(M,M)$ Pareto-dominates both NE? No,
$(T,L)$ gives $(3,3) > (2.5, 2.5)$. In any case, the key principle is that
the gap between the ``reward'' and ``punishment'' equilibria creates room
for enforcing non-NE play.
\end{example}

%% SECTION 4: INFINITELY REPEATED GAMES

\section{Infinitely Repeated Games}
\label{sec:infinitely-repeated}

To circumvent the unraveling argument, we consider infinite repetition
with discounting.

\begin{definition}[Infinitely Repeated Game]
\label{def:infinitely-repeated}
Given a stage game $G$ and a discount factor $\delta \in (0,1)$, the
\emph{infinitely repeated game} $G^\infty(\delta)$ is defined by:
\begin{enumerate}[label=(\alph*)]
    \item In each period $t \in \{1, 2, 3, \ldots\}$, players simultaneously
    choose actions $a^t \in A$.
    \item The history at period $t$ is
    $h^t = (a^1, \ldots, a^{t-1}) \in A^{t-1}$, with $h^1 = \emptyset$.
    A strategy for player $i$ is a function
    $s_i = (s_i^t)_{t=1}^\infty$ with $s_i^t : A^{t-1} \to A_i$.
    \item Player $i$'s payoff under strategy profile $s$ is the
    \emph{normalized discounted sum}:
    \[
        u_i(s) = (1 - \delta) \sum_{t=1}^{\infty} \delta^{t-1}
        g_i(a^t).
    \]
\end{enumerate}
The normalization factor $(1-\delta)$ ensures that payoffs lie in the same
range as stage-game payoffs: if $a^t = a$ for all $t$, then $u_i = g_i(a)$.
\end{definition}

\begin{remark}
Every history $h^t$ begins a subgame that is strategically identical to
$G^\infty(\delta)$ (the continuation game has the same structure and the
same discount factor). This self-similar structure is central to the
analysis of trigger strategies.
\end{remark}

%% SECTION 5: FEASIBLE AND INDIVIDUALLY RATIONAL PAYOFFS

\section{Feasible and Individually Rational Payoffs}
\label{sec:feasible-ir}

\begin{definition}[Feasible Payoff Set]
\label{def:feasible-payoffs}
The \emph{feasible payoff set} of the stage game $G$ is
\[
    V = \conv\bigl\{g(a) : a \in A\bigr\}
    = \conv\bigl\{(g_1(a), \ldots, g_n(a)) : a \in A\bigr\}
    \subset \R^n,
\]
the convex hull of all achievable stage-game payoff vectors.
\end{definition}

\begin{remark}
In the infinitely repeated game, any $v \in V$ is achievable by cycling
through action profiles with appropriate frequencies. If
$v = \sum_{k=1}^K \lambda_k\, g(a^{(k)})$ with $\lambda_k \geq 0$ and
$\sum_k \lambda_k = 1$, then playing $a^{(k)}$ in a periodic cycle with
frequency $\lambda_k$ yields average payoff $v$. As $\delta \to 1$, the
normalized discounted payoff converges to this time-average.
\end{remark}

\begin{definition}[Minimax Value and Individually Rational Payoffs]
\label{def:minimax-ir}
The \emph{minimax value} of player $i$ in pure strategies is
\[
    \underline{v}_i = \min_{a_{-i} \in A_{-i}}
    \max_{a_i \in A_i} g_i(a_i, a_{-i}).
\]
The \emph{mixed-strategy minimax value} is
\[
    \underline{v}_i^{\mathrm{mix}} = \min_{\alpha_{-i} \in \Delta(A_{-i})}
    \max_{a_i \in A_i}
    \E_{\alpha_{-i}}[g_i(a_i, a_{-i})],
\]
where $\Delta(A_{-i}) = \prod_{j \neq i} \Delta(A_j)$. A payoff vector
$v \in \R^n$ is \emph{individually rational} if
$v_i \geq \underline{v}_i^{\mathrm{mix}}$ for all $i \in N$.
\end{definition}

The minimax value is the lowest payoff that the opponents can force upon
player $i$, even when $i$ best-responds. No equilibrium can give $i$ less
than this amount.

\begin{proposition}[NE Payoff Lower Bound]
\label{prop:ne-payoff-lower-bound}
In any Nash equilibrium of $G^\infty(\delta)$, each player $i$ receives
payoff $u_i \geq \underline{v}_i^{\mathrm{mix}}$.
\end{proposition}

\begin{proof}
In any NE $\sigma$, player $i$ can deviate to a strategy that, in each
period $t$, plays
$\argmax_{a_i \in A_i} \E_{\sigma_{-i}^t}[g_i(a_i, a_{-i}^t)]$, where
$\sigma_{-i}^t$ is the marginal distribution over opponents' period-$t$
actions induced by the equilibrium strategy. This best-response strategy
guarantees a stage payoff of at least $\underline{v}_i^{\mathrm{mix}}$ in
every period (since $i$ can guarantee this value against \emph{any}
opponents' mixed action profile, and the equilibrium marginal is one such
profile). Hence the deviation yields total payoff at least
$\underline{v}_i^{\mathrm{mix}}$. Since $\sigma$ is a NE, the equilibrium
payoff must be at least as high as any deviation payoff:
$u_i(\sigma) \geq \underline{v}_i^{\mathrm{mix}}$.
\end{proof}

\begin{definition}[Full Dimensionality Condition]
\label{def:full-dim}
The stage game $G$ satisfies the \emph{full dimensionality condition} if the
feasible payoff set $V = \conv\{g(a) : a \in A\}$ has dimension $n$ (the
number of players), i.e., $V$ has nonempty interior in $\R^n$.
\end{definition}

\begin{remark}
The full dimensionality condition is equivalent to requiring that there
exist $n+1$ action profiles $a^{(0)}, a^{(1)}, \ldots, a^{(n)} \in A$
such that the vectors $g(a^{(1)}) - g(a^{(0)}), \ldots,
g(a^{(n)}) - g(a^{(0)})$ are linearly independent in $\R^n$. This
condition is generic (it holds for ``almost all'' games) but fails in
certain symmetric or degenerate games. For two-player games, it simply
requires that the feasible payoff set is not contained in a line.
\end{remark}

%% SECTION 6: COOPERATION IN REPEATED PD

\section{Cooperation in the Repeated Prisoner's Dilemma}
\label{sec:cooperation-pd}

\begin{example}[The Prisoner's Dilemma]
\label{ex:pd}
Consider the standard Prisoner's Dilemma with payoffs:
\[
\begin{array}{c|cc}
  & C & D \\ \hline
C & 3,3 & 0,4 \\
D & 4,0 & 1,1
\end{array}
\]
The unique stage-game NE is $(D,D)$ with payoffs $(1,1)$. The minimax value
for each player is $\underline{v}_i = 1$ (since mutual defection already
holds each player to their minimax). The Pareto-optimal outcome is $(C,C)$
with payoffs $(3,3)$.
\end{example}

\begin{definition}[Grim Trigger Strategy]
\label{def:grim-trigger}
Given a target action profile $a^c \in A$, the \emph{grim trigger strategy}
for player $i$ is:
\[
    s_i^t(h^t) = \begin{cases}
        a_i^c & \text{if } a^{\tau} = a^c
            \text{ for all } \tau < t, \\
        m_i   & \text{otherwise},
    \end{cases}
\]
where $m_i$ is player $i$'s component of a minimax punishment profile. The
strategy cooperates as long as no deviation has ever occurred, and punishes
by minimaxing forever after any deviation.
\end{definition}

\begin{proposition}[Grim Trigger Sustains Cooperation]
\label{prop:grim-trigger-pd}
In the infinitely repeated Prisoner's Dilemma
(Example~\ref{ex:pd}) with discount factor $\delta$, the grim trigger
strategy profile with $a^c = (C,C)$ and punishment $(D,D)$ is a subgame
perfect equilibrium if and only if $\delta \geq 1/3$.
\end{proposition}

\begin{proof}
We verify the one-shot deviation principle: in any subgame, no player
benefits from a single-period deviation followed by reversion to the
prescribed strategy.

\textbf{Cooperation subgame} (no deviation has occurred). If player $i$
cooperates, the continuation payoff is $3$ (cooperation forever). If
player $i$ deviates to $D$ in the current period, the payoff is:
\[
    u_i^{\mathrm{dev}} = (1-\delta) \cdot 4 + \delta \cdot 1
    = 4 - 3\delta,
\]
since the opponent cooperates this period (giving $i$ a stage payoff of
$4$), but grim trigger punishment begins next period, yielding $1$ forever.
Cooperation is preferred when $3 \geq 4 - 3\delta$, i.e.,
$\delta \geq 1/3$.

\textbf{Punishment subgame} (a deviation has occurred). Both players play
$(D,D)$ forever. Since $(D,D)$ is a stage-game NE, no player can
profitably deviate in any single period: deviating to $C$ gives
$(1-\delta) \cdot 0 + \delta \cdot 1 = \delta < 1$, while staying with
$D$ gives $1$.

Since no profitable one-shot deviation exists in any subgame when
$\delta \geq 1/3$, the grim trigger profile is an SPE. Conversely, if
$\delta < 1/3$, the deviation in the cooperation phase is profitable, so
grim trigger is not an NE (and hence not an SPE).
\end{proof}

\begin{rlconnection}[title={Discount Factors in RL}]
The discount factor $\delta$ in repeated games is analogous to the discount
factor $\gamma$ in MDPs. If $\gamma$ is too small (agents are too myopic),
cooperative equilibria cannot be sustained. This directly parallels the
threshold $\delta \geq 1/3$: myopic agents with small $\gamma$ will
inevitably learn to defect.
\end{rlconnection}

%% SECTION 7: TIT-FOR-TAT

\section{Tit-for-Tat}
\label{sec:tit-for-tat}

\begin{definition}[Tit-for-Tat]
\label{def:tit-for-tat}
In a two-player game, the \emph{tit-for-tat} strategy for player $i$ is:
\[
    s_i^1 = C, \qquad s_i^t(h^t) = a_j^{t-1} \quad \text{for } t \geq 2,
\]
where $j \neq i$. That is, start by cooperating, then replicate the
opponent's previous action.
\end{definition}

\begin{proposition}[Tit-for-Tat as NE in Repeated PD]
\label{prop:tft-ne}
In the infinitely repeated Prisoner's Dilemma
(Example~\ref{ex:pd}), the tit-for-tat profile is a Nash equilibrium for
$\delta \geq 1/3$.
\end{proposition}

\begin{proof}
Under the tit-for-tat profile, both players cooperate in every period,
yielding payoff $3$ for each. Consider player $i$'s best deviation.

\textbf{One-period deviation at time $t$.} If $i$ defects at time $t$
while $j$ cooperates (since $j$ copies $i$'s previous action, which was
$C$), player $i$ receives $4$ in period $t$. In period $t+1$, player $j$
defects (copying $i$'s defection). If $i$ returns to cooperation in period
$t+1$, $i$ receives $0$ (cooperating against $j$'s defection). In period
$t+2$, $j$ cooperates again (copying $i$'s cooperation in $t+1$),
restoring mutual cooperation.

The discounted payoff from this deviation path, starting from period $t$,
is:
\[
    (1-\delta)\bigl[4 + \delta \cdot 0 + \delta^2 \cdot 3
    + \delta^3 \cdot 3 + \cdots\bigr]
    = (1-\delta)\Bigl[4 + \frac{3\delta^2}{1-\delta}\Bigr]
    = 4(1-\delta) + 3\delta^2.
\]
Cooperation gives payoff $3$. The deviation is unprofitable when
$3 \geq 4(1-\delta) + 3\delta^2$, i.e.,
$3\delta^2 - 4\delta + 1 \leq 0$, i.e.,
$(3\delta - 1)(\delta - 1) \leq 0$. Since $\delta < 1$, this holds iff
$\delta \geq 1/3$.

\textbf{Permanent defection.} If $i$ defects forever starting at $t$, the
payoff is $(1-\delta)[4 + \delta \cdot 1 + \delta^2 \cdot 1 + \cdots]
= 4(1-\delta) + \delta = 4 - 3\delta$, and $3 \geq 4 - 3\delta$ iff
$\delta \geq 1/3$.

More generally, any deviation sequence can be analyzed period by period.
The key observation is that each defection yields at most a one-period gain
of $1$ (getting $4$ instead of $3$) followed by a one-period loss of at
least $3$ (getting $0$ instead of $3$ when the opponent retaliates), and for
$\delta \geq 1/3$ the discounted loss exceeds the gain.
\end{proof}

\begin{remark}
Unlike grim trigger, tit-for-tat is \emph{forgiving}: punishment lasts only
one period. However, tit-for-tat is generally \emph{not} subgame perfect in
the Prisoner's Dilemma. In the subgame following a deviation (where one
player defected and the other cooperated), tit-for-tat prescribes
alternating retaliation. In such a subgame, a one-shot deviation to
cooperation by the ``retaliating'' player would restore mutual cooperation
sooner, which can be profitable. Nevertheless, tit-for-tat's simplicity and
robustness have made it influential in evolutionary game theory and MARL
\citep{shoham2008}.
\end{remark}

%% SECTION 8: NASH FOLK THEOREM

\section{Nash Folk Theorem}
\label{sec:nash-folk}

The folk theorems characterize the set of equilibrium payoffs in infinitely
repeated games. The Nash folk theorem states that any feasible, strictly
individually rational payoff vector can be achieved in a Nash equilibrium
for patient enough players.

\begin{keyresult}[title={Nash Folk Theorem}]
\begin{theorem}[Nash Folk Theorem]
\label{thm:nash-folk}
Let $G$ be a finite normal-form game with feasible payoff set
$V = \conv\{g(a) : a \in A\}$. For any payoff vector $v \in V$ satisfying
$v_i > \underline{v}_i^{\mathrm{mix}}$ for all $i$ (strictly individually
rational), there exists $\underline{\delta} \in (0,1)$ such that for all
$\delta \in (\underline{\delta}, 1)$, there is a Nash equilibrium of
$G^\infty(\delta)$ with payoff vector $v$.
\end{theorem}
\end{keyresult}

\begin{proof}
The proof constructs a Nash equilibrium strategy profile that yields payoff
$v$ and deters deviations via permanent minimax punishment.

\textbf{Step 1: Achieving $v$ through action sequences.}
Since $v \in V = \conv\{g(a) : a \in A\}$, there exist action profiles
$a^{(1)}, \ldots, a^{(K)} \in A$ and weights
$\lambda_1, \ldots, \lambda_K \geq 0$ with $\sum_{k=1}^K \lambda_k = 1$
such that $v = \sum_{k=1}^K \lambda_k\, g(a^{(k)})$.

Construct a periodic cycle of length $L$: action profile $a^{(k)}$ is
played for $L_k = \lfloor \lambda_k L \rfloor$ periods (adjusting one
$L_k$ so that $\sum_k L_k = L$). Define the action sequence
$(a^t)_{t=1}^\infty$ by repeating this cycle. The time-average payoff of
this sequence is $\bar{v}(L) = \frac{1}{L}\sum_{k} L_k\, g(a^{(k)})$,
which satisfies $\|\bar{v}(L) - v\| \leq C/L$ for a constant $C$
depending only on the game. Fix $\varepsilon > 0$ small enough that
$v_i - \varepsilon > \underline{v}_i^{\mathrm{mix}}$ for all $i$, and
choose $L$ large enough that $\|\bar{v}(L) - v\| < \varepsilon/2$.

As $\delta \to 1$, the normalized discounted payoff of any periodic
sequence converges to its time-average (since
$(1-\delta)\sum_{t=1}^\infty \delta^{t-1} f(t)$ converges to
$\frac{1}{L}\sum_{t=1}^L f(t)$ for any $L$-periodic sequence $f$). Choose
$\underline{\delta}_1 \in (0,1)$ such that for $\delta >
\underline{\delta}_1$, the discounted payoff is within $\varepsilon/2$ of
the time-average. Then the discounted payoff is within $\varepsilon$ of $v$.

\textbf{Step 2: Punishment strategies.}
For each player $i$, let $\alpha_{-i}^{m,i} \in \Delta(A_{-i})$ be a
mixed minimax strategy profile:
$\max_{a_i} \E_{\alpha_{-i}^{m,i}}[g_i(a_i, a_{-i})]
= \underline{v}_i^{\mathrm{mix}}$.

\textbf{Step 3: Strategy construction.}
Define the following strategy for each player $j$:
\begin{itemize}
    \item \emph{Cooperation phase:} Play the periodic sequence
    $(a^t)_{t=1}^\infty$ from Step~1.
    \item \emph{Punishment phase:} If some player $i$ is the first to
    deviate from the prescribed sequence at period $\hat{t}$, then for all
    $t > \hat{t}$, all players $j \neq i$ play according to
    $\alpha_{-i}^{m,i}$ (punishing $i$ forever). Player $i$ plays any
    best response.
\end{itemize}

\textbf{Step 4: Verifying the Nash equilibrium condition.}
If player $i$ follows the prescribed strategy, she receives a payoff
within $\varepsilon$ of $v_i$ (for $\delta > \underline{\delta}_1$), hence
at least $v_i - \varepsilon > \underline{v}_i^{\mathrm{mix}}$.

If player $i$ deviates at period $\hat{t}$, she receives at most
$M = \max_{a,i} |g_i(a)|$ for finitely many periods (the deviation
period), then enters permanent punishment yielding at most
$\underline{v}_i^{\mathrm{mix}}$ per period. The deviation payoff is
bounded by:
\[
    u_i^{\mathrm{dev}} \leq (1-\delta)\sum_{\tau=1}^{\hat{t}}
    \delta^{\tau-1} M + \delta^{\hat{t}}\,
    \underline{v}_i^{\mathrm{mix}}
    \leq (1-\delta^{\hat{t}}) M + \delta^{\hat{t}}\,
    \underline{v}_i^{\mathrm{mix}}.
\]
The worst case for the deviator is $\hat{t} = 1$:
$u_i^{\mathrm{dev}} \leq (1-\delta) M + \delta\,
\underline{v}_i^{\mathrm{mix}}$. As $\delta \to 1$, this converges to
$\underline{v}_i^{\mathrm{mix}} < v_i - \varepsilon$. Hence there exists
$\underline{\delta}_2 \in (0,1)$ such that for
$\delta > \underline{\delta}_2$, the deviation payoff is strictly below
the equilibrium payoff.

Setting $\underline{\delta} = \max(\underline{\delta}_1,
\underline{\delta}_2)$ completes the proof.
\end{proof}

\begin{remark}
The equilibrium constructed above is Nash but \emph{not} necessarily
subgame perfect. The punishing players carry out permanent minimax
punishment, but they may have no incentive to do so: the punishment
strategy may not be a best response for the punishers in the punishment
phase. For example, if minimaxing player $i$ requires player $j$ to play
an action that gives $j$ a low stage payoff, $j$ might prefer to deviate
from the punishment. The subgame perfect folk theorem addresses this
shortcoming.
\end{remark}

%% SECTION 9: SUBGAME PERFECT FOLK THEOREM

\section{Subgame Perfect Folk Theorem}
\label{sec:spe-folk}

\begin{keyresult}[title={Subgame Perfect Folk Theorem}]
\begin{theorem}[Subgame Perfect Folk Theorem
    {\citep{fudenberg1986}}]
\label{thm:spe-folk}
Let $G$ be a finite $n$-player game satisfying the full dimensionality
condition (Definition~\ref{def:full-dim}). For any feasible payoff vector
$v \in V$ with $v_i > \underline{v}_i^{\mathrm{mix}}$ for all $i$, there
exists $\underline{\delta} \in (0,1)$ such that for all
$\delta \in (\underline{\delta}, 1)$, there is a subgame perfect
equilibrium of $G^\infty(\delta)$ with payoff vector $v$.
\end{theorem}
\end{keyresult}

\begin{proof}
The proof is constructive. We design a strategy profile with multiple
phases---cooperation and player-specific punishment---arranged so that
every phase is self-enforcing.

\textbf{Step 1: Choosing reward payoffs.}
Fix $v \in V$ with $v_i > \underline{v}_i^{\mathrm{mix}}$ for all $i$.
Since $V$ has full dimension (nonempty interior in $\R^n$), for each
player $i$ there exists a feasible payoff vector $w^i \in V$ satisfying:
\begin{enumerate}[label=(\roman*)]
    \item $w^i_j > \underline{v}_j^{\mathrm{mix}}$ for all $j \neq i$
    (punishers receive strictly above their minimax values during the
    punishment of $i$);
    \item $w^i_i < v_i$ (player $i$'s payoff during her own punishment
    phase is below her cooperative payoff).
\end{enumerate}
The full dimensionality condition guarantees such vectors exist: since $V$
has nonempty interior, we can perturb any interior point of $V$ to
independently adjust each coordinate. Specifically, let $v^0$ be an
interior point of $V$ with $v^0_j > \underline{v}_j^{\mathrm{mix}}$ for
all $j$. For each $i$, define $w^i = v^0 - \eta e_i$, where $e_i$ is the
$i$-th standard basis vector and $\eta > 0$ is small enough that
$w^i \in V$, $w^i_i > \underline{v}_i^{\mathrm{mix}}$, and $w^i_i < v_i$.
This is possible because $v^0$ is interior and
$v_i > \underline{v}_i^{\mathrm{mix}}$.

For each $i$, let $m^i_{-i} \in \Delta(A_{-i})$ be a mixed minimax
strategy profile against player $i$, and let $m^i_i$ be any best response
by $i$ to $m^i_{-i}$.

\textbf{Step 2: Strategy construction (three-phase automaton).}

\emph{Phase~0 (Cooperation):} Play the action sequence yielding payoff $v$
(constructed as in the Nash folk theorem proof, Step~1).

\emph{Phase~$\mathrm{I}_i$ (Punishment of player $i$):} Lasts for $L$
periods (to be determined). During this phase, the opponents of $i$ play
the minimax strategy $m^i_{-i}$ each period, and player $i$ plays any best
response $m^i_i$ (yielding $i$ at most
$\underline{v}_i^{\mathrm{mix}}$ per period).

\emph{Phase~$\mathrm{II}_i$ (Reward after punishing $i$):} After the
$L$-period punishment of $i$, play switches to the action sequence yielding
payoff $w^i$ forever.

\emph{Transitions:}
\begin{itemize}
    \item If player $j$ is the first to deviate from the prescribed actions
    in Phase~0, transition to Phase~$\mathrm{I}_j$.
    \item If player $j \neq i$ deviates during Phase~$\mathrm{I}_i$,
    transition to Phase~$\mathrm{I}_j$ (punish the deviating punisher).
    \item If player $i$ deviates during her own punishment
    Phase~$\mathrm{I}_i$, restart Phase~$\mathrm{I}_i$ from the
    beginning.
    \item If any player $j$ deviates during Phase~$\mathrm{II}_i$,
    transition to Phase~$\mathrm{I}_j$.
\end{itemize}

\textbf{Step 3: Choosing the punishment length $L$.}
Let $M = \max_{a,i} |g_i(a)|$ be a bound on stage-game payoffs. We need
$L$ large enough that the punishment phase is a genuine deterrent. The
continuation payoff for player $i$ at the start of Phase~$\mathrm{I}_i$ is:
\[
    p_i(\delta, L) = (1 - \delta^L)\, \underline{v}_i^{\mathrm{mix}}
    + \delta^L\, w^i_i.
\]
Since $\underline{v}_i^{\mathrm{mix}} \leq w^i_i < v_i$, we have
$p_i(\delta, L) \leq w^i_i < v_i$ for all $L$ and $\delta$. Moreover, as
$L \to \infty$ with $\delta$ fixed, $p_i(\delta, L) \to
\underline{v}_i^{\mathrm{mix}}$. We can choose $L$ large enough that the
punishment is sufficiently severe.

\textbf{Step 4: Verifying sequential rationality.}

We apply the one-shot deviation principle: a strategy profile in a
discounted repeated game is an SPE if and only if no player has a
profitable single-period deviation at any history
\citep{fudenberg1991}.

\emph{Phase~0 -- Player $i$ considers deviating.}
Compliance yields payoff (within $\varepsilon$ of) $v_i$. A one-shot
deviation yields at most $M$ in the current period, followed by
Phase~$\mathrm{I}_i$:
\[
    u_i^{\mathrm{dev}} = (1-\delta)\, M + \delta\, p_i(\delta, L).
\]
As $\delta \to 1$, $(1-\delta)M \to 0$ and $\delta\, p_i(\delta, L) \to
p_i(1, L) = w^i_i < v_i$. Hence for $\delta$ close enough to $1$,
$u_i^{\mathrm{dev}} < v_i$ and deviation is unprofitable.

\emph{Phase~$\mathrm{I}_i$ -- Punisher $j \neq i$ considers deviating.}
Compliance means playing the minimax action for the remaining
$L' \leq L$ periods of Phase~$\mathrm{I}_i$, then entering
Phase~$\mathrm{II}_i$ with payoff $w^i_j >
\underline{v}_j^{\mathrm{mix}}$. A one-shot deviation yields at most $M$
in one period, then Phase~$\mathrm{I}_j$ begins:
\[
    u_j^{\mathrm{dev}} = (1-\delta)\, M + \delta\, p_j(\delta, L).
\]
The compliance payoff from the continuation of Phase~$\mathrm{I}_i$ plus
Phase~$\mathrm{II}_i$ converges to $w^i_j$ as $\delta \to 1$ (the
finite punishment phase becomes negligible). Since
$w^i_j > \underline{v}_j^{\mathrm{mix}} \geq p_j(\delta, L)$ (the
latter being a weighted average of $\underline{v}_j^{\mathrm{mix}}$ and
$w^j_j$, both at most close to $\underline{v}_j^{\mathrm{mix}}$ for
large $L$), compliance dominates deviation for $\delta$ large enough.

More precisely: player $j$'s compliance continuation payoff from
Phase~$\mathrm{I}_i$ with $L'$ periods remaining is
\[
    c_j = (1 - \delta^{L'})\, g_j^{\mathrm{I}_i} + \delta^{L'}\, w^i_j,
\]
where $g_j^{\mathrm{I}_i}$ is $j$'s per-period payoff during
Phase~$\mathrm{I}_i$ (from playing the minimax strategy against $i$).
As $\delta \to 1$, $c_j \to w^i_j >
\underline{v}_j^{\mathrm{mix}}$. The deviation payoff converges to
$p_j(1, L)$, which can be made close to
$\underline{v}_j^{\mathrm{mix}}$ by choosing $L$ large. Hence
compliance is strictly preferred for large $\delta$.

\emph{Phase~$\mathrm{I}_i$ -- Player $i$ considers deviating.}
Deviation restarts Phase~$\mathrm{I}_i$, gaining at most $M$ in one
period but adding $L$ more punishment periods. Compliance leads to
Phase~$\mathrm{II}_i$ after $L'$ periods. The compliance continuation
payoff is $(1 - \delta^{L'})\,\underline{v}_i^{\mathrm{mix}} +
\delta^{L'}\, w^i_i$. The deviation payoff is
$(1-\delta)M + \delta\, p_i(\delta, L)$. For large $\delta$, the
compliance payoff exceeds the deviation payoff because compliance
leads to Phase~$\mathrm{II}_i$ sooner (after $L'$ rather than $L$
additional periods).

\emph{Phase~$\mathrm{II}_i$ -- Any player $j$ considers deviating.}
This is identical to the Phase~0 analysis with $w^i$ replacing $v$. Since
$w^i_j > \underline{v}_j^{\mathrm{mix}}$ for all $j$, the argument from
Phase~0 applies.

\textbf{Step 5: Combining.}
For each of the finitely many deviation scenarios above, there is a
threshold $\delta_k \in (0,1)$ above which deviation is unprofitable.
Setting $\underline{\delta} = \max_k \delta_k$ completes the proof.
\end{proof}

\begin{remark}[Two-Player Games]
For two-player games, the full dimensionality condition is not needed: the
subgame perfect folk theorem holds for any two-player game
\citep{fudenberg1986}. The reason is that in two-player games, the only
punisher is the single opponent, and minimaxing the deviator is
automatically in the punisher's interest (or at least does not require
special incentivization) because Phase~$\mathrm{II}_i$ can reward the
punisher without needing to independently adjust a third player's payoff.
\end{remark}

\begin{remark}[Necessity of Full Dimensionality for $n \geq 3$]
\label{rem:full-dim-necessity}
For three or more players, the full dimensionality condition is essential.
Without it, the feasible payoff set lies in a lower-dimensional subspace,
and one cannot independently reward punishers while punishing the deviator.
\citet{fudenberg1986} provide a counterexample: a three-player game
violating the condition where certain feasible, strictly individually
rational payoffs cannot be sustained in any SPE for any $\delta$.
\end{remark}

%% SECTION 10: RL CONNECTIONS

\section{Connections to Reinforcement Learning}
\label{sec:rl-connections}

\begin{rlconnection}[title={Self-Play and Emergent Cooperation}]
In multi-agent RL, agents trained via self-play in iterated games
effectively search over strategies in a repeated game. The folk theorems
predict that:
\begin{enumerate}[label=(\roman*)]
    \item \textbf{Many equilibria exist.} For patient agents (large
    $\gamma$), the set of possible converged outcomes is enormous. Small
    changes in learning rates, initialization, or exploration schedules can
    lead to qualitatively different equilibria.
    \item \textbf{Cooperation can emerge.} Even in social dilemmas (e.g.,
    iterated Prisoner's Dilemma), self-play can discover cooperative
    strategies resembling grim trigger or tit-for-tat, provided the
    discount factor is large enough.
    \item \textbf{Punishment structures appear naturally.} Trained agents
    often develop implicit punishment mechanisms: they cooperate when the
    opponent cooperates and retaliate after defection, mirroring the
    trigger strategies in the folk theorem proofs.
\end{enumerate}
\end{rlconnection}

\begin{rlconnection}[title={Opponent Modeling}]
The folk theorem strategies require each player to condition on the
\emph{entire history} of past play. In RL, this corresponds to:
\begin{itemize}
    \item \textbf{Recurrent architectures} (LSTMs, GRUs) that maintain a
    hidden state summarizing interaction history.
    \item \textbf{Opponent modeling}: explicitly building a model of the
    opponent's strategy from observed actions, analogous to detecting
    whether the opponent is ``cooperating'' or ``punishing'' in the folk
    theorem construction.
    \item \textbf{Theory of mind}: higher-order reasoning about what the
    opponent believes about one's own strategy, corresponding to the
    recursive reasoning in SPE verification.
\end{itemize}
\end{rlconnection}

\begin{rlconnection}[title={The Discount Factor and Cooperation}]
The threshold $\underline{\delta}$ in the folk theorems has a direct analog
in MARL. Empirically, increasing the discount factor $\gamma$ in multi-agent
training makes cooperative outcomes more likely, consistent with the
theoretical prediction. Conversely, highly myopic agents ($\gamma$ near $0$)
behave as if playing a one-shot game, converging to the stage-game NE. This
provides a principled way to influence emergent behavior through
hyperparameter selection \citep{shoham2008}.
\end{rlconnection}

%% SECTION 11: EXERCISES

\section{Exercises}
\label{sec:exercises}

\begin{exercise}[Unraveling in a Specific Game]
\label{ex:unraveling}
Consider the stage game with payoffs:
\[
\begin{array}{c|cc}
  & L & R \\ \hline
T & 2,2 & 0,3 \\
B & 3,0 & 1,1
\end{array}
\]
\begin{enumerate}[label=(\alph*)]
    \item Verify that $(B,R)$ is the unique Nash equilibrium.
    \item Apply the unraveling theorem to determine the unique SPE of the
    $100$-fold repetition $G^{100}$.
    \item Compute the SPE payoff for each player and compare it with the
    Pareto-optimal outcome $(T,L)$.
    \item Explain intuitively why the backward induction argument fails if
    the game is repeated infinitely.
\end{enumerate}
\end{exercise}

\begin{exercise}[Grim Trigger Threshold]
\label{ex:grim-threshold}
Consider a symmetric Prisoner's Dilemma with general payoffs:
\[
\begin{array}{c|cc}
  & C & D \\ \hline
C & b,b & 0, b+c \\
D & b+c, 0 & c, c
\end{array}
\]
where $b > c > 0$ (so mutual cooperation is preferred, but defection is
individually tempting).
\begin{enumerate}[label=(\alph*)]
    \item Compute the minimax value for each player.
    \item Derive the minimum discount factor $\underline{\delta}$ for grim
    trigger to sustain $(C,C)$ as an SPE.
    \item Show that $\underline{\delta} \to 1$ as $c \to b$, and interpret
    this result.
    \item For $b = 5, c = 2$, compute the numerical threshold.
\end{enumerate}
\end{exercise}

\begin{exercise}[Tit-for-Tat is Not SPE]
\label{ex:tft-not-spe}
Using the Prisoner's Dilemma of Example~\ref{ex:pd}:
\begin{enumerate}[label=(\alph*)]
    \item Consider the subgame reached after the history $(C, D)$ (player~1
    cooperated, player~2 defected in period~1). Describe the play under
    tit-for-tat from this point.
    \item Compute the payoffs for both players under tit-for-tat in this
    subgame.
    \item Show that player~2 can profitably deviate in this subgame
    (for suitable $\delta$), proving that tit-for-tat is not SPE.
    \item Explain the conceptual difference between NE and SPE in this
    context.
\end{enumerate}
\end{exercise}

\begin{exercise}[Feasible Set Computation]
\label{ex:feasible-set}
Consider the three-action game:
\[
\begin{array}{c|ccc}
  & L & C & R \\ \hline
T & 4,0 & 0,4 & 1,1 \\
M & 0,0 & 3,3 & 0,0 \\
B & 1,1 & 0,0 & 2,2
\end{array}
\]
\begin{enumerate}[label=(\alph*)]
    \item List all nine pure action payoff vectors.
    \item Sketch the feasible payoff set
    $V = \conv\{g(a) : a \in A\}$ in $\R^2$.
    \item Compute the mixed-strategy minimax value for each player.
    \item Identify the set of feasible, strictly individually rational
    payoff vectors.
    \item Does the full dimensionality condition hold? Why or why not?
\end{enumerate}
\end{exercise}

\begin{exercise}[Beyond Grim Trigger: Limited Punishment]
\label{ex:limited-punishment}
In the repeated Prisoner's Dilemma (Example~\ref{ex:pd}), consider a
\emph{$K$-period punishment} strategy: cooperate; upon any deviation,
defect for exactly $K$ periods, then return to cooperation.
\begin{enumerate}[label=(\alph*)]
    \item Compute the minimum discount factor $\underline{\delta}(K)$ for
    this strategy to be an SPE, as a function of $K$.
    \item Show that $\underline{\delta}(K) \to 1/3$ as $K \to \infty$
    (recovering the grim trigger threshold).
    \item Compare $\underline{\delta}(1)$ (single-period punishment) with
    $\underline{\delta}(\infty) = 1/3$. What is the cost of forgiveness?
    \item Discuss the trade-off between punishment length and the required
    patience level.
\end{enumerate}
\end{exercise}

\begin{exercise}[Folk Theorem Application]
\label{ex:folk-application}
In the game of Exercise~\ref{ex:feasible-set}:
\begin{enumerate}[label=(\alph*)]
    \item Using the Nash folk theorem, argue that the payoff vector $(3,3)$
    (from the action profile $(M,C)$) is achievable as a NE payoff for
    large $\delta$, even if $(M,C)$ is not a stage-game NE.
    \item Construct an explicit grim trigger strategy profile that yields
    payoff close to $(3,3)$.
    \item Argue that the payoff $(2,2)$ is also achievable as an SPE payoff
    for large $\delta$, using the subgame perfect folk theorem.
    \item Is the payoff $(0.5, 0.5)$ achievable as an equilibrium payoff?
    Justify your answer using the minimax lower bound.
\end{enumerate}
\end{exercise}

\bigskip

\noindent\textbf{Summary.} This lesson developed: (1) the finitely repeated
game model with the unraveling theorem showing that unique stage-game NE
leads to unique SPE; (2) the infinitely repeated game model with
discounting; (3) feasible and individually rational payoff sets, with the
minimax value providing a payoff lower bound in any NE; (4) grim trigger and
tit-for-tat as concrete mechanisms for sustaining cooperation; (5) the Nash
folk theorem, showing that any feasible, strictly individually rational
payoff is achievable in NE for patient players; and (6) the subgame perfect
folk theorem of \citet{fudenberg1986}, extending this to SPE under the full
dimensionality condition. These results are foundational for understanding
emergent behavior in multi-agent RL, where repeated interaction, discount
factors, and implicit punishment strategies play central roles.

\bigskip
\bibliography{game-theory-refs}

\end{document}
