\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 46: LP Formulation and Occupancy Measures}}
\fancyhead[R]{\thepage}
\title{Lesson 46: LP Formulation and Occupancy Measures}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% ============================================================
%% SECTION 1: MOTIVATION AND OVERVIEW
%% ============================================================

\section{Motivation and Overview}
\label{sec:motivation}

Lessons~42 and~45 developed two iterative algorithms for solving
discounted MDPs: value iteration and policy iteration. Both are
efficient and widely used, but they share a common limitation --
they cannot easily accommodate \emph{constraints}. Consider a
robot navigation MDP where the agent must reach a goal while
keeping the expected number of visits to dangerous states below a
threshold, or an energy management system that must minimize cost
subject to an average emission constraint. These constrained
optimization problems cannot be directly formulated as standard
Bellman equations.

Linear programming provides an alternative, fundamentally different
approach to MDPs. Rather than iterating on value functions or
policies, we cast the MDP as a linear program whose optimal solution
\emph{is} the optimal value function. The dual of this LP introduces
\emph{occupancy measures} -- probability distributions over
state-action pairs that encode how often each pair is visited under
a policy. The one-to-one correspondence between stationary policies
and occupancy measures transforms policy optimization into
optimization over a convex polytope, enabling the incorporation of
linear constraints naturally. This lesson develops the LP formulation,
the duality theory, and the occupancy measure framework, culminating
in the connection to constrained MDPs and inverse RL.

\begin{rlconnection}[title={Why the LP Formulation Matters for RL/ML}]
\begin{itemize}[nosep]
    \item \textbf{Constrained MDPs}: Adding linear constraints on
    occupancy measures directly handles safety constraints, resource
    budgets, and fairness requirements -- a key capability for
    real-world RL deployment.
    \item \textbf{Inverse RL}: The occupancy measure framework
    underpins inverse reinforcement learning, where the goal is to
    recover a reward function from observed expert behavior by
    matching occupancy measures.
    \item \textbf{Apprenticeship learning}: \citet{altman1999} and
    subsequent work on apprenticeship learning formulate imitation
    as matching the feature expectations (linear functions of
    occupancy measures) between the learner and the expert.
    \item \textbf{Convex optimization perspective}: The LP
    formulation reveals that MDP solving is a convex optimization
    problem, connecting RL theory to the rich toolkit of convex
    analysis (Lesson~17) and duality (Lesson~18).
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
MDP model, policies, value functions (Lesson~39). Bellman optimality
equations, the Bellman optimality operator $\Bellman^*$ (Lesson~41).
Value iteration (Lesson~42) and policy iteration (Lesson~45).
Linear programming basics: primal/dual LPs, strong duality, basic
feasible solutions, vertices of polytopes (Lesson~18 or equivalent
background in convex optimization).

\paragraph{Learning objectives.}
After completing this lesson, the student will be able to:
\begin{itemize}
    \item \emph{Derive} the primal LP formulation of a discounted
    MDP and \emph{prove} that its optimal solution equals $\Vstar$.
    \item \emph{Define} occupancy measures and \emph{prove} the
    flow conservation constraints they satisfy.
    \item \emph{Establish} the one-to-one correspondence between
    stationary policies and occupancy measures.
    \item \emph{Prove} LP strong duality for the MDP primal-dual pair.
    \item \emph{Formulate} constrained MDPs as linear programs with
    additional constraints on occupancy measures.
    \item \emph{Explain} how occupancy measures connect to inverse
    RL and apprenticeship learning.
\end{itemize}

%% ============================================================
%% SECTION 2: REFERENCES
%% ============================================================

\section{References}
\label{sec:references}

\begin{itemize}
    \item \citet[Chapter 6, Sections 6.9--6.10]{puterman2014} --
    Definitive treatment of the LP formulation of discounted MDPs,
    occupancy measures, and duality.
    \item \citet{manne1960} -- Original paper formulating MDPs as
    linear programs, establishing the connection between value
    functions and LP solutions.
    \item \citet{altman1999} -- Comprehensive treatment of
    constrained MDPs, including the LP approach and Lagrangian
    relaxation methods.
    \item \citet[Chapter 1]{bertsekas2012} -- LP formulation in
    the context of dynamic programming and optimal control.
    \item \citet{derman1970} -- Early treatment of the LP approach
    to MDPs with applications to operations research.
    \item \citet[Chapter 6]{suttonbarto2018} -- Discussion of the
    relationship between value functions, policies, and optimization.
\end{itemize}

%% ============================================================
%% SECTION 3: THE PRIMAL LP
%% ============================================================

\section{The Primal Linear Program}
\label{sec:primal-lp}

We work with a finite MDP
$\mathcal{M} = (\States, \Actions, P, R, \discount)$ with
$|\States| = S$, $|\Actions| = A$, and $\discount \in [0,1)$.
The key insight is that the Bellman optimality equation
$v(s) = \max_a [R(s,a) + \discount \sum_{s'} P(s'|s,a) v(s')]$
can be relaxed into a system of linear inequalities.

\begin{definition}[Bellman Feasibility]
\label{def:bellman-feasible}
A vector $v \in \R^S$ is \emph{Bellman superharmonic} (or
\emph{Bellman feasible}) if
\begin{equation}
\label{eq:bellman-ineq}
    v(s) \geq R(s,a) + \discount \sum_{s' \in \States}
    P(s' \mid s,a)\, v(s') \quad
    \text{for all } s \in \States, \; a \in \Actions.
\end{equation}
\end{definition}

\begin{lemma}[$\Vstar$ Is the Smallest Bellman Superharmonic Vector]
\label{lem:vstar-smallest}
The optimal value function $\Vstar$ satisfies the Bellman
feasibility constraints \eqref{eq:bellman-ineq}. Moreover, if
$v \in \R^S$ satisfies \eqref{eq:bellman-ineq}, then
$v(s) \geq \Vstar(s)$ for all $s$.
\end{lemma}

\begin{proof}
$\Vstar$ satisfies \eqref{eq:bellman-ineq} with equality at the
maximizing action: $\Vstar(s) = \max_a [R(s,a) + \discount \sum_{s'}
P(s'|s,a) \Vstar(s')] \geq R(s,a) + \discount \sum_{s'} P(s'|s,a)
\Vstar(s')$ for all $a$.

Now suppose $v$ satisfies \eqref{eq:bellman-ineq}. Then for all $s$:
\[
    v(s) \geq \max_{a \in \Actions} \left[ R(s,a) + \discount
    \sum_{s'} P(s'|s,a)\, v(s') \right] = (\Bellman^* v)(s).
\]
So $v \geq \Bellman^* v$. By the monotonicity of $\Bellman^*$
(Lesson~42, Proposition~8.1):
$\Bellman^* v \geq (\Bellman^*)^2 v \geq \cdots$. Wait -- this
goes the wrong direction. Let us argue more carefully.

From $v \geq \Bellman^* v$, by monotonicity of $\Bellman^*$:
$\Bellman^* v \geq (\Bellman^*)^2 v$. So
$v \geq \Bellman^* v \geq (\Bellman^*)^2 v \geq \cdots$
The sequence $\{(\Bellman^*)^n v\}$ is non-increasing. Since
$(\Bellman^*)^n v \to \Vstar$ by the Banach fixed-point theorem
(Lesson~42, Theorem~4.1), we get $v \geq \Vstar$.
\end{proof}

Lemma~\ref{lem:vstar-smallest} shows that $\Vstar$ minimizes any
objective that sums $v(s)$ with positive weights over the Bellman
feasible set. This motivates the primal LP.

\begin{definition}[Primal LP for Discounted MDPs]
\label{def:primal-lp}
Let $\alpha \in \R^S$ be any vector with $\alpha(s) > 0$ for all
$s$ and $\sum_s \alpha(s) = 1$ (an initial state distribution).
The \emph{primal LP} is:
\begin{equation}
\label{eq:primal-lp}
\begin{aligned}
    \text{minimize} \quad & c^\top v = \sum_{s \in \States}
    \alpha(s)\, v(s) \\
    \text{subject to} \quad & v(s) \geq R(s,a) + \discount
    \sum_{s'} P(s' \mid s,a)\, v(s'), \quad
    \forall\, s \in \States,\; a \in \Actions, \\
    & v \in \R^S.
\end{aligned}
\end{equation}
This LP has $S$ decision variables and $SA$ inequality constraints.
\end{definition}

\begin{theorem}[Primal LP Solves the MDP]
\label{thm:primal-optimal}
The optimal value of the primal LP \eqref{eq:primal-lp} equals
$\sum_s \alpha(s) \Vstar(s)$, and the unique optimal solution is
$v^* = \Vstar$.
\end{theorem}

\begin{proof}
By Lemma~\ref{lem:vstar-smallest}, $\Vstar$ is feasible and
$v \geq \Vstar$ for every feasible $v$. Since $\alpha(s) > 0$
for all $s$:
\[
    \sum_s \alpha(s) v(s) \geq \sum_s \alpha(s) \Vstar(s)
\]
for every feasible $v$, with equality if and only if $v = \Vstar$
(since $\alpha(s) > 0$ and $v(s) \geq \Vstar(s)$ imply that
equality of the sums requires $v(s) = \Vstar(s)$ for all $s$).
Therefore $\Vstar$ is the unique minimizer.
\end{proof}

\begin{keyresult}[title={MDP as a Linear Program}]
\label{kr:mdp-lp}
Finding the optimal value function $\Vstar$ of a discounted MDP
is equivalent to solving the linear program \eqref{eq:primal-lp}.
The LP has $S$ variables and $SA$ constraints. Once $\Vstar$ is
obtained, the optimal policy is recovered by identifying which
constraints are active (tight) at each state: $\optpolicy(s) =
\argmax_a [R(s,a) + \discount \sum_{s'} P(s'|s,a) \Vstar(s')]$.
\end{keyresult}

\begin{remark}
\label{rem:alpha-choice}
The choice of $\alpha$ does not affect the optimal \emph{solution}
$v^* = \Vstar$ (which is unique by Lemma~\ref{lem:vstar-smallest}),
only the optimal \emph{value} $\alpha^\top \Vstar$. Any strictly
positive $\alpha$ works. The uniform distribution
$\alpha(s) = 1/S$ is a common default.
\end{remark}

\begin{example}[Primal LP for a Two-State MDP]
\label{ex:primal-two-state}
Consider the two-state MDP with $\States = \{1, 2\}$,
$\Actions = \{a_1, a_2\}$, $\discount = 0.5$:
\begin{center}
\begin{tabular}{ccccc}
\toprule
$s$ & $a$ & $R(s,a)$ & $P(1 \mid s,a)$ & $P(2 \mid s,a)$ \\
\midrule
1 & $a_1$ & 5 & 0.5 & 0.5 \\
1 & $a_2$ & 10 & 0 & 1 \\
2 & $a_1$ & $-1$ & 0.8 & 0.2 \\
2 & $a_2$ & 2 & 0.3 & 0.7 \\
\bottomrule
\end{tabular}
\end{center}
With $\alpha = (0.5, 0.5)^\top$, the primal LP is:
\begin{align*}
    \min \quad & 0.5\, v_1 + 0.5\, v_2 \\
    \text{s.t.} \quad
    & v_1 \geq 5 + 0.5(0.5\, v_1 + 0.5\, v_2) \\
    & v_1 \geq 10 + 0.5(0 \cdot v_1 + 1 \cdot v_2) \\
    & v_2 \geq -1 + 0.5(0.8\, v_1 + 0.2\, v_2) \\
    & v_2 \geq 2 + 0.5(0.3\, v_1 + 0.7\, v_2)
\end{align*}
Simplifying:
\begin{align*}
    0.75\, v_1 - 0.25\, v_2 &\geq 5, \\
    v_1 - 0.5\, v_2 &\geq 10, \\
    -0.4\, v_1 + 0.9\, v_2 &\geq -1, \\
    -0.15\, v_1 + 0.65\, v_2 &\geq 2.
\end{align*}
From Lessons~41--42, $\Vstar \approx (13.04, 6.09)^\top$, achieved by
$\optpolicy(1) = a_2$, $\optpolicy(2) = a_2$. One can verify that
$\Vstar$ satisfies all four constraints, with the second and fourth
constraints tight (corresponding to the optimal actions).
\end{example}

\begin{warning}[title={LP Size and Practical Considerations}]
The primal LP has $S$ variables and $SA$ constraints. For modern LP
solvers, problems with up to $10^5$--$10^6$ variables and constraints
are tractable. However, for very large MDPs ($S > 10^6$), specialized
algorithms (value iteration, policy iteration) are more practical.
The LP formulation is most valuable for its theoretical insights
(duality, occupancy measures, constrained MDPs) rather than as a
computational workhorse.
\end{warning}

%% ============================================================
%% SECTION 4: THE DUAL LP AND OCCUPANCY MEASURES
%% ============================================================

\section{The Dual LP and Occupancy Measures}
\label{sec:dual-lp}

We now derive the dual of the primal LP \eqref{eq:primal-lp} and
show that it has a beautiful probabilistic interpretation in terms
of occupancy measures.

\begin{theorem}[Dual LP for Discounted MDPs]
\label{thm:dual-lp}
The dual of the primal LP \eqref{eq:primal-lp} is:
\begin{equation}
\label{eq:dual-lp}
\begin{aligned}
    \text{maximize} \quad & \sum_{s,a} R(s,a)\, x(s,a) \\
    \text{subject to} \quad & \sum_{a \in \Actions} x(s,a)
    = \alpha(s) + \discount \sum_{s',a'} P(s \mid s',a')\,
    x(s',a'), \quad \forall\, s \in \States, \\
    & x(s,a) \geq 0, \quad \forall\, s \in \States,\;
    a \in \Actions.
\end{aligned}
\end{equation}
This LP has $SA$ decision variables $x(s,a)$ and $S$ equality
constraints (plus $SA$ non-negativity constraints).
\end{theorem}

\begin{proof}
Rewrite the primal constraints as: for all $(s,a)$,
\[
    v(s) - \discount \sum_{s'} P(s'|s,a) v(s') \geq R(s,a).
\]
In matrix form with $v \in \R^S$: $(I - \discount P_a^\top) v \geq
r_a$ for each action $a$, where $P_a$ is the $S \times S$ matrix
with entries $P_a(s,s') = P(s'|s,a) \indicator[a]$.
More precisely, we associate a dual variable $x(s,a) \geq 0$
with each constraint. The Lagrangian is:
\begin{align*}
    L(v, x) &= \sum_s \alpha(s) v(s) - \sum_{s,a} x(s,a)
    \left[ v(s) - R(s,a) - \discount \sum_{s'} P(s'|s,a) v(s')
    \right] \\
    &= \sum_s v(s) \left[ \alpha(s) - \sum_a x(s,a)
    + \discount \sum_{s',a'} P(s|s',a') x(s',a') \right]
    + \sum_{s,a} R(s,a) x(s,a).
\end{align*}
For the dual to be bounded ($\inf_v L(v,x) > -\infty$), the
coefficient of each $v(s)$ must be zero:
\[
    \sum_a x(s,a) = \alpha(s) + \discount \sum_{s',a'}
    P(s|s',a') x(s',a'), \quad \forall\, s.
\]
Under this constraint, $L(v,x) = \sum_{s,a} R(s,a) x(s,a)$, which
we maximize. This gives the dual LP \eqref{eq:dual-lp}.
\end{proof}

\begin{definition}[Occupancy Measure]
\label{def:occupancy}
For a stationary policy $\policy$ and initial distribution $\alpha$,
the \emph{(discounted) occupancy measure} is
\begin{equation}
\label{eq:occupancy-def}
    x^\policy(s,a) = (1 - \discount) \sum_{t=0}^\infty \discount^t
    \Prob^\policy_\alpha(S_t = s, A_t = a),
\end{equation}
where $\Prob^\policy_\alpha$ denotes the probability under policy
$\policy$ with initial distribution $S_0 \sim \alpha$.
\end{definition}

\begin{remark}
\label{rem:normalization}
The normalization factor $(1-\discount)$ ensures that $x^\policy$
is a probability distribution: $\sum_{s,a} x^\policy(s,a) = 1$.
Some references omit the $(1-\discount)$ factor, in which case
$\sum_{s,a} x^\policy(s,a) = 1/(1-\discount)$. We adopt the
normalized convention.
\end{remark}

Without the normalization factor, define
$\tilde{x}^\policy(s,a) = \sum_{t=0}^\infty \discount^t
\Prob^\policy_\alpha(S_t = s, A_t = a)$, so that
$x^\policy = (1-\discount) \tilde{x}^\policy$. We will show that
$\tilde{x}^\policy$ is a feasible solution of the dual LP
\eqref{eq:dual-lp} (after appropriate rescaling).

\begin{proposition}[Occupancy Measures as Dual Feasible Points]
\label{prop:occupancy-feasible}
For any stationary policy $\policy$, the unnormalized occupancy
measure $\tilde{x}^\policy(s,a) = \sum_{t=0}^\infty \discount^t
\Prob^\policy_\alpha(S_t = s, A_t = a)$ satisfies the dual
constraints \eqref{eq:dual-lp}.
\end{proposition}

\begin{proof}
Clearly $\tilde{x}^\policy(s,a) \geq 0$. For the flow constraint,
define the state occupancy $\tilde{\mu}^\policy(s) = \sum_a
\tilde{x}^\policy(s,a) = \sum_{t=0}^\infty \discount^t
\Prob^\policy_\alpha(S_t = s)$. We need to show:
\[
    \tilde{\mu}^\policy(s) = \alpha(s) + \discount
    \sum_{s',a'} P(s|s',a') \tilde{x}^\policy(s',a').
\]
The left side is:
\begin{align*}
    \tilde{\mu}^\policy(s) &= \Prob_\alpha(S_0 = s) +
    \sum_{t=1}^\infty \discount^t \Prob^\policy_\alpha(S_t = s) \\
    &= \alpha(s) + \discount \sum_{t=0}^\infty \discount^t
    \sum_{s',a'} P(s|s',a') \Prob^\policy_\alpha(S_t = s',
    A_t = a') \\
    &= \alpha(s) + \discount \sum_{s',a'} P(s|s',a')
    \tilde{x}^\policy(s',a'),
\end{align*}
where the second equality uses the Markov property:
$\Prob^\policy_\alpha(S_{t+1} = s) = \sum_{s',a'} P(s|s',a')
\Prob^\policy_\alpha(S_t = s', A_t = a')$.
\end{proof}

\begin{proposition}[Dual Objective Equals Policy Value]
\label{prop:dual-value}
The dual objective evaluated at $\tilde{x}^\policy$ equals the
expected discounted return under $\policy$:
\[
    \sum_{s,a} R(s,a) \tilde{x}^\policy(s,a)
    = \sum_s \alpha(s) \Vpi(s).
\]
\end{proposition}

\begin{proof}
\begin{align*}
    \sum_{s,a} R(s,a) \tilde{x}^\policy(s,a)
    &= \sum_{s,a} R(s,a) \sum_{t=0}^\infty \discount^t
    \Prob^\policy_\alpha(S_t = s, A_t = a) \\
    &= \sum_{t=0}^\infty \discount^t \sum_{s,a} R(s,a)
    \Prob^\policy_\alpha(S_t = s, A_t = a) \\
    &= \sum_{t=0}^\infty \discount^t
    \E^\policy_\alpha[R(S_t, A_t)] \\
    &= \E^\policy_\alpha\!\left[\sum_{t=0}^\infty
    \discount^t R(S_t, A_t)\right]
    = \sum_s \alpha(s) \Vpi(s). \qedhere
\end{align*}
\end{proof}

%% ============================================================
%% SECTION 5: PROPERTIES OF OCCUPANCY MEASURES
%% ============================================================

\section{Properties of Occupancy Measures}
\label{sec:occupancy-properties}

\begin{theorem}[Characterization of the Occupancy Polytope]
\label{thm:occupancy-polytope}
Define the \emph{occupancy polytope}
\begin{equation}
\label{eq:occupancy-set}
    \mathcal{X}_\alpha = \left\{ x \in \R^{SA}_{\geq 0} :
    \sum_a x(s,a) = \alpha(s) + \discount \sum_{s',a'}
    P(s|s',a') x(s',a'), \;\; \forall\, s \in \States \right\}.
\end{equation}
Then:
\begin{enumerate}
    \item $\mathcal{X}_\alpha$ is a non-empty bounded convex
    polytope.
    \item $\sum_{s,a} x(s,a) = 1/(1-\discount)$ for every
    $x \in \mathcal{X}_\alpha$.
    \item The map $\policy \mapsto \tilde{x}^\policy$ is a
    bijection between stationary policies and $\mathcal{X}_\alpha$.
\end{enumerate}
\end{theorem}

\begin{proof}
\textbf{Part 1.} Non-emptiness: $\tilde{x}^\policy \in
\mathcal{X}_\alpha$ for any stationary policy $\policy$
(Proposition~\ref{prop:occupancy-feasible}). Convexity:
$\mathcal{X}_\alpha$ is defined by linear equalities and
non-negativity constraints, hence it is a polyhedron. Boundedness:
summing the flow constraints over $s$:
\[
    \sum_{s,a} x(s,a) = 1 + \discount \sum_{s,a} x(s,a),
\]
so $\sum_{s,a} x(s,a) = 1/(1-\discount)$. This proves Part 2 and
also implies boundedness (each $x(s,a) \leq 1/(1-\discount)$).

\textbf{Part 3 (surjectivity).} Given $x \in \mathcal{X}_\alpha$,
define $\mu(s) = \sum_a x(s,a)$ and the policy
\begin{equation}
\label{eq:policy-from-occupancy}
    \policy_x(a|s) = \begin{cases}
    x(s,a) / \mu(s) & \text{if } \mu(s) > 0, \\
    1/|\Actions| & \text{if } \mu(s) = 0.
    \end{cases}
\end{equation}
We need to show that $\tilde{x}^{\policy_x} = x$. Let
$y = \tilde{x}^{\policy_x}$. Define $\nu(s) = \sum_a y(s,a)$.
Then $y(s,a) = \policy_x(a|s) \nu(s)$ and
$\nu(s) = \alpha(s) + \discount \sum_{s'} P(s|s', \policy_x(s'))
\nu(s')$. Similarly, $x(s,a) = \policy_x(a|s) \mu(s)$ by
construction and $\mu(s) = \alpha(s) + \discount \sum_{s',a'}
P(s|s',a') x(s',a') = \alpha(s) + \discount \sum_{s'}
P(s|s', \policy_x(s')) \mu(s')$. So both $\mu$ and $\nu$ satisfy
the same linear system
$\mu = \alpha + \discount (P^{\policy_x})^\top \mu$, i.e.,
$(I - \discount (P^{\policy_x})^\top) \mu = \alpha$. Since
$\discount < 1$, this system has a unique solution
(Lesson~45, Theorem~\ref{thm:policy-eval-matrix} applies to the
transpose), so $\mu = \nu$ and hence $x = y$.

\textbf{Part 3 (injectivity).} If $\tilde{x}^{\policy_1} =
\tilde{x}^{\policy_2}$, then by \eqref{eq:policy-from-occupancy},
$\policy_1(a|s) = \tilde{x}^{\policy_1}(s,a) / \mu(s) =
\tilde{x}^{\policy_2}(s,a) / \mu(s) = \policy_2(a|s)$ for all
$s$ with $\mu(s) > 0$. For states with $\mu(s) = 0$ (never
visited), the policies may differ, but they yield the same
occupancy measure.
\end{proof}

\begin{keyresult}[title={Policy-Occupancy Correspondence}]
\label{kr:correspondence}
There is a one-to-one correspondence between stationary policies
and points in the occupancy polytope $\mathcal{X}_\alpha$:
\[
    \policy \;\longleftrightarrow\; \tilde{x}^\policy, \qquad
    \policy_x(a|s) = \frac{x(s,a)}{\sum_{a'} x(s,a')}.
\]
Deterministic policies correspond to vertices of
$\mathcal{X}_\alpha$, and the optimal policy corresponds to the
vertex that maximizes the linear objective
$\sum_{s,a} R(s,a) x(s,a)$.
\end{keyresult}

\begin{proposition}[Deterministic Policies Are Vertices]
\label{prop:vertices}
The vertices (extreme points) of $\mathcal{X}_\alpha$ are
precisely the occupancy measures of deterministic stationary policies.
\end{proposition}

\begin{proof}
$\mathcal{X}_\alpha$ is a polytope in $\R^{SA}$ defined by $S$
equality constraints and $SA$ non-negativity constraints. A vertex
(basic feasible solution) has exactly $SA - S = S(A-1)$ active
non-negativity constraints, meaning that for each state $s$,
exactly $A - 1$ of the variables $\{x(s,a)\}_{a \in \Actions}$
are zero. Hence exactly one action per state has $x(s,a) > 0$,
which corresponds to a deterministic policy.

Conversely, for a deterministic policy $d$, the occupancy measure
$\tilde{x}^d(s,a) = 0$ for $a \neq d(s)$, giving exactly
$S(A-1)$ active non-negativity constraints. The remaining $S$
variables $\{\tilde{x}^d(s, d(s))\}_{s \in \States}$ are
uniquely determined by the $S$ flow constraints (since the linear
system is non-singular). Thus $\tilde{x}^d$ is a basic feasible
solution, hence a vertex.
\end{proof}

\begin{example}[Occupancy Measures for the Two-State MDP]
\label{ex:occupancy-two-state}
Consider the two-state MDP from Example~\ref{ex:primal-two-state}
with $\discount = 0.5$ and $\alpha = (0.5, 0.5)^\top$.

For the optimal policy $d^*(1) = a_2$, $d^*(2) = a_2$, the
transition matrix is $P^{d^*} = \bigl(\begin{smallmatrix}
0 & 1 \\ 0.3 & 0.7 \end{smallmatrix}\bigr)$. The unnormalized
state occupancy satisfies $(I - 0.5 (P^{d^*})^\top) \mu = \alpha$:
\[
    \begin{pmatrix} 1 & -0.15 \\ -0.5 & 0.65 \end{pmatrix}
    \mu = \begin{pmatrix} 0.5 \\ 0.5 \end{pmatrix}.
\]
Determinant $= 0.65 - 0.075 = 0.575$.
$\mu_1 = (0.5 \cdot 0.65 + 0.15 \cdot 0.5)/0.575
= 0.4/0.575 \approx 0.696$,
$\mu_2 = (1 \cdot 0.5 + 0.5 \cdot 0.5)/0.575
= 0.75/0.575 \approx 1.304$.
Check: $\mu_1 + \mu_2 = 2 = 1/(1-0.5)$. Correct.

The occupancy measure is $\tilde{x}^{d^*}(1, a_2) = 0.696$,
$\tilde{x}^{d^*}(2, a_2) = 1.304$, and
$\tilde{x}^{d^*}(s, a_1) = 0$ for both states (deterministic
policy).

The dual objective is $10 \cdot 0.696 + 2 \cdot 1.304 = 9.57$,
which should equal $\alpha^\top \Vstar = 0.5 \cdot 13.04 +
0.5 \cdot 6.09 = 9.57$. Confirmed.

The normalized occupancy is $x^{d^*} = (1-0.5) \tilde{x}^{d^*}$:
$x^{d^*}(1, a_2) = 0.348$, $x^{d^*}(2, a_2) = 0.652$. This is a
probability distribution summing to 1, representing the long-run
discounted frequency of visiting each state-action pair.
\end{example}

%% ============================================================
%% SECTION 6: LP STRONG DUALITY
%% ============================================================

\section{LP Strong Duality}
\label{sec:strong-duality}

\begin{theorem}[Strong Duality for the MDP LP]
\label{thm:strong-duality}
The primal LP \eqref{eq:primal-lp} and dual LP \eqref{eq:dual-lp}
satisfy strong duality:
\begin{equation}
\label{eq:strong-duality}
    \min_{v \text{ feasible}} \sum_s \alpha(s) v(s)
    = \max_{x \in \mathcal{X}_\alpha} \sum_{s,a} R(s,a) x(s,a)
    = \sum_s \alpha(s) \Vstar(s).
\end{equation}
The unique primal optimum is $v^* = \Vstar$. The dual optimum is
$x^* = \tilde{x}^{\optpolicy}$, the unnormalized occupancy measure
of any optimal policy $\optpolicy$.
\end{theorem}

\begin{proof}
Both the primal and dual are feasible: the primal is feasible at
$\Vstar$ and the dual is feasible at $\tilde{x}^\policy$ for any
$\policy$. Both are bounded: the primal is bounded below (by the
dual objective at any feasible $x$) and the dual is bounded above
(by the primal objective at $\Vstar$). By the LP strong duality
theorem (which applies to any LP pair where both are feasible and
bounded), the optimal values are equal.

By Theorem~\ref{thm:primal-optimal}, the primal optimum is
$\Vstar$. By Proposition~\ref{prop:dual-value}, the dual objective
at $\tilde{x}^{\optpolicy}$ equals $\sum_s \alpha(s) \Vstar(s)$,
which is the primal optimal value. Hence
$\tilde{x}^{\optpolicy}$ is dual optimal.
\end{proof}

\begin{remark}
\label{rem:complementary-slackness}
Complementary slackness gives additional insight. At the optimal
pair $(\Vstar, \tilde{x}^{\optpolicy})$:
\begin{itemize}[nosep]
    \item For each $(s,a)$: either $\tilde{x}^{\optpolicy}(s,a) = 0$
    or the constraint $\Vstar(s) = R(s,a) + \discount \sum_{s'}
    P(s'|s,a) \Vstar(s')$ is tight.
    \item This means the optimal policy only assigns positive
    probability to actions that are optimal (achieve the maximum in
    the Bellman equation) -- consistent with the greedy policy
    characterization.
\end{itemize}
\end{remark}

\begin{rlconnection}[title={LP Duality and Reward Shaping}]
The LP duality framework provides a principled way to understand
reward shaping. A potential-based shaping function
$\Phi : \States \to \R$ modifies the reward to
$\tilde{R}(s,a,s') = R(s,a) + \discount \Phi(s') - \Phi(s)$. In
the LP framework, this corresponds to translating the primal
variable $v \mapsto v - \Phi$, which shifts the constraints but
does not change the optimal policy. The occupancy measures (dual
variables) remain unchanged -- confirming that potential-based
shaping preserves the optimal policy, a result known as Ng's
reward shaping theorem.
\end{rlconnection}

%% ============================================================
%% SECTION 7: CONSTRAINED MDPs
%% ============================================================

\section{Constrained MDPs}
\label{sec:cmdp}

The LP formulation naturally accommodates constraints on the agent's
behavior, leading to constrained MDPs (CMDPs).

\begin{definition}[Constrained MDP]
\label{def:cmdp}
A \emph{constrained MDP} (CMDP) extends a standard MDP with $K$
auxiliary cost functions $C_k : \States \times \Actions \to \R$
($k = 1, \ldots, K$) and corresponding bounds $b_k$. The goal is:
\begin{equation}
\label{eq:cmdp}
    \max_\policy \sum_s \alpha(s) \Vpi(s) \quad
    \text{subject to} \quad
    \sum_s \alpha(s) V^\policy_{C_k}(s) \leq b_k, \quad
    k = 1, \ldots, K,
\end{equation}
where $V^\policy_{C_k}(s) = \E^\policy[\sum_{t=0}^\infty
\discount^t C_k(S_t, A_t) \mid S_0 = s]$ is the expected
discounted cost under cost function $C_k$.
\end{definition}

The key observation is that $\sum_s \alpha(s) V^\policy_{C_k}(s) =
\sum_{s,a} C_k(s,a) \tilde{x}^\policy(s,a)$ by the same argument
as Proposition~\ref{prop:dual-value}. Therefore, the CMDP can be
written entirely in terms of occupancy measures.

\begin{theorem}[LP Formulation of CMDPs]
\label{thm:cmdp-lp}
The constrained MDP \eqref{eq:cmdp} is equivalent to the linear
program:
\begin{equation}
\label{eq:cmdp-lp}
\begin{aligned}
    \text{maximize} \quad & \sum_{s,a} R(s,a)\, x(s,a) \\
    \text{subject to} \quad & \sum_a x(s,a) = \alpha(s) +
    \discount \sum_{s',a'} P(s|s',a') x(s',a'), \quad
    \forall\, s, \\
    & \sum_{s,a} C_k(s,a)\, x(s,a) \leq b_k, \quad
    k = 1, \ldots, K, \\
    & x(s,a) \geq 0, \quad \forall\, s, a.
\end{aligned}
\end{equation}
\end{theorem}

\begin{proof}
By the policy-occupancy correspondence
(Theorem~\ref{thm:occupancy-polytope}), optimizing over stationary
policies is equivalent to optimizing over $\mathcal{X}_\alpha$.
The objective $\sum_s \alpha(s) \Vpi(s) = \sum_{s,a} R(s,a)
\tilde{x}^\policy(s,a)$ and the constraints
$\sum_s \alpha(s) V^\policy_{C_k}(s) = \sum_{s,a} C_k(s,a)
\tilde{x}^\policy(s,a) \leq b_k$ are all linear in
$\tilde{x}^\policy$. Replacing $\tilde{x}^\policy$ by the
free variable $x \in \mathcal{X}_\alpha$ yields \eqref{eq:cmdp-lp}.
\end{proof}

\begin{keyresult}[title={Constrained MDPs via Linear Programming}]
\label{kr:cmdp}
Any finite constrained MDP with $K$ linear cost constraints can be
solved as a linear program with $SA$ variables and $S + K$ equality/
inequality constraints (plus non-negativity). The optimal policy may
be \emph{stochastic} -- it randomizes over at most $K + 1$ actions
at each state (since the LP optimum can lie on a face of dimension
at most $K$ above the unconstrained vertex).
\end{keyresult}

\begin{remark}
\label{rem:randomization}
In the unconstrained case ($K = 0$), the LP optimum is a vertex
of $\mathcal{X}_\alpha$, corresponding to a deterministic policy.
With $K$ constraints, the optimum may lie in the interior of a
face, requiring a mixture of at most $K + 1$ deterministic policies.
This is a fundamental result: constraints can force optimal policies
to be stochastic.
\end{remark}

\begin{example}[Constrained Two-State MDP]
\label{ex:cmdp}
Consider the two-state MDP from Example~\ref{ex:primal-two-state}
with an additional safety cost $C(s,a)$ representing the probability
of triggering an alarm:
\begin{center}
\begin{tabular}{ccc}
\toprule
$s$ & $a$ & $C(s,a)$ \\
\midrule
1 & $a_1$ & 0.1 \\
1 & $a_2$ & 0.5 \\
2 & $a_1$ & 0.2 \\
2 & $a_2$ & 0.3 \\
\bottomrule
\end{tabular}
\end{center}
We require the expected discounted alarm cost to be at most $b = 0.6$:
$\sum_{s,a} C(s,a) \tilde{x}(s,a) \leq 0.6$.

The unconstrained optimum uses $d^*(1) = a_2$, $d^*(2) = a_2$ with
$\tilde{x}^{d^*}(1, a_2) \approx 0.696$ and
$\tilde{x}^{d^*}(2, a_2) \approx 1.304$. The alarm cost is
$0.5 \cdot 0.696 + 0.3 \cdot 1.304 = 0.348 + 0.391 = 0.739 > 0.6$.
The constraint is violated, so the unconstrained optimal policy is
infeasible.

The constrained LP adds the constraint
$0.1\, x(1,a_1) + 0.5\, x(1,a_2) + 0.2\, x(2,a_1) +
0.3\, x(2,a_2) \leq 0.6$. Solving this modified LP yields a
stochastic policy that randomizes between $a_1$ and $a_2$ at
one or both states.
\end{example}

\begin{rlconnection}[title={Constrained MDPs in Safe RL}]
The CMDP framework is the theoretical foundation for \emph{safe RL},
where agents must satisfy safety or resource constraints:
\begin{itemize}[nosep]
    \item \textbf{Constrained Policy Optimization (CPO)} solves
    an approximate CMDP at each iteration using trust-region
    methods combined with Lagrangian relaxation.
    \item \textbf{Lagrangian relaxation}: The CMDP can be relaxed
    by introducing multipliers $\lambda_k \geq 0$ for each
    constraint, yielding the unconstrained problem
    $\max_\policy [\sum_s \alpha(s) \Vpi(s) - \sum_k \lambda_k
    (\sum_s \alpha(s) V^\policy_{C_k}(s) - b_k)]$. Under strong
    duality (which holds for CMDPs), the saddle-point problem
    $\max_\policy \min_{\lambda \geq 0}$ yields the CMDP solution.
    \item \textbf{Primal-dual algorithms}: Practical safe RL
    algorithms alternate between policy updates and dual variable
    updates, implementing a stochastic primal-dual method on the
    CMDP Lagrangian.
\end{itemize}
\end{rlconnection}

%% ============================================================
%% SECTION 8: CONNECTION TO INVERSE RL
%% ============================================================

\section{Connection to Inverse RL and Apprenticeship Learning}
\label{sec:inverse-rl}

The occupancy measure framework provides a natural bridge to
\emph{inverse reinforcement learning} (IRL), where the goal is
to learn a reward function from observed expert behavior.

\begin{definition}[Feature Expectations]
\label{def:feature-expectations}
Given a feature map $\phi : \States \times \Actions \to \R^d$, the
\emph{feature expectation} of a policy $\policy$ is
\begin{equation}
\label{eq:feature-expect}
    \bar{\phi}^\policy = \sum_{s,a} \phi(s,a)\,
    \tilde{x}^\policy(s,a) = \E^\policy_\alpha\!\left[
    \sum_{t=0}^\infty \discount^t \phi(S_t, A_t)\right].
\end{equation}
\end{definition}

\begin{proposition}[Linear Reward and Feature Expectations]
\label{prop:linear-reward}
If the reward function is linear in features,
$R(s,a) = w^\top \phi(s,a)$ for some weight vector $w \in \R^d$,
then
\[
    \sum_s \alpha(s) \Vpi(s) = w^\top \bar{\phi}^\policy.
\]
\end{proposition}

\begin{proof}
Direct computation:
$\sum_s \alpha(s) \Vpi(s) = \sum_{s,a} R(s,a)
\tilde{x}^\policy(s,a) = \sum_{s,a} w^\top \phi(s,a)
\tilde{x}^\policy(s,a) = w^\top \bar{\phi}^\policy$.
\end{proof}

\begin{remark}
\label{rem:irl-connection}
In inverse RL, we observe an expert policy $\policy_E$ (through
demonstrations) and wish to find a reward function $R$ under which
$\policy_E$ is optimal. If $R(s,a) = w^\top \phi(s,a)$, then
$\policy_E$ is optimal iff $w^\top \bar{\phi}^{\policy_E} \geq
w^\top \bar{\phi}^\policy$ for all $\policy$. Equivalently, the
expert's feature expectation $\bar{\phi}^{\policy_E}$ must lie on
the boundary of the set $\{\bar{\phi}^\policy : \policy \text{
stationary}\}$ in the direction $w$.

This geometric view leads to algorithms that find $w$ by searching
for a hyperplane separating $\bar{\phi}^{\policy_E}$ from the
feature expectations of other policies -- the basis of
\emph{apprenticeship learning} and \emph{maximum margin IRL}.
\end{remark}

\begin{example}[Feature Matching in IRL]
\label{ex:irl}
Consider a grid-world MDP with features
$\phi(s,a) = (\indicator[\text{$s$ is near obstacle}],
\indicator[\text{$s$ is at goal}], 1)^\top$. An expert
demonstrates trajectories, from which we estimate
$\bar{\phi}^{\policy_E} \approx (0.1, 0.8, 5.0)^\top$.

An apprenticeship learning algorithm seeks a policy $\policy$
whose feature expectations $\bar{\phi}^\policy$ are close to
$\bar{\phi}^{\policy_E}$. In the occupancy measure framework,
this is:
\[
    \min_{x \in \mathcal{X}_\alpha}
    \left\|\sum_{s,a} \phi(s,a) x(s,a) - \bar{\phi}^{\policy_E}
    \right\|_2^2.
\]
This is a quadratic program over the occupancy polytope, which is
tractable and yields a policy that approximately matches the
expert's behavior.
\end{example}

%% ============================================================
%% SECTION 9: EXERCISES
%% ============================================================

\section{Exercises}
\label{sec:exercises}

\begin{exercise}[Primal LP Construction]
\textnormal{(\(*\))} \\
Consider the MDP with $\States = \{1, 2, 3\}$,
$\Actions = \{a, b\}$, $\discount = 0.9$:
\begin{center}
\begin{tabular}{cccccc}
\toprule
$s$ & $a$ & $R(s,a)$ & $P(1|s,a)$ & $P(2|s,a)$ & $P(3|s,a)$ \\
\midrule
1 & $a$ & 2 & 0.5 & 0.3 & 0.2 \\
1 & $b$ & 0 & 0 & 0.5 & 0.5 \\
2 & $a$ & 1 & 0.4 & 0.4 & 0.2 \\
2 & $b$ & 3 & 0 & 0 & 1 \\
3 & $a$ & $-1$ & 0.1 & 0.1 & 0.8 \\
3 & $b$ & 0 & 0 & 0.5 & 0.5 \\
\bottomrule
\end{tabular}
\end{center}
\begin{enumerate}[label=(\alph*)]
    \item Write out the primal LP with $\alpha = (1/3, 1/3, 1/3)$.
    \item For each deterministic policy, compute its value function
    and verify that the optimal one solves the LP.
\end{enumerate}
\end{exercise}

\begin{exercise}[Occupancy Measure Computation]
\textnormal{(\(*\))} \\
For the two-state MDP in Example~\ref{ex:primal-two-state} with
$\alpha = (0.5, 0.5)$:
\begin{enumerate}[label=(\alph*)]
    \item Compute the unnormalized occupancy measure
    $\tilde{x}^d$ for the policy $d(1) = a_1$, $d(2) = a_2$.
    \item Verify the flow conservation constraints.
    \item Compute the dual objective and compare with
    $\alpha^\top V^d$.
\end{enumerate}
\end{exercise}

\begin{exercise}[Vertices of the Occupancy Polytope]
\textnormal{(\(**\))} \\
For a two-state, two-action MDP with $\discount = 0.5$ and
$\alpha = (0.5, 0.5)$:
\begin{enumerate}[label=(\alph*)]
    \item How many vertices does $\mathcal{X}_\alpha$ have?
    \item Write out the four flow conservation constraints
    (two equalities, four non-negativity constraints).
    \item Enumerate all vertices by solving the linear system with
    each combination of two zero variables.
    \item Identify which vertex corresponds to which deterministic
    policy.
\end{enumerate}
\end{exercise}

\begin{exercise}[CMDP with a Budget Constraint]
\textnormal{(\(**\))} \\
Consider the MDP from Exercise~1 with an additional fuel cost
$C(s,a)$: $C(1,a) = 1$, $C(1,b) = 3$, $C(2,a) = 2$,
$C(2,b) = 1$, $C(3,a) = 0$, $C(3,b) = 2$. The budget is
$b = 3$.
\begin{enumerate}[label=(\alph*)]
    \item Write the CMDP as an LP.
    \item Check whether the unconstrained optimal policy satisfies
    the fuel constraint.
    \item If not, describe qualitatively how the constrained
    optimal policy differs from the unconstrained one.
\end{enumerate}
\end{exercise}

\begin{exercise}[Stochastic Policies from CMDPs]
\textnormal{(\(***\))} \\
Prove that for a CMDP with $K$ cost constraints, the optimal policy
randomizes over at most $K + 1$ actions at each state.

\textit{Hint:} Use the fact that the optimal solution of an LP lies
on a face of the feasible polytope, and count the number of
additional constraints (beyond the original $SA$ non-negativity
constraints) that can become active.
\end{exercise}

\begin{exercise}[Feature Expectations and IRL]
\textnormal{(\(***\))} \\
Consider a three-state, two-action MDP with features
$\phi(s,a) \in \R^2$. An expert policy $\policy_E$ has feature
expectation $\bar{\phi}^{\policy_E} = (3.0, 1.5)^\top$.
\begin{enumerate}[label=(\alph*)]
    \item Show that $\bar{\phi}^\policy$ is a linear function of
    the occupancy measure $\tilde{x}^\policy$.
    \item Prove that the set
    $\mathcal{F} = \{\bar{\phi}^\policy : \policy \text{ stationary}\}$
    is a convex polytope whose vertices correspond to deterministic
    policies.
    \item Explain geometrically why there always exists a weight
    vector $w$ such that $\policy_E$ is optimal for the reward
    $R(s,a) = w^\top \phi(s,a)$, provided $\bar{\phi}^{\policy_E}$
    lies on the boundary of $\mathcal{F}$.
\end{enumerate}
\end{exercise}

%% ============================================================
%% SECTION 10: SUMMARY
%% ============================================================

\section{Summary}
\label{sec:summary}

\begin{itemize}
    \item The \emph{primal LP} formulation casts finding $\Vstar$ as
    minimizing $\alpha^\top v$ subject to the Bellman feasibility
    constraints $v(s) \geq R(s,a) + \discount \sum_{s'} P(s'|s,a)
    v(s')$ for all $(s,a)$. The unique optimal solution is $\Vstar$.
    \item The \emph{dual LP} introduces occupancy measure variables
    $x(s,a) \geq 0$ and maximizes $\sum_{s,a} R(s,a) x(s,a)$
    subject to flow conservation constraints. The optimal dual
    solution is the occupancy measure of the optimal policy.
    \item \emph{Occupancy measures}
    $x^\policy(s,a) = (1-\discount) \sum_t \discount^t
    \Prob^\policy(S_t = s, A_t = a)$ encode the discounted
    visitation frequencies. They satisfy flow conservation, sum to
    1 (when normalized), and are non-negative.
    \item There is a \emph{one-to-one correspondence} between
    stationary policies and points in the occupancy polytope:
    $\policy(a|s) = x(s,a) / \sum_{a'} x(s,a')$.
    Deterministic policies correspond to vertices.
    \item \emph{LP strong duality} holds: the primal and dual
    optimal values are equal. Complementary slackness confirms that
    optimal occupancy measures concentrate on optimal actions.
    \item \emph{Constrained MDPs} are naturally handled by adding
    linear constraints $\sum_{s,a} C_k(s,a) x(s,a) \leq b_k$ to
    the dual LP. The optimal constrained policy may be stochastic,
    randomizing over at most $K + 1$ actions.
    \item \emph{Feature expectations} $\bar{\phi}^\policy =
    \sum_{s,a} \phi(s,a) \tilde{x}^\policy(s,a)$ provide the
    foundation for inverse RL and apprenticeship learning, where
    policy optimization is reduced to matching feature statistics.
\end{itemize}

Since this is the final lesson in the MDP subdomain, we provide a
summary of all eight MDP lessons.

\begin{center}
\begin{tabular}{clp{6.5cm}}
\toprule
\textbf{Lesson} & \textbf{Title} & \textbf{Key Results} \\
\midrule
39 & MDP Model and Policies &
    MDP tuple $(\States, \Actions, P, R, \discount)$; policy
    classes (Markov, stationary, deterministic); value functions
    $\Vpi$, $\Qpi$; Bellman expectation equations \\[4pt]
40 & Finite-Horizon MDPs &
    Backward induction; time-dependent optimal policy; dynamic
    programming principle \\[4pt]
41 & Discounted Reward and Bellman Equations &
    Bellman operators $\Bellman^\policy$, $\Bellman^*$;
    $\discount$-contraction; unique fixed points $\Vpi$, $\Vstar$;
    policy improvement theorem; greedy optimality \\[4pt]
42 & Contraction and Value Iteration &
    VI algorithm $v^{(n+1)} = \Bellman^* v^{(n)}$; geometric
    convergence at rate $\discount$; a priori/a posteriori bounds;
    approximate policy extraction \\[4pt]
43 & VI Error Bounds and Monotonicity &
    Span seminorm bounds; monotone VI sequences; improved stopping
    criteria; early policy convergence \\[4pt]
44 & Average Reward MDPs &
    Gain-bias equations; unichain theory; relative value iteration;
    $\discount \to 1$ limit; Blackwell optimality \\[4pt]
45 & Policy Iteration &
    Policy evaluation ($O(S^3)$ solve); policy improvement theorem;
    Howard's improvement; finite convergence ($\leq A^S$); modified
    PI; VI vs.\ PI comparison \\[4pt]
46 & LP Formulation and Occupancy Measures &
    Primal/dual LP; occupancy measures; policy-occupancy
    correspondence; strong duality; constrained MDPs; inverse
    RL connection \\
\bottomrule
\end{tabular}
\end{center}

%% ============================================================
%% SECTION 11: FURTHER READING
%% ============================================================

\section{Further Reading}
\label{sec:further-reading}

\begin{itemize}
    \item \textbf{Lesson~24 (TD Learning)} extends the MDP theory
    developed in Lessons~39--46 to the model-free setting, where
    the transition kernel $P$ and reward function $R$ are unknown.
    \item \textbf{Lesson~25 (Policy Gradient)} develops policy
    optimization methods that parameterize the policy directly,
    using gradient ascent on the expected return. The occupancy
    measure perspective provides an alternative derivation of the
    policy gradient theorem.
    \item \citet[Chapters 6 and 8--9]{puterman2014} provides the
    definitive treatment of LP formulations for both discounted and
    average reward MDPs, including multichain theory and sensitivity
    analysis.
    \item \citet{manne1960} is the original paper establishing the
    LP formulation; it is concise and elegant.
    \item \citet{altman1999} is the standard reference for
    constrained MDPs, covering LP methods, Lagrangian duality,
    sensitivity analysis, and applications.
    \item \citet{derman1970} provides an early treatment with
    applications to inventory and queuing problems.
    \item Ng and Russell (2000) introduced inverse RL from the
    occupancy measure perspective; Abbeel and Ng (2004) developed
    apprenticeship learning via feature expectation matching.
    These works directly build on the LP/occupancy framework of
    this lesson.
    \item For connections to convex optimization, see the treatment
    of LP duality in Lesson~18 (Convex Optimization and Duality).
\end{itemize}

\bibliography{mdp-refs}
\end{document}
