\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 45: Policy Iteration}}
\fancyhead[R]{\thepage}
\title{Lesson 45: Policy Iteration}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% ============================================================
%% SECTION 1: MOTIVATION AND OVERVIEW
%% ============================================================

\section{Motivation and Overview}
\label{sec:motivation}

Lesson~42 introduced value iteration as the natural algorithm arising
from the Banach fixed-point theorem: repeatedly apply the Bellman
optimality operator $\Bellman^*$ until convergence. While value
iteration is conceptually simple and widely used, its convergence rate
$\discount$ per iteration means that for $\discount$ close to 1 the
algorithm can require hundreds or thousands of iterations. Consider a
logistics MDP with a discount factor $\discount = 0.999$: the a priori
bound from Lesson~42 predicts roughly 10\,000 iterations to reach
$\varepsilon = 0.01$ accuracy. Can we do fundamentally better?

Policy iteration, introduced by \citet{howard1960}, takes a
radically different approach. Rather than iterating on value functions,
it iterates on \emph{policies}: evaluate the current policy exactly
(by solving a linear system), then improve the policy by acting
greedily with respect to the computed values. Because the number of
deterministic stationary policies is finite ($A^S$), this process must
terminate -- and it terminates at the optimal policy. In practice,
policy iteration converges in remarkably few iterations (often far
fewer than $A^S$), with each iteration making a large improvement.
The tools developed in this lesson -- the policy evaluation step,
the policy improvement theorem, Howard's improvement principle, and
modified policy iteration -- complete the algorithmic toolkit for
solving known MDPs.

\begin{rlconnection}[title={Why Policy Iteration Matters for RL/ML}]
\begin{itemize}[nosep]
    \item \textbf{Actor-critic methods}: The actor-critic architecture
    directly mirrors policy iteration: the critic performs approximate
    policy evaluation, and the actor performs (approximate) policy
    improvement.
    \item \textbf{Generalized policy iteration (GPI)}: Sutton and Barto
    identify the GPI principle -- alternating evaluation and
    improvement -- as the unifying idea behind nearly all RL algorithms.
    \item \textbf{Exact planning}: For moderate-sized MDPs with known
    models, policy iteration is often the algorithm of choice due to
    its rapid convergence, independent of $\discount$.
    \item \textbf{Natural policy gradient}: The natural policy gradient
    can be interpreted as a soft version of policy improvement in a
    parameterized policy space, connecting PI to modern deep RL.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
MDP model, stationary policies, value functions $\Vpi$ and $\Qpi$
(Lesson~39). Bellman expectation operator $\Bellman^\policy$, Bellman
optimality operator $\Bellman^*$, contraction properties, the policy
improvement theorem (Lesson~41, Theorem~7.3). Value iteration
algorithm, convergence, error bounds (Lesson~42). Linear systems and
matrix inversion (Lessons~5--6).

\paragraph{Learning objectives.}
After completing this lesson, the student will be able to:
\begin{itemize}
    \item \emph{Derive} the policy evaluation step as the solution of
    a linear system $v = (I - \discount P^\policy)^{-1} r^\policy$.
    \item \emph{State} and \emph{prove} the policy improvement theorem
    in its general form.
    \item \emph{Prove} that policy iteration converges to the optimal
    policy in at most $A^S$ iterations.
    \item \emph{Explain} Howard's improvement principle: strict
    improvement at every non-terminal iteration.
    \item \emph{Describe} modified policy iteration and \emph{prove}
    its convergence.
    \item \emph{Compare} the computational trade-offs between value
    iteration and policy iteration.
\end{itemize}

%% ============================================================
%% SECTION 2: REFERENCES
%% ============================================================

\section{References}
\label{sec:references}

\begin{itemize}
    \item \citet[Chapter 6, Sections 6.4--6.5]{puterman2014} --
    Definitive treatment of policy iteration, modified policy
    iteration, and convergence analysis for discounted MDPs.
    \item \citet{howard1960} -- Original monograph introducing
    policy iteration and proving its finite convergence.
    \item \citet[Chapter 1, Sections 1.3--1.4]{bertsekas2012} --
    Policy iteration, modified policy iteration, and connections
    to approximate dynamic programming.
    \item \citet[Chapter 4]{suttonbarto2018} -- Accessible
    presentation of policy iteration and the generalized policy
    iteration principle.
    \item \citet{denardo1967} -- Unified contraction framework
    placing both value iteration and policy iteration within a
    common operator-theoretic setting.
\end{itemize}

%% ============================================================
%% SECTION 3: POLICY EVALUATION
%% ============================================================

\section{Policy Evaluation}
\label{sec:policy-evaluation}

We work throughout with a finite MDP
$\mathcal{M} = (\States, \Actions, P, R, \discount)$ with
$|\States| = S$, $|\Actions| = A$,
$R_{\max} = \max_{s,a} |R(s,a)|$, and $\discount \in [0,1)$.
For a deterministic stationary policy $d : \States \to \Actions$,
we write $P^d \in \R^{S \times S}$ for the transition matrix
$P^d(s,s') = P(s' \mid s, d(s))$ and $r^d \in \R^S$ for the reward
vector $r^d(s) = R(s, d(s))$.

\begin{definition}[Policy Evaluation Problem]
\label{def:policy-eval}
Given a deterministic stationary policy $d$, the \emph{policy
evaluation problem} is to compute the value function
$V^d \in \R^S$ defined by
\[
    V^d(s) = \E\!\left[\sum_{t=0}^\infty \discount^t R(S_t, d(S_t))
    \;\middle|\; S_0 = s\right].
\]
\end{definition}

From Lesson~41, $V^d$ is the unique fixed point of the Bellman
expectation operator $\Bellman^d$:
\begin{equation}
\label{eq:bellman-eval}
    V^d = r^d + \discount P^d V^d.
\end{equation}
Rearranging gives the linear system
$(I - \discount P^d) V^d = r^d$.

\begin{theorem}[Policy Evaluation via Matrix Inversion]
\label{thm:policy-eval-matrix}
For any deterministic stationary policy $d$ and discount factor
$\discount \in [0,1)$, the matrix $I - \discount P^d$ is invertible
and
\begin{equation}
\label{eq:policy-eval-closed}
    V^d = (I - \discount P^d)^{-1} r^d.
\end{equation}
Moreover, $(I - \discount P^d)^{-1} = \sum_{k=0}^\infty
\discount^k (P^d)^k$, the \emph{discounted fundamental matrix}.
\end{theorem}

\begin{proof}
Since $P^d$ is a stochastic matrix, its spectral radius satisfies
$\rho(P^d) \leq 1$ (all eigenvalues have modulus at most 1). Thus
$\rho(\discount P^d) \leq \discount < 1$, which implies that
$I - \discount P^d$ is invertible and the Neumann series converges:
\[
    (I - \discount P^d)^{-1} = \sum_{k=0}^\infty (\discount P^d)^k
    = \sum_{k=0}^\infty \discount^k (P^d)^k.
\]
Substituting into \eqref{eq:bellman-eval}:
$V^d = (I - \discount P^d)^{-1} r^d$.
\end{proof}

\begin{remark}
\label{rem:eval-methods}
There are three practical methods for policy evaluation:
\begin{enumerate}[nosep]
    \item \textbf{Direct solve:} Solve the $S \times S$ linear
    system $(I - \discount P^d) v = r^d$ using Gaussian elimination
    or LU decomposition. Cost: $O(S^3)$.
    \item \textbf{Iterative (successive approximation):} Compute
    $v^{(j+1)} = r^d + \discount P^d v^{(j)}$ starting from any
    $v^{(0)}$. This converges geometrically at rate $\discount$
    since $\Bellman^d$ is a $\discount$-contraction (Lesson~41).
    Cost per iteration: $O(S^2)$.
    \item \textbf{Neumann series truncation:} Compute
    $V^d \approx \sum_{k=0}^{m-1} \discount^k (P^d)^k r^d$.
    This is equivalent to $m$ steps of iterative evaluation from
    $v^{(0)} = 0$.
\end{enumerate}
For small to moderate $S$, the direct solve is preferred. For large
$S$, iterative methods or sparse solvers are necessary.
\end{remark}

\begin{example}[Policy Evaluation on a Three-State MDP]
\label{ex:eval-three-state}
Consider an MDP with $\States = \{1, 2, 3\}$,
$\Actions = \{L, R\}$, $\discount = 0.9$, and transition/reward
structure:
\begin{center}
\begin{tabular}{cccccc}
\toprule
$s$ & $a$ & $R(s,a)$ & $P(1 \mid s,a)$ & $P(2 \mid s,a)$ &
$P(3 \mid s,a)$ \\
\midrule
1 & L & 2 & 0.5 & 0.3 & 0.2 \\
1 & R & 0 & 0 & 0.5 & 0.5 \\
2 & L & 1 & 0.4 & 0.4 & 0.2 \\
2 & R & 3 & 0 & 0 & 1 \\
3 & L & $-1$ & 0.1 & 0.1 & 0.8 \\
3 & R & 0 & 0 & 0.5 & 0.5 \\
\bottomrule
\end{tabular}
\end{center}

Evaluate policy $d(1) = L$, $d(2) = R$, $d(3) = L$. The induced
chain has $r^d = (2, 3, -1)^\top$ and
\[
    P^d = \begin{pmatrix}
    0.5 & 0.3 & 0.2 \\
    0 & 0 & 1 \\
    0.1 & 0.1 & 0.8
    \end{pmatrix}.
\]
The linear system $(I - 0.9 P^d) v = r^d$ is:
\[
    \begin{pmatrix}
    0.55 & -0.27 & -0.18 \\
    0 & 1 & -0.9 \\
    -0.09 & -0.09 & 0.28
    \end{pmatrix}
    \begin{pmatrix} v_1 \\ v_2 \\ v_3 \end{pmatrix}
    = \begin{pmatrix} 2 \\ 3 \\ -1 \end{pmatrix}.
\]
Solving (e.g., by back-substitution after row reduction) yields
approximately $V^d \approx (15.52, 15.38, 9.31)^\top$.
\end{example}

We now establish a key structural property of the policy evaluation
matrix that will be used in the analysis of policy iteration.

\begin{proposition}[Non-negativity of the Discounted Fundamental Matrix]
\label{prop:fundamental-nonneg}
For any deterministic stationary policy $d$ and $\discount \in [0,1)$,
$(I - \discount P^d)^{-1}$ has all non-negative entries.
\end{proposition}

\begin{proof}
By Theorem~\ref{thm:policy-eval-matrix},
$(I - \discount P^d)^{-1} = \sum_{k=0}^\infty \discount^k (P^d)^k$.
Since $P^d$ has non-negative entries and $\discount \geq 0$, every
term $\discount^k (P^d)^k$ has non-negative entries, and the sum
of non-negative matrices is non-negative.
\end{proof}

This non-negativity property will be essential in proving that policy
improvement preserves value ordering.

%% ============================================================
%% SECTION 4: THE POLICY IMPROVEMENT THEOREM
%% ============================================================

\section{The Policy Improvement Theorem}
\label{sec:policy-improvement}

The policy improvement theorem, first established in
\citet{howard1960}, is the theoretical engine driving policy iteration.
We restate and prove it here in a self-contained form.

\begin{theorem}[Policy Improvement Theorem]
\label{thm:pi-theorem}
Let $d$ be a deterministic stationary policy and define the improved
policy $d'$ by
\begin{equation}
\label{eq:greedy-improve}
    d'(s) = \argmax_{a \in \Actions} Q^d(s,a)
    = \argmax_{a \in \Actions} \left[ R(s,a) + \discount
    \sum_{s' \in \States} P(s' \mid s,a)\, V^d(s') \right]
\end{equation}
for all $s \in \States$ (breaking ties arbitrarily). Then:
\begin{enumerate}
    \item $V^{d'}(s) \geq V^d(s)$ for all $s \in \States$.
    \item $V^{d'} = V^d$ if and only if $d$ is already optimal
    (i.e., $V^d = \Vstar$).
\end{enumerate}
\end{theorem}

\begin{proof}
\textbf{Part 1.} By definition of $d'$, for every state $s$:
\begin{equation}
\label{eq:improve-step}
    Q^d(s, d'(s)) = \max_{a \in \Actions} Q^d(s,a)
    \geq Q^d(s, d(s)) = V^d(s).
\end{equation}
In operator notation, $\Bellman^{d'} V^d \geq V^d$ componentwise,
where the inequality is \eqref{eq:improve-step}. Define the vector
$\delta = \Bellman^{d'} V^d - V^d \geq 0$. We now show that
$V^{d'} \geq V^d$.

Applying $\Bellman^{d'}$ repeatedly using its monotonicity
(Lesson~42, Proposition~8.1 for $\Bellman^\policy$):
\begin{align}
    V^d &\leq \Bellman^{d'} V^d  \label{eq:chain1} \\
    &\leq (\Bellman^{d'})^2 V^d  \label{eq:chain2} \\
    &\leq \cdots \nonumber \\
    &\leq \lim_{n \to \infty} (\Bellman^{d'})^n V^d = V^{d'}.
    \label{eq:chain-limit}
\end{align}
The limit in \eqref{eq:chain-limit} equals $V^{d'}$ because
$\Bellman^{d'}$ is a $\discount$-contraction with unique fixed
point $V^{d'}$ (Lesson~41), so $(\Bellman^{d'})^n v \to V^{d'}$
for any starting $v$.

\textbf{Part 2 (``only if'' direction).} Suppose $V^{d'} = V^d$.
From \eqref{eq:improve-step}, $\max_a Q^d(s,a) = V^d(s)$ for all
$s$. This is precisely the Bellman optimality equation, so
$V^d = \Vstar$.

\textbf{Part 2 (``if'' direction).} If $V^d = \Vstar$, then $d$ is
optimal, and $d'$ (being greedy w.r.t.\ $\Vstar$) is also optimal.
Hence $V^{d'} = \Vstar = V^d$.
\end{proof}

\begin{keyresult}[title={Policy Improvement Guarantee}]
\label{kr:pi-guarantee}
The policy improvement theorem guarantees that the greedy policy
$d'$ with respect to $V^d$ is at least as good as $d$ in every
state. If $d$ is not optimal, then the improvement is \emph{strict}
in at least one state (by Howard's improvement principle,
Theorem~\ref{thm:howard-strict} below).
\end{keyresult}

We now provide an alternative, algebraic proof of the policy
improvement theorem that uses the matrix structure directly and
gives additional insight.

\begin{proposition}[Algebraic Proof of Policy Improvement]
\label{prop:algebraic-pi}
Under the conditions of Theorem~\ref{thm:pi-theorem},
\begin{equation}
\label{eq:algebraic-pi}
    V^{d'} - V^d = (I - \discount P^{d'})^{-1} \delta,
\end{equation}
where $\delta = \Bellman^{d'} V^d - V^d \geq 0$.
\end{proposition}

\begin{proof}
Since $V^{d'} = r^{d'} + \discount P^{d'} V^{d'}$ and
$\Bellman^{d'} V^d = r^{d'} + \discount P^{d'} V^d$:
\begin{align*}
    V^{d'} - V^d &= r^{d'} + \discount P^{d'} V^{d'} - V^d \\
    &= (\Bellman^{d'} V^d - V^d) + \discount P^{d'}(V^{d'} - V^d) \\
    &= \delta + \discount P^{d'}(V^{d'} - V^d).
\end{align*}
Rearranging: $(I - \discount P^{d'})(V^{d'} - V^d) = \delta$, so
$V^{d'} - V^d = (I - \discount P^{d'})^{-1} \delta$.
By Proposition~\ref{prop:fundamental-nonneg},
$(I - \discount P^{d'})^{-1}$ has non-negative entries and
$\delta \geq 0$, so $V^{d'} - V^d \geq 0$.
\end{proof}

\begin{remark}
\label{rem:algebraic-insight}
The algebraic form \eqref{eq:algebraic-pi} shows that policy
improvement has a ``cascading'' effect: a strict advantage
$\delta(s) > 0$ at state $s$ propagates to all states reachable
from $s$ under the new policy $d'$, weighted by the discounted
visitation probabilities encoded in $(I - \discount P^{d'})^{-1}$.
\end{remark}

\begin{example}[Policy Improvement Step]
\label{ex:pi-step}
Using the three-state MDP from Example~\ref{ex:eval-three-state}
with $V^d \approx (15.52, 15.38, 9.31)^\top$ under
$d(1) = L, d(2) = R, d(3) = L$:

Compute $Q^d(s,a)$ for the alternative actions:
\begin{align*}
    Q^d(1, R) &= 0 + 0.9(0.5 \cdot 15.38 + 0.5 \cdot 9.31)
    = 0.9 \cdot 12.345 = 11.11, \\
    Q^d(1, L) &= V^d(1) = 15.52.
\end{align*}
So $d'(1) = L$ (no change).
\begin{align*}
    Q^d(2, L) &= 1 + 0.9(0.4 \cdot 15.52 + 0.4 \cdot 15.38
    + 0.2 \cdot 9.31) \\
    &= 1 + 0.9(6.208 + 6.152 + 1.862) = 1 + 12.80 = 13.80, \\
    Q^d(2, R) &= V^d(2) = 15.38.
\end{align*}
So $d'(2) = R$ (no change).
\begin{align*}
    Q^d(3, R) &= 0 + 0.9(0.5 \cdot 15.38 + 0.5 \cdot 9.31)
    = 11.11, \\
    Q^d(3, L) &= V^d(3) = 9.31.
\end{align*}
So $d'(3) = R$ (improvement!). The improvement vector is
$\delta = (0, 0, 11.11 - 9.31)^\top = (0, 0, 1.80)^\top$.
The strict improvement at state 3 will cascade to other states
via the reachability structure.
\end{example}

%% ============================================================
%% SECTION 5: THE POLICY ITERATION ALGORITHM
%% ============================================================

\section{The Policy Iteration Algorithm}
\label{sec:pi-algorithm}

\begin{definition}[Policy Iteration]
\label{def:pi}
The \emph{policy iteration} (PI) algorithm generates a sequence of
deterministic stationary policies $\{d^{(k)}\}_{k \geq 0}$:
\begin{enumerate}[nosep]
    \item \textbf{Initialize:} Choose any deterministic stationary
    policy $d^{(0)}$.
    \item \textbf{Policy evaluation:} Compute $V^{d^{(k)}}$ by
    solving
    \begin{equation}
    \label{eq:pi-eval}
        V^{d^{(k)}} = (I - \discount P^{d^{(k)}})^{-1}
        r^{d^{(k)}}.
    \end{equation}
    \item \textbf{Policy improvement:} For each $s \in \States$,
    set
    \begin{equation}
    \label{eq:pi-improve}
        d^{(k+1)}(s) = \argmax_{a \in \Actions}
        \left[ R(s,a) + \discount \sum_{s'} P(s' \mid s,a)\,
        V^{d^{(k)}}(s') \right].
    \end{equation}
    \item \textbf{Stopping:} If $d^{(k+1)} = d^{(k)}$, stop and
    output $d^{(k)}$. Otherwise, set $k \leftarrow k+1$ and go to
    step 2.
\end{enumerate}
\end{definition}

\begin{remark}
\label{rem:pi-complexity}
Each iteration of policy iteration consists of two phases:
\begin{itemize}[nosep]
    \item \textbf{Policy evaluation:} Solving an $S \times S$ linear
    system costs $O(S^3)$ by direct methods. If sparse solvers are
    used, this can be reduced, but the worst case is cubic.
    \item \textbf{Policy improvement:} Computing the greedy action at
    each state requires $O(SA)$ operations (evaluating $A$ actions at
    each of $S$ states, each requiring an $O(S)$ inner product).
    The total cost is $O(S^2 A)$.
\end{itemize}
Thus each PI iteration costs $O(S^3 + S^2 A)$, compared to
$O(S^2 A)$ per VI iteration.
\end{remark}

\begin{rlconnection}[title={Generalized Policy Iteration}]
\citet{suttonbarto2018} identify \emph{generalized policy iteration}
(GPI) as the overarching principle: any RL algorithm that interleaves
policy evaluation (estimating $V^\policy$ or $Q^\policy$) and policy
improvement (making the policy greedy or nearly greedy w.r.t.\ the
estimated values) is performing GPI. Model-free methods like SARSA,
Q-learning, and actor-critic all follow this pattern, differing only
in how they approximate the evaluation step.
\end{rlconnection}

%% ============================================================
%% SECTION 6: FINITE CONVERGENCE OF PI
%% ============================================================

\section{Finite Convergence of Policy Iteration}
\label{sec:pi-convergence}

\begin{theorem}[Finite Convergence of Policy Iteration]
\label{thm:pi-finite}
Policy iteration (Definition~\ref{def:pi}) terminates after at most
$|\Pi^{\mathrm{SD}}| = A^S$ iterations, where
$\Pi^{\mathrm{SD}}$ denotes the set of all deterministic stationary
policies. Upon termination, the output policy $d^{(k)}$ is optimal:
$V^{d^{(k)}} = \Vstar$.
\end{theorem}

\begin{proof}
By the policy improvement theorem (Theorem~\ref{thm:pi-theorem}),
$V^{d^{(k+1)}}(s) \geq V^{d^{(k)}}(s)$ for all $s$ and $k$.

\textbf{Claim:} No policy repeats, i.e., $d^{(j)} \neq d^{(k)}$
for $j \neq k$.

Suppose for contradiction that $d^{(j)} = d^{(k)}$ for some
$j < k$. Then $V^{d^{(j)}} = V^{d^{(k)}}$. Since
the value sequence is non-decreasing, we must have
$V^{d^{(i)}} = V^{d^{(j)}}$ for all $j \leq i \leq k$.
In particular, $V^{d^{(j+1)}} = V^{d^{(j)}}$, which by
Theorem~\ref{thm:pi-theorem} implies $d^{(j)}$ is optimal and the
algorithm would have terminated at iteration $j$ (since
$d^{(j+1)} = d^{(j)}$), contradicting $j < k$.

Since there are exactly $A^S$ deterministic stationary policies and
no policy repeats, the algorithm terminates in at most $A^S$
iterations.

At termination, $d^{(k+1)} = d^{(k)}$, so by
Theorem~\ref{thm:pi-theorem}, $V^{d^{(k)}} = \Vstar$ and
$d^{(k)}$ is optimal.
\end{proof}

\begin{warning}[title={The $A^S$ Bound Is Rarely Tight}]
The bound $A^S$ is a worst-case guarantee. In practice, policy
iteration typically converges in a small number of iterations,
often $O(S)$ or even $O(\log S)$. However, there exist pathological
MDPs where PI requires an exponential number of iterations. Despite
this, such worst cases are extremely rare, and PI is considered
highly efficient in practice. The exact complexity of policy
iteration remains an open problem in computational complexity.
\end{warning}

\begin{theorem}[Strict Monotonicity of Value Sequence]
\label{thm:pi-strict-mono}
If policy iteration does not terminate at iteration $k$ (i.e.,
$d^{(k+1)} \neq d^{(k)}$), then
$V^{d^{(k+1)}}(s) > V^{d^{(k)}}(s)$ for at least one state
$s \in \States$.
\end{theorem}

\begin{proof}
If $d^{(k+1)} \neq d^{(k)}$, then there exists some state $s_0$
where $d^{(k+1)}(s_0) \neq d^{(k)}(s_0)$. At this state, the
greedy action $d^{(k+1)}(s_0)$ achieves a strictly higher
action-value:
\[
    Q^{d^{(k)}}(s_0, d^{(k+1)}(s_0))
    > Q^{d^{(k)}}(s_0, d^{(k)}(s_0)) = V^{d^{(k)}}(s_0),
\]
where the strict inequality holds because if equality held for all
maximizing actions, then $d^{(k)}(s_0)$ would also be a maximizer
and the tie-breaking rule would not change the action.

More precisely, since $d^{(k+1)}(s_0) \neq d^{(k)}(s_0)$ and
$d^{(k+1)}(s_0)$ is chosen to maximize $Q^{d^{(k)}}(s_0, \cdot)$,
we have
$Q^{d^{(k)}}(s_0, d^{(k+1)}(s_0)) \geq Q^{d^{(k)}}(s_0,
d^{(k)}(s_0)) = V^{d^{(k)}}(s_0)$.

So $\delta(s_0) = (\Bellman^{d^{(k+1)}} V^{d^{(k)}})(s_0) -
V^{d^{(k)}}(s_0) \geq 0$, and by the argument above
$\delta(s_0) \geq 0$. By the algebraic form
(Proposition~\ref{prop:algebraic-pi}):
\[
    V^{d^{(k+1)}} - V^{d^{(k)}} =
    (I - \discount P^{d^{(k+1)}})^{-1} \delta \geq 0.
\]
Since $\delta \geq 0$ and $\delta \neq 0$ (we have
$\delta(s_0) \geq 0$ and the improvement at $s_0$ is non-trivial),
and $(I - \discount P^{d^{(k+1)}})^{-1}$ has strictly positive
diagonal entries (the $(s_0, s_0)$ entry is at least 1), we get
$V^{d^{(k+1)}}(s_0) > V^{d^{(k)}}(s_0)$.
\end{proof}

We now state Howard's improvement principle, which strengthens the
strict monotonicity result.

\begin{theorem}[Howard's Improvement Principle]
\label{thm:howard-strict}
If $d^{(k+1)} \neq d^{(k)}$ in policy iteration, then for every
state $s$ that is reachable from a changed state under the new
policy $d^{(k+1)}$, the value function strictly improves:
$V^{d^{(k+1)}}(s) > V^{d^{(k)}}(s)$.
\end{theorem}

\begin{proof}
Let $C = \{s \in \States : d^{(k+1)}(s) \neq d^{(k)}(s)\}$
be the set of changed states, and define
$\delta = \Bellman^{d^{(k+1)}} V^{d^{(k)}} - V^{d^{(k)}} \geq 0$.
For $s \in C$, $\delta(s) \geq 0$, and for $s \notin C$,
$\delta(s) = 0$ (since the action did not change).

By Proposition~\ref{prop:algebraic-pi}:
\[
    V^{d^{(k+1)}}(s) - V^{d^{(k)}}(s) =
    \sum_{s'} \left[(I - \discount P^{d^{(k+1)}})^{-1}\right]_{s,s'}
    \delta(s').
\]
The entry $[(I - \discount P^{d^{(k+1)}})^{-1}]_{s,s'}
= \sum_{t=0}^\infty \discount^t (P^{d^{(k+1)}})^t_{s,s'}$ is the
discounted probability of visiting $s'$ starting from $s$ under
$d^{(k+1)}$. This entry is strictly positive if and only if $s'$
is reachable from $s$ under $d^{(k+1)}$.

Therefore, if $s$ can reach some state $s' \in C$ under $d^{(k+1)}$,
and $\delta(s') > 0$, then $V^{d^{(k+1)}}(s) > V^{d^{(k)}}(s)$.
Since $\delta(s') \geq 0$ for $s' \in C$ and $d^{(k+1)}$ is greedy,
$\delta(s') > 0$ for at least one $s' \in C$ unless $V^{d^{(k)}}$
already satisfies the Bellman optimality equation (in which case the
algorithm would have terminated). Thus the value strictly improves at
every state reachable from $C$.
\end{proof}

\begin{keyresult}[title={Howard's Improvement Principle}]
At each iteration of policy iteration, either:
\begin{enumerate}[nosep]
    \item The algorithm terminates with the optimal policy, or
    \item The value function strictly improves at every state that
    can reach (under the new policy) a state where the action
    changed.
\end{enumerate}
In ergodic MDPs (where every state is reachable from every other
state under any policy), this means the value improves at
\emph{every} state at every non-terminal iteration.
\end{keyresult}

\begin{example}[Complete Policy Iteration Trace]
\label{ex:pi-trace}
We trace policy iteration on the two-state MDP from Lesson~39 with
$\States = \{s_1, s_2\}$, $\Actions = \{a_1, a_2\}$,
$\discount = 0.5$:
\begin{center}
\begin{tabular}{ccccc}
\toprule
$s$ & $a$ & $R(s,a)$ & $P(s_1 \mid s,a)$ & $P(s_2 \mid s,a)$ \\
\midrule
$s_1$ & $a_1$ & 5 & 0.5 & 0.5 \\
$s_1$ & $a_2$ & 10 & 0 & 1 \\
$s_2$ & $a_1$ & $-1$ & 0.8 & 0.2 \\
$s_2$ & $a_2$ & 2 & 0.3 & 0.7 \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Iteration 0.} Start with $d^{(0)}(s_1) = a_1$,
$d^{(0)}(s_2) = a_1$.

\emph{Evaluation:}
$P^{d^{(0)}} = \bigl(\begin{smallmatrix} 0.5 & 0.5 \\
0.8 & 0.2 \end{smallmatrix}\bigr)$,
$r^{d^{(0)}} = (5, -1)^\top$.
Solving $(I - 0.5 P^{d^{(0)}}) v = r^{d^{(0)}}$:
\[
    \begin{pmatrix} 0.75 & -0.25 \\ -0.4 & 0.9 \end{pmatrix}
    v = \begin{pmatrix} 5 \\ -1 \end{pmatrix}.
\]
Determinant $= 0.675 - 0.1 = 0.575$. Solving:
$v_1 = (5 \cdot 0.9 + 0.25)/0.575 = 4.75/0.575 \approx 8.26$,
$v_2 = (0.75 \cdot (-1) + 0.4 \cdot 5)/0.575 = 1.25/0.575
\approx 2.17$.
So $V^{d^{(0)}} \approx (8.26, 2.17)^\top$.

\emph{Improvement:}
$Q^{d^{(0)}}(s_1, a_2) = 10 + 0.5 \cdot 2.17 = 11.09 > 8.26$,
so $d^{(1)}(s_1) = a_2$.
$Q^{d^{(0)}}(s_2, a_2) = 2 + 0.5(0.3 \cdot 8.26 + 0.7 \cdot 2.17)
= 2 + 0.5(2.478 + 1.519) = 4.00 > 2.17$,
so $d^{(1)}(s_2) = a_2$.

\textbf{Iteration 1.} $d^{(1)}(s_1) = a_2$, $d^{(1)}(s_2) = a_2$.

\emph{Evaluation:}
$P^{d^{(1)}} = \bigl(\begin{smallmatrix} 0 & 1 \\
0.3 & 0.7 \end{smallmatrix}\bigr)$,
$r^{d^{(1)}} = (10, 2)^\top$.
Solving $(I - 0.5 P^{d^{(1)}}) v = r^{d^{(1)}}$:
\[
    \begin{pmatrix} 1 & -0.5 \\ -0.15 & 0.65 \end{pmatrix}
    v = \begin{pmatrix} 10 \\ 2 \end{pmatrix}.
\]
Determinant $= 0.65 - 0.075 = 0.575$.
$v_1 = (10 \cdot 0.65 + 0.5 \cdot 2)/0.575 = 7.5/0.575
\approx 13.04$,
$v_2 = (1 \cdot 2 + 0.15 \cdot 10)/0.575 = 3.5/0.575
\approx 6.09$.
So $V^{d^{(1)}} \approx (13.04, 6.09)^\top$.

\emph{Improvement:}
$Q^{d^{(1)}}(s_1, a_1) = 5 + 0.5(0.5 \cdot 13.04 + 0.5 \cdot 6.09)
= 5 + 4.78 = 9.78 < 13.04$, so $d^{(2)}(s_1) = a_2$.
$Q^{d^{(1)}}(s_2, a_1) = -1 + 0.5(0.8 \cdot 13.04 + 0.2 \cdot 6.09)
= -1 + 5.82 = 4.82 < 6.09$, so $d^{(2)}(s_2) = a_2$.

Since $d^{(2)} = d^{(1)}$, the algorithm terminates.
$\optpolicy = d^{(1)}$ with $\Vstar \approx (13.04, 6.09)^\top$.
Policy iteration converged in 2 iterations (1 improvement step).
\end{example}

%% ============================================================
%% SECTION 7: MODIFIED POLICY ITERATION
%% ============================================================

\section{Modified Policy Iteration}
\label{sec:modified-pi}

Exact policy evaluation requires solving an $S \times S$ linear
system at each iteration, which costs $O(S^3)$. When $S$ is large,
this can be the bottleneck. Modified policy iteration replaces the
exact solve with a fixed number of Bellman backups, interpolating
between value iteration and exact policy iteration.

\begin{definition}[Modified Policy Iteration]
\label{def:modified-pi}
\emph{Modified policy iteration} with parameter $m \geq 1$ generates
a sequence of value functions $\{v^{(k)}\}$ and policies
$\{d^{(k)}\}$:
\begin{enumerate}[nosep]
    \item \textbf{Initialize:} Choose $v^{(0)} \in \R^S$ and set
    $d^{(1)}$ as the greedy policy w.r.t.\ $v^{(0)}$.
    \item \textbf{Truncated evaluation:} Starting from
    $v^{(k,0)} = v^{(k-1)}$, perform $m$ Bellman backups:
    \begin{equation}
    \label{eq:trunc-eval}
        v^{(k,j+1)} = \Bellman^{d^{(k)}} v^{(k,j)}
        = r^{d^{(k)}} + \discount P^{d^{(k)}} v^{(k,j)},
        \quad j = 0, 1, \ldots, m-1.
    \end{equation}
    Set $v^{(k)} = v^{(k,m)}$.
    \item \textbf{Policy improvement:} Set $d^{(k+1)}$ as the greedy
    policy w.r.t.\ $v^{(k)}$.
    \item \textbf{Stopping:} If $d^{(k+1)} = d^{(k)}$ or
    $\|v^{(k)} - v^{(k-1)}\|_\infty < \varepsilon(1-\discount)/
    \discount$, stop. Otherwise, increment $k$ and go to step 2.
\end{enumerate}
\end{definition}

\begin{proposition}[Extremes of Modified PI]
\label{prop:mpi-extremes}
Modified policy iteration with $m = 1$ and $v^{(0)} = 0$ reduces
to value iteration. In the limit $m \to \infty$, it reduces to
exact policy iteration.
\end{proposition}

\begin{proof}
\textbf{Case $m = 1$:} At each iteration, we perform one Bellman
backup $v^{(k)} = \Bellman^{d^{(k)}} v^{(k-1)}$ and then form
$d^{(k+1)} = \text{greedy}(v^{(k)})$. But
$\Bellman^{d^{(k)}} v^{(k-1)} = r^{d^{(k)}} + \discount P^{d^{(k)}}
v^{(k-1)}$ where $d^{(k)} = \argmax_a [R(s,a) + \discount
\sum_{s'} P(s'|s,a) v^{(k-1)}(s')]$.
Thus $(\Bellman^{d^{(k)}} v^{(k-1)})(s)
= \max_a [R(s,a) + \discount \sum_{s'} P(s'|s,a) v^{(k-1)}(s')]
= (\Bellman^* v^{(k-1)})(s)$,
so $v^{(k)} = \Bellman^* v^{(k-1)}$, which is exactly value iteration.

\textbf{Case $m \to \infty$:} Since $\Bellman^{d^{(k)}}$ is a
$\discount$-contraction with fixed point $V^{d^{(k)}}$, the
iterates $v^{(k,j)} \to V^{d^{(k)}}$ as $j \to \infty$. In the
limit, we evaluate $V^{d^{(k)}}$ exactly, recovering policy iteration.
\end{proof}

\begin{theorem}[Convergence of Modified Policy Iteration]
\label{thm:mpi-convergence}
For any fixed $m \geq 1$ and any initialization $v^{(0)}$, modified
policy iteration produces value functions satisfying
\[
    \|v^{(k)} - \Vstar\|_\infty \to 0 \quad \text{as } k \to \infty.
\]
Moreover, for each $k \geq 1$:
\begin{equation}
\label{eq:mpi-bound}
    \|v^{(k)} - \Vstar\|_\infty \leq \discount^m
    \|v^{(k-1)} - \Vstar\|_\infty.
\end{equation}
\end{theorem}

\begin{proof}
Since $d^{(k)}$ is greedy w.r.t.\ $v^{(k-1)}$, we have
$\Bellman^{d^{(k)}} v^{(k-1)} = \Bellman^* v^{(k-1)}$. After $m$
Bellman backups starting from $v^{(k,0)} = v^{(k-1)}$:
\begin{align*}
    \|v^{(k)} - V^{d^{(k)}}\|_\infty
    &= \|(\Bellman^{d^{(k)}})^m v^{(k-1)} - V^{d^{(k)}}\|_\infty \\
    &\leq \discount^m \|v^{(k-1)} - V^{d^{(k)}}\|_\infty,
\end{align*}
since $\Bellman^{d^{(k)}}$ is a $\discount$-contraction with fixed
point $V^{d^{(k)}}$.

Now, $V^{d^{(k)}} \geq V^{d^{(k-1)}}$ by the policy improvement
theorem (since $d^{(k)}$ is greedy w.r.t.\ $V^{d^{(k-1)}}$, and
a fortiori w.r.t.\ the less-accurate $v^{(k-1)}$). However, we
also need to bound $\|v^{(k-1)} - V^{d^{(k)}}\|_\infty$.

Using the triangle inequality:
\[
    \|v^{(k-1)} - V^{d^{(k)}}\|_\infty
    \leq \|v^{(k-1)} - \Vstar\|_\infty +
    \|V^{d^{(k)}} - \Vstar\|_\infty.
\]
Since $V^{d^{(k)}} \leq \Vstar$, we get
$\|v^{(k-1)} - V^{d^{(k)}}\|_\infty \leq
2\|v^{(k-1)} - \Vstar\|_\infty$.

On the other hand, using $\Bellman^{d^{(k)}} v^{(k-1)} = \Bellman^*
v^{(k-1)}$:
\begin{align*}
    \|v^{(k)} - \Vstar\|_\infty
    &= \|(\Bellman^{d^{(k)}})^m v^{(k-1)} - \Vstar\|_\infty \\
    &\leq \|(\Bellman^{d^{(k)}})^m v^{(k-1)}
    - (\Bellman^{d^{(k)}})^m \Vstar\|_\infty
    + \|(\Bellman^{d^{(k)}})^m \Vstar - \Vstar\|_\infty.
\end{align*}
The first term is $\leq \discount^m
\|v^{(k-1)} - \Vstar\|_\infty$ by contraction. For the second term,
$(\Bellman^{d^{(k)}})^m \Vstar \to V^{d^{(k)}}$ and
$\|V^{d^{(k)}} - \Vstar\|_\infty \leq
\|v^{(k-1)} - \Vstar\|_\infty \cdot 2\discount/(1-\discount)$
by the approximate policy bound (Lesson~42, Theorem~5.1). However,
for a cleaner bound, note that
\begin{align*}
    \|(\Bellman^{d^{(k)}})^m \Vstar - \Vstar\|_\infty
    &\leq \discount^m \|\Bellman^{d^{(k)}} \Vstar -
    \Vstar\|_\infty / (1-\discount) \cdot (1-\discount) \\
    &= \discount^m \|\Vstar - \Vstar\|_\infty \cdot \ldots
\end{align*}
We use a more direct approach. Note
$\Bellman^* \Vstar = \Vstar$ and
$\Bellman^{d^{(k)}} \Vstar \leq \Bellman^* \Vstar = \Vstar$.
Also $\Bellman^{d^{(k)}} V^{d^{(k)}} = V^{d^{(k)}} \leq \Vstar$.
By monotonicity:
\[
    V^{d^{(k)}} \leq (\Bellman^{d^{(k)}})^m \Vstar \leq \Vstar.
\]
Therefore:
\begin{align*}
    \|v^{(k)} - \Vstar\|_\infty
    &\leq \discount^m \|v^{(k-1)} - \Vstar\|_\infty
    + \|\Vstar - V^{d^{(k)}}\|_\infty.
\end{align*}
Since $\|\Vstar - V^{d^{(k)}}\|_\infty \leq
\frac{2\discount}{1-\discount} \|v^{(k-1)} - \Vstar\|_\infty$
when $d^{(k)}$ is greedy w.r.t.\ an $\|v^{(k-1)} - \Vstar\|_\infty$-
approximation to $\Vstar$ (by Lesson~42, Theorem~5.1), the overall
bound is:
\[
    \|v^{(k)} - \Vstar\|_\infty \leq
    \left(\discount^m + \frac{2\discount}{1-\discount}\right)
    \|v^{(k-1)} - \Vstar\|_\infty.
\]
However, this factor exceeds 1 for large $\discount$. Instead, we
establish the simpler geometric convergence directly. Since
$\Bellman^{d^{(k)}} v^{(k-1)} = \Bellman^* v^{(k-1)}$ and
$\Bellman^* \Vstar = \Vstar$:
\begin{align*}
    v^{(k,1)} &= \Bellman^* v^{(k-1)}, \\
    \|v^{(k,1)} - \Vstar\|_\infty
    &= \|\Bellman^* v^{(k-1)} - \Bellman^* \Vstar\|_\infty
    \leq \discount \|v^{(k-1)} - \Vstar\|_\infty.
\end{align*}
For subsequent backups ($j \geq 1$),
$v^{(k,j+1)} = \Bellman^{d^{(k)}} v^{(k,j)}$ and
\begin{align*}
    \|v^{(k,j+1)} - \Vstar\|_\infty
    &\leq \|\Bellman^{d^{(k)}} v^{(k,j)} - \Bellman^{d^{(k)}}
    \Vstar\|_\infty + \|\Bellman^{d^{(k)}} \Vstar - \Vstar\|_\infty \\
    &\leq \discount \|v^{(k,j)} - \Vstar\|_\infty
    + \|\Bellman^{d^{(k)}} \Vstar - \Vstar\|_\infty.
\end{align*}
Since $\Bellman^{d^{(k)}} \Vstar \leq \Bellman^* \Vstar = \Vstar$
and $\Bellman^{d^{(k)}} \Vstar \geq V^{d^{(k)}}$ is not immediate
in this form, we instead observe that
$\|v^{(k,j+1)} - V^{d^{(k)}}\|_\infty \leq \discount
\|v^{(k,j)} - V^{d^{(k)}}\|_\infty$ (pure contraction toward
$V^{d^{(k)}}$), giving after $m$ steps:
\[
    \|v^{(k)} - V^{d^{(k)}}\|_\infty \leq \discount^m
    \|v^{(k-1)} - V^{d^{(k)}}\|_\infty.
\]
Since the policies $d^{(k)}$ improve monotonically and there are
finitely many, the sequence eventually stabilizes at $\optpolicy$,
after which $v^{(k)} \to \Vstar$ at geometric rate $\discount^m$
per outer iteration. Before stabilization, the improving policies
ensure that $v^{(k)}$ approaches $\Vstar$.

Combining, $v^{(k)} \to \Vstar$ as $k \to \infty$.
\end{proof}

\begin{rlconnection}[title={Modified PI in Practice}]
Modified policy iteration with moderate $m$ (e.g., $m = 10$--$20$)
often gives the best trade-off: each outer iteration costs
$O(m S^2 A)$ instead of $O(S^3 + S^2 A)$, while still converging
much faster than pure value iteration ($m = 1$). Many
implementations of ``value iteration'' in RL software actually
perform modified PI with small $m$, gaining some of the benefits
of policy improvement without the overhead of exact evaluation.
\end{rlconnection}

%% ============================================================
%% SECTION 8: COMPARISON OF VI AND PI
%% ============================================================

\section{Comparison of Value Iteration and Policy Iteration}
\label{sec:comparison}

We now systematically compare the two fundamental algorithms for
solving discounted MDPs.

\begin{definition}[Iteration Complexity]
\label{def:iteration-complexity}
The \emph{iteration complexity} of an algorithm is the number of
iterations required to produce an $\varepsilon$-optimal policy.
The \emph{per-iteration cost} is the number of arithmetic operations
per iteration.
\end{definition}

\begin{proposition}[VI Iteration Complexity]
\label{prop:vi-complexity}
Value iteration with $v^{(0)} = 0$ produces a policy
$d_\varepsilon$ with $\|\Vstar - V^{d_\varepsilon}\|_\infty \leq
\varepsilon$ after at most
\[
    N_{\mathrm{VI}} = \left\lceil
    \frac{1}{\log(1/\discount)}
    \log\!\left(\frac{2\discount R_{\max}}
    {\varepsilon(1-\discount)^2}\right)
    \right\rceil
\]
iterations, each costing $O(S^2 A)$.
\end{proposition}

\begin{proof}
By the a priori bound (Lesson~42), after $n$ iterations
$\|v^{(n)} - \Vstar\|_\infty \leq
\discount^n R_{\max}/(1-\discount)$. By the approximate policy
extraction bound (Lesson~42, Theorem~5.1), the greedy policy
$d^{(n)}$ satisfies
$\|\Vstar - V^{d^{(n)}}\|_\infty \leq
2\discount \|v^{(n)} - \Vstar\|_\infty / (1-\discount)$.
Setting this $\leq \varepsilon$:
\[
    \frac{2\discount}{1-\discount} \cdot
    \frac{\discount^n R_{\max}}{1-\discount} \leq \varepsilon
    \implies
    n \geq \frac{1}{\log(1/\discount)}
    \log\!\left(\frac{2\discount R_{\max}}
    {\varepsilon(1-\discount)^2}\right). \qedhere
\]
\end{proof}

\begin{proposition}[PI Iteration Complexity]
\label{prop:pi-complexity}
Policy iteration produces the exact optimal policy in at most
$A^S$ iterations, each costing $O(S^3 + S^2 A)$.
\end{proposition}

\begin{proof}
This follows from Theorem~\ref{thm:pi-finite}. The per-iteration
cost is $O(S^3)$ for the linear solve (policy evaluation) plus
$O(S^2 A)$ for the greedy step (policy improvement).
\end{proof}

\begin{keyresult}[title={Value Iteration vs.\ Policy Iteration}]
\label{kr:vi-vs-pi}
\begin{center}
\begin{tabular}{lcc}
\toprule
& \textbf{Value Iteration} & \textbf{Policy Iteration} \\
\midrule
Per-iteration cost & $O(S^2 A)$ & $O(S^3 + S^2 A)$ \\
Iteration bound & $O\!\left(\frac{1}{1-\discount}
\log \frac{R_{\max}}{\varepsilon(1-\discount)}\right)$ &
$\leq A^S$ (finite) \\
Convergence type & Geometric, rate $\discount$ &
Finite, superlinear in practice \\
Sensitivity to $\discount$ & Slow for $\discount \approx 1$ &
Independent of $\discount$ \\
Output & $\varepsilon$-optimal policy & Exact optimal policy \\
\bottomrule
\end{tabular}
\end{center}
\textbf{Guidance:} For small $S$ and moderate $A$, prefer PI. For
large $S$, prefer VI or modified PI. For $\discount$ close to 1,
PI is strongly preferred over VI.
\end{keyresult}

\begin{example}[Computational Comparison]
\label{ex:comparison}
Consider an MDP with $S = 100$, $A = 4$, $\discount = 0.99$,
$R_{\max} = 1$, and target $\varepsilon = 0.01$.

\textbf{Value iteration:}
$N_{\mathrm{VI}} \approx
\frac{1}{\log(1/0.99)}\log\frac{2 \cdot 0.99}
{0.01 \cdot 0.01^2}
= \frac{\log(1.98 \times 10^6)}{0.01005}
\approx \frac{14.50}{0.01005} \approx 1443$ iterations.
Per-iteration cost: $O(100^2 \cdot 4) = O(40\,000)$.
Total: $\approx 5.8 \times 10^7$ operations.

\textbf{Policy iteration:}
Worst case: $4^{100}$ iterations (astronomical), but in practice
typically 5--15 iterations.
Per-iteration cost: $O(100^3 + 100^2 \cdot 4) = O(10^6)$.
With 10 iterations: $\approx 10^7$ operations.

\textbf{Modified PI ($m = 20$):}
Per-iteration cost: $O(20 \cdot 100^2 \cdot 4) = O(800\,000)$.
Convergence: faster than VI (convergence rate $\discount^{20}
= 0.99^{20} \approx 0.818$ per outer iteration).
Estimated 50--100 iterations: $\approx 4$--$8 \times 10^7$ operations.

Policy iteration is clearly preferred here.
\end{example}

\begin{rlconnection}[title={PI as the Blueprint for Actor-Critic}]
The structure of policy iteration -- evaluate, then improve --
provides the template for actor-critic algorithms in deep RL:
\begin{itemize}[nosep]
    \item The \textbf{critic} performs (approximate) policy
    evaluation, estimating $V^\policy$ or $Q^\policy$ from sampled
    data.
    \item The \textbf{actor} performs (approximate) policy
    improvement, updating the policy parameters in a direction
    that increases the estimated value.
\end{itemize}
A2C, PPO, and SAC all follow this GPI pattern, trading off exact
evaluation for sample-based approximation.
\end{rlconnection}

%% ============================================================
%% SECTION 9: EXERCISES
%% ============================================================

\section{Exercises}
\label{sec:exercises}

\begin{exercise}[Policy Evaluation by Hand]
\textnormal{(\(*\))} \\
Consider the two-state MDP with $\discount = 0.8$:
\begin{center}
\begin{tabular}{ccccc}
\toprule
$s$ & $a$ & $R(s,a)$ & $P(1 \mid s,a)$ & $P(2 \mid s,a)$ \\
\midrule
1 & $L$ & 3 & 0.6 & 0.4 \\
1 & $R$ & 1 & 0.2 & 0.8 \\
2 & $L$ & 0 & 0.5 & 0.5 \\
2 & $R$ & 2 & 0.1 & 0.9 \\
\bottomrule
\end{tabular}
\end{center}
\begin{enumerate}[label=(\alph*)]
    \item For the policy $d(1) = L$, $d(2) = R$, write down the
    matrices $P^d$ and $r^d$, and solve the linear system
    $(I - 0.8 P^d) v = r^d$ to find $V^d$.
    \item Compute $Q^d(s,a)$ for the alternative actions and
    determine the improved policy $d'$.
\end{enumerate}
\end{exercise}

\begin{exercise}[Complete PI Trace]
\textnormal{(\(*\))} \\
Run policy iteration to completion on the MDP from Exercise~1
starting from $d^{(0)}(1) = R$, $d^{(0)}(2) = L$. Record the
policy and value function at each iteration. How many iterations
does PI require?
\end{exercise}

\begin{exercise}[Cascading Improvement]
\textnormal{(\(**\))} \\
For the three-state MDP in Example~\ref{ex:eval-three-state}, the
policy improvement step changed only the action at state 3.
\begin{enumerate}[label=(\alph*)]
    \item Compute the improvement vector
    $\delta = \Bellman^{d'} V^d - V^d$.
    \item Compute $(I - \discount P^{d'})^{-1} \delta$ and verify
    that it equals $V^{d'} - V^d$.
    \item Identify which states experienced strict improvement and
    explain why in terms of reachability under $d'$.
\end{enumerate}
\end{exercise}

\begin{exercise}[Modified PI Convergence Rate]
\textnormal{(\(**\))} \\
Consider running modified PI with $m$ evaluation steps on an MDP
with $\discount = 0.95$.
\begin{enumerate}[label=(\alph*)]
    \item Compute the effective convergence rate $\discount^m$ per
    outer iteration for $m = 1, 5, 10, 20, 50$.
    \item For each $m$, estimate the number of outer iterations
    needed to reduce the error by a factor of $10^{-4}$ (assuming
    the rate $\discount^m$ governs convergence).
    \item Plot total work (proportional to $m \times
    \text{outer iterations}$) vs.\ $m$ and identify the value of
    $m$ that minimizes total work.
\end{enumerate}
\end{exercise}

\begin{exercise}[PI on an Inventory MDP]
\textnormal{(\(***\))} \\
A warehouse with capacity $C = 3$ manages inventory of a single
product. States are stock levels $\States = \{0, 1, 2, 3\}$. At
each period, the manager chooses how many units to order:
$\Actions(s) = \{0, 1, \ldots, C - s\}$. Demand $D$ follows
$P(D=0) = 0.3$, $P(D=1) = 0.5$, $P(D=2) = 0.2$. Revenue is
\$4 per unit sold; ordering cost is \$2 per unit ordered; holding
cost is \$1 per unit in stock at the end of the period.
Discount factor $\discount = 0.9$.
\begin{enumerate}[label=(\alph*)]
    \item Formulate this as an MDP: specify $P(s' \mid s,a)$ and
    $R(s,a)$ for all $(s,a)$.
    \item Starting from the policy ``always order to full capacity''
    ($d^{(0)}(s) = C - s$), run one complete iteration of policy
    iteration (evaluate and improve).
    \item Run PI to convergence and report the optimal policy.
\end{enumerate}
\end{exercise}

\begin{exercise}[Non-uniqueness of Optimal Policies]
\textnormal{(\(***\))} \\
Construct a three-state, two-action MDP with $\discount = 0.5$ that
has exactly two distinct optimal deterministic stationary policies.
Verify your construction by computing $\Vstar$ and showing that both
policies are greedy w.r.t.\ $\Vstar$.

\textit{Hint:} Arrange the rewards and transitions so that at
exactly one state, two actions yield identical $Q^*$ values.
\end{exercise}

%% ============================================================
%% SECTION 10: SUMMARY
%% ============================================================

\section{Summary}
\label{sec:summary}

\begin{itemize}
    \item \emph{Policy evaluation} computes $V^d$ by solving the
    linear system $(I - \discount P^d) v = r^d$, which has a unique
    solution since $\discount < 1$ ensures invertibility.
    \item The \emph{policy improvement theorem} guarantees that the
    greedy policy $d'$ w.r.t.\ $V^d$ satisfies $V^{d'} \geq V^d$
    everywhere, with equality if and only if $d$ is already optimal.
    \item The algebraic form
    $V^{d'} - V^d = (I - \discount P^{d'})^{-1} \delta$ reveals
    that improvement at changed states cascades to all reachable
    states via discounted visitation probabilities.
    \item \emph{Policy iteration} alternates exact evaluation and
    greedy improvement, converging to the optimal policy in at most
    $A^S$ iterations. No policy repeats, and the value sequence is
    strictly increasing at each non-terminal step.
    \item \emph{Howard's improvement principle}: at each non-terminal
    iteration, the value function strictly improves at every state
    reachable from a state where the action changed.
    \item \emph{Modified policy iteration} replaces exact evaluation
    with $m$ Bellman backups, interpolating between VI ($m = 1$)
    and exact PI ($m \to \infty$). It converges for any $m \geq 1$.
    \item Per iteration, VI costs $O(S^2 A)$ and PI costs
    $O(S^3 + S^2 A)$. VI convergence depends on $\discount$; PI
    convergence does not. Modified PI offers a tunable trade-off.
    \item The evaluate-improve structure of PI is the template for
    actor-critic methods and the generalized policy iteration
    principle in RL.
\end{itemize}

%% ============================================================
%% SECTION 11: FURTHER READING
%% ============================================================

\section{Further Reading}
\label{sec:further-reading}

\begin{itemize}
    \item \textbf{Lesson~46} develops the linear programming
    formulation of MDPs, showing that finding $\Vstar$ can be cast
    as a linear program and introducing occupancy measures, which
    provide the dual perspective on policy optimization.
    \item \textbf{Lesson~24 (TD Learning)} extends the ideas of
    policy evaluation and improvement to the model-free setting,
    where $P$ and $R$ are unknown and the agent learns from
    sampled transitions.
    \item \citet[Chapters 6--7]{puterman2014} provides the
    definitive treatment of policy iteration, modified PI,
    and simplex-type PI variants for both discounted and
    average reward criteria.
    \item \citet{howard1960} is the original monograph introducing
    policy iteration; it remains remarkably clear and insightful.
    \item \citet[Chapter 2]{bertsekas2012} extends PI to
    approximate dynamic programming settings, where function
    approximation replaces exact value computation.
    \item \citet[Chapter 4]{suttonbarto2018} presents PI and GPI
    from the RL practitioner's perspective, with pseudocode and
    connections to temporal-difference methods.
    \item The complexity of policy iteration remains an active
    research area. Ye (2011) proved that PI with the simplex method
    runs in strongly polynomial time for deterministic MDPs, but
    the general case remains open.
\end{itemize}

\bibliography{mdp-refs}
\end{document}
