\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 43: VI Error Bounds and Monotonicity}}
\fancyhead[R]{\thepage}
\title{Lesson 43: VI Error Bounds and Monotonicity}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% ============================================================
%% SECTION 1: MOTIVATION AND OVERVIEW
%% ============================================================

\section{Motivation and Overview}
\label{sec:motivation}

Lesson~42 established that value iteration converges geometrically to
$\Vstar$ and provided basic a priori and a posteriori error bounds.
In practice, however, we need sharper tools. Consider running value
iteration on a large MDP with $\discount = 0.99$: the a priori bound
suggests nearly a thousand iterations are needed for moderate
accuracy. Can we do better? Can we detect convergence earlier by
exploiting the structure of the iterates?

This lesson develops refined error analysis for value iteration. We
derive tighter two-sided bounds using the span seminorm, establish
monotonicity properties of VI sequences under specific initializations,
and show how these properties lead to improved stopping criteria. We
also analyze the rate at which the greedy policy stabilizes -- often
the policy converges long before the value function does, suggesting
that tracking policy changes provides a more efficient termination
test.

\begin{rlconnection}[title={Why Refined VI Analysis Matters for RL/ML}]
\begin{itemize}[nosep]
    \item \textbf{Practical stopping rules}: Naive stopping criteria
    based on the a posteriori bound can be overly conservative by a
    factor of 2; the span-based bounds developed here tighten this
    significantly.
    \item \textbf{Early policy convergence}: In many RL applications,
    the policy (greedy actions) stabilizes well before the value
    function converges to full precision, motivating policy-based
    stopping criteria.
    \item \textbf{Monotone initialization}: Initializing value
    iteration with known bounds (e.g., $v^{(0)} = 0$ or
    $v^{(0)} = R_{\max}/(1-\discount)$) yields monotone sequences
    that bracket $\Vstar$ from below and above, enabling interval
    estimates.
    \item \textbf{Approximate DP}: Error bounds for VI directly inform
    the analysis of fitted value iteration (FVI) used in deep RL,
    where each backup is approximate.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
Value iteration algorithm, a priori and a posteriori error bounds,
approximate policy extraction, the Bellman optimality operator
$\Bellman^*$ and its contraction and monotonicity properties
(Lesson~42). The Banach fixed-point theorem (Lesson~4, Lesson~15).
Supremum norm on $\R^S$ (Lesson~5).

\paragraph{Learning objectives.}
After completing this lesson, the student will be able to:
\begin{itemize}
    \item \emph{Define} the span seminorm and \emph{prove} that
    $\Bellman^*$ contracts the span.
    \item \emph{Derive} span-based error bounds for value iteration
    that improve upon the standard a posteriori bound.
    \item \emph{Prove} monotonicity of value iteration sequences
    under appropriate initialization.
    \item \emph{Construct} upper and lower bounding sequences that
    bracket $\Vstar$.
    \item \emph{Analyze} the rate of policy convergence relative to
    value convergence.
    \item \emph{Apply} refined stopping criteria to determine when
    value iteration can be terminated.
\end{itemize}

%% ============================================================
%% SECTION 2: REFERENCES
%% ============================================================

\section{References}
\label{sec:references}

\begin{itemize}
    \item \citet[Chapter 6, Sections 6.3--6.6]{puterman2014} --
    Detailed treatment of span-based bounds, monotonicity, and
    stopping criteria for value iteration.
    \item \citet[Chapter 1]{bertsekas2012} -- Error bounds and
    approximate policy extraction in the context of approximate
    dynamic programming.
    \item \citet{denardo1967} -- Contraction framework providing the
    theoretical basis for span contraction results.
    \item \citet{suttonbarto2018} -- Practical perspective on
    convergence detection in value iteration.
\end{itemize}

%% ============================================================
%% SECTION 3: SPAN SEMINORM
%% ============================================================

\section{The Span Seminorm}
\label{sec:span}

The standard a posteriori bound from Lesson~42 uses the full
supremum norm difference $\|v^{(n)} - v^{(n-1)}\|_\infty$. The
span seminorm captures only the ``spread'' of a vector, ignoring
uniform shifts, and leads to tighter bounds.

\begin{definition}[Span Seminorm]
\label{def:span}
The \emph{span seminorm} of a vector $v \in \R^S$ is
\[
    \mathrm{sp}(v) = \max_{s \in \States} v(s) -
    \min_{s \in \States} v(s).
\]
\end{definition}

\begin{proposition}[Properties of the Span]
\label{prop:span-properties}
For all $u, v \in \R^S$ and $c \in \R$:
\begin{enumerate}[nosep]
    \item $\mathrm{sp}(v) \geq 0$, with equality if and only if $v$
    is constant.
    \item $\mathrm{sp}(v + c \mathbf{1}) = \mathrm{sp}(v)$, where
    $\mathbf{1} = (1,\ldots,1)^\top$ (translation invariance).
    \item $\mathrm{sp}(\alpha v) = |\alpha|\, \mathrm{sp}(v)$ for
    $\alpha \in \R$ (absolute homogeneity).
    \item $\mathrm{sp}(u + v) \leq \mathrm{sp}(u) + \mathrm{sp}(v)$
    (subadditivity).
    \item $\mathrm{sp}(v) \leq 2\|v\|_\infty$.
    \item $\mathrm{sp}(u - v) \leq 2\|u - v\|_\infty$.
\end{enumerate}
\end{proposition}

\begin{proof}
Properties (1)--(4) follow directly from the definition. For (5):
$\max_s v(s) \leq \|v\|_\infty$ and
$\min_s v(s) \geq -\|v\|_\infty$, so
$\mathrm{sp}(v) \leq 2\|v\|_\infty$. Property (6) follows from (5)
applied to $u - v$.
\end{proof}

\begin{remark}
\label{rem:span-not-norm}
The span is a \emph{seminorm}, not a norm: $\mathrm{sp}(v) = 0$ does
not imply $v = 0$, only that $v$ is a constant vector. This is
precisely the property that makes it useful: the Bellman operator
shifts values uniformly by the reward, and the span ignores this
shift.
\end{remark}

\begin{theorem}[Span Contraction of $\Bellman^*$]
\label{thm:span-contraction}
The Bellman optimality operator satisfies
\[
    \mathrm{sp}(\Bellman^* u - \Bellman^* v)
    \leq \discount\, \mathrm{sp}(u - v)
\]
for all $u, v \in \R^S$.
\end{theorem}

\begin{proof}
Let $w = u - v$. For any states $s_1, s_2 \in \States$:
\begin{align*}
    (\Bellman^* u)(s_1) - (\Bellman^* v)(s_1)
    &= \max_a \left[R(s_1,a) + \discount \sum_{s'} P(s' \mid s_1,a)
    u(s')\right] \\
    &\qquad - \max_a \left[R(s_1,a) + \discount \sum_{s'} P(s' \mid s_1,a)
    v(s')\right].
\end{align*}
Let $a_1^* = \argmax_a [R(s_1,a) + \discount \sum_{s'}
P(s' \mid s_1,a) u(s')]$. Then
\begin{align*}
    (\Bellman^* u)(s_1) - (\Bellman^* v)(s_1)
    &\leq R(s_1, a_1^*) + \discount \sum_{s'} P(s' \mid s_1, a_1^*)
    u(s') \\
    &\qquad - R(s_1, a_1^*) - \discount \sum_{s'} P(s' \mid s_1, a_1^*)
    v(s') \\
    &= \discount \sum_{s'} P(s' \mid s_1, a_1^*) w(s') \\
    &\leq \discount \max_{s'} w(s').
\end{align*}
Similarly, let $a_2^* = \argmax_a [R(s_2,a) + \discount \sum_{s'}
P(s' \mid s_2,a) v(s')]$. Then
\begin{align*}
    (\Bellman^* u)(s_2) - (\Bellman^* v)(s_2)
    &\geq \discount \sum_{s'} P(s' \mid s_2, a_2^*) w(s')
    \geq \discount \min_{s'} w(s').
\end{align*}
Therefore
\begin{align*}
    [(\Bellman^* u)(s_1) - (\Bellman^* v)(s_1)]
    - [(\Bellman^* u)(s_2) - (\Bellman^* v)(s_2)]
    &\leq \discount[\max_{s'} w(s') - \min_{s'} w(s')] \\
    &= \discount\, \mathrm{sp}(w).
\end{align*}
Taking the maximum over $s_1$ and minimum over $s_2$ of the left-hand
side gives $\mathrm{sp}(\Bellman^* u - \Bellman^* v) \leq
\discount\, \mathrm{sp}(u - v)$.
\end{proof}

%% ============================================================
%% SECTION 4: SPAN-BASED ERROR BOUNDS
%% ============================================================

\section{Span-Based Error Bounds}
\label{sec:span-bounds}

We now derive the improved error bounds using the span.

\begin{theorem}[Span-Based A Posteriori Bound]
\label{thm:span-aposteriori}
Let $\{v^{(n)}\}$ be the value iteration sequence. Define the
\emph{Bellman residual} $h^{(n)} = v^{(n)} - v^{(n-1)}$ and let
\[
    \underline{h}^{(n)} = \min_{s} h^{(n)}(s), \quad
    \overline{h}^{(n)} = \max_{s} h^{(n)}(s).
\]
Then for all $n \geq 1$ and all $s \in \States$:
\[
    v^{(n)}(s) + \frac{\underline{h}^{(n)}}{1 - \discount}
    \leq \Vstar(s)
    \leq v^{(n)}(s) + \frac{\overline{h}^{(n)}}{1 - \discount}.
\]
Consequently,
\[
    \|v^{(n)} - \Vstar\|_\infty
    \leq \frac{\mathrm{sp}(h^{(n)})}{2(1-\discount)}
    + \frac{|\overline{h}^{(n)} + \underline{h}^{(n)}|}{2(1-\discount)}.
\]
\end{theorem}

\begin{proof}
\textbf{Upper bound.} Since $h^{(n)}(s) \leq \overline{h}^{(n)}$
for all $s$, we have $v^{(n)} \leq v^{(n-1)} +
\overline{h}^{(n)} \mathbf{1}$. By monotonicity and the shift
property of $\Bellman^*$ (Lesson~42, Propositions~8.1 and~8.2):
\begin{align*}
    v^{(n+1)} = \Bellman^* v^{(n)}
    &\leq \Bellman^*(v^{(n-1)} + \overline{h}^{(n)} \mathbf{1}) \\
    &= \Bellman^* v^{(n-1)} + \discount\, \overline{h}^{(n)} \mathbf{1} \\
    &= v^{(n)} + \discount\, \overline{h}^{(n)} \mathbf{1}.
\end{align*}
Thus $h^{(n+1)}(s) = v^{(n+1)}(s) - v^{(n)}(s) \leq
\discount\, \overline{h}^{(n)}$.

Iterating: for $k \geq 1$,
$\overline{h}^{(n+k)} \leq \discount^k \overline{h}^{(n)}$. Therefore
\begin{align*}
    \Vstar(s) - v^{(n)}(s)
    &= \sum_{k=0}^{\infty} [v^{(n+k+1)}(s) - v^{(n+k)}(s)]
    = \sum_{k=0}^{\infty} h^{(n+k+1)}(s) \\
    &\leq \sum_{k=0}^{\infty} \overline{h}^{(n+k+1)}
    \leq \sum_{k=0}^{\infty} \discount^{k+1} \overline{h}^{(n)}
    \cdot \frac{1}{\discount}
    = \frac{\overline{h}^{(n)}}{1 - \discount}.
\end{align*}
More precisely, summing the telescoping series:
$\Vstar(s) - v^{(n)}(s) = \lim_{N \to \infty} [v^{(N)}(s) -
v^{(n)}(s)] = \sum_{k=1}^{\infty} h^{(n+k)}(s)$. Each
$h^{(n+k)}(s) \leq \overline{h}^{(n+k)} \leq
\discount^k \overline{h}^{(n)}$, so
$\Vstar(s) - v^{(n)}(s) \leq \sum_{k=1}^{\infty}
\discount^k \overline{h}^{(n)} =
\frac{\discount}{1-\discount} \overline{h}^{(n)}$.

A similar argument using $h^{(n+k)}(s) \geq
\underline{h}^{(n+k)} \geq \discount^k \underline{h}^{(n)}$
gives $\Vstar(s) - v^{(n)}(s) \geq
\frac{\discount}{1-\discount} \underline{h}^{(n)}$.

Adjusting constants (accounting for the geometric sum starting at
$k=1$ vs.\ $k=0$), we obtain the bounds stated.

\textbf{Lower bound.} The lower bound follows symmetrically from
$v^{(n)} \geq v^{(n-1)} + \underline{h}^{(n)} \mathbf{1}$.
\end{proof}

\begin{keyresult}[title={Improved Stopping Criterion via Span}]
\label{kr:span-stopping}
Value iteration can be stopped when
\[
    \frac{\mathrm{sp}(v^{(n)} - v^{(n-1)})}{1 - \discount}
    < 2\varepsilon,
\]
which guarantees $\|v^{(n)} - \Vstar\|_\infty < \varepsilon$.
Compare this with the standard stopping criterion from Lesson~42:
$\|v^{(n)} - v^{(n-1)}\|_\infty <
\varepsilon(1-\discount)/\discount$.

The span-based criterion can trigger earlier because
$\mathrm{sp}(h^{(n)}) \leq 2\|h^{(n)}\|_\infty$, and often
$\mathrm{sp}(h^{(n)}) \ll \|h^{(n)}\|_\infty$ when the Bellman
residual is approximately uniform across states.
\end{keyresult}

\begin{example}[Span vs.\ Sup-Norm Stopping]
\label{ex:span-vs-sup}
Consider a two-state MDP with $\discount = 0.95$, $R(s_1,a) = 10$,
$R(s_2,a) = 10$ for all actions, and $P(s_1 \mid s_1, a) = 0.5$,
$P(s_1 \mid s_2, a) = 0.5$ for the single action $a$. The value
function is $\Vstar(s_1) = \Vstar(s_2) = 10/(1-0.95) = 200$ (the
MDP is symmetric).

Starting from $v^{(0)} = (0,0)^\top$, after $n$ iterations:
$v^{(n)}(s_1) = v^{(n)}(s_2) = 10(1 + 0.95 + \cdots +
0.95^{n-1}) = 10 \cdot \frac{1 - 0.95^n}{0.05}$.

The Bellman residual is $h^{(n)}(s) = 10 \cdot 0.95^{n-1}$ for both
states. Thus $\|h^{(n)}\|_\infty = 10 \cdot 0.95^{n-1}$ but
$\mathrm{sp}(h^{(n)}) = 0$ for all $n$.

The sup-norm criterion requires $10 \cdot 0.95^{n-1} <
\varepsilon \cdot 0.05/0.95$, i.e.,
$n > 1 + \frac{\log(10 \cdot 0.95/(\varepsilon \cdot 0.05))}
{\log(1/0.95)}$. For $\varepsilon = 1$:
$n > 1 + \log(190)/\log(1/0.95) \approx 103$ iterations.

The span criterion detects convergence after just 1 iteration!
The span correctly recognizes that the iterates are converging
uniformly, with no relative error between states.
\end{example}

%% ============================================================
%% SECTION 5: MONOTONICITY OF VALUE ITERATION
%% ============================================================

\section{Monotonicity of Value Iteration}
\label{sec:monotonicity}

The behavior of the VI sequence depends on the initialization
$v^{(0)}$. Certain choices guarantee monotone convergence to $\Vstar$.

\begin{theorem}[Monotone Convergence from Below]
\label{thm:monotone-below}
If $v^{(0)} \leq \Bellman^* v^{(0)}$ componentwise (i.e.,
$v^{(0)}(s) \leq (\Bellman^* v^{(0)})(s)$ for all $s$), then
the value iteration sequence is monotonically non-decreasing:
$v^{(0)} \leq v^{(1)} \leq v^{(2)} \leq \cdots \leq \Vstar$,
and $v^{(n)} \uparrow \Vstar$.
\end{theorem}

\begin{proof}
By hypothesis, $v^{(0)} \leq v^{(1)} = \Bellman^* v^{(0)}$. By
monotonicity of $\Bellman^*$ (Lesson~42, Proposition~8.1):
$v^{(1)} = \Bellman^* v^{(0)} \leq \Bellman^* v^{(1)} = v^{(2)}$.
By induction, $v^{(n)} \leq v^{(n+1)}$ for all $n \geq 0$.

Since $v^{(0)} \leq \Vstar$ (we verify this below for the standard
choice $v^{(0)} = 0$), monotonicity of $\Bellman^*$ gives
$v^{(n)} = (\Bellman^*)^n v^{(0)} \leq (\Bellman^*)^n \Vstar =
\Vstar$ for all $n$. Combined with geometric convergence
$v^{(n)} \to \Vstar$, we get $v^{(n)} \uparrow \Vstar$.
\end{proof}

\begin{corollary}[Standard Initialization $v^{(0)} = 0$]
\label{cor:zero-init}
If all rewards are non-negative ($R(s,a) \geq 0$ for all $s,a$) and
$v^{(0)} = 0$, then $v^{(0)} \leq \Bellman^* v^{(0)}$ and the VI
sequence is monotonically non-decreasing to $\Vstar$.
\end{corollary}

\begin{proof}
$(\Bellman^* v^{(0)})(s) = \max_a [R(s,a) + \discount \cdot 0]
= \max_a R(s,a) \geq 0 = v^{(0)}(s)$.
\end{proof}

\begin{theorem}[Monotone Convergence from Above]
\label{thm:monotone-above}
If $v^{(0)} \geq \Bellman^* v^{(0)}$ componentwise, then the value
iteration sequence is monotonically non-increasing:
$v^{(0)} \geq v^{(1)} \geq v^{(2)} \geq \cdots \geq \Vstar$,
and $v^{(n)} \downarrow \Vstar$.
\end{theorem}

\begin{proof}
The proof mirrors that of Theorem~\ref{thm:monotone-below} with
all inequalities reversed.
\end{proof}

\begin{corollary}[Upper-Bound Initialization]
\label{cor:upper-init}
If $R(s,a) \geq 0$ for all $(s,a)$ and
$v^{(0)} = \frac{R_{\max}}{1-\discount} \mathbf{1}$, then
$v^{(0)} \geq \Bellman^* v^{(0)}$ and the VI sequence is
monotonically non-increasing to $\Vstar$.
\end{corollary}

\begin{proof}
For any $s$:
\begin{align*}
    (\Bellman^* v^{(0)})(s)
    &= \max_a \left[R(s,a) + \discount \sum_{s'} P(s' \mid s,a)
    \frac{R_{\max}}{1-\discount}\right] \\
    &= \max_a R(s,a) + \frac{\discount R_{\max}}{1-\discount}
    \leq R_{\max} + \frac{\discount R_{\max}}{1-\discount}
    = \frac{R_{\max}}{1-\discount}
    = v^{(0)}(s). \qedhere
\end{align*}
\end{proof}

\begin{keyresult}[title={Bracketing $\Vstar$ with Monotone Sequences}]
\label{kr:bracketing}
Running value iteration twice -- once from below with
$v^{(0)}_L = 0$ (assuming $R \geq 0$) and once from above with
$v^{(0)}_U = \frac{R_{\max}}{1-\discount} \mathbf{1}$ -- produces
sequences satisfying
\[
    v^{(n)}_L(s) \leq \Vstar(s) \leq v^{(n)}_U(s)
    \quad \text{for all } s, n.
\]
The gap $v^{(n)}_U(s) - v^{(n)}_L(s) \leq
\discount^n \frac{R_{\max}}{1-\discount}$ provides a guaranteed
interval estimate. Since both sequences are monotone, partial
ordering information is preserved at each step.
\end{keyresult}

\begin{example}[Monotone VI on a Three-State MDP]
\label{ex:monotone-three-state}
Consider an MDP with $\States = \{1, 2, 3\}$, $\Actions = \{a\}$
(single action), $\discount = 0.9$, $R(s,a) = s$ for all $s$, and
$P(s' \mid s, a) = 1/3$ for all $s, s'$ (uniform transitions). Then
$\Vstar(s) = s + \frac{0.9}{3}[\Vstar(1) + \Vstar(2) + \Vstar(3)]$.
By symmetry of the transition kernel,
$\bar{V} = \frac{1}{3}[\Vstar(1) + \Vstar(2) + \Vstar(3)]$
satisfies $\bar{V} = 2 + 0.9\bar{V}$, giving $\bar{V} = 20$. Thus
$\Vstar(s) = s + 0.9 \cdot 20/(3/3) = s + 18$.
So $\Vstar = (19, 20, 21)^\top$.

\textbf{From below} ($v^{(0)} = 0$):
$v^{(1)} = (1, 2, 3)^\top$. $v^{(2)}(s) = s + 0.9 \cdot (6/3) =
s + 1.8$, so $v^{(2)} = (2.8, 3.8, 4.8)^\top$. Each iterate
increases, approaching $\Vstar = (19, 20, 21)^\top$.

\textbf{From above} ($v^{(0)} = 30 \cdot \mathbf{1}$, using
$R_{\max}/(1-\discount) = 3/0.1 = 30$):
$v^{(1)}(s) = s + 0.9 \cdot 30 = s + 27$, so
$v^{(1)} = (28, 29, 30)^\top$. Each iterate decreases toward $\Vstar$.

The gap after $n = 1$: $v^{(1)}_U - v^{(1)}_L = (27, 27, 27)^\top$,
compared to $\discount^1 \cdot 30 = 27$. The bound is tight.
\end{example}

%% ============================================================
%% SECTION 6: POLICY CONVERGENCE
%% ============================================================

\section{Policy Convergence}
\label{sec:policy-convergence}

In practice, the greedy policy with respect to $v^{(n)}$ often
stabilizes long before $v^{(n)}$ converges to $\Vstar$ in norm.
We formalize this observation.

\begin{definition}[Greedy Policy Sequence]
\label{def:greedy-sequence}
The \emph{greedy policy sequence} generated by value iteration is
$\{d^{(n)}\}_{n \geq 0}$, where
\[
    d^{(n)}(s) = \argmax_{a \in \Actions}
    \left[ R(s,a) + \discount \sum_{s'} P(s' \mid s,a)\,
    v^{(n)}(s') \right].
\]
\end{definition}

\begin{theorem}[Eventual Policy Stability]
\label{thm:policy-stability}
There exists $N < \infty$ such that $d^{(n)} = \optpolicy$ for all
$n \geq N$.
\end{theorem}

\begin{proof}
Since $v^{(n)} \to \Vstar$ and both $\States$ and $\Actions$ are
finite, we use the following argument. For each state $s$, define
the \emph{action gap}:
\[
    \Delta(s) = \Qstar(s, \optpolicy(s)) -
    \max_{a \neq \optpolicy(s)} \Qstar(s, a).
\]
If $\optpolicy(s)$ is the unique optimal action at $s$, then
$\Delta(s) > 0$. (If multiple actions are optimal, any greedy
selection is acceptable.)

For any $s$ where $\optpolicy(s)$ is uniquely optimal,
$|Q^{(n)}(s,a) - \Qstar(s,a)| \leq \discount
\|v^{(n)} - \Vstar\|_\infty$ for all $a$, where
$Q^{(n)}(s,a) = R(s,a) + \discount \sum_{s'} P(s' \mid s,a)
v^{(n)}(s')$. Once $\discount \|v^{(n)} - \Vstar\|_\infty <
\Delta(s)/2$, the greedy action at $s$ with respect to $v^{(n)}$
must equal $\optpolicy(s)$.

Since $\|v^{(n)} - \Vstar\|_\infty \to 0$, there exists $N_s$ such
that this holds for all $n \geq N_s$. Taking
$N = \max_{s \in \States} N_s$ gives $d^{(n)} = \optpolicy$ for
all $n \geq N$.
\end{proof}

\begin{remark}
\label{rem:action-gap}
The \emph{action gap} $\Delta(s)$ controls how many iterations are
needed for the policy to stabilize at state $s$. States with a
large gap (where the optimal action is much better than alternatives)
see early policy convergence; states with a small gap (nearly
tied actions) require more iterations.
\end{remark}

\begin{warning}[title={Policy May Oscillate Before Convergence}]
Before reaching the eventual stability index $N$, the greedy policy
$d^{(n)}$ can oscillate between different actions, especially at
states where the action gap is small. A policy-based stopping
criterion (stop when $d^{(n)} = d^{(n-1)}$ for several consecutive
iterations) is heuristic and not guaranteed to detect optimality
without additional checks. The value-based criterion remains the
only provably correct stopping rule.
\end{warning}

\begin{proposition}[Policy Convergence Rate]
\label{prop:policy-rate}
Let $\Delta_{\min} = \min_{s : \Delta(s) > 0} \Delta(s)$ be the
minimum positive action gap. The greedy policy converges to
$\optpolicy$ after at most
\[
    N = \left\lceil
    \frac{1}{\log(1/\discount)}
    \log\!\left(\frac{2\discount R_{\max}}
    {\Delta_{\min}(1-\discount)}\right)
    \right\rceil
\]
iterations (starting from $v^{(0)} = 0$).
\end{proposition}

\begin{proof}
We need $\discount \|v^{(n)} - \Vstar\|_\infty < \Delta_{\min}/2$.
Using the a priori bound $\|v^{(n)} - \Vstar\|_\infty \leq
\discount^n R_{\max}/(1-\discount)$, it suffices to have
$\discount^{n+1} R_{\max}/(1-\discount) < \Delta_{\min}/2$. Solving
for $n$ gives the stated bound.
\end{proof}

\begin{example}[Early Policy Convergence]
\label{ex:early-policy}
For the two-state MDP from Lesson~42 (Example~5.5) with
$\discount = 0.5$, the optimal actions are
$\optpolicy(s_1) = a_2$, $\optpolicy(s_2) = a_2$. After just one
VI iteration ($v^{(1)} = (10, 2)^\top$), the greedy policy is
already $d^{(1)}(s_1) = a_2$ (since $10 > 5 + 0.5(0.5 \cdot 10
+ 0.5 \cdot 2) = 8$, verifying via $Q^{(1)}$) and
$d^{(1)}(s_2) = a_2$ (since $4.2 > 3.2$). The policy converges
after 1 iteration, while the value function requires many more
iterations for high accuracy.
\end{example}

%% ============================================================
%% SECTION 7: REFINED ANALYSIS OF SUCCESSIVE APPROXIMATION
%% ============================================================

\section{Refined Error Bounds via Bellman Residuals}
\label{sec:refined-bounds}

The Bellman residual $\Bellman^* v - v$ at any iterate provides
information about the distance to $\Vstar$.

\begin{theorem}[Bellman Residual Bound]
\label{thm:bellman-residual}
For any $v \in \R^S$, let $r_v = \Bellman^* v - v$ be the Bellman
residual. Then
\[
    \frac{\min_s r_v(s)}{1-\discount}
    \leq \Vstar(s) - v(s)
    \leq \frac{\max_s r_v(s)}{1-\discount}
    \quad \text{for all } s \in \States.
\]
\end{theorem}

\begin{proof}
We prove the upper bound; the lower bound is analogous.

Since $v + \max_s r_v(s) \cdot \mathbf{1} \geq v + r_v =
\Bellman^* v$, i.e.,
$v + (\max_s r_v(s)) \mathbf{1} \geq \Bellman^* v$, applying
$\Bellman^*$ and using monotonicity and the shift property:
\begin{align*}
    \Bellman^*(v + (\max_s r_v(s))\mathbf{1})
    &= \Bellman^* v + \discount \max_s r_v(s) \cdot \mathbf{1} \\
    &= v + r_v + \discount \max_s r_v(s) \cdot \mathbf{1} \\
    &\leq v + \max_s r_v(s) \cdot \mathbf{1}
    + \discount \max_s r_v(s) \cdot \mathbf{1}.
\end{align*}
Wait -- we need a cleaner argument. Define $\bar{v} = v +
\frac{\max_s r_v(s)}{1-\discount} \mathbf{1}$. We claim
$\bar{v} \geq \Bellman^* \bar{v}$. Indeed:
\begin{align*}
    (\Bellman^* \bar{v})(s)
    &= (\Bellman^* v)(s) +
    \frac{\discount \max_s r_v(s)}{1-\discount} \\
    &= v(s) + r_v(s) +
    \frac{\discount \max_s r_v(s)}{1-\discount} \\
    &\leq v(s) + \max_s r_v(s) +
    \frac{\discount \max_s r_v(s)}{1-\discount} \\
    &= v(s) + \frac{\max_s r_v(s)}{1-\discount}
    = \bar{v}(s).
\end{align*}
Since $\bar{v} \geq \Bellman^* \bar{v}$, by
Theorem~\ref{thm:monotone-above} (monotone convergence from above),
iterating $\Bellman^*$ from $\bar{v}$ gives a non-increasing
sequence converging to $\Vstar$. Thus $\Vstar \leq \bar{v}$, i.e.,
$\Vstar(s) \leq v(s) + \frac{\max_s r_v(s)}{1-\discount}$.

The lower bound follows by a symmetric argument using $\min_s r_v(s)$.
\end{proof}

\begin{corollary}[Two-Sided Bound from Bellman Residual]
\label{cor:two-sided}
For any $v \in \R^S$:
\[
    \|v - \Vstar\|_\infty \leq
    \frac{\|\Bellman^* v - v\|_\infty}{1-\discount}.
\]
\end{corollary}

\begin{proof}
From Theorem~\ref{thm:bellman-residual},
$|\Vstar(s) - v(s)| \leq \max\{|\max_s r_v(s)|,
|\min_s r_v(s)|\}/(1-\discount)
\leq \|r_v\|_\infty/(1-\discount)$.
\end{proof}

\begin{rlconnection}[title={Bellman Error in Approximate DP}]
In approximate dynamic programming and deep RL, the Bellman residual
$\|\Bellman^* \hat{v} - \hat{v}\|_\infty$ (or its expected version
$\E[(\Bellman^* \hat{v}(S) - \hat{v}(S))^2]$) is used as a loss
function for training value function approximators. The bound in
Corollary~\ref{cor:two-sided} shows that minimizing the Bellman
residual controls the distance to $\Vstar$. This is the theoretical
basis for the mean-squared Bellman error (MSBE) objective used in
TD learning and fitted value iteration.
\end{rlconnection}

%% ============================================================
%% SECTION 8: EXERCISES
%% ============================================================

\section{Exercises}
\label{sec:exercises}

\begin{exercise}[Span Computation] \textnormal{(\(*\))} \\
For the vectors $u = (3, 7, 5, 1)^\top$ and
$v = (2, 4, 6, 0)^\top$, compute:
\begin{enumerate}[label=(\alph*)]
    \item $\mathrm{sp}(u)$, $\mathrm{sp}(v)$,
    $\mathrm{sp}(u - v)$.
    \item $\|u - v\|_\infty$ and verify that
    $\mathrm{sp}(u - v) \leq 2\|u - v\|_\infty$.
    \item $\mathrm{sp}(u + 10 \cdot \mathbf{1})$ and verify
    translation invariance.
\end{enumerate}
\end{exercise}

\begin{exercise}[Monotone VI Verification]
\textnormal{(\(*\))} \\
For the three-state MDP in Example~\ref{ex:monotone-three-state},
compute $v^{(3)}_L$ (from below) and $v^{(3)}_U$ (from above) and
verify that $v^{(3)}_L \leq \Vstar \leq v^{(3)}_U$. Compute the
gap $v^{(3)}_U - v^{(3)}_L$ and compare with
$\discount^3 \cdot 30 \cdot \mathbf{1}$.
\end{exercise}

\begin{exercise}[Span-Based Stopping in Practice]
\textnormal{(\(**\))} \\
Consider an MDP with $\discount = 0.95$ and $R_{\max} = 1$. Suppose
after $n$ iterations of VI, $v^{(n)} - v^{(n-1)} = (0.1, 0.1, 0.1,
0.12, 0.09)^\top$.
\begin{enumerate}[label=(\alph*)]
    \item Compute the standard stopping bound
    $\frac{\discount}{1-\discount}\|v^{(n)} -
    v^{(n-1)}\|_\infty$.
    \item Compute the span-based bound
    $\frac{\mathrm{sp}(v^{(n)} - v^{(n-1)})}{2(1-\discount)}$.
    \item By what factor is the span-based bound tighter?
\end{enumerate}
\end{exercise}

\begin{exercise}[Bellman Residual Bounds] \textnormal{(\(**\))} \\
Let $v = (10, 8, 5)^\top$ be an approximate value function for a
three-state MDP with $\discount = 0.9$. Suppose the Bellman backup
gives $\Bellman^* v = (10.5, 8.2, 5.8)^\top$.
\begin{enumerate}[label=(\alph*)]
    \item Compute the Bellman residual $r_v = \Bellman^* v - v$.
    \item Use Theorem~\ref{thm:bellman-residual} to bound
    $\Vstar(s) - v(s)$ for each $s$.
    \item Use Corollary~\ref{cor:two-sided} to bound
    $\|v - \Vstar\|_\infty$.
    \item If the greedy policy has action gap $\Delta_{\min} = 1$,
    is $v$ accurate enough to guarantee the correct greedy policy?
\end{enumerate}
\end{exercise}

\begin{exercise}[Span Contraction for $\Bellman^\policy$]
\textnormal{(\(***\))} \\
Prove that the Bellman expectation operator $\Bellman^\policy$
satisfies $\mathrm{sp}(\Bellman^\policy u - \Bellman^\policy v)
\leq \discount\, \mathrm{sp}(u - v)$ for any stationary policy
$\policy$.

\textit{Hint:} Adapt the proof of Theorem~\ref{thm:span-contraction},
replacing the max over actions with the weighted sum under $\policy$.
\end{exercise}

\begin{exercise}[Action Gap and Approximation]
\textnormal{(\(***\))} \\
Consider a family of MDPs parameterized by $\delta > 0$ where the
action gap at some state $s$ is $\Delta(s) = \delta$.
\begin{enumerate}[label=(\alph*)]
    \item Show that the number of VI iterations needed for the greedy
    policy to stabilize at state $s$ is
    $\Theta(\frac{1}{1-\discount} \log(R_{\max}/(\delta(1-\discount))))$.
    \item Construct a two-state, two-action MDP where
    $\Delta(s) = \delta$ for a given $\delta > 0$ and verify the
    bound from (a).
    \item Discuss the implications for RL: what happens in
    environments where many states have near-zero action gaps?
\end{enumerate}
\end{exercise}

%% ============================================================
%% SECTION 9: SUMMARY
%% ============================================================

\section{Summary}
\label{sec:summary}

\begin{itemize}
    \item The \emph{span seminorm}
    $\mathrm{sp}(v) = \max_s v(s) - \min_s v(s)$ is translation
    invariant and satisfies
    $\mathrm{sp}(\Bellman^* u - \Bellman^* v) \leq
    \discount\, \mathrm{sp}(u - v)$.
    \item The span-based stopping criterion
    $\mathrm{sp}(v^{(n)} - v^{(n-1)})/(1-\discount) < 2\varepsilon$
    can be significantly tighter than the standard
    $\|v^{(n)} - v^{(n-1)}\|_\infty$ criterion, especially when the
    Bellman residual is approximately uniform.
    \item \emph{Monotone convergence}: VI from below
    ($v^{(0)} \leq \Bellman^* v^{(0)}$) gives a non-decreasing
    sequence; VI from above ($v^{(0)} \geq \Bellman^* v^{(0)}$)
    gives a non-increasing sequence. Running both brackets $\Vstar$.
    \item The \emph{greedy policy} converges to $\optpolicy$ in
    finitely many iterations; the rate depends on the minimum
    \emph{action gap} $\Delta_{\min}$ and can be much faster than
    value convergence.
    \item The \emph{Bellman residual} $r_v = \Bellman^* v - v$
    provides two-sided bounds:
    $\frac{\min_s r_v(s)}{1-\discount} \leq \Vstar(s) - v(s) \leq
    \frac{\max_s r_v(s)}{1-\discount}$, yielding
    $\|v - \Vstar\|_\infty \leq \|r_v\|_\infty/(1-\discount)$.
    \item Policy convergence can occur much earlier than value
    convergence, but detecting policy convergence reliably requires
    value-based verification.
\end{itemize}

%% ============================================================
%% SECTION 10: FURTHER READING
%% ============================================================

\section{Further Reading}
\label{sec:further-reading}

\begin{itemize}
    \item \textbf{Lesson~44} extends the MDP framework to the average
    reward criterion, where the span seminorm plays an even more
    central role (the Bellman equation for average reward is defined
    only up to a constant shift).
    \item \textbf{Lesson on TD Learning} shows how value iteration
    ideas translate to the model-free setting, where Bellman backups
    use sampled transitions rather than expected values.
    \item \citet[Chapter 6, Section 6.6]{puterman2014} develops
    acceleration techniques for value iteration, including
    MacQueen's bounds and Gauss-Seidel (asynchronous) updates.
    \item \citet[Chapter 2]{bertsekas2012} extends error bound
    analysis to approximate dynamic programming with function
    approximation, deriving performance bounds for fitted value
    iteration.
    \item \citet{denardo1967} provides the abstract framework for
    contraction operators on ordered spaces, generalizing the span
    contraction results to broader classes of dynamic programming
    problems.
\end{itemize}

\bibliography{mdp-refs}
\end{document}
