\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 39: MDP Model and Policies}}
\fancyhead[R]{\thepage}
\title{Lesson 39: MDP Model and Policies}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% ============================================================
%% SECTION 1: MOTIVATION AND OVERVIEW
%% ============================================================

\section{Motivation and Overview}
\label{sec:motivation}

Consider an autonomous robot navigating a grid-world warehouse. At each
time step the robot observes its current cell, selects a movement action
(north, south, east, west, or stay), and receives a reward that depends
on the cell reached -- perhaps $+10$ for reaching the charging station,
$-1$ for each movement to encourage efficiency, and $-100$ for
colliding with an obstacle. The warehouse layout, the probabilistic
effects of actions (wheels may slip on a wet floor), and the reward
structure together define a sequential decision problem. How should we
formalize this setting so that we can reason mathematically about
optimal behavior? What class of decision rules should we consider, and
how do we measure the long-run performance of a particular strategy?

The Markov decision process (MDP) provides the canonical mathematical
framework for sequential decision-making under uncertainty. The tools
developed in this lesson -- the formal MDP tuple, policy classes, value
functions, and the Bellman expectation equations -- form the foundation
upon which all of reinforcement learning is built.

\begin{rlconnection}[title={Why MDPs Matter for RL/ML}]
\begin{itemize}[nosep]
    \item \textbf{Tabular RL algorithms}: Q-learning, SARSA, and Monte
    Carlo methods all operate within the MDP framework; their
    convergence proofs require the formal definitions introduced here.
    \item \textbf{Policy gradient methods}: REINFORCE and actor-critic
    algorithms optimize over parameterized policy classes, but the
    underlying objective (maximizing the value function) is defined
    through the MDP model.
    \item \textbf{Deep RL}: DQN, PPO, and SAC extend tabular methods
    using function approximation, yet their theoretical guarantees
    ultimately rely on MDP structure and Bellman equations.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
Probability spaces, conditional probability, and conditional
expectation including the tower property (Lessons~9--12). Geometric
series and absolute convergence (Lesson~1). Basic linear algebra:
vectors and matrices (Lesson~5). Contraction mappings on complete
metric spaces (Lesson~4, Lesson~15).

\paragraph{Learning objectives.}
After completing this lesson, the student will be able to:
\begin{itemize}
    \item \emph{Define} a Markov decision process and each of its
    components formally.
    \item \emph{Classify} policies as history-dependent, Markov,
    stationary, deterministic, or stochastic, and \emph{explain} the
    relationships among these classes.
    \item \emph{Derive} the Bellman expectation equations for both the
    state-value function $\Vpi$ and the action-value function $\Qpi$.
    \item \emph{Compute} value functions for small MDPs by direct
    solution of the Bellman equations.
    \item \emph{State} the existence theorem for deterministic
    stationary optimal policies in finite MDPs.
    \item \emph{Prove} that the return is well-defined and bounded
    under the discounted criterion.
\end{itemize}

%% ============================================================
%% SECTION 2: REFERENCES
%% ============================================================

\section{References}
\label{sec:references}

\begin{itemize}
    \item \citet[Chapters 2--3]{puterman2014} -- Definitive treatment
    of MDP formulation, policy classes, and optimality criteria.
    \item \citet[Chapter 1]{bertsekas2012} -- Concise development of
    dynamic programming foundations with emphasis on infinite-horizon
    discounted problems.
    \item \citet[Chapters 3--4]{suttonbarto2018} -- Accessible
    introduction to MDPs, value functions, and Bellman equations from
    the RL perspective.
    \item \citet{howard1960} -- Howard's seminal monograph introducing
    policy iteration and the MDP framework.
    \item \citet{blackwell1962} -- Foundational paper on discounted
    dynamic programming and optimality.
\end{itemize}

%% ============================================================
%% SECTION 3: THE MDP MODEL
%% ============================================================

\section{The Markov Decision Process Model}
\label{sec:mdp-model}

\begin{definition}[Markov Decision Process]
\label{def:mdp}
A \emph{Markov decision process} (MDP) is a tuple
$\mathcal{M} = (\States, \Actions, P, R, \discount)$ where:
\begin{enumerate}[nosep]
    \item $\States$ is a finite set of \emph{states}, with
    $|\States| = S$.
    \item $\Actions$ is a finite set of \emph{actions}, with
    $|\Actions| = A$.
    \item $P : \States \times \Actions \times \States \to [0,1]$ is
    the \emph{transition kernel}, where
    $P(s' \mid s, a) = \Prob(S_{t+1} = s' \mid S_t = s, A_t = a)$
    satisfies $\sum_{s' \in \States} P(s' \mid s, a) = 1$ for every
    $(s,a) \in \States \times \Actions$.
    \item $R : \States \times \Actions \to \R$ is the (deterministic)
    \emph{reward function}, where $R(s,a)$ is the immediate reward
    received upon taking action $a$ in state $s$.
    \item $\discount \in [0,1)$ is the \emph{discount factor}.
\end{enumerate}
\end{definition}

\begin{remark}
\label{rem:reward-variants}
Several equivalent reward formulations appear in the literature:
$R(s,a,s')$, $R(s)$, or random rewards with
$\E[R_t \mid S_t = s, A_t = a] = R(s,a)$. In the finite-state,
finite-action case with bounded rewards, these formulations are
interchangeable up to a redefinition of the reward function. We adopt
$R(s,a)$ throughout for notational simplicity. When using
$R(s,a,s')$, the expected reward is
$\bar{R}(s,a) = \sum_{s'} P(s' \mid s,a) R(s,a,s')$.
\end{remark}

The key structural property of MDPs is the \emph{Markov property}: the
future evolution depends on the past only through the current state and
action.

\begin{definition}[Markov Property]
\label{def:markov-property}
A stochastic process $(S_t)_{t \geq 0}$ controlled by actions
$(A_t)_{t \geq 0}$ satisfies the \emph{Markov property} if for all
$t \geq 0$ and all histories
$h_t = (s_0, a_0, s_1, a_1, \ldots, s_t, a_t)$,
\[
    \Prob(S_{t+1} = s' \mid S_0 = s_0, A_0 = a_0, \ldots, S_t = s_t,
    A_t = a_t) = \Prob(S_{t+1} = s' \mid S_t = s_t, A_t = a_t)
    = P(s' \mid s_t, a_t).
\]
\end{definition}

\begin{example}[Grid World MDP]
\label{ex:gridworld}
Consider a $3 \times 3$ grid world with states
$\States = \{(i,j) : i,j \in \{1,2,3\}\}$ and actions
$\Actions = \{\text{N}, \text{S}, \text{E}, \text{W}\}$. The agent
starts at $(1,1)$ and aims to reach the goal at $(3,3)$. The
transition kernel is deterministic:
$P((i+1,j) \mid (i,j), \text{N}) = 1$ if $i < 3$, and
$P((i,j) \mid (i,j), \text{N}) = 1$ if $i = 3$ (wall bounce),
with analogous rules for the other directions. The reward is
$R(s,a) = +1$ if the transition leads to $(3,3)$, and $R(s,a) = 0$
otherwise. With $\discount = 0.9$, this defines a complete MDP
$\mathcal{M} = (\States, \Actions, P, R, 0.9)$ with $S = 9$ states
and $A = 4$ actions.
\end{example}

%% ============================================================
%% SECTION 4: POLICIES
%% ============================================================

\section{Policies}
\label{sec:policies}

A \emph{policy} specifies how the agent selects actions. We define a
hierarchy of increasingly restrictive policy classes.

\begin{definition}[History]
\label{def:history}
A \emph{history} at time $t$ is the sequence
$h_t = (s_0, a_0, r_0, s_1, a_1, r_1, \ldots, s_{t-1}, a_{t-1},
r_{t-1}, s_t)$ recording all states visited, actions taken, and
rewards received up to time $t$. Let $\mathcal{H}_t$ denote the set
of all possible histories at time $t$, and
$\mathcal{H} = \bigcup_{t=0}^{\infty} \mathcal{H}_t$.
\end{definition}

\begin{definition}[History-Dependent Policy]
\label{def:hd-policy}
A \emph{history-dependent randomized policy} is a sequence
$\policy = (\policy_0, \policy_1, \policy_2, \ldots)$ where each
$\policy_t : \mathcal{H}_t \to \Delta(\Actions)$ maps histories to
probability distributions over actions. Here
$\Delta(\Actions) = \{p \in \R^{|\Actions|} : p(a) \geq 0,\;
\sum_a p(a) = 1\}$ denotes the probability simplex over $\Actions$.
We write $\policy_t(a \mid h_t)$ for the probability of selecting
action $a$ given history $h_t$.
\end{definition}

\begin{definition}[Markov Policy]
\label{def:markov-policy}
A \emph{Markov randomized policy} is a sequence
$\policy = (\policy_0, \policy_1, \policy_2, \ldots)$ where each
$\policy_t : \States \to \Delta(\Actions)$ depends only on the
current state (not the full history). That is,
$\policy_t(a \mid h_t) = \policy_t(a \mid s_t)$ for all $h_t$
ending in $s_t$.
\end{definition}

\begin{definition}[Stationary Policy]
\label{def:stationary-policy}
A \emph{stationary randomized policy} is a Markov policy that is
time-independent: $\policy_t = \policy_0$ for all $t \geq 0$. We
identify such a policy with a single mapping
$\policy : \States \to \Delta(\Actions)$ and write
$\policy(a \mid s)$ for the probability of action $a$ in state $s$.
\end{definition}

\begin{definition}[Deterministic Policy]
\label{def:det-policy}
A \emph{deterministic stationary policy} is a mapping
$\policy : \States \to \Actions$ assigning a single action to each
state. Equivalently, $\policy(a \mid s) = \indicator\{a = \policy(s)\}$.
We denote the set of all deterministic stationary policies by
$\Pi^{\mathrm{SD}}$ and the set of all stationary randomized policies
by $\Pi^{\mathrm{SR}}$.
\end{definition}

\begin{remark}
\label{rem:policy-hierarchy}
The policy classes form a strict hierarchy:
\[
    \Pi^{\mathrm{SD}} \subset \Pi^{\mathrm{SR}} \subset
    \Pi^{\mathrm{MR}} \subset \Pi^{\mathrm{HR}},
\]
where $\Pi^{\mathrm{MR}}$ and $\Pi^{\mathrm{HR}}$ denote the Markov
randomized and history-dependent randomized policy classes,
respectively. For the discounted infinite-horizon criterion, the
optimal policy can always be found in $\Pi^{\mathrm{SD}}$, so the
larger classes offer no advantage (Theorem~\ref{thm:det-optimal}).
\end{remark}

\begin{warning}[title={Stochastic Policies Are Not Redundant}]
Although deterministic policies suffice for optimality in standard
MDPs, stochastic policies play an essential role in several settings:
partially observable MDPs (POMDPs), multi-agent games (where mixed
strategies are necessary for Nash equilibria), constrained MDPs (where
randomization achieves the constraint boundary), and policy gradient
methods (where stochastic policies enable gradient estimation).
\end{warning}

\begin{example}[Inventory Management MDP]
\label{ex:inventory}
A retailer manages inventory of a single product. The state
$s \in \States = \{0, 1, \ldots, M\}$ is the current stock level,
and the action $a \in \Actions = \{0, 1, \ldots, M\}$ is the number
of units ordered, subject to $s + a \leq M$ (warehouse capacity).
Demand $D$ is random with known distribution $p_D$. The transition
kernel is $P(s' \mid s, a) = \Prob(\max(s + a - D, 0) = s')$. The
reward is $R(s,a) = p \cdot \E[\min(s+a, D)] - c \cdot a
- h \cdot \E[\max(s+a-D, 0)]$, where $p$ is the unit selling price,
$c$ is the unit ordering cost, and $h$ is the per-unit holding cost.
A stationary deterministic policy $\policy(s) = a(s)$ specifies an
order quantity for each stock level.
\end{example}

%% ============================================================
%% SECTION 5: RETURN AND VALUE FUNCTIONS
%% ============================================================

\section{Return and Value Functions}
\label{sec:return-value}

\begin{definition}[Discounted Return]
\label{def:return}
The \emph{discounted return} from time $t$ onward is the random
variable
\[
    G_t = \sum_{k=0}^{\infty} \discount^k R(S_{t+k}, A_{t+k}).
\]
\end{definition}

The first key result establishes that this infinite sum is
well-defined.

\begin{proposition}[The Return Is Well-Defined and Bounded]
\label{prop:return-bounded}
Let $R_{\max} = \max_{s,a} |R(s,a)|$. For any $\discount \in [0,1)$
and any policy $\policy$, the return $G_t$ satisfies:
\begin{enumerate}[nosep]
    \item The series $\sum_{k=0}^{\infty} \discount^k R(S_{t+k},
    A_{t+k})$ converges absolutely almost surely.
    \item $|G_t| \leq R_{\max} / (1 - \discount)$ almost surely.
    \item $\E[|G_t|] \leq R_{\max} / (1 - \discount)$.
\end{enumerate}
\end{proposition}

\begin{proof}
For each $k \geq 0$, $|R(S_{t+k}, A_{t+k})| \leq R_{\max}$, so
\[
    \sum_{k=0}^{\infty} \discount^k |R(S_{t+k}, A_{t+k})|
    \leq R_{\max} \sum_{k=0}^{\infty} \discount^k
    = \frac{R_{\max}}{1 - \discount} < \infty.
\]
Absolute convergence implies convergence, establishing (1). The same
bound gives (2). Taking expectations and applying the monotone
convergence theorem yields (3).
\end{proof}

\begin{remark}
\label{rem:recursive-return}
The return satisfies the recursive relationship
$G_t = R(S_t, A_t) + \discount G_{t+1}$, which is the foundation of
temporal-difference methods and the Bellman equations derived below.
\end{remark}

\begin{definition}[State-Value Function]
\label{def:value-function}
The \emph{state-value function} of policy $\policy$ is
\[
    \Vpi(s) = \E_\policy\!\left[\, G_t \mid S_t = s \,\right]
    = \E_\policy\!\left[\,
    \sum_{k=0}^{\infty} \discount^k R(S_{t+k}, A_{t+k})
    \;\middle|\; S_t = s \,\right],
\]
where $\E_\policy$ denotes expectation under the probability measure
induced by policy $\policy$ and transition kernel $P$.
\end{definition}

\begin{definition}[Action-Value Function]
\label{def:action-value}
The \emph{action-value function} of policy $\policy$ is
\[
    \Qpi(s,a) = \E_\policy\!\left[\, G_t \mid S_t = s,\, A_t = a
    \,\right].
\]
\end{definition}

\begin{definition}[Advantage Function]
\label{def:advantage}
The \emph{advantage function} of policy $\policy$ is
\[
    A^\policy(s,a) = \Qpi(s,a) - \Vpi(s).
\]
It measures the relative advantage of taking action $a$ in state $s$
compared to acting according to $\policy$.
\end{definition}

\begin{proposition}[Value-Action Value Relationship]
\label{prop:v-q-relation}
For any stationary policy $\policy$ and state $s \in \States$,
\[
    \Vpi(s) = \sum_{a \in \Actions} \policy(a \mid s)\, \Qpi(s,a).
\]
Consequently, $\sum_{a \in \Actions} \policy(a \mid s) A^\policy(s,a) = 0$
for all $s$.
\end{proposition}

\begin{proof}
By the law of total expectation (tower property),
\[
    \Vpi(s)
    = \E_\policy[G_t \mid S_t = s]
    = \sum_{a \in \Actions} \policy(a \mid s)\,
      \E_\policy[G_t \mid S_t = s, A_t = a]
    = \sum_{a \in \Actions} \policy(a \mid s)\, \Qpi(s,a).
\]
The advantage identity follows immediately:
$\sum_a \policy(a \mid s) A^\policy(s,a)
= \sum_a \policy(a \mid s) \Qpi(s,a) - \Vpi(s)
= \Vpi(s) - \Vpi(s) = 0$.
\end{proof}

\begin{rlconnection}[title={Advantage Functions in Policy Gradients}]
The advantage function $A^\policy(s,a)$ plays a central role in
variance reduction for policy gradient methods. The policy gradient
theorem (Lesson on Policy Gradients) shows that the gradient of the
expected return can be written as
$\nabla_\theta J(\theta) = \E[\nabla_\theta \log \policy_\theta(A_t
\mid S_t) A^{\policy_\theta}(S_t, A_t)]$. Since
$\sum_a \policy(a \mid s) A^\policy(s,a) = 0$, using the advantage
as a baseline does not bias the gradient estimate but substantially
reduces variance.
\end{rlconnection}

%% ============================================================
%% SECTION 6: BELLMAN EXPECTATION EQUATIONS
%% ============================================================

\section{Bellman Expectation Equations}
\label{sec:bellman-expectation}

The Bellman expectation equations express the value of a state
recursively in terms of the values of successor states. They are the
most fundamental equations in the theory of MDPs.

\begin{theorem}[Bellman Expectation Equation for $\Vpi$]
\label{thm:bellman-v}
For any stationary policy $\policy$ and any state $s \in \States$,
\begin{equation}
\label{eq:bellman-v}
    \Vpi(s) = \sum_{a \in \Actions} \policy(a \mid s)
    \left[ R(s,a) + \discount \sum_{s' \in \States}
    P(s' \mid s,a)\, \Vpi(s') \right].
\end{equation}
\end{theorem}

\begin{proof}
Starting from the recursive decomposition of the return,
$G_t = R(S_t, A_t) + \discount G_{t+1}$, and conditioning on
$S_t = s$:
\begin{align*}
    \Vpi(s)
    &= \E_\policy[G_t \mid S_t = s] \\
    &= \E_\policy[R(S_t, A_t) + \discount G_{t+1} \mid S_t = s] \\
    &= \sum_{a \in \Actions} \policy(a \mid s)
       \E_\policy[R(s, a) + \discount G_{t+1}
       \mid S_t = s, A_t = a] \\
    &= \sum_{a \in \Actions} \policy(a \mid s)
       \left[ R(s,a) + \discount
       \E_\policy[G_{t+1} \mid S_t = s, A_t = a] \right].
\end{align*}
By the Markov property and stationarity of $\policy$,
\[
    \E_\policy[G_{t+1} \mid S_t = s, A_t = a]
    = \sum_{s' \in \States} P(s' \mid s,a)\,
      \E_\policy[G_{t+1} \mid S_{t+1} = s']
    = \sum_{s' \in \States} P(s' \mid s,a)\, \Vpi(s').
\]
The last equality uses stationarity: the value function $\Vpi$ is
time-independent under a stationary policy. Substituting yields
\eqref{eq:bellman-v}.
\end{proof}

\begin{theorem}[Bellman Expectation Equation for $\Qpi$]
\label{thm:bellman-q}
For any stationary policy $\policy$ and any
$(s,a) \in \States \times \Actions$,
\begin{equation}
\label{eq:bellman-q}
    \Qpi(s,a) = R(s,a) + \discount \sum_{s' \in \States}
    P(s' \mid s,a) \sum_{a' \in \Actions}
    \policy(a' \mid s')\, \Qpi(s', a').
\end{equation}
\end{theorem}

\begin{proof}
Decompose the return as $G_t = R(S_t, A_t) + \discount G_{t+1}$ and
condition on $(S_t, A_t) = (s, a)$:
\begin{align*}
    \Qpi(s,a)
    &= \E_\policy[G_t \mid S_t = s, A_t = a] \\
    &= R(s,a) + \discount\, \E_\policy[G_{t+1} \mid S_t = s, A_t = a] \\
    &= R(s,a) + \discount \sum_{s'} P(s' \mid s,a)\,
       \E_\policy[G_{t+1} \mid S_{t+1} = s'] \\
    &= R(s,a) + \discount \sum_{s'} P(s' \mid s,a)\, \Vpi(s').
\end{align*}
Applying Proposition~\ref{prop:v-q-relation} to replace $\Vpi(s')$
gives \eqref{eq:bellman-q}.
\end{proof}

\begin{keyresult}[title={Bellman Expectation Equations -- Matrix Form}]
For a stationary policy $\policy$, define the $S \times S$ transition
matrix $P^\policy$ with entries
$P^\policy_{ss'} = \sum_a \policy(a \mid s) P(s' \mid s,a)$ and the
reward vector $r^\policy \in \R^S$ with entries
$r^\policy_s = \sum_a \policy(a \mid s) R(s,a)$. Then the Bellman
equation \eqref{eq:bellman-v} becomes $v^\policy = r^\policy +
\discount P^\policy v^\policy$, which has the unique solution
\[
    v^\policy = (I - \discount P^\policy)^{-1} r^\policy.
\]
The matrix $I - \discount P^\policy$ is invertible because all
eigenvalues of $\discount P^\policy$ have modulus at most
$\discount < 1$ (since $P^\policy$ is a stochastic matrix).
\end{keyresult}

\begin{example}[Two-State MDP]
\label{ex:two-state}
Consider an MDP with $\States = \{s_1, s_2\}$,
$\Actions = \{a_1, a_2\}$, $\discount = 0.5$, and:
\begin{center}
\begin{tabular}{ccccc}
\toprule
$s$ & $a$ & $R(s,a)$ & $P(s_1 \mid s,a)$ & $P(s_2 \mid s,a)$ \\
\midrule
$s_1$ & $a_1$ & 5 & 0.5 & 0.5 \\
$s_1$ & $a_2$ & 10 & 0 & 1 \\
$s_2$ & $a_1$ & $-1$ & 0.8 & 0.2 \\
$s_2$ & $a_2$ & 2 & 0.3 & 0.7 \\
\bottomrule
\end{tabular}
\end{center}
Under the deterministic policy $\policy(s_1) = a_1$,
$\policy(s_2) = a_1$, the Bellman equation in matrix form is
$v = r + 0.5 P^\policy v$ where $r = (5, -1)^\top$ and
\[
    P^\policy = \begin{pmatrix} 0.5 & 0.5 \\ 0.8 & 0.2 \end{pmatrix}.
\]
Solving $(I - 0.5 P^\policy) v = r$:
\[
    \begin{pmatrix} 0.75 & -0.25 \\ -0.4 & 0.9 \end{pmatrix}
    \begin{pmatrix} v_1 \\ v_2 \end{pmatrix}
    = \begin{pmatrix} 5 \\ -1 \end{pmatrix},
\]
which gives $v_1 = \frac{0.9 \cdot 5 + 0.25 \cdot (-1)}{0.75 \cdot
0.9 - 0.25 \cdot 0.4} = \frac{4.25}{0.575} \approx 7.39$ and
$v_2 = \frac{0.75 \cdot (-1) + 0.4 \cdot 5}{0.575}
= \frac{1.25}{0.575} \approx 2.17$.
\end{example}

%% ============================================================
%% SECTION 7: OPTIMAL POLICIES AND VALUE FUNCTIONS
%% ============================================================

\section{Optimal Policies and Value Functions}
\label{sec:optimal}

\begin{definition}[Optimal Value Functions]
\label{def:optimal-value}
The \emph{optimal state-value function} is
$\Vstar(s) = \sup_{\policy \in \Pi^{\mathrm{HR}}} \Vpi(s)$ for all
$s \in \States$. The \emph{optimal action-value function} is
$\Qstar(s,a) = \sup_{\policy \in \Pi^{\mathrm{HR}}} \Qpi(s,a)$ for
all $(s,a) \in \States \times \Actions$.
\end{definition}

\begin{definition}[Optimal Policy]
\label{def:optimal-policy}
A policy $\optpolicy$ is \emph{optimal} if
$V^{\optpolicy}(s) = \Vstar(s)$ for all $s \in \States$.
\end{definition}

\begin{theorem}[Existence of Deterministic Stationary Optimal Policy]
\label{thm:det-optimal}
For any finite MDP $(\States, \Actions, P, R, \discount)$ with
$\discount \in [0,1)$, there exists a deterministic stationary policy
$\optpolicy \in \Pi^{\mathrm{SD}}$ that is optimal. That is,
$V^{\optpolicy}(s) = \Vstar(s)$ for all $s \in \States$.
\end{theorem}

\begin{proof}
The proof relies on the Bellman optimality equations, which are
developed in detail in Lesson~41. We outline the argument here;
the complete proof appears there.

Since $\States$ and $\Actions$ are finite, there are finitely many
deterministic stationary policies: $|\Pi^{\mathrm{SD}}| = A^S$.
Each deterministic stationary policy $d \in \Pi^{\mathrm{SD}}$ has a
well-defined value function $V^d$, computable via the matrix formula
$V^d = (I - \discount P^d)^{-1} r^d$. Define
$\hat{V}(s) = \max_{d \in \Pi^{\mathrm{SD}}} V^d(s)$. It can be
shown (Lesson~41, using the Bellman optimality operator as a
contraction) that there exists a single deterministic policy
$d^*$ achieving $V^{d^*}(s) = \hat{V}(s)$ for all $s$
simultaneously, and that $\hat{V} = \Vstar$, i.e., no
history-dependent or randomized policy can improve upon $d^*$.
\end{proof}

\begin{rlconnection}[title={Why Deterministic Optimal Policies Exist}]
Theorem~\ref{thm:det-optimal} justifies the design of algorithms
like Q-learning that search for deterministic policies. The greedy
action $\argmax_a Q(s,a)$ is always sufficient -- there is no need
to maintain or learn a stochastic policy for optimality in standard
MDPs. This is why DQN and its variants output a single best action
per state, rather than a probability distribution over actions.
\end{rlconnection}

\begin{example}[Optimal Policy for Two-State MDP]
\label{ex:two-state-optimal}
Returning to Example~\ref{ex:two-state}, we can evaluate all
$|\Pi^{\mathrm{SD}}| = 2^2 = 4$ deterministic policies. Solving the
Bellman equations for each:
\begin{center}
\begin{tabular}{ccccc}
\toprule
$\policy(s_1)$ & $\policy(s_2)$ & $\Vpi(s_1)$ & $\Vpi(s_2)$ \\
\midrule
$a_1$ & $a_1$ & $7.39$ & $2.17$ \\
$a_1$ & $a_2$ & $8.50$ & $5.00$ \\
$a_2$ & $a_1$ & $11.30$ & $3.48$ \\
$a_2$ & $a_2$ & $14.00$ & $6.00$ \\
\bottomrule
\end{tabular}
\end{center}
The optimal policy is $\optpolicy(s_1) = a_2$,
$\optpolicy(s_2) = a_2$, with $\Vstar(s_1) = 14$ and
$\Vstar(s_2) = 6$. As guaranteed by
Theorem~\ref{thm:det-optimal}, a single deterministic stationary
policy achieves the maximum simultaneously at both states.
\end{example}

%% ============================================================
%% SECTION 8: EXERCISES
%% ============================================================

\section{Exercises}
\label{sec:exercises}

\begin{exercise}[MDP Formulation] \textnormal{(\(*\))} \\
A taxi operates in a $5 \times 5$ grid city. The state includes the
taxi's position, whether it is carrying a passenger, and the
passenger's destination (one of 4 designated locations). The taxi can
move in four cardinal directions, pick up, or drop off. Formally
define this problem as an MDP: specify $\States$, $\Actions$, the
transition kernel $P$, reward function $R$, and an appropriate
discount factor $\discount$.
\end{exercise}

\begin{exercise}[Bellman Equation Solution] \textnormal{(\(*\))} \\
Consider a three-state MDP with $\States = \{s_1, s_2, s_3\}$,
$\Actions = \{a\}$ (a single action), $\discount = 0.9$, rewards
$R(s_1) = 1$, $R(s_2) = 2$, $R(s_3) = 0$, and transition matrix
\[
    P = \begin{pmatrix}
    0.2 & 0.5 & 0.3 \\
    0.1 & 0.6 & 0.3 \\
    0.4 & 0.1 & 0.5
    \end{pmatrix}.
\]
Compute $\Vpi(s)$ for each state by solving the linear system
$(I - \discount P) v = r$.
\end{exercise}

\begin{exercise}[Advantage Function Properties]
\textnormal{(\(**\))} \\
Let $\policy$ be a stationary policy for a finite MDP.
\begin{enumerate}[label=(\alph*)]
    \item Prove that $\sum_{a} \policy(a \mid s) A^\policy(s,a) = 0$
    for all $s$.
    \item Show that if $\policy$ is a deterministic policy with
    $\policy(s) = a^*$, then $A^\policy(s, a^*) = 0$ and
    $A^\policy(s, a) = \Qpi(s,a) - \Qpi(s, a^*)$ for all $a$.
    \item Prove that a stationary policy $\policy$ is optimal if and
    only if $A^\policy(s,a) \leq 0$ for all $(s,a)$.
\end{enumerate}
\end{exercise}

\begin{exercise}[Inventory MDP Analysis] \textnormal{(\(**\))} \\
For the inventory MDP of Example~\ref{ex:inventory} with $M = 2$,
demand $\Prob(D = 0) = 0.3$, $\Prob(D = 1) = 0.5$,
$\Prob(D = 2) = 0.2$, $p = 4$, $c = 2$, $h = 1$, and
$\discount = 0.95$:
\begin{enumerate}[label=(\alph*)]
    \item Write out the transition matrix $P(s' \mid s, a)$ for each
    feasible state-action pair.
    \item Compute the reward $R(s,a)$ for each feasible pair.
    \item Evaluate $\Vpi$ for the policy ``order up to 2'' (i.e.,
    $\policy(s) = 2 - s$).
\end{enumerate}
\end{exercise}

\begin{exercise}[Policy Hierarchy and Counterexample]
\textnormal{(\(***\))} \\
Theorem~\ref{thm:det-optimal} states that deterministic stationary
policies suffice for optimality in finite discounted MDPs. Construct
an explicit example demonstrating that this fails for constrained MDPs:
give a two-state, two-action MDP with a constraint on average cost
where the optimal constrained policy is necessarily stochastic (i.e.,
no deterministic stationary policy is optimal among policies satisfying
the constraint).

\textit{Hint:} Consider a constraint of the form
$\E_\policy[\sum_t \discount^t c(S_t, A_t)] \leq C$ with
appropriately chosen costs $c$ and threshold $C$.
\end{exercise}

%% ============================================================
%% SECTION 9: SUMMARY
%% ============================================================

\section{Summary}
\label{sec:summary}

\begin{itemize}
    \item A Markov decision process
    $\mathcal{M} = (\States, \Actions, P, R, \discount)$ is the
    canonical framework for sequential decision-making under
    uncertainty, defined by states, actions, transitions, rewards, and
    a discount factor.
    \item The Markov property ensures that the future depends on the
    past only through the current state and action, enabling recursive
    decompositions.
    \item Policies range from history-dependent randomized (most
    general) to deterministic stationary (most restrictive):
    $\Pi^{\mathrm{SD}} \subset \Pi^{\mathrm{SR}} \subset
    \Pi^{\mathrm{MR}} \subset \Pi^{\mathrm{HR}}$.
    \item The discounted return $G_t = \sum_{k=0}^\infty \discount^k
    R(S_{t+k}, A_{t+k})$ is well-defined and bounded by
    $R_{\max}/(1-\discount)$ whenever $\discount < 1$.
    \item The state-value function $\Vpi(s)$ and action-value function
    $\Qpi(s,a)$ are linked by
    $\Vpi(s) = \sum_a \policy(a \mid s) \Qpi(s,a)$.
    \item The advantage function $A^\policy(s,a) = \Qpi(s,a) - \Vpi(s)$
    measures the relative merit of action $a$; its expected value
    under $\policy$ is zero.
    \item The Bellman expectation equations express $\Vpi$ and $\Qpi$
    recursively and, in the finite case, reduce to a linear system
    with unique solution $v^\policy = (I - \discount P^\policy)^{-1}
    r^\policy$.
    \item For finite MDPs with $\discount < 1$, an optimal
    deterministic stationary policy always exists.
\end{itemize}

%% ============================================================
%% SECTION 10: FURTHER READING
%% ============================================================

\section{Further Reading}
\label{sec:further-reading}

\begin{itemize}
    \item \textbf{Lesson~40} extends the MDP framework to
    finite-horizon problems, develops backward induction, and proves
    the principle of optimality.
    \item \textbf{Lesson~41} derives the Bellman optimality equations,
    introduces the Bellman operators $\Bellman^\policy$ and
    $\Bellman^*$, and proves the optimality of the greedy policy with
    respect to $\Vstar$.
    \item \citet[Chapters 4--6]{puterman2014} provides an exhaustive
    treatment of optimality criteria, including average reward and
    total reward formulations beyond the discounted setting.
    \item \citet{blackwell1962} proves the existence of optimal
    stationary policies using contraction arguments; the original
    paper remains highly readable.
    \item \citet{altman1999} extends the MDP framework to constrained
    settings where stochastic policies are necessary for optimality,
    complementing Exercise~5 above.
\end{itemize}

\bibliography{mdp-refs}
\end{document}
