\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 44: Average Reward MDPs}}
\fancyhead[R]{\thepage}
\title{Lesson 44: Average Reward MDPs}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% ============================================================
%% SECTION 1: MOTIVATION AND OVERVIEW
%% ============================================================

\section{Motivation and Overview}
\label{sec:motivation}

Consider a data center scheduling agent that manages server workloads
continuously, with no natural endpoint. The agent routes incoming
requests to servers, balancing throughput against latency. Should the
agent use a discount factor $\discount = 0.99$? Or $\discount = 0.999$?
The choice of $\discount$ affects both the optimal policy and the
convergence rate of algorithms, yet in many continuous-operation
settings there is no intrinsic reason to discount future performance.
The agent should maximize its \emph{long-run average reward per step}
-- the rate of reward accumulation over an infinite horizon.

The average reward criterion eliminates the discount factor entirely,
replacing the discounted return
$\sum_{t=0}^\infty \discount^t R(S_t, A_t)$ with the long-run
average $\lim_{T \to \infty} \frac{1}{T} \sum_{t=0}^{T-1}
R(S_t, A_t)$. This leads to a fundamentally different mathematical
structure: the Bellman equation acquires a scalar offset (the
\emph{gain}), value functions are defined only up to an additive
constant, and the analysis requires ergodicity conditions on the
Markov chain. This lesson develops the theory of average reward MDPs,
culminating in the gain-bias equations and a brief treatment of
Blackwell optimality.

\begin{rlconnection}[title={Why Average Reward MDPs Matter for RL/ML}]
\begin{itemize}[nosep]
    \item \textbf{Continuing tasks}: Many real-world RL applications
    (network routing, inventory management, robot operation) have no
    natural episode boundary; the average reward criterion is their
    natural objective.
    \item \textbf{R-learning}: The R-learning algorithm is the
    average-reward analogue of Q-learning, directly estimating the
    gain and relative values without a discount factor.
    \item \textbf{Sensitivity to $\discount$}: Discounted algorithms
    suffer from sensitivity to the choice of $\discount$; the
    average reward criterion avoids this hyperparameter entirely.
    \item \textbf{Blackwell optimality}: A Blackwell optimal policy
    is simultaneously optimal for all discount factors sufficiently
    close to 1, connecting discounted and average reward theory.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
MDP model, stationary policies, Bellman operators, contraction
properties, value iteration, policy iteration (Lessons~39--43).
Markov chains, stationary distributions, ergodicity (Lesson~12).
Linear algebra: eigenvalues, matrix inverses, rank (Lessons~5--6).

\paragraph{Learning objectives.}
After completing this lesson, the student will be able to:
\begin{itemize}
    \item \emph{Define} the average reward criterion and the gain of
    a policy.
    \item \emph{State} and \emph{prove} the gain-bias (average
    reward Bellman) equations for unichain MDPs.
    \item \emph{Classify} MDPs as unichain, multichain, or
    communicating, and \emph{explain} why the classification matters.
    \item \emph{Derive} the relationship between average reward and
    discounted reward as $\discount \to 1$.
    \item \emph{Define} Blackwell optimality and \emph{state}
    Blackwell's theorem on the existence of Blackwell optimal
    policies.
    \item \emph{Compute} the gain and bias for small unichain MDPs.
\end{itemize}

%% ============================================================
%% SECTION 2: REFERENCES
%% ============================================================

\section{References}
\label{sec:references}

\begin{itemize}
    \item \citet[Chapters 8--9]{puterman2014} -- Comprehensive
    treatment of average reward MDPs, gain-bias equations, unichain
    and multichain theory, and Blackwell optimality.
    \item \citet[Chapter 7]{bertsekas2012} -- Average cost dynamic
    programming, including the relative value iteration algorithm.
    \item \citet{blackwell1962} -- Original paper proving the
    existence of discount-optimal policies that are simultaneously
    optimal for all large enough discount factors.
    \item \citet{howard1960} -- Introduction of the gain and bias
    framework and policy iteration for average reward.
    \item \citet{derman1970} -- Foundational treatment of average
    reward in finite Markov decision processes.
\end{itemize}

%% ============================================================
%% SECTION 3: MARKOV CHAIN CLASSIFICATION
%% ============================================================

\section{Markov Chain Classification}
\label{sec:classification}

The average reward criterion requires careful analysis of the
long-run behavior of Markov chains induced by policies. We begin
by recalling the relevant classification.

\begin{definition}[Induced Markov Chain]
\label{def:induced-chain}
A stationary policy $\policy$ on an MDP
$\mathcal{M} = (\States, \Actions, P, R, \discount)$ induces a
Markov chain with transition matrix $P^\policy \in \R^{S \times S}$,
where $P^\policy_{ss'} = \sum_{a \in \Actions} \policy(a \mid s)
P(s' \mid s, a)$.
\end{definition}

\begin{definition}[Communication Classes]
\label{def:communication}
State $s$ \emph{communicates with} state $s'$ (written $s
\leftrightarrow s'$) if there exist integers $m, n \geq 1$ such that
$(P^\policy)^m_{ss'} > 0$ and $(P^\policy)^n_{s's} > 0$. A
\emph{communication class} (or \emph{recurrent class}) is a maximal
set of mutually communicating states. States not in any recurrent
class are \emph{transient}.
\end{definition}

\begin{definition}[Chain Types]
\label{def:chain-types}
A Markov chain with transition matrix $P^\policy$ is:
\begin{enumerate}[nosep]
    \item \emph{Irreducible} (or \emph{ergodic}) if all states form
    a single communication class.
    \item \emph{Unichain} if it has exactly one recurrent class
    (plus possibly transient states).
    \item \emph{Multichain} if it has two or more recurrent classes.
\end{enumerate}
An MDP is \emph{unichain} if every deterministic stationary policy
induces a unichain Markov chain. An MDP is \emph{communicating} if
for every pair of states $s, s'$, there exists a deterministic
stationary policy $d$ such that $s'$ is reachable from $s$ under $d$.
\end{definition}

\begin{example}[Unichain vs.\ Multichain]
\label{ex:chain-types}
Consider $\States = \{1, 2, 3\}$ with two deterministic policies:

Policy $d_1$: $P^{d_1} = \bigl(\begin{smallmatrix}
0 & 1 & 0 \\ 0 & 0.5 & 0.5 \\ 1 & 0 & 0
\end{smallmatrix}\bigr)$. All states communicate (irreducible).

Policy $d_2$: $P^{d_2} = \bigl(\begin{smallmatrix}
1 & 0 & 0 \\ 0 & 0 & 1 \\ 0 & 1 & 0
\end{smallmatrix}\bigr)$. State 1 is an absorbing recurrent class;
$\{2, 3\}$ is another recurrent class. This is multichain.

If these are the only two deterministic policies, the MDP is
\emph{not} unichain (since $d_2$ is multichain). However, it is
communicating, because from state 1 we can reach states 2 and 3
under $d_1$.
\end{example}

\begin{proposition}[Stationary Distribution for Unichain Policies]
\label{prop:stationary-dist}
If policy $\policy$ induces a unichain Markov chain, then there
exists a unique stationary distribution
$\mu^\policy \in \Delta(\States)$ satisfying
$(\mu^\policy)^\top P^\policy = (\mu^\policy)^\top$. Moreover,
$\mu^\policy(s) > 0$ for all states $s$ in the recurrent class,
and $\mu^\policy(s) = 0$ for transient states.
\end{proposition}

\begin{proof}
By the Perron-Frobenius theorem, a stochastic matrix with a single
recurrent class has eigenvalue 1 with algebraic multiplicity 1. The
corresponding left eigenvector, normalized to sum to 1, is the
unique stationary distribution. States in the recurrent class have
positive weight because the restriction of $P^\policy$ to the
recurrent class is irreducible.
\end{proof}

%% ============================================================
%% SECTION 4: AVERAGE REWARD CRITERION
%% ============================================================

\section{The Average Reward Criterion}
\label{sec:average-reward}

\begin{definition}[Average Reward / Gain]
\label{def:gain}
The \emph{gain} (or \emph{average reward}) of a stationary policy
$\policy$ starting from state $s$ is
\[
    \rho^\policy(s) = \liminf_{T \to \infty}
    \frac{1}{T} \E_\policy\!\left[
    \sum_{t=0}^{T-1} R(S_t, A_t) \;\middle|\; S_0 = s \right].
\]
\end{definition}

\begin{remark}
\label{rem:liminf}
We use $\liminf$ rather than $\lim$ because the limit may not exist
for general (multichain) MDPs. For unichain MDPs, the limit exists
and equals $\rho^\policy(s)$ for all $s$ (see
Theorem~\ref{thm:gain-unichain} below).
\end{remark}

\begin{theorem}[Gain of a Unichain Policy]
\label{thm:gain-unichain}
If policy $\policy$ induces a unichain Markov chain with stationary
distribution $\mu^\policy$, then the gain is independent of the
initial state:
\[
    \rho^\policy(s) = \rho^\policy
    := \sum_{s' \in \States} \mu^\policy(s')
    \sum_{a \in \Actions} \policy(a \mid s') R(s', a)
    = (\mu^\policy)^\top r^\policy
    \quad \text{for all } s \in \States,
\]
where $r^\policy \in \R^S$ has entries
$r^\policy_s = \sum_a \policy(a \mid s) R(s,a)$.
\end{theorem}

\begin{proof}
For a unichain policy, the ergodic theorem for finite Markov chains
gives
\[
    \lim_{T \to \infty} \frac{1}{T} \sum_{t=0}^{T-1}
    \E_\policy[R(S_t, A_t) \mid S_0 = s]
    = \sum_{s'} \mu^\policy(s') r^\policy_{s'}
    = (\mu^\policy)^\top r^\policy,
\]
regardless of the initial state $s$. This is because
$(P^\policy)^t_{ss'} \to \mu^\policy(s')$ as $t \to \infty$ for
all $s$ (convergence to stationarity in the unichain case), so the
Cesaro average converges to the stationary expectation.
\end{proof}

\begin{definition}[Optimal Gain]
\label{def:optimal-gain}
The \emph{optimal gain} is
$\rho^* = \sup_{\policy} \rho^\policy$. For unichain MDPs, the
supremum is over all stationary policies (the gain is
state-independent).
\end{definition}

\begin{example}[Average Reward of a Simple MDP]
\label{ex:avg-simple}
Consider $\States = \{1, 2\}$, $\Actions = \{a\}$ (single action),
with $P(1 \mid 1, a) = 0.3$, $P(2 \mid 1, a) = 0.7$,
$P(1 \mid 2, a) = 0.6$, $P(2 \mid 2, a) = 0.4$. Rewards:
$R(1, a) = 5$, $R(2, a) = 1$.

The stationary distribution satisfies $\mu(1) \cdot 0.7 =
\mu(2) \cdot 0.6$, so $\mu(1)/\mu(2) = 6/7$. Normalizing:
$\mu(1) = 6/13$, $\mu(2) = 7/13$.

The gain is $\rho = \mu(1) R(1,a) + \mu(2) R(2,a) =
\frac{6}{13} \cdot 5 + \frac{7}{13} \cdot 1 =
\frac{30 + 7}{13} = \frac{37}{13} \approx 2.85$.
\end{example}

%% ============================================================
%% SECTION 5: GAIN-BIAS EQUATIONS
%% ============================================================

\section{The Gain-Bias Equations}
\label{sec:gain-bias}

The average reward analogue of the Bellman equations involves two
quantities: the gain $\rho^\policy$ (a scalar) and the \emph{bias}
$h^\policy$ (a vector), also called the \emph{relative value
function} or \emph{differential value function}.

\begin{definition}[Bias / Relative Value Function]
\label{def:bias}
For a unichain policy $\policy$ with gain $\rho^\policy$, the
\emph{bias} (or \emph{relative value function}) is
\[
    h^\policy(s) = \E_\policy\!\left[
    \sum_{t=0}^{\infty} \left(R(S_t, A_t) - \rho^\policy\right)
    \;\middle|\; S_0 = s \right].
\]
\end{definition}

\begin{remark}
\label{rem:bias-convergence}
The series in the bias definition converges because $R(S_t, A_t) -
\rho^\policy$ has zero mean under the stationary distribution, and
the deviations decay geometrically due to ergodicity. Formally,
$|\E_\policy[R(S_t, A_t) - \rho^\policy \mid S_0 = s]| \leq
C \lambda^t$ for some $C > 0$ and $\lambda \in (0,1)$ depending on
the mixing rate of the chain.
\end{remark}

\begin{theorem}[Gain-Bias Equations for Policy Evaluation]
\label{thm:gain-bias-eval}
For a unichain policy $\policy$, the gain $\rho^\policy \in \R$ and
bias $h^\policy \in \R^S$ satisfy the \emph{gain-bias equations}
(or \emph{Poisson equations}):
\begin{equation}
\label{eq:gain-bias}
    \rho^\policy + h^\policy(s) = \sum_{a \in \Actions}
    \policy(a \mid s) \left[ R(s,a) + \sum_{s' \in \States}
    P(s' \mid s,a)\, h^\policy(s') \right]
    \quad \text{for all } s \in \States.
\end{equation}
In matrix form: $\rho^\policy \mathbf{1} + h^\policy =
r^\policy + P^\policy h^\policy$, or equivalently
$(I - P^\policy) h^\policy = r^\policy -
\rho^\policy \mathbf{1}$.
\end{theorem}

\begin{proof}
Starting from the bias definition, for any $s$:
\begin{align*}
    h^\policy(s)
    &= \E_\policy\!\left[
    \sum_{t=0}^{\infty} (R(S_t, A_t) - \rho^\policy)
    \;\middle|\; S_0 = s \right] \\
    &= \E_\policy[R(S_0, A_0) - \rho^\policy \mid S_0 = s]
    + \E_\policy\!\left[
    \sum_{t=1}^{\infty} (R(S_t, A_t) - \rho^\policy)
    \;\middle|\; S_0 = s \right] \\
    &= \sum_a \policy(a \mid s) R(s,a) - \rho^\policy
    + \sum_{s'} P^\policy_{ss'}\,
    \E_\policy\!\left[
    \sum_{t=0}^{\infty} (R(S_t, A_t) - \rho^\policy)
    \;\middle|\; S_0 = s' \right] \\
    &= r^\policy_s - \rho^\policy
    + \sum_{s'} P^\policy_{ss'}\, h^\policy(s').
\end{align*}
The third equality uses the Markov property and stationarity.
Rearranging gives \eqref{eq:gain-bias}.
\end{proof}

\begin{remark}
\label{rem:bias-uniqueness}
The gain $\rho^\policy$ is unique, but the bias $h^\policy$ is
determined only up to an additive constant: if $({\rho}, h)$
solves \eqref{eq:gain-bias}, then so does $({\rho}, h + c\mathbf{1})$
for any $c \in \R$. This is because $(I - P^\policy)\mathbf{1} = 0$
(the rows of $P^\policy$ sum to 1). To obtain a unique solution, one
typically imposes a normalization such as
$\sum_s \mu^\policy(s) h^\policy(s) = 0$ or $h^\policy(s_0) = 0$
for a reference state $s_0$.
\end{remark}

\begin{keyresult}[title={Gain-Bias Equations (Poisson Equations)}]
For a unichain policy $\policy$ with gain $\rho^\policy$ and bias
$h^\policy$:
\[
    \rho^\policy + h^\policy(s)
    = \sum_a \policy(a \mid s) \left[ R(s,a) + \sum_{s'}
    P(s' \mid s,a)\, h^\policy(s') \right],
    \quad s \in \States.
\]
Compared to the discounted Bellman equation
$\Vpi(s) = \sum_a \policy(a \mid s)[R(s,a) + \discount \sum_{s'}
P(s' \mid s,a) \Vpi(s')]$, the gain $\rho^\policy$ replaces
the discount factor's role: it is the constant rate subtracted from
immediate rewards to make the remaining value bounded.
\end{keyresult}

\begin{example}[Computing Gain and Bias]
\label{ex:gain-bias-compute}
Return to Example~\ref{ex:avg-simple}: $\States = \{1, 2\}$,
$r = (5, 1)^\top$, $P = \bigl(\begin{smallmatrix}
0.3 & 0.7 \\ 0.6 & 0.4 \end{smallmatrix}\bigr)$, and we
computed $\rho = 37/13$.

The gain-bias equations are:
\begin{align*}
    \frac{37}{13} + h(1) &= 5 + 0.3 h(1) + 0.7 h(2), \\
    \frac{37}{13} + h(2) &= 1 + 0.6 h(1) + 0.4 h(2).
\end{align*}
Rearranging: $0.7 h(1) - 0.7 h(2) = 5 - 37/13 = 28/13$ and
$-0.6 h(1) + 0.6 h(2) = 1 - 37/13 = -24/13$.

From the first: $h(1) - h(2) = 28/(13 \cdot 0.7) = 40/13$.
Setting $h(2) = 0$ (normalization): $h(1) = 40/13 \approx 3.08$.

The bias $h(1) > h(2)$ reflects that state 1 is ``better than
average'' (reward 5 $>$ gain $37/13 \approx 2.85$), so starting
there yields a transient bonus.
\end{example}

%% ============================================================
%% SECTION 6: AVERAGE REWARD OPTIMALITY
%% ============================================================

\section{Average Reward Optimality Equations}
\label{sec:avg-optimality}

\begin{theorem}[Average Reward Bellman Optimality Equations]
\label{thm:avg-optimality}
For a unichain MDP, there exist a scalar $\rho^*$ (the optimal gain)
and a vector $h^* \in \R^S$ (the optimal bias) satisfying:
\begin{equation}
\label{eq:avg-bellman-opt}
    \rho^* + h^*(s) = \max_{a \in \Actions}
    \left[ R(s,a) + \sum_{s' \in \States} P(s' \mid s,a)\, h^*(s')
    \right]
    \quad \text{for all } s \in \States.
\end{equation}
The optimal gain $\rho^*$ is unique. The optimal bias $h^*$ is unique
up to an additive constant.
\end{theorem}

\begin{proof}
The proof uses the connection to discounted MDPs via the Laurent
series expansion (developed in
Theorem~\ref{thm:discount-to-average}). We provide an outline.

\textbf{Step 1: Existence.}
Consider the discounted optimal value function $\Vstar_\discount$ as
a function of $\discount$. For a unichain MDP, define
\[
    \rho^* = \lim_{\discount \uparrow 1} (1-\discount)
    \Vstar_\discount(s)
\]
(this limit exists and is independent of $s$ for unichain MDPs) and
\[
    h^*(s) = \lim_{\discount \uparrow 1}
    [\Vstar_\discount(s) - \Vstar_\discount(s_0)]
\]
for some reference state $s_0$. Taking the limit
$\discount \uparrow 1$ in the discounted Bellman optimality equation
$\Vstar_\discount(s) = \max_a [R(s,a) + \discount \sum_{s'}
P(s' \mid s,a) \Vstar_\discount(s')]$ yields
\eqref{eq:avg-bellman-opt}.

\textbf{Step 2: Uniqueness of $\rho^*$.}
Suppose $(\rho_1, h_1)$ and $(\rho_2, h_2)$ both satisfy
\eqref{eq:avg-bellman-opt} with $\rho_1 > \rho_2$. Let $d_2$ be the
greedy policy w.r.t.\ $h_2$, i.e., the maximizer in
\eqref{eq:avg-bellman-opt} for $(\rho_2, h_2)$. Then for all $s$:
$\rho_1 + h_1(s) \geq R(s, d_2(s)) + \sum_{s'} P(s' \mid s,
d_2(s)) h_1(s') = \rho_2 + h_2(s) + \sum_{s'} P(s' \mid s,
d_2(s))[h_1(s') - h_2(s')]$.
Summing with the stationary distribution $\mu^{d_2}$ and using
$\sum_s \mu^{d_2}(s) \sum_{s'} P(s' \mid s, d_2(s)) f(s') =
\sum_{s'} \mu^{d_2}(s') f(s')$ for any $f$, we get a contradiction
unless $\rho_1 = \rho_2$.
\end{proof}

\begin{definition}[Average Reward Greedy Policy]
\label{def:avg-greedy}
Given a solution $(\rho^*, h^*)$ to \eqref{eq:avg-bellman-opt}, the
\emph{average reward optimal policy} is
\[
    \optpolicy(s) = \argmax_{a \in \Actions}
    \left[ R(s,a) + \sum_{s'} P(s' \mid s,a)\, h^*(s') \right].
\]
\end{definition}

\begin{theorem}[Optimality of the Greedy Policy]
\label{thm:avg-greedy-optimal}
For a unichain MDP, the greedy policy $\optpolicy$ with respect to
$h^*$ achieves the optimal gain: $\rho^{\optpolicy} = \rho^*$.
\end{theorem}

\begin{proof}
Since $\optpolicy$ achieves the max in \eqref{eq:avg-bellman-opt},
$h^*$ satisfies the gain-bias equation
$\rho^* + h^*(s) = R(s, \optpolicy(s)) + \sum_{s'} P(s' \mid s,
\optpolicy(s)) h^*(s')$ for all $s$. Multiplying by
$\mu^{\optpolicy}(s)$ and summing:
$\rho^* = \sum_s \mu^{\optpolicy}(s) R(s, \optpolicy(s)) =
\rho^{\optpolicy}$.
\end{proof}

\begin{rlconnection}[title={Gain-Bias Equations and R-Learning}]
The R-learning algorithm directly estimates $\rho^*$ and $h^*$ using
stochastic approximation. The update rules are:
\begin{align*}
    h(s_t, a_t) &\leftarrow h(s_t, a_t) + \alpha_t
    [R(s_t, a_t) - \hat{\rho} + \max_{a'} h(s_{t+1}, a') -
    h(s_t, a_t)], \\
    \hat{\rho} &\leftarrow \hat{\rho} + \beta_t
    [R(s_t, a_t) + \max_{a'} h(s_{t+1}, a') -
    \max_{a'} h(s_t, a') - \hat{\rho}].
\end{align*}
This is the average-reward analogue of Q-learning, with the
gain estimate $\hat{\rho}$ replacing the discount factor.
\end{rlconnection}

%% ============================================================
%% SECTION 7: CONNECTION TO DISCOUNTED REWARD
%% ============================================================

\section{Connection to the Discounted Criterion}
\label{sec:discount-connection}

The average reward criterion can be viewed as the limiting case of
the discounted criterion as $\discount \to 1$.

\begin{theorem}[Discount-to-Average Limit]
\label{thm:discount-to-average}
For a unichain MDP with optimal discounted value $\Vstar_\discount$
and optimal average reward $\rho^*$:
\begin{enumerate}[nosep]
    \item $\lim_{\discount \uparrow 1} (1-\discount)
    \Vstar_\discount(s) = \rho^*$ for all $s \in \States$.
    \item The limit $h^*(s) = \lim_{\discount \uparrow 1}
    [\Vstar_\discount(s) - \Vstar_\discount(s_0)]$ exists for any
    reference state $s_0$ and satisfies the average reward
    Bellman optimality equation \eqref{eq:avg-bellman-opt}.
\end{enumerate}
\end{theorem}

\begin{proof}
\textbf{Part 1:} The discounted value satisfies
$\|\Vstar_\discount\|_\infty \leq R_{\max}/(1-\discount)$, so
$(1-\discount)\|\Vstar_\discount\|_\infty \leq R_{\max}$, and
$(1-\discount)\Vstar_\discount$ is bounded.

For a fixed deterministic policy $d$, the discounted value is
$V^d_\discount = (I - \discount P^d)^{-1} r^d$. Using the
Neumann series:
$(I - \discount P^d)^{-1} = \sum_{t=0}^{\infty} \discount^t
(P^d)^t$.

For a unichain policy $d$, $(P^d)^t \to \mathbf{1}(\mu^d)^\top$
as $t \to \infty$, where $\mu^d$ is the stationary distribution.
Therefore:
\begin{align*}
    (1-\discount) V^d_\discount(s)
    &= (1-\discount) \sum_{t=0}^{\infty} \discount^t
    [(P^d)^t r^d]_s \\
    &\to (\mu^d)^\top r^d = \rho^d
    \quad \text{as } \discount \to 1.
\end{align*}
The interchange of limit and summation is justified by dominated
convergence (the partial sums are bounded by $R_{\max}$).

Since $\Vstar_\discount(s) = \max_d V^d_\discount(s)$ and the max
is over finitely many deterministic policies:
\begin{align*}
    \lim_{\discount \to 1} (1-\discount) \Vstar_\discount(s)
    = \max_d \lim_{\discount \to 1} (1-\discount) V^d_\discount(s)
    = \max_d \rho^d = \rho^*.
\end{align*}
The interchange of max and limit is valid because the max is over a
finite set.

\textbf{Part 2:} The centered function
$\tilde{V}_\discount(s) = \Vstar_\discount(s) -
\Vstar_\discount(s_0)$ satisfies $\|\tilde{V}_\discount\|_\infty
\leq 2R_{\max}/(1-\discount)$, but the span is bounded:
$\mathrm{sp}(\tilde{V}_\discount) = \mathrm{sp}(\Vstar_\discount)$.

For a unichain MDP, $\mathrm{sp}(\Vstar_\discount)$ remains bounded
as $\discount \to 1$ (this uses the span contraction from
Lesson~43). By compactness (Bolzano-Weierstrass on $\R^S$), there
exists a convergent subsequence, and one can show the limit satisfies
\eqref{eq:avg-bellman-opt} by passing to the limit in the discounted
Bellman equation. Uniqueness of $\rho^*$ ensures the full limit
exists.
\end{proof}

\begin{example}[Discounted to Average Reward]
\label{ex:discount-to-avg}
For the MDP in Example~\ref{ex:avg-simple}, the discounted value
under the single policy with $\discount = 0.99$ is
$V_\discount(1) = (I - 0.99P)^{-1} r$. With
$r = (5, 1)^\top$ and $P = \bigl(\begin{smallmatrix}
0.3 & 0.7 \\ 0.6 & 0.4\end{smallmatrix}\bigr)$:
\[
    I - 0.99P = \begin{pmatrix}
    0.703 & -0.693 \\ -0.594 & 0.604
    \end{pmatrix}.
\]
The determinant is $0.703 \cdot 0.604 - 0.693 \cdot 0.594
= 0.4246 - 0.4117 = 0.0129$. Solving:
$V_{0.99}(1) = (0.604 \cdot 5 + 0.693 \cdot 1)/0.0129
= 3.713/0.0129 \approx 287.8$.

Then $(1-0.99) V_{0.99}(1) = 0.01 \cdot 287.8 \approx 2.88
\approx 37/13 = \rho$. As $\discount \to 1$, the convergence
becomes exact.
\end{example}

\begin{warning}[title={Average Reward Requires Ergodicity}]
The limit $(1-\discount)\Vstar_\discount(s) \to \rho^*$ independent
of $s$ relies on the unichain assumption. For multichain MDPs, the
limit can depend on the initial state, and the average reward
optimality equations are more complex. In such cases, additional
structure (e.g., communicating MDP assumption) or state-dependent
gain is needed.
\end{warning}

%% ============================================================
%% SECTION 8: RELATIVE VALUE ITERATION
%% ============================================================

\section{Relative Value Iteration}
\label{sec:rvi}

Value iteration cannot be directly applied to the average reward
setting because there is no contraction: the Bellman operator
$(\Bellman_{avg} h)(s) = \max_a [R(s,a) + \sum_{s'} P(s' \mid s,a)
h(s')]$ (with no discount) satisfies
$\|\Bellman_{avg} u - \Bellman_{avg} v\|_\infty \leq
\|u - v\|_\infty$ (non-expansive, but not a contraction). However,
\emph{relative value iteration} resolves this by centering the
iterates.

\begin{definition}[Relative Value Iteration]
\label{def:rvi}
\emph{Relative value iteration} (RVI) generates sequences
$\{h^{(n)}\}$ and $\{\rho^{(n)}\}$ by:
\begin{enumerate}[nosep]
    \item \textbf{Initialize:} Choose $h^{(0)} \in \R^S$.
    \item \textbf{Iterate:} For $n = 0, 1, 2, \ldots$,
    \begin{align*}
        w^{(n+1)}(s) &= \max_{a \in \Actions}
        \left[ R(s,a) + \sum_{s'} P(s' \mid s,a)\, h^{(n)}(s')
        \right], \\
        \rho^{(n+1)} &= w^{(n+1)}(s_0) \quad
        \text{(for a fixed reference state } s_0\text{)}, \\
        h^{(n+1)}(s) &= w^{(n+1)}(s) - \rho^{(n+1)}.
    \end{align*}
\end{enumerate}
\end{definition}

\begin{remark}
\label{rem:rvi-centering}
The subtraction $h^{(n+1)}(s) = w^{(n+1)}(s) - w^{(n+1)}(s_0)$
ensures $h^{(n+1)}(s_0) = 0$ at every iteration, preventing the
iterates from diverging. This is necessary because the undiscounted
Bellman operator shifts values by the gain at each step.
\end{remark}

\begin{theorem}[Convergence of Relative Value Iteration]
\label{thm:rvi-convergence}
For a unichain MDP, relative value iteration converges:
$\rho^{(n)} \to \rho^*$ and $h^{(n)} \to h^*$ (up to the
normalization $h^*(s_0) = 0$).
\end{theorem}

\begin{proof}
The span contraction from Lesson~43 is the key tool. Since
$\mathrm{sp}(\Bellman_{avg} u - \Bellman_{avg} v) \leq
\mathrm{sp}(u - v)$ for the undiscounted operator (with contraction
coefficient 1), the convergence analysis requires the stronger
result that the span contraction coefficient is strictly less than 1
for unichain MDPs, a result known as the
\emph{span condition}. We outline the argument.

Define $\sigma = \max_{a_1, a_2} \min_{s'} [P(s' \mid \cdot, a_1)
+ P(s' \mid \cdot, a_2)]$ (the \emph{scrambling coefficient}). For
a communicating MDP, $\sigma > 0$, and the span contraction
coefficient of $\Bellman_{avg}$ is at most $1 - \sigma < 1$.
Standard results then give convergence in the span seminorm, which
implies convergence of $\rho^{(n)}$ and convergence of $h^{(n)}$ up
to a constant.

A complete proof can be found in \citet[Chapter 8]{puterman2014}.
\end{proof}

\begin{example}[Relative Value Iteration]
\label{ex:rvi}
For the MDP of Example~\ref{ex:avg-simple} with a single action,
starting from $h^{(0)} = (0, 0)^\top$ and $s_0 = 2$:

\textbf{Iteration 1:}
$w^{(1)}(1) = 5 + 0.3 \cdot 0 + 0.7 \cdot 0 = 5$,
$w^{(1)}(2) = 1 + 0.6 \cdot 0 + 0.4 \cdot 0 = 1$.
$\rho^{(1)} = w^{(1)}(2) = 1$.
$h^{(1)} = (5 - 1, 1 - 1)^\top = (4, 0)^\top$.

\textbf{Iteration 2:}
$w^{(2)}(1) = 5 + 0.3 \cdot 4 + 0.7 \cdot 0 = 6.2$,
$w^{(2)}(2) = 1 + 0.6 \cdot 4 + 0.4 \cdot 0 = 3.4$.
$\rho^{(2)} = 3.4$.
$h^{(2)} = (6.2 - 3.4, 0)^\top = (2.8, 0)^\top$.

\textbf{Iteration 3:}
$w^{(3)}(1) = 5 + 0.3 \cdot 2.8 = 5.84$,
$w^{(3)}(2) = 1 + 0.6 \cdot 2.8 = 2.68$.
$\rho^{(3)} = 2.68$.
$h^{(3)} = (3.16, 0)^\top$.

The gain sequence $1, 3.4, 2.68, \ldots$ oscillates toward
$\rho^* = 37/13 \approx 2.846$, and the bias
$h^{(n)}(1) \to 40/13 \approx 3.077$.
\end{example}

%% ============================================================
%% SECTION 9: BLACKWELL OPTIMALITY
%% ============================================================

\section{Blackwell Optimality}
\label{sec:blackwell}

Blackwell optimality provides a remarkable connection between the
discounted and average reward criteria.

\begin{definition}[Blackwell Optimality]
\label{def:blackwell}
A deterministic stationary policy $d^*$ is \emph{Blackwell optimal}
if there exists $\discount_0 \in (0,1)$ such that $d^*$ is
$\discount$-discount optimal for all
$\discount \in (\discount_0, 1)$. That is,
$V^{d^*}_\discount(s) = \Vstar_\discount(s)$ for all $s$ and all
$\discount > \discount_0$.
\end{definition}

\begin{theorem}[Blackwell's Theorem]
\label{thm:blackwell}
For any finite MDP, a Blackwell optimal policy exists. Moreover,
every Blackwell optimal policy is average-reward optimal.
\end{theorem}

\begin{proof}
\textbf{Existence:} For each deterministic stationary policy $d$, the
discounted value $V^d_\discount(s)$ is a rational function of
$\discount$ (since $V^d_\discount = (I - \discount P^d)^{-1} r^d$
and each entry of the inverse is a ratio of polynomials in $\discount$
via Cramer's rule).

For two policies $d_1, d_2$, the difference
$V^{d_1}_\discount(s) - V^{d_2}_\discount(s)$ is a rational function
of $\discount$ and therefore has finitely many zeros in $(0,1)$
(unless identically zero). This means the set of $\discount$ values
where $d_1$ and $d_2$ swap in terms of which is better is finite.

Since there are finitely many policies ($A^S$ total), the union of
all such swap points is finite. Therefore, there exists
$\discount_0 < 1$ such that for all $\discount > \discount_0$, the
set of discount-optimal policies does not change. Any policy that is
optimal for all $\discount > \discount_0$ is Blackwell optimal.

\textbf{Average-reward optimality:} If $d^*$ is Blackwell optimal,
then $(1-\discount) V^{d^*}_\discount(s) =
(1-\discount) \Vstar_\discount(s)$ for $\discount > \discount_0$.
Taking $\discount \to 1$:
$\rho^{d^*} = \lim (1-\discount) V^{d^*}_\discount(s) =
\lim (1-\discount) \Vstar_\discount(s) = \rho^*$.
\end{proof}

\begin{keyresult}[title={Blackwell Optimality}]
A Blackwell optimal policy exists for every finite MDP. It is
simultaneously:
\begin{itemize}[nosep]
    \item $\discount$-discount optimal for all $\discount$
    sufficiently close to 1,
    \item average-reward optimal ($\rho^{d^*} = \rho^*$).
\end{itemize}
Blackwell optimality is the strongest notion of optimality for
infinite-horizon MDPs, subsuming both discounted and average reward
optimality.
\end{keyresult}

\begin{example}[Blackwell Optimality in a Two-Action MDP]
\label{ex:blackwell}
Consider $\States = \{1\}$ (single state), $\Actions = \{a, b\}$
with $R(1, a) = 2$, $R(1, b) = 1$. Since there is only one state,
both actions lead back to state 1. For any $\discount$, the optimal
policy is $\optpolicy(1) = a$ (the higher-reward action). This
policy is trivially Blackwell optimal with $\discount_0 = 0$: it is
optimal for all discount factors.

For a more interesting example, consider $\States = \{1, 2\}$ with:
\begin{center}
\begin{tabular}{ccccc}
\toprule
$s$ & $a$ & $R(s,a)$ & $P(1 \mid s,a)$ & $P(2 \mid s,a)$ \\
\midrule
$1$ & $a$ & 0 & 0 & 1 \\
$1$ & $b$ & 1 & 1 & 0 \\
$2$ & $a$ & 3 & 1 & 0 \\
$2$ & $b$ & 3 & 1 & 0 \\
\bottomrule
\end{tabular}
\end{center}
Policy $d_1$: $(b, a)$ gives $P^{d_1} = \bigl(\begin{smallmatrix}
1 & 0 \\ 1 & 0 \end{smallmatrix}\bigr)$, $r^{d_1} = (1, 3)^\top$.
Stationary distribution: $\mu = (1, 0)$, gain $\rho^{d_1} = 1$.

Policy $d_2$: $(a, a)$ gives $P^{d_2} = \bigl(\begin{smallmatrix}
0 & 1 \\ 1 & 0 \end{smallmatrix}\bigr)$, $r^{d_2} = (0, 3)^\top$.
Stationary distribution: $\mu = (1/2, 1/2)$, gain
$\rho^{d_2} = 3/2$.

Policy $d_2$ has higher average reward. For the discounted case,
$V^{d_2}_\discount(1) = \frac{3\discount}{1-\discount^2} =
\frac{3\discount}{(1-\discount)(1+\discount)}$ and
$V^{d_1}_\discount(1) = \frac{1}{1-\discount}$. We have
$V^{d_2}_\discount(1) > V^{d_1}_\discount(1)$ iff
$\frac{3\discount}{1+\discount} > 1$ iff $\discount > 1/2$.
So $d_2$ is Blackwell optimal with $\discount_0 = 1/2$.
\end{example}

%% ============================================================
%% SECTION 10: EXERCISES
%% ============================================================

\section{Exercises}
\label{sec:exercises}

\begin{exercise}[Gain and Bias Computation] \textnormal{(\(*\))} \\
Consider a two-state MDP with $\States = \{1, 2\}$,
$\Actions = \{a, b\}$:
\begin{center}
\begin{tabular}{ccccc}
\toprule
$s$ & $a$ & $R(s,a)$ & $P(1 \mid s,a)$ & $P(2 \mid s,a)$ \\
\midrule
1 & $a$ & 4 & 0.5 & 0.5 \\
1 & $b$ & 2 & 0.8 & 0.2 \\
2 & $a$ & 1 & 0.3 & 0.7 \\
2 & $b$ & 3 & 0.6 & 0.4 \\
\bottomrule
\end{tabular}
\end{center}
\begin{enumerate}[label=(\alph*)]
    \item For the policy $d(1) = a$, $d(2) = b$, compute the
    stationary distribution $\mu^d$ and the gain $\rho^d$.
    \item Solve the gain-bias equations for $h^d$ (with
    normalization $h^d(2) = 0$).
\end{enumerate}
\end{exercise}

\begin{exercise}[Unichain Classification] \textnormal{(\(*\))} \\
For the MDP in Exercise~1, check whether each of the four
deterministic policies induces a unichain Markov chain. Is the MDP
unichain?
\end{exercise}

\begin{exercise}[Average Reward Optimality Equations]
\textnormal{(\(**\))} \\
For the MDP in Exercise~1:
\begin{enumerate}[label=(\alph*)]
    \item Write down the average reward Bellman optimality equations
    \eqref{eq:avg-bellman-opt}.
    \item Solve for $\rho^*$ and $h^*$ (with $h^*(2) = 0$) by
    trying each candidate optimal policy.
    \item Verify by computing the gain of all four deterministic
    policies.
\end{enumerate}
\end{exercise}

\begin{exercise}[Discount-to-Average Convergence]
\textnormal{(\(**\))} \\
For the MDP in Example~\ref{ex:avg-simple}:
\begin{enumerate}[label=(\alph*)]
    \item Compute $\Vstar_\discount$ for $\discount = 0.9$, $0.99$,
    $0.999$.
    \item Verify that $(1-\discount)\Vstar_\discount(s) \to
    \rho^* = 37/13$ for each $s$.
    \item Compute $\Vstar_\discount(1) - \Vstar_\discount(2)$ for
    each $\discount$ and verify convergence to
    $h^*(1) - h^*(2) = 40/13$.
\end{enumerate}
\end{exercise}

\begin{exercise}[Blackwell Optimality Threshold]
\textnormal{(\(***\))} \\
For the MDP in Example~\ref{ex:blackwell}:
\begin{enumerate}[label=(\alph*)]
    \item Compute $V^d_\discount(s)$ for all four deterministic
    policies and both states, as functions of $\discount$.
    \item Determine the Blackwell threshold $\discount_0$ (the
    largest $\discount$ at which the optimal policy changes).
    \item Verify that the Blackwell optimal policy is indeed
    average-reward optimal.
\end{enumerate}
\end{exercise}

\begin{exercise}[Multichain Counterexample]
\textnormal{(\(***\))} \\
Construct a three-state, two-action MDP that is multichain (at least
one deterministic policy has two recurrent classes) and show that the
average reward can depend on the initial state. Specifically, find
a policy $\policy$ and two states $s_1, s_2$ such that
$\rho^\policy(s_1) \neq \rho^\policy(s_2)$.

\textit{Hint:} Use absorbing states to create separate recurrent
classes with different reward rates.
\end{exercise}

%% ============================================================
%% SECTION 11: SUMMARY
%% ============================================================

\section{Summary}
\label{sec:summary}

\begin{itemize}
    \item The \emph{average reward criterion} evaluates policies by
    their long-run reward rate
    $\rho^\policy = \lim_{T \to \infty} \frac{1}{T}
    \E[\sum_{t=0}^{T-1} R(S_t, A_t)]$, eliminating the discount
    factor as a hyperparameter.
    \item For \emph{unichain} MDPs (every policy induces a chain
    with a single recurrent class), the gain $\rho^\policy$ is
    independent of the initial state and equals the expected reward
    under the stationary distribution:
    $\rho^\policy = (\mu^\policy)^\top r^\policy$.
    \item The \emph{gain-bias equations} (Poisson equations)
    $\rho^\policy + h^\policy(s) = \sum_a \policy(a \mid s)
    [R(s,a) + \sum_{s'} P(s' \mid s,a) h^\policy(s')]$ are the
    average-reward analogue of the discounted Bellman equations.
    The bias $h^\policy(s)$ measures the transient advantage of
    starting in state $s$.
    \item The bias is unique only up to an additive constant,
    reflecting that only \emph{relative} values matter in the
    average reward setting.
    \item The average reward Bellman optimality equation
    $\rho^* + h^*(s) = \max_a [R(s,a) + \sum_{s'} P(s' \mid s,a)
    h^*(s')]$ characterizes the optimal gain and bias.
    \item \emph{Relative value iteration} centers iterates at a
    reference state to prevent divergence, enabling iterative
    solution of the average reward optimality equations.
    \item The discounted and average reward criteria are connected:
    $(1-\discount)\Vstar_\discount \to \rho^*$ as
    $\discount \to 1$, and the centered discounted values converge
    to the bias.
    \item A \emph{Blackwell optimal} policy is simultaneously
    discount-optimal for all $\discount$ sufficiently close to 1
    and average-reward optimal. Blackwell optimal policies exist
    for all finite MDPs.
\end{itemize}

%% ============================================================
%% SECTION 12: FURTHER READING
%% ============================================================

\section{Further Reading}
\label{sec:further-reading}

\begin{itemize}
    \item \textbf{Lesson on TD Learning} extends the ideas from
    this lesson to the model-free setting, where the transition
    kernel is unknown and the agent learns from sampled trajectories.
    \item \textbf{Lesson on Stochastic Approximation} (Lesson~19)
    provides the convergence framework for R-learning and other
    average-reward RL algorithms.
    \item \citet[Chapters 8--10]{puterman2014} provides the
    definitive treatment of average reward MDPs, including multichain
    theory, sensitive-discount optimality, and Laurent series
    expansions.
    \item \citet{blackwell1962} is the original paper establishing
    Blackwell optimality; it remains concise and readable.
    \item \citet{howard1960} introduced the gain-bias framework and
    developed policy iteration for average reward, establishing the
    computational foundations.
    \item \citet[Chapter 7]{bertsekas2012} treats average cost
    problems with connections to stochastic shortest path formulations
    and semi-Markov decision processes.
    \item \citet{derman1970} provides an accessible treatment of
    average reward theory with applications to queuing and inventory
    problems.
\end{itemize}

\bibliography{mdp-refs}
\end{document}
