\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 65: Communication in Multi-Agent RL}}
\fancyhead[R]{\thepage}
\title{Lesson 65: Communication in Multi-Agent RL}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% ============================================================
%% SECTION 1: MOTIVATION AND OVERVIEW
%% ============================================================
\section{Motivation and Overview}
\label{sec:motivation}

Consider a team of autonomous search-and-rescue drones deployed
over a disaster zone.  Each drone has a limited field of view and
can only observe a small region of the terrain at any time step.
When one drone spots a survivor, it cannot carry out the rescue
alone---it needs teammates to converge on the location, perhaps
bringing medical supplies or forming a relay to a base station.
The challenge is immediate: the drone that detects the survivor
must somehow \emph{inform} its teammates, who are distributed
across the zone, each processing only their own local
observations.  But what should the message contain?  The raw
observation?  A compressed encoding?  A directive?  And to whom
should the message be sent---all drones (broadcast) or only the
nearest ones (targeted)?

This scenario exposes a fundamental tension in multi-agent
reinforcement learning: decentralized agents operating under
partial observability can, in principle, achieve near-centralized
performance---\emph{if} they communicate effectively.  The tools
developed in this lesson will let us formalize communication
channels within the Dec-POMDP framework, analyze when
communication provably improves team performance, and derive
end-to-end learning algorithms for communication protocols.

\begin{rlconnection}[title={Why Communication Matters for RL/ML}]
\begin{itemize}[nosep]
  \item \textbf{Robot swarm coordination.}  Swarms of robots
    performing exploration, construction, or foraging rely on
    learned communication to coordinate without centralized
    control.
  \item \textbf{Autonomous vehicle negotiation.}  Vehicles at
    intersections without traffic signals must exchange intentions
    to avoid collisions, requiring compact communication protocols.
  \item \textbf{Distributed sensor networks.}  Bandwidth-limited
    sensors must decide what information to transmit to a fusion
    center, balancing informativeness against channel cost.
  \item \textbf{Emergent language in AI.}  Studying how RL agents
    develop communication protocols from scratch connects to
    fundamental questions about the origins of language.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
Stochastic game formalism and Markov perfect equilibrium
(Lesson~61, Definitions~3.1--3.4).  Cooperative MARL, Dec-POMDP
framework, and centralized training with decentralized execution
(Lesson~62, Sections~3--5).  Policy gradient theorem and
REINFORCE (Lessons~54--55).  Neural network forward and backward
propagation (Phase~03, Lessons~59--66).

\paragraph{Learning objectives.}
After completing this lesson, you will be able to:
\begin{itemize}[nosep]
  \item \textbf{Define} a communication-augmented Dec-POMDP and
    \textbf{classify} communication protocols by channel type,
    topology, and learning method.
  \item \textbf{Derive} the CommNet gradient and \textbf{explain}
    why end-to-end backpropagation through continuous communication
    is well-defined.
  \item \textbf{Compare} DIAL and RIAL for discrete communication,
    identifying the variance trade-off between reparameterization
    and REINFORCE-style gradients.
  \item \textbf{Formalize} TarMAC's attention-based targeted
    communication mechanism and \textbf{prove} that it subsumes
    CommNet as a special case.
  \item \textbf{Analyze} conditions under which emergent
    communication arises and \textbf{evaluate} metrics for
    compositionality.
  \item \textbf{Justify} when communication improves team
    performance over non-communicative baselines using the value of
    communication framework.
\end{itemize}

%% ============================================================
%% SECTION 2: REFERENCES
%% ============================================================
\section{References}
\label{sec:references}

\begin{itemize}
  \item \citet{sukhbaatar2016} --- CommNet: continuous
    communication via mean-pooled hidden states with end-to-end
    backpropagation.
  \item \citet{foerster2016} --- DIAL and RIAL: differentiable
    and reinforced approaches to discrete inter-agent
    communication.
  \item \citet{das2019} --- TarMAC: targeted multi-agent
    communication using soft attention over agent messages.
  \item \citet{zhang2021} --- Comprehensive survey of multi-agent
    RL theory and algorithms, including communication methods.
  \item \citet{hernandezleal2019} --- Critical survey of deep
    MARL covering communication architectures and open problems.
\end{itemize}

%% ============================================================
%% SECTION 3: COMMUNICATION-AUGMENTED DEC-POMDP
%% ============================================================
\section{Communication-Augmented Dec-POMDP}
\label{sec:comm-dec-pomdp}

We begin by extending the standard Dec-POMDP framework of
Lesson~62 with explicit communication channels.  The key idea
is that agents can exchange messages before (or simultaneously
with) selecting actions, and these messages become part of the
information available for decision-making.

\begin{definition}[Message Space]
\label{def:message-space}
For a multi-agent system with agent set $\Players = \{1, \ldots,
N\}$, the \emph{message space} of agent $i$ is a set
$\mathcal{M}_i$.  A message $m_i \in \mathcal{M}_i$ is a signal
that agent $i$ can transmit at each time step.  The message may
be continuous ($\mathcal{M}_i \subseteq \R^d$) or discrete
($\mathcal{M}_i = \{1, \ldots, M\}$).
\end{definition}

\begin{definition}[Communication-Augmented Dec-POMDP]
\label{def:comm-dec-pomdp}
A \emph{communication-augmented Dec-POMDP} (Comm-Dec-POMDP) is a
tuple
\[
  \mathcal{G} = \bigl(\Players, \States, \{\Actions_i\}_{i \in
  \Players}, \Trans, \{r_i\}_{i \in \Players}, \{\Omega_i\}_{i
  \in \Players}, \{O_i\}_{i \in \Players}, \{\mathcal{M}_i\}_{i
  \in \Players}, \mathcal{C}, \discount \bigr),
\]
where $\States$ is the state space, $\Actions_i$ is agent $i$'s
action space, $\Trans$ is the transition kernel, $r_i$ is agent
$i$'s reward, $\Omega_i$ is agent $i$'s observation space,
$O_i : \States \to \simplex(\Omega_i)$ is the observation
function, $\mathcal{M}_i$ is the message space, $\mathcal{C}$
is the \emph{communication protocol}, and $\discount \in [0,1)$
is the discount factor.

The communication protocol $\mathcal{C}$ specifies:
\begin{enumerate}[nosep]
  \item \textbf{Timing:} whether messages are sent before or
    simultaneously with actions;
  \item \textbf{Topology:} a directed graph $G = (\Players, E)$
    where $(i,j) \in E$ means agent $i$ can send messages to
    agent $j$;
  \item \textbf{Bandwidth:} constraints on $|\mathcal{M}_i|$ or
    $\dim(\mathcal{M}_i)$.
\end{enumerate}

Agent $i$'s policy maps its local observation and received
messages to a joint action-message output:
\[
  \policyi : \Omega_i \times \prod_{j : (j,i) \in E}
  \mathcal{M}_j \;\longrightarrow\; \simplex(\Actions_i \times
  \mathcal{M}_i).
\]
\end{definition}

\begin{remark}[Reduction to Centralized POMDP]
\label{rem:centralized-reduction}
Consider a Comm-Dec-POMDP with broadcast topology ($E = \Players
\times \Players \setminus \{(i,i)\}$) and message spaces
satisfying $|\mathcal{M}_i| \geq |\Omega_i|$ for all $i$.  Then
each agent can transmit its full observation, and after one round
of communication every agent has access to the joint observation
$(o_1, \ldots, o_N)$.  The resulting information structure is
equivalent to a centralized POMDP.
\end{remark}

\begin{proposition}[Communication Reduces Complexity]
\label{prop:complexity-reduction}
The optimal finite-horizon Dec-POMDP problem is NEXP-complete
\citep{busoniu2008}.  If communication channels satisfy
$|\mathcal{M}_i| \geq |\Omega_i|$ for all $i \in \Players$ with
broadcast topology, then the Comm-Dec-POMDP reduces to a
centralized POMDP, which is PSPACE-complete.
\end{proposition}

\begin{proof}
Under broadcast communication with $|\mathcal{M}_i| \geq
|\Omega_i|$, agent $i$ can encode its observation $o_i$ into
a message $m_i = \text{enc}(o_i)$ with a bijective encoding.
After one communication round, every agent $j$ receives
$\{m_i\}_{i \neq j}$ and can decode to obtain $\{o_i\}_{i \neq
j}$.  Combined with its own observation $o_j$, agent $j$ has
access to the full joint observation $(o_1, \ldots, o_N)$.

When all agents observe the same information, the joint policy
optimization can be cast as a single-agent POMDP with state space
$\States$, action space $\Actions_1 \times \cdots \times
\Actions_N$, and observation space $\Omega_1 \times \cdots \times
\Omega_N$.  The finite-horizon POMDP is PSPACE-complete (Papadimitriou
and Tsitsiklis, 1987), so the Comm-Dec-POMDP inherits this
complexity class under sufficient communication.
\end{proof}

The following taxonomy organizes the main axes along which
communication protocols differ.

\begin{definition}[Communication Taxonomy]
\label{def:comm-taxonomy}
Communication protocols in MARL are classified along three axes:
\begin{enumerate}[nosep]
  \item \textbf{Channel type:} \emph{Continuous}
    ($\mathcal{M}_i \subseteq \R^d$) vs.\ \emph{discrete}
    ($\mathcal{M}_i$ finite).
  \item \textbf{Topology:} \emph{Broadcast} (all-to-all) vs.\
    \emph{targeted} (agent-specific recipients) vs.\
    \emph{bandwidth-limited} (fixed number of recipients).
  \item \textbf{Learning method:} \emph{End-to-end
    differentiable} vs.\ \emph{reinforcement-based} vs.\
    \emph{pre-defined} (hand-crafted protocol).
\end{enumerate}
\end{definition}

\begin{rlconnection}[title={Bridging Centralized and Decentralized
    Performance}]
Communication is the mechanism that interpolates between two
extremes: fully decentralized execution (no communication, each
agent acts on local observations alone) and fully centralized
control (a single controller with access to all observations).
The communication bandwidth determines where in this spectrum
the system operates.  In practice, even a small amount of learned
communication can dramatically close the gap between decentralized
and centralized performance \citep{sukhbaatar2016}.
\end{rlconnection}

We now turn to the three major algorithmic paradigms for learning
communication protocols: continuous communication (Section~\ref{sec:commnet}),
discrete communication (Section~\ref{sec:dial-rial}), and
targeted communication (Section~\ref{sec:tarmac}).

%% ============================================================
%% SECTION 4: CONTINUOUS COMMUNICATION: COMMNET
%% ============================================================
\section{Continuous Communication: CommNet}
\label{sec:commnet}

CommNet \citep{sukhbaatar2016} is the canonical continuous
communication architecture.  All agents share a single neural
network, and messages are the hidden states of other agents,
averaged and fed back as input.

\subsection{Architecture}

Consider $N$ agents with shared parameters $\theta$.  At each
time step $t$, the CommNet computation proceeds in $K$
communication rounds.

\begin{definition}[CommNet]
\label{def:commnet}
The CommNet architecture with $K$ communication rounds is defined
by the following recurrence.  For each agent $i \in \Players$:
\begin{enumerate}[nosep]
  \item \textbf{Encoding:}
    $\mathbf{h}_i^{(0)} = f_{\mathrm{enc}}(o_i; \theta_{\mathrm{enc}})$,
    where $f_{\mathrm{enc}} : \Omega_i \to \R^d$.
  \item \textbf{Communication rounds} ($k = 1, \ldots, K$):
    \begin{align}
      \mathbf{c}_i^{(k)} &= \frac{1}{N-1} \sum_{j \neq i}
        \mathbf{h}_j^{(k-1)},
        \label{eq:commnet-msg} \\
      \mathbf{h}_i^{(k)} &= \sigma\!\bigl(W_1^{(k)}
        \mathbf{h}_i^{(k-1)} + W_2^{(k)}
        \mathbf{c}_i^{(k)}\bigr),
        \label{eq:commnet-update}
    \end{align}
    where $W_1^{(k)}, W_2^{(k)} \in \R^{d \times d}$ are
    learnable weight matrices and $\sigma$ is an element-wise
    nonlinearity (e.g., ReLU or tanh).
  \item \textbf{Decoding:}
    $\policyi(\cdot \mid o_i, \mathbf{c}_i) =
    \mathrm{softmax}\!\bigl(f_{\mathrm{dec}}(\mathbf{h}_i^{(K)};
    \theta_{\mathrm{dec}})\bigr)$.
\end{enumerate}
\end{definition}

The ``message'' in CommNet is the mean of other agents' hidden
states---a continuous vector in $\R^d$.  This design ensures
that the entire computation graph, from observations to actions,
is a composition of differentiable operations.

\subsection{End-to-End Differentiability}

\begin{theorem}[CommNet Gradient]
\label{thm:commnet-gradient}
Let $J(\theta) = \E\bigl[\sum_{t=0}^{T} \discount^t \sum_{i
\in \Players} r_i(s_t, \mathbf{a}_t)\bigr]$ be the team
objective under CommNet with parameters $\theta = \{
\theta_{\mathrm{enc}}, \{W_1^{(k)}, W_2^{(k)}\}_{k=1}^K,
\theta_{\mathrm{dec}}\}$.  Then $\nabla_\theta J(\theta)$ is
well-defined and can be computed by standard backpropagation
through the communication rounds.
\end{theorem}

\begin{proof}
We show that every operation in the CommNet computation graph is
differentiable with respect to $\theta$.

\textbf{Step 1: Encoding.}  The encoder $f_{\mathrm{enc}}$ is a
neural network with differentiable activation functions, so
$\partial \mathbf{h}_i^{(0)} / \partial \theta_{\mathrm{enc}}$
exists.

\textbf{Step 2: Communication rounds.}  Fix round $k$.  The
communication message \eqref{eq:commnet-msg} is a linear
combination (mean) of the hidden states $\{\mathbf{h}_j^{(k-1)}
\}_{j \neq i}$, which is differentiable.  The hidden state
update \eqref{eq:commnet-update} is a composition of an affine
map $(W_1^{(k)}, W_2^{(k)})$ and the nonlinearity $\sigma$.  If
$\sigma$ is differentiable almost everywhere (as with ReLU or
tanh), then $\mathbf{h}_i^{(k)}$ is differentiable with respect
to both $\mathbf{h}_i^{(k-1)}$ and the parameters $W_1^{(k)},
W_2^{(k)}$.

\textbf{Step 3: Induction.}  By induction on $k$, each
$\mathbf{h}_i^{(k)}$ is differentiable with respect to $\theta$.
The key observation is that the mean operation in
\eqref{eq:commnet-msg} couples all agents: $\partial
\mathbf{c}_i^{(k)} / \partial \mathbf{h}_j^{(k-1)} = \frac{1}{N-1}
I_d$ for $j \neq i$.  This coupling propagates gradients across
agents.

\textbf{Step 4: Decoding and objective.}  The decoder
$f_{\mathrm{dec}}$ and softmax are differentiable.  For
deterministic action selection, the gradient flows directly.
For stochastic policies, we apply the reparameterization trick
(for continuous actions) or the log-derivative trick combined
with the fact that the logits are differentiable in $\theta$.

The overall gradient $\nabla_\theta J$ is obtained by
backpropagating through the temporal sequence and through the
communication rounds at each time step.  Since the composition
of differentiable functions is differentiable, the gradient is
well-defined.
\end{proof}

\begin{keyresult}[title={End-to-End Differentiable Communication}]
CommNet's use of continuous messages (mean-pooled hidden states)
makes the entire multi-agent system differentiable with respect to
shared parameters.  Standard backpropagation computes the gradient
without requiring REINFORCE-style estimators for the communication
component, yielding low-variance gradient estimates.
\end{keyresult}

\subsection{Properties and Limitations}

\begin{example}[Three-Agent Hallway Navigation]
\label{ex:hallway}
Consider three agents in a hallway with three rooms.  Each agent
observes only its current room.  The team receives reward $+1$
when all agents are in different rooms at the same time step.
Without communication, agents must resort to randomized
strategies.

With CommNet ($K=1$), each agent broadcasts its encoded
observation.  The received message $\mathbf{c}_i^{(1)} =
\frac{1}{2}(\mathbf{h}_j^{(0)} + \mathbf{h}_k^{(0)})$ encodes
the average position of the other two agents.  This suffices for
each agent to infer a ``free'' room and navigate to it, achieving
near-optimal coordination.
\end{example}

\begin{remark}[Limitations of Mean-Pooled Communication]
\label{rem:commnet-limits}
CommNet has two principal limitations:
\begin{enumerate}[nosep]
  \item \textbf{No targeted messaging.}  Every agent receives
    the same aggregate message (the mean of all others' hidden
    states), so agent-specific information is lost.
  \item \textbf{Permutation symmetry.}  The mean operation is
    invariant to permutations of agents, which can be beneficial
    (scalability) but also limits expressiveness when agent
    identities matter.
\end{enumerate}
These limitations motivate the targeted communication mechanisms
of Section~\ref{sec:tarmac}.
\end{remark}

%% ============================================================
%% SECTION 5: DISCRETE COMMUNICATION: DIAL AND RIAL
%% ============================================================
\section{Discrete Communication: DIAL and RIAL}
\label{sec:dial-rial}

In many real-world deployments, communication channels carry
discrete signals: binary flags, categorical indices, or natural
language tokens.  The fundamental challenge is that discrete
message selection (e.g., $\argmax$) is not differentiable, which
breaks the end-to-end gradient path exploited by CommNet.
\citet{foerster2016} proposed two approaches: RIAL (reinforced)
and DIAL (differentiable).

\subsection{RIAL: Reinforced Inter-Agent Learning}

\begin{definition}[RIAL Policy]
\label{def:rial}
In RIAL, agent $i$'s policy produces both an environment action
and a discrete message:
\[
  \policyi(\actioni, m_i \mid o_i, \mathbf{m}_{-i}^{\mathrm{recv}})
  = \policyi^a(\actioni \mid o_i, \mathbf{m}_{-i}^{\mathrm{recv}})
  \;\cdot\; \policyi^m(m_i \mid o_i,
  \mathbf{m}_{-i}^{\mathrm{recv}}),
\]
where $m_i \in \{1, \ldots, M\}$ is a discrete message,
$\mathbf{m}_{-i}^{\mathrm{recv}}$ denotes the messages received
from other agents, and the action and message components may share
a common hidden representation.
\end{definition}

In RIAL, the message $m_i$ is treated as an additional action.
The gradient for the message component is estimated via the
policy gradient (REINFORCE):
\begin{equation}
  \nabla_\theta J^{\mathrm{RIAL}} = \E\Bigl[
    \sum_{i \in \Players} \nabla_\theta \log \policyi(\actioni,
    m_i \mid \cdot\,; \theta) \; \hat{A}_i \Bigr],
  \label{eq:rial-gradient}
\end{equation}
where $\hat{A}_i$ is an advantage estimate.  Crucially, there is
\emph{no gradient flowing through the communication channel}:
agent $j$ treats the received message $m_i$ as part of the
environment, not as a function of $\theta$.

\subsection{DIAL: Differentiable Inter-Agent Learning}

\begin{definition}[Discretize/Regularize Unit (DRU)]
\label{def:dru}
The DRU maps a continuous pre-message $\tilde{m} \in \R^d$ to a
channel-compatible signal:
\[
  \mathrm{DRU}_\tau(\tilde{m}) =
  \begin{cases}
    \sigma(\tilde{m} / \tau), & \text{during training
      (temperature } \tau > 0\text{)}, \\[3pt]
    \indicator[\tilde{m} > 0], & \text{during execution
      (hard discretization)},
  \end{cases}
\]
where $\sigma$ denotes the logistic sigmoid applied
element-wise, $\tau$ is a temperature parameter, and
$\indicator[\cdot > 0]$ is the element-wise indicator function.
\end{definition}

\begin{definition}[DIAL]
\label{def:dial}
In DIAL, agent $i$ produces a continuous pre-message
$\tilde{m}_i = g_i(o_i, \mathbf{m}_{-i}^{\mathrm{recv}};
\theta)$ via a neural network.  During training, the message
transmitted is $m_i = \mathrm{DRU}_\tau(\tilde{m}_i)$, which
is differentiable in $\theta$.  During execution, the message
is discretized: $m_i = \indicator[\tilde{m}_i > 0]$.
\end{definition}

The DIAL gradient for the communication component is:
\begin{equation}
  \frac{\partial J}{\partial \theta} \;\ni\;
  \frac{\partial J}{\partial a_j}
  \frac{\partial a_j}{\partial m_i}
  \frac{\partial m_i}{\partial \tilde{m}_i}
  \frac{\partial \tilde{m}_i}{\partial \theta},
  \label{eq:dial-gradient-chain}
\end{equation}
where $\partial m_i / \partial \tilde{m}_i = \sigma'(\tilde{m}_i
/ \tau) / \tau$ during training.  This is a
reparameterization-like gradient that passes through the
communication channel.

\subsection{Variance Comparison}

\begin{proposition}[DIAL Has Lower Variance Than RIAL]
\label{prop:dial-variance}
For the communication component of the gradient, DIAL provides
lower-variance estimates than RIAL.  Specifically, let
$g^{\mathrm{DIAL}}$ and $g^{\mathrm{RIAL}}$ denote the gradient
estimators for the message parameters.  Then
\[
  \Var[g^{\mathrm{DIAL}}] \leq \Var[g^{\mathrm{RIAL}}].
\]
\end{proposition}

\begin{proof}
The RIAL gradient estimator for the message component has the
REINFORCE form:
\[
  g^{\mathrm{RIAL}} = \nabla_\theta \log \policyi^m(m_i \mid
  \cdot\,; \theta) \;\cdot\; \hat{A}_i.
\]
This is the product of the score function and the advantage,
and its variance includes the term $\Var[\hat{A}_i] \cdot
\E[\|\nabla_\theta \log \policyi^m\|^2]$ plus cross-terms.

The DIAL gradient estimator uses the pathwise derivative:
\[
  g^{\mathrm{DIAL}} = \frac{\partial J}{\partial m_i}
  \cdot \frac{\partial \mathrm{DRU}_\tau(\tilde{m}_i)}
  {\partial \tilde{m}_i} \cdot \frac{\partial \tilde{m}_i}
  {\partial \theta}.
\]
Each factor is computed by backpropagation through deterministic
operations (given the sampled trajectory).  The variance arises
only from the stochasticity of the trajectory, not from a
log-probability score function.

By the general result that pathwise (reparameterization)
estimators have lower variance than score-function (REINFORCE)
estimators when both are applicable \citep{suttonbarto2018}, we
conclude $\Var[g^{\mathrm{DIAL}}] \leq \Var[g^{\mathrm{RIAL}}]$.
\end{proof}

\subsection{Comparison Summary}

\begin{center}
\begin{tabular}{lcccc}
\toprule
& \textbf{Gradient type} & \textbf{Variance} &
\textbf{Channel} & \textbf{Execution} \\
\midrule
RIAL & REINFORCE (score) & High & Discrete &
  Discrete \\
DIAL & Pathwise (reparam.) & Low & Continuous
  (train) & Discrete \\
\bottomrule
\end{tabular}
\end{center}

\begin{warning}[Training-Execution Gap in DIAL]
During training, DIAL uses the continuous relaxation
$\sigma(\tilde{m}/\tau)$, but during execution, messages are
discretized to $\indicator[\tilde{m} > 0]$.  If the temperature
$\tau$ is not annealed toward zero during training, the learned
policy may depend on continuous message values that are lost
upon discretization, causing performance degradation at test
time.  Careful scheduling of $\tau \to 0$ is essential.
\end{warning}

\begin{example}[Two-Agent Switch Riddle]
\label{ex:switch}
Two agents are placed in separate rooms, each containing a switch
(up or down).  Agent 1 observes the state of both switches but
can only flip switch 1.  Agent 2 can flip switch 2 but observes
nothing.  The team receives reward $+1$ if both switches end up
in the same position.

Without communication, agent 2 must guess.  With a 1-bit
communication channel ($m_1 \in \{0,1\}$), agent 1 can signal
the state of switch 2, and agent 2 can act optimally.  RIAL
learns this by reinforcing the correct message; DIAL learns it
by backpropagating the reward signal through the sigmoid-relaxed
channel.  In experiments \citep{foerster2016}, DIAL converges
faster due to its lower-variance gradient.
\end{example}

%% ============================================================
%% SECTION 6: TARGETED COMMUNICATION: TARMAC
%% ============================================================
\section{Targeted Communication: TarMAC}
\label{sec:tarmac}

CommNet broadcasts a uniform average to all agents, ignoring
the fact that different agents may need different information.
TarMAC \citep{das2019} introduces an attention mechanism that
allows agents to send targeted messages, weighted by relevance.

\subsection{Architecture}

\begin{definition}[TarMAC]
\label{def:tarmac}
In TarMAC, each agent $i$ produces three quantities from its
hidden state $\mathbf{h}_i$:
\begin{align}
  \mathbf{v}_i &= f_v(\mathbf{h}_i; \theta_v)
    \in \R^{d_v}, & &\text{(value/message)},
    \label{eq:tarmac-value} \\
  \mathbf{k}_i &= f_k(\mathbf{h}_i; \theta_k)
    \in \R^{d_k}, & &\text{(key/signature)},
    \label{eq:tarmac-key} \\
  \mathbf{q}_i &= f_q(\mathbf{h}_i; \theta_q)
    \in \R^{d_k}, & &\text{(query)}.
    \label{eq:tarmac-query}
\end{align}
The attention weight from agent $i$ to agent $j$ is:
\begin{equation}
  \alpha_{j \leftarrow i} = \frac{\exp(\mathbf{q}_j^\top
  \mathbf{k}_i / \sqrt{d_k})}{\sum_{l \neq j} \exp(
  \mathbf{q}_j^\top \mathbf{k}_l / \sqrt{d_k})},
  \label{eq:tarmac-attention}
\end{equation}
and the aggregated message received by agent $j$ is:
\begin{equation}
  \mathbf{c}_j = \sum_{i \neq j} \alpha_{j \leftarrow i}
  \;\mathbf{v}_i.
  \label{eq:tarmac-message}
\end{equation}
Agent $j$ then conditions its policy on
$(\mathbf{h}_j, \mathbf{c}_j)$.
\end{definition}

This is precisely the scaled dot-product attention mechanism
from transformer architectures, applied across agents rather
than across sequence positions.  Multi-head variants extend this
by computing $H$ independent attention heads and concatenating
the results.

\subsection{Subsumption of CommNet}

\begin{proposition}[TarMAC Subsumes CommNet]
\label{prop:tarmac-subsumes}
If the attention weights in TarMAC are uniform, i.e.,
$\alpha_{j \leftarrow i} = 1/(N-1)$ for all $i \neq j$, and
the value function is the identity $\mathbf{v}_i =
\mathbf{h}_i$, then TarMAC reduces to CommNet's mean
communication.
\end{proposition}

\begin{proof}
Under uniform attention, the aggregated message
\eqref{eq:tarmac-message} becomes
\[
  \mathbf{c}_j = \sum_{i \neq j} \frac{1}{N-1} \;\mathbf{v}_i
  = \frac{1}{N-1} \sum_{i \neq j} \mathbf{h}_i,
\]
where the last equality uses $\mathbf{v}_i = \mathbf{h}_i$.
This is exactly the CommNet communication
\eqref{eq:commnet-msg}.
\end{proof}

\begin{keyresult}[title={Attention-Based Targeted Communication}]
TarMAC's attention mechanism allows each agent to selectively
attend to the most relevant messages, generalizing CommNet's
uniform broadcast.  The attention weights are learned end-to-end,
enabling the system to discover which agents should communicate
with which.  As a strict generalization of CommNet, TarMAC
achieves at least the same representational power.
\end{keyresult}

\begin{rlconnection}[title={Attention and Scalability}]
The attention-based communication in TarMAC mirrors the
self-attention mechanism in transformers.  This connection
suggests that advances in efficient attention (sparse attention,
linear attention) can be directly applied to scale multi-agent
communication to large agent populations.  Furthermore,
attention weights provide interpretability: examining
$\alpha_{j \leftarrow i}$ reveals which agents are communicating
most, aiding in debugging and understanding emergent strategies.
\end{rlconnection}

\begin{example}[Predator-Prey with Targeted Communication]
\label{ex:predator-prey}
Consider four predators pursuing a single prey in a grid world.
Each predator observes a $5 \times 5$ patch centered on itself.
When a predator is near the prey but needs help to capture it
(capture requires two predators adjacent to the prey
simultaneously), it should communicate primarily with the
nearest other predator, not all of them.

With TarMAC, the predator near the prey learns to produce a
high-attention key $\mathbf{k}_i$ that matches the query
$\mathbf{q}_j$ of the nearest predator $j$.  The attention
weight $\alpha_{j \leftarrow i}$ becomes large, routing the
positional information to the most relevant teammate.
Distant predators receive low attention weight and are not
distracted by irrelevant messages.
\end{example}

%% ============================================================
%% SECTION 7: EMERGENT COMMUNICATION AND LANGUAGE GROUNDING
%% ============================================================
\section{Emergent Communication and Language Grounding}
\label{sec:emergent}

Sections~\ref{sec:commnet}--\ref{sec:tarmac} treated
communication as a mechanism for improving team performance.
A deeper question is whether agents can develop a structured
\emph{language} from scratch---a communication protocol with
compositional, interpretable structure.

\begin{definition}[Emergent Communication]
\label{def:emergent-comm}
\emph{Emergent communication} refers to a communication protocol
that arises purely from multi-agent RL training, without any
pre-defined language, message semantics, or explicit supervision
on the communication channel.  The protocol is shaped solely by
the reward signal.
\end{definition}

\subsection{Referential Games}

The standard testbed for emergent communication is the
\emph{referential game} \citep{lazaridou2017}.

\begin{definition}[Referential Game]
\label{def:referential-game}
A referential game consists of:
\begin{itemize}[nosep]
  \item A \emph{speaker} agent $S$ that observes a target object
    $x \in \mathcal{X}$ (e.g., an image);
  \item A message space $\mathcal{M}$ (typically a sequence of
    discrete symbols);
  \item A \emph{listener} agent $L$ that receives the message
    $m$ and must identify $x$ from a set of candidates
    $\{x_1, \ldots, x_C\}$;
  \item Reward: $r = \indicator[L \text{ selects the correct
    } x]$.
\end{itemize}
\end{definition}

\subsection{Compositionality}

\begin{definition}[Compositionality]
\label{def:compositionality}
An emergent language is \emph{compositional} if it exhibits
systematic structure: the meaning of a complex message is
determined by the meanings of its parts and the rules for
combining them.  Formally, let $\mathcal{X}$ be a structured
input space (e.g., objects described by attributes such as
color, shape, size) and let $\phi : \mathcal{X} \to
\mathcal{M}$ be the speaker's encoding.  The language is
compositional if there exists a decomposition
\[
  \phi(x) = (\phi_1(\text{attr}_1(x)), \;\phi_2(\text{attr}_2(x)),
  \;\ldots)
\]
where each $\phi_k$ maps the $k$-th attribute independently
to a message component.
\end{definition}

\begin{definition}[Topographic Similarity]
\label{def:topo-sim}
The \emph{topographic similarity} (topo-sim) of a communication
protocol $\phi$ is the Spearman correlation between distances
in meaning space and distances in message space:
\[
  \rho_{\mathrm{topo}} = \mathrm{Spearman}\!\bigl(\{d_{\mathcal{X}}
  (x, x')\}_{x \neq x'}, \;\{d_{\mathcal{M}}(\phi(x),
  \phi(x'))\}_{x \neq x'}\bigr),
\]
where $d_{\mathcal{X}}$ and $d_{\mathcal{M}}$ are distance
metrics on the input and message spaces, respectively.  A
topo-sim close to $+1$ indicates that semantically similar
inputs are mapped to similar messages (compositionality).
\end{definition}

\subsection{Conditions for Emergence}

\begin{theorem}[Conditions for Non-Trivial Communication]
\label{thm:emergence-conditions}
In a cooperative multi-agent setting with aligned incentives,
non-trivial emergent communication (i.e., communication that
improves team reward over non-communicative baselines) arises
when the following conditions hold:
\begin{enumerate}[nosep]
  \item \textbf{Partial observability:} no single agent observes
    the full state;
  \item \textbf{Complementary information:} different agents have
    access to different relevant features of the state;
  \item \textbf{Action interdependence:} the optimal action for
    at least one agent depends on information held by another;
  \item \textbf{Sufficient channel capacity:}
    $|\mathcal{M}_i|$ is large enough to encode the relevant
    information.
\end{enumerate}
\end{theorem}

\begin{proof}
We prove this by establishing necessity.

\textbf{Necessity of (1):} If all agents observe the full state
$s$, then each agent can independently compute the optimal action
$a_i^*(s)$ without any communication.  Hence $\text{VoC} = 0$
(see Definition~\ref{def:voc}).

\textbf{Necessity of (2):} If agents have identical observations
($o_i = o_j$ for all $i, j$), then no agent possesses
information that others lack.  Any message is redundant since the
receiver already has the same information as the sender.

\textbf{Necessity of (3):} If agent $i$'s optimal action does not
depend on the observations of any other agent---i.e., $a_i^*(o_i)
= \argmax_{a_i} \E[r_i \mid o_i, a_i]$ regardless of $a_{-i}$
and $o_{-i}$---then communication cannot improve $i$'s policy.
If this holds for all agents, $\text{VoC} = 0$.

\textbf{Necessity of (4):} If the channel capacity is zero
($|\mathcal{M}_i| = 1$), no information can be transmitted.
More generally, the channel must support encoding of the relevant
information with sufficient fidelity.

\textbf{Sufficiency:} When all four conditions hold, there
exists at least one state $s$ and pair of agents $(i,j)$ such
that agent $j$ can improve its action by receiving information
from agent $i$ that is not available in $o_j$.  A communication
protocol mapping $o_i$ to a message $m_i$ that encodes this
information yields strictly positive value of communication.
\end{proof}

\begin{warning}[Interpretability of Emergent Protocols]
Emergent communication protocols are typically not interpretable
by humans.  Agents may exploit artifacts of the channel (e.g.,
encoding information in the \emph{timing} or \emph{magnitude}
of messages) rather than developing a human-readable language.
Compositionality, when it emerges, often requires explicit
pressure such as limited channel bandwidth or structured input
distributions \citep{lazaridou2017}.
\end{warning}

\begin{rlconnection}[title={Emergent Communication and Language
    Evolution}]
The study of emergent communication in MARL connects to deep
questions in linguistics and cognitive science about how human
languages evolved.  Referential games serve as simplified models
of the naming game studied in evolutionary linguistics.
Computational experiments have shown that compositionality---a
hallmark of human language---can emerge under appropriate
pressures (limited bandwidth, generalization requirements),
bridging MARL and computational linguistics.
\end{rlconnection}

%% ============================================================
%% SECTION 8: THEORETICAL ANALYSIS OF COMMUNICATION BENEFITS
%% ============================================================
\section{Theoretical Analysis of Communication Benefits}
\label{sec:theory}

We now formalize the question: \emph{when does communication
provably help?}

\begin{definition}[Value of Communication]
\label{def:voc}
For a Dec-POMDP $\mathcal{G}$, let $V^*_{\mathrm{comm}}$ denote
the optimal team value achievable with a communication channel
$\mathcal{C}$, and let $V^*_{\mathrm{no\text{-}comm}}$ denote
the optimal team value without communication.  The \emph{value
of communication} is
\[
  \mathrm{VoC}(\mathcal{C}) = V^*_{\mathrm{comm}} -
  V^*_{\mathrm{no\text{-}comm}} \;\geq\; 0.
\]
\end{definition}

The non-negativity $\mathrm{VoC} \geq 0$ follows because any
non-communicative policy is a special case of a communicative
policy (with all messages set to a constant).

\begin{proposition}[Zero Value of Communication]
\label{prop:voc-zero}
$\mathrm{VoC}(\mathcal{C}) = 0$ if there exists, for each
agent $i$, a sufficient statistic $T_i(o_i)$ for the optimal
action $a_i^*$ that depends only on $o_i$.  Formally, if
\[
  \policyi^*(\actioni \mid o_i) = \policyi^*(\actioni \mid o_i,
  m_{-i}) \quad \text{for all } m_{-i} \in \prod_{j \neq i}
  \mathcal{M}_j,
\]
for every agent $i$, then communication has zero value.
\end{proposition}

\begin{proof}
If agent $i$'s optimal policy is invariant to received messages
for every $i$, then the optimal communicative policy can be
achieved by a non-communicative policy (simply ignore all
messages).  Hence $V^*_{\mathrm{comm}} =
V^*_{\mathrm{no\text{-}comm}}$.
\end{proof}

\begin{proposition}[Positive Value of Communication]
\label{prop:voc-positive}
$\mathrm{VoC}(\mathcal{C}) > 0$ if and only if the following
three conditions hold simultaneously:
\begin{enumerate}[nosep]
  \item \textbf{Partial observability:} there exists a state
    $s \in \States$ and agent $i$ such that $o_i$ does not
    determine $s$;
  \item \textbf{Complementary information:} there exist agents
    $i, j$ and states $s, s'$ such that $o_i(s) \neq o_i(s')$
    but $o_j(s) = o_j(s')$ (agent $i$ distinguishes states that
    agent $j$ cannot);
  \item \textbf{Action interdependence:} agent $j$'s optimal
    action differs between $s$ and $s'$:
    $a_j^*(s) \neq a_j^*(s')$.
\end{enumerate}
\end{proposition}

\begin{proof}
\textbf{(If):}  Under conditions (1)--(3), agent $j$ faces
states $s, s'$ that it cannot distinguish ($o_j(s) = o_j(s')$)
but that require different optimal actions.  Without
communication, $j$ must choose a single action for both states.
With communication, agent $i$ (who can distinguish $s$ from
$s'$) sends a message encoding this distinction.  Agent $j$
conditions on the message and selects the state-appropriate
action.  The expected reward strictly increases, so
$\mathrm{VoC} > 0$.

\textbf{(Only if):}  If (1) fails, all agents observe the full
state and communication is redundant.  If (2) fails, no agent
has information that another lacks.  If (3) fails, the optimal
action is the same in all states that only communication could
distinguish, so the additional information is useless.
\end{proof}

\begin{theorem}[Communication Complexity Bound]
\label{thm:comm-complexity}
Consider a cooperative Comm-Dec-POMDP with $N$ agents, finite
state space $|\States|$, and horizon $T$.  An $\epsilon$-optimal
joint policy (achieving value within $\epsilon$ of $V^*_{\mathrm{comm}}$)
can be achieved with at most
\[
  B = O\!\left(|\States| \log \frac{T}{\epsilon}\right)
\]
bits of communication per agent per time step, assuming broadcast
topology.
\end{theorem}

\begin{proof}[Proof sketch]
The key insight is that the relevant information for optimal
decision-making is the sufficient statistic for the joint
belief state.

\textbf{Step 1: Belief state representation.}  The joint belief
$b_t \in \simplex(\States)$ requires at most $|\States| - 1$
real numbers to specify.  Quantizing each coordinate to
$\epsilon/(T \cdot |\States|)$ precision (to ensure total error
over $T$ steps is bounded by $\epsilon$) requires
$\log(T \cdot |\States| / \epsilon)$ bits per coordinate.

\textbf{Step 2: Sufficient communication.}  If each agent
broadcasts its local belief update (contribution to the joint
belief), the team can reconstruct the joint belief.  Each
agent's contribution involves at most $|\States|$ coordinates,
each requiring $O(\log(T/\epsilon))$ bits.

\textbf{Step 3: Total.}  The total communication per agent per
step is $O(|\States| \log(T/\epsilon))$ bits.  With this
precision, the team can implement a policy that is
$\epsilon$-close to the optimal centralized policy.
\end{proof}

\begin{keyresult}[title={Conditions for Communication Value}]
The value of communication is positive if and only if agents
have complementary partial observations and interdependent
optimal actions.  The communication bandwidth required for
$\epsilon$-optimal performance scales as $O(|\States|
\log(T/\epsilon))$ bits per agent per step---logarithmic in the
horizon and the accuracy requirement, but linear in the state
space size.
\end{keyresult}

\begin{example}[VoC in the Switch Riddle]
\label{ex:voc-switch}
Return to the switch riddle of Example~\ref{ex:switch}.
The state space is $\States = \{(\text{up}, \text{up}),
(\text{up}, \text{down}), (\text{down}, \text{up}),
(\text{down}, \text{down})\}$.

\textbf{Without communication:}  Agent 2 observes nothing,
so its policy must be independent of the state.  The best
strategy achieves reward $1/2$ in expectation (guess randomly
whether to flip).

\textbf{With 1-bit communication:}  Agent 1 sends $m_1 =
\indicator[\text{switch 2 is up}]$.  Agent 2 conditions on
$m_1$ and acts optimally, achieving reward $1$.  Hence
$\mathrm{VoC} = 1 - 1/2 = 1/2 > 0$.

All three conditions of Proposition~\ref{prop:voc-positive}
hold: (1) agent 2 has no observation; (2) agent 1 observes
what agent 2 cannot; (3) agent 2's optimal action differs
based on the unobserved switch state.
\end{example}

\begin{example}[VoC = 0 in Independent Tasks]
\label{ex:voc-zero}
Consider $N$ agents each solving an independent bandit problem:
agent $i$ observes its own bandit state $o_i$ and selects arm
$\actioni$.  The team reward is $\sum_i r_i(o_i, \actioni)$
with no interaction.  Since each $r_i$ depends only on $(o_i,
\actioni)$, the optimal policy for agent $i$ is $a_i^*(o_i)$
regardless of other agents' observations.  Condition (3) of
Proposition~\ref{prop:voc-positive} fails, so $\mathrm{VoC} =
0$.
\end{example}

%% ============================================================
%% SECTION 9: SURVEY AND COMPARISON OF METHODS
%% ============================================================
\section{Survey and Comparison of Methods}
\label{sec:survey}

We conclude the technical development with a comparative overview
of the major communication methods in MARL.

\subsection{Summary Table}

\begin{center}
\small
\begin{tabular}{lccccc}
\toprule
\textbf{Method} & \textbf{Channel} & \textbf{Topology} &
\textbf{Gradient} & \textbf{Scalability} & \textbf{Key
Innovation} \\
\midrule
CommNet & Continuous & Broadcast & Backprop &
  Moderate & Mean-pooled hidden states \\
RIAL & Discrete & Broadcast & REINFORCE &
  Moderate & Message as RL action \\
DIAL & Discrete (exec.) & Broadcast & Backprop (train) &
  Moderate & DRU continuous relaxation \\
TarMAC & Continuous & Targeted & Backprop &
  Good & Attention over messages \\
IC3Net & Continuous & Gated & Backprop &
  Good & Gating when to communicate \\
ATOC & Continuous & Dynamic & Backprop &
  Good & Attention + dynamic groups \\
\bottomrule
\end{tabular}
\end{center}

\noindent
IC3Net \citep{singh2019} extends CommNet with a learned gating
mechanism that decides \emph{when} to communicate, reducing
unnecessary bandwidth usage.  ATOC \citep{jiang2018} uses an
attention mechanism to form dynamic communication groups based on
agent proximity and relevance.

\subsection{Method Selection Guidelines}

\begin{itemize}[nosep]
  \item \textbf{CommNet:}  Appropriate for simple cooperative
    tasks with a moderate number of agents ($N \leq 20$) where
    all agents benefit from a shared summary of the group state.
  \item \textbf{RIAL:}  When the execution channel is strictly
    discrete and training-execution consistency is paramount.
    Accepts higher gradient variance in exchange for no
    train-test gap.
  \item \textbf{DIAL:}  When discrete execution is required but
    sample efficiency is critical.  The continuous relaxation
    during training significantly accelerates learning.
  \item \textbf{TarMAC:}  Large agent populations ($N \geq 10$)
    where agents have heterogeneous roles and targeted
    communication matters.  The attention mechanism provides
    natural scalability and interpretability.
  \item \textbf{IC3Net:}  Mixed cooperative-competitive settings
    where agents should selectively withhold communication from
    opponents.
\end{itemize}

\subsection{Open Problems}

Several fundamental challenges remain:
\begin{enumerate}[nosep]
  \item \textbf{Scalability to thousands of agents.}  Current
    methods scale to tens or hundreds of agents.  Mean-field
    approximations (Lesson~64) may help, but integrating them
    with learned communication is largely unexplored.
  \item \textbf{Adversarial robustness.}  If agents can
    intercept or corrupt messages, communication protocols must
    be robust.  Connections to Byzantine fault tolerance and
    robust game theory are open.
  \item \textbf{Communication cost.}  Most methods assume free
    communication.  Incorporating explicit bandwidth costs into
    the objective function changes the optimization landscape
    significantly.
  \item \textbf{Interpretability.}  Developing communication
    protocols that are both performant and human-interpretable
    remains a grand challenge connecting MARL to natural language
    processing.
\end{enumerate}

%% ============================================================
%% SECTION 10: EXERCISES
%% ============================================================
\section{Exercises}
\label{sec:exercises}

\begin{exercise}[CommNet Mean Observation]
\textnormal{(\(*\))} \\
Consider a CommNet with $K = 1$ communication round and identity
encoder ($f_{\mathrm{enc}}(o_i) = o_i$ with observations
$o_i \in \R^d$).  Show that if $W_1^{(1)} = 0$ and
$W_2^{(1)} = I_d$ (identity), and $\sigma$ is the identity
function, then each agent's hidden state after one round equals
the mean of all \emph{other} agents' observations:
$\mathbf{h}_i^{(1)} = \frac{1}{N-1} \sum_{j \neq i} o_j$.
\end{exercise}

\begin{exercise}[RIAL Policy Gradient]
\textnormal{(\(*\))} \\
Consider two agents ($\Players = \{1, 2\}$) with binary
message space $\mathcal{M}_i = \{0, 1\}$ and binary action
space $\Actions_i = \{0, 1\}$.  Agent 1 observes $o_1$ and
sends message $m_1$.  Agent 2 observes $o_2$ and receives
$m_1$.  Write the RIAL policy gradient $\nabla_\theta J$ for
this game, expanding all terms in $\policyi(\actioni, m_i
\mid o_i, m_{-i}^{\mathrm{recv}})$ explicitly.  Identify which
terms contribute to learning the communication protocol.
\end{exercise}

\begin{exercise}[TarMAC Subsumes CommNet]
\textnormal{(\(**\))} \\
Provide a complete proof that TarMAC (Definition~\ref{def:tarmac})
reduces to CommNet (Definition~\ref{def:commnet}) when:
(a)~$\mathbf{v}_i = \mathbf{h}_i$ (identity value function),
(b)~$\mathbf{q}_j = \mathbf{q}$ for all $j$ (uniform query),
and (c)~$\mathbf{k}_i = \mathbf{k}$ for all $i$ (uniform key).
Verify that conditions (b) and (c) imply uniform attention
weights, and substitute into the TarMAC aggregation formula.
\end{exercise}

\begin{exercise}[DRU Gradient Analysis]
\textnormal{(\(**\))} \\
For the DRU (Definition~\ref{def:dru}) with temperature $\tau$,
compute:
\begin{enumerate}[nosep,label=(\alph*)]
  \item The gradient $\partial \mathrm{DRU}_\tau(\tilde{m}) /
    \partial \tilde{m}$ during training for scalar $\tilde{m}
    \in \R$.
  \item Show that as $\tau \to 0^+$, the DRU output approaches
    the hard threshold $\indicator[\tilde{m} > 0]$ and the
    gradient approaches a scaled Dirac delta at
    $\tilde{m} = 0$.
  \item Compare the variance of the DIAL gradient (using the
    DRU) with the REINFORCE gradient for a single binary
    message, using a concrete example with a known reward
    function.
\end{enumerate}
\textit{Hint:} For part (b), use the fact that
$\sigma(x/\tau) \to \indicator[x > 0]$ pointwise as
$\tau \to 0^+$, and $\sigma'(x/\tau)/\tau \to \delta(x)$.
\end{exercise}

\begin{exercise}[Constructing a Dec-POMDP with VoC = 0]
\textnormal{(\(***\))} \\
Construct a Dec-POMDP with two agents, partial observability
(neither agent observes the full state), and yet
$\mathrm{VoC} = 0$.  Specifically:
\begin{enumerate}[nosep,label=(\alph*)]
  \item Define a state space $\States$, observation functions
    $O_1, O_2$, and reward function such that each agent's
    observations are incomplete but its optimal action depends
    only on its own observation.
  \item Prove formally that $\mathrm{VoC} = 0$ by showing that
    condition (3) of Proposition~\ref{prop:voc-positive} fails.
  \item Explain intuitively why partial observability alone is
    not sufficient for communication to help.
\end{enumerate}
\end{exercise}

\begin{exercise}[Communication Protocol Design]
\textnormal{(\(***\))} \\
Consider three agents in a grid world.  Agent 1 observes the
target location $\ell \in \{1, \ldots, 9\}$ (one of 9 grid
cells).  Agents 2 and 3 observe nothing about the target.  All
agents can send messages from $\mathcal{M}_i = \{0, 1\}^b$
($b$-bit messages).
\begin{enumerate}[nosep,label=(\alph*)]
  \item What is the minimum value of $b$ such that agent 1 can
    communicate the target location exactly?  Justify using
    information-theoretic arguments.
  \item Design an explicit communication protocol (mapping from
    $\ell$ to messages) for the minimum $b$.
  \item Suppose communication is noisy: each bit is flipped
    independently with probability $p < 1/2$.  Compute the
    probability of correct target identification as a function
    of $b$ and $p$, and determine the minimum $b$ to achieve
    error probability at most $0.01$ for $p = 0.1$.
\end{enumerate}
\textit{Hint:} For (c), use repetition coding and apply
Hoeffding's inequality.
\end{exercise}

%% ============================================================
%% SECTION 11: SUMMARY
%% ============================================================
\section{Summary}
\label{sec:summary}

\begin{itemize}
  \item A \emph{communication-augmented Dec-POMDP} extends the
    standard Dec-POMDP with message spaces, a communication
    protocol, and policies that condition on received messages
    as well as local observations
    (Definition~\ref{def:comm-dec-pomdp}).
  \item With sufficient bandwidth and broadcast topology,
    communication reduces the Dec-POMDP from NEXP-complete to
    PSPACE-complete by enabling centralized information
    aggregation (Proposition~\ref{prop:complexity-reduction}).
  \item \textbf{CommNet} achieves end-to-end differentiable
    communication by passing the mean of agents' hidden states
    as a continuous message, enabling standard backpropagation
    through the communication channel
    (Theorem~\ref{thm:commnet-gradient}).
  \item \textbf{RIAL} treats discrete messages as reinforcement
    learning actions, using REINFORCE-style gradients; \textbf{DIAL}
    uses a continuous relaxation (DRU) during training for
    lower-variance gradients, discretizing only at execution
    (Proposition~\ref{prop:dial-variance}).
  \item \textbf{TarMAC} introduces attention-based targeted
    communication, allowing each agent to selectively attend to
    the most relevant messages.  TarMAC strictly generalizes
    CommNet (Proposition~\ref{prop:tarmac-subsumes}).
  \item \emph{Emergent communication} arises when agents develop
    protocols from scratch via RL training; compositionality can
    be measured by topographic similarity
    (Definition~\ref{def:topo-sim}).
  \item Non-trivial communication emerges under four necessary
    conditions: partial observability, complementary information,
    action interdependence, and sufficient channel capacity
    (Theorem~\ref{thm:emergence-conditions}).
  \item The \emph{value of communication} $\mathrm{VoC}$ is
    positive if and only if agents have complementary
    observations and interdependent optimal actions
    (Proposition~\ref{prop:voc-positive}); achieving
    $\epsilon$-optimal performance requires
    $O(|\States|\log(T/\epsilon))$ bits per agent per step
    (Theorem~\ref{thm:comm-complexity}).
  \item Key open problems include scalability to thousands of
    agents, adversarial robustness of communication, explicit
    communication cost modeling, and interpretability of learned
    protocols.
\end{itemize}

%% ============================================================
%% SECTION 12: FURTHER READING
%% ============================================================
\section{Further Reading}
\label{sec:further-reading}

\begin{itemize}
  \item \textbf{Phase 06 applications.}  Lesson~66 (continuous
    control) and Lesson~70 (multi-vehicle coordination) apply
    communication-based MARL to aerospace scenarios, including
    formation control with limited bandwidth and
    communication-aware collision avoidance.
  \item \textbf{Learning when to communicate.}
    \citet{singh2019} introduce IC3Net, which augments CommNet
    with a learned gating mechanism that decides at each time
    step whether an agent should communicate, significantly
    reducing bandwidth in cooperative-competitive settings.
  \item \textbf{Attentional communication.}  \citet{jiang2018}
    propose ATOC, which uses an attention mechanism to form
    dynamic communication groups, and demonstrate scalability
    advantages in large-scale cooperative navigation.
  \item \textbf{Bidirectionally-coordinated nets.}
    \citet{peng2017} develop BiCNet for StarCraft micromanagement,
    using bidirectional recurrent communication for real-time
    coordination among heterogeneous units.
  \item \textbf{Emergent language and NLP.}  \citet{lazaridou2017}
    study multi-agent cooperation and the emergence of natural
    language properties.  The connection between emergent
    communication and computational pragmatics is surveyed in
    the broader NLP literature on language grounding.
  \item \textbf{Comprehensive surveys.}  \citet{zhang2021}
    provide a theoretical treatment of MARL including
    communication, while \citet{hernandezleal2019} offer a
    critical perspective on deep MARL methods and their
    limitations.
  \item \textbf{Historical note.}  The study of communication in
    multi-agent systems predates deep RL, with roots in
    distributed AI and the speech act theory of Austin and Searle.
    \citet{shoham2008} cover the logical and game-theoretic
    foundations of multi-agent communication protocols.
\end{itemize}

\bibliography{communication-refs}

\end{document}
