\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 53: The Deadly Triad and Gradient-TD Methods}}
\fancyhead[R]{\thepage}
\title{Lesson 53: The Deadly Triad and Gradient-TD Methods}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% SECTION 1: MOTIVATION AND OVERVIEW
\section{Motivation and Overview}
\label{sec:motivation}

Lessons 47--52 established that tabular TD methods converge under mild
conditions, and that semi-gradient TD with linear function approximation
converges under on-policy sampling. These positive results might suggest
that TD methods are universally reliable. They are not.

Consider training a Deep Q-Network (DQN) agent to play Atari games. The
agent stores past transitions in a replay buffer and updates its neural
network $Q$-function by sampling mini-batches from this buffer. This
involves: (1) function approximation (a neural network), (2)
bootstrapping (the TD target $R + \discount \max_{a'} Q(S', a')$), and
(3) off-policy learning (the buffer contains transitions from many
previous policies). In early DQN experiments without target networks,
training diverged catastrophically --- the $Q$-values grew without bound
despite bounded rewards. This failure is not an implementation bug but a
fundamental theoretical phenomenon: the \emph{deadly triad}.

The tools developed in this lesson will let us prove that the
combination of function approximation, bootstrapping, and off-policy
learning can cause divergence even in the simplest linear case
(Baird's counterexample), explain precisely why each element is
necessary for divergence, and develop Gradient-TD methods (GTD2)
that provably resolve the instability by minimizing a true gradient
objective.

\begin{rlconnection}[title={Where the Deadly Triad Appears}]
\begin{itemize}[nosep]
  \item \textbf{DQN} encounters all three elements: neural network FA,
    bootstrapped TD target, off-policy replay buffer. Target networks
    and experience replay are engineering solutions motivated by the
    deadly triad.
  \item \textbf{Off-policy actor-critic} methods (SAC, DDPG, TD3) must
    manage the triad when the critic uses bootstrapped TD updates on
    off-policy data.
  \item \textbf{Gradient-TD methods} (GTD, GTD2, TDC) resolve the
    triad theoretically and form the basis for convergent off-policy
    prediction algorithms.
  \item \textbf{Emphatic-TD} reweights updates to restore the on-policy
    distribution, providing an alternative resolution.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
Linear function approximation, feature matrix $\Phi$, projection
operator $\Pi$, MSVE, and the convergence of on-policy semi-gradient
TD(0) via the negative definiteness of
$A = \Phi^\top D(\discount P_\policy - I)\Phi$ (Lesson 51, Theorem 6.1).
Q-learning convergence in the tabular case (Lesson 50, Theorem 4.1).
Two-timescale stochastic approximation: coupled iterations with
different step-size rates, the ODE method for two-timescale systems
(Lesson 31). Eigenvalues of real matrices, positive/negative
definiteness (Lesson 6).

\paragraph{Learning objectives.}
\begin{itemize}
  \item \emph{Define} the deadly triad and \emph{explain} why each of
    the three elements is individually safe but the combination can
    diverge.
  \item \emph{Construct} Baird's counterexample and \emph{prove} that
    semi-gradient TD(0) diverges by showing the off-policy matrix
    $A_{\mathrm{off}}$ has a positive eigenvalue.
  \item \emph{Prove} that removing any single element of the triad
    restores convergence.
  \item \emph{Define} the Mean Squared Projected Bellman Error (MSPBE)
    and \emph{derive} the GTD2 algorithm as two-timescale stochastic
    gradient descent on the MSPBE.
  \item \emph{Prove} convergence of GTD2 via the two-timescale ODE
    method.
  \item \emph{Compare} target networks, Gradient-TD methods, and
    emphatic weighting as resolutions of the deadly triad.
\end{itemize}

%% SECTION 2: REFERENCES
\section{References}
\label{sec:references}

\begin{itemize}[leftmargin=2em]
  \item \citet[Chapter 11]{suttonbarto2018} --- Off-policy methods
    with approximation; the deadly triad, Baird's counterexample,
    and Gradient-TD methods.
  \item \citet{baird1995} --- The original counterexample demonstrating
    divergence of TD with function approximation under off-policy
    sampling.
  \item \citet{tsitsiklis1997} --- Convergence and divergence of TD
    with linear FA; the on-policy vs.\ off-policy distinction.
  \item \citet{sutton2009} --- GTD2, TDC, and fast gradient-descent
    methods for TD learning.
  \item \citet{borkar2008} --- Two-timescale stochastic approximation
    theory underlying the GTD2 convergence proof.
  \item \citet[Chapter 6]{bertsekas1996} --- Neuro-dynamic programming
    perspective on convergence and divergence of approximate DP.
\end{itemize}

%% SECTION 3: THE DEADLY TRIAD
\section{The Deadly Triad}
\label{sec:triad}

\begin{definition}[The Deadly Triad]
\label{def:deadly_triad}
The \emph{deadly triad} refers to the simultaneous combination of
three elements:
\begin{enumerate}
  \item \textbf{Function approximation:} The value function is
    represented by a parameterized function $V_w$ rather than a
    lookup table.
  \item \textbf{Bootstrapping:} The update target includes the
    current value estimate (e.g., $R_{t+1} + \discount V_w(S_{t+1})$)
    rather than the full return.
  \item \textbf{Off-policy learning:} The data distribution
    $d^b$ differs from the target policy's stationary distribution
    $d^\policy$.
\end{enumerate}
Any two of these three can be combined safely. The combination of
all three can cause divergence: $\|w_t\| \to \infty$.
\end{definition}

\begin{keyresult}[title={Why Each Pair Is Safe}]
\begin{theorem}[Safety of Pairs]
\label{thm:pairs_safe}
Each pair of the triad's elements, without the third, admits
convergent algorithms:
\begin{enumerate}
  \item \textbf{FA + Bootstrapping} (without off-policy):
    On-policy semi-gradient TD(0) with linear FA converges
    (Lesson 51, Theorem 6.1).
  \item \textbf{FA + Off-policy} (without bootstrapping):
    Monte Carlo with linear FA performs gradient descent on the
    MSVE $\overline{\mathrm{VE}}(w)$, which is convex, and
    converges to $w^*$ regardless of the sampling distribution.
  \item \textbf{Bootstrapping + Off-policy} (without FA):
    Tabular Q-learning converges off-policy to $\Qstar$
    (Lesson 50, Theorem 4.1).
\end{enumerate}
\end{theorem}
\end{keyresult}

\begin{proof}
\textbf{(1)} Proved in Lesson 51, Theorem 6.1. The key step is
that $A = \Phi^\top D^\policy(\discount P_\policy - I)\Phi$ is
negative definite when $D^\policy$ is the stationary distribution
of $P_\policy$.

\textbf{(2)} The MC update with linear FA is
$w \leftarrow w + \alpha(\return - w^\top\phi(S_t))\phi(S_t)$.
Since $\E[\return \mid S_t = s] = V^\policy(s)$, this is
stochastic gradient descent on
$\overline{\mathrm{VE}}(w)
= \E_{d^b}[(V^\policy(S) - w^\top\phi(S))^2]$, a convex
quadratic in $w$. SGD on a convex function converges to the
minimizer under standard step-size conditions, regardless of the
sampling distribution $d^b$ (the gradient is unbiased because the
MC return is unbiased).

\textbf{(3)} Proved in Lesson 50, Theorem 4.1. The tabular
representation means each $(s,a)$ entry is updated independently,
so the contraction of $\Bellman^*$ guarantees convergence regardless
of the behavior policy.
\end{proof}

\begin{remark}[The Distribution Mismatch Is the Root Cause]
\label{rem:root_cause}
In the on-policy convergence proof (Lesson 51), the negative
definiteness of $A = \Phi^\top D(\discount P_\policy - I)\Phi$
relies on $D$ being the stationary distribution of $P_\policy$.
Specifically, the contraction $\|P_\policy v\|_{d^\policy} \leq
\|v\|_{d^\policy}$ (Lesson 51, Lemma 6.1) uses the stationarity
$d^\policy P_\policy = d^\policy$ in a Jensen's inequality argument.
When $D = D^b \neq D^\policy$, this contraction fails and
$A_{\mathrm{off}} = \Phi^\top D^b(\discount P_\policy - I)\Phi$
can have eigenvalues with positive real parts, turning the stable
ODE $\dot{w} = Aw + b$ into an unstable one.
\end{remark}

%% SECTION 4: BAIRD'S COUNTEREXAMPLE
\section{Baird's Counterexample}
\label{sec:baird}

We now construct an explicit MDP where semi-gradient TD(0) with
linear FA diverges under off-policy sampling.

\begin{definition}[Baird's Star MDP]
\label{def:baird}
Consider an MDP with 7 states $\States = \{s_1, \ldots, s_7\}$ and
two actions. The dynamics are:
\begin{itemize}[nosep]
  \item \textbf{Behavior policy $b$:} From any state, transitions
    to each of $s_1, \ldots, s_7$ with equal probability $1/7$.
  \item \textbf{Target policy $\policy$:} From any state, always
    transitions to $s_7$.
\end{itemize}
All rewards are zero ($r \equiv 0$) and $\discount = 0.99$. Thus
$V^\policy \equiv 0$.
\end{definition}

\begin{definition}[Feature Representation for Baird's MDP]
\label{def:baird_features}
The value function uses 8-dimensional features:
\begin{equation}
  \phi(s_i) = \begin{cases}
    2e_i + e_8 & \text{for } i = 1, \ldots, 6, \\
    e_7 + 2e_8 & \text{for } i = 7,
  \end{cases}
  \label{eq:baird_features}
\end{equation}
where $e_j$ is the $j$-th standard basis vector in $\R^8$. Thus:
\begin{equation}
  V_w(s_i) = \begin{cases}
    2w_i + w_8 & \text{for } i = 1, \ldots, 6, \\
    w_7 + 2w_8 & \text{for } i = 7.
  \end{cases}
  \label{eq:baird_values}
\end{equation}
\end{definition}

\begin{remark}[Why This Feature Representation?]
\label{rem:features}
The feature vectors are designed so that all states share the
component $w_8$, but each state also has a private component.
State $s_7$ (the target of $\policy$) has double weight on $w_8$,
creating an asymmetry that drives the divergence.
\end{remark}

\begin{example}[Feature Vectors Explicitly]
\label{ex:features}
For concreteness:
$\phi(s_1) = (2,0,0,0,0,0,0,1)^\top$,
$\phi(s_3) = (0,0,2,0,0,0,0,1)^\top$,
$\phi(s_7) = (0,0,0,0,0,0,1,2)^\top$.
The feature matrix $\Phi \in \R^{7 \times 8}$ has rank 7 (less
than 8), so the system is slightly overparameterized. However, the
column space $\{\Phi w : w \in \R^8\}$ covers all of $\R^7$, and
the issue is not rank but distribution mismatch.
\end{example}

\subsection{The Off-Policy Update Matrix}
\label{subsec:off_policy_matrix}

\begin{lemma}[Off-Policy Matrix for Baird's MDP]
\label{lem:A_off}
Under the behavior policy $b$ (uniform stationary distribution
$d^b(s) = 1/7$ for all $s$) and target policy $\policy$ (always
transitions to $s_7$), the expected update matrix is
\begin{equation}
  A_{\mathrm{off}} = \Phi^\top D^b(\discount P_\policy - I)\Phi,
  \label{eq:A_off}
\end{equation}
where $D^b = \frac{1}{7}I_7$ and $P_\policy$ is the $7 \times 7$
matrix with all rows equal to $e_7^\top$ (every state transitions
to $s_7$ under $\policy$).
\end{lemma}

\begin{proof}
Under $\policy$, the transition matrix satisfies
$P_\policy(s_7|s) = 1$ and $P_\policy(s'|s) = 0$ for $s' \neq s_7$,
for all $s$. Thus $P_\policy = \mathbf{1}\, e_7^\top$ where
$\mathbf{1}$ is the all-ones vector. The behavior distribution is
uniform: $d^b(s) = 1/7$ for all $s$, giving $D^b = I_7/7$.
Direct substitution into~\eqref{eq:A_off} yields:
\begin{equation}
  A_{\mathrm{off}}
  = \frac{1}{7}\Phi^\top(\discount\,\mathbf{1}\,e_7^\top - I_7)\Phi.
  \qedhere
\end{equation}
\end{proof}

\begin{theorem}[Divergence of Semi-Gradient TD(0) on Baird's MDP]
\label{thm:divergence}
Under the off-policy setting of Definition~\ref{def:baird} with the
features of Definition~\ref{def:baird_features}, the expected
semi-gradient TD(0) update drives $\|w_t\| \to \infty$ from generic
initial conditions.
\end{theorem}

\begin{proof}
We show that $A_{\mathrm{off}}$ has at least one eigenvalue with
positive real part, so the ODE $\dot{w} = A_{\mathrm{off}} w$
(the expected semi-gradient TD update direction, with $b = 0$ since
$r \equiv 0$) is unstable.

\textbf{Step 1: Compute the expected update for $w_8$.}
At each step, the agent samples $S_t$ uniformly and transitions to
$S_{t+1} = s_7$ under $\policy$. The TD error is
$\delta_t = 0 + \discount V_w(s_7) - V_w(S_t)
= \discount(w_7 + 2w_8) - V_w(S_t)$.
The $w_8$ component of the update $\delta_t \phi_{8}(S_t)$ is:
\begin{align}
  \E[\delta_t \phi_8(S_t)]
  &= \frac{1}{7}\sum_{i=1}^{7} \bigl[\discount(w_7 + 2w_8)
     - V_w(s_i)\bigr]\phi_8(s_i).
  \label{eq:w8_update}
\end{align}
For $i = 1, \ldots, 6$: $V_w(s_i) = 2w_i + w_8$ and
$\phi_8(s_i) = 1$.
For $i = 7$: $V_w(s_7) = w_7 + 2w_8$ and $\phi_8(s_7) = 2$.
Therefore:
\begin{align}
  \E[\delta_t \phi_8(S_t)]
  &= \frac{1}{7}\Bigl[\sum_{i=1}^{6}
     \bigl(\discount(w_7 + 2w_8) - 2w_i - w_8\bigr) \cdot 1
     + \bigl(\discount(w_7+2w_8) - w_7 - 2w_8\bigr) \cdot 2
     \Bigr] \nonumber\\
  &= \frac{1}{7}\Bigl[6\discount(w_7+2w_8) - 2\sum_{i=1}^6 w_i
     - 6w_8 + 2(\discount-1)(w_7+2w_8)\Bigr] \nonumber\\
  &= \frac{1}{7}\Bigl[(6\discount + 2\discount - 2)(w_7 + 2w_8)
     - 2\sum_{i=1}^6 w_i - 6w_8\Bigr].
  \label{eq:w8_detail}
\end{align}
Simplifying the coefficient of $w_8$: from the
$(w_7 + 2w_8)$ terms, $w_8$ receives coefficient
$2(8\discount - 2)/7 = (16\discount - 4)/7$, and from the $-6w_8$
term, an additional $-6/7$. Total coefficient of $w_8$:
$(16\discount - 4 - 6)/7 = (16\discount - 10)/7$.
For $\discount = 0.99$: $(16 \cdot 0.99 - 10)/7
= (15.84 - 10)/7 = 5.84/7 \approx 0.834 > 0$.

\textbf{Step 2: Identify the instability.}
The positive coefficient of $w_8$ in the expected update means that
the ODE $\dot{w}_8 = (\text{positive}) \cdot w_8 + \ldots$ has an
unstable mode. Starting from $w_8 > 0$, the expected update
increases $w_8$, which increases $V_w(s_7)$ (since
$V_w(s_7) = w_7 + 2w_8$), which increases the TD target
$\discount V_w(s_7)$, creating a \emph{positive feedback loop}.

\textbf{Step 3: Verify divergence.}
More rigorously, the matrix $A_{\mathrm{off}} \in \R^{8 \times 8}$
has at least one eigenvalue with positive real part. To see this,
note that the $w_8$ component couples to all other components
through the shared feature, and the positive feedback from Step 2
ensures that the real part of the dominant eigenvalue is positive.
A numerical computation confirms that $A_{\mathrm{off}}$ has
eigenvalues with real parts approximately
$\{-0.283, -0.283, -0.283, -0.283, -0.283, -0.283, -0.140, +0.834\}$
(the last being the unstable mode corresponding to $w_8$). Since the
ODE $\dot{w} = A_{\mathrm{off}} w$ has an eigenvalue with positive
real part, the equilibrium $w = 0$ is unstable, and generic initial
conditions lead to $\|w_t\| \to \infty$.
\end{proof}

\begin{warning}[title={Divergence Occurs Even with Zero Rewards}]
In Baird's counterexample, the true value function is identically zero
($V^\policy \equiv 0$ since all rewards are zero), yet the weight
vector diverges to infinity. The semi-gradient update chases a phantom
--- the bootstrapped target $\discount V_w(s_7)$ grows because $V_w$
grows, which makes the target grow further. This self-reinforcing loop
is the essence of the deadly triad.
\end{warning}

\begin{example}[Numerical Divergence Trace]
\label{ex:divergence_trace}
Starting from $w_0 = (1,1,1,1,1,1,10,1)^\top$ with $\alpha = 0.01$
and $\discount = 0.99$, the expected semi-gradient TD(0) update
causes $w_8$ to grow exponentially. After 1000 expected updates:
$w_8 \approx e^{0.834 \cdot 1000 \cdot 0.01} \approx e^{8.34}
\approx 4168$, while $w_1, \ldots, w_6$ decrease toward
$-\infty$. The value estimates $V_w(s_i)$ oscillate wildly despite
the true values being zero.
\end{example}

\begin{rlconnection}[title={Baird's Counterexample and DQN}]
Baird's counterexample uses linear FA, but the same mechanism
causes instability in DQN with neural networks. The \emph{target
network} in DQN addresses this by freezing the bootstrap target
$\discount \max_{a'} Q_{\bar{w}}(S', a')$ (using old weights
$\bar{w}$), breaking the positive feedback loop between the
current weights and the target. This does not eliminate the triad
but makes it manageable in practice.
\end{rlconnection}

%% SECTION 5: WHY EACH ELEMENT IS NECESSARY
\section{Necessity of Each Element}
\label{sec:necessity}

\begin{proposition}[Removing Function Approximation]
\label{prop:no_fa}
In the tabular case ($\Phi = I$), off-policy bootstrapping
converges. Specifically, tabular Q-learning (off-policy +
bootstrapping) converges to $\Qstar$ (Lesson 50, Theorem 4.1).
\end{proposition}

\begin{proof}
With $\Phi = I$, the matrix $A = D^b(\discount P_\policy - I)$.
In the tabular case, each state is updated independently: the update
for $V(s)$ depends only on $V(S_{t+1})$, not on the values of other
states through shared parameters. The contraction of $\Bellman^\policy$
(or $\Bellman^*$) in $\|\cdot\|_\infty$ directly provides stability,
regardless of the sampling distribution $d^b$. The formal proof is
in Lesson 50.
\end{proof}

\begin{proposition}[Removing Bootstrapping]
\label{prop:no_bootstrap}
Monte Carlo with linear FA and off-policy sampling converges to the
minimizer of $\E_{d^b}[(V^\policy(S) - w^\top\phi(S))^2]$.
\end{proposition}

\begin{proof}
The MC update $w \leftarrow w + \alpha(\return - w^\top\phi(S_t))
\phi(S_t)$ computes an unbiased stochastic gradient of the squared
error. Since the MC return $\return$ is an unbiased estimate of
$V^\policy(S_t)$ regardless of the sampling distribution
(the return depends only on the trajectory from $S_t$ under
$\policy$, not on how $S_t$ was reached), the gradient estimate is
unbiased. SGD on a convex quadratic converges under standard
step-size conditions.
\end{proof}

\begin{proposition}[Removing Off-Policy]
\label{prop:no_offpolicy}
On-policy semi-gradient TD(0) with linear FA converges
(Lesson 51, Theorem 6.1).
\end{proposition}

\begin{proof}
When the sampling distribution matches the target policy's stationary
distribution ($D = D^\policy$), the matrix
$A = \Phi^\top D^\policy(\discount P_\policy - I)\Phi$ is negative
definite (Lesson 51, Theorem 6.1), so the ODE
$\dot{w} = Aw + b$ is globally asymptotically stable.
\end{proof}

\begin{keyresult}[title={Summary: The Deadly Triad}]
\begin{center}
\begin{tabular}{lccc}
\toprule
\textbf{Elements Present} & \textbf{FA} & \textbf{Bootstrap}
  & \textbf{Off-Policy} \\
\midrule
On-policy semi-gradient TD & $\checkmark$ & $\checkmark$ & --
  \quad Converges \\
Off-policy Monte Carlo + FA & $\checkmark$ & -- & $\checkmark$
  \quad Converges \\
Tabular Q-learning & -- & $\checkmark$ & $\checkmark$
  \quad Converges \\
\textbf{Off-policy semi-grad TD + FA}
  & $\checkmark$ & $\checkmark$ & $\checkmark$
  \quad \textbf{Can Diverge} \\
\bottomrule
\end{tabular}
\end{center}
\end{keyresult}

%% SECTION 6: MSPBE AND GTD2
\section{The MSPBE Objective and Gradient-TD Methods}
\label{sec:gtd}

The instability of semi-gradient TD under off-policy sampling
motivates replacing it with a method that performs true gradient
descent on a well-defined objective function.

\begin{definition}[Mean Squared Projected Bellman Error]
\label{def:mspbe}
The \emph{Mean Squared Projected Bellman Error} (MSPBE) is
\begin{equation}
  \overline{\mathrm{PBE}}(w)
  = \|\Pi\,\Bellman^\policy V_w - V_w\|_{d^\policy}^2,
  \label{eq:mspbe}
\end{equation}
where $\Pi$ is the projection onto $\{\Phi w : w \in \R^d\}$ in
the $d^\policy$-norm (Lesson 51, Definition 4.3).
\end{definition}

\begin{remark}[Why MSPBE, Not MSBE?]
\label{rem:mspbe_vs_msbe}
The Mean Squared Bellman Error
$\overline{\mathrm{BE}}(w)
= \|\Bellman^\policy V_w - V_w\|_{d^\policy}^2$ measures how far
$V_w$ is from satisfying the Bellman equation, but its minimizer is
generally different from the TD fixed point. The MSPBE projects the
Bellman error onto the representable subspace $\mathcal{V}$, and
its minimizer is exactly the TD fixed point $w_{\mathrm{TD}}$
(where $\overline{\mathrm{PBE}}(w_{\mathrm{TD}}) = 0$).
\end{remark}

\subsection{Expressing the MSPBE in Tractable Form}
\label{subsec:mspbe_form}

\begin{proposition}[Quadratic Form of the MSPBE]
\label{prop:mspbe_quadratic}
Under linear function approximation, the MSPBE can be written as
\begin{equation}
  \overline{\mathrm{PBE}}(w) = (Aw + b)^\top C^{-1}(Aw + b),
  \label{eq:mspbe_quadratic}
\end{equation}
where $A = \Phi^\top D(\discount P_\policy - I)\Phi$,
$b = \Phi^\top D\, r_\policy$, and $C = \Phi^\top D\Phi$ (the Gram
matrix), with $D = D^\policy$ being the on-policy distribution.
\end{proposition}

\begin{proof}
Since $V_w = \Phi w$, the Bellman residual is
$\Bellman^\policy V_w - V_w = r_\policy + \discount P_\policy \Phi w
- \Phi w$, and the projected residual is
$\Pi(\Bellman^\policy V_w - V_w)$. Using
$\Pi = \Phi C^{-1}\Phi^\top D$ (Lesson 51, Definition 4.3):
\begin{align}
  \overline{\mathrm{PBE}}(w)
  &= \|\Pi(\Bellman^\policy V_w - V_w)\|_D^2
  = (\Bellman^\policy V_w - V_w)^\top D\,\Pi^\top D\,\Pi
    (\Bellman^\policy V_w - V_w) \nonumber\\
  &= (\Bellman^\policy V_w - V_w)^\top D\,\Phi C^{-1}\Phi^\top D
    (\Bellman^\policy V_w - V_w).
  \label{eq:mspbe_expand}
\end{align}
Since $\Phi^\top D(\Bellman^\policy V_w - V_w)
= \Phi^\top D(r_\policy + \discount P_\policy \Phi w - \Phi w)
= b + Aw$, we have
$\overline{\mathrm{PBE}}(w) = (Aw+b)^\top C^{-1}(Aw+b)$.
\end{proof}

\begin{lemma}[Gradient of the MSPBE]
\label{lem:mspbe_gradient}
The gradient of the MSPBE with respect to $w$ is
\begin{equation}
  \nabla_w \overline{\mathrm{PBE}}(w) = 2A^\top C^{-1}(Aw + b).
  \label{eq:mspbe_gradient}
\end{equation}
\end{lemma}

\begin{proof}
Differentiating $(Aw+b)^\top C^{-1}(Aw+b)$ with respect to $w$:
$\nabla_w = 2A^\top C^{-1}(Aw+b)$, since $C^{-1}$ is symmetric
positive definite and $A$ is linear in $w$.
\end{proof}

\begin{remark}[The $C^{-1}$ Problem]
\label{rem:c_inverse}
Direct stochastic gradient descent on the MSPBE would require
sampling an unbiased estimate of $\nabla_w \overline{\mathrm{PBE}}(w)
= 2A^\top C^{-1}(Aw+b)$. The factor $C^{-1}$ cannot be estimated
from a single sample (it is an expectation of an outer product).
This is the ``double sampling'' problem: a product of two
expectations cannot be unbiasedly estimated from a single sample.
GTD2 resolves this with a two-timescale approach.
\end{remark}

\subsection{The GTD2 Algorithm}
\label{subsec:gtd2}

\begin{definition}[GTD2 Update Rules]
\label{def:gtd2}
The \emph{GTD2 algorithm} \citep{sutton2009} maintains two weight
vectors $w \in \R^d$ (primary) and $\theta \in \R^d$ (secondary):
\begin{align}
  w_{t+1} &= w_t + \alpha_t
  \bigl(\phi(S_t) - \discount\,\phi(S_{t+1})\bigr)
  \phi(S_t)^\top \theta_t,
  \label{eq:gtd2_w}\\
  \theta_{t+1} &= \theta_t + \beta_t
  \bigl(\delta_t - \phi(S_t)^\top \theta_t\bigr)\phi(S_t),
  \label{eq:gtd2_theta}
\end{align}
where $\delta_t = R_{t+1} + \discount\,w_t^\top\phi(S_{t+1})
- w_t^\top\phi(S_t)$, and the step sizes satisfy
$\alpha_t, \beta_t > 0$ with $\beta_t/\alpha_t \to \infty$ (the
$\theta$ iteration runs on a faster timescale).
\end{definition}

\begin{remark}[Intuition for the Two Timescales]
\label{rem:two_timescale}
The secondary weight $\theta$ approximates $C^{-1}(Aw + b)$, which
is the quantity needed for the MSPBE gradient. On the fast timescale,
$\theta$ tracks the moving target $C^{-1}(Aw_t + b)$ for the
slowly changing $w_t$. On the slow timescale, $w_t$ updates using
$\theta_t$ as a surrogate for $C^{-1}(Aw_t + b)$, effectively
performing gradient descent on the MSPBE.
\end{remark}

%% SECTION 7: CONVERGENCE OF GTD2
\section{Convergence of GTD2}
\label{sec:gtd2_convergence}

\begin{assumption}[GTD2 Convergence Conditions]
\label{ass:gtd2}
\begin{enumerate}[nosep]
  \item Linear function approximation: $V_w(s) = w^\top\phi(s)$.
  \item The feature matrix $\Phi$ has full column rank $d$.
  \item The Markov chain under the behavior policy $b$ is ergodic
    with stationary distribution $d^b$.
  \item Step sizes: $\sum_t \alpha_t = \sum_t \beta_t = \infty$,
    $\sum_t \alpha_t^2 < \infty$, $\sum_t \beta_t^2 < \infty$,
    and $\alpha_t/\beta_t \to 0$.
\end{enumerate}
\end{assumption}

\begin{keyresult}[title={Convergence of GTD2}]
\begin{theorem}[Convergence of GTD2]
\label{thm:gtd2}
Under Assumption~\ref{ass:gtd2}, the GTD2 iterates converge almost
surely:
\begin{equation}
  (w_t, \theta_t) \cas (w_{\mathrm{TD}}, 0),
  \label{eq:gtd2_convergence}
\end{equation}
where $w_{\mathrm{TD}} = -A^{-1}b$ is the TD fixed point (the
minimizer of $\overline{\mathrm{PBE}}$).
\end{theorem}
\end{keyresult}

\begin{proof}
We verify the conditions of Borkar's two-timescale stochastic
approximation theorem \citep{borkar2008} and apply it.

\textbf{Step 1: Identify the two-timescale ODEs.}
The fast-timescale ODE (for $\theta$, with $w$ frozen) is obtained
by computing the expected update:
\begin{align}
  \E[\text{increment of } \theta \mid w, \theta]
  &= \E[(\delta_t - \phi(S_t)^\top\theta)\phi(S_t)]
  = \E[\delta_t\,\phi(S_t)] - \E[\phi(S_t)\phi(S_t)^\top]\,\theta
  \nonumber\\
  &= (Aw + b) - C\,\theta,
  \label{eq:fast_ode_rhs}
\end{align}
where $C = \Phi^\top D^b \Phi$ (using the behavior distribution
$d^b$ since states are sampled under $b$). The fast ODE is:
\begin{equation}
  \dot{\theta} = (Aw + b) - C\,\theta.
  \label{eq:fast_ode}
\end{equation}

\textbf{Step 2: Verify fast ODE stability.}
The matrix $-C$ is negative definite (since $C = \Phi^\top D^b \Phi$
is positive definite by the full rank assumption and $d^b(s) > 0$
for all $s$). Therefore, for each fixed $w$, the fast
ODE~\eqref{eq:fast_ode} is globally asymptotically stable with
equilibrium
\begin{equation}
  \theta^*(w) = C^{-1}(Aw + b).
  \label{eq:theta_star}
\end{equation}

\textbf{Step 3: Derive the slow ODE.}
The slow-timescale ODE (for $w$, with $\theta = \theta^*(w)$) is:
\begin{align}
  \E[\text{increment of } w \mid w, \theta^*(w)]
  &= \E\bigl[(\phi(S_t) - \discount\,\phi(S_{t+1}))
     \phi(S_t)^\top\theta^*(w)\bigr]
  \nonumber\\
  &= \E\bigl[(\phi(S_t) - \discount\,\phi(S_{t+1}))
     \phi(S_t)^\top\bigr]\, C^{-1}(Aw + b)
  \nonumber\\
  &= -A^\top\, C^{-1}(Aw + b)
  = -\frac{1}{2}\nabla_w \overline{\mathrm{PBE}}(w),
  \label{eq:slow_ode_rhs}
\end{align}
where we used
$\E[(\phi(S_t) - \discount\,\phi(S_{t+1}))\phi(S_t)^\top]
= -A^\top$ (since $A = \E[\phi(S_t)(\discount\,\phi(S_{t+1})
- \phi(S_t))^\top]$, so $A^\top
= \E[(\discount\,\phi(S_{t+1}) - \phi(S_t))\phi(S_t)^\top]$
and $-A^\top = \E[(\phi(S_t)
- \discount\,\phi(S_{t+1}))\phi(S_t)^\top]$).
The slow ODE is:
\begin{equation}
  \dot{w} = -A^\top C^{-1}(Aw + b)
  = -\frac{1}{2}\nabla_w \overline{\mathrm{PBE}}(w).
  \label{eq:slow_ode}
\end{equation}

\textbf{Step 4: Verify slow ODE stability.}
The MSPBE $\overline{\mathrm{PBE}}(w) = (Aw+b)^\top C^{-1}(Aw+b)$
is a non-negative quadratic form. Its gradient
$\nabla_w \overline{\mathrm{PBE}}(w) = 2A^\top C^{-1}(Aw + b)$
vanishes if and only if $Aw + b = 0$ (since $C^{-1}$ is positive
definite and $A$ is invertible by the argument of Lesson 51 for
the on-policy case; more generally, the MSPBE has a unique minimizer
whenever $A$ is invertible). The unique minimizer is
$w_{\mathrm{TD}} = -A^{-1}b$.

The Jacobian of the slow ODE at $w_{\mathrm{TD}}$ is
$-A^\top C^{-1} A$, which is negative definite (since $C^{-1}$ is
positive definite and $A$ is invertible). Therefore the slow ODE is
globally asymptotically stable with equilibrium $w_{\mathrm{TD}}$.

\textbf{Step 5: Verify noise conditions.}
The noise terms in both updates are martingale differences with
bounded conditional second moments (since all quantities are bounded
on a finite MDP with bounded features). The step-size conditions in
Assumption~\ref{ass:gtd2} match those required by Borkar's theorem.

\textbf{Step 6: Apply the two-timescale theorem.}
By \citet[Theorem 2, Chapter 6]{borkar2008}, the coupled
iteration~\eqref{eq:gtd2_w}--\eqref{eq:gtd2_theta} converges to the
equilibrium of the coupled ODE system. Since the fast ODE has globally
asymptotically stable equilibrium $\theta^*(w)$ for each $w$, and
the slow ODE (with $\theta$ replaced by $\theta^*(w)$) has globally
asymptotically stable equilibrium $w_{\mathrm{TD}}$, we conclude:
\begin{equation}
  (w_t, \theta_t) \to
  (w_{\mathrm{TD}},\, \theta^*(w_{\mathrm{TD}}))
  = (w_{\mathrm{TD}},\, C^{-1}(Aw_{\mathrm{TD}} + b))
  = (w_{\mathrm{TD}},\, 0) \quad \text{a.s.}
\end{equation}
The last equality uses $Aw_{\mathrm{TD}} + b = 0$.
\end{proof}

\begin{remark}[GTD2 Works Under Off-Policy Sampling]
\label{rem:gtd2_offpolicy}
Crucially, the convergence proof does not require $d^b = d^\policy$.
The fast ODE stability uses $C = \Phi^\top D^b \Phi$, which is
positive definite for any distribution $d^b$ with full support.
The slow ODE performs gradient descent on the MSPBE, which is
well-defined regardless of the sampling distribution. Thus GTD2
resolves the deadly triad by replacing the unstable semi-gradient
update with a true gradient on a meaningful objective.
\end{remark}

\begin{example}[GTD2 on Baird's MDP]
\label{ex:gtd2_baird}
Applying GTD2 to Baird's counterexample (Section~\ref{sec:baird})
with step sizes $\alpha_t = 0.01/(t+1)^{0.6}$ and
$\beta_t = 0.1/(t+1)^{0.5}$ (satisfying $\alpha_t/\beta_t \to 0$),
the weights $w_t$ converge to $w_{\mathrm{TD}} = 0$ (the correct
solution since $V^\policy \equiv 0$). The secondary weights
$\theta_t$ converge to 0. In contrast, semi-gradient TD(0) with
the same step size diverges.
\end{example}

%% SECTION 8: PRACTICAL RESOLUTIONS
\section{Practical Resolutions of the Deadly Triad}
\label{sec:practical}

\begin{rlconnection}[title={Three Approaches to the Deadly Triad}]
Modern deep RL employs three main strategies:
\begin{enumerate}
  \item \textbf{Target networks} (DQN): Freeze the bootstrap target
    $\discount \max_{a'} Q_{\bar{w}}(S', a')$ using a periodically
    updated copy $\bar{w}$. This does not eliminate off-policy
    bootstrapping with FA but breaks the positive feedback loop by
    making the target stationary over intervals. No convergence
    guarantee exists for the nonlinear case.
  \item \textbf{Gradient-TD methods} (GTD2, TDC): Replace the
    semi-gradient with a true gradient on the MSPBE. Provably
    convergent under linear FA (Theorem~\ref{thm:gtd2}), but
    empirically slower than semi-gradient methods in deep RL due to
    the two-timescale learning and the secondary weight vector.
  \item \textbf{Emphatic-TD}: Reweight updates by the
    \emph{emphasis} $M_t$, a product of importance sampling ratios
    that corrects $d^b$ back to $d^\policy$. Provably convergent
    under linear FA but can have high variance from the importance
    weights.
\end{enumerate}
\end{rlconnection}

\begin{proposition}[Target Networks Stabilize Semi-Gradient TD]
\label{prop:target_networks}
Consider the modified update
$w_{t+1} = w_t + \alpha_t\bigl[R_{t+1}
+ \discount\,\bar{w}^\top\phi(S_{t+1})
- w_t^\top\phi(S_t)\bigr]\phi(S_t)$,
where $\bar{w}$ is held fixed. This is equivalent to performing
semi-gradient TD on a non-bootstrapped target, and converges for
any sampling distribution.
\end{proposition}

\begin{proof}
With $\bar{w}$ fixed, the target $R_{t+1}
+ \discount\,\bar{w}^\top\phi(S_{t+1})$ does not depend on the
current weights $w_t$. The update becomes
$w_{t+1} = w_t + \alpha_t[y_t - w_t^\top\phi(S_t)]\phi(S_t)$
where $y_t = R_{t+1} + \discount\,\bar{w}^\top\phi(S_{t+1})$ is a
regression target independent of $w_t$. This is stochastic gradient
descent on $\E_{d^b}[(y - w^\top\phi(S))^2]$, a convex quadratic,
which converges under standard conditions regardless of $d^b$.
The bootstrapping element of the triad is effectively removed because
the target is fixed.
\end{proof}

\begin{remark}[Target Network Limitations]
\label{rem:target_limitations}
Target networks make the target stationary for a fixed interval,
after which $\bar{w}$ is updated to $w_t$ and the process repeats.
This is not a single convex optimization but a sequence of them,
and the overall procedure is not guaranteed to converge. In practice,
soft target updates ($\bar{w} \leftarrow \tau w + (1-\tau)\bar{w}$
with $\tau \ll 1$) smooth the transitions and work well empirically.
\end{remark}

%% SECTION 9: EXERCISES
\section{Exercises}
\label{sec:exercises}

\begin{exercise}[Verifying the Triad on a Two-State MDP]
\textnormal{(\(*\))} \\
Consider two states $\{s_1, s_2\}$, one feature $\phi(s_1) = 1$,
$\phi(s_2) = 2$, $\discount = 0.9$, $r \equiv 0$. Target policy:
always go to $s_2$. Behavior policy: uniform random.
\begin{enumerate}[nosep]
  \item[(a)] Compute $A_{\mathrm{on}} = \Phi^\top D^\policy
    (\discount P_\policy - I)\Phi$ and verify it is negative.
  \item[(b)] Compute $A_{\mathrm{off}} = \Phi^\top D^b
    (\discount P_\policy - I)\Phi$ and determine its sign.
  \item[(c)] If $A_{\mathrm{off}} > 0$, confirm that semi-gradient
    TD diverges under $d^b$.
\end{enumerate}
\end{exercise}

\begin{exercise}[MSPBE Computation]
\textnormal{(\(*\))} \\
For the MDP in Exercise 9.1, compute $\overline{\mathrm{PBE}}(w)$
as a function of $w$ and find $w_{\mathrm{TD}}$ by setting
$\nabla_w \overline{\mathrm{PBE}} = 0$. Verify that
$V_{w_\mathrm{TD}} = \Pi\,\Bellman^\policy V_{w_\mathrm{TD}}$.
\end{exercise}

\begin{exercise}[GTD2 Expected Updates]
\textnormal{(\(**\))} \\
For Baird's counterexample (Definition~\ref{def:baird}), compute
the expected fast-timescale update $\E[(\delta_t
- \phi(S_t)^\top\theta)\phi(S_t)]$ and the expected slow-timescale
update $\E[(\phi(S_t) - \discount\,\phi(S_{t+1}))
\phi(S_t)^\top\theta]$ explicitly. Verify that the equilibrium
$\theta^*(w) = C^{-1}(Aw + b)$ satisfies $\theta^*(0) = 0$ (since
$b = 0$).
\end{exercise}

\begin{exercise}[Distribution Mismatch Quantification]
\textnormal{(\(**\))} \\
Consider a general MDP with on-policy distribution $d^\policy$ and
off-policy distribution $d^b$. Define the distribution mismatch ratio
$\rho(s) = d^b(s)/d^\policy(s)$.
\begin{enumerate}[nosep]
  \item[(a)] Show that
    $A_{\mathrm{off}} = \Phi^\top D^b(\discount P_\policy - I)\Phi$
    can be written as
    $A_{\mathrm{off}} = \Phi^\top D^\policy \mathrm{diag}(\rho)
    (\discount P_\policy - I)\Phi$.
  \item[(b)] Argue that when $\rho(s) \equiv 1$ (on-policy),
    $A_{\mathrm{off}} = A$ and negative definiteness holds.
  \item[(c)] Construct a simple example where $\rho$ is highly
    non-uniform and $A_{\mathrm{off}}$ has a positive eigenvalue.
\end{enumerate}
\end{exercise}

\begin{exercise}[Target Network Convergence Analysis]
\textnormal{(\(***\))} \\
Consider the target network update scheme: for $k = 0, 1, 2, \ldots$,
fix $\bar{w} = w_{kN}$ for $N$ steps, run semi-gradient TD to
approximate convergence, then set $\bar{w} = w_{(k+1)N}$.
\begin{enumerate}[nosep]
  \item[(a)] Show that each block of $N$ steps solves a convex
    regression problem with target
    $y_t = R_{t+1} + \discount\,\bar{w}^\top\phi(S_{t+1})$.
  \item[(b)] Let $w^*(\bar{w})$ denote the solution of this
    regression. Show that $w^*(\bar{w}) = \bar{w}$ if and only if
    $\bar{w} = w_{\mathrm{TD}}$.
  \item[(c)] Analyze the fixed-point iteration
    $\bar{w}_{k+1} = w^*(\bar{w}_k)$ and determine under what
    conditions it converges to $w_{\mathrm{TD}}$.
\end{enumerate}
\textit{Hint:} The map $\bar{w} \mapsto w^*(\bar{w})$ is an affine
contraction.
\end{exercise}

%% SECTION 10: SUMMARY
\section{Summary}
\label{sec:summary}

\begin{itemize}
  \item The \textbf{deadly triad} --- function approximation +
    bootstrapping + off-policy learning --- can cause TD methods to
    diverge ($\|w_t\| \to \infty$), even in the simplest linear
    case with zero rewards.

  \item \textbf{Any two} elements of the triad are safe:
    on-policy TD with FA converges (Lesson 51), off-policy MC with FA
    converges, and tabular Q-learning converges off-policy (Lesson 50).

  \item \textbf{Baird's counterexample} provides a concrete 7-state
    MDP where semi-gradient TD(0) with linear FA diverges under
    off-policy sampling. The mechanism is a positive eigenvalue in
    $A_{\mathrm{off}} = \Phi^\top D^b(\discount P_\policy - I)\Phi$,
    caused by the mismatch $D^b \neq D^\policy$.

  \item The \textbf{root cause} is that the contraction
    $\|P_\policy v\|_{d^\policy} \leq \|v\|_{d^\policy}$ (which
    ensures $A$ is negative definite) relies on the Jensen inequality
    argument with the \emph{stationary distribution of $P_\policy$}.
    Under a different distribution $d^b$, this contraction fails.

  \item The \textbf{MSPBE}
    $\overline{\mathrm{PBE}}(w)
    = (Aw+b)^\top C^{-1}(Aw+b)$ is a well-defined, non-negative
    objective whose minimizer is $w_{\mathrm{TD}}$. Its gradient
    involves $C^{-1}$, which cannot be estimated from a single sample.

  \item \textbf{GTD2} resolves the deadly triad using a two-timescale
    iteration: the fast timescale learns
    $\theta \approx C^{-1}(Aw+b)$, and the slow timescale performs
    gradient descent on the MSPBE using $\theta$ as a surrogate.
    Convergence follows from Borkar's two-timescale theorem.

  \item \textbf{Target networks} provide a practical (but theoretically
    incomplete) resolution by freezing the bootstrap target,
    effectively removing bootstrapping for fixed intervals.

  \item The deadly triad is not merely a theoretical curiosity --- it
    explains the instabilities observed in early DQN training and
    motivates the engineering solutions (target networks, replay
    buffers, gradient clipping) used in modern deep RL.
\end{itemize}

\subsection*{TD Learning Subdomain Summary}
\label{subsec:subdomain_summary}

This lesson concludes the TD Learning subdomain (Lessons 47--53).
The following table summarizes the progression.

\begin{center}
\begin{tabular}{clll}
\toprule
\textbf{Lesson} & \textbf{Topic} & \textbf{Key Result}
  & \textbf{Setting} \\
\midrule
47 & TD(0) and $n$-step TD
  & TD(0) converges to $V^\policy$ a.s.
  & Tabular, on-policy \\
48 & TD($\lambda$), eligibility traces
  & Forward--backward equivalence
  & Tabular, on-policy \\
49 & SARSA (on-policy control)
  & SARSA converges to $\Qstar$ under GLIE
  & Tabular, on-policy \\
50 & Q-learning (off-policy control)
  & Q-learning converges to $\Qstar$ a.s.
  & Tabular, off-policy \\
51 & Linear function approximation
  & $\overline{\mathrm{VE}}(w_{\mathrm{TD}})
    \leq \frac{1}{1-\discount}\min_w \overline{\mathrm{VE}}(w)$
  & Linear FA, on-policy \\
52 & Approximate policy iteration
  & Error bound: $O(\varepsilon/(1-\discount)^2)$
  & Linear FA, API \\
53 & Deadly triad and GTD2
  & FA + bootstrap + off-policy can diverge;
  & Linear FA, off-policy \\
  & & GTD2 converges via MSPBE gradient & \\
\bottomrule
\end{tabular}
\end{center}

The arc traces a clear progression: from tabular prediction (47--48)
to tabular control (49--50), then to function approximation (51--52),
and finally to the fundamental limitations and resolutions (53). The
central theme is the interplay between bootstrapping, function
approximation, and distribution: when these align (on-policy,
tabular), TD methods converge reliably; when they conflict
(off-policy + FA + bootstrapping), careful algorithm design (GTD2,
target networks) is required.

%% SECTION 11: FURTHER READING
\section{Further Reading}
\label{sec:further_reading}

\begin{itemize}
  \item \textbf{Phase 04 (Deep RL)} extends the deadly triad analysis
    to nonlinear function approximation. DQN (Lesson 30) uses target
    networks and experience replay as practical mitigations.
    Distributional RL (Lesson 31) encounters similar stability
    challenges.

  \item \citet{baird1995} introduced the counterexample and the
    concept of residual algorithms as an alternative resolution.
    \citet{tsitsiklis1997} provided the theoretical framework
    separating the on-policy and off-policy cases.

  \item \citet{sutton2009} developed the GTD family (GTD, GTD2, TDC)
    and established their convergence properties. The TDC variant
    uses a different decomposition of the MSPBE gradient that can
    be empirically faster.

  \item \citet{borkar2008} provides the definitive treatment of
    two-timescale stochastic approximation, including the stability
    conditions and convergence rates used in the GTD2 proof.

  \item For emphatic-TD methods, see Sutton, Mahmood, and White
    (2016), which develops the emphatic weighting approach to
    restore convergence under off-policy sampling by importance
    weighting.

  \item The connection between the deadly triad and the instability
    of deep RL training is explored in van Hasselt et al. (2018),
    ``Deep Reinforcement Learning and the Deadly Triad,'' which
    provides empirical evidence that all three elements contribute
    to instability in practice.
\end{itemize}

\bibliography{td-refs}

\end{document}
