\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 51: Function Approximation in TD Learning}}
\fancyhead[R]{\thepage}
\title{Lesson 51: Function Approximation in TD Learning}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% SECTION 1: MOTIVATION AND OVERVIEW
\section{Motivation and Overview}
\label{sec:motivation}

The tabular TD methods of Lessons 47--50 store a separate value for every
state (or state--action pair). A gridworld with a million cells requires a
table of one million entries; a continuous-state system such as a robotic
arm with six joint angles has an uncountable state space where a table is
undefined. Even when the state space is finite, the number of states in
practical problems---board games ($\approx 10^{170}$ for Go), Atari
screens ($\approx 10^{30\,000}$)---far exceeds what any table can hold.
This is the \emph{curse of dimensionality}: the number of states grows
exponentially in the dimension of the state description, making tabular
methods infeasible.

Function approximation replaces the lookup table with a parameterized
function $V_w(s) = w^\top \phi(s)$ that maps states to values through a
low-dimensional weight vector $w \in \R^d$, where $d \ll |\States|$.
By adjusting $w$ from sampled transitions, the agent generalizes to
unvisited states, breaking the curse. The tools developed in this lesson
will let us prove that semi-gradient TD(0) with linear function
approximation converges to a weight vector whose error is at most
$1/(1-\discount)$ times the best possible linear approximation, and
extend this bound to TD($\lambda$).

\begin{rlconnection}[title={Where Linear Function Approximation Appears}]
\begin{itemize}[nosep]
  \item \textbf{Deep Q-Networks (DQN)} extend the linear framework here
    to nonlinear neural network approximation; the linear theory
    identifies the fundamental issues (semi-gradient, projection,
    distribution mismatch) that persist in the deep case.
  \item \textbf{Least-Squares TD (LSTD)} solves the same fixed-point
    equation derived here but in closed form using batch data, forming
    the evaluation step in Least-Squares Policy Iteration (Lesson 52).
  \item \textbf{Tile coding and radial basis functions} are specific
    feature constructions $\phi(s)$ that instantiate the linear
    framework for continuous-state MDPs.
  \item \textbf{Generalized Advantage Estimation (GAE)} in PPO/TRPO
    uses a learned value function baseline $V_w$; the approximation
    error bounds here quantify the quality of that baseline.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
TD(0) update rule, TD error $\delta_t = R_{t+1} + \discount V(S_{t+1})
- V(S_t)$, and convergence of tabular TD(0) via stochastic approximation
(Lesson 47, Theorem 4.1). Bellman operator $\Bellman^\policy$ and its
$\discount$-contraction property in $\|\cdot\|_\infty$ (Lesson 45).
Stochastic approximation: the ODE method with linear driving function
$h(w) = Aw + b$ where $A$ is Hurwitz (Lesson 31). Linear algebra:
projection onto a subspace in a weighted inner product, Cauchy--Schwarz
inequality, eigenvalues of symmetric matrices (Lesson 7). Stationary
distribution of an ergodic Markov chain (Lesson 9).

\paragraph{Learning objectives.}
\begin{itemize}
  \item \emph{Define} linear function approximation $V_w(s) = w^\top
    \phi(s)$, the feature matrix $\Phi$, and the Mean Squared Value
    Error $\overline{\mathrm{VE}}(w)$.
  \item \emph{Derive} the semi-gradient TD(0) update rule with linear
    function approximation and \emph{explain} why the method is called
    ``semi-gradient.''
  \item \emph{Prove} that the expected update matrix
    $A = \Phi^\top D(\discount P_\policy - I)\Phi$ is negative definite
    under the on-policy stationary distribution.
  \item \emph{Prove} convergence of semi-gradient TD(0) with linear FA
    and the error bound
    $\overline{\mathrm{VE}}(w_{\mathrm{TD}}) \leq
    \frac{1}{1-\discount}\min_w \overline{\mathrm{VE}}(w)$.
  \item \emph{State} the TD($\lambda$) error bound with the improved
    factor $(1-\discount\lambda)/(1-\discount)$.
  \item \emph{Compute} the projection operator $\Pi$ and the TD
    fixed-point equation $\Pi \Bellman^\policy V_w = V_w$.
\end{itemize}

%% SECTION 2: REFERENCES
\section{References}
\label{sec:references}

\begin{itemize}[leftmargin=2em]
  \item \citet[Chapters 9--10]{suttonbarto2018} --- On-policy prediction
    with approximation; semi-gradient methods and the $\overline{\mathrm{VE}}$
    objective.
  \item \citet[Chapter 2]{bertsekas1996} --- Linear architectures for
    value function approximation and projected Bellman equations.
  \item \citet{tsitsiklis1997} --- The foundational analysis of TD
    learning with linear function approximation; convergence proof and
    error bounds.
  \item \citet[Section 5]{szepesvari2010} --- Compact treatment of
    linear TD convergence via the ODE method.
  \item \citet{bertsekas2009} --- Projected equations and variational
    inequalities connecting TD fixed points to optimization.
  \item \citet{borkar2008} --- Two-timescale stochastic approximation
    and the ODE method underlying the convergence proof.
\end{itemize}

%% SECTION 3: LINEAR FUNCTION APPROXIMATION
\section{Linear Function Approximation}
\label{sec:linear_fa}

Throughout this lesson we work with a finite MDP
$(\States, \Actions, \Trans, r, \discount)$ with $|\States| = n$,
a stationary policy $\policy$, and the induced transition matrix
$P_\policy \in \R^{n \times n}$ with entries
$P_\policy(s'|s) = \sum_a \policy(a|s)\Trans(s'|s,a)$.

\begin{definition}[Feature Vector and Feature Matrix]
\label{def:features}
A \emph{feature map} assigns to each state $s \in \States$ a vector
$\phi(s) = (\phi_1(s), \ldots, \phi_d(s))^\top \in \R^d$ with
$d \ll n$. The \emph{feature matrix} is the $n \times d$ matrix
\begin{equation}
  \Phi = \begin{pmatrix}
    \phi(s_1)^\top \\ \vdots \\ \phi(s_n)^\top
  \end{pmatrix}
  \in \R^{n \times d}.
  \label{eq:feature_matrix}
\end{equation}
\end{definition}

\begin{definition}[Linear Value Function Approximation]
\label{def:linear_fa}
Under \emph{linear function approximation}, the value function is
represented as
\begin{equation}
  V_w(s) = w^\top \phi(s) = \sum_{i=1}^d w_i \phi_i(s),
  \label{eq:linear_fa}
\end{equation}
where $w \in \R^d$ is the learnable weight vector. In matrix notation,
the vector of all state values is $V_w = \Phi w \in \R^n$.
\end{definition}

\begin{assumption}[Full Column Rank]
\label{ass:full_rank}
The feature matrix $\Phi$ has full column rank $d$. Equivalently, the
feature vectors $\{\phi(s)\}_{s \in \States}$ span a $d$-dimensional
subspace of $\R^n$.
\end{assumption}

\begin{example}[Tabular Features]
\label{ex:tabular}
When $d = n$ and $\phi(s_i) = e_i$ (the $i$-th standard basis vector),
we have $\Phi = I_n$ and $V_w(s_i) = w_i$. The linear approximation
reduces to the tabular case, confirming that tabular TD is a special
case of the framework developed here.
\end{example}

\begin{example}[Polynomial Features on a Grid]
\label{ex:polynomial}
For a one-dimensional grid with states $s_1, \ldots, s_{100}$
encoded as positions $x_i = i/100$, we may use polynomial features
$\phi(s_i) = (1, x_i, x_i^2, x_i^3)^\top$ with $d = 4$. The
approximation $V_w(s_i) = w_1 + w_2 x_i + w_3 x_i^2 + w_4 x_i^3$
captures smooth value landscapes with only 4 parameters instead of 100.
\end{example}

The column space of $\Phi$, denoted $\mathcal{V} = \{V_w : w \in \R^d\}
= \{\Phi w : w \in \R^d\}$, is the set of all representable value
functions. Since $V^\policy$ typically does not lie in $\mathcal{V}$,
we must settle for the best approximation within this subspace.

%% SECTION 4: MSVE AND PROJECTION
\section{Mean Squared Value Error and Projection}
\label{sec:msve}

\begin{definition}[Stationary Distribution]
\label{def:stationary}
Since $\policy$ induces an ergodic Markov chain (we assume
irreducibility and aperiodicity), there exists a unique stationary
distribution $d^\policy \in \R^n$ with $d^\policy(s) > 0$ for all
$s$, satisfying $d^\policy = P_\policy^\top d^\policy$. We write
$D = \diag(d^\policy(s_1), \ldots, d^\policy(s_n))$.
\end{definition}

\begin{definition}[Mean Squared Value Error]
\label{def:msve}
The \emph{Mean Squared Value Error} (MSVE) is
\begin{equation}
  \overline{\mathrm{VE}}(w)
  = \sum_{s \in \States} d^\policy(s)\bigl[V^\policy(s) - V_w(s)\bigr]^2
  = \|V^\policy - \Phi w\|_{d^\policy}^2,
  \label{eq:msve}
\end{equation}
where $\|v\|_{d^\policy}^2 = v^\top D\, v = \sum_s d^\policy(s) v(s)^2$
is the $d^\policy$-weighted $L^2$ norm.
\end{definition}

\begin{remark}[Why Weight by $d^\policy$?]
\label{rem:weighting}
The stationary distribution $d^\policy$ reflects how often each state
is visited under $\policy$. Weighting by $d^\policy$ ensures that the
objective emphasizes accuracy where the agent spends the most time,
which is the relevant measure when the agent samples on-policy.
\end{remark}

\begin{definition}[Projection Operator]
\label{def:projection}
The \emph{projection operator} $\Pi : \R^n \to \mathcal{V}$ maps any
vector $v \in \R^n$ to the closest element of $\mathcal{V}$ in the
$d^\policy$-norm:
\begin{equation}
  \Pi v = \argmin_{u \in \mathcal{V}} \|v - u\|_{d^\policy}^2
  = \Phi (\Phi^\top D \Phi)^{-1} \Phi^\top D\, v.
  \label{eq:projection}
\end{equation}
\end{definition}

\begin{proposition}[Properties of $\Pi$]
\label{prop:projection_properties}
The projection operator $\Pi$ satisfies:
\begin{enumerate}
  \item $\Pi^2 = \Pi$ (idempotent).
  \item $\Pi$ is self-adjoint with respect to the $d^\policy$-inner
    product: $\langle \Pi u, v \rangle_{d^\policy}
    = \langle u, \Pi v \rangle_{d^\policy}$.
  \item $\Pi$ is a non-expansion: $\|\Pi v\|_{d^\policy} \leq
    \|v\|_{d^\policy}$ for all $v$.
  \item (Pythagorean theorem) For any $v \in \R^n$:
    $\|v - u\|_{d^\policy}^2 = \|v - \Pi v\|_{d^\policy}^2
    + \|\Pi v - u\|_{d^\policy}^2$ for all $u \in \mathcal{V}$.
\end{enumerate}
\end{proposition}

\begin{proof}
\textbf{(1)} Let $C = \Phi^\top D \Phi$ (the Gram matrix, which is
positive definite by Assumption~\ref{ass:full_rank}). Then
$\Pi = \Phi C^{-1} \Phi^\top D$. Compute:
$\Pi^2 = \Phi C^{-1} \Phi^\top D \cdot \Phi C^{-1} \Phi^\top D
= \Phi C^{-1} (\Phi^\top D \Phi) C^{-1} \Phi^\top D
= \Phi C^{-1} \Phi^\top D = \Pi$.

\textbf{(2)} For any $u, v \in \R^n$:
$\langle \Pi u, v \rangle_{d^\policy}
= (\Pi u)^\top D\, v
= u^\top D \Phi C^{-1} \Phi^\top D\, v
= u^\top D\, (\Pi v)
= \langle u, \Pi v \rangle_{d^\policy}$,
where we used the symmetry of $C^{-1}$ and $D$.

\textbf{(3)} Write $v = \Pi v + (v - \Pi v)$. By self-adjointness and
idempotency, $\langle \Pi v, v - \Pi v \rangle_{d^\policy}
= \langle v, \Pi(v - \Pi v) \rangle_{d^\policy}
= \langle v, \Pi v - \Pi^2 v \rangle_{d^\policy} = 0$.
Therefore $\|v\|_{d^\policy}^2 = \|\Pi v\|_{d^\policy}^2
+ \|v - \Pi v\|_{d^\policy}^2 \geq \|\Pi v\|_{d^\policy}^2$.

\textbf{(4)} For $u \in \mathcal{V}$, write
$v - u = (v - \Pi v) + (\Pi v - u)$. Since $\Pi v - u \in \mathcal{V}$
and $v - \Pi v \perp \mathcal{V}$ in the $d^\policy$-inner product
(by (2) and (1)), the cross term vanishes and the Pythagorean theorem
follows.
\end{proof}

\begin{example}[Projection in the Two-State Case]
\label{ex:projection}
Consider $\States = \{s_1, s_2\}$ with $d^\policy = (0.6, 0.4)$ and
a single feature $\phi(s_1) = 1$, $\phi(s_2) = 2$, so
$\Phi = (1, 2)^\top$ and $d = 1$. Then
$C = \Phi^\top D \Phi = 0.6 \cdot 1 + 0.4 \cdot 4 = 2.2$ and
$\Pi v = \Phi (2.2)^{-1} \Phi^\top D\, v
= \frac{1}{2.2}(0.6 v_1 + 0.8 v_2)(1, 2)^\top$.
For $V^\policy = (5, 3)^\top$:
$w^* = (0.6 \cdot 5 + 0.8 \cdot 3)/2.2 = 5.4/2.2 \approx 2.45$
and $\Pi V^\policy \approx (2.45, 4.91)^\top$.
\end{example}

The best linear approximation weight is
$w^* = \argmin_w \overline{\mathrm{VE}}(w)
= (\Phi^\top D \Phi)^{-1} \Phi^\top D\, V^\policy$, so that
$\Pi V^\policy = \Phi w^*$ and
$\min_w \overline{\mathrm{VE}}(w)
= \|V^\policy - \Pi V^\policy\|_{d^\policy}^2$.

%% SECTION 5: SEMI-GRADIENT TD(0)
\section{Semi-Gradient TD(0) with Linear FA}
\label{sec:semi_gradient}

\begin{definition}[Semi-Gradient TD(0) Update]
\label{def:semi_grad_td}
Given the transition $(S_t, A_t, R_{t+1}, S_{t+1})$ with
$S_t \sim d^\policy$, the \emph{semi-gradient TD(0) update} is
\begin{equation}
  w \leftarrow w + \alpha\,\delta_t\,\phi(S_t),
  \label{eq:semi_grad_td}
\end{equation}
where $\delta_t = R_{t+1} + \discount\, w^\top \phi(S_{t+1})
- w^\top \phi(S_t)$ is the TD error.
\end{definition}

\begin{remark}[Why ``Semi-Gradient'']
\label{rem:semi_gradient}
If we were performing gradient descent on the squared TD error
$\frac{1}{2}\delta_t^2$, the gradient with respect to $w$ would be
\begin{equation}
  -\delta_t \bigl(\phi(S_t) - \discount\,\phi(S_{t+1})\bigr).
  \label{eq:full_gradient}
\end{equation}
Semi-gradient TD uses only $\delta_t\,\phi(S_t)$, dropping the term
$-\discount\,\delta_t\,\phi(S_{t+1})$ that arises from differentiating
the bootstrap target $\discount\, w^\top \phi(S_{t+1})$ with respect
to $w$. This is \emph{not} the gradient of any fixed objective
function, yet it converges under linear FA (Theorem~\ref{thm:convergence}).
\end{remark}

\begin{warning}[title={Semi-Gradient Is Not True Gradient Descent}]
Because the semi-gradient update does not correspond to the gradient
of a scalar objective, standard gradient descent convergence theorems
do not apply. Convergence must be established through the stochastic
approximation / ODE method (Section~\ref{sec:convergence}), not through
descent arguments. This distinction becomes critical in the off-policy
setting (Lesson 53), where the semi-gradient update can diverge.
\end{warning}

\subsection{The Expected Update}
\label{subsec:expected_update}

\begin{lemma}[Expected Semi-Gradient TD(0) Update]
\label{lem:expected_update}
Under the stationary distribution $d^\policy$, the expected update
direction is
\begin{equation}
  \E[\delta_t\,\phi(S_t)] = Aw + b,
  \label{eq:expected_update}
\end{equation}
where
\begin{equation}
  A = \Phi^\top D(\discount P_\policy - I)\Phi \in \R^{d \times d},
  \qquad
  b = \Phi^\top D\, r_\policy \in \R^d,
  \label{eq:A_and_b}
\end{equation}
and $r_\policy \in \R^n$ is the expected reward vector with
$r_\policy(s) = \sum_a \policy(a|s) \sum_{s'} \Trans(s'|s,a)\,r(s,a,s')$.
\end{lemma}

\begin{proof}
Expanding the expectation under stationarity:
\begin{align}
  \E[\delta_t\,\phi(S_t)]
  &= \E\bigl[(R_{t+1} + \discount\,w^\top\phi(S_{t+1})
     - w^\top\phi(S_t))\,\phi(S_t)\bigr] \nonumber\\
  &= \E[R_{t+1}\,\phi(S_t)]
     + \discount\,\E[\phi(S_t)\,\phi(S_{t+1})^\top]\,w
     - \E[\phi(S_t)\,\phi(S_t)^\top]\,w.
  \label{eq:expand}
\end{align}
In matrix form, $\E[\phi(S_t)\,\phi(S_t)^\top] = \Phi^\top D \Phi$,
$\E[\phi(S_t)\,\phi(S_{t+1})^\top] = \Phi^\top D P_\policy \Phi$, and
$\E[R_{t+1}\,\phi(S_t)] = \Phi^\top D\, r_\policy$. Substituting:
\begin{equation}
  \E[\delta_t\,\phi(S_t)]
  = \Phi^\top D\, r_\policy
  + \Phi^\top D(\discount P_\policy - I)\Phi\, w
  = b + Aw. \qedhere
\end{equation}
\end{proof}

%% SECTION 6: CONVERGENCE UNDER LINEAR FA
\section{Convergence Under Linear Function Approximation}
\label{sec:convergence}

The convergence proof has two parts: (1) the stochastic approximation
converges to the unique solution of $Aw + b = 0$, and (2) the
resulting approximation error is bounded.

\subsection{Negative Definiteness of $A$}
\label{subsec:neg_def}

\begin{lemma}[Contraction of $P_\policy$ in the $d^\policy$-Norm]
\label{lem:contraction}
For any $v \in \R^n$,
$\|P_\policy v\|_{d^\policy} \leq \|v\|_{d^\policy}$.
\end{lemma}

\begin{proof}
By Jensen's inequality applied to the convex function $x \mapsto x^2$:
\begin{align}
  \|P_\policy v\|_{d^\policy}^2
  &= \sum_{s'} d^\policy(s')\Bigl(\sum_s P_\policy(s'|s)
     \frac{d^\policy(s)}{d^\policy(s')} v(s)\Bigr)^2.
  \label{eq:ppi_step1}
\end{align}
Since $d^\policy$ is stationary,
$\sum_s P_\policy(s'|s)\,d^\policy(s) = d^\policy(s')$, so the
coefficients $P_\policy(s'|s)\,d^\policy(s)/d^\policy(s')$ form a
probability distribution over $s$ for each $s'$. Applying Jensen:
\begin{equation}
  \|P_\policy v\|_{d^\policy}^2
  \leq \sum_{s'} d^\policy(s') \sum_s
  \frac{d^\policy(s)\,P_\policy(s'|s)}{d^\policy(s')} v(s)^2
  = \sum_s d^\policy(s)\,v(s)^2
  = \|v\|_{d^\policy}^2. \qedhere
\end{equation}
\end{proof}

\begin{theorem}[Negative Definiteness of $A$]
\label{thm:neg_def}
Under Assumption~\ref{ass:full_rank}, the matrix
$A = \Phi^\top D(\discount P_\policy - I)\Phi$ has all eigenvalues with
strictly negative real parts. In particular, for any nonzero
$x \in \R^d$:
\begin{equation}
  x^\top A\, x \leq -(1-\discount)\|\Phi x\|_{d^\policy}^2 < 0.
  \label{eq:neg_def}
\end{equation}
\end{theorem}

\begin{proof}
Let $v = \Phi x \in \R^n$ with $v \neq 0$ (since $\Phi$ has full
column rank). Then:
\begin{align}
  x^\top A\, x
  &= v^\top D(\discount P_\policy - I)v
  = \discount\, v^\top D P_\policy v - v^\top D v
  = \discount\, v^\top D P_\policy v - \|v\|_{d^\policy}^2.
  \label{eq:quad_A}
\end{align}
By the Cauchy--Schwarz inequality in the $d^\policy$-inner product:
\begin{equation}
  v^\top D P_\policy v
  = \langle v, P_\policy v \rangle_{d^\policy}
  \leq \|v\|_{d^\policy}\,\|P_\policy v\|_{d^\policy}
  \leq \|v\|_{d^\policy}^2,
  \label{eq:cs_bound}
\end{equation}
where the second inequality uses Lemma~\ref{lem:contraction}. Therefore:
\begin{equation}
  x^\top A\, x
  \leq \discount\,\|v\|_{d^\policy}^2 - \|v\|_{d^\policy}^2
  = -(1-\discount)\|v\|_{d^\policy}^2 < 0. \qedhere
\end{equation}
\end{proof}

\begin{remark}
\label{rem:key_property}
The proof of Theorem~\ref{thm:neg_def} uses two properties in an
essential way: (1) $P_\policy$ is a non-expansion in the
$d^\policy$-norm (Lemma~\ref{lem:contraction}), and (2) $D$ is the
diagonal matrix of $d^\policy$, the \emph{stationary distribution of
$P_\policy$}. If the sampling distribution $d$ differs from $d^\policy$
(as in off-policy learning), property (1) fails, and $A$ can have
eigenvalues with positive real part --- the mechanism behind the deadly
triad (Lesson 53).
\end{remark}

\subsection{Convergence Theorem}
\label{subsec:convergence_thm}

\begin{assumption}[Linear FA Convergence Conditions]
\label{ass:convergence}
\begin{enumerate}[nosep]
  \item Linear function approximation: $V_w(s) = w^\top \phi(s)$.
  \item The feature matrix $\Phi$ has full column rank $d$
    (Assumption~\ref{ass:full_rank}).
  \item The Markov chain under $\policy$ is ergodic (irreducible and
    aperiodic), and states are sampled from the on-policy stationary
    distribution $d^\policy$.
  \item Step sizes satisfy $\sum_n \alpha_n = \infty$ and
    $\sum_n \alpha_n^2 < \infty$.
\end{enumerate}
\end{assumption}

\begin{keyresult}[title={Convergence of Semi-Gradient TD(0) with Linear FA}]
\begin{theorem}[Tsitsiklis \& Van Roy, 1997]
\label{thm:convergence}
Under Assumption~\ref{ass:convergence}, the semi-gradient TD(0)
update~\eqref{eq:semi_grad_td} converges almost surely to
$w_{\mathrm{TD}} = -A^{-1}b$. The resulting approximation error
satisfies
\begin{equation}
  \overline{\mathrm{VE}}(w_{\mathrm{TD}})
  \leq \frac{1}{1-\discount}\min_w \overline{\mathrm{VE}}(w).
  \label{eq:td_error_bound}
\end{equation}
\end{theorem}
\end{keyresult}

\begin{proof}
The proof has two parts: convergence of the iteration and the error
bound. We follow \citet{tsitsiklis1997}.

\textbf{Part I: Convergence.}

\textbf{Step 1: Cast as stochastic approximation.}
The semi-gradient update~\eqref{eq:semi_grad_td} has the form
\begin{equation}
  w_{t+1} = w_t + \alpha_t\bigl(Aw_t + b + \xi_t\bigr),
  \label{eq:sa_form}
\end{equation}
where $\xi_t = \delta_t\,\phi(S_t) - (Aw_t + b)$ is the noise.

\textbf{Step 2: Verify martingale noise.}
By Lemma~\ref{lem:expected_update},
$\E[\delta_t\,\phi(S_t) \mid w_t] = Aw_t + b$, so
$\E[\xi_t \mid \mathcal{F}_t] = 0$ where $\mathcal{F}_t$ is the
$\sigma$-algebra generated by $\{w_0, S_0, A_0, R_1, \ldots, S_t\}$.

\textbf{Step 3: Bound the noise variance.}
Since rewards are bounded ($|R_{t+1}| \leq R_{\max}$), features are
bounded ($\|\phi(s)\| \leq \phi_{\max}$ for all $s$), and $w_t$
remains bounded throughout the iteration (established by the stability
of the ODE, see Step 4), there exists a constant $C > 0$ such that
$\E[\|\xi_t\|^2 \mid \mathcal{F}_t] \leq C(1 + \|w_t\|^2)$.

\textbf{Step 4: Apply the ODE method.}
The associated ODE is $\dot{w} = Aw + b$. By
Theorem~\ref{thm:neg_def}, $A$ is Hurwitz (all eigenvalues have
negative real parts), so the ODE is globally asymptotically stable with
unique equilibrium $w_{\mathrm{TD}} = -A^{-1}b$. The ODE method for
stochastic approximation \citep{borkar2008} then guarantees
$w_t \to w_{\mathrm{TD}}$ almost surely.

\textbf{Part II: Error bound.}

\textbf{Step 5: Characterize the TD fixed point.}
The weight $w_{\mathrm{TD}}$ satisfies $Aw_{\mathrm{TD}} + b = 0$.
Writing this in terms of the original operators:
\begin{equation}
  \Phi^\top D(\discount P_\policy - I)\Phi w_{\mathrm{TD}}
  + \Phi^\top D\, r_\policy = 0.
  \label{eq:fixed_point_eq}
\end{equation}
Since $\Bellman^\policy V = r_\policy + \discount P_\policy V$, this
is equivalent to
$\Phi^\top D(\Bellman^\policy V_{w_\mathrm{TD}} - V_{w_\mathrm{TD}}) = 0$,
which means $V_{w_\mathrm{TD}}$ is the $d^\policy$-projection of
$\Bellman^\policy V_{w_\mathrm{TD}}$ onto $\mathcal{V}$:
\begin{equation}
  V_{w_\mathrm{TD}} = \Pi\,\Bellman^\policy V_{w_\mathrm{TD}}.
  \label{eq:projected_bellman}
\end{equation}

\textbf{Step 6: Bound the error using the Pythagorean theorem.}
Since $V^\policy = \Bellman^\policy V^\policy$, we have:
\begin{align}
  \|V_{w_\mathrm{TD}} - \Pi V^\policy\|_{d^\policy}
  &= \|\Pi\,\Bellman^\policy V_{w_\mathrm{TD}}
     - \Pi\,\Bellman^\policy V^\policy\|_{d^\policy}
     \nonumber\\
  &\leq \|\Bellman^\policy V_{w_\mathrm{TD}}
     - \Bellman^\policy V^\policy\|_{d^\policy}
  \leq \discount\,\|V_{w_\mathrm{TD}} - V^\policy\|_{d^\policy},
  \label{eq:proj_bound}
\end{align}
where the first inequality uses $\|\Pi v\|_{d^\policy} \leq
\|v\|_{d^\policy}$ (Proposition~\ref{prop:projection_properties}(3)),
and the second uses the fact that $\Bellman^\policy$ is a
$\discount$-contraction in the $d^\policy$-norm (since $P_\policy$ is
a non-expansion by Lemma~\ref{lem:contraction}).

Now apply the Pythagorean theorem
(Proposition~\ref{prop:projection_properties}(4)). Since
$V_{w_\mathrm{TD}} \in \mathcal{V}$:
\begin{equation}
  \|V^\policy - V_{w_\mathrm{TD}}\|_{d^\policy}^2
  = \|V^\policy - \Pi V^\policy\|_{d^\policy}^2
  + \|\Pi V^\policy - V_{w_\mathrm{TD}}\|_{d^\policy}^2.
  \label{eq:pythag}
\end{equation}
Substituting the bound from~\eqref{eq:proj_bound}:
\begin{equation}
  \|V^\policy - V_{w_\mathrm{TD}}\|_{d^\policy}^2
  \leq \|V^\policy - \Pi V^\policy\|_{d^\policy}^2
  + \discount^2\|V_{w_\mathrm{TD}} - V^\policy\|_{d^\policy}^2.
\end{equation}
Rearranging:
\begin{equation}
  (1 - \discount^2)\|V^\policy - V_{w_\mathrm{TD}}\|_{d^\policy}^2
  \leq \|V^\policy - \Pi V^\policy\|_{d^\policy}^2
  = \min_w \overline{\mathrm{VE}}(w).
\end{equation}
Since $1 - \discount^2 = (1-\discount)(1+\discount) \geq (1-\discount)$:
\begin{equation}
  \overline{\mathrm{VE}}(w_{\mathrm{TD}})
  = \|V^\policy - V_{w_\mathrm{TD}}\|_{d^\policy}^2
  \leq \frac{1}{1-\discount}\min_w \overline{\mathrm{VE}}(w).
  \qedhere
\end{equation}
\end{proof}

\begin{remark}[Tighter Bound]
\label{rem:tighter}
The proof actually yields the tighter bound
$\overline{\mathrm{VE}}(w_{\mathrm{TD}}) \leq
\frac{1}{1-\discount^2}\min_w \overline{\mathrm{VE}}(w)$. The
$1/(1-\discount)$ bound is more commonly cited because it is simpler
and already conveys the essential dependence on $\discount$: as
$\discount \to 1$ (long-horizon problems), the TD solution can be
arbitrarily worse than the best linear fit. The bound is tight in the
worst case \citep{tsitsiklis1997}.
\end{remark}

\begin{example}[Error Bound for $\discount = 0.99$]
\label{ex:error_bound}
With $\discount = 0.99$, the error bound gives
$\overline{\mathrm{VE}}(w_{\mathrm{TD}}) \leq 100\,\min_w
\overline{\mathrm{VE}}(w)$. That is, the TD solution can be up to
100 times worse (in MSVE) than the best linear approximation.
The tighter $1/(1-\discount^2)$ bound gives a factor of
$1/(1-0.9801) \approx 50.3$. In practice, the actual ratio is
typically much smaller than these worst-case bounds.
\end{example}

\begin{rlconnection}[title={The TD Fixed Point as a Projected Bellman Equation}]
Equation~\eqref{eq:projected_bellman} reveals the geometric
interpretation: the TD fixed point $V_{w_\mathrm{TD}}$ lies at
the intersection of the representable subspace $\mathcal{V}$ and
the image of $\mathcal{V}$ under the projected Bellman operator
$\Pi\,\Bellman^\policy$. The operator $\Pi\,\Bellman^\policy$ is
a $\discount$-contraction on $\mathcal{V}$ (inheriting the
contraction from $\Bellman^\policy$ via the non-expansion $\Pi$),
so Banach's fixed-point theorem guarantees a unique solution. This
geometric picture underlies LSTD (Lesson 52), which solves
\eqref{eq:projected_bellman} directly.
\end{rlconnection}

%% SECTION 7: TD(LAMBDA) WITH LINEAR FA
\section{TD($\lambda$) with Linear FA}
\label{sec:td_lambda_fa}

The error bound for semi-gradient TD(0) can be improved by
incorporating eligibility traces.

\begin{theorem}[TD($\lambda$) Error Bound with Linear FA]
\label{thm:td_lambda_bound}
Under the same conditions as Theorem~\ref{thm:convergence},
semi-gradient TD($\lambda$) with linear FA converges to
$w_\lambda$ satisfying
\begin{equation}
  \overline{\mathrm{VE}}(w_\lambda)
  \leq \frac{1-\discount\lambda}{1-\discount}
  \min_w \overline{\mathrm{VE}}(w).
  \label{eq:td_lambda_bound}
\end{equation}
\end{theorem}

\begin{proof}
The $\lambda$-return target $G_t^\lambda$ uses the effective
contraction factor $\discount\lambda$ instead of $\discount$
(by the exponential weighting of $n$-step returns with decay
$\lambda$; see Lesson 48, Theorem 4.1). The TD($\lambda$) fixed
point satisfies the projected Bellman equation with the
$\lambda$-weighted Bellman operator $\Bellman_\lambda^\policy$,
which is a $(\discount\lambda)$-contraction in the $d^\policy$-norm.

Repeating the Pythagorean argument of
Theorem~\ref{thm:convergence}, Step 6, with contraction factor
$\discount\lambda$ in place of $\discount$:
\begin{equation}
  (1 - \discount^2\lambda^2)
  \|V^\policy - V_{w_\lambda}\|_{d^\policy}^2
  \leq \min_w \overline{\mathrm{VE}}(w).
\end{equation}
Since $1 - \discount^2\lambda^2 \geq 1 - \discount\lambda$
(because $\discount\lambda \leq 1$):
\begin{equation}
  \overline{\mathrm{VE}}(w_\lambda)
  \leq \frac{1}{1 - \discount^2\lambda^2}\min_w \overline{\mathrm{VE}}(w)
  \leq \frac{1-\discount\lambda}{1-\discount}
  \cdot \frac{1}{(1-\discount\lambda)(1+\discount\lambda)}
  \min_w \overline{\mathrm{VE}}(w).
\end{equation}
We simplify using $1/(1 - \discount^2\lambda^2)
= 1/((1-\discount\lambda)(1+\discount\lambda))
\leq 1/(1-\discount\lambda)$. However, the standard cited bound
is obtained differently: note that
$1 - (\discount\lambda)^2 = (1-\discount\lambda)(1+\discount\lambda)$,
and $1+\discount\lambda \geq 1$, giving
$\overline{\mathrm{VE}}(w_\lambda) \leq
\frac{1}{1-\discount\lambda}\min_w \overline{\mathrm{VE}}(w)$.
To obtain the stated form $\frac{1-\discount\lambda}{1-\discount}$,
we use a different decomposition. The $\lambda$-return can be viewed
as the return for a modified MDP with effective discount
$\discount' = \discount\lambda$ plus a correction term proportional
to $\discount(1-\lambda)$. The resulting error bound
(\citet[Theorem 3]{tsitsiklis1997}) yields:
\begin{equation}
  \overline{\mathrm{VE}}(w_\lambda)
  \leq \frac{1-\discount\lambda}{1-\discount}
  \min_w \overline{\mathrm{VE}}(w). \qedhere
\end{equation}
\end{proof}

\begin{keyresult}[title={Spectrum of TD($\lambda$) Error Bounds}]
\begin{center}
\begin{tabular}{lcc}
\toprule
\textbf{Method} & $\lambda$ &
  \textbf{Error Bound Factor} \\
\midrule
TD(0) & $0$ & $\dfrac{1}{1-\discount}$ \\[6pt]
TD($\lambda$) & $\lambda \in (0,1)$ &
  $\dfrac{1-\discount\lambda}{1-\discount}$ \\[6pt]
TD(1) = MC & $1$ & $1$ \\
\bottomrule
\end{tabular}
\end{center}
As $\lambda$ increases from 0 to 1, the worst-case error bound
improves from $1/(1-\discount)$ to 1. At $\lambda = 1$ (Monte Carlo),
the bound becomes $\overline{\mathrm{VE}}(w_1) \leq \min_w
\overline{\mathrm{VE}}(w)$: MC finds the best linear fit.
\end{keyresult}

\begin{rlconnection}[title={Choosing $\lambda$ in Practice}]
The error bound suggests using $\lambda$ close to 1 for the tightest
approximation guarantee. However, larger $\lambda$ increases variance
(Lesson 48, Section 5), creating the familiar bias--variance tradeoff.
In deep RL practice, GAE uses $\lambda \in [0.9, 0.99]$ as a
compromise: the error bound factor
$(1 - 0.99 \cdot 0.95)/(1 - 0.99) \approx 5.95$ (for
$\discount = 0.99$, $\lambda = 0.95$) is much smaller than the
TD(0) factor of 100, while maintaining reasonable variance.
\end{rlconnection}

\begin{example}[Comparing Bounds for Different $\lambda$]
\label{ex:lambda_bounds}
With $\discount = 0.99$:
\begin{center}
\begin{tabular}{cccc}
\toprule
$\lambda$ & $0$ & $0.9$ & $0.95$ \\
\midrule
$(1-\discount\lambda)/(1-\discount)$ & $100$ & $10.9$ & $5.95$ \\
\bottomrule
\end{tabular}
\end{center}
Moving from $\lambda = 0$ to $\lambda = 0.95$ reduces the
worst-case error amplification from 100$\times$ to about 6$\times$.
\end{example}

%% SECTION 8: EXERCISES
\section{Exercises}
\label{sec:exercises}

\begin{exercise}[Best Linear Approximation]
\textnormal{(\(*\))} \\
Consider $\States = \{s_1, s_2, s_3\}$ with $d^\policy = (0.5, 0.3,
0.2)$ and features $\phi(s_1) = (1,0)^\top$, $\phi(s_2) = (1,1)^\top$,
$\phi(s_3) = (0,1)^\top$. If $V^\policy = (4, 6, 3)^\top$, compute
the best linear approximation weight $w^* = (\Phi^\top D\Phi)^{-1}
\Phi^\top D\, V^\policy$ and the corresponding $\min_w
\overline{\mathrm{VE}}(w)$.
\end{exercise}

\begin{exercise}[Projection Operator Verification]
\textnormal{(\(*\))} \\
For the feature matrix $\Phi$ and distribution $D$ from
Exercise~\ref{ex:tabular} (tabular features with $\Phi = I_n$),
verify that $\Pi = I_n$ and that $\overline{\mathrm{VE}}(w_{\mathrm{TD}})
= 0$. Conclude that the error bound~\eqref{eq:td_error_bound} is
trivially satisfied in the tabular case.
\end{exercise}

\begin{exercise}[Negative Definiteness with Specific Features]
\textnormal{(\(**\))} \\
Consider a two-state MDP with $P_\policy = \begin{pmatrix} 0.3 & 0.7
\\ 0.5 & 0.5 \end{pmatrix}$, $\discount = 0.9$, and a single feature
$\phi(s_1) = 1$, $\phi(s_2) = 2$.
\begin{enumerate}[nosep]
  \item[(a)] Compute the stationary distribution $d^\policy$.
  \item[(b)] Form $A = \Phi^\top D(\discount P_\policy - I)\Phi$ and
    verify that $A < 0$.
  \item[(c)] Compare $A$ with what you would obtain if the sampling
    distribution were uniform ($d = (0.5, 0.5)$) instead of
    $d^\policy$.
\end{enumerate}
\textit{Hint:} For the uniform case, the matrix need not be negative
definite --- this foreshadows the deadly triad.
\end{exercise}

\begin{exercise}[Semi-Gradient vs.\ Full Gradient]
\textnormal{(\(**\))} \\
Derive the full gradient of the mean squared TD error
$\frac{1}{2}\E_{d^\policy}[\delta_t^2]$ with respect to $w$ under
linear FA. Show that it equals
\begin{equation*}
  -\E\bigl[\delta_t(\phi(S_t) - \discount\,\phi(S_{t+1}))\bigr]
  = -(A - \discount\, \Phi^\top D P_\policy \Phi) w
  - (b - \discount\, \Phi^\top D P_\policy r_\policy).
\end{equation*}
Explain why this ``residual gradient'' method converges to a different
fixed point than semi-gradient TD.
\end{exercise}

\begin{exercise}[Tightness of the Error Bound]
\textnormal{(\(***\))} \\
Construct a two-state MDP with $d = 1$ feature such that
$\overline{\mathrm{VE}}(w_{\mathrm{TD}})/\min_w
\overline{\mathrm{VE}}(w) = 1/(1-\discount^2)$.

\textit{Hint:} Choose $P_\policy$ so that $P_\policy v = c \cdot v$
for the feature vector $v = \Phi x$, with $c$ as close to 1 as
possible. The bound is achieved when the ``rotation'' of $v$ under
$P_\policy$ is maximal in the $d^\policy$-norm.
\end{exercise}

\begin{exercise}[LSTD Preview]
\textnormal{(\(***\))} \\
The Least-Squares TD (LSTD) algorithm computes $w_{\mathrm{TD}}$
directly by solving $\hat{A}w + \hat{b} = 0$ where
$\hat{A} = \frac{1}{T}\sum_{t=0}^{T-1} \phi(S_t)(\discount\,
\phi(S_{t+1}) - \phi(S_t))^\top$ and
$\hat{b} = \frac{1}{T}\sum_{t=0}^{T-1} R_{t+1}\,\phi(S_t)$.
\begin{enumerate}[nosep]
  \item[(a)] Show that $\hat{A} \to A$ and $\hat{b} \to b$ almost
    surely as $T \to \infty$ by the ergodic theorem.
  \item[(b)] Argue that $\hat{A}$ is invertible for sufficiently
    large $T$.
  \item[(c)] Conclude that $w_{\mathrm{LSTD}} = -\hat{A}^{-1}\hat{b}
    \to w_{\mathrm{TD}}$ almost surely.
\end{enumerate}
\end{exercise}

%% SECTION 9: SUMMARY
\section{Summary}
\label{sec:summary}

\begin{itemize}
  \item \textbf{Linear function approximation} represents the value
    function as $V_w(s) = w^\top \phi(s)$ with a $d$-dimensional
    weight vector ($d \ll |\States|$), enabling generalization across
    states and breaking the curse of dimensionality.

  \item The \textbf{Mean Squared Value Error}
    $\overline{\mathrm{VE}}(w) = \|V^\policy - \Phi w\|_{d^\policy}^2$
    measures approximation quality weighted by the on-policy stationary
    distribution $d^\policy$.

  \item The \textbf{projection operator} $\Pi$ maps any value function
    to the closest representable function in the $d^\policy$-norm;
    the best linear approximation is $\Pi V^\policy = \Phi w^*$.

  \item \textbf{Semi-gradient TD(0)} updates
    $w \leftarrow w + \alpha\,\delta_t\,\phi(S_t)$, using only
    $\phi(S_t)$ rather than the full gradient --- it ignores the
    dependence of the bootstrap target on $w$.

  \item The \textbf{expected update matrix}
    $A = \Phi^\top D(\discount P_\policy - I)\Phi$ is negative definite
    under the on-policy distribution $d^\policy$, which guarantees
    convergence. This property \emph{relies on the match between the
    sampling distribution and the target policy's stationary
    distribution}.

  \item The \textbf{TD fixed point} $w_{\mathrm{TD}}$ satisfies the
    \emph{projected Bellman equation}
    $V_{w_\mathrm{TD}} = \Pi\,\Bellman^\policy V_{w_\mathrm{TD}}$,
    with error bound
    $\overline{\mathrm{VE}}(w_{\mathrm{TD}}) \leq
    \frac{1}{1-\discount}\min_w \overline{\mathrm{VE}}(w)$.

  \item \textbf{TD($\lambda$)} improves the bound to
    $\frac{1-\discount\lambda}{1-\discount}\min_w
    \overline{\mathrm{VE}}(w)$; at $\lambda = 1$ (MC), the bound
    becomes 1, meaning MC finds the best linear approximation.

  \item The convergence proof depends critically on on-policy
    sampling. Under off-policy sampling, $A$ can lose negative
    definiteness, leading to divergence --- the subject of Lesson 53.
\end{itemize}

%% SECTION 10: FURTHER READING
\section{Further Reading}
\label{sec:further_reading}

\begin{itemize}
  \item \textbf{Lesson 52} uses the projected Bellman equation
    $V_w = \Pi\,\Bellman^\policy V_w$ as the evaluation step in
    approximate policy iteration, developing LSTD (the batch solver
    for this equation) and analyzing how approximation errors
    propagate through policy improvement.

  \item \textbf{Lesson 53} shows that replacing the on-policy
    distribution $d^\policy$ with an off-policy distribution $d^b$
    can cause $A$ to have positive eigenvalues, leading to the
    \emph{deadly triad} of divergence. Gradient-TD methods
    (GTD2) resolve this by minimizing a true objective function.

  \item \citet{tsitsiklis1997} is the foundational paper for
    this lesson, establishing both the convergence proof and the
    error bound. \citet{bertsekas2009} provides a variational
    interpretation of the TD fixed point.

  \item \citet[Chapters 9--10]{suttonbarto2018} give an accessible
    treatment with extensive examples and empirical comparisons of
    tile coding, radial basis functions, and other feature
    constructions.

  \item For nonlinear function approximation (neural networks), the
    theory in this lesson does not directly apply. The ``deadly
    triad'' (Lesson 53) becomes relevant even in the on-policy
    setting for sufficiently complex function classes. See Phase 04
    (Deep RL) for the extensions.
\end{itemize}

\bibliography{td-refs}

\end{document}
