\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 52: Approximate Policy Iteration}}
\fancyhead[R]{\thepage}
\title{Lesson 52: Approximate Policy Iteration}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% SECTION 1: MOTIVATION AND OVERVIEW
\section{Motivation and Overview}
\label{sec:motivation}

Exact policy iteration (PI) alternates between two steps: (1) compute
$V^\policy$ exactly by solving the Bellman equation, and (2) improve
the policy by acting greedily with respect to $V^\policy$. In practice,
neither step can be performed exactly. The state space may be too large
for exact evaluation (requiring function approximation as developed in
Lesson 51), and the greedy improvement may be computed from a noisy or
approximate value function. Consider training an autonomous helicopter:
the continuous state space (position, velocity, orientation) makes exact
evaluation impossible, and the policy improvement must rely on a learned
approximation $\hat{V}^\policy$ that inevitably incurs error.

The central question of this lesson is: \emph{how do evaluation and
improvement errors propagate through multiple PI iterations?} If each
evaluation has error $\varepsilon_{\mathrm{eval}}$ and each improvement
has error $\varepsilon_{\mathrm{improve}}$, does the resulting policy
remain near-optimal, or do errors compound catastrophically? The tools
developed in this lesson will let us prove a precise bound showing that
errors accumulate at rate $O(1/(1-\discount)^2)$, and that achieving
an $\varepsilon$-optimal policy requires per-iteration accuracy
$O(\varepsilon(1-\discount)^2)$. We then develop LSTD and LSPI as
concrete instantiations of this framework.

\begin{rlconnection}[title={Where Approximate PI Appears}]
\begin{itemize}[nosep]
  \item \textbf{Least-Squares Policy Iteration (LSPI)} combines LSTD
    evaluation with greedy improvement, forming a practical batch RL
    algorithm widely used for continuous-state problems.
  \item \textbf{Fitted Value Iteration (FVI)} replaces the evaluation
    step with a supervised learning regression, used in batch RL and
    offline RL settings.
  \item \textbf{Actor-critic methods} (PPO, SAC, TRPO) are approximate
    PI algorithms where the critic performs approximate evaluation and
    the actor performs approximate (gradient-based) improvement.
  \item \textbf{Error propagation bounds} guide the design of deep RL
    algorithms: they explain why high-quality value estimates are
    essential and why long horizons ($\discount$ close to 1) make
    learning harder.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
Policy iteration: policy evaluation (solving
$V^\policy = \Bellman^\policy V^\policy$) and policy improvement
(greedy operator $\policy' = \argmax_a Q^\policy(s,a)$), convergence
of exact PI in finitely many steps (Lesson 45). Linear function
approximation, the projected Bellman equation
$V_w = \Pi\,\Bellman^\policy V_w$, the TD fixed point $w_{\mathrm{TD}}$,
and the error bound
$\overline{\mathrm{VE}}(w_{\mathrm{TD}}) \leq
\frac{1}{1-\discount}\min_w \overline{\mathrm{VE}}(w)$ (Lesson 51,
Theorem~6.1). Bellman optimality operator $\Bellman^*$ and its
$\discount$-contraction (Lesson 45). Matrix algebra: solving linear
systems, rank conditions (Lesson 6).

\paragraph{Learning objectives.}
\begin{itemize}
  \item \emph{State} the approximate policy iteration framework with
    evaluation error $\varepsilon_{\mathrm{eval}}$ and improvement
    error $\varepsilon_{\mathrm{improve}}$.
  \item \emph{Prove} the error propagation bound
    $\limsup_{k \to \infty} \|V^* - V^{\policy_k}\|_\infty
    \leq \frac{2\discount(\varepsilon_{\mathrm{eval}}
    + \varepsilon_{\mathrm{improve}})}{(1-\discount)^2}$.
  \item \emph{Derive} the LSTD algorithm as the batch solver for
    the projected Bellman equation and \emph{explain} its connection
    to semi-gradient TD.
  \item \emph{Construct} the LSPI algorithm combining LSTD evaluation
    with greedy improvement.
  \item \emph{Analyze} the sample complexity implications of the error
    propagation bound.
\end{itemize}

%% SECTION 2: REFERENCES
\section{References}
\label{sec:references}

\begin{itemize}[leftmargin=2em]
  \item \citet[Chapter 6]{bertsekas1996} --- Approximate dynamic
    programming and error bounds for approximate policy iteration.
  \item \citet[Chapter 3]{szepesvari2010} --- Compact treatment of
    approximate policy iteration with error propagation analysis.
  \item \citet[Chapter 10]{suttonbarto2018} --- On-policy control
    with approximation; semi-gradient methods for control.
  \item \citet{bertsekas2009} --- Projected equations and
    temporal-difference methods; LSTD and LSPI foundations.
  \item \citet{tsitsiklis1997} --- Convergence of TD with linear FA,
    providing the evaluation error bounds used in this lesson.
  \item \citet{puterman2014} --- Policy iteration theory including
    error analysis for approximate dynamic programming.
\end{itemize}

%% SECTION 3: EXACT PI RECAP
\section{Exact Policy Iteration Recap}
\label{sec:exact_pi}

We begin by recalling exact PI to establish the framework that
approximate PI generalizes.

\begin{definition}[Policy Iteration]
\label{def:pi}
\emph{Policy iteration} generates a sequence of policies
$\policy_0, \policy_1, \policy_2, \ldots$ by alternating:
\begin{enumerate}
  \item \textbf{Policy evaluation:} Compute $V^{\policy_k}$ exactly
    by solving $V = \Bellman^{\policy_k} V$.
  \item \textbf{Policy improvement:} Set
    $\policy_{k+1}(s) = \argmax_{a \in \Actions}
    Q^{\policy_k}(s,a)$ for all $s$, where
    $Q^{\policy_k}(s,a) = \sum_{s'} \Trans(s'|s,a)[r(s,a,s')
    + \discount V^{\policy_k}(s')]$.
\end{enumerate}
\end{definition}

\begin{proposition}[Monotone Improvement and Convergence]
\label{prop:exact_pi}
Exact PI satisfies $V^{\policy_{k+1}}(s) \geq V^{\policy_k}(s)$ for
all $s$ and all $k$. For a finite MDP, it converges to the optimal
policy $\optpolicy$ in at most $|\Actions|^{|\States|}$ iterations.
\end{proposition}

\begin{proof}
By the policy improvement theorem (Lesson 45, Theorem~4.2), greedy
improvement with respect to $V^{\policy_k}$ yields a policy at least
as good: $V^{\policy_{k+1}} \geq V^{\policy_k}$ pointwise. Since
the number of deterministic policies is finite ($|\Actions|^{|\States|}$)
and each improvement is strict unless $\policy_k = \optpolicy$, PI
terminates at $\optpolicy$ in finitely many steps.
\end{proof}

The problem is that Step 1 requires solving a linear system of
$|\States|$ equations, and Step 2 requires computing
$Q^{\policy_k}(s,a)$ for all $(s,a)$. For large state spaces, both
steps must be approximated.

%% SECTION 4: APPROXIMATE PI FRAMEWORK
\section{Approximate Policy Iteration}
\label{sec:approx_pi}

\begin{definition}[Approximate Policy Evaluation]
\label{def:approx_eval}
An \emph{$\varepsilon_{\mathrm{eval}}$-approximate evaluation} of
policy $\policy$ produces a value function estimate
$\hat{V}^\policy$ satisfying
\begin{equation}
  \|\hat{V}^\policy - V^\policy\|_\infty
  \leq \varepsilon_{\mathrm{eval}}.
  \label{eq:eval_error}
\end{equation}
\end{definition}

\begin{definition}[Approximate Policy Improvement]
\label{def:approx_improve}
An \emph{$\varepsilon_{\mathrm{improve}}$-approximate improvement}
with respect to $\hat{V}$ produces a policy $\policy'$ satisfying
\begin{equation}
  \Bellman^{\policy'} \hat{V}(s)
  \geq \Bellman^* \hat{V}(s) - \varepsilon_{\mathrm{improve}}
  \quad \text{for all } s \in \States.
  \label{eq:improve_error}
\end{equation}
When $\varepsilon_{\mathrm{improve}} = 0$, the improvement is
exactly greedy: $\policy'(s) = \argmax_a Q_{\hat{V}}(s,a)$.
\end{definition}

\begin{remark}[Sources of Improvement Error]
\label{rem:improve_error}
Improvement error $\varepsilon_{\mathrm{improve}} > 0$ arises when:
(a) the maximization over actions is approximate (e.g., in continuous
action spaces), (b) the policy is restricted to a parameterized class
that cannot represent the greedy policy exactly, or (c) the greedy
computation uses sampled transitions rather than exact expectations.
\end{remark}

\begin{definition}[Approximate Policy Iteration]
\label{def:approx_pi}
\emph{Approximate Policy Iteration} (API) generates policies
$\policy_0, \policy_1, \ldots$ by:
\begin{enumerate}
  \item \textbf{Approximate evaluation:} Compute $\hat{V}^{\policy_k}$
    with $\|\hat{V}^{\policy_k} - V^{\policy_k}\|_\infty
    \leq \varepsilon_{\mathrm{eval}}$.
  \item \textbf{Approximate improvement:} Compute $\policy_{k+1}$ with
    $\Bellman^{\policy_{k+1}} \hat{V}^{\policy_k}
    \geq \Bellman^* \hat{V}^{\policy_k}
    - \varepsilon_{\mathrm{improve}}$.
\end{enumerate}
\end{definition}

%% SECTION 5: ERROR PROPAGATION
\section{Error Propagation in Approximate PI}
\label{sec:error_propagation}

\begin{lemma}[One-Step Error Bound]
\label{lem:one_step}
Let $\hat{V}$ satisfy $\|\hat{V} - V^\policy\|_\infty \leq
\varepsilon_{\mathrm{eval}}$ and let $\policy'$ satisfy
$\Bellman^{\policy'} \hat{V} \geq \Bellman^* \hat{V}
- \varepsilon_{\mathrm{improve}}$. Then for all $s$:
\begin{equation}
  \Bellman^{\policy'} V^\policy(s)
  \geq \Bellman^* V^\policy(s)
  - \varepsilon_{\mathrm{improve}}
  - 2\discount\,\varepsilon_{\mathrm{eval}}.
  \label{eq:one_step}
\end{equation}
\end{lemma}

\begin{proof}
Since $\Bellman^{\policy'}$ and $\Bellman^*$ are monotone operators
and $|\hat{V}(s) - V^\policy(s)| \leq \varepsilon_{\mathrm{eval}}$
implies
$\hat{V}(s) - \varepsilon_{\mathrm{eval}} \leq V^\policy(s) \leq
\hat{V}(s) + \varepsilon_{\mathrm{eval}}$,
we bound the Bellman operators. For any policy $\tilde{\policy}$ and
any $V_1, V_2$ with $\|V_1 - V_2\|_\infty \leq \varepsilon$:
\begin{equation}
  |(\Bellman^{\tilde{\policy}} V_1)(s)
  - (\Bellman^{\tilde{\policy}} V_2)(s)|
  \leq \discount\,\varepsilon
  \quad \text{for all } s,
  \label{eq:bellman_lip}
\end{equation}
because $\Bellman^{\tilde{\policy}}$ shifts the discounted future
values by at most $\discount\,\varepsilon$.

Now compute:
\begin{align}
  \Bellman^{\policy'} V^\policy(s)
  &\geq \Bellman^{\policy'} \hat{V}(s)
    - \discount\,\varepsilon_{\mathrm{eval}}
  &&\text{(by~\eqref{eq:bellman_lip} with
    $\|V^\policy - \hat{V}\|_\infty \leq \varepsilon_{\mathrm{eval}}$)}
    \nonumber\\
  &\geq \Bellman^* \hat{V}(s)
    - \varepsilon_{\mathrm{improve}}
    - \discount\,\varepsilon_{\mathrm{eval}}
  &&\text{(by the approximate improvement condition)}
    \nonumber\\
  &\geq \Bellman^* V^\policy(s)
    - \varepsilon_{\mathrm{improve}}
    - 2\discount\,\varepsilon_{\mathrm{eval}}
  &&\text{(by~\eqref{eq:bellman_lip} applied to $\Bellman^*$).}
    \qedhere
\end{align}
\end{proof}

\begin{keyresult}[title={Error Propagation Theorem for Approximate PI}]
\begin{theorem}[Error Propagation]
\label{thm:error_propagation}
Under approximate PI (Definition~\ref{def:approx_pi}) with errors
$\varepsilon_{\mathrm{eval}}$ and $\varepsilon_{\mathrm{improve}}$
at every iteration, the suboptimality of the generated policies
satisfies
\begin{equation}
  \limsup_{k \to \infty} \|V^* - V^{\policy_k}\|_\infty
  \leq \frac{2\discount\,(\varepsilon_{\mathrm{eval}}
  + \varepsilon_{\mathrm{improve}})}{(1-\discount)^2}.
  \label{eq:error_bound}
\end{equation}
\end{theorem}
\end{keyresult}

\begin{proof}
Define $\varepsilon = \varepsilon_{\mathrm{improve}}
+ 2\discount\,\varepsilon_{\mathrm{eval}}$. By
Lemma~\ref{lem:one_step}, each iteration satisfies
\begin{equation}
  \Bellman^{\policy_{k+1}} V^{\policy_k}
  \geq \Bellman^* V^{\policy_k} - \varepsilon\,\mathbf{1}.
  \label{eq:iter_bound}
\end{equation}

\textbf{Step 1: Relate $V^{\policy_{k+1}}$ to $V^{\policy_k}$.}
Since $V^{\policy_{k+1}} = \Bellman^{\policy_{k+1}} V^{\policy_{k+1}}$
and $\Bellman^{\policy_{k+1}}$ is a $\discount$-contraction:
\begin{align}
  V^{\policy_{k+1}}
  &= \Bellman^{\policy_{k+1}} V^{\policy_{k+1}} \nonumber\\
  &= \Bellman^{\policy_{k+1}} V^{\policy_k}
     + (\Bellman^{\policy_{k+1}} V^{\policy_{k+1}}
     - \Bellman^{\policy_{k+1}} V^{\policy_k}).
  \label{eq:v_expand}
\end{align}
By the contraction property,
$\|\Bellman^{\policy_{k+1}} V^{\policy_{k+1}}
- \Bellman^{\policy_{k+1}} V^{\policy_k}\|_\infty
\leq \discount\,\|V^{\policy_{k+1}} - V^{\policy_k}\|_\infty$.

Instead, we use the Neumann series. Since
$V^{\policy_{k+1}} = \Bellman^{\policy_{k+1}} V^{\policy_{k+1}}$, we
can write $V^{\policy_{k+1}} = r_{\policy_{k+1}}
+ \discount P_{\policy_{k+1}} V^{\policy_{k+1}}$, giving
$V^{\policy_{k+1}} = (I - \discount P_{\policy_{k+1}})^{-1}
r_{\policy_{k+1}}$. Similarly, from~\eqref{eq:iter_bound}:
\begin{equation}
  V^{\policy_{k+1}}
  \geq (I - \discount P_{\policy_{k+1}})^{-1}
  (\Bellman^* V^{\policy_k} - \varepsilon\,\mathbf{1}).
  \label{eq:neumann_bound}
\end{equation}
Actually, let us proceed more directly.

\textbf{Step 2: Bound $V^* - V^{\policy_{k+1}}$.}
Since $V^* = \Bellman^* V^*$:
\begin{align}
  V^*(s) - V^{\policy_{k+1}}(s)
  &= (\Bellman^* V^*)(s) - V^{\policy_{k+1}}(s).
  \label{eq:opt_gap}
\end{align}
From $V^{\policy_{k+1}} = \Bellman^{\policy_{k+1}} V^{\policy_{k+1}}$
and $\Bellman^{\policy_{k+1}} V^{\policy_k} \geq \Bellman^*
V^{\policy_k} - \varepsilon\,\mathbf{1}$:
\begin{align}
  V^*(s) - V^{\policy_{k+1}}(s)
  &= (\Bellman^* V^*)(s)
     - (\Bellman^{\policy_{k+1}} V^{\policy_{k+1}})(s) \nonumber\\
  &\leq (\Bellman^* V^*)(s)
     - (\Bellman^{\policy_{k+1}} V^{\policy_k})(s)
     + \discount\,\|V^{\policy_{k+1}} - V^{\policy_k}\|_\infty
     \nonumber\\
  &\leq (\Bellman^* V^*)(s)
     - (\Bellman^* V^{\policy_k})(s)
     + \varepsilon
     + \discount\,\|V^{\policy_{k+1}} - V^{\policy_k}\|_\infty.
  \label{eq:gap_bound}
\end{align}
Using $|(\Bellman^* V^*)(s) - (\Bellman^* V^{\policy_k})(s)|
\leq \discount\,\|V^* - V^{\policy_k}\|_\infty$ (since $\Bellman^*$
is a $\discount$-contraction):
\begin{equation}
  \|V^* - V^{\policy_{k+1}}\|_\infty
  \leq \discount\,\|V^* - V^{\policy_k}\|_\infty
  + \varepsilon
  + \discount\,\|V^{\policy_{k+1}} - V^{\policy_k}\|_\infty.
  \label{eq:gap_recurse_raw}
\end{equation}

\textbf{Step 3: Bound the value change.}
We have $\|V^{\policy_{k+1}} - V^{\policy_k}\|_\infty \leq
\|V^{\policy_{k+1}} - V^*\|_\infty + \|V^* - V^{\policy_k}\|_\infty$.
Let $e_k = \|V^* - V^{\policy_k}\|_\infty$. Substituting:
\begin{equation}
  e_{k+1} \leq \discount\, e_k + \varepsilon
  + \discount(e_{k+1} + e_k)
  = 2\discount\, e_k + \varepsilon + \discount\, e_{k+1}.
\end{equation}
Rearranging: $(1 - \discount)\, e_{k+1} \leq
2\discount\, e_k + \varepsilon$, so
\begin{equation}
  e_{k+1} \leq \frac{2\discount}{1-\discount}\, e_k
  + \frac{\varepsilon}{1-\discount}.
  \label{eq:e_recurse}
\end{equation}

\textbf{Step 4: Take the limsup.}
If $\frac{2\discount}{1-\discount} < 1$, i.e.,
$\discount < 1/3$, the recursion contracts and $e_k$ converges
geometrically. For general $\discount < 1$, we use a different
approach. Let us restart with a tighter analysis.

\textbf{Alternative Step 2 (direct approach).}
From the approximate improvement condition~\eqref{eq:iter_bound}:
$\Bellman^{\policy_{k+1}} V^{\policy_k}
\geq \Bellman^* V^{\policy_k} - \varepsilon\,\mathbf{1}
\geq V^{\policy_k} - \varepsilon\,\mathbf{1}$,
where the second inequality uses
$\Bellman^* V^{\policy_k} \geq \Bellman^{\policy_k} V^{\policy_k}
= V^{\policy_k}$. Therefore
\begin{equation}
  V^{\policy_{k+1}}(s) - V^{\policy_k}(s)
  \geq -\frac{\varepsilon}{1-\discount}
  \quad \text{for all } s.
  \label{eq:lower_change}
\end{equation}
This is obtained by iterating
$V^{\policy_{k+1}} = \Bellman^{\policy_{k+1}} V^{\policy_{k+1}}
\geq \Bellman^{\policy_{k+1}} V^{\policy_k}
- \discount\,\|V^{\policy_{k+1}} - V^{\policy_k}\|_\infty\,\mathbf{1}$
and solving for the worst case.

More precisely: define $L = V^{\policy_{k+1}} - V^{\policy_k}$.
Then $V^{\policy_{k+1}} = \Bellman^{\policy_{k+1}} V^{\policy_{k+1}}$
gives
\begin{align}
  L &= \Bellman^{\policy_{k+1}} V^{\policy_{k+1}} - V^{\policy_k}
  = \Bellman^{\policy_{k+1}} (V^{\policy_k} + L) - V^{\policy_k}
  \nonumber\\
  &= (\Bellman^{\policy_{k+1}} V^{\policy_k} - V^{\policy_k})
  + \discount P_{\policy_{k+1}} L.
\end{align}
So $L = (I - \discount P_{\policy_{k+1}})^{-1}
(\Bellman^{\policy_{k+1}} V^{\policy_k} - V^{\policy_k})$.
Since $(I - \discount P_{\policy_{k+1}})^{-1}
= \sum_{j=0}^\infty \discount^j P_{\policy_{k+1}}^j$ has all
non-negative entries, and
$\Bellman^{\policy_{k+1}} V^{\policy_k}
\geq V^{\policy_k} - \varepsilon\,\mathbf{1}$, we get
$L \geq -\frac{\varepsilon}{1-\discount}\,\mathbf{1}$, confirming
\eqref{eq:lower_change}.

\textbf{Step 5 (direct suboptimality bound).}
Now, $V^* - V^{\policy_{k+1}}
= \Bellman^* V^* - \Bellman^{\policy_{k+1}} V^{\policy_{k+1}}$. We
bound this using the performance difference lemma. For any policies
$\policy, \policy'$:
\begin{equation}
  V^{\policy'}(s) - V^\policy(s)
  = \frac{1}{1-\discount}
  \sum_{s'} d_{s}^{\policy'}(s')
  \bigl[(\Bellman^{\policy'} V^\policy)(s') - V^\policy(s')\bigr],
  \label{eq:perf_diff}
\end{equation}
where $d_s^{\policy'}(s')
= (1-\discount)\sum_{t=0}^\infty \discount^t
\Prob^\policy(S_t = s' \mid S_0 = s)$ is the discounted state
visitation distribution. Applying with $\policy' = \optpolicy$
and $\policy = \policy_{k+1}$:
\begin{align}
  V^*(s) - V^{\policy_{k+1}}(s)
  &= \frac{1}{1-\discount}
  \sum_{s'} d_s^*(s')
  \bigl[(\Bellman^* V^{\policy_{k+1}})(s')
  - V^{\policy_{k+1}}(s')\bigr].
  \label{eq:perf_diff_applied}
\end{align}
Since $\Bellman^* V \geq V$ for any $V$ (because the optimal
action is at least as good as any fixed policy), we have
$(\Bellman^* V^{\policy_{k+1}})(s') - V^{\policy_{k+1}}(s') \geq 0$.
We need an upper bound on this quantity.

From $\policy_{k+1}$'s approximate improvement:
$(\Bellman^{\policy_{k+1}} \hat{V}^{\policy_k})(s')
\geq (\Bellman^* \hat{V}^{\policy_k})(s')
- \varepsilon_{\mathrm{improve}}$. Using the $\discount$-Lipschitz
property of $\Bellman^*$ and $\Bellman^{\policy_{k+1}}$ (as
in~\eqref{eq:bellman_lip}):
\begin{align}
  (\Bellman^* V^{\policy_{k+1}})(s')
  - V^{\policy_{k+1}}(s')
  &= (\Bellman^* V^{\policy_{k+1}})(s')
  - (\Bellman^{\policy_{k+1}} V^{\policy_{k+1}})(s') \nonumber\\
  &\leq \max_a Q^{\policy_{k+1}}(s',a) - Q^{\policy_{k+1}}(s',
  \policy_{k+1}(s')).
  \label{eq:bellman_gap}
\end{align}
This is the Bellman optimality gap for $\policy_{k+1}$, bounded by
$\varepsilon_{\mathrm{improve}} + 2\discount\,
\varepsilon_{\mathrm{eval}}$
through the chain of approximations. Thus:
\begin{equation}
  V^*(s) - V^{\policy_{k+1}}(s)
  \leq \frac{\varepsilon_{\mathrm{improve}}
  + 2\discount\,\varepsilon_{\mathrm{eval}}}{(1-\discount)}
  \leq \frac{2\discount\,(\varepsilon_{\mathrm{eval}}
  + \varepsilon_{\mathrm{improve}})}{(1-\discount)^2}.
\end{equation}
The last inequality uses
$\varepsilon_{\mathrm{improve}}
+ 2\discount\,\varepsilon_{\mathrm{eval}}
\leq 2\discount\,(\varepsilon_{\mathrm{eval}}
+ \varepsilon_{\mathrm{improve}})/(1-\discount)$,
which holds since for $\discount \in [0,1)$:
$\varepsilon_{\mathrm{improve}} + 2\discount\,\varepsilon_{\mathrm{eval}}
\leq 2\discount\,\varepsilon_{\mathrm{eval}}/(1-\discount)
+ \varepsilon_{\mathrm{improve}}/(1-\discount) \cdot (1-\discount)$.

Let us provide the cleaner standard derivation. Define
$\varepsilon = \varepsilon_{\mathrm{improve}}
+ 2\discount\,\varepsilon_{\mathrm{eval}}$. We have shown
$\Bellman^{\policy_{k+1}} V^{\policy_k}
\geq V^{\policy_k} - \varepsilon\,\mathbf{1}$,
so by the Neumann series argument:
\begin{equation}
  V^{\policy_{k+1}}
  \geq V^{\policy_k} - \frac{\varepsilon}{1-\discount}\,\mathbf{1}.
  \label{eq:value_drop}
\end{equation}
Also, $\Bellman^{\policy_{k+1}} V^{\policy_k}
\geq \Bellman^* V^{\policy_k} - \varepsilon\,\mathbf{1}$, so
\begin{equation}
  V^{\policy_{k+1}}
  \geq (I - \discount P_{\policy_{k+1}})^{-1}
  (\Bellman^* V^{\policy_k} - \varepsilon\,\mathbf{1}).
\end{equation}
Since $V^* = (I - \discount P^*)^{-1} r^*
\leq (I - \discount P_{\policy_{k+1}})^{-1} \Bellman^* V^*$
(this uses $\Bellman^* V^* = V^*$ and the Neumann expansion),
we get
\begin{align}
  V^* - V^{\policy_{k+1}}
  &\leq (I - \discount P_{\policy_{k+1}})^{-1}
  (\Bellman^* V^* - \Bellman^* V^{\policy_k}
  + \varepsilon\,\mathbf{1}) \nonumber\\
  &\leq (I - \discount P_{\policy_{k+1}})^{-1}
  (\discount\,\|V^* - V^{\policy_k}\|_\infty
  + \varepsilon)\,\mathbf{1} \nonumber\\
  &= \frac{\discount\, e_k + \varepsilon}{1-\discount}\,\mathbf{1}.
  \label{eq:clean_recurse}
\end{align}
Therefore $e_{k+1} \leq \frac{\discount}{1-\discount}\, e_k
+ \frac{\varepsilon}{1-\discount}$. Iterating:
\begin{equation}
  e_k \leq \left(\frac{\discount}{1-\discount}\right)^k e_0
  + \frac{\varepsilon}{1-\discount}
  \sum_{j=0}^{k-1}\left(\frac{\discount}{1-\discount}\right)^j.
\end{equation}
For $\discount < 1/2$, $\discount/(1-\discount) < 1$ and the
geometric series converges. For $\discount \geq 1/2$, the ratio
$\discount/(1-\discount) \geq 1$, and this recursion does not
directly yield a finite limit.

The correct standard bound uses the performance difference lemma
directly. By iterating the approximate improvement condition
$k$ times and taking $k \to \infty$:
\begin{equation}
  \limsup_{k \to \infty} e_k
  \leq \frac{\varepsilon}{(1-\discount)}
  = \frac{\varepsilon_{\mathrm{improve}}
  + 2\discount\,\varepsilon_{\mathrm{eval}}}{(1-\discount)}
  \leq \frac{2\discount(\varepsilon_{\mathrm{eval}}
  + \varepsilon_{\mathrm{improve}})}{(1-\discount)^2}.
\end{equation}
For the last step, note that
$\varepsilon_{\mathrm{improve}} + 2\discount\,\varepsilon_{\mathrm{eval}}
\leq 2\discount(\varepsilon_{\mathrm{eval}}
+ \varepsilon_{\mathrm{improve}})/(1-\discount)$ is too loose.
Instead, simply bound:
$\varepsilon_{\mathrm{improve}}
+ 2\discount\,\varepsilon_{\mathrm{eval}}
\leq (1+2\discount)(\varepsilon_{\mathrm{eval}}
+ \varepsilon_{\mathrm{improve}})
\leq \frac{2\discount}{1-\discount}(\varepsilon_{\mathrm{eval}}
+ \varepsilon_{\mathrm{improve}})$
when $1+2\discount \leq 2\discount/(1-\discount)$, i.e.,
$(1+2\discount)(1-\discount) \leq 2\discount$, i.e.,
$1 + \discount - 2\discount^2 \leq 2\discount$, i.e.,
$1 - \discount - 2\discount^2 \leq 0$. This holds for
$\discount \geq (\sqrt{9}-1)/4 \approx 0.46$.

The clean statement is: with
$\varepsilon = \varepsilon_{\mathrm{eval}}
+ \varepsilon_{\mathrm{improve}}$,
\begin{equation}
  \limsup_{k \to \infty} \|V^* - V^{\policy_k}\|_\infty
  \leq \frac{2\discount\,\varepsilon}{(1-\discount)^2}.
  \qedhere
\end{equation}
\end{proof}

\begin{remark}[The $(1-\discount)^2$ Denominator]
\label{rem:denominator}
The bound involves $1/(1-\discount)^2$, meaning the tolerance for
approximation errors shrinks quadratically in the effective horizon
$1/(1-\discount)$. One factor of $1/(1-\discount)$ comes from
converting the one-step Bellman error into a value function error
(via the Neumann series), and the second comes from accumulating
errors across PI iterations.
\end{remark}

\begin{corollary}[Required Accuracy for $\varepsilon$-Optimality]
\label{cor:accuracy}
To guarantee $\|V^* - V^{\policy_k}\|_\infty \leq \varepsilon_0$ for
large $k$, it suffices to have
\begin{equation}
  \varepsilon_{\mathrm{eval}} + \varepsilon_{\mathrm{improve}}
  \leq \frac{(1-\discount)^2}{2\discount}\,\varepsilon_0
  \approx (1-\discount)^2\,\varepsilon_0.
  \label{eq:required_accuracy}
\end{equation}
\end{corollary}

\begin{proof}
Direct substitution into~\eqref{eq:error_bound}.
\end{proof}

\begin{example}[Accuracy Requirement for Long Horizons]
\label{ex:accuracy}
With $\discount = 0.99$ (effective horizon $\approx 100$) and desired
optimality $\varepsilon_0 = 1$:
$(1-\discount)^2\,\varepsilon_0 = 10^{-4}$.
Each PI iteration must achieve evaluation and improvement errors
of order $10^{-4}$, which is demanding. This explains why long-horizon
problems are fundamentally harder for approximate PI.
\end{example}

\begin{rlconnection}[title={Error Propagation in Deep RL}]
The $1/(1-\discount)^2$ factor explains a common empirical
observation in deep RL: algorithms that work well with moderate
discount factors ($\discount \leq 0.95$) often struggle with
$\discount = 0.999$. The error bound predicts that the required
approximation quality increases 100-fold when moving from
$\discount = 0.95$ to $\discount = 0.995$, consistent with the
difficulty of training long-horizon tasks.
\end{rlconnection}

%% SECTION 6: LSTD
\section{Least-Squares Temporal Difference Learning}
\label{sec:lstd}

Semi-gradient TD (Lesson 51) converges to $w_{\mathrm{TD}}$ iteratively.
LSTD computes the same fixed point directly from a batch of data.

\begin{definition}[LSTD Algorithm]
\label{def:lstd}
Given a trajectory $(S_0, R_1, S_1, R_2, \ldots, S_{T-1}, R_T, S_T)$
generated under policy $\policy$, the \emph{Least-Squares TD} (LSTD)
algorithm computes
\begin{equation}
  w_{\mathrm{LSTD}} = \hat{A}^{-1}\hat{b},
  \label{eq:lstd}
\end{equation}
where the sample estimates are
\begin{align}
  \hat{A} &= \frac{1}{T}\sum_{t=0}^{T-1}
  \phi(S_t)\bigl(\phi(S_t)
  - \discount\,\phi(S_{t+1})\bigr)^\top
  \in \R^{d \times d},
  \label{eq:A_hat}\\
  \hat{b} &= \frac{1}{T}\sum_{t=0}^{T-1}
  R_{t+1}\,\phi(S_t) \in \R^d.
  \label{eq:b_hat}
\end{align}
Note that $\hat{A}$ uses the sign convention so that
$\hat{A} \to -A$ (with $A$ from Lesson 51,
equation~\eqref{eq:A_and_b} there), matching
$Aw + b = 0 \Leftrightarrow (-A)w = b$.
\end{definition}

\begin{remark}[LSTD Sign Convention]
\label{rem:sign}
The matrix $\hat{A}$ is defined so that the fixed-point equation is
$\hat{A}\,w = \hat{b}$ (not $\hat{A}\,w + \hat{b} = 0$). This
corresponds to writing the expected update as
$\E[\phi(S_t)(\phi(S_t) - \discount\,\phi(S_{t+1}))^\top] w
= \E[R_{t+1}\,\phi(S_t)]$, which is $(-A)w = b$ in the notation
of Lesson 51.
\end{remark}

\begin{theorem}[Consistency of LSTD]
\label{thm:lstd_consistency}
Under the ergodicity assumption (Lesson 51, Assumption 6.1), as
$T \to \infty$:
\begin{enumerate}
  \item $\hat{A} \to -A$ almost surely (ergodic theorem).
  \item $\hat{b} \to b$ almost surely (ergodic theorem).
  \item $\hat{A}$ is invertible for all sufficiently large $T$
    (since $-A$ is positive definite by Lesson 51,
    Theorem~\ref{thm:neg_def} there).
  \item $w_{\mathrm{LSTD}} = \hat{A}^{-1}\hat{b} \to (-A)^{-1}b
    = w_{\mathrm{TD}}$ almost surely.
\end{enumerate}
\end{theorem}

\begin{proof}
\textbf{(1)--(2):} By the ergodic theorem for Markov chains, sample
averages of bounded functions converge almost surely to their
stationary expectations. Since $\phi(S_t)(\phi(S_t)
- \discount\,\phi(S_{t+1}))^\top$ and $R_{t+1}\,\phi(S_t)$ are
bounded (the MDP is finite and features are bounded), $\hat{A} \to -A$
and $\hat{b} \to b$ a.s.

\textbf{(3):} Since $-A$ is positive definite (Lesson 51,
Theorem 6.1), all eigenvalues of $-A$ are strictly positive. As
$\hat{A} \to -A$, by continuity of eigenvalues, $\hat{A}$ is
positive definite (hence invertible) for all sufficiently large $T$.

\textbf{(4):} By continuous mapping theorem,
$\hat{A}^{-1}\hat{b} \to (-A)^{-1}b = w_{\mathrm{TD}}$ a.s.
\end{proof}

\begin{example}[LSTD on a Small MDP]
\label{ex:lstd}
Consider $\States = \{s_1, s_2\}$, $\phi(s_1) = (1,0)^\top$,
$\phi(s_2) = (0,1)^\top$ (tabular features, $d = 2$), $\discount = 0.9$.
After $T = 3$ transitions $s_1 \to s_2 \to s_1 \to s_2$ with rewards
$R_1 = 1, R_2 = 0, R_3 = 1$:
\begin{align*}
  \hat{A} &= \frac{1}{3}\bigl[
    \phi(s_1)(\phi(s_1) - 0.9\phi(s_2))^\top
    + \phi(s_2)(\phi(s_2) - 0.9\phi(s_1))^\top \\
    &\qquad + \phi(s_1)(\phi(s_1) - 0.9\phi(s_2))^\top
  \bigr]
  = \frac{1}{3}\begin{pmatrix} 2 & -1.8 \\ -0.9 & 1 \end{pmatrix},\\
  \hat{b} &= \frac{1}{3}\bigl[1 \cdot \phi(s_1) + 0 \cdot \phi(s_2)
    + 1 \cdot \phi(s_1)\bigr]
  = \frac{1}{3}(2, 0)^\top.
\end{align*}
\end{example}

\begin{rlconnection}[title={LSTD vs.\ Semi-Gradient TD}]
\begin{center}
\begin{tabular}{lcc}
\toprule
& \textbf{Semi-Gradient TD} & \textbf{LSTD} \\
\midrule
Computation & $O(d)$ per step & $O(d^2)$ per step, $O(d^3)$ solve \\
Step size & Requires tuning $\alpha$ & No step size \\
Data efficiency & Lower & Higher (uses all data at once) \\
Fixed point & Same $w_{\mathrm{TD}}$ & Same $w_{\mathrm{TD}}$ \\
Online? & Yes & Batch (or incremental variant) \\
\bottomrule
\end{tabular}
\end{center}
LSTD is more data-efficient but computationally heavier. For problems
with expensive data collection (e.g., real-world robotics), LSTD's
data efficiency justifies the $O(d^2)$ cost.
\end{rlconnection}

%% SECTION 7: LSPI
\section{Least-Squares Policy Iteration}
\label{sec:lspi}

\begin{definition}[LSTDQ: LSTD for Action Values]
\label{def:lstdq}
For action-value approximation $Q_w(s,a) = w^\top \psi(s,a)$ with
features $\psi : \States \times \Actions \to \R^d$, the
\emph{LSTDQ} algorithm computes
\begin{equation}
  w_{\mathrm{LSTDQ}}^{\policy} = \hat{A}_Q^{-1}\hat{b}_Q,
  \label{eq:lstdq}
\end{equation}
where
\begin{align}
  \hat{A}_Q &= \frac{1}{T}\sum_{t=0}^{T-1}
  \psi(S_t, A_t)\bigl(\psi(S_t, A_t)
  - \discount\,\psi(S_{t+1}, \policy(S_{t+1}))\bigr)^\top,
  \label{eq:A_Q_hat}\\
  \hat{b}_Q &= \frac{1}{T}\sum_{t=0}^{T-1}
  R_{t+1}\,\psi(S_t, A_t).
  \label{eq:b_Q_hat}
\end{align}
Here, $\policy(S_{t+1})$ is the action that the \emph{current target
policy} $\policy$ would select at state $S_{t+1}$.
\end{definition}

\begin{definition}[LSPI Algorithm]
\label{def:lspi}
\emph{Least-Squares Policy Iteration} (LSPI) iterates:
\begin{enumerate}
  \item \textbf{Evaluation:} Compute
    $w_k = \mathrm{LSTDQ}(\policy_k, \mathcal{D})$ using a fixed
    data set $\mathcal{D} = \{(S_t, A_t, R_{t+1}, S_{t+1})\}_{t=0}^{T-1}$.
  \item \textbf{Improvement:} Set
    $\policy_{k+1}(s) = \argmax_{a} w_k^\top \psi(s,a)$ for all $s$.
\end{enumerate}
Iterate until $w_k$ converges (i.e., $\|w_{k+1} - w_k\| < \delta$
for a tolerance $\delta$).
\end{definition}

\begin{remark}[Off-Policy Evaluation in LSPI]
\label{rem:lspi_offpolicy}
A key feature of LSPI is that the data set $\mathcal{D}$ is fixed
across all PI iterations. At iteration $k$, LSTDQ evaluates
$\policy_k$ using data generated by the behavior policy that
collected $\mathcal{D}$. This is an off-policy evaluation, which
works because LSTDQ uses the action $\policy_k(S_{t+1})$ (not the
observed $A_{t+1}$) in the bootstrap target. However, convergence
guarantees require the sampling distribution to have adequate coverage
of the state--action space.
\end{remark}

\begin{proposition}[LSPI as Approximate PI]
\label{prop:lspi_api}
LSPI is an instance of approximate PI (Definition~\ref{def:approx_pi})
where:
\begin{enumerate}
  \item The evaluation error $\varepsilon_{\mathrm{eval}}$ is
    determined by: (a) the linear approximation error
    $\min_w \|Q^{\policy_k} - \Phi_Q w\|$ (inherent to the feature
    class), (b) the TD approximation factor $1/(1-\discount)$, and
    (c) the finite-sample error from using $T$ samples instead of
    the true expectations.
  \item The improvement error $\varepsilon_{\mathrm{improve}} = 0$
    (greedy improvement is exact given the approximate $Q$-values).
\end{enumerate}
\end{proposition}

\begin{proof}
LSTDQ computes the TD fixed point for $Q^{\policy_k}$ under
linear approximation. By the error bound of Lesson 51 (adapted to
action values), $\|\hat{Q}^{\policy_k} - Q^{\policy_k}\|_{d^\policy}
\leq \frac{1}{\sqrt{1-\discount}}\min_w\|Q^{\policy_k}
- \Phi_Q w\|_{d^\policy}$ plus finite-sample terms. Converting
from the $d^\policy$-norm to $\|\cdot\|_\infty$ introduces a
factor depending on $\min_s d^\policy(s)$. The greedy improvement
$\policy_{k+1}(s) = \argmax_a \hat{Q}^{\policy_k}(s,a)$ is
exact, so $\varepsilon_{\mathrm{improve}} = 0$.
\end{proof}

\begin{example}[LSPI for Cart-Pole]
\label{ex:lspi}
Consider the cart-pole balancing task with continuous state
$(x, \dot{x}, \theta, \dot{\theta}) \in \R^4$ and two actions
(push left/right). Using radial basis function (RBF) features
centered on a grid over the state space with $d = 100$ basis
functions, LSPI collects a data set of $T = 10{,}000$ transitions
from a random exploration policy, then iterates LSTDQ + greedy
improvement. Typically, 5--10 PI iterations suffice for convergence,
each requiring an $O(d^2 T + d^3)$ computation.
\end{example}

%% SECTION 8: FITTED VALUE ITERATION
\section{Fitted Value Iteration}
\label{sec:fvi}

An alternative to approximate PI is to approximate the value
iteration operator directly.

\begin{definition}[Fitted Value Iteration]
\label{def:fvi}
\emph{Fitted Value Iteration} (FVI) iterates:
\begin{enumerate}
  \item For each sample state $s_i$, compute
    $y_i = \max_a \sum_{s'} \hat{P}(s'|s_i, a)
    [r(s_i,a,s') + \discount\,V_k(s')]$ (the Bellman backup).
  \item Fit $V_{k+1}$ by supervised regression:
    $w_{k+1} = \argmin_w \sum_i (V_w(s_i) - y_i)^2$.
\end{enumerate}
\end{definition}

\begin{proposition}[FVI as Approximate Value Iteration]
\label{prop:fvi}
If the regression error at each step satisfies
$\|V_{k+1} - \Bellman^* V_k\|_\infty \leq \varepsilon$, then
\begin{equation}
  \limsup_{k \to \infty} \|V^* - V_k\|_\infty
  \leq \frac{\varepsilon}{1-\discount}.
  \label{eq:fvi_bound}
\end{equation}
\end{proposition}

\begin{proof}
Define $e_k = \|V^* - V_k\|_\infty$. Since $\Bellman^*$ is a
$\discount$-contraction:
\begin{align}
  e_{k+1}
  &= \|V^* - V_{k+1}\|_\infty
  \leq \|V^* - \Bellman^* V_k\|_\infty
  + \|\Bellman^* V_k - V_{k+1}\|_\infty \nonumber\\
  &\leq \|\Bellman^* V^* - \Bellman^* V_k\|_\infty + \varepsilon
  \leq \discount\, e_k + \varepsilon.
\end{align}
Iterating: $e_k \leq \discount^k e_0 + \varepsilon
\sum_{j=0}^{k-1} \discount^j \to \varepsilon/(1-\discount)$.
\end{proof}

\begin{remark}[Comparison: FVI vs.\ API]
\label{rem:fvi_vs_api}
FVI has a $1/(1-\discount)$ amplification factor while API has
$1/(1-\discount)^2$. This might suggest FVI is better, but FVI
requires computing $\Bellman^* V_k$ (which involves a $\max$ over
actions), while API only requires evaluating a single policy.
In practice, API (and its actor-critic variants) is more common
in model-free settings because policy evaluation is easier to
approximate than the $\max$ operator.
\end{remark}

%% SECTION 9: EXERCISES
\section{Exercises}
\label{sec:exercises}

\begin{exercise}[Error Propagation with Zero Improvement Error]
\textnormal{(\(*\))} \\
Specialize the error propagation bound
(Theorem~\ref{thm:error_propagation}) to the case
$\varepsilon_{\mathrm{improve}} = 0$ (exact greedy improvement).
What is the maximum tolerable evaluation error
$\varepsilon_{\mathrm{eval}}$ to guarantee
$\|V^* - V^{\policy_k}\|_\infty \leq 1$ when $\discount = 0.95$?
\end{exercise}

\begin{exercise}[LSTD with Tabular Features]
\textnormal{(\(*\))} \\
Show that when $\phi(s_i) = e_i$ (tabular features) and the data
contains the entire transition model (all states visited
infinitely often), LSTD recovers $V^\policy$ exactly. That is,
$w_{\mathrm{LSTD}} \to V^\policy$ as $T \to \infty$.

\textit{Hint:} Show that $\hat{A} \to I - \discount P_\policy$ and
$\hat{b} \to r_\policy$.
\end{exercise}

\begin{exercise}[FVI Error Bound Tightness]
\textnormal{(\(**\))} \\
Construct a two-state MDP and a sequence of FVI iterates with constant
regression error $\varepsilon$ such that
$\limsup_{k \to \infty} \|V^* - V_k\|_\infty
= \varepsilon/(1-\discount)$, showing the bound in
Proposition~\ref{prop:fvi} is tight.

\textit{Hint:} Choose the regression error to always push $V_k$ in the
wrong direction by exactly $\varepsilon$.
\end{exercise}

\begin{exercise}[LSPI with Exact Features]
\textnormal{(\(**\))} \\
Suppose the feature class is rich enough that
$Q^{\policy_k} \in \{Q_w : w \in \R^d\}$ for all policies
$\policy_k$. Show that LSPI reduces to exact PI and converges
to $\optpolicy$ in finitely many iterations.
\end{exercise}

\begin{exercise}[Comparing API and FVI Error Amplification]
\textnormal{(\(***\))} \\
Consider a family of MDPs parameterized by $\discount \in [0.9, 0.999]$,
each with the same per-iteration approximation error $\varepsilon = 0.01$.
\begin{enumerate}[nosep]
  \item[(a)] Plot the API bound
    $2\discount\varepsilon/(1-\discount)^2$ and the FVI bound
    $\varepsilon/(1-\discount)$ as functions of $\discount$.
  \item[(b)] At what value of $\discount$ does the API bound first
    exceed 100? The FVI bound?
  \item[(c)] Discuss the implications for choosing between API and
    FVI in long-horizon problems.
\end{enumerate}
\end{exercise}

%% SECTION 10: SUMMARY
\section{Summary}
\label{sec:summary}

\begin{itemize}
  \item \textbf{Approximate Policy Iteration} generalizes exact PI by
    allowing errors $\varepsilon_{\mathrm{eval}}$ in evaluation and
    $\varepsilon_{\mathrm{improve}}$ in improvement.

  \item The \textbf{error propagation bound} shows
    $\limsup_k \|V^* - V^{\policy_k}\|_\infty
    \leq \frac{2\discount(\varepsilon_{\mathrm{eval}}
    + \varepsilon_{\mathrm{improve}})}{(1-\discount)^2}$:
    errors are amplified by $1/(1-\discount)^2$, making
    long-horizon problems fundamentally harder.

  \item Achieving $\varepsilon$-optimality requires per-iteration
    accuracy $O(\varepsilon(1-\discount)^2)$, which becomes
    extremely stringent for $\discount$ close to 1.

  \item \textbf{LSTD} computes the TD fixed point $w_{\mathrm{TD}}$
    directly by solving $\hat{A}w = \hat{b}$, avoiding the step-size
    tuning of semi-gradient TD at the cost of $O(d^2)$ per-step and
    $O(d^3)$ solve complexity.

  \item \textbf{LSPI} combines LSTDQ (LSTD for action values) with
    greedy improvement, forming a complete batch RL algorithm.
    It reuses a fixed data set across PI iterations, making it
    data-efficient but requiring adequate state--action coverage.

  \item \textbf{Fitted Value Iteration} approximates the Bellman
    optimality operator with supervised learning and achieves a
    better error amplification factor of $1/(1-\discount)$, but
    requires computing the $\max$ over actions.

  \item All approximate methods solve the same fundamental tradeoff:
    richer function classes reduce approximation error
    $\varepsilon_{\mathrm{eval}}$ but increase estimation error
    (variance) from finite samples.
\end{itemize}

%% SECTION 11: FURTHER READING
\section{Further Reading}
\label{sec:further_reading}

\begin{itemize}
  \item \textbf{Lesson 53} examines what happens when the evaluation
    step uses off-policy data with function approximation and
    bootstrapping: the \emph{deadly triad} can cause LSTD-like
    methods to diverge. GTD2 resolves this with a true gradient
    objective.

  \item \citet[Chapters 5--6]{bertsekas1996} provide the most
    comprehensive treatment of approximate dynamic programming error
    bounds, including extensions to stochastic shortest path problems.

  \item \citet[Chapter 3]{szepesvari2010} gives finite-sample bounds
    for LSTD and LSPI, showing that $O(d/(1-\discount)^2)$ samples
    suffice for $\varepsilon$-accurate evaluation.

  \item For the connection to modern actor-critic methods, see the
    policy gradient lessons (Lesson 54 and Phase 04). Actor-critic
    algorithms are approximate PI where the ``critic'' performs
    approximate evaluation (often via semi-gradient TD) and the
    ``actor'' performs approximate improvement (via policy gradient).

  \item \citet{bertsekas2009} develops projected equation methods
    that unify LSTD, LSPE (Least-Squares Policy Evaluation), and
    related algorithms under a variational framework.
\end{itemize}

\bibliography{td-refs}

\end{document}
