\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 47: TD(0) and n-Step TD}}
\fancyhead[R]{\thepage}
\title{Lesson 47: TD(0) and n-Step TD}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% SECTION 1: MOTIVATION AND OVERVIEW
\section{Motivation and Overview}
\label{sec:motivation}

Consider an agent navigating a large gridworld with thousands of states,
receiving a reward only upon reaching a distant goal. After each episode,
the Monte Carlo approach updates every visited state using the realized
return---but those returns are sums of dozens of random transitions,
producing estimates with enormous variance. Worse, the agent must wait
until the episode terminates before any learning occurs. In continuing
(non-episodic) environments, Monte Carlo methods cannot be applied at all.

Temporal-difference (TD) learning resolves both problems by
\emph{bootstrapping}: updating value estimates using other value estimates
rather than waiting for a complete return. The simplest variant, TD(0),
adjusts the estimate of each state immediately after a single transition,
using the observed reward and the current estimate of the next state as a
proxy for the full future return. The $n$-step extension interpolates
between the one-step bootstrap of TD(0) and the full-episode return of
Monte Carlo, providing a tunable bias--variance tradeoff. The tools
developed in this lesson will let us prove that these bootstrapped updates
converge to the true value function despite the apparent circularity of
updating estimates from estimates.

\begin{rlconnection}[title={Where TD(0) and $n$-Step TD Appear}]
\begin{itemize}[nosep]
  \item \textbf{TD(0)} is the prediction engine inside SARSA and
    Q-learning (Lesson 49), where the TD error drives value updates.
  \item \textbf{$n$-step returns} underpin Generalized Advantage
    Estimation (GAE), the variance reduction technique used by PPO and
    TRPO for policy gradient estimation.
  \item \textbf{TD errors} form the building blocks of eligibility traces
    and TD($\lambda$) (Lesson 48), connecting the forward and backward
    views of credit assignment.
  \item \textbf{Bias--variance analysis} of TD targets guides
    hyperparameter selection ($n$, $\lambda$) in every practical deep RL
    implementation.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
Markov decision processes: state-value function $\Vpi(s) =
\E_\policy[\sum_{t=0}^\infty \discount^t R_{t+1} \mid S_0 = s]$,
Bellman expectation equation, Bellman operator $\Bellman^\policy$ and its
$\discount$-contraction property in $\|\cdot\|_\infty$ (Lessons 45--46).
Stochastic approximation: the Robbins--Monro theorem with martingale
noise, step-size conditions $\sum \alpha_n = \infty$,
$\sum \alpha_n^2 < \infty$ (Lesson 31). Conditional expectation and the
law of total variance (Lesson 7).

\paragraph{Learning objectives.}
\begin{itemize}
  \item \emph{State} the TD(0) update rule and explain how it differs
    from the Monte Carlo update through bootstrapping.
  \item \emph{Prove} that the TD error $\delta_t^\policy$ under the true
    value function is a martingale difference sequence.
  \item \emph{Derive} the convergence of TD(0) by casting it as a
    stochastic approximation of the Bellman equation.
  \item \emph{Analyze} the bias--variance tradeoff between TD(0) and
    Monte Carlo targets.
  \item \emph{Define} the $n$-step return and \emph{prove} the
    telescoping decomposition into TD errors.
  \item \emph{Prove} convergence of $n$-step TD for any fixed $n \geq 1$.
\end{itemize}

%% SECTION 2: REFERENCES
\section{References}
\label{sec:references}

\begin{itemize}[leftmargin=2em]
  \item \citet[Chapter 6]{suttonbarto2018} --- Introduction to TD
    prediction; TD(0) algorithm and examples.
  \item \citet[Chapter 7]{suttonbarto2018} --- $n$-step bootstrapping
    methods and the bias--variance spectrum.
  \item \citet[Chapter 4]{bertsekas1996} --- Convergence of temporal
    difference methods via stochastic approximation.
  \item \citet[Section 4]{szepesvari2010} --- Compact treatment of TD
    convergence under the stochastic approximation framework.
  \item \citet{tsitsiklis1994} --- Asynchronous stochastic approximation
    and the convergence machinery for TD and Q-learning.
  \item \citet{sutton1988} --- The original paper formalizing TD learning
    and the TD($\lambda$) family.
\end{itemize}

%% SECTION 3: MONTE CARLO PREDICTION BASELINE
\section{Monte Carlo Prediction as Baseline}
\label{sec:mc}

We begin with Monte Carlo (MC) prediction to establish the baseline
against which TD methods will be compared. Throughout this lesson, we
work with a finite MDP $(\States, \Actions, \Trans, r, \discount)$ with
$\discount \in [0,1)$, a stationary policy $\policy$, and the return
$\return = \sum_{k=0}^\infty \discount^k R_{t+k+1}$.

\begin{definition}[Monte Carlo Update]
\label{def:mc_update}
Given an observed return $\return$ from state $S_t$, the
\emph{every-visit MC update} is
\begin{equation}
  V(S_t) \leftarrow V(S_t) + \alpha\bigl[\return - V(S_t)\bigr].
  \label{eq:mc_update}
\end{equation}
\end{definition}

\begin{proposition}[Convergence of MC Prediction]
\label{prop:mc_convergence}
Suppose every state $s \in \States$ is visited infinitely often and the
step sizes satisfy $\sum_n \alpha_n(s) = \infty$,
$\sum_n \alpha_n^2(s) < \infty$ for each $s$. Then the MC
update~\eqref{eq:mc_update} converges to $\Vpi(s)$ almost surely.
\end{proposition}

\begin{proof}
Since $\E_\policy[\return \mid S_t = s] = \Vpi(s)$, the MC update is a
Robbins--Monro iteration with driving function
$h(V(s)) = \Vpi(s) - V(s)$ and noise $w_t = \return - \Vpi(s)$. The
noise has zero conditional mean: $\E[w_t \mid S_t = s] = 0$. The mapping
$h$ has unique root $V(s) = \Vpi(s)$, and the step-size conditions are
satisfied by assumption. By the Robbins--Monro theorem (Lesson 31),
$V_n(s) \to \Vpi(s)$ almost surely.
\end{proof}

\begin{lemma}[Variance of the MC Return]
\label{lem:mc_variance}
For deterministic rewards $r(s,a,s')$, the variance of the return
satisfies
\begin{equation}
  \Var_\policy(\return \mid S_t = s)
  = \discount^2 \E_\policy\bigl[\Var_\policy(G_{t+1} \mid S_{t+1})
    \,\big|\, S_t = s\bigr]
  + \Var_\policy\!\bigl(r(S_t, A_t, S_{t+1})
    + \discount \Vpi(S_{t+1}) \,\big|\, S_t = s\bigr).
  \label{eq:mc_variance}
\end{equation}
\end{lemma}

\begin{proof}
Write $\return = R_{t+1} + \discount G_{t+1}$. By the law of total
variance conditioned on $(S_t, A_t, S_{t+1})$:
\[
  \Var_\policy(\return \mid S_t)
  = \E_\policy\bigl[\Var_\policy(\return \mid S_t, A_t, S_{t+1})
    \,\big|\, S_t\bigr]
  + \Var_\policy\!\bigl(\E_\policy[\return \mid S_t, A_t, S_{t+1}]
    \,\big|\, S_t\bigr).
\]
For deterministic rewards, the inner conditional variance reduces to
$\discount^2 \Var_\policy(G_{t+1} \mid S_{t+1})$ and the inner
conditional expectation is $r(S_t, A_t, S_{t+1}) + \discount \Vpi(S_{t+1})$,
yielding~\eqref{eq:mc_variance}.
\end{proof}

\begin{warning}[title={MC Variance Accumulates Over the Entire Trajectory}]
Equation~\eqref{eq:mc_variance} shows that $\Var_\policy(\return \mid
S_t)$ includes a $\discount^2$-weighted contribution from
$\Var_\policy(G_{t+1} \mid S_{t+1})$, which in turn depends on
$\Var_\policy(G_{t+2} \mid S_{t+2})$, and so on. The variance
accumulates from \emph{all} future time steps, growing substantially
when $\discount$ is close to 1. This motivates the bootstrapping
approach of TD methods.
\end{warning}

The recursive structure of~\eqref{eq:mc_variance} makes clear that the
MC target $\return$ incorporates randomness from the entire trajectory.
We now introduce TD(0), which eliminates all future variance beyond
a single transition.

%% SECTION 4: TD(0) PREDICTION
\section{TD(0) Prediction}
\label{sec:td0}

\begin{definition}[TD Error]
\label{def:td_error}
Given a value estimate $V$, the \emph{temporal-difference error} at
time $t$ is
\begin{equation}
  \delta_t = R_{t+1} + \discount V(S_{t+1}) - V(S_t).
  \label{eq:td_error}
\end{equation}
When $V = \Vpi$, we write
$\delta_t^\policy = R_{t+1} + \discount \Vpi(S_{t+1}) - \Vpi(S_t)$.
\end{definition}

\begin{definition}[TD(0) Update Rule]
\label{def:td0_update}
The \emph{TD(0) update} modifies the value estimate upon observing the
transition $(S_t, A_t, R_{t+1}, S_{t+1})$:
\begin{equation}
  V(S_t) \leftarrow V(S_t) + \alpha\,\delta_t
  = V(S_t) + \alpha\bigl[R_{t+1} + \discount V(S_{t+1}) - V(S_t)\bigr].
  \label{eq:td0_update}
\end{equation}
\end{definition}

\begin{remark}[Bootstrapping]
\label{rem:bootstrapping}
The key distinction from MC is that TD(0) uses the
\emph{bootstrapped target} $R_{t+1} + \discount V(S_{t+1})$ instead
of the full return $\return$. This replaces all future randomness
(from $G_{t+1}$) with the current estimate $V(S_{t+1})$, reducing
variance at the cost of introducing bias when $V \neq \Vpi$.
\end{remark}

\subsection{Properties of the TD Error}
\label{subsec:td_error_properties}

\begin{lemma}[Unbiasedness of the TD Error]
\label{lem:td_unbiased}
Under policy $\policy$, the TD error computed with the true value
function has zero conditional expectation:
\begin{equation}
  \E_\policy\bigl[\delta_t^\policy \mid S_t = s\bigr] = 0
  \quad \text{for all } s \in \States.
  \label{eq:td_unbiased}
\end{equation}
That is, $\{\delta_t^\policy\}$ is a martingale difference sequence with
respect to the filtration generated by $(S_0, A_0, R_1, S_1, \ldots)$.
\end{lemma}

\begin{proof}
Compute the conditional expectation directly:
\begin{align}
  \E_\policy\bigl[\delta_t^\policy \mid S_t = s\bigr]
  &= \E_\policy\bigl[R_{t+1} + \discount \Vpi(S_{t+1})
     - \Vpi(S_t) \mid S_t = s\bigr] \nonumber\\
  &= \E_\policy\bigl[R_{t+1} + \discount \Vpi(S_{t+1})
     \mid S_t = s\bigr] - \Vpi(s) \nonumber\\
  &= \sum_{a} \policy(a \mid s) \sum_{s'} \Trans(s' \mid s, a)
     \bigl[r(s,a,s') + \discount \Vpi(s')\bigr] - \Vpi(s) \nonumber\\
  &= (\Bellman^\policy \Vpi)(s) - \Vpi(s)
   = \Vpi(s) - \Vpi(s) = 0,
  \label{eq:td_unbiased_proof}
\end{align}
where the last equality uses the fact that $\Vpi$ is the fixed point
of $\Bellman^\policy$.
\end{proof}

\begin{keyresult}[title={Return as Telescoping Sum of TD Errors}]
\begin{lemma}[Telescoping Decomposition]
\label{lem:return_telescope}
Under the true value function $\Vpi$, the prediction error decomposes as
\begin{equation}
  \return - \Vpi(S_t) = \sum_{k=0}^{\infty} \discount^k
  \delta_{t+k}^\policy.
  \label{eq:return_telescope}
\end{equation}
\end{lemma}
\end{keyresult}

\begin{proof}
Expand the right-hand side:
\begin{align}
  \sum_{k=0}^{\infty} \discount^k \delta_{t+k}^\policy
  &= \sum_{k=0}^{\infty} \discount^k
     \bigl[R_{t+k+1} + \discount \Vpi(S_{t+k+1})
     - \Vpi(S_{t+k})\bigr] \nonumber\\
  &= \sum_{k=0}^{\infty} \discount^k R_{t+k+1}
     + \sum_{k=0}^{\infty} \discount^{k+1} \Vpi(S_{t+k+1})
     - \sum_{k=0}^{\infty} \discount^k \Vpi(S_{t+k}).
     \label{eq:telescope_expand}
\end{align}
Re-index the second sum with $j = k + 1$:
\begin{align}
  &= \return
     + \sum_{j=1}^{\infty} \discount^j \Vpi(S_{t+j})
     - \sum_{k=0}^{\infty} \discount^k \Vpi(S_{t+k}) \nonumber\\
  &= \return
     + \sum_{k=1}^{\infty} \discount^k \Vpi(S_{t+k})
     - \Vpi(S_t)
     - \sum_{k=1}^{\infty} \discount^k \Vpi(S_{t+k}) \nonumber\\
  &= \return - \Vpi(S_t),
  \label{eq:telescope_final}
\end{align}
where the telescoping cancellation eliminates all intermediate value
terms. The boundary term $\lim_{k \to \infty} \discount^k \Vpi(S_{t+k})
= 0$ since $\discount < 1$ and $\Vpi$ is bounded on the finite state
space.
\end{proof}

\begin{rlconnection}[title={TD Errors as Prediction Errors}]
Lemma~\ref{lem:return_telescope} shows that the return minus the value
--- the total prediction error --- equals the sum of discounted one-step
TD errors. This decomposition is the mathematical foundation for:
\begin{itemize}[nosep]
  \item The forward view of TD($\lambda$) (Lesson 48), where
    exponentially weighted sums of TD errors define the $\lambda$-return.
  \item Generalized Advantage Estimation (GAE), which uses truncated
    versions of this sum for low-variance policy gradient estimation.
\end{itemize}
\end{rlconnection}

\subsection{Convergence of TD(0)}
\label{subsec:td0_convergence}

We now prove the central result of this lesson: TD(0) converges to the
true value function almost surely.

\begin{assumption}[TD(0) Convergence Conditions]
\label{ass:td0}
We assume:
\begin{enumerate}[nosep]
  \item The MDP has finite state and action spaces.
  \item The policy $\policy$ is stationary.
  \item Every state $s \in \States$ is visited infinitely often.
  \item The step sizes satisfy $\sum_{n=0}^\infty \alpha_n(s) = \infty$
    and $\sum_{n=0}^\infty \alpha_n^2(s) < \infty$ for each
    $s \in \States$.
\end{enumerate}
\end{assumption}

\begin{keyresult}[title={Convergence of TD(0)}]
\begin{theorem}[Convergence of TD(0)]
\label{thm:td0_convergence}
Under Assumption~\ref{ass:td0}, the TD(0)
update~\eqref{eq:td0_update} converges to $\Vpi$ almost surely:
\begin{equation}
  V_n(s) \cas \Vpi(s) \quad \text{for all } s \in \States.
\end{equation}
\end{theorem}
\end{keyresult}

\begin{proof}
We cast TD(0) as a stochastic approximation of the Bellman equation and
verify all conditions of the Robbins--Monro theorem
(Lesson 31, Theorem 3.1). The proof follows
\citet{tsitsiklis1994} and \citet{jaakkola1994}.

\textbf{Step 1: Express as stochastic approximation.}
Fix a state $s$ and let $n_k(s)$ denote the time of the $k$-th visit
to $s$. At these visits, the TD(0) update reads
\begin{equation}
  V_{k+1}(s) = V_k(s) + \alpha_k(s)\bigl[R_{n_k+1}
  + \discount V_k(S_{n_k+1}) - V_k(s)\bigr].
  \label{eq:td0_sa}
\end{equation}
Decompose the update target into its expected value plus noise:
\begin{equation}
  R_{n_k+1} + \discount V_k(S_{n_k+1})
  = (\Bellman^\policy V_k)(s) + w_k(s),
  \label{eq:td0_decomp}
\end{equation}
where $w_k(s) = R_{n_k+1} + \discount V_k(S_{n_k+1})
- (\Bellman^\policy V_k)(s)$.

\textbf{Step 2: Verify martingale noise.}
Let $\mathcal{F}_k$ denote the filtration containing all information
up to the $k$-th visit to $s$ (including $V_k$ and $S_{n_k} = s$).
Then:
\begin{align}
  \E[w_k(s) \mid \mathcal{F}_k]
  &= \E\bigl[R_{n_k+1} + \discount V_k(S_{n_k+1})
     \mid S_{n_k} = s, V_k\bigr]
     - (\Bellman^\policy V_k)(s) \nonumber\\
  &= \sum_a \policy(a \mid s) \sum_{s'} \Trans(s' \mid s, a)
     \bigl[r(s,a,s') + \discount V_k(s')\bigr]
     - (\Bellman^\policy V_k)(s) \nonumber\\
  &= (\Bellman^\policy V_k)(s) - (\Bellman^\policy V_k)(s)
   = 0.
  \label{eq:td0_noise_mg}
\end{align}
Thus $\{w_k(s)\}$ is a martingale difference sequence with respect
to $\{\mathcal{F}_k\}$.

\textbf{Step 3: Identify the expected update direction.}
Substituting~\eqref{eq:td0_decomp} into~\eqref{eq:td0_sa}:
\begin{equation}
  V_{k+1}(s) = V_k(s) + \alpha_k(s)
  \bigl[(\Bellman^\policy V_k)(s) - V_k(s) + w_k(s)\bigr].
  \label{eq:td0_sa_clean}
\end{equation}
The deterministic driving function is
$h(V)(s) = (\Bellman^\policy V)(s) - V(s)$, which has the unique root
$V = \Vpi$ (the fixed point of $\Bellman^\policy$).

\textbf{Step 4: Verify stability via contraction.}
Since $\Bellman^\policy$ is a $\discount$-contraction in
$\|\cdot\|_\infty$ (Lessons 45--46), for any $\alpha \in (0, 1]$:
\begin{align}
  \|V + \alpha\, h(V) - \Vpi\|_\infty
  &= \|(1 - \alpha) V + \alpha\, \Bellman^\policy V - \Vpi\|_\infty
     \nonumber\\
  &= \|(1 - \alpha)(V - \Vpi)
     + \alpha(\Bellman^\policy V - \Bellman^\policy \Vpi)\|_\infty
     \nonumber\\
  &\leq (1 - \alpha)\|V - \Vpi\|_\infty
     + \alpha\,\discount\,\|V - \Vpi\|_\infty \nonumber\\
  &= \bigl(1 - \alpha(1 - \discount)\bigr)\|V - \Vpi\|_\infty.
  \label{eq:td0_contraction}
\end{align}
Since $\alpha(1 - \discount) > 0$, the expected update contracts the
error toward zero, confirming that $h$ drives iterates toward $\Vpi$.

\textbf{Step 5: Bound the noise variance.}
By finiteness of the MDP, there exists $R_{\max} < \infty$ such that
$|R_{t+1}| \leq R_{\max}$. The value iterates remain bounded
throughout: $\|V_k\|_\infty \leq R_{\max}/(1 - \discount)$, which can
be established by induction using~\eqref{eq:td0_contraction}. Therefore:
\begin{equation}
  \E\bigl[\|w_k(s)\|^2 \mid \mathcal{F}_k\bigr]
  \leq C\bigl(1 + \|V_k\|_\infty^2\bigr)
  \label{eq:td0_noise_bound}
\end{equation}
for a constant $C$ depending on $R_{\max}$, $\discount$, and
$|\States|$.

\medskip
All five conditions of the Robbins--Monro theorem are verified: (1)
step sizes, (2) martingale noise, (3) unique root of $h$, (4) stability
via contraction, (5) bounded noise variance. Therefore
$V_n(s) \to \Vpi(s)$ almost surely for all $s \in \States$.
\end{proof}

\begin{remark}[Convergence Despite Bootstrapping Bias]
\label{rem:bootstrap_convergence}
At any finite time, $V(S_{t+1}) \neq \Vpi(S_{t+1})$, so the TD
target $R_{t+1} + \discount V(S_{t+1})$ is a biased estimate of
$\Vpi(S_t)$. Why does TD(0) still converge? The contraction property
(Step 4) provides the answer: the expected update direction
$\Bellman^\policy V - V$ always points toward $\Vpi$, and the
contraction ensures that the bootstrapping loop is stable --- the
bias shrinks as $V$ approaches $\Vpi$, creating a virtuous cycle.
\end{remark}

\begin{example}[Two-State MDP]
\label{ex:two_state}
Consider an MDP with $\States = \{A, B\}$, a single action,
deterministic transitions $A \to B \to A \to \cdots$, rewards
$r(A) = 1$, $r(B) = 0$, and $\discount = 0.9$. The true values are
$\Vpi(A) = 100/19 \approx 5.26$ and $\Vpi(B) \approx 4.74$.
Starting from $V_0(A) = V_0(B) = 0$ with $\alpha = 0.1$, the first
TD(0) update at $A$ gives $V_1(A) = 0.1(1 + 0) = 0.1$, gradually
refining toward the true values through bootstrapping.
\end{example}

%% SECTION 5: BIAS-VARIANCE ANALYSIS
\section{Bias--Variance Analysis}
\label{sec:bias_variance}

We now make precise the tradeoff between bias and variance that
distinguishes TD(0) from MC.

\begin{proposition}[Bias of the TD(0) Target]
\label{prop:td0_bias}
The conditional expectation of the TD(0) target is
\begin{equation}
  \E_\policy\bigl[R_{t+1} + \discount V(S_{t+1}) \mid S_t = s\bigr]
  = (\Bellman^\policy V)(s),
  \label{eq:td0_bias}
\end{equation}
which equals $\Vpi(s)$ if and only if $V = \Vpi$. Thus the TD(0)
target is \emph{biased} whenever $V \neq \Vpi$.
\end{proposition}

\begin{proof}
Direct computation:
\begin{equation}
  \E_\policy\bigl[R_{t+1} + \discount V(S_{t+1}) \mid S_t = s\bigr]
  = \sum_a \policy(a \mid s) \sum_{s'} \Trans(s' \mid s, a)
    \bigl[r(s,a,s') + \discount V(s')\bigr]
  = (\Bellman^\policy V)(s).
\end{equation}
Since $\Vpi$ is the unique fixed point of $\Bellman^\policy$,
$(\Bellman^\policy V)(s) = \Vpi(s)$ for all $s$ if and only if
$V = \Vpi$.
\end{proof}

\begin{proposition}[Variance Reduction by TD(0)]
\label{prop:td0_variance}
For deterministic rewards, the variance of the TD(0) target satisfies
\begin{equation}
  \Var_\policy\!\bigl(R_{t+1} + \discount V(S_{t+1}) \mid S_t = s\bigr)
  = \Var_\policy\!\bigl(r(s, A_t, S_{t+1}) + \discount V(S_{t+1})
    \mid S_t = s\bigr),
  \label{eq:td0_variance}
\end{equation}
which depends only on the randomness of the next transition
$(A_t, S_{t+1})$. In contrast, $\Var_\policy(\return \mid S_t = s)$
depends on all future transitions.
\end{proposition}

\begin{proof}
Since $R_{t+1} + \discount V(S_{t+1})$ is fully determined by
$(S_t, A_t, S_{t+1})$, its variance involves only the randomness of
the next action and transition. Using the law of total variance with
$V = \Vpi$:
\begin{equation}
  \Var_\policy(\return \mid S_t)
  = \underbrace{\Var_\policy\!\bigl(R_{t+1} + \discount \Vpi(S_{t+1})
    \mid S_t\bigr)}_{\text{one-step variance}}
  + \discount^2 \underbrace{\E_\policy\bigl[\Var_\policy(G_{t+1}
    \mid S_{t+1}) \mid S_t\bigr]}_{\text{future variance } \geq 0}.
  \label{eq:variance_decomp}
\end{equation}
The TD(0) target captures only the first term, eliminating the
non-negative future variance entirely.
\end{proof}

\begin{keyresult}[title={Bias--Variance Tradeoff: MC vs TD(0)}]
\begin{center}
\begin{tabular}{lcc}
\toprule
& \textbf{Bias} & \textbf{Variance} \\
\midrule
MC ($\return$) & Zero & High (all future steps) \\
TD(0) ($R_{t+1} + \discount V(S_{t+1})$)
  & Nonzero (when $V \neq \Vpi$) & Low (one step) \\
\bottomrule
\end{tabular}
\end{center}
The $n$-step TD methods developed next provide a continuum between
these two extremes.
\end{keyresult}

\begin{example}[Variance Comparison]
\label{ex:variance}
In a chain with $\discount = 0.99$ (effective horizon $\approx 100$)
and per-step reward variance $\sigma^2$, the MC return variance is
approximately $\sum_{k=0}^{99} \discount^{2k}\sigma^2 \approx
50\sigma^2$, while the TD(0) target variance is approximately
$\sigma^2$ --- a 50-fold reduction.
\end{example}

%% SECTION 6: N-STEP TD PREDICTION
\section{$n$-Step TD Prediction}
\label{sec:nstep}

The $n$-step TD method interpolates between TD(0) (one-step bootstrap)
and MC (full return) by using $n$ actual rewards before bootstrapping.

\begin{definition}[$n$-Step Return]
\label{def:nstep_return}
The \emph{$n$-step return} from time $t$ is
\begin{equation}
  G_t^{(n)} = \sum_{k=0}^{n-1} \discount^k R_{t+k+1}
  + \discount^n V(S_{t+n}).
  \label{eq:nstep_return}
\end{equation}
For $n = 1$, this reduces to the TD(0) target
$R_{t+1} + \discount V(S_{t+1})$. In the limit $n \to \infty$
(with episode length $T$), the bootstrap term
$\discount^n V(S_{t+n}) \to 0$ and $G_t^{(n)} \to \return$.
\end{definition}

\begin{definition}[$n$-Step TD Update]
\label{def:nstep_update}
The \emph{$n$-step TD update} is
\begin{equation}
  V(S_t) \leftarrow V(S_t)
  + \alpha\bigl[G_t^{(n)} - V(S_t)\bigr].
  \label{eq:nstep_update}
\end{equation}
\end{definition}

\begin{example}[$n$-Step Returns in a Chain]
\label{ex:nstep_chain}
Consider a chain $A \to B \to C \to D$ with $\discount = 0.9$,
rewards $r(A) = 1, r(B) = 0, r(C) = 2$, and estimates
$V(B) = 3, V(C) = 4, V(D) = 2$. From state $A$:
$G_A^{(1)} = 1 + 0.9(3) = 3.7$,
$G_A^{(2)} = 1 + 0 + 0.81(4) = 4.24$,
$G_A^{(3)} = 1 + 0 + 1.62 + 0.729(2) = 4.078$.
Each uses more actual rewards and less bootstrapping.
\end{example}

\subsection{Telescoping Decomposition}
\label{subsec:nstep_telescope}

\begin{lemma}[$n$-Step Return as Telescoping Sum of TD Errors]
\label{lem:nstep_telescope}
The $n$-step prediction error decomposes as
\begin{equation}
  G_t^{(n)} - V(S_t) = \sum_{k=0}^{n-1} \discount^k \delta_{t+k},
  \label{eq:nstep_telescope}
\end{equation}
where $\delta_{t+k} = R_{t+k+1} + \discount V(S_{t+k+1}) - V(S_{t+k})$.
\end{lemma}

\begin{proof}
Expand the right-hand side:
\begin{align}
  \sum_{k=0}^{n-1} \discount^k \delta_{t+k}
  &= \sum_{k=0}^{n-1} \discount^k
     \bigl[R_{t+k+1} + \discount V(S_{t+k+1}) - V(S_{t+k})\bigr]
     \nonumber\\
  &= \sum_{k=0}^{n-1} \discount^k R_{t+k+1}
     + \sum_{k=0}^{n-1} \discount^{k+1} V(S_{t+k+1})
     - \sum_{k=0}^{n-1} \discount^k V(S_{t+k}).
     \label{eq:nstep_expand}
\end{align}
Re-index the second sum with $j = k + 1$:
\begin{align}
  &= \sum_{k=0}^{n-1} \discount^k R_{t+k+1}
     + \sum_{j=1}^{n} \discount^j V(S_{t+j})
     - \sum_{k=0}^{n-1} \discount^k V(S_{t+k}) \nonumber\\
  &= \sum_{k=0}^{n-1} \discount^k R_{t+k+1}
     + \discount^n V(S_{t+n}) - V(S_t)
   = G_t^{(n)} - V(S_t),
  \label{eq:nstep_telescope_final}
\end{align}
where the telescoping cancellation eliminates all intermediate value
terms $\discount^k V(S_{t+k})$ for $k = 1, \ldots, n-1$.
\end{proof}

\begin{remark}[Connecting to the Infinite Telescoping]
\label{rem:nstep_infinite}
Lemma~\ref{lem:nstep_telescope} is the finite-$n$ analogue of
Lemma~\ref{lem:return_telescope}. As $n \to \infty$, the $n$-step
decomposition~\eqref{eq:nstep_telescope} approaches the infinite
decomposition~\eqref{eq:return_telescope}, with the bootstrap term
$\discount^n V(S_{t+n}) \to 0$ absorbed into the tail of the TD error
sum.
\end{remark}

\subsection{Convergence of $n$-Step TD}
\label{subsec:nstep_convergence}

\begin{theorem}[Convergence of $n$-Step TD]
\label{thm:nstep_convergence}
Under Assumption~\ref{ass:td0}, the $n$-step TD
update~\eqref{eq:nstep_update} converges to $\Vpi$ almost surely
for any fixed $n \geq 1$.
\end{theorem}

\begin{proof}
Define the $n$-step Bellman operator
$\Bellman_n^\policy \colon \R^{|\States|} \to \R^{|\States|}$ by
\begin{equation}
  (\Bellman_n^\policy V)(s)
  = \E_\policy\bigl[G_t^{(n)} \mid S_t = s\bigr]
  = \E_\policy\!\left[\sum_{k=0}^{n-1} \discount^k R_{t+k+1}
    + \discount^n V(S_{t+n}) \;\middle|\; S_t = s\right].
  \label{eq:nstep_operator}
\end{equation}

\textbf{Step 1: $\Bellman_n^\policy$ is a $\discount^n$-contraction.}
The operator $\Bellman_n^\policy$ is the $n$-fold composition
$(\Bellman^\policy)^n$. Since $\Bellman^\policy$ is a
$\discount$-contraction in $\|\cdot\|_\infty$, we have for any
$V_1, V_2$:
\begin{equation}
  \|\Bellman_n^\policy V_1 - \Bellman_n^\policy V_2\|_\infty
  = \|(\Bellman^\policy)^n V_1 - (\Bellman^\policy)^n V_2\|_\infty
  \leq \discount^n \|V_1 - V_2\|_\infty.
\end{equation}
The unique fixed point of $\Bellman_n^\policy$ is $\Vpi$
(since $\Bellman^\policy \Vpi = \Vpi$ implies
$(\Bellman^\policy)^n \Vpi = \Vpi$).

\textbf{Step 2: Cast as stochastic approximation.}
The $n$-step TD update at the $k$-th visit to state $s$ has the form
\begin{equation}
  V_{k+1}(s) = V_k(s)
  + \alpha_k(s)\bigl[(\Bellman_n^\policy V_k)(s) - V_k(s)
    + w_k^{(n)}(s)\bigr],
\end{equation}
where $w_k^{(n)}(s) = G_t^{(n)} - (\Bellman_n^\policy V_k)(s)$ is
zero-mean noise (by the same argument as Step 2 in the TD(0) proof).

\textbf{Step 3: Verify remaining conditions.}
The driving function $h_n(V)(s) = (\Bellman_n^\policy V)(s) - V(s)$
has unique root $\Vpi$. The $\discount^n$-contraction ensures stability
(indeed, the contraction rate is \emph{stronger} than TD(0)). The
noise variance is bounded since rewards are bounded and $n$ is finite.
Step-size conditions hold by assumption.

All conditions of the Robbins--Monro theorem are satisfied, so
$V_k(s) \to \Vpi(s)$ almost surely for all $s$.
\end{proof}

\begin{rlconnection}[title={The Spectrum from TD(0) to MC}]
The $n$-step TD family provides a continuum parameterized by $n$:
\begin{itemize}[nosep]
  \item $n = 1$: TD(0) --- maximal bias, minimal variance, fastest
    per-step computation.
  \item $n = T$ (episode length): MC --- zero bias, maximal variance,
    requires complete episodes.
  \item Intermediate $n$: trades off bias and variance; optimal $n$
    depends on the problem.
\end{itemize}
In deep RL practice, moderate values of $n$ (typically 5--20) combined
with value function baselines yield the best empirical performance.
\end{rlconnection}

\begin{example}[Contraction Rate vs.\ Variance]
\label{ex:nstep_bias}
For $\discount = 0.99$: TD(0) contracts at rate $0.99$, while 10-step
TD contracts at rate $0.99^{10} \approx 0.904$ (faster). But the
10-step target has higher variance from 10 random transitions. The
optimal $n$ balances faster contraction against noisier updates.
\end{example}

%% SECTION 7: EXERCISES
\section{Exercises}
\label{sec:exercises}

\begin{exercise}[TD Error Variance]
\textnormal{(\(*\))} \\
Consider a finite MDP with deterministic rewards. Show that the variance
of the TD error under the true value function satisfies
\[
  \Var_\policy(\delta_t^\policy \mid S_t = s)
  = \Var_\policy\!\bigl(R_{t+1} + \discount \Vpi(S_{t+1})
    \mid S_t = s\bigr).
\]
Interpret this as the one-step variance that TD(0) uses.
\end{exercise}

\begin{exercise}[MC as $n$-Step TD]
\textnormal{(\(*\))} \\
For an episodic MDP with episode length $T$ starting from time $t = 0$,
verify directly from Definition~\ref{def:nstep_return} that
$G_0^{(T)} = \return[0]$ when $V(S_T) = 0$ (terminal state has zero
value).
\end{exercise}

\begin{exercise}[Contraction Rate of the $n$-Step Operator]
\textnormal{(\(**\))} \\
Let $\discount = 0.95$ and $n = 5$. Compute the contraction rate
$\discount^n$ and determine how many applications of $\Bellman_n^\policy$
are needed to reduce the initial error $\|V_0 - \Vpi\|_\infty$ by a
factor of $10^{-3}$.

\textit{Hint:} Solve $\discount^{nk} \leq 10^{-3}$ for $k$.
\end{exercise}

\begin{exercise}[Stochastic Approximation Decomposition for $n = 2$]
\textnormal{(\(**\))} \\
Write out the full stochastic approximation decomposition for the
2-step TD update: express $G_t^{(2)} - V(S_t)$ as
$(\Bellman_2^\policy V)(S_t) - V(S_t) + w_t^{(2)}$, and verify
explicitly that $\E[w_t^{(2)} \mid S_t, V] = 0$.
\end{exercise}

\begin{exercise}[Bias Quantification]
\textnormal{(\(***\))} \\
Let $V$ be an arbitrary value function estimate with
$\|V - \Vpi\|_\infty \leq \varepsilon$. Show that the bias of the
$n$-step return satisfies
\[
  \bigl|\E_\policy\bigl[G_t^{(n)} \mid S_t = s\bigr]
  - \Vpi(s)\bigr| \leq \discount^n \varepsilon
  \quad \text{for all } s \in \States.
\]
Conclude that larger $n$ reduces the bias geometrically.

\textit{Hint:} Use the fact that $\Bellman_n^\policy$ is a
$\discount^n$-contraction with fixed point $\Vpi$.
\end{exercise}

\begin{exercise}[Variance of the $n$-Step Return]
\textnormal{(\(***\))} \\
Assuming $V = \Vpi$ and deterministic rewards, derive a recursive
formula for $\Var_\policy(G_t^{(n)} \mid S_t = s)$ in terms of one-step
transition variances. Show that $\Var_\policy(G_t^{(n)} \mid S_t)$ is
non-decreasing in $n$, confirming that longer lookback increases
variance.

\textit{Hint:} Use the telescoping decomposition
$G_t^{(n)} - \Vpi(S_t) = \sum_{k=0}^{n-1} \discount^k
\delta_{t+k}^\policy$ and the fact that the $\delta_{t+k}^\policy$
are martingale differences.
\end{exercise}

%% SECTION 8: SUMMARY
\section{Summary}
\label{sec:summary}

\begin{itemize}
  \item The \textbf{TD error}
    $\delta_t = R_{t+1} + \discount V(S_{t+1}) - V(S_t)$ measures the
    one-step prediction discrepancy; under the true value function, it
    forms a \emph{martingale difference sequence}
    ($\E[\delta_t^\policy \mid S_t] = 0$).

  \item The \textbf{telescoping decomposition}
    $\return - \Vpi(S_t) = \sum_{k=0}^\infty \discount^k
    \delta_{t+k}^\policy$ decomposes the total prediction error into
    a sum of discounted one-step TD errors --- the foundation for
    TD($\lambda$) and GAE.

  \item \textbf{TD(0)} converges to $\Vpi$ almost surely under standard
    conditions, proved by casting it as a stochastic approximation of
    the Bellman equation: the Bellman operator's contraction property
    provides stability, and the martingale noise has bounded variance.

  \item \textbf{Bias--variance tradeoff}: MC targets are unbiased but
    have high variance (accumulated over all future steps); TD(0)
    targets have low variance (one step) but are biased when
    $V \neq \Vpi$.

  \item The \textbf{$n$-step return}
    $G_t^{(n)} = \sum_{k=0}^{n-1} \discount^k R_{t+k+1}
    + \discount^n V(S_{t+n})$ interpolates between TD(0) ($n=1$) and
    MC ($n \to \infty$), with contraction rate $\discount^n$.

  \item \textbf{$n$-step TD converges} for any fixed $n \geq 1$ via
    the same stochastic approximation framework, with the $n$-step
    Bellman operator $\Bellman_n^\policy = (\Bellman^\policy)^n$
    providing a $\discount^n$-contraction.
\end{itemize}

%% SECTION 9: FURTHER READING
\section{Further Reading}
\label{sec:further_reading}

\begin{itemize}
  \item \textbf{Lesson 48} extends the $n$-step framework to
    TD($\lambda$), which takes a geometrically weighted average of all
    $n$-step returns, and develops the theory of eligibility traces for
    efficient incremental computation.

  \item \textbf{Lesson 49} applies TD errors to the control setting,
    developing SARSA, Q-learning, and Expected SARSA with convergence
    proofs.

  \item \citet{sutton1988} is the original paper introducing TD learning
    and proving convergence for the linear case. The stochastic
    approximation approach used here follows \citet{tsitsiklis1994} and
    \citet{jaakkola1994}.

  \item \citet[Chapter 7]{suttonbarto2018} provides extensive empirical
    comparisons of $n$-step methods on benchmark tasks, illustrating the
    bias--variance tradeoff in practice.

  \item \citet{borkar2008} gives the definitive treatment of stochastic
    approximation via the ODE method, providing the theoretical
    foundations underlying all convergence proofs in this lesson.

  \item For the extension to function approximation, where TD methods
    can diverge (the ``deadly triad''), see \citet{tsitsiklis1997}
    and \citet{baird1995}.
\end{itemize}

\bibliography{td-refs}

\end{document}
