\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 49: SARSA -- On-Policy TD Control}}
\fancyhead[R]{\thepage}
\title{Lesson 49: SARSA -- On-Policy TD Control}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% SECTION 1: MOTIVATION AND OVERVIEW
\section{Motivation and Overview}
\label{sec:motivation}

Lessons 47 and 48 developed temporal-difference methods for
\emph{prediction}: estimating the value function $\Vpi$ of a fixed
policy $\policy$. But the ultimate goal of reinforcement learning is
\emph{control} --- finding a policy that maximizes cumulative reward.
Consider a robot navigating a factory floor, where each wrong turn
costs time and battery. The robot cannot afford to evaluate every
candidate policy to completion; it must improve its behavior
\emph{while interacting with the environment}. This requires
simultaneously evaluating the current policy and improving it --- a
challenge known as the \emph{exploration--exploitation dilemma}.

The transition from prediction to control introduces two fundamental
difficulties. First, the agent must learn \emph{action-value} functions
$\Qpi(s,a)$ rather than state-value functions $\Vpi(s)$, because
computing the greedy policy from $\Vpi$ requires a model of the
environment, while computing it from $\Qpi$ does not. Second, the
agent must maintain exploration to ensure all state--action pairs are
sampled, while simultaneously converging to a greedy policy. SARSA
resolves both difficulties by applying the TD(0) update to
action-value functions under an $\varepsilon$-greedy policy that
becomes greedier over time. The tools developed in this lesson will let
us formalize the SARSA algorithm, state the GLIE convergence
conditions, and prove that SARSA converges to the optimal
action-value function $\Qstar$.

\begin{rlconnection}[title={Where SARSA and On-Policy Control Appear}]
\begin{itemize}[nosep]
  \item \textbf{SARSA($\lambda$)} extends SARSA with eligibility
    traces (Lesson 48), propagating TD errors backward through
    recently visited state--action pairs for faster credit assignment.
  \item \textbf{Expected SARSA} (Lesson 50) replaces the sampled
    next action with an expectation, reducing variance while
    maintaining the on-policy structure.
  \item \textbf{Actor--critic methods} generalize the on-policy
    evaluation--improvement loop of SARSA to the function approximation
    setting, using a critic (value estimator) and an actor (policy).
  \item \textbf{Safe reinforcement learning}: SARSA's on-policy
    nature means it accounts for the cost of exploration, often
    yielding safer behavior than off-policy alternatives like
    Q-learning.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
TD(0) update rule and TD error $\delta_t$ (Lesson 47,
Definitions 4.1--4.2). Convergence of TD(0) via stochastic
approximation: the Bellman operator $\Bellman^\policy$ as a
$\discount$-contraction, martingale noise, and the Robbins--Monro
theorem (Lesson 47, Theorem 4.7). Bellman expectation and optimality
equations for action-value functions $\Qpi$ and $\Qstar$ (Lessons
45--46). Step-size conditions $\sum_n \alpha_n = \infty$,
$\sum_n \alpha_n^2 < \infty$ (Lesson 31).

\paragraph{Learning objectives.}
\begin{itemize}
  \item \emph{Define} the action-value Bellman operator and verify
    that it is a $\discount$-contraction in $\|\cdot\|_\infty$.
  \item \emph{State} the SARSA update rule and explain the acronym
    $(S_t, A_t, R_{t+1}, S_{t+1}, A_{t+1})$.
  \item \emph{Define} $\varepsilon$-greedy policies and the GLIE
    (Greedy in the Limit with Infinite Exploration) condition.
  \item \emph{Prove} the convergence of SARSA to $\Qstar$ under
    GLIE by decomposing the update into a Q-learning component
    plus a vanishing exploration perturbation.
  \item \emph{Compute} SARSA updates on small MDPs and compare
    the learned values with those of a purely greedy agent.
  \item \emph{Explain} why the on-policy nature of SARSA leads to
    different behavior than off-policy Q-learning in stochastic
    environments.
\end{itemize}

%% SECTION 2: REFERENCES
\section{References}
\label{sec:references}

\begin{itemize}[leftmargin=2em]
  \item \citet[Chapter 6, Sections 6.4--6.5]{suttonbarto2018} ---
    SARSA, on-policy TD control, and the cliff-walking example
    comparing SARSA and Q-learning.
  \item \citet[Chapter 4]{bertsekas1996} --- Convergence of TD
    control methods via stochastic approximation.
  \item \citet{singh2000} --- The definitive convergence proof for
    single-step on-policy RL algorithms, including SARSA under GLIE.
  \item \citet[Section 4.2]{szepesvari2010} --- Compact treatment of
    TD control convergence.
  \item \citet{rummery1994} --- The original paper introducing SARSA
    (under the name ``modified connectionist Q-learning'').
  \item \citet{tsitsiklis1994} --- Asynchronous stochastic
    approximation framework underlying the convergence analysis.
\end{itemize}

%% SECTION 3: FROM PREDICTION TO CONTROL
\section{From Prediction to Control}
\label{sec:prediction_to_control}

The TD prediction methods of Lessons 47--48 estimate $\Vpi$ for a
fixed policy $\policy$. To find an optimal policy, we employ the
\emph{Generalized Policy Iteration} (GPI) framework: alternate between
\emph{policy evaluation} (estimating a value function) and
\emph{policy improvement} (making the policy greedy with respect to
the estimated values).

\begin{definition}[Generalized Policy Iteration]
\label{def:gpi}
\emph{Generalized Policy Iteration (GPI)} refers to any algorithm
that interleaves partial policy evaluation with policy improvement.
The evaluation step updates the value function toward $\Qpi$
(or $\Vpi$), and the improvement step adjusts the policy to be
greedier with respect to the current value estimate.
\end{definition}

For model-free control, we must work with action-value functions
rather than state-value functions.

\begin{remark}[Why Action Values?]
\label{rem:why_q}
Given $\Vpi(s)$, selecting the greedy action requires solving
$\argmax_a \sum_{s'} \Trans(s' \mid s,a)[r(s,a,s') + \discount
\Vpi(s')]$, which requires knowledge of the transition kernel
$\Trans$. In contrast, given $\Qpi(s,a)$, the greedy action is
simply $\argmax_a \Qpi(s,a)$ --- no model needed. This is why all
model-free TD control methods learn $Q$-values.
\end{remark}

\subsection{Action-Value Bellman Operators}
\label{subsec:bellman_q}

\begin{definition}[Action-Value Bellman Expectation Operator]
\label{def:bellman_q}
For a policy $\policy$, the \emph{action-value Bellman expectation
operator} $\Bellman^\policy_Q \colon \R^{|\States||\Actions|}
\to \R^{|\States||\Actions|}$ is defined by
\begin{equation}
  (\Bellman^\policy_Q Q)(s,a)
  = \sum_{s'} \Trans(s' \mid s,a) \left[
    r(s,a,s') + \discount \sum_{a'} \policy(a' \mid s') Q(s',a')
  \right].
  \label{eq:bellman_q}
\end{equation}
Its unique fixed point is $\Qpi$.
\end{definition}

\begin{definition}[Bellman Optimality Operator on $Q$-Values]
\label{def:bellman_opt_q}
The \emph{Bellman optimality operator} on action values is
\begin{equation}
  (\Bellman^*_Q Q)(s,a)
  = \sum_{s'} \Trans(s' \mid s,a) \left[
    r(s,a,s') + \discount \max_{a'} Q(s',a')
  \right].
  \label{eq:bellman_opt_q}
\end{equation}
Its unique fixed point is $\Qstar$.
\end{definition}

\begin{lemma}[Contraction of Action-Value Bellman Operators]
\label{lem:bellman_q_contraction}
Both $\Bellman^\policy_Q$ and $\Bellman^*_Q$ are
$\discount$-contractions in $\|\cdot\|_\infty$:
\begin{equation}
  \|\Bellman^\policy_Q Q_1 - \Bellman^\policy_Q Q_2\|_\infty
  \leq \discount \|Q_1 - Q_2\|_\infty, \qquad
  \|\Bellman^*_Q Q_1 - \Bellman^*_Q Q_2\|_\infty
  \leq \discount \|Q_1 - Q_2\|_\infty.
  \label{eq:q_contraction}
\end{equation}
\end{lemma}

\begin{proof}
We prove the result for $\Bellman^\policy_Q$; the argument for
$\Bellman^*_Q$ is analogous (using $|\max_a f(a) - \max_a g(a)|
\leq \max_a |f(a) - g(a)|$). For any $(s,a)$:
\begin{align}
  |(\Bellman^\policy_Q Q_1)(s,a) - (\Bellman^\policy_Q Q_2)(s,a)|
  &= \discount \left|
    \sum_{s'} \Trans(s' \mid s,a) \sum_{a'} \policy(a' \mid s')
    \bigl[Q_1(s',a') - Q_2(s',a')\bigr]
  \right| \nonumber\\
  &\leq \discount \sum_{s'} \Trans(s' \mid s,a) \sum_{a'}
    \policy(a' \mid s') |Q_1(s',a') - Q_2(s',a')| \nonumber\\
  &\leq \discount \|Q_1 - Q_2\|_\infty
    \underbrace{\sum_{s'} \Trans(s' \mid s,a)
    \sum_{a'} \policy(a' \mid s')}_{= 1}.
  \label{eq:q_contraction_proof}
\end{align}
Taking the supremum over $(s,a)$ gives the result.
\end{proof}

The contraction properties established in
Lemma~\ref{lem:bellman_q_contraction} will be the key stability
condition in the convergence proofs that follow.

%% SECTION 4: EXPLORATION AND EPSILON-GREEDY POLICIES
\section{Exploration and $\varepsilon$-Greedy Policies}
\label{sec:exploration}

\begin{definition}[$\varepsilon$-Greedy Policy]
\label{def:epsilon_greedy}
Given an action-value function $Q$, the \emph{$\varepsilon$-greedy
policy} assigns
\begin{equation}
  \policy_\varepsilon(a \mid s) = \begin{cases}
    1 - \varepsilon + \frac{\varepsilon}{|\Actions|}
      & \text{if } a = \argmax_{a'} Q(s,a'), \\[4pt]
    \frac{\varepsilon}{|\Actions|}
      & \text{otherwise,}
  \end{cases}
  \label{eq:eps_greedy}
\end{equation}
where $\varepsilon \in (0,1]$ and ties in $\argmax$ are broken
arbitrarily but consistently.
\end{definition}

\begin{proposition}[$\varepsilon$-Greedy Policy Improvement]
\label{prop:eps_improvement}
Let $Q = \Qpi$ for some policy $\policy$, and let
$\policy_\varepsilon$ be the $\varepsilon$-greedy policy with
respect to $Q$. Then
$\Qpi(s, \policy_\varepsilon) \geq \Vpi(s)$ for all $s$,
where $\Qpi(s, \policy_\varepsilon)
= \sum_a \policy_\varepsilon(a \mid s) \Qpi(s,a)$.
\end{proposition}

\begin{proof}
Since $\policy_\varepsilon$ places probability at least
$\varepsilon/|\Actions|$ on every action and the remaining
$1 - \varepsilon$ on the greedy action:
\begin{align}
  \sum_a \policy_\varepsilon(a \mid s) \Qpi(s,a)
  &= \frac{\varepsilon}{|\Actions|} \sum_a \Qpi(s,a)
     + (1 - \varepsilon) \max_a \Qpi(s,a) \nonumber\\
  &\geq \frac{\varepsilon}{|\Actions|} \sum_a \Qpi(s,a)
     + (1 - \varepsilon) \sum_a \policy(a \mid s) \Qpi(s,a)
     \nonumber\\
  &\geq \sum_a \policy(a \mid s) \Qpi(s,a) = \Vpi(s),
  \label{eq:eps_improvement}
\end{align}
where the first inequality uses
$\max_a \Qpi(s,a) \geq \sum_a \policy(a \mid s) \Qpi(s,a)$
and the second uses $\varepsilon/|\Actions| \leq \policy(a \mid s)$
for the non-greedy actions under a sufficiently exploratory policy.
More precisely, the improvement follows from convexity: the
$\varepsilon$-greedy policy shifts probability mass toward higher
$Q$-values.
\end{proof}

\begin{definition}[GLIE Condition]
\label{def:glie}
A sequence of policies $\{\policy_t\}_{t=0}^\infty$ is
\emph{Greedy in the Limit with Infinite Exploration (GLIE)} if:
\begin{enumerate}[nosep]
  \item Every state--action pair is visited infinitely often:
    $\sum_{t=0}^\infty \indicator\{S_t = s, A_t = a\} = \infty$
    almost surely for all $(s,a) \in \States \times \Actions$.
  \item The policy converges to the greedy policy:
    $\policy_t(a \mid s) \to \indicator\{a = \argmax_{a'} Q(s,a')\}$
    almost surely for all $s \in \States$.
\end{enumerate}
\end{definition}

\begin{example}[Standard GLIE Schedule]
\label{ex:glie_schedule}
The $\varepsilon_t$-greedy policy with $\varepsilon_t = 1/t$
satisfies GLIE. Condition 2 holds since $\varepsilon_t \to 0$.
For Condition 1, since $\varepsilon_t > 0$ for all $t$ and the
MDP is finite, every state--action pair reachable under any
positive-probability policy is visited infinitely often.
Another common choice is $\varepsilon_t = c/(c + t)$ for some
constant $c > 0$.
\end{example}

\begin{warning}[title={Constant $\varepsilon$ Does Not Satisfy GLIE}]
Using a fixed $\varepsilon > 0$ violates Condition 2 of GLIE: the
policy never becomes fully greedy. With constant $\varepsilon$,
SARSA converges not to $\Qstar$ but to $Q^{\policy_\varepsilon}$,
the value function of the $\varepsilon$-greedy policy. The gap
$\|\Qstar - Q^{\policy_\varepsilon}\|_\infty$ is bounded by
$O(\varepsilon)$ but does not vanish.
\end{warning}

With the exploration framework established, we now define the SARSA
algorithm.

%% SECTION 5: THE SARSA ALGORITHM
\section{The SARSA Algorithm}
\label{sec:sarsa}

\begin{definition}[SARSA Update Rule]
\label{def:sarsa}
Given the quintuple $(S_t, A_t, R_{t+1}, S_{t+1}, A_{t+1})$ where
actions are drawn from the current $\varepsilon$-greedy policy
$A_t \sim \policy_t(\cdot \mid S_t)$ and
$A_{t+1} \sim \policy_t(\cdot \mid S_{t+1})$, the \emph{SARSA
update} is
\begin{equation}
  Q(S_t, A_t) \leftarrow Q(S_t, A_t)
  + \alpha_t\bigl[R_{t+1} + \discount Q(S_{t+1}, A_{t+1})
  - Q(S_t, A_t)\bigr].
  \label{eq:sarsa}
\end{equation}
The name ``SARSA'' is an acronym for the quintuple
$(\mathbf{S}_t, \mathbf{A}_t, \mathbf{R}_{t+1},
\mathbf{S}_{t+1}, \mathbf{A}_{t+1})$ used in the update.
\end{definition}

\begin{remark}[SARSA as TD(0) for Action Values]
\label{rem:sarsa_td}
Comparing with the TD(0) update for state values
(Lesson 47, Definition 4.2):
$V(S_t) \leftarrow V(S_t) + \alpha[\delta_t]$ where
$\delta_t = R_{t+1} + \discount V(S_{t+1}) - V(S_t)$,
the SARSA update is the direct analogue for action values with
TD error
\begin{equation}
  \delta_t = R_{t+1} + \discount Q(S_{t+1}, A_{t+1})
  - Q(S_t, A_t).
  \label{eq:sarsa_td_error}
\end{equation}
The only structural difference is that the bootstrap target uses the
action-value at the \emph{next state--action pair} rather than the
next state alone.
\end{remark}

\begin{keyresult}[title={SARSA: On-Policy TD Control}]
The SARSA algorithm performs GPI by interleaving single-step
action-value evaluation (the TD update~\eqref{eq:sarsa}) with
policy improvement (selecting $\varepsilon$-greedy actions). It is
\emph{on-policy}: the action $A_{t+1}$ in the bootstrap target is
drawn from the \emph{same} policy that the agent follows. This
means SARSA learns the value of the policy it is actually executing,
including the cost of exploratory actions.
\end{keyresult}

\begin{example}[SARSA Updates on a Grid]
\label{ex:sarsa_grid}
Consider a $3 \times 1$ grid with states $\{L, C, R\}$, actions
$\{$left, right$\}$, deterministic transitions (left/right move
accordingly, with walls keeping the agent in place), reward $+1$ at
$R$ and $0$ elsewhere, $\discount = 0.9$, and $Q_0 \equiv 0$.
Starting in $C$ with $\varepsilon = 0.1$ and $\alpha = 0.1$:

Suppose the trajectory begins $S_0 = C, A_0 = \text{right}$ (greedy
or exploratory), $R_1 = 1$, $S_1 = R, A_1 = \text{left}$ (action
chosen by $\policy_0$). Then:
\[
  Q_1(C, \text{right}) = 0 + 0.1[1 + 0.9 \cdot Q_0(R, \text{left})
  - 0] = 0.1[1 + 0] = 0.1.
\]
After several episodes, $Q(C, \text{right})$ grows toward
$\Qstar(C, \text{right}) = 1 + 0.9 \cdot 1/(1-0.9) = 10$.
\end{example}

\begin{rlconnection}[title={On-Policy Safety in Cliff Walking}]
The classic cliff-walking example
\citep[Section 6.5]{suttonbarto2018} illustrates a key consequence
of SARSA's on-policy nature. In a gridworld with a cliff along the
bottom edge (falling off gives $-100$ reward), SARSA learns to take
a \emph{safe path} far from the cliff because its value estimates
account for the $\varepsilon$-probability of exploratory actions
that could step off the edge. Q-learning, being off-policy, learns
the \emph{optimal path} along the cliff edge --- which is optimal
only if the agent executes the greedy policy perfectly, without
exploration. During learning, Q-learning's agent falls off the cliff
much more frequently.
\end{rlconnection}

We now turn to the convergence analysis of SARSA.

%% SECTION 6: CONVERGENCE OF SARSA
\section{Convergence of SARSA}
\label{sec:convergence}

\begin{assumption}[SARSA Convergence Conditions]
\label{ass:sarsa}
\begin{enumerate}[nosep]
  \item The MDP has finite state space $\States$ and finite action
    space $\Actions$.
  \item Rewards are bounded: $|r(s,a,s')| \leq R_{\max}$ for all
    $(s,a,s')$.
  \item The policy sequence $\{\policy_t\}$ satisfies GLIE
    (Definition~\ref{def:glie}).
  \item Step sizes satisfy $\sum_n \alpha_n(s,a) = \infty$ and
    $\sum_n \alpha_n^2(s,a) < \infty$ for each
    $(s,a) \in \States \times \Actions$.
\end{enumerate}
\end{assumption}

\begin{keyresult}[title={Convergence of SARSA (Singh et al.\ 2000)}]
\begin{theorem}[Convergence of SARSA]
\label{thm:sarsa}
Under Assumption~\ref{ass:sarsa}, the SARSA iterates converge to the
optimal action-value function almost surely:
\begin{equation}
  Q_t(s,a) \cas \Qstar(s,a)
  \quad \text{for all } (s,a) \in \States \times \Actions.
  \label{eq:sarsa_convergence}
\end{equation}
\end{theorem}
\end{keyresult}

\begin{proof}
We follow the approach of \citet{singh2000}, decomposing the SARSA
update into a Q-learning component plus a vanishing exploration
perturbation.

\textbf{Step 1: Decompose the SARSA target.}
At time $t$, the SARSA update target is
$R_{t+1} + \discount Q_t(S_{t+1}, A_{t+1})$. Add and subtract
$\discount \max_{a'} Q_t(S_{t+1}, a')$:
\begin{equation}
  R_{t+1} + \discount Q_t(S_{t+1}, A_{t+1})
  = \underbrace{R_{t+1} + \discount \max_{a'} Q_t(S_{t+1}, a')}
    _{\text{Q-learning target}}
  + \underbrace{\discount\bigl[Q_t(S_{t+1}, A_{t+1})
    - \max_{a'} Q_t(S_{t+1}, a')\bigr]}_{\xi_t}.
  \label{eq:sarsa_decomp}
\end{equation}
The perturbation $\xi_t$ captures the difference between following
the current $\varepsilon_t$-greedy policy and acting greedily.

\textbf{Step 2: Express as stochastic approximation with perturbation.}
At the $k$-th visit to state--action pair $(s,a)$, the SARSA update
can be written as
\begin{equation}
  Q_{k+1}(s,a) = Q_k(s,a)
  + \alpha_k(s,a)\bigl[(\Bellman^*_Q Q_k)(s,a) - Q_k(s,a)
  + w_k(s,a) + \xi_k(s,a)\bigr],
  \label{eq:sarsa_sa}
\end{equation}
where $w_k(s,a)$ is the zero-mean stochastic approximation noise
(the difference between the sampled Q-learning target and its
expectation $(\Bellman^*_Q Q_k)(s,a)$), and $\xi_k(s,a)$ is the
exploration perturbation from Step 1.

\textbf{Step 3: Verify that the noise is a martingale difference.}
Let $\mathcal{F}_k$ denote the filtration containing all information
up to the $k$-th visit to $(s,a)$. The noise
$w_k(s,a) = R_{t+1} + \discount \max_{a'} Q_k(S_{t+1}, a')
- (\Bellman^*_Q Q_k)(s,a)$ satisfies
\begin{equation}
  \E[w_k(s,a) \mid \mathcal{F}_k]
  = (\Bellman^*_Q Q_k)(s,a) - (\Bellman^*_Q Q_k)(s,a) = 0,
  \label{eq:noise_mg}
\end{equation}
by the definition of $\Bellman^*_Q$. The variance is bounded by
$\E[w_k^2 \mid \mathcal{F}_k] \leq C$ for a constant $C$ depending
on $R_{\max}$ and $\discount$, since rewards and $Q$-iterates
are bounded (the latter by $R_{\max}/(1-\discount)$, established
by induction using the contraction property).

\textbf{Step 4: Bound the exploration perturbation.}
Under the $\varepsilon_t$-greedy policy $\policy_t$, the conditional
expectation of $\xi_t$ given $\mathcal{F}_k$ and $S_{t+1}$ is:
\begin{align}
  \E[\xi_t \mid \mathcal{F}_k, S_{t+1}]
  &= \discount \sum_{a'} \policy_t(a' \mid S_{t+1}) Q_k(S_{t+1}, a')
     - \discount \max_{a'} Q_k(S_{t+1}, a') \nonumber\\
  &= \discount\!\left[
    (1-\varepsilon_t) \max_{a'} Q_k(S_{t+1}, a')
    + \frac{\varepsilon_t}{|\Actions|}
      \sum_{a'} Q_k(S_{t+1}, a')
    - \max_{a'} Q_k(S_{t+1}, a')
  \right] \nonumber\\
  &= \discount\,\varepsilon_t \left[
    \frac{1}{|\Actions|} \sum_{a'} Q_k(S_{t+1}, a')
    - \max_{a'} Q_k(S_{t+1}, a')
  \right].
  \label{eq:xi_bound}
\end{align}
Since $Q_k$ is bounded by $R_{\max}/(1-\discount)$:
\begin{equation}
  |\E[\xi_t \mid \mathcal{F}_k, S_{t+1}]|
  \leq \discount\,\varepsilon_t \cdot
  \frac{2R_{\max}}{1 - \discount}
  =: C_\xi\,\varepsilon_t.
  \label{eq:xi_bound_const}
\end{equation}

\textbf{Step 5: Apply perturbed stochastic approximation.}
The iteration~\eqref{eq:sarsa_sa} is a perturbed stochastic
approximation of the Bellman optimality equation with driving
function $h(Q)(s,a) = (\Bellman^*_Q Q)(s,a) - Q(s,a)$. The
unperturbed iteration (Q-learning) converges to $\Qstar$ by the
Robbins--Monro theorem, since $\Bellman^*_Q$ is a
$\discount$-contraction with unique fixed point $\Qstar$
(Lemma~\ref{lem:bellman_q_contraction}).

For the perturbation, we need $\sum_k \alpha_k(s,a)
|\E[\xi_k \mid \mathcal{F}_k]| < \infty$ almost surely. By
\eqref{eq:xi_bound_const}:
\begin{equation}
  \sum_k \alpha_k(s,a) |\E[\xi_k \mid \mathcal{F}_k]|
  \leq C_\xi \sum_k \alpha_k(s,a)\,\varepsilon_{t_k},
  \label{eq:perturbation_sum}
\end{equation}
where $t_k$ is the time of the $k$-th visit to $(s,a)$. Under the
GLIE condition, $\varepsilon_t \to 0$. For the standard schedule
$\alpha_n = 1/n$ and $\varepsilon_n = 1/n$:
$\sum_k \alpha_k \varepsilon_{t_k} \leq \sum_n 1/n^2 < \infty$.
More generally, the GLIE condition requires $\varepsilon_t$
to decrease fast enough that the perturbation series converges.

By the perturbed Robbins--Monro theorem: if the unperturbed
stochastic approximation converges (which it does --- this is
Q-learning) and the perturbation satisfies
$\sum_k \alpha_k |\E[\xi_k \mid \mathcal{F}_k]| < \infty$ a.s.,
then the perturbed iteration converges to the same limit.

Since $\Bellman^*_Q$ is a $\discount$-contraction with unique
fixed point $\Qstar$, and the perturbation is summable, the error
$\|Q_t - \Qstar\|_\infty$ converges to zero almost surely.
\end{proof}

\begin{remark}[The Role of GLIE]
\label{rem:glie_role}
The GLIE condition serves two purposes in the convergence proof.
Condition 1 (infinite exploration) ensures all state--action pairs
are visited infinitely often, which is needed for the step-size
conditions to be meaningful. Condition 2 ($\varepsilon_t \to 0$)
ensures that the exploration perturbation $\xi_t$ vanishes, so
that SARSA asymptotically behaves like Q-learning. Without
Condition 2, the perturbation remains $O(\varepsilon)$ and SARSA
converges to $Q^{\policy_\varepsilon}$ rather than $\Qstar$.
\end{remark}

\begin{example}[Verifying the Perturbation Bound]
\label{ex:perturbation}
Consider an MDP with $|\Actions| = 4$, $R_{\max} = 1$,
$\discount = 0.9$. Then $C_\xi = 0.9 \cdot 2/(1-0.9) = 18$.
With $\varepsilon_{100} = 0.01$, the perturbation magnitude at
step 100 is bounded by $18 \cdot 0.01 = 0.18$, a small fraction
of the typical $Q$-value scale $R_{\max}/(1-\discount) = 10$.
As $\varepsilon_t \to 0$, this bound vanishes.
\end{example}

\begin{rlconnection}[title={SARSA Convergence via Two-Timescale Analysis}]
An alternative view of SARSA convergence uses two-timescale
stochastic approximation (Lesson 32). If the exploration rate
$\varepsilon_t$ decreases on a \emph{faster} timescale than the
step size $\alpha_t$ (i.e., $\varepsilon_t/\alpha_t \to 0$), then
the policy converges to greedy ``fast'' while the $Q$-values update
``slowly,'' and the system tracks the equilibrium of Q-learning.
This perspective, developed by \citet{borkar2008}, provides a more
general framework for analyzing policy improvement algorithms with
diminishing exploration.
\end{rlconnection}

%% SECTION 7: ON-POLICY VS. OFF-POLICY BEHAVIOR
\section{On-Policy Behavior of SARSA}
\label{sec:on_policy}

\begin{proposition}[SARSA with Fixed $\varepsilon$]
\label{prop:sarsa_fixed_eps}
If the exploration rate is held constant at $\varepsilon > 0$ (so
the GLIE condition is violated), and all other conditions of
Assumption~\ref{ass:sarsa} hold, then SARSA converges to
$Q^{\policy_\varepsilon}$, the action-value function of the
$\varepsilon$-greedy policy, rather than $\Qstar$.
\end{proposition}

\begin{proof}
With fixed $\varepsilon$, the SARSA update targets
$\Bellman^{\policy_\varepsilon}_Q Q$, the Bellman expectation
operator of the $\varepsilon$-greedy policy. Since the policy is
now fixed, SARSA is simply TD(0) applied to action values under
$\policy_\varepsilon$. By the convergence of TD(0)
(Lesson 47, Theorem 4.7, adapted to action values):
$Q_t \to Q^{\policy_\varepsilon}$ almost surely.
\end{proof}

\begin{example}[On-Policy vs.\ Off-Policy in a Stochastic Grid]
\label{ex:on_off_policy}
Consider a $4 \times 4$ grid with stochastic wind: with probability
0.3, the agent is pushed one cell upward regardless of its action.
Near a cliff at the top row, the greedy-optimal path runs along the
top (shortest distance to the goal), but exploratory actions during
learning risk falling off. SARSA with $\varepsilon = 0.1$ learns
to avoid the top row, assigning lower $Q$-values to states near
the cliff because it accounts for the 10\% chance of a random
exploratory action pushing the agent over the edge. The converged
$Q^{\policy_{0.1}}$ values reflect the true cost of the exploration
policy, yielding a safer learned behavior.
\end{example}

This on-policy property distinguishes SARSA from Q-learning (Lesson
50) and is the principal reason for choosing SARSA in safety-critical
applications.

%% SECTION 8: EXERCISES
\section{Exercises}
\label{sec:exercises}

\begin{exercise}[SARSA TD Error Properties]
\textnormal{(\(*\))} \\
Let $\delta_t = R_{t+1} + \discount Q(S_{t+1}, A_{t+1})
- Q(S_t, A_t)$ be the SARSA TD error. If $Q = \Qpi$ for the
current (stationary) policy $\policy$, show that
$\E_\policy[\delta_t \mid S_t, A_t] = 0$.
\end{exercise}

\begin{exercise}[SARSA on a Three-State MDP]
\textnormal{(\(*\))} \\
Consider $\States = \{A, B, C\}$ with $\Actions = \{1, 2\}$,
deterministic transitions: action 1 in $A$ leads to $B$, action 2
in $A$ leads to $C$, all actions in $B$ and $C$ lead to $A$.
Rewards: $r(A,1) = 0$, $r(A,2) = 1$, $r(B,\cdot) = 0$,
$r(C,\cdot) = 2$. Let $\discount = 0.5$, $\alpha = 0.5$,
$Q_0 \equiv 0$, $\varepsilon = 0.1$. Suppose the first episode
generates the trajectory $(A,2,1,C,1,0,A)$ and then
$(A,1,0,B,2,0,A)$. Compute the SARSA updates for each step.
\end{exercise}

\begin{exercise}[GLIE Schedule Design]
\textnormal{(\(**\))} \\
Prove that the schedule $\varepsilon_t = c/(c+t)$ for $c > 0$
satisfies GLIE on any finite, communicating MDP. Specifically:
\begin{enumerate}[nosep]
  \item[(a)] Show that $\varepsilon_t \to 0$ (Condition 2).
  \item[(b)] Show that for any $(s,a)$, the probability that
    $(s,a)$ is visited at time $t$ is bounded below by a constant
    times $\varepsilon_t/|\Actions|$, and conclude from
    $\sum_t \varepsilon_t = \infty$ that $(s,a)$ is visited
    infinitely often (Condition 1).
\end{enumerate}
\end{exercise}

\begin{exercise}[Perturbation Bound Tightness]
\textnormal{(\(**\))} \\
In the convergence proof of Theorem~\ref{thm:sarsa}, the
perturbation bound~\eqref{eq:xi_bound_const} uses the worst case
$|\Actions|^{-1}\sum_{a'} Q(s',a') - \max_{a'} Q(s',a')$.
\begin{enumerate}[nosep]
  \item[(a)] Show that this quantity is always non-positive and
    bounded in magnitude by $2R_{\max}/(1-\discount)$.
  \item[(b)] Give an example where the bound is tight (achieved
    with equality).
  \item[(c)] Show that the bound can be improved to
    $R_{\max}(|\Actions|-1)/((1-\discount)|\Actions|)$ by a
    more careful analysis.
\end{enumerate}
\end{exercise}

\begin{exercise}[SARSA with Fixed $\varepsilon$: Value Gap]
\textnormal{(\(***\))} \\
Let $\Qstar$ denote the optimal action-value function and
$Q^{\policy_\varepsilon}$ the action-value function of the
$\varepsilon$-greedy policy with respect to $\Qstar$. Show that
\begin{equation*}
  \|\Qstar - Q^{\policy_\varepsilon}\|_\infty
  \leq \frac{2\varepsilon\,R_{\max}}{(1-\discount)^2}.
\end{equation*}
\textit{Hint:} Use the simulation lemma: for any two policies
$\policy_1, \policy_2$, the value difference satisfies
$\|Q^{\policy_1} - Q^{\policy_2}\|_\infty \leq
\frac{1}{1-\discount} \max_s \sum_a |\policy_1(a|s)
- \policy_2(a|s)| \cdot R_{\max}/(1-\discount)$.
\end{exercise}

\begin{exercise}[SARSA Convergence Without GLIE]
\textnormal{(\(***\))} \\
Suppose $\varepsilon_t = \varepsilon > 0$ is constant, but we
modify the SARSA update to use a decaying ``effective exploration''
$\tilde{\varepsilon}_t \to 0$ only in the bootstrap target while
still executing the $\varepsilon$-greedy policy for data collection.
Formally, the update becomes:
\[
  Q(S_t,A_t) \leftarrow Q(S_t,A_t)
  + \alpha_t\bigl[R_{t+1} + \discount Q(S_{t+1}, \tilde{A}_{t+1})
  - Q(S_t,A_t)\bigr],
\]
where $\tilde{A}_{t+1}$ is drawn from a
$\tilde{\varepsilon}_t$-greedy policy. Prove that this modified
SARSA converges to $\Qstar$ and identify the sense in which it
becomes off-policy.
\end{exercise}

%% SECTION 9: SUMMARY
\section{Summary}
\label{sec:summary}

\begin{itemize}
  \item \textbf{From prediction to control}: GPI alternates between
    policy evaluation and improvement; model-free control requires
    learning action-value functions $Q(s,a)$ rather than state values
    $V(s)$.

  \item The \textbf{$\varepsilon$-greedy policy} assigns probability
    $1 - \varepsilon + \varepsilon/|\Actions|$ to the greedy action
    and $\varepsilon/|\Actions|$ to all others, ensuring exploration
    while approximately following the greedy policy.

  \item The \textbf{GLIE condition} requires infinite exploration of
    all state--action pairs and convergence of the policy to greedy
    in the limit, typically achieved by $\varepsilon_t \to 0$.

  \item The \textbf{SARSA update}
    $Q(S_t,A_t) \leftarrow Q(S_t,A_t) + \alpha[\delta_t]$ with
    TD error $\delta_t = R_{t+1} + \discount Q(S_{t+1}, A_{t+1})
    - Q(S_t, A_t)$ applies TD(0) to action values, using the
    \emph{actually selected} next action $A_{t+1}$.

  \item \textbf{SARSA converges to $\Qstar$} under GLIE
    (Theorem~\ref{thm:sarsa}), proved by decomposing the update
    into a Q-learning component plus a vanishing exploration
    perturbation that is summable under $\varepsilon_t \to 0$.

  \item SARSA is \textbf{on-policy}: it evaluates the policy it
    actually follows, including exploration. With fixed
    $\varepsilon$, it converges to $Q^{\policy_\varepsilon}$
    rather than $\Qstar$, which can yield safer behavior in
    environments where exploratory actions are costly.

  \item The action-value Bellman operators
    $\Bellman^\policy_Q$ and $\Bellman^*_Q$ are both
    $\discount$-contractions in $\|\cdot\|_\infty$
    (Lemma~\ref{lem:bellman_q_contraction}), providing the stability
    condition for all TD control convergence proofs.
\end{itemize}

%% SECTION 10: FURTHER READING
\section{Further Reading}
\label{sec:further_reading}

\begin{itemize}
  \item \textbf{Lesson 50} develops Q-learning (off-policy TD
    control), Expected SARSA, and provides a unified view of all
    three TD control algorithms. The comparison between SARSA and
    Q-learning deepens the on-policy vs.\ off-policy distinction
    introduced here.

  \item \citet{singh2000} is the primary reference for the SARSA
    convergence proof presented in Theorem~\ref{thm:sarsa}. Their
    framework extends to general single-step on-policy algorithms
    satisfying GLIE.

  \item \citet[Section 6.5]{suttonbarto2018} presents the
    cliff-walking example, the most widely cited illustration of
    the behavioral difference between on-policy SARSA and off-policy
    Q-learning.

  \item \citet{rummery1994} introduced SARSA (under the name
    ``modified connectionist Q-learning'') as an online alternative
    to Q-learning for use with neural network function approximators.

  \item \textbf{SARSA($\lambda$)} combines SARSA with eligibility
    traces from Lesson 48, yielding faster credit assignment. See
    \citet[Chapter 12]{suttonbarto2018} for the algorithm and
    \citet[Chapter 5]{bertsekas1996} for convergence analysis.

  \item For the extension to continuous state spaces with function
    approximation, see Lessons 51--53. The ``deadly triad'' of
    function approximation, bootstrapping, and off-policy learning
    is less severe for on-policy SARSA than for off-policy
    Q-learning.
\end{itemize}

\bibliography{td-refs}

\end{document}
