\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 22: Fixed Point Theorems}}
\fancyhead[R]{\thepage}
\title{Lesson 22: Fixed Point Theorems}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% ============================================================
%% SECTION 1: MOTIVATION AND OVERVIEW
%% ============================================================

\section{Motivation and Overview}
\label{sec:motivation}

The equation $T(x) = x$---find a point that a map sends to itself---is
one of the most universal formulations in mathematics.  Differential
equations, integral equations, game-theoretic equilibria, and the
Bellman equation of dynamic programming can all be cast as fixed point
problems.  The theorems in this lesson provide conditions under which
such equations have solutions, and in the most favorable case (the
Banach contraction theorem) they also supply an algorithm and a
convergence rate.

We develop four fixed point theorems of increasing generality:
\begin{enumerate}[leftmargin=2em,nosep]
    \item The \textbf{Banach Contraction Mapping Theorem}, which
          guarantees existence, uniqueness, and geometric convergence
          of iterates for contractions on complete metric spaces.
    \item The \textbf{Brouwer Fixed Point Theorem}, which guarantees
          existence (but not uniqueness) for continuous self-maps of
          compact convex subsets of $\R^n$.
    \item The \textbf{Schauder Fixed Point Theorem}, which extends
          Brouwer to infinite-dimensional Banach spaces by requiring
          the map to be compact.
    \item The \textbf{Kakutani Fixed Point Theorem}, which extends
          Brouwer to set-valued maps (correspondences), providing the
          standard route to Nash equilibrium existence.
\end{enumerate}

\begin{rlconnection}[title={Fixed Point Theorems in Reinforcement Learning}]
\begin{enumerate}[leftmargin=2em,nosep]
    \item \textbf{Value iteration.}  The Bellman optimality operator
          $T$ is a $\gamma$-contraction on the space of bounded value
          functions.  By the Banach contraction theorem, the iterates
          $V_{n+1} = TV_n$ converge geometrically to the unique
          optimal value function $V^*$, regardless of initialization.
    \item \textbf{Policy iteration.}  The policy evaluation step
          solves $V^\pi = T^\pi V^\pi$, which is again a contraction
          fixed point problem.
    \item \textbf{Stationary distributions.}  Brouwer's and
          Schauder's theorems guarantee the existence of stationary
          distributions for transition operators on compact state
          spaces, since the simplex of probability distributions is
          compact and convex.
    \item \textbf{Nash equilibrium.}  The best-response
          correspondence in a finite game satisfies the hypotheses of
          Kakutani's theorem on the product of mixed strategy
          simplices.  The resulting fixed point is a Nash equilibrium
          ---the foundational existence result of game theory and
          multi-agent RL.
\end{enumerate}
\end{rlconnection}

\medskip
\noindent\textbf{Prerequisites.}  Metric spaces, completeness, and
compactness (Lessons 15 and 18).  Banach spaces and bounded linear
operators (Lessons 16--17).  The Baire Category Theorem (Lesson 18)
is used in one remark but is not essential.

\medskip
\noindent\textbf{Outline.}
Section~\ref{sec:references} lists references.
Section~\ref{sec:banach} proves the Banach Contraction Mapping
Theorem with the geometric convergence rate.
Section~\ref{sec:brouwer} proves the Brouwer Fixed Point Theorem via
the retraction argument.
Section~\ref{sec:schauder} extends Brouwer to infinite dimensions
through the Schauder Fixed Point Theorem.
Section~\ref{sec:kakutani} proves the Kakutani Fixed Point Theorem
for set-valued maps.
Section~\ref{sec:hierarchy} summarizes the relationship between the
four theorems.
Section~\ref{sec:exercises} provides exercises.
Section~\ref{sec:summary} summarizes the key results.
Section~\ref{sec:further} offers further reading.

%% ============================================================
%% SECTION 2: REFERENCES
%% ============================================================

\section{References}
\label{sec:references}

The Banach contraction mapping theorem is proved in
\citet{kreyszig1989}, Section~5.1, and \citet{granas2003}, Chapter~1.
The original paper is \citet{banach1922}.  Brouwer's theorem appears
in \citet{granas2003}, Chapters~6--10, with the original formulation
in \citet{brouwer1911}.  The Schauder theorem is presented in
\citet{granas2003}, Chapter~8; the original is
\citet{schauder1930}.  Kakutani's theorem and its application to game
theory are in \citet{kakutani1941} and \citet{granas2003},
Chapter~10.  For the connection to Markov decision processes, see
\citet{puterman2014}, Chapter~6.  General treatments of fixed point
theory in functional analysis appear in \citet{brezis2011},
\citet{rudin1991}, and \citet{dunford1988}.

%% ============================================================
%% SECTION 3: BANACH CONTRACTION MAPPING THEOREM
%% ============================================================

\section{The Banach Contraction Mapping Theorem}
\label{sec:banach}

\begin{definition}[Contraction Mapping]
\label{def:contraction}
Let $(X, d)$ be a metric space.  A map $T \colon X \to X$ is a
\emph{contraction} (or \emph{contraction mapping}) if there exists a
constant $\gamma \in [0, 1)$ such that
\[
    d(Tx, Ty) \leq \gamma \, d(x, y)
    \quad \text{for all } x, y \in X.
\]
The constant $\gamma$ is called the \emph{contraction factor}
(or \emph{Lipschitz constant}).  A contraction is automatically
continuous (in fact, Lipschitz continuous).
\end{definition}

\begin{remark}
\label{rem:nonexpansive}
A map satisfying $d(Tx, Ty) \leq d(x, y)$ (i.e., $\gamma = 1$) is
called \emph{nonexpansive}.  Nonexpansive maps on complete metric
spaces need not have fixed points (consider a translation on $\R$),
and even when they do, iterates need not converge.  The strict
inequality $\gamma < 1$ is essential for the contraction theorem.
\end{remark}

\begin{theorem}[Banach Contraction Mapping Theorem]
\label{thm:banach}
Let $(X, d)$ be a nonempty complete metric space and
$T \colon X \to X$ a contraction with factor $\gamma \in [0, 1)$.
Then:
\begin{enumerate}[label=(\roman*)]
    \item $T$ has a unique fixed point $x^* \in X$ (i.e.,
          $Tx^* = x^*$).
    \item For any initial point $x_0 \in X$, the iterates
          $x_{n+1} = Tx_n$ converge to $x^*$.
    \item The convergence is geometric: for all $n \geq 0$,
          \[
              d(x_n, x^*) \leq \frac{\gamma^n}{1 - \gamma}
              \, d(x_0, Tx_0).
          \]
\end{enumerate}
\end{theorem}

\begin{proof}
\textbf{Existence and convergence.}
Fix $x_0 \in X$ and define $x_{n+1} = Tx_n$ for $n \geq 0$.
By the contraction property applied inductively,
\[
    d(x_{n+1}, x_n) = d(Tx_n, Tx_{n-1})
    \leq \gamma\, d(x_n, x_{n-1})
    \leq \cdots \leq \gamma^n\, d(x_1, x_0).
\]
For $m > n \geq 0$, the triangle inequality gives
\begin{align}
    d(x_m, x_n)
    &\leq \sum_{k=n}^{m-1} d(x_{k+1}, x_k)
    \leq \sum_{k=n}^{m-1} \gamma^k\, d(x_1, x_0)
    \notag\\
    &\leq \gamma^n \sum_{j=0}^{\infty} \gamma^j\, d(x_1, x_0)
    = \frac{\gamma^n}{1 - \gamma}\, d(x_1, x_0).
    \label{eq:cauchy-estimate}
\end{align}
Since $\gamma < 1$, the right-hand side tends to $0$ as
$n \to \infty$, so $(x_n)$ is a Cauchy sequence.  By completeness of
$X$, there exists $x^* \in X$ with $x_n \to x^*$.

Since $T$ is Lipschitz (hence continuous),
\[
    Tx^* = T\!\left(\lim_{n \to \infty} x_n\right)
    = \lim_{n \to \infty} Tx_n
    = \lim_{n \to \infty} x_{n+1}
    = x^*.
\]

\textbf{Uniqueness.}
Suppose $Tx^* = x^*$ and $Ty^* = y^*$.  Then
$d(x^*, y^*) = d(Tx^*, Ty^*) \leq \gamma\, d(x^*, y^*)$.
Since $\gamma < 1$, this forces $(1 - \gamma)\, d(x^*, y^*) \leq 0$,
so $d(x^*, y^*) = 0$ and $x^* = y^*$.

\textbf{Geometric convergence rate.}
Let $m \to \infty$ in \eqref{eq:cauchy-estimate}:
\[
    d(x_n, x^*)
    = \lim_{m \to \infty} d(x_m, x_n)
    \leq \frac{\gamma^n}{1 - \gamma}\, d(x_1, x_0)
    = \frac{\gamma^n}{1 - \gamma}\, d(x_0, Tx_0).
    \qedhere
\]
\end{proof}

\begin{keyresult}[title={Banach Contraction Mapping Theorem}]
In a complete metric space, every contraction $T$ with factor
$\gamma < 1$ has a \emph{unique} fixed point $x^*$.  Iterating $T$
from any starting point converges to $x^*$ at the geometric rate
$d(x_n, x^*) \leq \frac{\gamma^n}{1 - \gamma}\, d(x_0, Tx_0)$.
After $n \geq \frac{1}{1 - \gamma} \log \frac{d(x_0, Tx_0)}
{(1-\gamma)\varepsilon}$ iterations,
$d(x_n, x^*) \leq \varepsilon$.
\end{keyresult}

\begin{rlconnection}[title={Value Iteration as a Contraction}]
Let $\mathcal{S}$ be a finite state space and consider the Banach
space $(B(\mathcal{S}), \|\cdot\|_\infty)$ of bounded real-valued
functions on $\mathcal{S}$.  The \textbf{Bellman optimality operator}
\[
    (TV)(s) = \max_{a \in \mathcal{A}}
    \Bigl[ R(s,a) + \gamma \sum_{s'} P(s' \mid s, a)\, V(s') \Bigr]
\]
is a $\gamma$-contraction in $\|\cdot\|_\infty$ (this is proved in
Exercise~\ref{ex:bellman}).  By the Banach contraction theorem:
\begin{itemize}[nosep]
    \item There exists a unique $V^*$ with $TV^* = V^*$ (the Bellman
          equation).
    \item Value iteration $V_{n+1} = TV_n$ converges to $V^*$ from
          any initialization.
    \item The iteration count for $\varepsilon$-optimality is
          $O\!\left(\frac{1}{1-\gamma}
          \log \frac{1}{(1-\gamma)\varepsilon}\right)$.
\end{itemize}
Policy evaluation ($T^\pi$ instead of $T$) is likewise a contraction,
so the same guarantees apply.
\end{rlconnection}

\begin{example}[Picard--Lindelof via Contraction]
\label{ex:picard}
Consider the initial value problem $y'(t) = f(t, y(t))$, $y(0) = y_0$,
where $f$ is Lipschitz in $y$ with constant $L$.  The equivalent
integral equation $y(t) = y_0 + \int_0^t f(s, y(s))\, ds$ defines an
operator $T$ on $C([0, h])$ for $h > 0$.  On a sufficiently short
interval ($h < 1/L$), the operator $T$ is a contraction in
$\|\cdot\|_\infty$, giving existence and uniqueness of the solution.
More precisely, by using the weighted norm
$\|y\|_w = \sup_{t \in [0,h]} e^{-2Lt} |y(t)|$, one can show $T$ is
a contraction with factor $1/2$ on any interval, thereby recovering
the global Picard--Lindelof theorem.
\end{example}

%% ============================================================
%% SECTION 4: BROUWER FIXED POINT THEOREM
%% ============================================================

\section{The Brouwer Fixed Point Theorem}
\label{sec:brouwer}

The Banach contraction theorem requires a quantitative condition
($\gamma < 1$) that is too strong for many applications.  The Brouwer
theorem drops the contraction requirement entirely, asking only that
the map be continuous and the domain be compact and convex.  The price
is that we lose uniqueness and any constructive iteration scheme.

\begin{theorem}[Brouwer Fixed Point Theorem]
\label{thm:brouwer}
Let $C \subseteq \R^n$ be a nonempty compact convex set and
$f \colon C \to C$ a continuous map.  Then $f$ has a fixed point:
there exists $x^* \in C$ with $f(x^*) = x^*$.
\end{theorem}

We prove the theorem for $C = \overline{B}^n = \{x \in \R^n :
\|x\| \leq 1\}$ (the closed unit ball).  The general case follows
because every nonempty compact convex subset of $\R^n$ is
homeomorphic to a closed ball of the appropriate dimension.

The proof proceeds by contradiction, using the no-retraction lemma.

\begin{definition}[Retraction]
\label{def:retraction}
Let $A \subseteq X$ be a subspace of a topological space $X$.  A
\emph{retraction} of $X$ onto $A$ is a continuous map
$r \colon X \to A$ such that $r(a) = a$ for all $a \in A$.
\end{definition}

\begin{lemma}[No-Retraction Lemma]
\label{lem:no-retraction}
There is no retraction of the closed ball $\overline{B}^n$ onto the
sphere $S^{n-1} = \{x \in \R^n : \|x\| = 1\}$.
\end{lemma}

\begin{proof}
We give the proof for $n = 2$ via a smooth approximation and degree
argument; the general case is analogous.

Suppose $r \colon \overline{B}^2 \to S^1$ is a retraction.  By the
Weierstrass approximation theorem, $r$ can be uniformly approximated
by smooth maps.  Composing with the normalization map
$x \mapsto x / \|x\|$ (which is smooth away from the origin), one
obtains a smooth map $\tilde{r} \colon \overline{B}^2 \to S^1$ with
$\tilde{r}|_{S^1}$ homotopic to the identity on $S^1$.

The \emph{degree} of the identity map $\mathrm{id} \colon S^1 \to S^1$
is $1$.  However, any smooth map $\overline{B}^2 \to S^1$ that
extends to the whole disk must have degree $0$ on the boundary
(since the disk is contractible, any map from it to $S^1$ can be
lifted to a map into $\R$, and such lifted maps have winding number
$0$).

Since $\tilde{r}|_{S^1}$ is homotopic to $\mathrm{id}_{S^1}$ (which
has degree $1$) but must have degree $0$, we reach a contradiction.
Hence no such retraction exists.
\end{proof}

\begin{proof}[Proof of the Brouwer Fixed Point Theorem
  (Theorem~\ref{thm:brouwer})]
Suppose for contradiction that $f(x) \neq x$ for all
$x \in \overline{B}^n$.  For each $x \in \overline{B}^n$, consider
the ray starting at $f(x)$ and passing through $x$.  Since
$f(x) \neq x$, this ray is well-defined.  Let $r(x)$ be the unique
point where this ray intersects the sphere $S^{n-1}$ on the side of
$x$ (i.e., $r(x) = x + t(x - f(x))$ for the unique $t \geq 0$ with
$\|r(x)\| = 1$).

\textbf{Continuity of $r$.}
The parameter $t = t(x)$ is determined by the equation
$\|x + t(x - f(x))\|^2 = 1$, which expands to a quadratic in $t$
with coefficients depending continuously on $x$ (since $f$ is
continuous).  The discriminant is strictly positive (because $f(x)
\neq x$ guarantees the ray from $f(x)$ through $x$ exits the ball),
so $t(x)$ is a continuous function of $x$ by the continuous
dependence of roots of polynomials on their coefficients.  Hence $r$
is continuous.

\textbf{$r$ is a retraction.}
For $x \in S^{n-1}$, we have $\|x\| = 1$, so $t = 0$ satisfies
$\|x + 0 \cdot (x - f(x))\| = \|x\| = 1$.  Thus $r(x) = x$ for all
$x \in S^{n-1}$.

Therefore $r \colon \overline{B}^n \to S^{n-1}$ is a retraction,
contradicting Lemma~\ref{lem:no-retraction}.  The assumption that $f$
has no fixed point must be false.
\end{proof}

\begin{remark}
\label{rem:brouwer-nonconstructive}
The Brouwer theorem is a pure existence result and provides no
algorithm for finding the fixed point.  Computing approximate Brouwer
fixed points is PPAD-complete---widely believed to be intractable.
When the contraction structure of the Banach theorem is available, it
provides the far more useful algorithmic guarantee.
\end{remark}

\begin{rlconnection}[title={Stationary Distributions via Brouwer}]
Let $P$ be the transition matrix of a Markov chain on a finite state
space $\mathcal{S}$ with $|\mathcal{S}| = n$.  A stationary
distribution is a vector $\mu \in \Delta_n$ (the probability simplex)
satisfying $\mu^\top P = \mu^\top$.  Equivalently, $\mu$ is a fixed
point of the map $f(\mu) = \mu^\top P$ restricted to $\Delta_n$.
Since $\Delta_n$ is a nonempty compact convex subset of $\R^n$ and
$f$ is linear (hence continuous) with $f(\Delta_n) \subseteq
\Delta_n$ (the image of a probability vector under a stochastic
matrix is again a probability vector), the Brouwer theorem guarantees
the existence of at least one stationary distribution.
\end{rlconnection}

%% ============================================================
%% SECTION 5: SCHAUDER FIXED POINT THEOREM
%% ============================================================

\section{The Schauder Fixed Point Theorem}
\label{sec:schauder}

In infinite-dimensional Banach spaces, the closed unit ball is not
compact (by Riesz's lemma), so the Brouwer theorem does not directly
apply.  The Schauder theorem overcomes this obstacle by requiring the
map to be \emph{compact}---a condition that recovers enough
finite-dimensional structure to invoke Brouwer.

\begin{definition}[Compact Operator]
\label{def:compact-op}
Let $X$ be a Banach space and $C \subseteq X$.  A map
$T \colon C \to X$ is \emph{compact} if $T$ is continuous and maps
bounded subsets of $C$ to relatively compact subsets of $X$ (i.e.,
subsets whose closure is compact).
\end{definition}

\begin{remark}
\label{rem:compact-finite-rank}
Every continuous map with finite-dimensional range is compact (since
bounded subsets of finite-dimensional spaces are relatively compact).
Compact operators are the ``next best thing'' to
finite-dimensional operators, and they arise naturally as integral
operators with continuous kernels (by the Arzela--Ascoli theorem).
\end{remark}

\begin{theorem}[Schauder Fixed Point Theorem]
\label{thm:schauder}
Let $X$ be a Banach space, $C \subseteq X$ a nonempty closed bounded
convex set, and $T \colon C \to C$ a compact map.  Then $T$ has a
fixed point.
\end{theorem}

\begin{proof}
Since $T$ is compact and $C$ is bounded, the set $K = \overline{T(C)}$
is compact.  We approximate $T$ by finite-dimensional maps and apply
the Brouwer theorem.

\textbf{Step 1: Schauder projections.}
For each $n \in \N$, cover $K$ by finitely many open balls of radius
$1/n$: there exist points $y_1^{(n)}, \ldots, y_{k_n}^{(n)} \in T(C)$
such that $K \subseteq \bigcup_{i=1}^{k_n} B(y_i^{(n)}, 1/n)$.
Define the weight functions
\[
    \lambda_i^{(n)}(x)
    = \max\!\bigl(0,\; 1/n - \|x - y_i^{(n)}\|\bigr),
    \qquad i = 1, \ldots, k_n.
\]
For $x \in K$, at least one ball covers $x$, so
$\sum_{i=1}^{k_n} \lambda_i^{(n)}(x) > 0$.  The
\emph{Schauder projection} $P_n \colon K \to X$ is
\[
    P_n(x)
    = \frac{\sum_{i=1}^{k_n} \lambda_i^{(n)}(x)\, y_i^{(n)}}
           {\sum_{i=1}^{k_n} \lambda_i^{(n)}(x)}.
\]
Then $P_n$ is continuous, its range lies in
$C_n = \conv\{y_1^{(n)}, \ldots, y_{k_n}^{(n)}\} \subseteq C$ (since
$C$ is convex and $y_i^{(n)} \in T(C) \subseteq C$), and
\begin{equation}
\label{eq:schauder-approx}
    \|P_n(x) - x\| < 1/n
    \quad \text{for all } x \in K.
\end{equation}
Indeed, $\lambda_i^{(n)}(x) > 0$ only when
$\|x - y_i^{(n)}\| < 1/n$, so $P_n(x)$ is a convex combination of
points within distance $1/n$ of $x$.

\textbf{Step 2: Finite-dimensional fixed points.}
The set $C_n$ is a compact convex subset of a finite-dimensional
subspace of $X$.  Define $T_n \colon C_n \to C_n$ by
$T_n = P_n \circ T|_{C_n}$.  This is a continuous self-map of a
compact convex set in finite dimensions.  By the Brouwer Fixed Point
Theorem~\ref{thm:brouwer}, there exists $x_n \in C_n$ with
$T_n(x_n) = x_n$, i.e.,
\[
    x_n = P_n(Tx_n).
\]
By~\eqref{eq:schauder-approx},
$\|x_n - Tx_n\| = \|P_n(Tx_n) - Tx_n\| < 1/n$.

\textbf{Step 3: Passing to the limit.}
The sequence $(x_n)_{n \geq 1}$ lies in $C$ (which is bounded), so
$(Tx_n)_{n \geq 1}$ lies in $T(C)$, whose closure $K$ is compact.
Hence $(Tx_n)$ has a convergent subsequence, say
$Tx_{n_j} \to y \in K$.  Since
$\|x_{n_j} - Tx_{n_j}\| < 1/n_j \to 0$, we also have
$x_{n_j} \to y$.  Since $C$ is closed, $y \in C$.  By continuity
of $T$,
\[
    Ty = T\!\left(\lim_{j \to \infty} x_{n_j}\right)
    = \lim_{j \to \infty} Tx_{n_j} = y.
    \qedhere
\]
\end{proof}

\begin{remark}
\label{rem:schauder-variants}
A more general version states: if $C$ is a nonempty convex subset of
a Banach space (not necessarily closed or bounded) and
$T \colon C \to C$ is continuous with $T(C)$ relatively compact, then
$T$ has a fixed point.  The proof is similar---the closure of $T(C)$
is compact, and the argument of Steps 1--3 applies with minor
modifications.
\end{remark}

\begin{rlconnection}[title={Schauder and Continuous MDPs}]
In continuous-state MDPs, the state space $\mathcal{S}$ is a compact
subset of $\R^d$ and value functions live in the Banach space
$C(\mathcal{S})$.  When the transition kernel has a continuous
density, the Bellman evaluation operator $T^\pi$ maps bounded sets of
continuous functions to equicontinuous families (by the integral
smoothing effect), so $T^\pi$ is compact by the Arzela--Ascoli
theorem.  The Schauder theorem then guarantees the existence of a
fixed point $V^\pi \in C(\mathcal{S})$---the value function of
policy $\pi$.  When the discount factor $\gamma < 1$, the Banach
contraction theorem also applies and gives the stronger conclusion of
uniqueness and geometric convergence; the Schauder theorem is
relevant in undiscounted settings where the contraction property
fails.
\end{rlconnection}

%% ============================================================
%% SECTION 6: KAKUTANI FIXED POINT THEOREM
%% ============================================================

\section{The Kakutani Fixed Point Theorem}
\label{sec:kakutani}

The Brouwer and Schauder theorems apply to single-valued maps.  In
game theory and mathematical economics, one frequently encounters
\emph{set-valued maps} (correspondences), such as the best-response
correspondence that assigns to each opponent strategy profile the set
of optimal strategies.  Kakutani's theorem extends Brouwer's result to
this setting.

\begin{definition}[Set-Valued Map / Correspondence]
\label{def:correspondence}
A \emph{set-valued map} (or \emph{correspondence}) from a set $X$ to
a set $Y$ is a map $\Phi \colon X \rightrightarrows Y$ that assigns
to each $x \in X$ a subset $\Phi(x) \subseteq Y$.  A \emph{fixed
point} of $\Phi \colon X \rightrightarrows X$ is a point $x^*$ with
$x^* \in \Phi(x^*)$.
\end{definition}

\begin{definition}[Upper Hemicontinuity]
\label{def:uhc}
A set-valued map $\Phi \colon X \rightrightarrows Y$ between
topological spaces is \emph{upper hemicontinuous} (u.h.c.) if for
every open set $V \subseteq Y$, the upper inverse
$\Phi^+(V) = \{x \in X : \Phi(x) \subseteq V\}$ is open in $X$.

Equivalently, when $X$ and $Y$ are metric spaces and $\Phi$ has
compact values: $\Phi$ is u.h.c.\ if and only if it has a closed
graph, meaning that $x_n \to x$, $y_n \in \Phi(x_n)$, and
$y_n \to y$ together imply $y \in \Phi(x)$.
\end{definition}

\begin{remark}
\label{rem:lhc}
A correspondence is \emph{lower hemicontinuous} if for every open
$V \subseteq Y$, the lower inverse
$\Phi^-(V) = \{x \in X : \Phi(x) \cap V \neq \emptyset\}$ is open.
Kakutani's theorem requires upper hemicontinuity; lower
hemicontinuity is not needed.
\end{remark}

\begin{theorem}[Kakutani Fixed Point Theorem]
\label{thm:kakutani}
Let $C \subseteq \R^n$ be a nonempty compact convex set and
$\Phi \colon C \rightrightarrows C$ a set-valued map satisfying:
\begin{enumerate}[label=(\roman*)]
    \item For every $x \in C$, the set $\Phi(x)$ is nonempty and
          convex.
    \item $\Phi$ is upper hemicontinuous.
\end{enumerate}
Then $\Phi$ has a fixed point: there exists $x^* \in C$ with
$x^* \in \Phi(x^*)$.
\end{theorem}

\begin{proof}
We reduce to the Brouwer theorem by approximating $\Phi$ with
continuous single-valued maps.

\textbf{Step 1: Approximate selections.}
For each $k \in \N$, since $C$ is compact, we can cover it by finitely
many open balls $\{B(x_i, 1/k)\}_{i=1}^{m_k}$ with centers
$x_i \in C$.  Let $\{\lambda_i\}_{i=1}^{m_k}$ be a continuous
partition of unity subordinate to this cover, i.e., each
$\lambda_i \colon C \to [0,1]$ is continuous,
$\supp(\lambda_i) \subseteq B(x_i, 1/k)$, and
$\sum_{i=1}^{m_k} \lambda_i(x) = 1$ for all $x \in C$.

For each $i$, choose any $y_i \in \Phi(x_i)$ (possible since
$\Phi(x_i)$ is nonempty).  Define the continuous function
\[
    f_k(x) = \sum_{i=1}^{m_k} \lambda_i(x)\, y_i.
\]
Since each $y_i \in C$ and $C$ is convex, $f_k(x)$ is a convex
combination of points in $C$, so $f_k \colon C \to C$.

\textbf{Step 2: Apply the Brouwer theorem.}
By Theorem~\ref{thm:brouwer}, each $f_k$ has a fixed point
$x_k \in C$ with $f_k(x_k) = x_k$.

\textbf{Step 3: Pass to the limit.}
Since $C$ is compact, $(x_k)$ has a convergent subsequence.  Passing
to this subsequence (still denoted $(x_k)$), let $x_k \to x^*$.  We
show $x^* \in \Phi(x^*)$.

Let $V$ be any open convex set with $\Phi(x^*) \subseteq V$.  By
upper hemicontinuity, $\Phi^+(V) = \{x \in C : \Phi(x) \subseteq V\}$
is open and contains $x^*$.  Since $x_k \to x^*$, for all
sufficiently large $k$:
\begin{itemize}[nosep]
    \item $x_k \in \Phi^+(V)$, so $\Phi(x_k) \subseteq V$.
    \item $\|x_k - x^*\| < 1/k$.
\end{itemize}
For such large $k$, if $\lambda_i(x_k) > 0$ then
$\|x_k - x_i\| < 1/k$, hence $\|x_i - x^*\| < 2/k$.  For $k$
large enough, $x_i$ lies in $\Phi^+(V)$ as well, so
$\Phi(x_i) \subseteq V$, and in particular $y_i \in V$.

Therefore $x_k = f_k(x_k) = \sum_i \lambda_i(x_k)\, y_i$ is a convex
combination of points $y_i \in V$.  Since $V$ is convex, $x_k \in V$.

Since this holds for every open convex $V \supseteq \Phi(x^*)$ and
$x_k \to x^*$, we conclude $x^* \in \overline{V}$ for every such $V$.
Because $\Phi(x^*)$ is compact and convex,
\[
    x^* \in \bigcap
    \bigl\{\overline{V} : V \supseteq \Phi(x^*),\;
    V \text{ open and convex}\bigr\}
    = \Phi(x^*).
    \qedhere
\]
\end{proof}

\begin{rlconnection}[title={Nash Equilibrium Existence via Kakutani}]
Consider a finite $n$-player game.  Player $i$ has a finite action
set $A_i$ and a payoff function $u_i \colon \prod_{j=1}^n A_j \to \R$.
A \emph{mixed strategy} for player $i$ is $\sigma_i \in \Delta(A_i)$
(the simplex of probability distributions over $A_i$).  The
\emph{best-response correspondence} of player $i$ is
\[
    BR_i(\sigma_{-i})
    = \argmax_{\sigma_i \in \Delta(A_i)}
    \sum_{a \in A} \biggl(\prod_{j=1}^n \sigma_j(a_j)\biggr) u_i(a),
\]
where $\sigma_{-i} = (\sigma_j)_{j \neq i}$.  Define the joint
best-response $\Phi(\sigma) = \prod_{i=1}^n BR_i(\sigma_{-i})$ on the
compact convex set $C = \prod_{i=1}^n \Delta(A_i)$.

One verifies:
\begin{enumerate}[leftmargin=2em,nosep]
    \item Each $BR_i(\sigma_{-i})$ is nonempty and convex (it is the
          set of maximizers of a linear function over a simplex).
    \item $\Phi$ is upper hemicontinuous (by the Berge Maximum
          Theorem: the objective is continuous and the constraint set
          is constant).
\end{enumerate}

By Kakutani's theorem, $\Phi$ has a fixed point $\sigma^*$, which
is precisely a \textbf{Nash equilibrium}: every player is
best-responding to the others.  This is the original proof of
\citet{kakutani1941} as applied by Nash (1950).
\end{rlconnection}

%% ============================================================
%% SECTION 7: HIERARCHY OF FIXED POINT THEOREMS
%% ============================================================

\section{Hierarchy and Comparison}
\label{sec:hierarchy}

The four fixed point theorems form a natural hierarchy, summarized in
the following table.

\begin{center}
\small
\begin{tabular}{llll}
\toprule
\textbf{Theorem} & \textbf{Setting} & \textbf{Map requirement}
    & \textbf{Conclusion} \\
\midrule
Banach & Complete metric space
    & Contraction ($\gamma < 1$)
    & Unique FP + algorithm \\
Brouwer & Compact convex $\subseteq \R^n$
    & Continuous, single-valued
    & Existence \\
Schauder & Closed bdd.\ convex in Banach sp.
    & Compact (continuous + bdd $\to$ rel.\ cpt.)
    & Existence \\
Kakutani & Compact convex $\subseteq \R^n$
    & u.h.c., convex-valued corresp.
    & Existence \\
\bottomrule
\end{tabular}
\end{center}

\begin{remark}
\label{rem:hierarchy}
The logical dependencies are:
\[
    \text{Brouwer} \;\xrightarrow{\text{compact approx.}}\;
    \text{Schauder}
    \qquad \text{and} \qquad
    \text{Brouwer} \;\xrightarrow{\text{approx.\ selections}}\;
    \text{Kakutani}.
\]
The Banach theorem is independent of the others: it uses completeness
and a metric contraction condition, while the topological theorems
use compactness and convexity.  The Banach theorem gives uniqueness
and an algorithm; the others give only existence.
\end{remark}

\begin{warning}[title={Choosing the Right Theorem}]
\begin{itemize}[nosep]
    \item Use \textbf{Banach} when you have a complete metric space
          and a contraction---this is the case for value iteration
          with $\gamma < 1$.
    \item Use \textbf{Brouwer} for existence arguments in
          finite-dimensional compact convex sets (e.g., stationary
          distributions, equilibria in simplex-valued problems).
    \item Use \textbf{Schauder} when working in infinite-dimensional
          spaces with compact operators (e.g., integral operators on
          function spaces).
    \item Use \textbf{Kakutani} when the ``solution map'' is
          set-valued (e.g., best-response correspondences in game
          theory).
\end{itemize}
\end{warning}

%% ============================================================
%% SECTION 8: EXERCISES
%% ============================================================

\section{Exercises}
\label{sec:exercises}

\begin{exercise}[Bellman Operator Is a Contraction (**)]\label{ex:bellman}
Let $\mathcal{S}$ and $\mathcal{A}$ be finite sets,
$R \colon \mathcal{S} \times \mathcal{A} \to \R$ a reward function,
$P(\cdot \mid s, a)$ a transition kernel, and $\gamma \in (0, 1)$ a
discount factor.  Define the Bellman optimality operator $T$ on the
space $B(\mathcal{S})$ of bounded functions on $\mathcal{S}$ (equipped
with $\|\cdot\|_\infty$) by
\[
    (TV)(s)
    = \max_{a \in \mathcal{A}}
    \Bigl[ R(s,a) + \gamma \sum_{s' \in \mathcal{S}}
    P(s' \mid s, a)\, V(s') \Bigr].
\]
\begin{enumerate}[label=(\alph*)]
    \item Prove that $T$ is a $\gamma$-contraction:
          $\|TV - TW\|_\infty \leq \gamma\, \|V - W\|_\infty$.

          \emph{Hint:}  For each $s$, let $a_s^*$ achieve the maximum
          for $(TV)(s)$.  Then
          $(TV)(s) - (TW)(s) \leq \gamma \sum_{s'} P(s' \mid s, a_s^*)
          (V(s') - W(s'))$.  By symmetry, obtain the reverse
          inequality and take the supremum over $s$.
    \item Deduce from the Banach contraction theorem that value
          iteration converges geometrically and that $V^*$ is the
          unique solution of the Bellman equation $TV = V$.
    \item How many iterations $n$ are needed to guarantee
          $\|V_n - V^*\|_\infty \leq \varepsilon$?  Express the
          answer in terms of $\gamma$, $\varepsilon$, and
          $\|V_1 - V_0\|_\infty$.
\end{enumerate}
\end{exercise}

\begin{exercise}[Contractive Iterate (**)]\label{ex:contractive-iterate}
Let $(X, d)$ be a complete metric space and $T \colon X \to X$ a
continuous map such that $T^N$ is a contraction for some $N \geq 1$.
\begin{enumerate}[label=(\alph*)]
    \item Show that $T^N$ has a unique fixed point $x^*$.
    \item Show that $Tx^*$ is also a fixed point of $T^N$, and
          conclude $Tx^* = x^*$.
    \item Deduce that $T$ itself has a unique fixed point.
\end{enumerate}
\emph{Application:}  The Picard operator for an ODE on $[0, h]$ may
not be a contraction in $\|\cdot\|_\infty$, but $T^N$ is for $N$
large enough.  This exercise shows uniqueness still holds.
\end{exercise}

\begin{exercise}[Brouwer in One Dimension (*)]\label{ex:brouwer-1d}
\begin{enumerate}[label=(\alph*)]
    \item Prove the Brouwer Fixed Point Theorem for $n = 1$ directly:
          if $f \colon [a, b] \to [a, b]$ is continuous, then $f$ has
          a fixed point.

          \emph{Hint:}  Apply the intermediate value theorem to
          $g(x) = f(x) - x$.
    \item Show by example that the conclusion fails if $[a,b]$ is
          replaced by an open interval $(a, b)$.
    \item Show by example that the conclusion fails if $[a,b]$ is
          replaced by a non-convex compact set
          $K = [0, 1] \cup [2, 3]$.
\end{enumerate}
\end{exercise}

\begin{exercise}[Schauder and Integral Equations (**)]\label{ex:schauder-integral}
Let $K \colon [0,1] \times [0,1] \to \R$ be continuous and define
$T \colon C[0,1] \to C[0,1]$ by
$(Tf)(t) = \int_0^1 K(t,s)\, f(s)\, ds$.
\begin{enumerate}[label=(\alph*)]
    \item Show that $T$ maps the closed ball
          $\overline{B}(0, R) = \{f \in C[0,1] : \|f\|_\infty
          \leq R\}$ into itself for $R$ sufficiently large, provided
          we add a continuous inhomogeneity:
          $(Tf)(t) = g(t) + \int_0^1 K(t,s)\, f(s)\, ds$.
          Find the condition on $R$ in terms of $\|g\|_\infty$ and
          $\|K\|_\infty$.
    \item Show that $T(\overline{B}(0, R))$ is equicontinuous.

          \emph{Hint:} Use uniform continuity of $K$ on
          $[0,1]^2$.
    \item Conclude via the Arzela--Ascoli theorem that $T$ is a
          compact operator.
    \item Apply the Schauder theorem to deduce that the integral
          equation $f = Tf$ has a solution.
\end{enumerate}
\end{exercise}

\begin{exercise}[Kakutani and Two-Player Zero-Sum Games (**)]
\label{ex:kakutani-zerosum}
Let $A \in \R^{m \times n}$ be a payoff matrix for a two-player
zero-sum game.  Player 1 chooses $p \in \Delta_m$ and Player 2
chooses $q \in \Delta_n$, with payoff $p^\top A q$ to Player 1.
\begin{enumerate}[label=(\alph*)]
    \item Write the best-response correspondences $BR_1(q)$ and
          $BR_2(p)$ explicitly.
    \item Verify the hypotheses of the Kakutani theorem for the joint
          best-response
          $\Phi(p, q) = BR_1(q) \times BR_2(p)$ on
          $\Delta_m \times \Delta_n$.
    \item Conclude that a Nash equilibrium (saddle point) exists.
    \item Explain how this relates to the minimax theorem
          $\max_p \min_q p^\top A q = \min_q \max_p p^\top A q$.
\end{enumerate}
\end{exercise}

\begin{exercise}[No Fixed Point Without Compactness (*)]\label{ex:no-fp}
Give an example of a continuous map $f \colon C \to C$ on a closed
convex but \emph{unbounded} subset $C$ of $\R^n$ that has no fixed
point.  This shows that compactness in the Brouwer theorem cannot be
dropped.

\emph{Hint:} Consider a translation on $C = \R$.
\end{exercise}

\begin{exercise}[Policy Iteration and Contraction (***)]\label{ex:policy-iter}
In the MDP setting of Exercise~\ref{ex:bellman}, \emph{policy
iteration} generates a sequence of policies $\pi_0, \pi_1, \ldots$ by:
(i) evaluate $V^{\pi_k}$ as the unique fixed point of $T^{\pi_k}$
(where $(T^\pi V)(s) = R(s, \pi(s)) + \gamma \sum_{s'} P(s' \mid s,
\pi(s))\, V(s')$), and (ii) improve:
$\pi_{k+1}(s) = \argmax_{a} [R(s,a) + \gamma \sum_{s'} P(s' \mid
s, a)\, V^{\pi_k}(s')]$.
\begin{enumerate}[label=(\alph*)]
    \item Show that $T^{\pi}$ is a $\gamma$-contraction for any
          policy $\pi$, so $V^{\pi}$ is well-defined and unique.
    \item Show that $V^{\pi_{k+1}} \geq V^{\pi_k}$ componentwise.

          \emph{Hint:}
          $T^{\pi_{k+1}} V^{\pi_k} \geq T^{\pi_k} V^{\pi_k}
          = V^{\pi_k}$.  Now iterate.
    \item Conclude that policy iteration converges to $V^*$ in
          finitely many steps (there are finitely many deterministic
          policies).
\end{enumerate}
\end{exercise}

%% ============================================================
%% SECTION 9: SUMMARY
%% ============================================================

\section{Summary}
\label{sec:summary}

\begin{itemize}[leftmargin=2em,nosep]
    \item \textbf{Contraction.}  A map $T$ on a metric space with
          $d(Tx, Ty) \leq \gamma\, d(x, y)$ for some $\gamma < 1$.

    \item \textbf{Banach Contraction Mapping Theorem.}  On a complete
          metric space, every contraction has a unique fixed point.
          The iterates $x_{n+1} = Tx_n$ converge from any starting
          point at geometric rate
          $d(x_n, x^*) \leq \frac{\gamma^n}{1-\gamma}\, d(x_0, Tx_0)$.
          This is the theoretical foundation of value iteration.

    \item \textbf{Brouwer Fixed Point Theorem.}  Every continuous
          self-map of a nonempty compact convex subset of $\R^n$ has a
          fixed point.  Proved via the no-retraction lemma.

    \item \textbf{Compact operator.}  A continuous map that sends
          bounded sets to relatively compact sets.

    \item \textbf{Schauder Fixed Point Theorem.}  Every compact
          self-map of a nonempty closed bounded convex subset of a
          Banach space has a fixed point.  Proved by approximating
          with finite-dimensional maps (Schauder projections) and
          applying Brouwer.

    \item \textbf{Set-valued map (correspondence).}
          $\Phi \colon X \rightrightarrows Y$ assigns a subset
          $\Phi(x) \subseteq Y$ to each $x$.  Upper hemicontinuity
          requires $\{x : \Phi(x) \subseteq V\}$ to be open for
          every open $V$.

    \item \textbf{Kakutani Fixed Point Theorem.}  An upper
          hemicontinuous correspondence with nonempty convex values on
          a compact convex subset of $\R^n$ has a fixed point.  This
          is the theorem used to prove Nash equilibrium existence.

    \item \textbf{Hierarchy.}  Banach gives uniqueness and an
          algorithm; Brouwer/Schauder/Kakutani give only existence.
          Brouwer implies Schauder (via compact approximation) and
          Kakutani (via approximate selections).
\end{itemize}

%% ============================================================
%% SECTION 10: FURTHER READING
%% ============================================================

\section{Further Reading}
\label{sec:further}

The definitive reference on fixed point theory is
\citet{granas2003}, which covers all four theorems with detailed
proofs, historical notes, and numerous applications.  The Banach
contraction theorem in the context of metric spaces is also treated
in \citet{kreyszig1989}, Section 5.1, and \citet{rudin1991}.  For a
proof of the Brouwer theorem via Sperner's lemma (a combinatorial
alternative to the retraction argument), see
\citet{granas2003}, Chapter 6.

The connection between the Banach contraction theorem and Markov
decision processes is developed rigorously in
\citet{puterman2014}, Chapter 6.  The application of Kakutani's
theorem to Nash equilibrium existence follows \citet{kakutani1941}.

The preceding lesson (Lesson 21) proved the four fundamental theorems
of functional analysis (Hahn--Banach, Uniform Boundedness, Open
Mapping, Closed Graph).  The next topic in the curriculum (Lesson 23)
begins the optimization sequence with convex sets and their
structural properties.

\bibliography{functional-analysis-refs}

\end{document}
