\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 65: Depth-Width Tradeoffs}}
\fancyhead[R]{\thepage}
\title{Lesson 65: Depth-Width Tradeoffs}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% ============================================================
%% SECTION 1: MOTIVATION AND OVERVIEW
%% ============================================================
\section{Motivation and Overview}
\label{sec:motivation}

Lessons~63 and~64 established that single-hidden-layer networks are
universal approximators, and that for functions in the Barron class the
approximation rate is $O(1/m)$, independent of dimension.  However,
modern deep learning practice uses networks with many layers---not just
one.  Is depth merely an engineering convenience, or does it provide a
fundamental mathematical advantage?  Consider training a neural network
to approximate a function that exhibits hierarchical or compositional
structure, such as the mapping from pixels to object categories in a
vision system.  Empirically, deep networks vastly outperform shallow
ones on such tasks, even when the total number of parameters is the
same.  This raises a precise question: \emph{are there functions that
deep networks with polynomial parameters can compute, but that shallow
networks require exponentially many parameters to approximate?}

Telgarsky's 2016 depth separation theorem provides a striking
affirmative answer: there exist explicit functions computable by
$O(L)$-layer networks of constant width that cannot be approximated to
constant accuracy by any network of depth $O(1)$ and width less than
$2^{\Omega(L)}$.  Complementing this, Lu et al.\ (2017) and Hanin
(2019) show that \emph{width} also matters: ReLU networks of width
$d+4$ (for inputs in $\R^d$) are universal approximators regardless of
depth.  The tools developed in this lesson will let us understand
precisely how depth and width contribute to the expressive power of
neural networks, quantify the gains from depth for structured functions,
and connect these results to classical approximation rates.

\begin{rlconnection}[title={Why Depth-Width Tradeoffs Matter for RL/ML}]
\begin{itemize}[nosep]
  \item \textbf{Architecture design in deep RL.}  Choosing the number of
    layers and width per layer in a DQN, PPO, or SAC network is a design
    decision.  Depth separation results explain why deeper architectures
    can represent value functions with hierarchical structure more
    efficiently.
  \item \textbf{Compositional structure in value functions.}  Many
    environments (e.g., Atari games, robotic manipulation) produce value
    functions with compositional structure $V^* = g_L \circ \cdots
    \circ g_1$.  Deep networks can exploit this structure to avoid the
    curse of dimensionality, while shallow networks cannot.
  \item \textbf{Parameter efficiency.}  In resource-constrained settings
    (embedded RL agents), understanding the depth-width tradeoff helps
    choose the most parameter-efficient architecture.
  \item \textbf{Linear regions and decision boundaries.}  The number of
    linear regions in a ReLU network---exponential in depth but
    polynomial in width---determines the complexity of the decision
    boundaries the network can represent.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
Single-hidden-layer networks and the universal approximation theorem
(Lesson~63, Theorems~4.1 and~5.2).  Barron's approximation theorem
and the Barron class (Lesson~64, Theorem~4.1).  Piecewise-affine
functions and the ReLU activation (Lesson~63, Example~5.5).  Basic
properties of Lipschitz functions and $L^p$ norms (Lessons~2--4).

\paragraph{Learning objectives.}
After completing this lesson, you will be able to:
\begin{itemize}[nosep]
  \item \textbf{Define} deep neural networks with $L$ layers and width
    $W$, the function class $\Fclass(L, W, \sigmoid)$, and
    \textbf{compare} parameter counts of shallow vs.\ deep architectures.
  \item \textbf{State and prove} Telgarsky's depth separation theorem,
    including the sawtooth construction and the lower bound on the
    width required by shallow networks.
  \item \textbf{State} the width-sufficiency results of Lu et al.\
    and Hanin, and \textbf{explain} the role of minimal width in
    universal approximation.
  \item \textbf{Derive} bounds on the number of linear regions of
    ReLU networks as a function of depth and width.
  \item \textbf{Apply} approximation rate results for smooth and
    compositionally structured functions to determine the advantage of
    deep over shallow architectures.
  \item \textbf{Justify} architecture choices in deep RL using
    depth-width tradeoff theory.
\end{itemize}

%% ============================================================
%% SECTION 2: REFERENCES
%% ============================================================
\section{References}
\label{sec:references}

\begin{itemize}
  \item \citet{telgarsky2016} --- The depth separation theorem: benefits
    of depth via the sawtooth construction.
  \item \citet{lu2017} --- Width-$(d+4)$ ReLU networks are universal
    approximators; the expressive power from the width perspective.
  \item \citet{hanin2019} --- Universal approximation with bounded width
    $d+1$ for $L^p$ and width $d+1$ for $C(K)$ with one extra unit.
  \item \citet{pinkus1999} --- Comprehensive survey of MLP approximation
    theory, including classical rate results.
  \item \citet{montufar2014} --- Counting linear regions of deep
    ReLU networks; the combinatorial source of depth advantage.
  \item \citet{yarotsky2017} --- Approximation rates for deep ReLU
    networks applied to Sobolev-class functions.
\end{itemize}

%% ============================================================
%% SECTION 3: DEEP NEURAL NETWORKS
%% ============================================================
\section{Deep Neural Networks}
\label{sec:deep-networks}

We formalize the class of functions computed by multi-layer
(deep) neural networks, extending the single-hidden-layer definition
from Lesson~63.

\begin{definition}[$L$-Layer Neural Network]
\label{def:deep-network}
Let $\sigmoid \colon \R \to \R$ be a fixed activation function.  An
\emph{$L$-layer feedforward neural network} with layer widths
$n_0, n_1, \ldots, n_L$ is a function
$f \colon \R^{n_0} \to \R^{n_L}$ defined recursively by
\begin{align}
  h_0 &= x, \notag \\
  h_\ell &= \sigmoid(W_\ell h_{\ell-1} + b_\ell),
    \qquad \ell = 1, \ldots, L-1, \label{eq:hidden-layers} \\
  f(x) &= W_L h_{L-1} + b_L, \notag
\end{align}
where $W_\ell \in \R^{n_\ell \times n_{\ell-1}}$ is the weight matrix
and $b_\ell \in \R^{n_\ell}$ is the bias vector for layer $\ell$.
The activation $\sigmoid$ is applied componentwise.  The integer $L$
is called the \emph{depth} and the last layer (without activation) is
the \emph{output layer}.
\end{definition}

\begin{definition}[Width and Depth]
\label{def:width-depth}
Given an $L$-layer network with layer widths $n_0, n_1, \ldots, n_L$:
\begin{enumerate}[nosep,label=(\roman*)]
  \item The \emph{depth} is $L$ (the number of affine transformations).
  \item The \emph{width} is $W = \max_{1 \le \ell \le L-1} n_\ell$
    (the maximum hidden-layer width).
  \item The \emph{total number of parameters} is
    $\sum_{\ell=1}^{L} (n_\ell \cdot n_{\ell-1} + n_\ell)$.
\end{enumerate}
\end{definition}

\begin{definition}[Deep Network Function Class]
\label{def:function-class}
For depth $L$, width $W$, input dimension $d$, output dimension $1$,
and activation $\sigmoid$, define
\[
  \Fclass(L, W, \sigmoid) = \bigl\{ f \colon \R^d \to \R :
  f \text{ is an $L$-layer network with width } \le W
  \text{ and activation } \sigmoid \bigr\}.
\]
When $\sigmoid = \text{ReLU}$, we write $\Fclass(L, W)$ for brevity.
\end{definition}

\begin{example}[Parameter Count: Shallow vs.\ Deep]
\label{ex:param-count}
Consider input dimension $d = 100$ and scalar output.
\begin{itemize}[nosep]
  \item \textbf{Shallow network} (depth $L = 2$, width $m$):
    The single hidden layer has $m(d + 1) = 101m$ parameters (weights
    plus biases), and the output layer has $m + 1$ parameters.  Total:
    $m(d+2) + 1 = 102m + 1$.  For $m = 1000$: $102{,}001$ parameters.
  \item \textbf{Deep network} (depth $L = 10$, width $W = 100$):
    The first hidden layer has $100(100 + 1) = 10{,}100$ parameters.
    Each of the 8 intermediate hidden layers has $100(100 + 1) = 10{,}100$
    parameters.  The output layer has $100 + 1 = 101$ parameters.
    Total: $10{,}100 + 8 \times 10{,}100 + 101 = 91{,}001$ parameters.
\end{itemize}
Both architectures have roughly $10^5$ parameters, but as we shall see,
the deep network can represent exponentially more complex functions.
\end{example}

\begin{remark}
\label{rem:single-vs-deep}
A single-hidden-layer network with $m$ neurons is the special case
$L = 2$, $n_0 = d$, $n_1 = m$, $n_2 = 1$.  The function class
$\Fclass_m^{\sigmoid}$ from Lesson~63 coincides with $\Fclass(2, m,
\sigmoid)$.  All results of Lessons~63--64 concern $L = 2$.
\end{remark}

The central question is: for a fixed parameter budget, does distributing
parameters across many layers (large $L$, moderate $W$) offer any
advantage over concentrating them in a single layer (small $L$, large
$W$)?  The next section provides a dramatic affirmative answer.

%% ============================================================
%% SECTION 4: DEPTH SEPARATION
%% ============================================================
\section{Depth Separation}
\label{sec:depth-separation}

The depth separation theorem of \citet{telgarsky2016} shows that
depth provides an exponential advantage in expressiveness.  The
key construction uses iterated piecewise-linear maps---the ``sawtooth''
function---that oscillate with a frequency that grows exponentially
with depth.

\subsection{The Sawtooth Function}
\label{sec:sawtooth}

\begin{definition}[Tent Map]
\label{def:tent-map}
The \emph{tent map} $t \colon \R \to \R$ is defined by
\[
  t(x) = \begin{cases}
    2x & \text{if } 0 \le x \le 1/2, \\
    2(1-x) & \text{if } 1/2 < x \le 1, \\
    0 & \text{otherwise},
  \end{cases}
\]
or equivalently $t(x) = 2 \min(x, 1-x)$ for $x \in [0,1]$.
\end{definition}

\begin{lemma}[ReLU Representation of the Tent Map]
\label{lem:tent-relu}
The tent map $t$ can be computed by a ReLU network of depth $2$ and
width $2$.  Specifically,
\[
  t(x) = 2\max(0, x) - 4\max(0, x - 1/2) + 2\max(0, x - 1).
\]
This uses $3$ ReLU units, but can be reduced to $2$ hidden units when
restricted to $[0,1]$ (since the third term vanishes).
\end{lemma}

\begin{proof}
For $x \in [0, 1/2]$: $\max(0, x) = x$, $\max(0, x - 1/2) = 0$,
$\max(0, x - 1) = 0$, so the expression equals $2x = t(x)$.
For $x \in (1/2, 1]$: $\max(0, x) = x$, $\max(0, x - 1/2) = x - 1/2$,
$\max(0, x - 1) = 0$, so the expression equals
$2x - 4(x - 1/2) = 2x - 4x + 2 = 2(1-x) = t(x)$.
\end{proof}

\begin{definition}[Sawtooth Function]
\label{def:sawtooth}
For $L \ge 1$, the \emph{$L$-fold sawtooth function}
$s_L \colon [0,1] \to [0,1]$ is the $L$-fold composition of the
tent map:
\[
  s_L = \underbrace{t \circ t \circ \cdots \circ t}_{L \text{ times}}.
\]
We set $s_0(x) = x$ (the identity).
\end{definition}

\begin{lemma}[Properties of the Sawtooth]
\label{lem:sawtooth-properties}
The function $s_L$ has the following properties:
\begin{enumerate}[nosep,label=(\roman*)]
  \item $s_L$ is continuous and piecewise linear on $[0,1]$.
  \item $s_L$ maps $[0,1]$ onto $[0,1]$.
  \item $s_L$ has exactly $2^L$ linear pieces (intervals of monotonicity),
    each of length $2^{-L}$.
  \item $s_L$ oscillates between $0$ and $1$ exactly $2^{L-1}$ times
    (it has $2^{L-1}$ local maxima and $2^{L-1}$ local minima on $(0,1)$,
    counting endpoints).
  \item $s_L$ has Lipschitz constant $2^L$.
\end{enumerate}
\end{lemma}

\begin{proof}
We prove these by induction on $L$.

\textbf{Base case ($L = 1$):} $s_1 = t$ has $2 = 2^1$ linear pieces
(on $[0,1/2]$ and $[1/2,1]$), maps $[0,1]$ onto $[0,1]$, has one
local maximum (at $x = 1/2$), and has Lipschitz constant $2 = 2^1$.

\textbf{Inductive step:}  Assume $s_{L-1}$ has $2^{L-1}$ linear pieces,
each of length $2^{-(L-1)}$.  On each such piece, $s_{L-1}$ is an
affine function mapping the piece onto $[0,1]$ (by property (ii)).
Composing with $t$ replaces each linear piece with two sub-pieces
(the tent map folds each $[0,1]$ traversal into two pieces), giving
$2 \cdot 2^{L-1} = 2^L$ linear pieces.  The length of each piece is
halved to $2^{-L}$.  The Lipschitz constant of $t \circ s_{L-1}$ is
$\text{Lip}(t) \cdot \text{Lip}(s_{L-1}) = 2 \cdot 2^{L-1} = 2^L$.
The oscillation count follows from the folding: each application of
$t$ doubles the number of maxima.
\end{proof}

\begin{example}[Sawtooth Functions for Small $L$]
\label{ex:sawtooth-small}
\begin{itemize}[nosep]
  \item $s_1 = t$: a single tent on $[0,1]$, $2$ linear pieces.
  \item $s_2 = t \circ t$: a ``W''-shaped function with $4$ linear pieces,
    oscillating between $0$ and $1$ twice.
  \item $s_3 = t \circ t \circ t$: $8$ linear pieces, $4$ oscillations.
  \item $s_{10}$: $1024$ linear pieces, $512$ oscillations---a rapidly
    oscillating function that looks like a dense sawtooth wave.
\end{itemize}
\end{example}

\begin{proposition}[Sawtooth as a Deep Network]
\label{prop:sawtooth-deep}
For any $L \ge 1$, the sawtooth function $s_L$ belongs to
$\Fclass(L+1, 2)$: it can be computed by a ReLU network of depth
$L + 1$ and width $2$.
\end{proposition}

\begin{proof}
By Lemma~\ref{lem:tent-relu}, the tent map $t$ is computed by a
depth-$2$, width-$2$ ReLU network.  Since $s_L = t \circ t \circ
\cdots \circ t$ ($L$ times), we compose $L$ copies of this network.
The composition of $L$ depth-$2$ networks yields a network of depth
$L + 1$ (collapsing adjacent linear layers) with width $2$ at every
hidden layer.  The total number of parameters is $O(L)$.
\end{proof}

\subsection{The Depth Separation Theorem}

\begin{lemma}[Linear Pieces of Shallow ReLU Networks]
\label{lem:shallow-pieces}
Any function $f \in \Fclass(2, m)$ (a single-hidden-layer ReLU network
with $m$ hidden units and one-dimensional input) is piecewise linear
with at most $m + 1$ linear pieces.
\end{lemma}

\begin{proof}
Each hidden unit computes $\max(0, w_j x + b_j)$, which is piecewise
linear with exactly $2$ pieces (and one breakpoint at $x = -b_j/w_j$
when $w_j \neq 0$).  The output $f(x) = a_0 + \sum_{j=1}^m a_j
\max(0, w_j x + b_j)$ is a sum of $m$ piecewise-linear functions
on $\R$, each with one breakpoint.  The set of breakpoints has at
most $m$ elements, so $f$ is piecewise linear with at most $m + 1$
pieces.
\end{proof}

\begin{lemma}[Oscillation Lower Bound]
\label{lem:oscillation-bound}
Let $f, g \colon [0,1] \to \R$ be continuous functions.  If $f$ has $k$
points where $f$ crosses a threshold $\tau$ (i.e., $f$ changes from
$f < \tau$ to $f > \tau$ or vice versa), and $\|f - g\|_\infty \le
\varepsilon < (\max f - \min f)/4$, then $g$ must cross the threshold
$\tau$ at least $k - 2$ times (for $\varepsilon$ sufficiently small
relative to the oscillation amplitude).

More precisely, if $f$ achieves values $0$ and $1$ on alternating
subintervals and $\|f - g\|_{L^1} \le \delta$ with $\delta < 1/4$,
then $g$ must have at least as many sign changes from below $1/2$ to
above $1/2$ as $f$ does, minus a correction depending on $\delta$.
\end{lemma}

\begin{proof}
Since $f$ oscillates between $0$ and $1$ with $k/2$ full oscillations,
there are $k/2$ disjoint subintervals $I_j$ of $[0,1]$ on which
$f \ge 3/4$ and $k/2$ disjoint subintervals $J_j$ on which $f \le 1/4$,
with the $I_j$ and $J_j$ alternating.  If $\|f - g\|_{L^1} < \delta$
with $\delta$ small enough, then on most of each $I_j$, $g$ must also
be close to $1$ (by Markov's inequality applied interval by interval),
and on most of each $J_j$, $g$ must be close to $0$.  For $g$ to
transition between these values, it must cross $1/2$ at least once
between each pair of consecutive intervals, giving at least $k - 1$
crossings.  A piecewise linear function with $p$ pieces can cross any
threshold at most $p - 1$ times.  Therefore $p - 1 \ge k - 1$, i.e.,
$p \ge k$.
\end{proof}

\begin{theorem}[Telgarsky, 2016 --- Depth Separation]
\label{thm:depth-separation}
For any $L \ge 2$, there exists a function $f_L \colon [0,1] \to [0,1]$
such that:
\begin{enumerate}[nosep,label=(\roman*)]
  \item $f_L \in \Fclass(L+1, 2)$: $f_L$ is computed by a ReLU network
    of depth $L+1$ and width $2$ (with $O(L)$ total parameters).
  \item For any $g \in \Fclass(2, m)$ (a depth-$2$ network with $m$
    hidden units),
    \[
      \|f_L - g\|_{L^1[0,1]} \ge \frac{1}{8}
      \qquad \text{whenever } m \le 2^{L/2} - 1.
    \]
\end{enumerate}
That is, approximating $f_L$ within constant $L^1$ error requires
a shallow network of width at least $2^{\Omega(L)}$.
\end{theorem}

\begin{proof}
\textbf{Construction.}  Set $f_L = s_L$, the $L$-fold sawtooth function
(Definition~\ref{def:sawtooth}).  By
Proposition~\ref{prop:sawtooth-deep}, $f_L \in \Fclass(L+1, 2)$,
establishing (i).

\textbf{Lower bound.}  By Lemma~\ref{lem:sawtooth-properties},
$f_L = s_L$ has $2^L$ linear pieces and oscillates between $0$ and $1$
with $2^{L-1}$ full oscillations.  We claim that any good approximation
$g$ must also have many linear pieces.

Consider the $2^{L-1}$ subintervals on which $s_L$ rises from $0$ to
$1$ (the ``up-ramps'').  Each up-ramp has length $2^{-L}$ and maps
$[a_j, a_j + 2^{-L}]$ linearly onto $[0, 1]$.  Similarly, the
$2^{L-1}$ ``down-ramps'' each map an interval of length $2^{-L}$
linearly from $1$ to $0$.

Now suppose $g \in \Fclass(2, m)$ satisfies $\|f_L - g\|_{L^1} < 1/8$.
By Lemma~\ref{lem:shallow-pieces}, $g$ is piecewise linear with at most
$m + 1$ pieces.  We show this forces $m \ge 2^{L/2}$.

On each of the $2^{L-1}$ subintervals where $s_L \ge 3/4$ (the top
halves of the up-ramps, totaling length $2^{L-1} \cdot 2^{-(L+1)} =
1/4$), if $g$ is far from $s_L$ on a fraction larger than $1/2$ of this
region, the $L^1$ error exceeds $(1/4)(1/4)(1/2) = 1/32$ from these
intervals alone.  Similarly for the intervals where $s_L \le 1/4$.
For $\|f_L - g\|_{L^1} < 1/8$, the function $g$ must be above $1/2$ on
most of the intervals where $s_L$ is near $1$, and below $1/2$ on most
of the intervals where $s_L$ is near $0$.  Since these intervals
alternate, $g$ must cross the level $1/2$ at least $\Omega(2^{L-1})$
times.

Each linear piece of $g$ can cross the level $1/2$ at most once.
Therefore $g$ must have at least $\Omega(2^{L-1})$ linear pieces,
which means $m + 1 \ge 2^{L-1}$, i.e., $m \ge 2^{L-1} - 1$.  The
sharper bound $m \ge 2^{L/2} - 1$ stated in the theorem follows from
a more refined analysis of the $L^1$ error on individual pieces: a
piecewise linear function with $p$ pieces on $[0,1]$ can approximate
a function with $k$ oscillations of unit amplitude only if $p \ge
\sqrt{k}$ when measured in $L^1$ (since the $L^1$ error per
oscillation period that is not tracked by a linear piece is
$\Omega(1/p)$, giving total error $\Omega(k/p)$; setting this less
than $1/8$ requires $p \ge 8k$, but the $\sqrt{k}$ bound comes from
a counting argument: each piece of $g$ can ``cancel'' at most $O(p)$
of the oscillations, so $p^2 \ge k$ is needed).  The exact constant
and the exponent $L/2$ in the lower bound are established by
\citet{telgarsky2016}.
\end{proof}

\begin{keyresult}[title={Telgarsky's Depth Separation}]
There exist functions computable by ReLU networks of depth $O(L)$ and
width $O(1)$ that require width $2^{\Omega(L)}$ to be approximated
(in $L^1$ norm) by any network of depth $O(1)$.  Depth provides an
\textbf{exponential} advantage in expressiveness over width.
\end{keyresult}

\begin{example}[Depth-2 vs.\ Depth-20]
\label{ex:depth-comparison}
The sawtooth $s_{20}$ has $2^{20} \approx 10^6$ linear pieces and is
computed by a depth-$21$ network with width $2$ ($O(20)$ parameters).
Approximating it within $L^1$ error $1/8$ with a single-hidden-layer
network requires width at least $2^{10} = 1024$.  The deep network
uses roughly $60$ parameters; the shallow network needs at least
$2 \times 1024 + 1025 = 3073$ parameters.  The ratio is roughly
$50{\times}$.
\end{example}

\begin{example}[Depth Separation in Higher Dimensions]
\label{ex:higher-dim}
Telgarsky's construction is one-dimensional, but it extends to higher
dimensions by applying the sawtooth along a coordinate projection.
\citet{eldan2016} proved a depth separation result in $\R^d$ with
a different construction: there exists a radial function on $\R^d$
expressible by a depth-3 network of polynomial width, for which any
depth-2 network requires width $\Omega(2^d)$ to approximate.  This
shows that depth separation is not merely a one-dimensional phenomenon.
\end{example}

\begin{rlconnection}[title={Depth Separation and RL Architectures}]
In practice, deep RL agents use networks with $3$--$10$ hidden layers,
not single-hidden-layer networks.  The depth separation theorem
provides theoretical justification: if the value function $V^*$ or
the policy $\pi^*$ has structure resembling iterated folding (which is
plausible for environments with hierarchical dynamics), then deep
networks can represent it with polynomially fewer parameters than
shallow ones.  This is consistent with the empirical observation that
deeper networks improve performance in complex environments (e.g.,
Atari, MuJoCo) even when the total parameter count is controlled.
\end{rlconnection}

The depth separation theorem shows that depth provides an exponential
advantage.  We now ask the complementary question: is \emph{width}
also important?

%% ============================================================
%% SECTION 5: WIDTH SEPARATION AND WIDTH SUFFICIENCY
%% ============================================================
\section{Width Separation and Width Sufficiency}
\label{sec:width}

While Telgarsky's theorem shows that depth can be exponentially
more efficient than width, the following results of Lu et al.\ and
Hanin show that sufficient width also guarantees universal
approximation, even at fixed depth.

\subsection{Width Sufficiency}

\begin{theorem}[Lu et al., 2017 --- Width Sufficiency for ReLU Networks]
\label{thm:lu}
Let $d \ge 1$.  The class of ReLU networks with input dimension $d$
and width $w \ge d + 4$ is a universal approximator of $L^1(\R^d,
\mu)$ for any compactly supported probability measure $\mu$ on $\R^d$:
for any $f \in L^1(\mu)$ and $\varepsilon > 0$, there exists a ReLU
network of width $w$ (and some finite depth $L$) such that
\[
  \int_{\R^d} |f(x) - g(x)| \, d\mu(x) < \varepsilon.
\]
\end{theorem}

\begin{proof}[Proof sketch]
The proof of \citet{lu2017} proceeds by showing that width-$(d+4)$
networks can implement certain ``building block'' operations:
\begin{enumerate}[nosep]
  \item \textbf{Step functions:} A width-$(d+4)$ ReLU network of
    sufficient depth can approximate indicator functions of
    hyperrectangles to arbitrary accuracy.
  \item \textbf{Lebesgue integral:} Since simple functions (finite
    linear combinations of indicator functions of measurable sets)
    are dense in $L^1(\mu)$, it suffices to approximate these.
    Hyperrectangle indicators generate the Borel $\sigma$-algebra,
    so step (1) suffices.
  \item \textbf{Propagation:} The key technical argument shows that
    a ReLU network of width $d + 4$ can (a) compute the identity
    map on $\R^d$ (using $d$ units), (b) use $2$ additional units to
    implement a tent-map fold, and (c) use $2$ more units for
    bookkeeping.  By composing many layers, the network can construct
    arbitrarily fine-grained step functions.
\end{enumerate}
The full proof requires careful analysis of how ReLU units propagate
information through layers; we refer to \citet{lu2017} for details.
\end{proof}

\begin{theorem}[Hanin, 2019 --- Minimal Width]
\label{thm:hanin}
Let $K \subseteq \R^d$ be compact and let $\mu$ be a probability
measure on $K$ with full support.
\begin{enumerate}[nosep,label=(\roman*)]
  \item For $1 \le p < \infty$: the class of ReLU networks of width
    $w \ge d + 1$ is dense in $L^p(K, \mu)$.
  \item For uniform approximation: the class of ReLU networks of width
    $w \ge d + 1$ is dense in $C(K)$.
  \item The bound $w \ge d + 1$ is tight: width-$d$ ReLU networks
    are \emph{not} dense in $L^p(K, \mu)$ for any $p \ge 1$.
\end{enumerate}
\end{theorem}

\begin{proof}[Proof sketch]
(iii) Width-$d$ ReLU networks compute functions $f \colon \R^d \to \R$
whose level sets $\{x : f(x) = c\}$ have a specific geometric
structure imposed by the bottleneck layer of width $d$.  Specifically,
at any hidden layer of width exactly $d$, the network factors through
a piecewise-linear map $\R^d \to \R^d$, which has determinant zero on
a measure-zero set and is injective almost everywhere.  This prevents
the network from ``folding'' the input space enough to approximate
functions with certain topological properties.  The detailed argument
uses the piecewise-linear structure and degree theory; see
\citet{hanin2019}.

(i)--(ii) The positive results follow by showing that width-$(d+1)$
networks can implement the identity map $\R^d \to \R^d$ while
retaining one extra unit for folding.  The folding operation, iterated
over many layers, generates arbitrarily complex piecewise-linear
functions, analogous to the sawtooth construction in
Section~\ref{sec:depth-separation} but in higher dimensions.
\end{proof}

\begin{example}[Minimal Width in One Dimension]
\label{ex:minimal-width-1d}
For $d = 1$, the minimal width for universal approximation is
$d + 1 = 2$.  Width-$1$ networks compute functions of the form
$f(x) = a_L \max(0, a_{L-1} \max(0, \cdots) + b_{L-1}) + b_L$, which
are piecewise linear with at most $2$ pieces regardless of depth (each
ReLU applied to a monotone function can add at most one breakpoint, but
with width $1$, only one breakpoint is produced at each layer, and
subsequent layers may remove breakpoints).  Therefore width-$1$ networks
are not universal, confirming $w \ge 2 = d + 1$.
\end{example}

\subsection{Linear Regions of ReLU Networks}

The depth-versus-width advantage for ReLU networks can be understood
through the lens of \emph{linear regions}: the number of connected
regions of the input space on which the network computes a single
affine function.

\begin{definition}[Linear Region]
\label{def:linear-region}
A \emph{linear region} of a piecewise-affine function $f \colon
\R^d \to \R$ is a maximal connected subset of $\R^d$ on which $f$ is
affine.
\end{definition}

\begin{theorem}[Linear Region Bounds]
\label{thm:linear-regions}
Let $f \in \Fclass(L, W)$ be a ReLU network with depth $L$ and width
$W$ on $\R^d$.
\begin{enumerate}[nosep,label=(\roman*)]
  \item \textbf{Upper bound:} $f$ has at most
    $\binom{W}{d}^{L-1} \cdot 2^W$ linear regions.  When $W \ge d$,
    this simplifies to at most $\bigl( \frac{eW}{d} \bigr)^{d(L-1)}
    \cdot 2^W$.  In particular, the number of linear regions is
    $O(W^{d(L-1)})$ --- exponential in $L$ and polynomial in $W$
    (for fixed $d$).
  \item \textbf{Lower bound:} There exist ReLU networks of depth $L$
    and width $W$ with at least $\lfloor W/d \rfloor^{d(L-1)}$
    linear regions.
\end{enumerate}
\end{theorem}

\begin{proof}[Proof sketch]
(i) Each hidden layer has $W$ ReLU units.  Each unit partitions
$\R^d$ into two half-spaces (the unit's active and inactive regions).
The arrangement of $W$ hyperplanes in $\R^d$ creates at most
$\sum_{i=0}^{d} \binom{W}{i} \le \binom{W}{d} + \binom{W}{d-1} +
\cdots$ regions (by Zaslavsky's theorem for hyperplane arrangements).
When $W \ge d$, this is at most $\bigl( \frac{eW}{d} \bigr)^d$.  Each
layer can multiply the number of linear regions by this factor, so
$L - 1$ hidden layers give at most $\bigl( \frac{eW}{d} \bigr)^{d(L-1)}$
regions.  The last hidden layer contributes at most $2^W$ (with a
looser bound that is tighter when $W$ is small).

(ii) The lower bound is achieved by networks where each layer folds
the input space maximally.  The construction in \citet{montufar2014}
uses each group of $d$ units to implement a fold (doubling the number
of regions), and $W/d$ such groups per layer give $\lfloor W/d
\rfloor^d$ new regions per layer.
\end{proof}

\begin{example}[Linear Regions: Depth vs.\ Width]
\label{ex:linear-regions}
For $d = 10$, a network with $W = 100$ and $L = 2$ (shallow) has at
most $\binom{100}{10} \approx 1.7 \times 10^{13}$ linear regions.
A network with $W = 20$ and $L = 10$ (deep) can have up to
$\binom{20}{10}^9 \approx (184{,}756)^9 \approx 10^{47}$ linear
regions.  The deep network has far fewer parameters ($\approx 4000$)
than the shallow one ($\approx 11{,}000$) but can create exponentially
more linear regions.
\end{example}

\begin{keyresult}[title={Depth vs.\ Width in Linear Region Count}]
For ReLU networks with input dimension $d$, the maximum number of linear
regions is:
\begin{itemize}[nosep]
  \item \textbf{Exponential in depth}: $\Theta(W^d)^{L-1}$ regions for
    depth-$L$, width-$W$ networks.
  \item \textbf{Polynomial in width}: $O(W^d)$ regions for
    single-hidden-layer (depth-$2$) networks with $W$ units.
\end{itemize}
Depth provides exponentially more linear regions per parameter than width.
\end{keyresult}

\begin{warning}[title={Linear Regions Do Not Equal Approximation Quality}]
A large number of linear regions is \emph{necessary} for approximating
highly oscillatory functions but is not \emph{sufficient} for good
approximation.  The regions must be arranged to match the target
function.  The optimization landscape (how gradient descent navigates
the parameter space to find the right arrangement) is a separate and
harder question, addressed by the neural tangent kernel theory in
Lesson~66.
\end{warning}

We now quantify the approximation rates that deep networks achieve
for specific function classes.

%% ============================================================
%% SECTION 6: APPROXIMATION RATES FOR SMOOTH FUNCTIONS
%% ============================================================
\section{Approximation Rates for Smooth Functions}
\label{sec:approx-rates}

The depth separation results of Section~\ref{sec:depth-separation}
use pathological functions (sawtooths) to prove worst-case separation.
In this section, we study approximation rates for natural function
classes---Sobolev spaces and compositionally structured functions---and
show that deep networks achieve provably better rates than shallow ones.

\subsection{Classical Approximation Rates}
\label{sec:classical-rates}

We recall the classical approximation rate for functions in Sobolev
spaces, previously stated in Lesson~64 (Proposition~5.1).

\begin{definition}[Sobolev Space]
\label{def:sobolev}
For $s \ge 1$ and $1 \le p \le \infty$, the \emph{Sobolev space}
$\mathcal{W}^{s,p}([0,1]^d)$ consists of all functions
$f \colon [0,1]^d \to \R$ whose partial derivatives up to order $s$
exist in the weak sense and belong to $L^p$, equipped with the norm
\[
  \|f\|_{\mathcal{W}^{s,p}} = \sum_{|\alpha| \le s}
  \|D^\alpha f\|_{L^p}.
\]
The Sobolev ball of radius $B$ is $\mathcal{W}^{s,p}_B = \{f :
\|f\|_{\mathcal{W}^{s,p}} \le B\}$.
\end{definition}

\begin{proposition}[Classical Minimax Rates]
\label{prop:classical-minimax}
For any nonlinear approximation scheme using $m$ parameters (including
neural networks, splines, wavelets), the minimax $L^2$ approximation
rate over $\mathcal{W}^{s,2}_B([0,1]^d)$ satisfies
\[
  \inf_{\hat{f}_m} \sup_{f \in \mathcal{W}^{s,2}_B}
  \|f - \hat{f}_m\|_{L^2}
  = \Theta(B \cdot m^{-s/d}).
\]
The rate $m^{-s/d}$ suffers the \emph{curse of dimensionality}: for
fixed smoothness $s$ and large $d$, the rate becomes arbitrarily slow.
\end{proposition}

\begin{proof}
The lower bound follows from Kolmogorov $n$-widths or metric entropy
arguments; see \citet{pinkus1999}.  The upper bound is achieved by
tensor-product B-splines or wavelet expansions with $m$ terms.
\end{proof}

\subsection{Deep Network Rates for Sobolev Functions}

Deep ReLU networks can match the classical minimax rate with an
explicit construction.

\begin{theorem}[Yarotsky, 2017]
\label{thm:yarotsky}
For $f \in \mathcal{W}^{s,\infty}([0,1]^d)$ with
$\|f\|_{\mathcal{W}^{s,\infty}} \le B$, and for any $\varepsilon > 0$,
there exists a ReLU network $g$ with depth $O(\log(1/\varepsilon))$
and total parameters $O(\varepsilon^{-d/s} \log(1/\varepsilon))$ such
that $\|f - g\|_{L^\infty} \le \varepsilon$.

The approximation rate as a function of the number of parameters $m$ is
\[
  \|f - g\|_{L^\infty} = O\bigl(m^{-s/d} \cdot (\log m)^{s/d}\bigr),
\]
which matches the classical rate $m^{-s/d}$ up to logarithmic factors.
\end{theorem}

\begin{proof}[Proof sketch]
The proof of \citet{yarotsky2017} proceeds in three steps:

\textbf{Step 1: Approximate monomials.}  Using the identity $xy =
\frac{1}{2}[(x+y)^2 - x^2 - y^2]$ and the fact that $x^2$ can be
approximated by a deep ReLU network of depth $O(\log(1/\varepsilon))$
with error $\varepsilon$ (via a binary expansion construction), one
shows that products $x_i x_j$ can be approximated.  By iteration,
arbitrary monomials $x_1^{k_1} \cdots x_d^{k_d}$ with $\sum k_j \le s$
can be approximated.

\textbf{Step 2: Local Taylor approximation.}  Partition $[0,1]^d$ into
$N^d$ cubes of side length $1/N$.  On each cube, approximate $f$ by its
Taylor polynomial of degree $s - 1$.  By Taylor's theorem, the local
error is $O(N^{-s})$.

\textbf{Step 3: Assemble.}  Using the ReLU network to implement the
partition of unity and the local polynomial approximations, the total
number of parameters is $O(N^d \cdot \text{poly}(\log N))$.  Setting
$N = \varepsilon^{-1/s}$ gives total parameters $m =
O(\varepsilon^{-d/s} \log(1/\varepsilon))$, which achieves
$\|f - g\|_{L^\infty} \le C \varepsilon$.
\end{proof}

\begin{remark}
\label{rem:depth-benefit-sobolev}
Theorem~\ref{thm:yarotsky} uses depth $O(\log(1/\varepsilon))$ to
approximate monomials via the squaring trick.  A shallow network
would need width $O(\varepsilon^{-d/s})$ without the logarithmic
depth factor.  The depth is essential for efficiently computing
multiplications, which are the building blocks of polynomial
approximation.
\end{remark}

\subsection{Compositional Functions and the Blessing of Depth}

The most dramatic advantage of depth appears for functions with
\emph{compositional structure}: functions that decompose as
$f = f_L \circ f_{L-1} \circ \cdots \circ f_1$ where each $f_\ell$
operates on a low-dimensional input.

\begin{definition}[Compositional Function]
\label{def:compositional}
A function $f \colon \R^d \to \R$ has \emph{$(L, d_0)$-compositional
structure} if there exist functions $g_1, g_2, \ldots, g_L$ with
$g_\ell \colon \R^{d_0} \to \R^{d_0}$ (for $\ell < L$) and
$g_L \colon \R^{d_0} \to \R$, each with $s$ bounded derivatives,
such that
\[
  f = g_L \circ g_{L-1} \circ \cdots \circ g_1 \circ P,
\]
where $P \colon \R^d \to \R^{d_0}$ is a fixed projection (or partition
of coordinates into groups of size $d_0$).
\end{definition}

\begin{theorem}[Depth Advantage for Compositional Functions]
\label{thm:compositional}
Let $f$ be an $(L, d_0)$-compositional function where each component
$g_\ell \in \mathcal{W}^{s,\infty}(\R^{d_0})$ with bounded norm.
Then:
\begin{enumerate}[nosep,label=(\roman*)]
  \item A \textbf{deep network} matching the compositional structure
    (depth $O(L \log(1/\varepsilon))$, width $O(d_0)$) achieves
    \[
      \|f - g\|_{L^\infty} \le \varepsilon
      \quad \text{with } O\bigl(L \cdot \varepsilon^{-d_0/s}
      \log(1/\varepsilon)\bigr) \text{ parameters}.
    \]
    The dependence on the intrinsic dimension $d_0$ replaces the
    ambient dimension $d$.
  \item A \textbf{shallow network} (depth $2$), unaware of the
    compositional structure, requires
    $\Omega(\varepsilon^{-d/s})$ parameters to achieve the same
    accuracy, suffering the curse of dimensionality in $d$.
\end{enumerate}
\end{theorem}

\begin{proof}[Proof sketch]
(i) Apply Theorem~\ref{thm:yarotsky} to each component $g_\ell$
separately, using a sub-network of depth $O(\log(1/\varepsilon))$ and
$O(\varepsilon^{-d_0/s} \log(1/\varepsilon))$ parameters.
Compose the $L$ sub-networks.  The errors accumulate multiplicatively
(due to the Lipschitz constants of the components), but since each
$g_\ell$ has bounded Lipschitz constant, the total error is
$O(L \varepsilon)$ (rescaling $\varepsilon$ absorbs the factor $L$).
The total parameter count is $L$ times the per-component cost.

(ii) A shallow network treats $f$ as a function on $\R^d$ without
exploiting the compositional structure, so the classical rate
$\Theta(\varepsilon^{-d/s})$ applies.
\end{proof}

\begin{example}[Vision-Like Compositional Structure]
\label{ex:vision}
Consider classifying images of size $32 \times 32$ pixels ($d = 1024$).
Suppose the classification function has a $3$-level hierarchy:
(1)~local edge detectors operate on $5 \times 5$ patches ($d_0 = 25$),
(2)~mid-level feature combiners aggregate edge outputs, and
(3)~a global classifier combines mid-level features.  Each level has
smoothness $s = 2$.

\begin{itemize}[nosep]
  \item \textbf{Shallow network:} Rate $\varepsilon^{-d/s} =
    \varepsilon^{-512}$.  For $\varepsilon = 0.01$: $10^{1024}$
    parameters.
  \item \textbf{Deep network:} Rate $L \cdot \varepsilon^{-d_0/s} =
    3 \cdot \varepsilon^{-12.5}$.  For $\varepsilon = 0.01$:
    $3 \times 10^{25}$ parameters.
\end{itemize}
The deep network reduces the exponent from $512$ to $12.5$---a
dramatic improvement.
\end{example}

\begin{rlconnection}[title={Hierarchical Structure in RL}]
Many RL environments exhibit compositional structure:
\begin{itemize}[nosep]
  \item \textbf{Atari games}: The screen is processed hierarchically
    (pixels $\to$ edges $\to$ objects $\to$ game state).
  \item \textbf{Robotic control}: Joint angles $\to$ limb positions
    $\to$ body configuration $\to$ task-relevant features.
  \item \textbf{Multi-agent systems}: Individual agent states $\to$
    local interactions $\to$ group behavior $\to$ global outcome.
\end{itemize}
In each case, the value function inherits the compositional structure
of the environment, and deep networks can exploit this structure via
Theorem~\ref{thm:compositional}, while shallow networks cannot.
\end{rlconnection}

\begin{remark}
\label{rem:pinkus-context}
The results of this section contextualize \citet{pinkus1999}, who
surveyed the approximation theory of the MLP model before the deep
learning revolution.  Pinkus focused on single-hidden-layer networks
and established the minimax rate $\Theta(m^{-s/d})$ for Sobolev
functions.  The key insight of the subsequent work by Telgarsky,
Yarotsky, and others is that \emph{depth} allows networks to
exploit structure (compositionality, multiplications, recursive
folding) in ways that are impossible for shallow networks.
\end{remark}

%% ============================================================
%% SECTION 7: EXERCISES
%% ============================================================
\section{Exercises}
\label{sec:exercises}

\begin{exercise}[Sawtooth Function Computation]
\textnormal{(\({}^*\))} \\
Compute $s_3(x)$ explicitly for $x \in [0, 1]$.  List all breakpoints,
verify that there are $2^3 = 8$ linear pieces, and sketch the graph.
\end{exercise}

\begin{exercise}[Linear Regions of a Specific Network]
\textnormal{(\({}^*\))} \\
Consider a ReLU network on $\R^2$ with depth $L = 3$ and width $W = 4$.
\begin{enumerate}[label=(\alph*)]
  \item Compute the upper bound on the number of linear regions from
    Theorem~\ref{thm:linear-regions}.
  \item Construct a specific network achieving at least $16$ linear
    regions.
\end{enumerate}
\end{exercise}

\begin{exercise}[Depth Separation for $L^2$ Approximation]
\textnormal{(\({}^{**}\))} \\
Show that $\|s_L\|_{L^2[0,1]}^2 = 1/3$ for all $L \ge 1$.
Using this, strengthen the depth separation bound of
Theorem~\ref{thm:depth-separation} to the $L^2$ norm: show that
any $g \in \Fclass(2, m)$ with $m < 2^{L/2}$ satisfies
$\|s_L - g\|_{L^2} \ge c$ for a constant $c > 0$.

\textit{Hint:} Use the orthogonality of $s_L$ on its linear pieces
and the fact that a piecewise linear function with $m+1$ pieces
has $\|g\|_{L^2}^2$ that can be computed explicitly.
\end{exercise}

\begin{exercise}[Approximating Products with Deep ReLU Networks]
\textnormal{(\({}^{**}\))} \\
The key building block in Yarotsky's construction
(Theorem~\ref{thm:yarotsky}) is approximating $x^2$ on $[0,1]$.
\begin{enumerate}[label=(\alph*)]
  \item Show that the sawtooth function $s_L$ satisfies
    $\|x - x^2 - \frac{1}{4} s_L(x)\|_{L^\infty[0,1]} = O(4^{-L})$
    by considering the map $g(x) = 4x(1-x)$ and its iterates.
    (Actually, use the identity $x^2 = x - \sum_{k=1}^L
    2^{-2k} s_k(x) + O(4^{-L})$.)
  \item Conclude that $x^2$ can be approximated to accuracy
    $\varepsilon$ by a ReLU network of depth $O(\log(1/\varepsilon))$
    and width $O(1)$.
  \item Using $xy = \frac{1}{2}[(x+y)^2 - x^2 - y^2]$, show that
    multiplication can be approximated by a ReLU network of depth
    $O(\log(1/\varepsilon))$ and width $O(1)$.
\end{enumerate}
\end{exercise}

\begin{exercise}[Width-$d$ Networks Are Not Universal]
\textnormal{(\({}^{**}\))} \\
Let $d = 2$ and consider a ReLU network of width $2$ and arbitrary
depth $L$.
\begin{enumerate}[label=(\alph*)]
  \item Show that each hidden layer computes a piecewise-affine map
    $\R^2 \to \R^2$.  Argue that the composition of $L$ such maps is
    also piecewise affine.
  \item Show that the resulting function $f \colon \R^2 \to \R$ has
    level sets $\{x : f(x) = c\}$ that are piecewise linear.
  \item Conclude that $f$ cannot approximate the function
    $g(x_1, x_2) = x_1^2 + x_2^2$ uniformly on $[-1,1]^2$, because
    $g$ has curved level sets.  This confirms that width $d = 2$ is
    insufficient for universal approximation in $C([-1,1]^2)$.
\end{enumerate}
\end{exercise}

\begin{exercise}[Deep RL Architecture Design]
\textnormal{(\({}^{***}\))} \\
Consider an RL environment where the state is a vector
$s = (s_1, \ldots, s_{100}) \in \R^{100}$ and the optimal value
function has the compositional structure
$V^*(s) = h(g_1(s_1, \ldots, s_{10}), g_2(s_{11}, \ldots, s_{20}),
\ldots, g_{10}(s_{91}, \ldots, s_{100}))$, where each $g_k \colon
\R^{10} \to \R$ and $h \colon \R^{10} \to \R$ are twice continuously
differentiable ($s = 2$).
\begin{enumerate}[label=(\alph*)]
  \item Apply Theorem~\ref{thm:compositional} to determine the
    parameter count needed by a deep network to achieve
    $\varepsilon$-approximation.
  \item Compute the parameter count for a shallow network.
  \item Design a specific network architecture (layer widths and depth)
    that exploits the compositional structure.  What is the total
    parameter count?
\end{enumerate}
\end{exercise}

%% ============================================================
%% SECTION 8: SUMMARY
%% ============================================================
\section{Summary}
\label{sec:summary}

\begin{itemize}
  \item An $L$-layer neural network applies $L$ affine transformations
    interleaved with activations; the \textbf{depth} $L$ and
    \textbf{width} $W = \max_\ell n_\ell$ are the key architectural
    parameters.

  \item The \textbf{sawtooth function} $s_L = t^{\circ L}$ (iterated
    tent maps) has $2^L$ linear pieces and can be computed by a
    depth-$(L+1)$, width-$2$ ReLU network with $O(L)$ parameters.

  \item \textbf{Telgarsky's depth separation theorem}: the sawtooth
    $s_L$ requires width $2^{\Omega(L)}$ to be approximated by a
    depth-$2$ network.  Depth provides an exponential advantage in
    expressiveness.

  \item \textbf{Width sufficiency}: ReLU networks of width $d + 4$
    (Lu et al.) or $d + 1$ (Hanin) are universal approximators
    regardless of the target function, provided depth is allowed to
    grow.  Width $d$ is provably insufficient.

  \item The number of \textbf{linear regions} of a ReLU network grows
    exponentially in depth but polynomially in width, providing a
    combinatorial explanation for depth's advantage.

  \item For \textbf{Sobolev-class functions}, deep ReLU networks
    achieve the classical minimax rate $O(m^{-s/d})$ using depth
    $O(\log(1/\varepsilon))$, which is needed to efficiently compute
    polynomial building blocks via the squaring trick.

  \item For \textbf{compositionally structured functions}
    $f = g_L \circ \cdots \circ g_1$, deep networks achieve
    approximation rates depending on the intrinsic dimension $d_0$
    of each component, not the ambient dimension $d$---avoiding the
    curse of dimensionality entirely.

  \item The depth-width tradeoff justifies the use of deep
    architectures in RL/ML: environments with hierarchical structure
    produce value functions that deep networks can approximate
    exponentially more efficiently than shallow networks.
\end{itemize}

%% ============================================================
%% SECTION 9: FURTHER READING
%% ============================================================
\section{Further Reading}
\label{sec:further-reading}

\begin{itemize}
  \item \textbf{Lesson 66} develops the neural tangent kernel
    framework, which addresses the optimization question: can
    gradient descent find good weights in the deep networks whose
    existence this lesson establishes?  It also provides Rademacher
    complexity bounds for deep networks that complement the
    approximation results here.

  \item \textbf{Lesson 67} (RNNs and BPTT) studies recurrent
    architectures, where ``depth'' takes the form of unrolling through
    time.  The depth separation results of this lesson apply (with
    modifications) to unrolled RNNs.

  \item \citet{eldan2016} proves a depth separation result in
    $\R^d$ showing that three-layer networks can represent functions
    that two-layer networks require exponential width to approximate.
    The construction uses a different technique (random feature
    functions) than Telgarsky's sawtooth.

  \item \citet{safran2017} extends depth-width tradeoffs to
    ``natural'' functions (not just adversarial constructions),
    showing that radial functions benefit from depth.

  \item \citet{poggio2017} provides a detailed analysis of when
    and why depth avoids the curse of dimensionality, focusing on
    compositional functions arising in vision and language.

  \item \citet{schmidt-hieber2020} establishes minimax-optimal
    rates for deep ReLU networks in nonparametric regression,
    showing that deep networks adapt to low-dimensional structure
    in the regression function.

  \item \textbf{Historical note.}  The question of depth versus width
    in neural networks dates back to the 1990s (e.g., the work of
    H\aa stad on Boolean circuit complexity inspired analogous
    questions for neural networks).  The modern depth separation
    results of Telgarsky (2016) and Eldan--Shamir (2016) provided the
    first unconditional separations for standard neural network
    architectures, settling a longstanding open question.
\end{itemize}

\bibliography{nn-refs}
\end{document}
