\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 68: LSTM and Gating Mechanisms}}
\fancyhead[R]{\thepage}
\title{Lesson 68: Long Short-Term Memory\\and Gating Mechanisms}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% ============================================================
%% SECTION 1: MOTIVATION AND OVERVIEW
%% ============================================================
\section{Motivation and Overview}
\label{sec:motivation}

In Lesson~67, we proved that the gradient of the loss with respect to
an earlier hidden state decays exponentially as $\gamma^{t-\tau}$ when
$\|\mathbf{U}\| < 1$.  This means a vanilla RNN cannot learn to
associate an input at time $\tau = 1$ with a target at time $t = 100$:
the gradient signal is attenuated by a factor of $\gamma^{99}$, which
is negligible for any $\gamma < 1$.  The root cause is the
\emph{multiplicative} nature of gradient flow through the recurrence
$\mathbf{h}_t = \tanh(\mathbf{W}\mathbf{x}_t +
\mathbf{U}\mathbf{h}_{t-1} + \mathbf{b})$: each backward step
multiplies by $\diag(1 - \mathbf{h}_t^2)\mathbf{U}$, and the product
of many such matrices either shrinks or grows.

What if we could replace this multiplicative chain with an
\emph{additive} update?  The Long Short-Term Memory (LSTM)
architecture, introduced by \citet{hochreiter1997}, does exactly this.
It introduces a \emph{cell state} $\mathbf{c}_t$ that is updated
additively: $\mathbf{c}_t = \mathbf{f}_t \odot \mathbf{c}_{t-1} +
\mathbf{i}_t \odot \tilde{\mathbf{c}}_t$.  When the forget gate
$\mathbf{f}_t \approx \mathbf{1}$, gradients flow through the cell
state with minimal decay, creating a ``gradient highway'' that allows
information to persist over hundreds of time steps.  The tools
developed in this lesson will let us formalize the LSTM architecture,
derive the BPTT algorithm through its gated structure, and prove that
the cell state update mitigates the vanishing gradient problem.

\begin{rlconnection}[title={Why LSTMs Matter for RL/ML}]
\begin{itemize}[nosep]
  \item \textbf{Recurrent policies in POMDPs.}  LSTM-based policies
    in PPO and A3C can maintain belief states over long episodes,
    enabling effective decision-making under partial observability.
  \item \textbf{Long-range credit assignment.}  The LSTM's ability to
    preserve gradient flow over hundreds of steps is critical for
    tasks where rewards are sparse or delayed.
  \item \textbf{Sequence modeling benchmarks.}  LSTMs achieve
    state-of-the-art results on speech recognition, machine
    translation, and language modeling---tasks that require capturing
    long-range dependencies.
  \item \textbf{Gating as a design principle.}  The gating mechanism
    introduced by LSTMs appears throughout deep RL: highway networks,
    residual connections, and attention mechanisms all use similar
    ideas of controlled information flow.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
Vanilla RNN architecture and BPTT (Lesson~67, Definitions~4.2--5.3
and Theorems~6.1--6.4).  Vanishing gradient problem and the Jacobian
product structure (Lesson~67, Section~7).  Sigmoid function
$\sigma(z) = 1/(1 + e^{-z})$ and its derivative
$\sigma'(z) = \sigma(z)(1 - \sigma(z))$ (Lesson~63, Section~3).
Hadamard (element-wise) product.

\paragraph{Learning objectives.}
After completing this lesson, you will be able to:
\begin{itemize}[nosep]
  \item \textbf{Define} the LSTM cell architecture, including the
    forget gate, input gate, candidate cell state, cell state
    update, and output gate.
  \item \textbf{Derive} the complete LSTM forward equations and
    \textbf{compute} the parameter count.
  \item \textbf{Derive} the BPTT algorithm through the LSTM,
    including the cell state gradient recurrence and gate gradients.
  \item \textbf{Prove} that the cell state gradient recurrence
    $\partial \mathbf{c}_t / \partial \mathbf{c}_{t-1} =
    \diag(\mathbf{f}_t)$ allows gradients to flow with minimal
    decay when $\mathbf{f}_t \approx \mathbf{1}$.
  \item \textbf{Compare} the gradient flow in LSTMs versus vanilla
    RNNs and \textbf{explain} why the additive cell state update
    solves the vanishing gradient problem.
  \item \textbf{Describe} LSTM variants (peephole connections,
    coupled input-forget gate) and \textbf{evaluate} their
    trade-offs.
\end{itemize}

%% ============================================================
%% SECTION 2: REFERENCES
%% ============================================================
\section{References}
\label{sec:references}

\begin{itemize}
  \item \citet{hochreiter1997} --- The original LSTM paper:
    architecture, gating mechanisms, and experimental validation
    on long-range dependency tasks.
  \item \citet{gers2000} --- Introduction of the forget gate, which
    was absent from the original LSTM and is now considered essential.
  \item \citet{greff2017} --- Comprehensive empirical study of LSTM
    variants, ablation studies, and practical guidelines.
  \item \citet[Chapter~10]{goodfellow2016} --- Textbook treatment of
    LSTM architecture and its relationship to gated recurrent
    networks.
  \item \citet{graves2012} --- Supervised sequence labelling with
    LSTMs, including CTC and bidirectional architectures.
\end{itemize}

%% ============================================================
%% SECTION 3: THE GATING PRINCIPLE
%% ============================================================
\section{The Gating Principle}
\label{sec:gating-principle}

The central idea behind the LSTM is to control information flow using
\emph{gates}: vectors whose components lie in $(0, 1)$, which act as
soft switches via element-wise multiplication.

\begin{definition}[Sigmoid Gate]
\label{def:sigmoid-gate}
A \emph{sigmoid gate} is a vector
$\mathbf{g} \in (0, 1)^m$ computed as:
\[
  \mathbf{g} = \sigma(\mathbf{W}_g \mathbf{x}_t +
    \mathbf{U}_g \mathbf{h}_{t-1} + \mathbf{b}_g),
\]
where $\sigma(z) = 1/(1 + e^{-z})$ is the logistic sigmoid applied
element-wise.
\end{definition}

\begin{definition}[Hadamard Product]
\label{def:hadamard}
For vectors $\mathbf{a}, \mathbf{b} \in \R^m$, the \emph{Hadamard
product} $\mathbf{a} \odot \mathbf{b} \in \R^m$ is defined
component-wise:
\[
  (\mathbf{a} \odot \mathbf{b})_j = a_j \cdot b_j,
  \quad j = 1, \ldots, m.
\]
In matrix form: $\mathbf{a} \odot \mathbf{b} =
\diag(\mathbf{a})\mathbf{b} = \diag(\mathbf{b})\mathbf{a}$.
\end{definition}

\begin{remark}[Gating as Soft Selection]
\label{rem:gating}
The operation $\mathbf{g} \odot \mathbf{v}$ where
$\mathbf{g} \in (0, 1)^m$ acts as a soft selection: when
$g_j \approx 1$, component $v_j$ passes through; when
$g_j \approx 0$, it is suppressed.  This allows the network to
\emph{learn} which information to keep, which to discard, and which
to update---all in a differentiable manner.
\end{remark}

\begin{proposition}[Derivative Properties]
\label{prop:activation-derivatives}
The activation functions used in LSTM gates satisfy:
\begin{align}
  \frac{d\sigma}{dz} &= \sigma(z)(1 - \sigma(z)),
    \label{eq:sigmoid-deriv} \\
  \frac{d\tanh}{dz} &= 1 - \tanh^2(z) = \mathrm{sech}^2(z).
    \label{eq:tanh-deriv}
\end{align}
Both derivatives can be expressed in terms of the function values,
enabling efficient computation during backpropagation.
\end{proposition}

\begin{proof}
For the sigmoid:
$\frac{d}{dz}\frac{1}{1+e^{-z}} = \frac{e^{-z}}{(1+e^{-z})^2}
= \frac{1}{1+e^{-z}} \cdot \frac{e^{-z}}{1+e^{-z}}
= \sigma(z)(1 - \sigma(z))$.
The $\tanh$ derivative follows from
$\tanh(z) = 2\sigma(2z) - 1$, giving
$\frac{d\tanh}{dz} = 2 \cdot 2\sigma(2z)(1-\sigma(2z)) =
1 - \tanh^2(z)$.
\end{proof}

We now construct the LSTM cell from four gating components.

%% ============================================================
%% SECTION 4: LSTM CELL ARCHITECTURE
%% ============================================================
\section{LSTM Cell Architecture}
\label{sec:lstm-cell}

The LSTM cell maintains two state vectors at each time step:
\begin{itemize}[nosep]
  \item The \emph{cell state} $\mathbf{c}_t \in \R^m$: long-term
    memory, updated additively.
  \item The \emph{hidden state} $\mathbf{h}_t \in (-1, 1)^m$:
    short-term memory and output, derived from the cell state.
\end{itemize}
Four components govern the cell update at each time step.

\subsection{The Forget Gate}

\begin{definition}[Forget Gate]
\label{def:forget-gate}
The \emph{forget gate} $\mathbf{f}_t \in (0, 1)^m$ at time $t$
determines which components of the previous cell state to retain:
\begin{equation}
  \boxed{\mathbf{f}_t = \sigma(\mathbf{W}_f \mathbf{x}_t +
    \mathbf{U}_f \mathbf{h}_{t-1} + \mathbf{b}_f)},
  \label{eq:forget-gate}
\end{equation}
where $\mathbf{W}_f \in \R^{m \times n}$,
$\mathbf{U}_f \in \R^{m \times m}$, and
$\mathbf{b}_f \in \R^m$.
\end{definition}

\begin{remark}[Interpretation]
When $f_t^{(j)} \approx 0$, the $j$-th component of the cell state
is ``forgotten'' (erased).  When $f_t^{(j)} \approx 1$, it is
preserved.  The forget gate was not part of the original LSTM
\citep{hochreiter1997} but was added by \citet{gers2000} and is now
considered essential.
\end{remark}

\subsection{The Input Gate and Candidate Cell State}

\begin{definition}[Input Gate]
\label{def:input-gate}
The \emph{input gate} $\mathbf{i}_t \in (0, 1)^m$ controls how much
of the new candidate information to incorporate:
\begin{equation}
  \boxed{\mathbf{i}_t = \sigma(\mathbf{W}_i \mathbf{x}_t +
    \mathbf{U}_i \mathbf{h}_{t-1} + \mathbf{b}_i)},
  \label{eq:input-gate}
\end{equation}
where $\mathbf{W}_i \in \R^{m \times n}$,
$\mathbf{U}_i \in \R^{m \times m}$, and
$\mathbf{b}_i \in \R^m$.
\end{definition}

\begin{definition}[Candidate Cell State]
\label{def:candidate}
The \emph{candidate cell state}
$\tilde{\mathbf{c}}_t \in (-1, 1)^m$ represents potential new
information:
\begin{equation}
  \boxed{\tilde{\mathbf{c}}_t = \tanh(\mathbf{W}_c \mathbf{x}_t +
    \mathbf{U}_c \mathbf{h}_{t-1} + \mathbf{b}_c)},
  \label{eq:candidate}
\end{equation}
where $\mathbf{W}_c \in \R^{m \times n}$,
$\mathbf{U}_c \in \R^{m \times m}$, and
$\mathbf{b}_c \in \R^m$.
\end{definition}

\subsection{Cell State Update}

\begin{definition}[Cell State Update]
\label{def:cell-update}
The cell state is updated by combining retained old information with
gated new information:
\begin{equation}
  \boxed{\mathbf{c}_t = \mathbf{f}_t \odot \mathbf{c}_{t-1} +
    \mathbf{i}_t \odot \tilde{\mathbf{c}}_t}.
  \label{eq:cell-update}
\end{equation}
\end{definition}

\begin{remark}[Additive Structure]
\label{rem:additive}
The cell state update~\eqref{eq:cell-update} is \emph{additive}:
$\mathbf{c}_t$ is a weighted sum of the previous cell state and a
new candidate.  This is the key architectural innovation.  Unlike the
vanilla RNN's multiplicative update
$\mathbf{h}_t = \tanh(\mathbf{U}\mathbf{h}_{t-1} + \cdots)$, the
additive structure allows the Jacobian
$\partial \mathbf{c}_t / \partial \mathbf{c}_{t-1}$ to be close to
the identity, enabling gradient flow over long distances.
\end{remark}

\subsection{The Output Gate and Hidden State}

\begin{definition}[Output Gate]
\label{def:output-gate}
The \emph{output gate} $\mathbf{o}_t \in (0, 1)^m$ determines which
components of the cell state are exposed as the hidden state:
\begin{equation}
  \boxed{\mathbf{o}_t = \sigma(\mathbf{W}_o \mathbf{x}_t +
    \mathbf{U}_o \mathbf{h}_{t-1} + \mathbf{b}_o)},
  \label{eq:output-gate}
\end{equation}
where $\mathbf{W}_o \in \R^{m \times n}$,
$\mathbf{U}_o \in \R^{m \times m}$, and
$\mathbf{b}_o \in \R^m$.
\end{definition}

\begin{definition}[Hidden State Output]
\label{def:hidden-output}
The hidden state is computed by filtering the cell state through the
output gate:
\begin{equation}
  \boxed{\mathbf{h}_t = \mathbf{o}_t \odot \tanh(\mathbf{c}_t)}.
  \label{eq:hidden-output}
\end{equation}
The $\tanh$ squashes the cell state to $(-1, 1)$, and the output
gate selects which components to expose.
\end{definition}

\begin{example}[LSTM as Selective Memory]
\label{ex:selective-memory}
Consider a language model processing the sentence ``The cat, which
was sitting on the mat, \textbf{was} sleeping.''  The LSTM must
remember that the subject is singular (``cat'') across the relative
clause to correctly predict the verb ``was'' (not ``were''):
\begin{itemize}[nosep]
  \item At ``cat'': the input gate stores subject number information
    in a cell state component.
  \item During the relative clause: the forget gate keeps this
    component close to $1$, preserving the information.
  \item At the verb position: the output gate exposes the number
    information to influence the prediction.
\end{itemize}
A vanilla RNN would lose this information due to the vanishing
gradient over the 7-word gap.
\end{example}

%% ============================================================
%% SECTION 5: COMPLETE LSTM FORWARD PASS
%% ============================================================
\section{Complete LSTM Forward Pass}
\label{sec:forward-pass}

\begin{keyresult}[title={LSTM Forward Equations}]
At each time step $t = 1, \ldots, T$:
\begin{align}
  \mathbf{f}_t &= \sigma(\mathbf{W}_f \mathbf{x}_t +
    \mathbf{U}_f \mathbf{h}_{t-1} + \mathbf{b}_f),
    && \text{[Forget Gate]} \label{eq:lstm-f} \\
  \mathbf{i}_t &= \sigma(\mathbf{W}_i \mathbf{x}_t +
    \mathbf{U}_i \mathbf{h}_{t-1} + \mathbf{b}_i),
    && \text{[Input Gate]} \label{eq:lstm-i} \\
  \tilde{\mathbf{c}}_t &= \tanh(\mathbf{W}_c \mathbf{x}_t +
    \mathbf{U}_c \mathbf{h}_{t-1} + \mathbf{b}_c),
    && \text{[Candidate]} \label{eq:lstm-c-tilde} \\
  \mathbf{c}_t &= \mathbf{f}_t \odot \mathbf{c}_{t-1} +
    \mathbf{i}_t \odot \tilde{\mathbf{c}}_t,
    && \text{[Cell Update]} \label{eq:lstm-c} \\
  \mathbf{o}_t &= \sigma(\mathbf{W}_o \mathbf{x}_t +
    \mathbf{U}_o \mathbf{h}_{t-1} + \mathbf{b}_o),
    && \text{[Output Gate]} \label{eq:lstm-o} \\
  \mathbf{h}_t &= \mathbf{o}_t \odot \tanh(\mathbf{c}_t),
    && \text{[Hidden State]} \label{eq:lstm-h}
\end{align}
with initial conditions $\mathbf{h}_0 = \mathbf{0} \in \R^m$ and
$\mathbf{c}_0 = \mathbf{0} \in \R^m$.
\end{keyresult}

\begin{proposition}[LSTM Parameter Count]
\label{prop:lstm-params}
For an LSTM with input dimension $n$ and hidden dimension $m$, the
total number of trainable parameters (excluding the output layer) is:
\begin{equation}
  P = 4(mn + m^2 + m) = 4m(n + m + 1).
  \label{eq:lstm-params}
\end{equation}
\end{proposition}

\begin{proof}
Each of the four components (forget gate, input gate, candidate,
output gate) has:
\begin{itemize}[nosep]
  \item Input weight matrix: $m \times n$ parameters,
  \item Recurrent weight matrix: $m \times m$ parameters,
  \item Bias vector: $m$ parameters.
\end{itemize}
Total: $4 \times (mn + m^2 + m) = 4m(n + m + 1)$.
\end{proof}

\begin{remark}[LSTM vs Vanilla RNN Parameters]
An LSTM has approximately $4\times$ more parameters than a vanilla
RNN of the same hidden dimension, due to the four separate
gate/candidate components.  Despite this overhead, the improved
gradient flow typically justifies the additional cost.
\end{remark}

\begin{definition}[Compact Matrix Formulation]
\label{def:compact}
For computational efficiency, define concatenated parameters:
\begin{equation}
  \mathbf{W} = \begin{pmatrix} \mathbf{W}_f \\ \mathbf{W}_i \\
    \mathbf{W}_c \\ \mathbf{W}_o \end{pmatrix} \in \R^{4m \times n},
  \quad
  \mathbf{U} = \begin{pmatrix} \mathbf{U}_f \\ \mathbf{U}_i \\
    \mathbf{U}_c \\ \mathbf{U}_o \end{pmatrix} \in \R^{4m \times m},
  \quad
  \mathbf{b} = \begin{pmatrix} \mathbf{b}_f \\ \mathbf{b}_i \\
    \mathbf{b}_c \\ \mathbf{b}_o \end{pmatrix} \in \R^{4m}.
  \label{eq:compact-params}
\end{equation}
All pre-activations are computed in a single operation:
\[
  \mathbf{z}_t = \mathbf{W}\mathbf{x}_t + \mathbf{U}\mathbf{h}_{t-1}
    + \mathbf{b} \in \R^{4m},
\]
then split into four segments of size $m$ and passed through the
appropriate activations ($\sigma$ for gates, $\tanh$ for candidate).
This reduces the number of matrix multiplications from 8 to 2.
\end{definition}

\begin{example}[LSTM Parameter Count]
\label{ex:param-count}
For an LSTM processing word embeddings of dimension $n = 300$ with
hidden size $m = 512$:
\[
  P = 4 \times 512 \times (300 + 512 + 1) = 4 \times 512 \times 813
    = 1{,}664{,}256 \approx 1.66\text{M parameters}.
\]
A vanilla RNN of the same dimensions has
$512 \times (300 + 512 + 1) = 416{,}256 \approx 0.42\text{M}$, about
$4\times$ fewer.
\end{example}

%% ============================================================
%% SECTION 6: BPTT THROUGH LSTM
%% ============================================================
\section{Backpropagation Through the LSTM}
\label{sec:bptt-lstm}

Given a sequence loss $\mathcal{L} = \sum_{t=1}^T \mathcal{L}_t$,
we derive the gradients by propagating errors backward through the
LSTM computational graph.  Define the error signals:
\begin{align}
  \boldsymbol{\delta}_t^h &= \frac{\partial \mathcal{L}}
    {\partial \mathbf{h}_t}, \label{eq:delta-h} \\
  \boldsymbol{\delta}_t^c &= \frac{\partial \mathcal{L}}
    {\partial \mathbf{c}_t}. \label{eq:delta-c}
\end{align}

\begin{theorem}[Cell State Gradient Recurrence]
\label{thm:cell-gradient}
The cell state error signal satisfies the recurrence:
\begin{equation}
  \boxed{\boldsymbol{\delta}_t^c = \boldsymbol{\delta}_t^h \odot
    \mathbf{o}_t \odot (1 - \tanh^2(\mathbf{c}_t)) +
    \boldsymbol{\delta}_{t+1}^c \odot \mathbf{f}_{t+1}}.
  \label{eq:cell-recurrence}
\end{equation}
\end{theorem}

\begin{proof}
The cell state $\mathbf{c}_t$ influences the total loss through two
paths:
\begin{enumerate}[nosep]
  \item Through the current hidden state:
    $\mathbf{c}_t \to \tanh(\mathbf{c}_t) \to \mathbf{h}_t$.
  \item Through the next cell state:
    $\mathbf{c}_t \to \mathbf{c}_{t+1}$ via the cell update equation.
\end{enumerate}

For path (1): since $\mathbf{h}_t = \mathbf{o}_t \odot
\tanh(\mathbf{c}_t)$,
\[
  \frac{\partial \mathbf{h}_t}{\partial \mathbf{c}_t}
    = \diag(\mathbf{o}_t) \diag(1 - \tanh^2(\mathbf{c}_t)).
\]
So the contribution is $\boldsymbol{\delta}_t^h \odot \mathbf{o}_t
\odot (1 - \tanh^2(\mathbf{c}_t))$.

For path (2): since $\mathbf{c}_{t+1} = \mathbf{f}_{t+1} \odot
\mathbf{c}_t + \mathbf{i}_{t+1} \odot \tilde{\mathbf{c}}_{t+1}$,
and in the standard LSTM (without peephole connections) the gates
$\mathbf{f}_{t+1}$, $\mathbf{i}_{t+1}$, and
$\tilde{\mathbf{c}}_{t+1}$ do not depend on $\mathbf{c}_t$:
\[
  \frac{\partial \mathbf{c}_{t+1}}{\partial \mathbf{c}_t}
    = \diag(\mathbf{f}_{t+1}).
\]
Combining both paths gives the recurrence~\eqref{eq:cell-recurrence}.
\end{proof}

\begin{keyresult}[title={LSTM Gradient Highway}]
The cell state gradient recurrence
$\boldsymbol{\delta}_t^c = (\text{local term}) +
\boldsymbol{\delta}_{t+1}^c \odot \mathbf{f}_{t+1}$
is the key to the LSTM's success.  The gradient flows backward
through the cell state by multiplication with the forget gate values
$\mathbf{f}_{t+1}$.  When $\mathbf{f}_t \approx \mathbf{1}$:
\[
  \boldsymbol{\delta}_1^c \approx \boldsymbol{\delta}_T^c \cdot
    \prod_{t=2}^{T} \diag(\mathbf{f}_t) \approx
    \boldsymbol{\delta}_T^c,
\]
achieving near-perfect gradient preservation over $T$ steps.
\end{keyresult}

The gate pre-activation gradients are used to compute weight
gradients.  Define the pre-activation error signals:
\begin{align}
  \boldsymbol{\delta}_t^f &= \boldsymbol{\delta}_t^c \odot
    \mathbf{c}_{t-1} \odot \mathbf{f}_t \odot (1 - \mathbf{f}_t),
    && \text{[Forget Gate]} \label{eq:delta-f} \\
  \boldsymbol{\delta}_t^i &= \boldsymbol{\delta}_t^c \odot
    \tilde{\mathbf{c}}_t \odot \mathbf{i}_t \odot
    (1 - \mathbf{i}_t),
    && \text{[Input Gate]} \label{eq:delta-i} \\
  \boldsymbol{\delta}_t^{\tilde{c}} &= \boldsymbol{\delta}_t^c
    \odot \mathbf{i}_t \odot (1 - \tilde{\mathbf{c}}_t^2),
    && \text{[Candidate]} \label{eq:delta-c-tilde} \\
  \boldsymbol{\delta}_t^o &= \boldsymbol{\delta}_t^h \odot
    \tanh(\mathbf{c}_t) \odot \mathbf{o}_t \odot
    (1 - \mathbf{o}_t).
    && \text{[Output Gate]} \label{eq:delta-o}
\end{align}

\begin{theorem}[LSTM Weight Gradients]
\label{thm:lstm-weight-grads}
The gradients with respect to the forget gate parameters are:
\begin{align}
  \frac{\partial \mathcal{L}}{\partial \mathbf{W}_f}
    &= \sum_{t=1}^{T} \boldsymbol{\delta}_t^f \mathbf{x}_t^\top,
    \label{eq:grad-Wf} \\
  \frac{\partial \mathcal{L}}{\partial \mathbf{U}_f}
    &= \sum_{t=1}^{T} \boldsymbol{\delta}_t^f
       \mathbf{h}_{t-1}^\top,
    \label{eq:grad-Uf} \\
  \frac{\partial \mathcal{L}}{\partial \mathbf{b}_f}
    &= \sum_{t=1}^{T} \boldsymbol{\delta}_t^f.
    \label{eq:grad-bf}
\end{align}
Analogous expressions hold for the input gate, candidate, and output
gate parameters (replacing the superscript $f$ with $i$,
$\tilde{c}$, and $o$ respectively).
\end{theorem}

\begin{proof}
Consider the forget gate at time $t$.  Let
$\mathbf{z}_t^f = \mathbf{W}_f \mathbf{x}_t +
\mathbf{U}_f \mathbf{h}_{t-1} + \mathbf{b}_f$ be the
pre-activation.  Then $\mathbf{f}_t = \sigma(\mathbf{z}_t^f)$
and the chain rule gives:
\[
  \frac{\partial \mathcal{L}}{\partial \mathbf{z}_t^f}
    = \frac{\partial \mathcal{L}}{\partial \mathbf{c}_t}
      \odot \frac{\partial \mathbf{c}_t}{\partial \mathbf{f}_t}
      \odot \frac{\partial \mathbf{f}_t}{\partial \mathbf{z}_t^f}
    = \boldsymbol{\delta}_t^c \odot \mathbf{c}_{t-1} \odot
      \mathbf{f}_t \odot (1 - \mathbf{f}_t)
    = \boldsymbol{\delta}_t^f.
\]
Since $\mathbf{z}_t^f = \mathbf{W}_f \mathbf{x}_t +
\mathbf{U}_f \mathbf{h}_{t-1} + \mathbf{b}_f$, the weight
gradients follow by the standard matrix calculus:
$\frac{\partial \mathcal{L}}{\partial \mathbf{W}_f} =
\sum_{t} \boldsymbol{\delta}_t^f \mathbf{x}_t^\top$.
The derivations for the other gates are identical in structure.
\end{proof}

\begin{remark}[Hidden State Error Signal]
The hidden state error $\boldsymbol{\delta}_t^h$ receives
contributions from the current loss $\mathcal{L}_t$ and from the
backpropagation through the gates at time $t + 1$ (since the gates
at $t + 1$ depend on $\mathbf{h}_t$).  Specifically:
\begin{equation}
  \boldsymbol{\delta}_t^h = \frac{\partial \mathcal{L}_t}
    {\partial \mathbf{h}_t} + \sum_{g \in \{f,i,c,o\}}
    \mathbf{U}_g^\top \boldsymbol{\delta}_{t+1}^g,
  \label{eq:delta-h-full}
\end{equation}
where the sum is over the four gate/candidate pre-activation
errors at time $t + 1$.
\end{remark}

%% ============================================================
%% SECTION 7: GRADIENT FLOW ANALYSIS
%% ============================================================
\section{Gradient Flow Analysis}
\label{sec:gradient-flow}

We now formally compare the gradient flow properties of the LSTM
with those of the vanilla RNN.

\begin{theorem}[LSTM Cell State Jacobian]
\label{thm:lstm-jacobian}
In the standard LSTM (without peephole connections), the Jacobian of
$\mathbf{c}_t$ with respect to $\mathbf{c}_{t-1}$ is:
\begin{equation}
  \frac{\partial \mathbf{c}_t}{\partial \mathbf{c}_{t-1}}
    = \diag(\mathbf{f}_t).
  \label{eq:lstm-jacobian}
\end{equation}
\end{theorem}

\begin{proof}
Taking the partial derivative of the cell state update
$\mathbf{c}_t = \mathbf{f}_t \odot \mathbf{c}_{t-1} +
\mathbf{i}_t \odot \tilde{\mathbf{c}}_t$ with respect to
$\mathbf{c}_{t-1}$:
\[
  \frac{\partial \mathbf{c}_t}{\partial \mathbf{c}_{t-1}}
    = \diag(\mathbf{f}_t)
      + \mathbf{c}_{t-1} \odot \frac{\partial \mathbf{f}_t}
        {\partial \mathbf{c}_{t-1}}
      + \tilde{\mathbf{c}}_t \odot
        \frac{\partial \mathbf{i}_t}{\partial \mathbf{c}_{t-1}}
      + \mathbf{i}_t \odot
        \frac{\partial \tilde{\mathbf{c}}_t}
          {\partial \mathbf{c}_{t-1}}.
\]
In the standard LSTM, the gates depend on $\mathbf{x}_t$ and
$\mathbf{h}_{t-1}$ but \emph{not directly} on $\mathbf{c}_{t-1}$,
so the last three terms vanish:
$\frac{\partial \mathbf{c}_t}{\partial \mathbf{c}_{t-1}} =
\diag(\mathbf{f}_t)$.
\end{proof}

\begin{corollary}[Long-Range Cell State Gradient]
\label{cor:long-range-lstm}
The gradient of $\mathcal{L}$ with respect to an earlier cell state
$\mathbf{c}_\tau$ ($\tau < t$) through the cell state pathway is:
\begin{equation}
  \frac{\partial \mathbf{c}_t}{\partial \mathbf{c}_\tau}
    = \prod_{i=\tau+1}^{t} \diag(\mathbf{f}_i).
  \label{eq:long-range-cell}
\end{equation}
\end{corollary}

\begin{proof}
By Theorem~\ref{thm:lstm-jacobian} and the chain rule:
$\frac{\partial \mathbf{c}_t}{\partial \mathbf{c}_\tau} =
\prod_{i=\tau+1}^{t} \frac{\partial \mathbf{c}_i}
{\partial \mathbf{c}_{i-1}} = \prod_{i=\tau+1}^{t}
\diag(\mathbf{f}_i)$.
\end{proof}

\begin{theorem}[Vanishing Gradient Mitigation]
\label{thm:vanishing-mitigation}
When $\mathbf{f}_t \approx \mathbf{1}$ for all $t$, the cell state
gradient satisfies:
\begin{equation}
  \frac{\partial \mathbf{c}_t}{\partial \mathbf{c}_\tau}
    \approx \mathbf{I},
  \label{eq:gradient-preservation}
\end{equation}
preventing the exponential decay characteristic of vanilla RNNs.
More precisely, for component $j$:
\[
  \left[\frac{\partial \mathbf{c}_t}{\partial \mathbf{c}_\tau}
  \right]_{jj} = \prod_{i=\tau+1}^{t} f_i^{(j)}.
\]
If $f_i^{(j)} \geq 1 - \epsilon$ for all $i$, then
$\prod_{i=\tau+1}^{t} f_i^{(j)} \geq (1-\epsilon)^{t-\tau}
\geq 1 - (t-\tau)\epsilon$ for small $\epsilon$.
\end{theorem}

\begin{proof}
When $\mathbf{f}_i \approx \mathbf{1}$ for all $i$, each diagonal
matrix $\diag(\mathbf{f}_i) \approx \mathbf{I}$.  The product of
$t - \tau$ such matrices is approximately $\mathbf{I}$.  For the
precise bound, use $(1-\epsilon)^k \geq 1 - k\epsilon$ (valid for
$0 \leq \epsilon \leq 1$ and $k \geq 1$, by Bernoulli's inequality).
\end{proof}

\begin{example}[LSTM vs Vanilla RNN Gradient Flow]
\label{ex:gradient-comparison}
Consider a sequence of length $T = 100$ with time gap $k = 99$.
\begin{itemize}[nosep]
  \item \textbf{Vanilla RNN} with $\|\mathbf{U}\| = 0.9$: gradient
    decays by $0.9^{99} \approx 3 \times 10^{-5}$ (effectively zero).
  \item \textbf{LSTM} with $f_i^{(j)} = 0.99$ for all $i$:
    gradient decays by $0.99^{99} \approx 0.37$ (substantial
    signal preserved).
  \item \textbf{LSTM} with $f_i^{(j)} = 1.0$: gradient decays by
    $1.0^{99} = 1$ (perfect preservation).
\end{itemize}
The LSTM's forget gate can maintain gradient flow over arbitrarily
long sequences by keeping $\mathbf{f}_t$ close to $\mathbf{1}$.
\end{example}

\begin{rlconnection}[title={Forget Gate Bias Initialization}]
In practice, the forget gate bias $\mathbf{b}_f$ is often
initialized to a positive value (e.g., $1$ or $2$) so that
$\sigma(\mathbf{b}_f) \approx 1$ at the start of training.  This
ensures that the LSTM begins with near-perfect gradient flow through
the cell state, allowing it to learn long-range dependencies from
the outset.  This trick, recommended by \citet{gers2000}, is widely
used in deep RL implementations.
\end{rlconnection}

\begin{warning}[title={Gradient Flow Through Hidden State}]
The analysis above concerns gradient flow through the
\emph{cell state} pathway.  Gradients also flow through the
\emph{hidden state} pathway via the gate recurrences
(Equation~\ref{eq:delta-h-full}), which involves multiplication by
the recurrent weight matrices $\mathbf{U}_g^\top$.  This pathway
can still suffer from vanishing gradients, but the cell state
highway provides an alternative route that bypasses this bottleneck.
\end{warning}

%% ============================================================
%% SECTION 8: LSTM VARIANTS
%% ============================================================
\section{LSTM Variants}
\label{sec:variants}

\subsection{Peephole Connections}

\begin{definition}[Peephole LSTM]
\label{def:peephole}
\emph{Peephole connections}, introduced by \citet{gers2000}, allow
the gates to access the cell state directly:
\begin{align}
  \mathbf{f}_t &= \sigma(\mathbf{W}_f \mathbf{x}_t +
    \mathbf{U}_f \mathbf{h}_{t-1} +
    \mathbf{V}_f \odot \mathbf{c}_{t-1} + \mathbf{b}_f),
    \label{eq:peep-f} \\
  \mathbf{i}_t &= \sigma(\mathbf{W}_i \mathbf{x}_t +
    \mathbf{U}_i \mathbf{h}_{t-1} +
    \mathbf{V}_i \odot \mathbf{c}_{t-1} + \mathbf{b}_i),
    \label{eq:peep-i} \\
  \mathbf{o}_t &= \sigma(\mathbf{W}_o \mathbf{x}_t +
    \mathbf{U}_o \mathbf{h}_{t-1} +
    \mathbf{V}_o \odot \mathbf{c}_t + \mathbf{b}_o),
    \label{eq:peep-o}
\end{align}
where $\mathbf{V}_f, \mathbf{V}_i, \mathbf{V}_o \in \R^m$ are
diagonal peephole weight vectors.
\end{definition}

\begin{remark}[Peephole Trade-offs]
Peephole connections add $3m$ parameters and make the Jacobian
$\partial \mathbf{c}_t / \partial \mathbf{c}_{t-1}$ more complex
(no longer purely $\diag(\mathbf{f}_t)$).  Empirical studies
\citep{greff2017} find that peephole connections provide marginal
benefit on most tasks and are not commonly used in modern
implementations.
\end{remark}

\subsection{Coupled Input-Forget Gate}

\begin{definition}[Coupled Gate LSTM]
\label{def:coupled}
The \emph{coupled input-forget gate} variant enforces the constraint
$\mathbf{i}_t = \mathbf{1} - \mathbf{f}_t$, reducing the cell state
update to:
\begin{equation}
  \mathbf{c}_t = \mathbf{f}_t \odot \mathbf{c}_{t-1} +
    (\mathbf{1} - \mathbf{f}_t) \odot \tilde{\mathbf{c}}_t.
  \label{eq:coupled}
\end{equation}
\end{definition}

\begin{remark}[Coupled Gate Interpretation]
The coupled gate makes an explicit trade-off: the network can either
retain old information ($f_t^{(j)} \approx 1$) or write new
information ($f_t^{(j)} \approx 0$), but not both simultaneously.
This reduces the parameter count by $m(n + m + 1)$ (one fewer gate)
while maintaining the core functionality.
\end{remark}

\begin{example}[Coupled Gate vs Standard LSTM]
\label{ex:coupled}
With the coupled gate, the cell state update
$\mathbf{c}_t = \mathbf{f}_t \odot \mathbf{c}_{t-1} +
(1 - \mathbf{f}_t) \odot \tilde{\mathbf{c}}_t$ is a convex
combination of the old cell state and the candidate.  This ensures
that the cell state norm does not grow unboundedly (since
$|\tilde{c}_t^{(j)}| < 1$), providing implicit regularization.
In the standard LSTM, $\mathbf{c}_t$ can grow without bound if
both $\mathbf{f}_t$ and $\mathbf{i}_t$ are simultaneously large.
\end{example}

%% ============================================================
%% SECTION 9: EXERCISES
%% ============================================================
\section{Exercises}
\label{sec:exercises}

\begin{exercise}[LSTM Forward Pass]
\textnormal{(\(\ast\))} \\
Consider a scalar LSTM ($n = m = 1$) with
$W_f = W_i = W_c = W_o = 1$, $U_f = U_i = U_c = U_o = 0.5$,
$b_f = 1$, and $b_i = b_c = b_o = 0$.  Starting from
$h_0 = c_0 = 0$, compute the cell state $c_1, c_2$ and hidden state
$h_1, h_2$ for inputs $x_1 = 1, x_2 = -1$.
\end{exercise}

\begin{exercise}[Cell State Gradient]
\textnormal{(\(\ast\ast\))} \\
For an LSTM with $m = 2$ and constant forget gate values
$\mathbf{f}_t = (0.95, 0.99)^\top$ for all $t$, compute the cell
state gradient $\partial \mathbf{c}_{100} / \partial \mathbf{c}_1$
and compare it with the gradient in a vanilla RNN with
$\|\mathbf{U}\| = 0.95$.  Which architecture preserves the gradient
signal better over 100 steps?
\end{exercise}

\begin{exercise}[LSTM vs Coupled Gate Parameter Count]
\textnormal{(\(\ast\))} \\
For an LSTM with input dimension $n = 256$ and hidden dimension
$m = 512$, compute the parameter count for: (a)~the standard LSTM,
(b)~the coupled input-forget gate variant.  What percentage of
parameters does the coupled variant save?
\end{exercise}

\begin{exercise}[Peephole Jacobian]
\textnormal{(\(\ast\ast\ast\))} \\
Derive the Jacobian $\partial \mathbf{c}_t / \partial
\mathbf{c}_{t-1}$ for the peephole LSTM
(Definition~\ref{def:peephole}).  Show that it is no longer
$\diag(\mathbf{f}_t)$ and identify the additional terms.  Under
what conditions do these terms improve or degrade gradient flow
compared to the standard LSTM?

\textit{Hint:} The forget and input gates now depend on
$\mathbf{c}_{t-1}$ through the peephole connections.
\end{exercise}

\begin{exercise}[Gradient Flow with Saturated Gates]
\textnormal{(\(\ast\ast\))} \\
Suppose the forget gate saturates at $\mathbf{f}_t = \mathbf{1}$
(all components exactly $1$).  Show that the cell state gradient
$\partial \mathbf{c}_t / \partial \mathbf{c}_\tau = \mathbf{I}$ for
all $\tau < t$.  Then explain why this ``perfect'' gradient flow
is problematic: what happens to the cell state values
$\|\mathbf{c}_t\|$ as $t$ grows, and how does this affect the
numerical stability of the network?
\end{exercise}

%% ============================================================
%% SECTION 10: SUMMARY
%% ============================================================
\section{Summary}
\label{sec:summary}

\begin{itemize}
  \item The LSTM introduces a \emph{cell state} $\mathbf{c}_t$
    updated additively:
    $\mathbf{c}_t = \mathbf{f}_t \odot \mathbf{c}_{t-1} +
    \mathbf{i}_t \odot \tilde{\mathbf{c}}_t$.  This additive
    structure is the key architectural innovation.
  \item Four gating components---forget gate $\mathbf{f}_t$, input
    gate $\mathbf{i}_t$, candidate $\tilde{\mathbf{c}}_t$, and
    output gate $\mathbf{o}_t$---control information flow using
    sigmoid-gated element-wise products.
  \item The cell state Jacobian is
    $\partial \mathbf{c}_t / \partial \mathbf{c}_{t-1} =
    \diag(\mathbf{f}_t)$, which is close to the identity when
    $\mathbf{f}_t \approx \mathbf{1}$, creating a ``gradient highway''
    that mitigates the vanishing gradient problem.
  \item BPTT through the LSTM involves the cell state gradient
    recurrence $\boldsymbol{\delta}_t^c =
    \boldsymbol{\delta}_t^h \odot \mathbf{o}_t \odot
    (1 - \tanh^2(\mathbf{c}_t)) +
    \boldsymbol{\delta}_{t+1}^c \odot \mathbf{f}_{t+1}$
    and four gate pre-activation errors.
  \item The LSTM has $4m(n + m + 1)$ parameters---approximately
    $4\times$ more than a vanilla RNN---due to the four separate
    gate/candidate components.
  \item Variants include peephole connections (gates access cell
    state directly) and the coupled input-forget gate
    ($\mathbf{i}_t = \mathbf{1} - \mathbf{f}_t$), which trades
    flexibility for fewer parameters.
  \item Initializing the forget gate bias to a positive value
    ($b_f \approx 1$--$2$) ensures good gradient flow from the
    start of training.
\end{itemize}

%% ============================================================
%% SECTION 11: FURTHER READING
%% ============================================================
\section{Further Reading}
\label{sec:further-reading}

\begin{itemize}
  \item Lesson~69 introduces the Gated Recurrent Unit (GRU), a
    simplified gated architecture with fewer parameters, and covers
    bidirectional RNNs, deep stacked RNNs, and the encoder-decoder
    framework.
  \item \citet{greff2017} provides a comprehensive empirical
    evaluation of LSTM variants (peephole, coupled gates,
    no-input-gate, no-forget-gate, etc.), finding that the
    forget gate and output activation are the most critical
    components.
  \item \citet{hochreiter1997} is the original LSTM paper and
    contains the detailed analysis of the \emph{constant error
    carousel}---the precursor to the modern cell state update.
  \item \citet{graves2012} covers practical applications of
    LSTMs to sequence labelling, including speech recognition
    and handwriting generation.
  \item The connection between LSTMs and residual networks
    (both use additive shortcuts to improve gradient flow)
    is explored in the deep RL literature on recurrent policies
    (Phase~04, Lessons~77--80).
\end{itemize}

\bibliography{nn-refs}
\end{document}
