\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 67: Recurrent Neural Networks and BPTT}}
\fancyhead[R]{\thepage}
\title{Lesson 67: Recurrent Neural Networks\\and Backpropagation Through Time}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% ============================================================
%% SECTION 1: MOTIVATION AND OVERVIEW
%% ============================================================
\section{Motivation and Overview}
\label{sec:motivation}

Consider an RL agent navigating a partially observable environment.
At each time step, the agent receives an observation $\mathbf{x}_t$
that depends on the full history of states and actions.  A feedforward
network maps each observation independently to an action, discarding
all temporal structure.  Yet the optimal policy often depends on the
\emph{sequence} of past observations---for instance, distinguishing
whether a partially visible object is moving left or right requires
comparing the current frame with previous ones.  How can a neural
network process variable-length sequences while maintaining a
fixed-size parameter set?

Recurrent Neural Networks (RNNs) answer this question by introducing
a \emph{hidden state} $\mathbf{h}_t$ that is updated at each time
step as a function of the current input and the previous hidden
state.  This recurrence compresses the entire history into a
fixed-dimensional vector, enabling the network to model temporal
dependencies.  The learning algorithm that trains RNNs---Backpropagation
Through Time (BPTT)---unfolds the recurrence into a deep feedforward
graph and applies the chain rule.  However, the repeated multiplication
of Jacobian matrices along this unfolded graph leads to a fundamental
pathology: gradients either \emph{vanish} or \emph{explode}
exponentially with the sequence length.  The tools developed in this
lesson will let us formalize the RNN architecture, derive the BPTT
algorithm, and prove sharp bounds on gradient decay and growth.

\begin{rlconnection}[title={Why RNNs and BPTT Matter for RL/ML}]
\begin{itemize}[nosep]
  \item \textbf{Partial observability in RL.}  When the agent cannot
    observe the full state, recurrent policies (e.g., LSTM-based
    actors in PPO) use hidden states to approximate belief states
    over the underlying MDP.
  \item \textbf{Sequence prediction.}  Language modeling, speech
    recognition, and time-series forecasting all require processing
    variable-length sequences---the core capability of RNNs.
  \item \textbf{Gradient pathology.}  The vanishing/exploding gradient
    problem discovered in vanilla RNNs motivates gated architectures
    (LSTM, GRU) studied in Lessons~68--69, and also informs gradient
    clipping strategies used throughout deep RL.
  \item \textbf{Credit assignment over time.}  BPTT is the temporal
    analogue of spatial backpropagation; understanding its limitations
    clarifies why long-horizon credit assignment remains challenging
    in deep RL.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
Feedforward neural networks and backpropagation (Lesson~63,
Section~3).  Matrix norms, spectral norm, and eigenvalues
(Lesson~8, Definitions~3.1--3.3).  Chain rule for vector-valued
functions and Jacobian matrices (Lesson~3, Theorem~2.3; Lesson~8,
Section~5).  Sequences and convergence (Lesson~1, Section~4).

\paragraph{Learning objectives.}
After completing this lesson, you will be able to:
\begin{itemize}[nosep]
  \item \textbf{Define} the vanilla RNN architecture and
    \textbf{derive} the forward equations for hidden state
    updates, output computation, and sequence loss.
  \item \textbf{Derive} the complete BPTT algorithm, including
    the hidden state error recurrence and weight gradient formulas.
  \item \textbf{Prove} that gradients in vanilla RNNs decay
    exponentially when the spectral norm $\|\mathbf{U}\| < 1$
    (vanishing gradient theorem).
  \item \textbf{Prove} that gradients can grow exponentially
    when $\|\mathbf{U}\| > 1$ (exploding gradient theorem).
  \item \textbf{Analyze} the eigenvalue structure of the recurrent
    weight matrix and its role in determining gradient behavior.
  \item \textbf{Explain} gradient clipping, orthogonal initialization,
    and truncated BPTT as mitigation strategies, and \textbf{evaluate}
    their limitations.
\end{itemize}

%% ============================================================
%% SECTION 2: REFERENCES
%% ============================================================
\section{References}
\label{sec:references}

\begin{itemize}
  \item \citet[Chapter~10]{goodfellow2016} --- Comprehensive treatment
    of RNN architectures, BPTT, and gradient flow analysis.
  \item \citet{elman1990} --- Original Elman network introducing the
    simple recurrent architecture with hidden state feedback.
  \item \citet{rumelhart1986} --- Backpropagation algorithm for
    feedforward networks; the foundation on which BPTT builds.
  \item \citet{bengio1994} --- Seminal analysis of why learning
    long-term dependencies with gradient descent is difficult.
  \item \citet{pascanu2013} --- Detailed analysis of the vanishing
    and exploding gradient problem with gradient clipping solutions.
  \item \citet{hochreiter1991} --- Early identification of the
    vanishing gradient problem in recurrent networks.
\end{itemize}

%% ============================================================
%% SECTION 3: SEQUENCE MODELING
%% ============================================================
\section{Sequence Modeling}
\label{sec:sequences}

\begin{definition}[Sequence]
\label{def:sequence}
A \emph{sequence} of length $T$ over a set $\mathcal{X}$ is an
ordered tuple
\[
  \mathbf{x}_{1:T} = (\mathbf{x}_1, \mathbf{x}_2, \ldots,
    \mathbf{x}_T), \quad
  \mathbf{x}_t \in \mathcal{X} \text{ for all } t \in \{1,\ldots,T\}.
\]
\end{definition}

The fundamental challenge in sequence modeling is to learn a function
that captures dependencies between elements at different positions.
Feedforward networks treat each input independently, discarding
positional information unless it is explicitly encoded.

\begin{definition}[Sequence-to-Sequence Map]
\label{def:seq2seq-map}
A \emph{sequence-to-sequence map} is a function
$F : \mathcal{X}^T \to \mathcal{Y}^T$ that maps an input sequence
$\mathbf{x}_{1:T}$ to an output sequence
$\mathbf{y}_{1:T} = (F(\mathbf{x}_{1:T})_1, \ldots,
F(\mathbf{x}_{1:T})_T)$.  The map is \emph{causal} if each output
$\mathbf{y}_t$ depends only on current and past inputs
$\mathbf{x}_1, \ldots, \mathbf{x}_t$.
\end{definition}

\begin{example}[Language Modeling]
\label{ex:language-model}
In language modeling, $\mathbf{x}_t$ is the one-hot encoding of the
$t$-th word and $\mathbf{y}_t = P(\mathbf{x}_{t+1} \mid
\mathbf{x}_{1:t})$ is the conditional distribution over the next word.
This is a causal sequence-to-sequence map: predicting the next word
requires only the past context.
\end{example}

\begin{rlconnection}[title={Sequences in RL}]
In reinforcement learning, the agent's trajectory
$(\mathbf{o}_1, a_1, r_1, \mathbf{o}_2, a_2, r_2, \ldots)$ is a
sequence.  Under partial observability (POMDPs), the optimal policy
$\pi^*$ may depend on the entire observation history
$\mathbf{o}_{1:t}$, not just the current observation $\mathbf{o}_t$.
An RNN can approximate this by encoding the history into a hidden
state $\mathbf{h}_t$ that serves as a surrogate for the belief state.
\end{rlconnection}

We now introduce the recurrence principle that allows a neural
network to process sequences with a fixed number of parameters.

%% ============================================================
%% SECTION 4: RNN ARCHITECTURE
%% ============================================================
\section{Recurrent Neural Network Architecture}
\label{sec:rnn-architecture}

\begin{definition}[Recurrence Relation]
\label{def:recurrence}
A \emph{recurrence relation} defines a hidden state $\mathbf{h}_t$
as a function of the current input $\mathbf{x}_t$ and the previous
hidden state $\mathbf{h}_{t-1}$:
\[
  \mathbf{h}_t = f(\mathbf{h}_{t-1}, \mathbf{x}_t; \boldsymbol{\theta}),
\]
where $f$ is a parameterized function and $\boldsymbol{\theta}$
represents the learnable parameters.
\end{definition}

\begin{definition}[Vanilla RNN]
\label{def:vanilla-rnn}
A \emph{vanilla recurrent neural network} (Elman network) with input
dimension $n$, hidden dimension $m$, and output dimension $k$ is
defined by the following equations.  At each time step $t = 1, \ldots, T$:
\begin{align}
  \mathbf{a}_t &= \mathbf{W}\mathbf{x}_t + \mathbf{U}\mathbf{h}_{t-1}
    + \mathbf{b}, \label{eq:pre-activation} \\
  \mathbf{h}_t &= \tanh(\mathbf{a}_t), \label{eq:hidden-state} \\
  \mathbf{o}_t &= \mathbf{V}\mathbf{h}_t + \mathbf{c},
    \label{eq:output-pre} \\
  \mathbf{y}_t &= g(\mathbf{o}_t), \label{eq:output}
\end{align}
where:
\begin{itemize}[nosep]
  \item $\mathbf{W} \in \R^{m \times n}$ is the input-to-hidden
    weight matrix,
  \item $\mathbf{U} \in \R^{m \times m}$ is the hidden-to-hidden
    (recurrent) weight matrix,
  \item $\mathbf{V} \in \R^{k \times m}$ is the hidden-to-output
    weight matrix,
  \item $\mathbf{b} \in \R^m$ and $\mathbf{c} \in \R^k$ are bias
    vectors,
  \item $g$ is an output activation function (identity for regression,
    softmax for classification),
  \item $\mathbf{h}_0 = \mathbf{0} \in \R^m$ (or learned).
\end{itemize}
\end{definition}

\begin{keyresult}[title={RNN Forward Equations}]
The complete RNN forward pass is:
\begin{align}
  \mathbf{h}_t &= \tanh(\mathbf{W}\mathbf{x}_t +
    \mathbf{U}\mathbf{h}_{t-1} + \mathbf{b}),
    \label{eq:rnn-hidden} \\
  \mathbf{y}_t &= g(\mathbf{V}\mathbf{h}_t + \mathbf{c}),
    \label{eq:rnn-output}
\end{align}
with initial condition $\mathbf{h}_0 = \mathbf{0} \in \R^m$.
\end{keyresult}

\begin{remark}[Parameter Sharing]
\label{rem:param-sharing}
The same parameters $(\mathbf{W}, \mathbf{U}, \mathbf{V},
\mathbf{b}, \mathbf{c})$ are used at every time step.  This provides:
(i)~the ability to process sequences of any length,
(ii)~a parameter count independent of sequence length, and
(iii)~statistical sharing of patterns across positions.
\end{remark}

\begin{proposition}[Parameter Count]
\label{prop:param-count}
For a vanilla RNN with input dimension $n$, hidden dimension $m$,
and output dimension $k$, the total number of trainable parameters is:
\[
  P = m(n + m + 1) + k(m + 1).
\]
\end{proposition}

\begin{proof}
The parameters are: $\mathbf{W} \in \R^{m \times n}$ ($mn$
entries), $\mathbf{U} \in \R^{m \times m}$ ($m^2$ entries),
$\mathbf{b} \in \R^m$ ($m$ entries), $\mathbf{V} \in \R^{k \times m}$
($km$ entries), $\mathbf{c} \in \R^k$ ($k$ entries).  Total:
$mn + m^2 + m + km + k = m(n + m + 1) + k(m + 1)$.
\end{proof}

\begin{definition}[Unfolded RNN]
\label{def:unfolded}
For a sequence of length $T$, the \emph{unfolded} RNN is the
composition:
\[
  \mathbf{h}_T = f(f(\cdots f(f(\mathbf{h}_0, \mathbf{x}_1),
    \mathbf{x}_2) \cdots, \mathbf{x}_{T-1}), \mathbf{x}_T),
\]
where $f(\mathbf{h}, \mathbf{x}) = \tanh(\mathbf{W}\mathbf{x} +
\mathbf{U}\mathbf{h} + \mathbf{b})$.  This can be viewed as a deep
feedforward network with $T$ layers, all sharing the same parameters.
\end{definition}

\begin{example}[Simple RNN Computation]
\label{ex:simple-rnn}
Consider a scalar RNN with $n = m = k = 1$, weights $W = 0.5$,
$U = 0.8$, $V = 1$, biases $b = c = 0$, and inputs
$x_1 = 1, x_2 = 0, x_3 = -1$.  Starting from $h_0 = 0$:
\begin{align*}
  h_1 &= \tanh(0.5 \cdot 1 + 0.8 \cdot 0) = \tanh(0.5) \approx 0.462, \\
  h_2 &= \tanh(0.5 \cdot 0 + 0.8 \cdot 0.462) = \tanh(0.370)
    \approx 0.354, \\
  h_3 &= \tanh(0.5 \cdot (-1) + 0.8 \cdot 0.354) = \tanh(-0.217)
    \approx -0.214.
\end{align*}
The hidden state retains a trace of past inputs: $h_3$ is influenced
by all three inputs despite receiving only $x_3$ directly.
\end{example}

The RNN architecture defines a causal sequence-to-sequence map.
We now define the loss function used to train it.

%% ============================================================
%% SECTION 5: LOSS FUNCTIONS
%% ============================================================
\section{Loss Functions for Sequences}
\label{sec:loss}

\begin{definition}[Sequence Loss]
\label{def:seq-loss}
For a sequence with target outputs $\hat{\mathbf{y}}_{1:T}$, the
\emph{total sequence loss} is:
\begin{equation}
  \mathcal{L}(\boldsymbol{\theta}) = \sum_{t=1}^{T}
    \mathcal{L}_t(\mathbf{y}_t, \hat{\mathbf{y}}_t),
  \label{eq:seq-loss}
\end{equation}
where $\mathcal{L}_t$ is the per-timestep loss function.
\end{definition}

\begin{definition}[Common Loss Functions]
\label{def:common-losses}
For regression tasks with continuous targets:
\[
  \mathcal{L}_t(\mathbf{y}_t, \hat{\mathbf{y}}_t)
    = \frac{1}{2} \|\mathbf{y}_t - \hat{\mathbf{y}}_t\|_2^2
    \quad \text{(mean squared error)}.
\]
For classification with softmax outputs and one-hot targets:
\[
  \mathcal{L}_t(\mathbf{y}_t, \hat{\mathbf{y}}_t)
    = -\sum_{i=1}^{k} \hat{y}_t^{(i)} \log y_t^{(i)}
    = -\log y_t^{(c_t)}
    \quad \text{(cross-entropy)},
\]
where $c_t$ is the true class index.
\end{definition}

\begin{proposition}[Softmax--Cross-Entropy Gradient]
\label{prop:softmax-ce}
When using softmax output with cross-entropy loss, the gradient of
$\mathcal{L}_t$ with respect to the pre-activation $\mathbf{o}_t$
simplifies to:
\begin{equation}
  \frac{\partial \mathcal{L}_t}{\partial \mathbf{o}_t}
    = \mathbf{y}_t - \hat{\mathbf{y}}_t.
  \label{eq:softmax-grad}
\end{equation}
\end{proposition}

\begin{proof}
For the $j$-th component with true class $c_t$:
\[
  \frac{\partial \mathcal{L}_t}{\partial o_t^{(j)}}
    = -\frac{\partial}{\partial o_t^{(j)}}
      \log \frac{e^{o_t^{(c_t)}}}{\sum_i e^{o_t^{(i)}}}
    = -\delta_{j c_t} + \frac{e^{o_t^{(j)}}}{\sum_i e^{o_t^{(i)}}}
    = y_t^{(j)} - \hat{y}_t^{(j)}.  \qedhere
\]
\end{proof}

With the forward pass and loss defined, we derive the algorithm
for computing gradients.

%% ============================================================
%% SECTION 6: BACKPROPAGATION THROUGH TIME
%% ============================================================
\section{Backpropagation Through Time}
\label{sec:bptt}

Backpropagation Through Time (BPTT) computes gradients of the
sequence loss $\mathcal{L} = \sum_{t=1}^T \mathcal{L}_t$ by
applying the chain rule to the unfolded computational graph.  The
key insight is that $\mathbf{h}_t$ influences the loss both directly
(through $\mathcal{L}_t$) and indirectly (through all future losses
$\mathcal{L}_{t+1}, \ldots, \mathcal{L}_T$ via the recurrence).

\begin{definition}[Output Error Signal]
\label{def:output-error}
The \emph{output error signal} at time $t$ is:
\begin{equation}
  \boldsymbol{\delta}_t^{(o)} = \frac{\partial \mathcal{L}_t}
    {\partial \mathbf{o}_t} \in \R^k.
  \label{eq:output-error}
\end{equation}
For softmax with cross-entropy:
$\boldsymbol{\delta}_t^{(o)} = \mathbf{y}_t - \hat{\mathbf{y}}_t$.
\end{definition}

\begin{definition}[Hidden State Error Signal]
\label{def:hidden-error}
The \emph{hidden state error signal} at time $t$ accumulates
contributions from all future losses:
\begin{equation}
  \boldsymbol{\delta}_t^{(h)} = \frac{\partial \mathcal{L}}
    {\partial \mathbf{h}_t} \in \R^m.
  \label{eq:hidden-error}
\end{equation}
\end{definition}

\begin{theorem}[Hidden State Error Recurrence]
\label{thm:hidden-error-recurrence}
The hidden state error signal satisfies the backward recurrence:
\begin{equation}
  \boldsymbol{\delta}_t^{(h)} = \mathbf{V}^\top
    \boldsymbol{\delta}_t^{(o)} + \mathbf{U}^\top
    \bigl(\boldsymbol{\delta}_{t+1}^{(h)} \odot
    (1 - \mathbf{h}_{t+1}^2)\bigr) \cdot \indicator_{t < T},
  \label{eq:hidden-recurrence}
\end{equation}
where $\odot$ denotes element-wise multiplication, $\mathbf{h}_{t+1}^2$
denotes element-wise squaring, and $(1 - \mathbf{h}_{t+1}^2)$ is the
derivative of $\tanh$ evaluated at $\mathbf{a}_{t+1}$.
\end{theorem}

\begin{proof}
The hidden state $\mathbf{h}_t$ influences the total loss through two
paths:
\begin{enumerate}[nosep]
  \item \emph{Direct path} through $\mathbf{y}_t$:
    $\mathbf{h}_t \to \mathbf{o}_t \to \mathbf{y}_t \to \mathcal{L}_t$.
  \item \emph{Indirect path} through $\mathbf{h}_{t+1}$ (when $t < T$):
    $\mathbf{h}_t \to \mathbf{a}_{t+1} \to \mathbf{h}_{t+1} \to
    \cdots \to \mathcal{L}$.
\end{enumerate}
For the direct path, since $\mathbf{o}_t = \mathbf{V}\mathbf{h}_t +
\mathbf{c}$:
\[
  \frac{\partial \mathcal{L}_t}{\partial \mathbf{h}_t}
    = \mathbf{V}^\top \frac{\partial \mathcal{L}_t}
      {\partial \mathbf{o}_t}
    = \mathbf{V}^\top \boldsymbol{\delta}_t^{(o)}.
\]
For the indirect path (when $t < T$), since
$\mathbf{a}_{t+1} = \mathbf{W}\mathbf{x}_{t+1} +
\mathbf{U}\mathbf{h}_t + \mathbf{b}$ and
$\mathbf{h}_{t+1} = \tanh(\mathbf{a}_{t+1})$:
\[
  \frac{\partial \mathcal{L}}{\partial \mathbf{h}_t}\bigg|_{\text{indirect}}
    = \frac{\partial \mathbf{a}_{t+1}^\top}{\partial \mathbf{h}_t}
      \frac{\partial \mathbf{h}_{t+1}^\top}{\partial \mathbf{a}_{t+1}}
      \frac{\partial \mathcal{L}}{\partial \mathbf{h}_{t+1}}
    = \mathbf{U}^\top \diag(1 - \mathbf{h}_{t+1}^2)
      \boldsymbol{\delta}_{t+1}^{(h)}.
\]
Combining both paths yields the recurrence.
\end{proof}

\begin{definition}[Pre-activation Error]
\label{def:pre-activation-error}
The error with respect to the pre-activation is:
\begin{equation}
  \boldsymbol{\delta}_t^{(a)} = \frac{\partial \mathcal{L}}
    {\partial \mathbf{a}_t}
    = \boldsymbol{\delta}_t^{(h)} \odot (1 - \mathbf{h}_t^2).
  \label{eq:pre-activation-error}
\end{equation}
\end{definition}

\begin{theorem}[BPTT Weight Gradients]
\label{thm:weight-gradients}
The gradients with respect to the weight matrices and bias vectors are:
\begin{align}
  \frac{\partial \mathcal{L}}{\partial \mathbf{V}}
    &= \sum_{t=1}^{T} \boldsymbol{\delta}_t^{(o)}
       \mathbf{h}_t^\top, \label{eq:grad-V} \\
  \frac{\partial \mathcal{L}}{\partial \mathbf{W}}
    &= \sum_{t=1}^{T} \boldsymbol{\delta}_t^{(a)}
       \mathbf{x}_t^\top, \label{eq:grad-W} \\
  \frac{\partial \mathcal{L}}{\partial \mathbf{U}}
    &= \sum_{t=1}^{T} \boldsymbol{\delta}_t^{(a)}
       \mathbf{h}_{t-1}^\top, \label{eq:grad-U} \\
  \frac{\partial \mathcal{L}}{\partial \mathbf{c}}
    &= \sum_{t=1}^{T} \boldsymbol{\delta}_t^{(o)},
       \label{eq:grad-c} \\
  \frac{\partial \mathcal{L}}{\partial \mathbf{b}}
    &= \sum_{t=1}^{T} \boldsymbol{\delta}_t^{(a)}.
       \label{eq:grad-b}
\end{align}
\end{theorem}

\begin{proof}
For $\mathbf{V}$: at each time $t$, $\mathbf{o}_t =
\mathbf{V}\mathbf{h}_t + \mathbf{c}$, so
$\frac{\partial \mathcal{L}_t}{\partial \mathbf{V}} =
\boldsymbol{\delta}_t^{(o)} \mathbf{h}_t^\top$.  Summing over $t$
gives~\eqref{eq:grad-V}.

For $\mathbf{W}$ and $\mathbf{U}$: since
$\mathbf{a}_t = \mathbf{W}\mathbf{x}_t +
\mathbf{U}\mathbf{h}_{t-1} + \mathbf{b}$, we have
$\frac{\partial \mathbf{a}_t}{\partial \mathbf{W}} = \mathbf{x}_t^\top$
and $\frac{\partial \mathbf{a}_t}{\partial \mathbf{U}} =
\mathbf{h}_{t-1}^\top$.  The total gradient accumulates over all
time steps:
\[
  \frac{\partial \mathcal{L}}{\partial \mathbf{W}}
    = \sum_{t=1}^{T} \frac{\partial \mathcal{L}}
      {\partial \mathbf{a}_t}
      \frac{\partial \mathbf{a}_t}{\partial \mathbf{W}}
    = \sum_{t=1}^{T} \boldsymbol{\delta}_t^{(a)} \mathbf{x}_t^\top.
\]
The derivation for $\mathbf{U}$, $\mathbf{b}$, and $\mathbf{c}$
is analogous.
\end{proof}

\begin{keyresult}[title={Complete BPTT Algorithm}]
\textbf{Forward Pass:}
\begin{enumerate}[nosep]
  \item Initialize $\mathbf{h}_0 = \mathbf{0}$.
  \item For $t = 1$ to $T$: compute $\mathbf{a}_t$,
    $\mathbf{h}_t$, $\mathbf{o}_t$, $\mathbf{y}_t$, and
    $\mathcal{L}_t$.
\end{enumerate}
\textbf{Backward Pass:}
\begin{enumerate}[nosep]
  \item Initialize $\boldsymbol{\delta}_{T+1}^{(h)} = \mathbf{0}$.
  \item For $t = T$ down to $1$:
    \begin{itemize}[nosep]
      \item $\boldsymbol{\delta}_t^{(o)} = \mathbf{y}_t -
        \hat{\mathbf{y}}_t$
      \item $\boldsymbol{\delta}_t^{(h)} = \mathbf{V}^\top
        \boldsymbol{\delta}_t^{(o)} + \mathbf{U}^\top
        (\boldsymbol{\delta}_{t+1}^{(h)} \odot
        (1 - \mathbf{h}_{t+1}^2))$
      \item $\boldsymbol{\delta}_t^{(a)} =
        \boldsymbol{\delta}_t^{(h)} \odot (1 - \mathbf{h}_t^2)$
      \item Accumulate weight gradients.
    \end{itemize}
\end{enumerate}
\end{keyresult}

\begin{example}[BPTT for a Two-Step Sequence]
\label{ex:bptt-two-step}
Consider $T = 2$ with $m = 2$, $k = 1$ (regression, so $g$ is
identity).  The total loss is $\mathcal{L} = \mathcal{L}_1 +
\mathcal{L}_2$.  The backward pass computes:
\begin{align*}
  \boldsymbol{\delta}_2^{(h)} &= \mathbf{V}^\top
    \boldsymbol{\delta}_2^{(o)}, \\
  \boldsymbol{\delta}_1^{(h)} &= \mathbf{V}^\top
    \boldsymbol{\delta}_1^{(o)} + \mathbf{U}^\top
    (\boldsymbol{\delta}_2^{(h)} \odot (1 - \mathbf{h}_2^2)).
\end{align*}
Note that $\boldsymbol{\delta}_1^{(h)}$ receives gradient information
from \emph{both} $\mathcal{L}_1$ and $\mathcal{L}_2$, reflecting
the influence of $\mathbf{h}_1$ on the entire future.
\end{example}

The BPTT algorithm reveals a critical structural property: the
backward pass involves repeated multiplication by $\mathbf{U}^\top$
scaled by the tanh derivative.  We now analyze the consequences.

%% ============================================================
%% SECTION 7: VANISHING AND EXPLODING GRADIENTS
%% ============================================================
\section{The Vanishing and Exploding Gradient Problem}
\label{sec:gradient-problem}

\begin{theorem}[Long-Range Gradient]
\label{thm:long-range}
The gradient of the loss at time $t$ with respect to the hidden
state at an earlier time $\tau < t$ is:
\begin{equation}
  \frac{\partial \mathcal{L}_t}{\partial \mathbf{h}_\tau}
    = \frac{\partial \mathcal{L}_t}{\partial \mathbf{h}_t}
      \prod_{i=\tau+1}^{t} \frac{\partial \mathbf{h}_i}
        {\partial \mathbf{h}_{i-1}}.
  \label{eq:long-range}
\end{equation}
\end{theorem}

\begin{proof}
By repeated application of the chain rule along the path
$\mathbf{h}_\tau \to \mathbf{h}_{\tau+1} \to \cdots \to
\mathbf{h}_t \to \mathcal{L}_t$:
\[
  \frac{\partial \mathcal{L}_t}{\partial \mathbf{h}_\tau}
    = \frac{\partial \mathcal{L}_t}{\partial \mathbf{h}_t}
      \cdot \frac{\partial \mathbf{h}_t}{\partial \mathbf{h}_{t-1}}
      \cdots \frac{\partial \mathbf{h}_{\tau+1}}
        {\partial \mathbf{h}_\tau}
    = \frac{\partial \mathcal{L}_t}{\partial \mathbf{h}_t}
      \prod_{i=\tau+1}^{t} \frac{\partial \mathbf{h}_i}
        {\partial \mathbf{h}_{i-1}}.  \qedhere
\]
\end{proof}

\begin{lemma}[Jacobian of Hidden State Transition]
\label{lem:jacobian}
The Jacobian of the hidden state transition is:
\begin{equation}
  \frac{\partial \mathbf{h}_t}{\partial \mathbf{h}_{t-1}}
    = \diag(1 - \mathbf{h}_t^2) \, \mathbf{U},
  \label{eq:jacobian}
\end{equation}
where $\diag(1 - \mathbf{h}_t^2)$ is the diagonal matrix with entries
$(1 - h_t^{(j)2})$.
\end{lemma}

\begin{proof}
Since $\mathbf{h}_t = \tanh(\mathbf{a}_t)$ and
$\mathbf{a}_t = \mathbf{W}\mathbf{x}_t + \mathbf{U}\mathbf{h}_{t-1}
+ \mathbf{b}$:
\[
  \frac{\partial \mathbf{h}_t}{\partial \mathbf{h}_{t-1}}
    = \frac{\partial \mathbf{h}_t}{\partial \mathbf{a}_t}
      \cdot \frac{\partial \mathbf{a}_t}{\partial \mathbf{h}_{t-1}}
    = \diag(1 - \mathbf{h}_t^2) \, \mathbf{U}.  \qedhere
\]
\end{proof}

\begin{theorem}[Vanishing Gradient]
\label{thm:vanishing}
Let $\gamma = \|\mathbf{U}\|$ be the spectral norm of $\mathbf{U}$.
If $\gamma < 1$, then:
\begin{equation}
  \left\|\frac{\partial \mathcal{L}_t}{\partial \mathbf{h}_\tau}
  \right\| \leq \left\|\frac{\partial \mathcal{L}_t}
    {\partial \mathbf{h}_t}\right\| \gamma^{t - \tau}.
  \label{eq:vanishing}
\end{equation}
Thus, gradients decay \textbf{exponentially} with the time gap
$(t - \tau)$.
\end{theorem}

\begin{proof}
Using the submultiplicativity of matrix norms and
Theorem~\ref{thm:long-range}:
\[
  \left\|\frac{\partial \mathcal{L}_t}{\partial \mathbf{h}_\tau}
  \right\| = \left\|\frac{\partial \mathcal{L}_t}
    {\partial \mathbf{h}_t}
    \prod_{i=\tau+1}^{t} \diag(1 - \mathbf{h}_i^2)\,
    \mathbf{U}\right\|
  \leq \left\|\frac{\partial \mathcal{L}_t}
    {\partial \mathbf{h}_t}\right\|
    \prod_{i=\tau+1}^{t} \|\diag(1 - \mathbf{h}_i^2)\|
    \cdot \|\mathbf{U}\|.
\]
Since $\tanh$ maps to $(-1, 1)$, we have
$|1 - h_i^{(j)2}| \leq 1$ for all components, so
$\|\diag(1 - \mathbf{h}_i^2)\| \leq 1$.  Therefore:
\[
  \left\|\frac{\partial \mathcal{L}_t}{\partial \mathbf{h}_\tau}
  \right\| \leq \left\|\frac{\partial \mathcal{L}_t}
    {\partial \mathbf{h}_t}\right\| \|\mathbf{U}\|^{t-\tau}
    = \left\|\frac{\partial \mathcal{L}_t}
    {\partial \mathbf{h}_t}\right\| \gamma^{t-\tau}.  \qedhere
\]
\end{proof}

\begin{keyresult}[title={Vanishing Gradient Bound}]
When $\gamma = \|\mathbf{U}\| < 1$, the gradient
$\|\partial \mathcal{L}_t / \partial \mathbf{h}_\tau\|$ decays at
rate $\gamma^{t-\tau}$.  For $\gamma = 0.9$ and a time gap of
$k = t - \tau$ steps:
\begin{center}
\begin{tabular}{ccccc}
\toprule
$k$ & 10 & 50 & 100 & 200 \\
\midrule
$\gamma^k$ & 0.35 & $5 \times 10^{-3}$ & $2.7 \times 10^{-5}$
  & $7 \times 10^{-10}$ \\
\bottomrule
\end{tabular}
\end{center}
Gradients become negligible after a few dozen time steps.
\end{keyresult}

\begin{theorem}[Exploding Gradient]
\label{thm:exploding}
If $\gamma = \|\mathbf{U}\| > 1$ and the Jacobian products do not
shrink sufficiently (i.e., the activation derivatives remain bounded
away from zero), then gradients can grow exponentially:
\begin{equation}
  \left\|\frac{\partial \mathcal{L}_t}{\partial \mathbf{h}_\tau}
  \right\| \geq C \cdot \gamma_{\min}^{t - \tau},
  \label{eq:exploding}
\end{equation}
for some constant $C > 0$, where $\gamma_{\min} > 0$ depends on the
minimum activation derivative along the trajectory.
\end{theorem}

\begin{proof}
If $\sigma_{\min}(\mathbf{U}) > 0$ denotes the smallest singular
value of $\mathbf{U}$ and the diagonal entries of
$\diag(1 - \mathbf{h}_i^2)$ are bounded below by some
$\alpha > 0$ (the network operates in the linear regime of $\tanh$),
then:
\[
  \left\|\prod_{i=\tau+1}^{t} \diag(1 - \mathbf{h}_i^2)\,
    \mathbf{U}\right\|
  \geq \prod_{i=\tau+1}^{t} \alpha \cdot \sigma_{\min}(\mathbf{U})
  = (\alpha \, \sigma_{\min}(\mathbf{U}))^{t-\tau}.
\]
When $\alpha \, \sigma_{\min}(\mathbf{U}) > 1$, the product grows
exponentially.  In practice, even $\|\mathbf{U}\| > 1$ often
suffices since the activation derivatives are not uniformly small.
\end{proof}

\begin{warning}[title={The Gradient Dilemma}]
The vanishing and exploding gradient problems create a fundamental
dilemma for vanilla RNNs:
\begin{itemize}[nosep]
  \item If $\|\mathbf{U}\| < 1$: long-range dependencies cannot be
    learned (gradients vanish).
  \item If $\|\mathbf{U}\| > 1$: training becomes unstable (gradients
    explode).
  \item If $\|\mathbf{U}\| = 1$: marginally stable but practically
    unachievable with fixed weights.
\end{itemize}
This means vanilla RNNs cannot effectively learn dependencies
spanning more than approximately 10--20 time steps.
\end{warning}

\begin{proposition}[Eigenvalue Characterization]
\label{prop:eigenvalue}
Let $\lambda_1, \ldots, \lambda_m$ be the eigenvalues of
$\mathbf{U}$.  The gradient behavior is characterized by the
spectral radius $\rho(\mathbf{U}) = \max_j |\lambda_j|$:
\begin{itemize}[nosep]
  \item If $|\lambda_{\max}| < 1$: vanishing gradients dominate.
  \item If $|\lambda_{\max}| > 1$: exploding gradient risk.
  \item If $|\lambda_{\max}| = 1$: marginally stable (ideal but
    fragile).
\end{itemize}
\end{proposition}

\begin{example}[Vanishing Gradient Demonstration]
\label{ex:vanishing-demo}
Consider $m = 2$ with
$\mathbf{U} = \begin{pmatrix} 0.5 & 0.3 \\ 0.2 & 0.4 \end{pmatrix}$.
The eigenvalues are $\lambda_1 \approx 0.72$ and
$\lambda_2 \approx 0.18$, so $\|\mathbf{U}\| < 1$.  After $k = 20$
steps, the gradient is attenuated by at most $0.72^{20} \approx
0.0013$---three orders of magnitude smaller.  The network effectively
cannot learn dependencies beyond $\sim$20 steps.
\end{example}

\begin{rlconnection}[title={Gradient Pathology in Deep RL}]
In recurrent policies for POMDPs, the vanishing gradient problem
limits the effective memory horizon.  An LSTM-based PPO agent can
maintain information over hundreds of steps (Lesson~68), while a
vanilla RNN policy is limited to $\sim$10--20 steps.  This is why
virtually all modern recurrent RL architectures use gated units.
\end{rlconnection}

Having identified the gradient pathology, we now discuss strategies
to mitigate it.

%% ============================================================
%% SECTION 8: MITIGATION STRATEGIES
%% ============================================================
\section{Mitigation Strategies}
\label{sec:mitigation}

\begin{definition}[Gradient Clipping]
\label{def:grad-clip}
\emph{Gradient clipping} rescales the gradient when its norm exceeds
a threshold $\theta > 0$:
\begin{equation}
  \bar{\mathbf{g}} = \begin{cases}
    \mathbf{g} & \text{if } \|\mathbf{g}\| \leq \theta, \\
    \frac{\theta}{\|\mathbf{g}\|} \mathbf{g}
      & \text{if } \|\mathbf{g}\| > \theta.
  \end{cases}
  \label{eq:grad-clip}
\end{equation}
\end{definition}

\begin{remark}[Clipping Properties]
Gradient clipping preserves the gradient direction while bounding its
magnitude.  It addresses the exploding gradient problem but does
\emph{not} solve the vanishing gradient problem.  The threshold
$\theta$ is a hyperparameter; \citet{pascanu2013} recommend values in
the range $[1, 10]$.
\end{remark}

\begin{proposition}[Orthogonal Initialization]
\label{prop:orthogonal}
If $\mathbf{U}$ is initialized as an orthogonal matrix
($\mathbf{U}^\top \mathbf{U} = \mathbf{I}$), then
$\|\mathbf{U}\| = 1$ and all singular values equal $1$.  This
provides ideal gradient flow at initialization:
$\|\diag(1 - \mathbf{h}_i^2) \, \mathbf{U}\| \leq 1$ with
equality when the network operates in the linear regime.
\end{proposition}

\begin{proof}
For an orthogonal matrix, all singular values are $1$, so
$\|\mathbf{U}\| = 1$.  The Jacobian product satisfies:
\[
  \left\|\prod_{i=\tau+1}^{t} \diag(1 - \mathbf{h}_i^2)\,
    \mathbf{U}\right\| \leq \prod_{i=\tau+1}^{t}
    \|\diag(1 - \mathbf{h}_i^2)\| \cdot \|\mathbf{U}\|
    \leq 1.
\]
When the pre-activations are near zero (as at initialization with
small inputs), $1 - h_i^{(j)2} \approx 1$, yielding
$\|\diag(1 - \mathbf{h}_i^2) \, \mathbf{U}\| \approx 1$.
\end{proof}

\begin{warning}[title={Orthogonality is Not Preserved}]
Although orthogonal initialization provides ideal gradient flow at
the start of training, gradient updates break the orthogonality
of $\mathbf{U}$.  After several training steps, the singular values
of $\mathbf{U}$ drift away from $1$, and the vanishing/exploding
gradient problem resurfaces.
\end{warning}

\begin{definition}[Truncated BPTT]
\label{def:truncated-bptt}
\emph{Truncated BPTT} with window size $k$ limits backpropagation to
the most recent $k$ time steps:
\begin{equation}
  \frac{\partial \mathcal{L}_t}{\partial \mathbf{h}_\tau}
    \approx \begin{cases}
      \text{computed} & \text{if } t - \tau \leq k, \\
      0 & \text{if } t - \tau > k.
    \end{cases}
  \label{eq:truncated-bptt}
\end{equation}
\end{definition}

\begin{remark}[Truncated BPTT Trade-off]
Truncated BPTT introduces bias by ignoring long-range dependencies
but reduces computational cost from $O(T)$ to $O(k)$ per step.  It
is a practical compromise when full BPTT is too expensive.  The
window size $k$ controls the bias--computation trade-off: larger $k$
captures longer dependencies but costs more.
\end{remark}

\begin{example}[Comparing Mitigation Strategies]
\label{ex:mitigation-compare}
Consider training a vanilla RNN on sequences of length $T = 100$:
\begin{itemize}[nosep]
  \item \emph{No mitigation}: gradients vanish for $t - \tau > 20$;
    the network can only learn short-range patterns.
  \item \emph{Gradient clipping} ($\theta = 5$): prevents numerical
    overflow from exploding gradients but does not help with vanishing
    gradients.
  \item \emph{Truncated BPTT} ($k = 20$): reduces computation but
    explicitly limits the memory horizon.
  \item \emph{Orthogonal initialization}: helps at the start of
    training but degrades as $\mathbf{U}$ drifts.
\end{itemize}
The fundamental solution requires redesigning the architecture, which
motivates gated RNNs (Lessons~68--69).
\end{example}

%% ============================================================
%% SECTION 9: EXERCISES
%% ============================================================
\section{Exercises}
\label{sec:exercises}

\begin{exercise}[RNN Forward Pass]
\textnormal{(\(\ast\))} \\
Consider a vanilla RNN with $n = 2$, $m = 3$, $k = 1$, and
\[
  \mathbf{W} = \begin{pmatrix} 1 & 0 \\ 0 & 1 \\ 1 & 1
    \end{pmatrix}, \quad
  \mathbf{U} = \begin{pmatrix} 0.5 & 0 & 0 \\ 0 & 0.5 & 0 \\
    0 & 0 & 0.5 \end{pmatrix}, \quad
  \mathbf{V} = \begin{pmatrix} 1 & 1 & 1 \end{pmatrix},
\]
with zero biases and $\mathbf{h}_0 = \mathbf{0}$.  Compute the
hidden states $\mathbf{h}_1, \mathbf{h}_2$ and outputs
$y_1, y_2$ for the input sequence
$\mathbf{x}_1 = (1, 0)^\top$, $\mathbf{x}_2 = (0, 1)^\top$.
\end{exercise}

\begin{exercise}[BPTT Derivation for Scalar RNN]
\textnormal{(\(\ast\ast\))} \\
For a scalar RNN ($n = m = k = 1$) with loss
$\mathcal{L} = \frac{1}{2}\sum_{t=1}^T (y_t - \hat{y}_t)^2$ and
$g = \text{identity}$, derive explicit expressions for
$\frac{\partial \mathcal{L}}{\partial W}$,
$\frac{\partial \mathcal{L}}{\partial U}$, and
$\frac{\partial \mathcal{L}}{\partial V}$ for a sequence of length
$T = 3$.  Verify that the gradient with respect to $U$ involves
contributions from all time steps.
\end{exercise}

\begin{exercise}[Gradient Norm Bound]
\textnormal{(\(\ast\ast\))} \\
Prove that for the vanilla RNN with $\tanh$ activation, the gradient
satisfies:
\[
  \left\|\frac{\partial \mathcal{L}_t}{\partial \mathbf{h}_\tau}
  \right\| \leq \left\|\frac{\partial \mathcal{L}_t}
    {\partial \mathbf{h}_t}\right\| \cdot
    \left(\max_j |1 - h_j^2| \cdot \|\mathbf{U}\|\right)^{t-\tau}.
\]
Under what conditions on the hidden state values does this become
a tighter bound than the one in Theorem~\ref{thm:vanishing}?

\textit{Hint:} Consider what happens when the hidden state components
are large in magnitude (near saturation).
\end{exercise}

\begin{exercise}[ReLU Activation in RNNs]
\textnormal{(\(\ast\ast\))} \\
Suppose we replace $\tanh$ with $\text{ReLU}(z) = \max(0, z)$ in the
vanilla RNN.  Derive the Jacobian
$\partial \mathbf{h}_t / \partial \mathbf{h}_{t-1}$ in this case.
Explain why the vanishing gradient problem persists (or worsens) and
why the exploding gradient problem becomes more severe.
\end{exercise}

\begin{exercise}[Optimal Memory Horizon]
\textnormal{(\(\ast\ast\ast\))} \\
Consider a vanilla RNN trained on a task where the target at time $T$
depends on the input at time $1$ (a ``copy'' task with delay $T-1$).
Using the vanishing gradient bound (Theorem~\ref{thm:vanishing}),
derive an upper bound on the sequence length $T$ for which the
gradient signal from time $1$ remains above a threshold
$\epsilon > 0$.  Express your answer in terms of $\gamma$ and
$\epsilon$.  For $\gamma = 0.95$ and
$\epsilon = 10^{-6}$, what is the maximum effective delay?

\textit{Hint:} Solve $\gamma^{T-1} \geq \epsilon$ for $T$.
\end{exercise}

%% ============================================================
%% SECTION 10: SUMMARY
%% ============================================================
\section{Summary}
\label{sec:summary}

\begin{itemize}
  \item A recurrent neural network maintains a hidden state
    $\mathbf{h}_t = \tanh(\mathbf{W}\mathbf{x}_t +
    \mathbf{U}\mathbf{h}_{t-1} + \mathbf{b})$ that compresses the
    entire input history into a fixed-dimensional vector.
  \item Parameter sharing across time steps allows RNNs to process
    sequences of arbitrary length with $O(m(n + m + 1) + k(m+1))$
    parameters, independent of sequence length.
  \item BPTT computes gradients by unfolding the recurrence into a
    deep feedforward graph and propagating errors backward through
    time, yielding weight gradients that accumulate over all time
    steps.
  \item The long-range gradient involves a product of $(t - \tau)$
    Jacobian matrices $\diag(1 - \mathbf{h}_i^2)\mathbf{U}$,
    leading to exponential decay or growth.
  \item When $\|\mathbf{U}\| < 1$, gradients vanish at rate
    $\gamma^{t-\tau}$, preventing the network from learning
    dependencies beyond $\sim$10--20 time steps.
  \item When $\|\mathbf{U}\| > 1$, gradients can explode, causing
    training instability and numerical overflow.
  \item Gradient clipping addresses exploding gradients by rescaling
    the gradient vector, while orthogonal initialization and
    truncated BPTT provide partial remedies.
  \item The fundamental solution to the vanishing gradient problem
    requires architectural innovation: gated recurrent networks
    (LSTM, GRU) create additive paths for gradient flow.
\end{itemize}

%% ============================================================
%% SECTION 11: FURTHER READING
%% ============================================================
\section{Further Reading}
\label{sec:further-reading}

\begin{itemize}
  \item Lesson~68 introduces the LSTM architecture, which solves the
    vanishing gradient problem through an additive cell state update
    and learnable gating mechanisms.
  \item Lesson~69 covers the GRU (a simplified gated architecture),
    bidirectional RNNs, deep stacked RNNs, and the encoder-decoder
    framework for sequence-to-sequence tasks.
  \item \citet{pascanu2013} provides a detailed analysis of gradient
    flow in RNNs, including second-order methods and the connection
    between gradient clipping and trust regions.
  \item \citet{bengio1994} gives the original theoretical analysis
    of why learning long-term dependencies is fundamentally difficult,
    including connections to dynamical systems theory.
  \item \citet[Chapter~10]{goodfellow2016} offers a comprehensive
    survey of recurrent architectures, training techniques, and
    practical considerations.
  \item \citet{graves2012} covers supervised sequence labelling with
    RNNs, including connectionist temporal classification (CTC) for
    speech and handwriting recognition.
\end{itemize}

\bibliography{nn-refs}
\end{document}
