\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 69: GRU and Sequence Architectures}}
\fancyhead[R]{\thepage}
\title{Lesson 69: Gated Recurrent Unit\\and Sequence Architectures}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% ============================================================
%% SECTION 1: MOTIVATION AND OVERVIEW
%% ============================================================
\section{Motivation and Overview}
\label{sec:motivation}

Lesson~67 established the vanilla RNN and exposed its fundamental
limitation: vanishing gradients prevent learning long-range
dependencies.  Lesson~68 showed how the LSTM resolves this through
an additive cell state update with learnable gates, at the cost of
$4\times$ more parameters.  A natural question arises: \emph{can we
achieve comparable gradient flow with a simpler gating mechanism?}

The Gated Recurrent Unit (GRU), introduced by \citet{cho2014},
answers affirmatively.  The GRU merges the cell state and hidden
state into a single vector and uses only two gates (reset and
update) instead of three.  Empirical comparisons \citep{chung2014}
show that the GRU achieves performance comparable to the LSTM on
many tasks while requiring fewer parameters and less computation.

Beyond the choice of recurrent cell, the power of recurrent
architectures is amplified by structural innovations:
\emph{bidirectional} processing captures both past and future
context, \emph{deep stacking} increases representational capacity,
and the \emph{encoder-decoder} framework enables variable-length
sequence-to-sequence mappings.  This lesson develops all four
extensions and concludes with a capstone summary of the entire
Recurrent Neural Networks subdomain.

\begin{rlconnection}[title={Why GRU and Sequence Architectures
    Matter for RL/ML}]
\begin{itemize}[nosep]
  \item \textbf{Lightweight recurrent policies.}  GRU-based policies
    offer a favorable trade-off between memory capacity and
    computational cost for on-policy RL algorithms like PPO that
    require many forward passes per training step.
  \item \textbf{Bidirectional context in offline RL.}  When
    processing pre-collected trajectories (offline RL, imitation
    learning), bidirectional RNNs can leverage both past and future
    context to improve state estimation.
  \item \textbf{Deep recurrent architectures.}  Stacked RNNs
    increase the capacity of recurrent value functions and policies,
    enabling richer representations in complex environments.
  \item \textbf{Sequence-to-sequence in RL.}  The encoder-decoder
    framework underlies trajectory transformers, goal-conditioned
    policies, and communication protocols in multi-agent RL.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
LSTM cell architecture and gating mechanisms (Lesson~68,
Definitions~4.1--4.7 and Theorem~7.1).  Vanishing gradient problem
and the cell state gradient highway (Lesson~68, Corollary~7.2 and
Theorem~7.3).  BPTT algorithm for vanilla RNNs (Lesson~67,
Section~6).

\paragraph{Learning objectives.}
After completing this lesson, you will be able to:
\begin{itemize}[nosep]
  \item \textbf{Define} the GRU architecture and \textbf{derive}
    its forward equations, including the reset and update gates.
  \item \textbf{Compare} the GRU and LSTM in terms of parameter
    count, gradient flow properties, and computational efficiency.
  \item \textbf{Derive} the BPTT equations for the GRU, including
    the hidden state gradient recurrence.
  \item \textbf{Define} bidirectional RNNs and \textbf{explain}
    when bidirectional processing is appropriate.
  \item \textbf{Define} deep (stacked) RNNs and \textbf{compute}
    their parameter counts.
  \item \textbf{Formalize} the encoder-decoder framework and
    \textbf{explain} the context vector bottleneck.
  \item \textbf{Synthesize} the subdomain arc from vanilla RNNs
    through gated architectures to sequence-level designs.
\end{itemize}

%% ============================================================
%% SECTION 2: REFERENCES
%% ============================================================
\section{References}
\label{sec:references}

\begin{itemize}
  \item \citet{cho2014} --- Original GRU paper: the encoder-decoder
    architecture with GRU cells for statistical machine translation.
  \item \citet{chung2014} --- Empirical comparison of GRU and LSTM
    on sequence modeling tasks (polyphonic music, speech).
  \item \citet{schuster1997} --- Bidirectional recurrent neural
    networks: architecture and training algorithm.
  \item \citet{sutskever2014} --- Sequence-to-sequence learning with
    neural networks: the encoder-decoder framework with LSTMs.
  \item \citet{bahdanau2015} --- Attention mechanism for
    encoder-decoder models, addressing the context vector bottleneck.
  \item \citet[Chapter~10]{goodfellow2016} --- Comprehensive
    treatment of recurrent architectures and sequence modeling.
\end{itemize}

%% ============================================================
%% SECTION 3: GATED RECURRENT UNIT
%% ============================================================
\section{Gated Recurrent Unit}
\label{sec:gru}

\subsection{GRU Architecture}

The GRU simplifies the LSTM by merging the cell state and hidden
state into a single vector $\mathbf{h}_t$ and using two gates
instead of three.

\begin{definition}[Reset Gate]
\label{def:reset-gate}
The \emph{reset gate} $\mathbf{r}_t \in (0, 1)^m$ controls how much
of the previous hidden state to ``forget'' when computing the
candidate:
\begin{equation}
  \boxed{\mathbf{r}_t = \sigma(\mathbf{W}_r \mathbf{x}_t +
    \mathbf{U}_r \mathbf{h}_{t-1} + \mathbf{b}_r)},
  \label{eq:reset-gate}
\end{equation}
where $\mathbf{W}_r \in \R^{m \times n}$,
$\mathbf{U}_r \in \R^{m \times m}$, and $\mathbf{b}_r \in \R^m$.
\end{definition}

\begin{definition}[Update Gate]
\label{def:update-gate}
The \emph{update gate} $\mathbf{z}_t \in (0, 1)^m$ controls the
balance between the previous hidden state and the candidate:
\begin{equation}
  \boxed{\mathbf{z}_t = \sigma(\mathbf{W}_z \mathbf{x}_t +
    \mathbf{U}_z \mathbf{h}_{t-1} + \mathbf{b}_z)},
  \label{eq:update-gate}
\end{equation}
where $\mathbf{W}_z \in \R^{m \times n}$,
$\mathbf{U}_z \in \R^{m \times m}$, and $\mathbf{b}_z \in \R^m$.
\end{definition}

\begin{definition}[GRU Candidate Hidden State]
\label{def:gru-candidate}
The \emph{candidate hidden state} $\tilde{\mathbf{h}}_t \in
(-1, 1)^m$ is computed using the reset gate to selectively
incorporate the previous hidden state:
\begin{equation}
  \boxed{\tilde{\mathbf{h}}_t = \tanh(\mathbf{W}_h \mathbf{x}_t +
    \mathbf{U}_h (\mathbf{r}_t \odot \mathbf{h}_{t-1}) +
    \mathbf{b}_h)},
  \label{eq:gru-candidate}
\end{equation}
where $\mathbf{W}_h \in \R^{m \times n}$,
$\mathbf{U}_h \in \R^{m \times m}$, and $\mathbf{b}_h \in \R^m$.
\end{definition}

\begin{definition}[GRU Hidden State Update]
\label{def:gru-update}
The hidden state is a convex combination of the previous hidden
state and the candidate:
\begin{equation}
  \boxed{\mathbf{h}_t = (1 - \mathbf{z}_t) \odot \mathbf{h}_{t-1}
    + \mathbf{z}_t \odot \tilde{\mathbf{h}}_t}.
  \label{eq:gru-hidden}
\end{equation}
\end{definition}

\begin{keyresult}[title={GRU Forward Equations}]
At each time step $t = 1, \ldots, T$:
\begin{align}
  \mathbf{r}_t &= \sigma(\mathbf{W}_r \mathbf{x}_t +
    \mathbf{U}_r \mathbf{h}_{t-1} + \mathbf{b}_r),
    && \text{[Reset Gate]} \label{eq:gru-r} \\
  \mathbf{z}_t &= \sigma(\mathbf{W}_z \mathbf{x}_t +
    \mathbf{U}_z \mathbf{h}_{t-1} + \mathbf{b}_z),
    && \text{[Update Gate]} \label{eq:gru-z} \\
  \tilde{\mathbf{h}}_t &= \tanh(\mathbf{W}_h \mathbf{x}_t +
    \mathbf{U}_h (\mathbf{r}_t \odot \mathbf{h}_{t-1}) +
    \mathbf{b}_h),
    && \text{[Candidate]} \label{eq:gru-h-tilde} \\
  \mathbf{h}_t &= (1 - \mathbf{z}_t) \odot \mathbf{h}_{t-1} +
    \mathbf{z}_t \odot \tilde{\mathbf{h}}_t.
    && \text{[Hidden State]} \label{eq:gru-h}
\end{align}
\end{keyresult}

\begin{remark}[GRU vs LSTM: Structural Comparison]
\label{rem:gru-vs-lstm}
The GRU and LSTM share the principle of gated information flow but
differ in key ways:
\begin{itemize}[nosep]
  \item The GRU has no separate cell state; $\mathbf{h}_t$ serves
    as both memory and output.
  \item The update gate $\mathbf{z}_t$ plays the role of both the
    LSTM's forget gate and input gate (coupled as
    $\mathbf{i} = 1 - \mathbf{f}$).
  \item The reset gate $\mathbf{r}_t$ has no direct LSTM analogue;
    it controls how much of $\mathbf{h}_{t-1}$ contributes to the
    candidate computation.
  \item The GRU hidden state update is a convex combination, so
    $\|\mathbf{h}_t\| \leq \max(\|\mathbf{h}_{t-1}\|,
    \|\tilde{\mathbf{h}}_t\|)$, providing implicit norm control.
\end{itemize}
\end{remark}

\begin{proposition}[GRU Parameter Count]
\label{prop:gru-params}
For a GRU with input dimension $n$ and hidden dimension $m$, the
total number of trainable parameters (excluding the output layer) is:
\begin{equation}
  P = 3(mn + m^2 + m) = 3m(n + m + 1).
  \label{eq:gru-params}
\end{equation}
\end{proposition}

\begin{proof}
Each of the three components (reset gate, update gate, candidate)
has $mn + m^2 + m$ parameters.  Total:
$3(mn + m^2 + m) = 3m(n + m + 1)$.
\end{proof}

\begin{example}[Parameter Comparison]
\label{ex:param-comparison}
For $n = 300$, $m = 512$ (word embedding to hidden state):
\begin{center}
\begin{tabular}{lcc}
\toprule
Architecture & Formula & Parameters \\
\midrule
Vanilla RNN & $m(n + m + 1)$ & 416,256 \\
GRU & $3m(n + m + 1)$ & 1,248,768 \\
LSTM & $4m(n + m + 1)$ & 1,664,256 \\
\bottomrule
\end{tabular}
\end{center}
The GRU has $75\%$ of the LSTM's parameters while often achieving
comparable performance.
\end{example}

\subsection{GRU Gradient Flow}

\begin{theorem}[GRU Hidden State Jacobian]
\label{thm:gru-jacobian}
The Jacobian of $\mathbf{h}_t$ with respect to $\mathbf{h}_{t-1}$
in the GRU is:
\begin{equation}
  \frac{\partial \mathbf{h}_t}{\partial \mathbf{h}_{t-1}}
    = \diag(1 - \mathbf{z}_t) + \mathbf{A}_t + \mathbf{B}_t
      + \mathbf{C}_t,
  \label{eq:gru-jacobian}
\end{equation}
where:
\begin{align*}
  \mathbf{A}_t &= \diag(\mathbf{z}_t \odot
    (1 - \tilde{\mathbf{h}}_t^2)) \, \mathbf{U}_h \,
    \diag(\mathbf{r}_t),
    && \text{(candidate through reset)} \\
  \mathbf{B}_t &= \diag(\tilde{\mathbf{h}}_t - \mathbf{h}_{t-1})
    \, \diag(\mathbf{z}_t \odot (1 - \mathbf{z}_t)) \,
    \mathbf{U}_z,
    && \text{(update gate)} \\
  \mathbf{C}_t &= \diag(\mathbf{z}_t \odot
    (1 - \tilde{\mathbf{h}}_t^2)) \, \mathbf{U}_h \,
    \diag(\mathbf{h}_{t-1} \odot \mathbf{r}_t \odot
    (1 - \mathbf{r}_t)) \, \mathbf{U}_r.
    && \text{(reset gate)}
\end{align*}
\end{theorem}

\begin{proof}
The hidden state update is $\mathbf{h}_t = (1 - \mathbf{z}_t) \odot
\mathbf{h}_{t-1} + \mathbf{z}_t \odot \tilde{\mathbf{h}}_t$.
Differentiating with respect to $\mathbf{h}_{t-1}$:
\[
  \frac{\partial \mathbf{h}_t}{\partial \mathbf{h}_{t-1}}
    = \diag(1 - \mathbf{z}_t)
    - \diag(\mathbf{h}_{t-1}) \frac{\partial \mathbf{z}_t}
      {\partial \mathbf{h}_{t-1}}
    + \diag(\tilde{\mathbf{h}}_t) \frac{\partial \mathbf{z}_t}
      {\partial \mathbf{h}_{t-1}}
    + \diag(\mathbf{z}_t) \frac{\partial \tilde{\mathbf{h}}_t}
      {\partial \mathbf{h}_{t-1}}.
\]
For the update gate: $\frac{\partial \mathbf{z}_t}
{\partial \mathbf{h}_{t-1}} = \diag(\mathbf{z}_t \odot
(1 - \mathbf{z}_t)) \mathbf{U}_z$.
For the candidate:
$\frac{\partial \tilde{\mathbf{h}}_t}{\partial \mathbf{h}_{t-1}}
= \diag(1 - \tilde{\mathbf{h}}_t^2) \mathbf{U}_h
(\diag(\mathbf{r}_t) + \diag(\mathbf{h}_{t-1})
\frac{\partial \mathbf{r}_t}{\partial \mathbf{h}_{t-1}})$,
where $\frac{\partial \mathbf{r}_t}{\partial \mathbf{h}_{t-1}} =
\diag(\mathbf{r}_t \odot (1 - \mathbf{r}_t)) \mathbf{U}_r$.
Collecting terms yields $\mathbf{A}_t$, $\mathbf{B}_t$, and
$\mathbf{C}_t$.
\end{proof}

\begin{corollary}[GRU Gradient Preservation]
\label{cor:gru-gradient}
When $\mathbf{z}_t \approx \mathbf{0}$ (update gate closed), the
dominant term in the Jacobian is $\diag(1 - \mathbf{z}_t) \approx
\mathbf{I}$, and the hidden state approximately copies forward:
$\mathbf{h}_t \approx \mathbf{h}_{t-1}$.  This provides a gradient
highway analogous to the LSTM's cell state pathway.
\end{corollary}

\begin{rlconnection}[title={GRU in Deep RL}]
In practice, GRUs and LSTMs perform comparably on most RL tasks.
The GRU's lower parameter count makes it attractive for:
\begin{itemize}[nosep]
  \item On-policy methods (PPO, A3C) where many forward passes are
    needed per gradient update.
  \item Memory-constrained settings (mobile robotics, embedded
    systems).
  \item Multi-agent RL where each agent has its own recurrent
    network.
\end{itemize}
\end{rlconnection}

%% ============================================================
%% SECTION 4: BIDIRECTIONAL RNNS
%% ============================================================
\section{Bidirectional Recurrent Neural Networks}
\label{sec:bidirectional}

\begin{definition}[Bidirectional RNN]
\label{def:bidirectional}
A \emph{bidirectional RNN} processes a sequence in both forward and
backward directions using two separate recurrent networks:
\begin{align}
  \overrightarrow{\mathbf{h}}_t &= f_{\text{forward}}(\mathbf{x}_t,
    \overrightarrow{\mathbf{h}}_{t-1}),
    \label{eq:bidir-forward} \\
  \overleftarrow{\mathbf{h}}_t &= f_{\text{backward}}(\mathbf{x}_t,
    \overleftarrow{\mathbf{h}}_{t+1}),
    \label{eq:bidir-backward}
\end{align}
where $f_{\text{forward}}$ and $f_{\text{backward}}$ are recurrent
cells (vanilla RNN, LSTM, or GRU) with independent parameters.
The outputs are concatenated:
\begin{equation}
  \mathbf{h}_t = [\overrightarrow{\mathbf{h}}_t;
    \overleftarrow{\mathbf{h}}_t] \in \R^{2m},
  \label{eq:bidir-concat}
\end{equation}
where $[\cdot;\cdot]$ denotes vertical concatenation.
\end{definition}

\begin{proposition}[Bidirectional RNN Parameter Count]
\label{prop:bidir-params}
A bidirectional RNN with cell type $C$ (having $P_C$ parameters per
direction) has $2P_C$ recurrent parameters plus the output layer
parameters.  The output layer maps from $2m$ dimensions instead of
$m$.
\end{proposition}

\begin{proof}
The forward and backward networks have independent parameters, so
the total recurrent parameters are $2P_C$.  The output layer
$\mathbf{V} \in \R^{k \times 2m}$ maps the concatenated hidden
state to the output.
\end{proof}

\begin{example}[Bidirectional Context]
\label{ex:bidir-context}
Consider named entity recognition in the sentence
``John visited Paris last summer.''  At the word ``Paris'':
\begin{itemize}[nosep]
  \item The forward hidden state $\overrightarrow{\mathbf{h}}_3$
    captures ``John visited''---suggesting a destination.
  \item The backward hidden state $\overleftarrow{\mathbf{h}}_3$
    captures ``last summer''---suggesting a temporal context.
\end{itemize}
The concatenated representation $[\overrightarrow{\mathbf{h}}_3;
\overleftarrow{\mathbf{h}}_3]$ captures both contexts, making it
easier to classify ``Paris'' as a location.
\end{example}

\begin{warning}[title={Bidirectional RNNs Require Full Sequences}]
Bidirectional RNNs can only be applied when the \emph{entire}
sequence is available before prediction (the backward pass requires
future inputs).  They are \textbf{not suitable} for:
\begin{itemize}[nosep]
  \item Online prediction (each output must be produced before
    seeing future inputs).
  \item Causal sequence-to-sequence tasks (language modeling,
    real-time RL).
  \item Streaming applications where inputs arrive one at a time.
\end{itemize}
They \textbf{are suitable} for tasks like sequence classification,
offline RL trajectory encoding, and post-hoc sequence labeling.
\end{warning}

\begin{remark}[Training Bidirectional RNNs]
\label{rem:bidir-training}
During training, the forward and backward networks are trained
independently on their respective directions.  The total gradient
at each parameter is the sum of gradients from the forward and
backward losses.  Computationally, the forward pass requires
two sequential scans of the sequence (one in each direction),
doubling the forward pass time compared to a unidirectional RNN.
\end{remark}

%% ============================================================
%% SECTION 5: DEEP (STACKED) RNNS
%% ============================================================
\section{Deep (Stacked) Recurrent Neural Networks}
\label{sec:deep-rnn}

\begin{definition}[Stacked RNN]
\label{def:stacked}
An $L$-layer \emph{stacked} (deep) RNN defines a hierarchy of
recurrent layers.  Let $\mathbf{h}_t^{(\ell)}$ denote the hidden
state at time $t$ and layer $\ell$.  The first layer processes the
input:
\begin{equation}
  \mathbf{h}_t^{(1)} = \text{Cell}^{(1)}(\mathbf{x}_t,
    \mathbf{h}_{t-1}^{(1)}),
  \label{eq:stacked-layer1}
\end{equation}
and subsequent layers use the hidden state of the layer below as
input:
\begin{equation}
  \mathbf{h}_t^{(\ell)} = \text{Cell}^{(\ell)}(
    \mathbf{h}_t^{(\ell-1)}, \mathbf{h}_{t-1}^{(\ell)}),
    \quad \ell = 2, \ldots, L,
  \label{eq:stacked-layerL}
\end{equation}
where $\text{Cell}^{(\ell)}$ is a recurrent cell (vanilla RNN,
LSTM, or GRU) with its own parameters.  The output is typically
taken from the top layer: $\mathbf{y}_t = g(\mathbf{V}
\mathbf{h}_t^{(L)} + \mathbf{c})$.
\end{definition}

\begin{proposition}[Stacked RNN Parameter Count]
\label{prop:stacked-params}
For an $L$-layer stacked RNN where each layer has hidden dimension
$m$ and the cell type has parameter count $P_C(n_\ell, m)$ for
input dimension $n_\ell$:
\begin{equation}
  P = P_C(n, m) + (L - 1) P_C(m, m) + k(m + 1),
  \label{eq:stacked-params}
\end{equation}
where the first term is the bottom layer (input dimension $n$), the
middle term covers layers $2$ through $L$ (input dimension $m$), and
the last term is the output layer.
\end{proposition}

\begin{proof}
Layer $1$ receives input of dimension $n$, contributing $P_C(n, m)$
parameters.  Layers $2, \ldots, L$ each receive hidden states of
dimension $m$, contributing $(L-1) P_C(m, m)$ parameters.  The
output layer $\mathbf{V} \in \R^{k \times m}$ and $\mathbf{c} \in
\R^k$ contribute $k(m + 1)$.
\end{proof}

\begin{example}[Stacked LSTM Parameter Count]
\label{ex:stacked-lstm}
A 3-layer stacked LSTM with $n = 300$, $m = 512$, $k = 10000$
(vocabulary size):
\begin{align*}
  \text{Layer 1:} &\quad 4 \times 512 \times (300 + 512 + 1)
    = 1{,}664{,}256, \\
  \text{Layers 2-3:} &\quad 2 \times 4 \times 512 \times
    (512 + 512 + 1) = 4{,}198{,}400, \\
  \text{Output:} &\quad 10000 \times (512 + 1) = 5{,}130{,}000. \\
  \text{Total:} &\quad 10{,}992{,}656 \approx 11\text{M}.
\end{align*}
\end{example}

\begin{remark}[Depth vs Width]
\label{rem:depth-vs-width}
Stacking layers adds \emph{depth} in the spatial dimension (across
layers at a fixed time step), complementing the temporal depth
inherent in the recurrence.  Each layer can learn increasingly
abstract representations:
\begin{itemize}[nosep]
  \item Layer 1: low-level features (character patterns, phonemes).
  \item Layer 2: mid-level features (word meanings, phrases).
  \item Layer 3: high-level features (sentence semantics, discourse).
\end{itemize}
In practice, 2--4 layers suffice for most tasks; deeper stacks
suffer from optimization difficulties similar to very deep
feedforward networks.
\end{remark}

\begin{rlconnection}[title={Deep Recurrent RL}]
In deep RL, stacked LSTMs or GRUs are used when the agent faces
complex partial observability.  For example, a 2-layer LSTM in an
Atari game with partial information: layer~1 processes visual
features (from a CNN) into temporal patterns, while layer~2
integrates these patterns into a belief state for action selection.
\end{rlconnection}

%% ============================================================
%% SECTION 6: ENCODER-DECODER FRAMEWORK
%% ============================================================
\section{Encoder-Decoder Framework}
\label{sec:encoder-decoder}

\begin{definition}[Encoder-Decoder Architecture]
\label{def:enc-dec}
The \emph{encoder-decoder} (sequence-to-sequence) architecture
\citep{sutskever2014,cho2014} maps a variable-length input sequence
$\mathbf{x}_{1:T}$ to a variable-length output sequence
$\mathbf{y}_{1:T'}$ (where $T' \neq T$ in general) through two
components:

\textbf{Encoder:} processes the input sequence and produces a
fixed-dimensional \emph{context vector}:
\begin{equation}
  \mathbf{h}_t^{\text{enc}} = f_{\text{enc}}(\mathbf{x}_t,
    \mathbf{h}_{t-1}^{\text{enc}}), \quad t = 1, \ldots, T,
  \label{eq:encoder}
\end{equation}
\begin{equation}
  \mathbf{q} = \phi(\mathbf{h}_T^{\text{enc}}),
  \label{eq:context}
\end{equation}
where $\phi$ is typically the identity ($\mathbf{q} =
\mathbf{h}_T^{\text{enc}}$).

\textbf{Decoder:} generates the output sequence conditioned on the
context vector:
\begin{equation}
  \mathbf{h}_t^{\text{dec}} = f_{\text{dec}}(\mathbf{y}_{t-1},
    \mathbf{h}_{t-1}^{\text{dec}}), \quad
    \mathbf{h}_0^{\text{dec}} = \mathbf{q},
  \label{eq:decoder}
\end{equation}
\begin{equation}
  \mathbf{y}_t = g(\mathbf{V} \mathbf{h}_t^{\text{dec}} +
    \mathbf{c}).
  \label{eq:decoder-output}
\end{equation}
\end{definition}

\begin{remark}[Context Vector as Information Bottleneck]
\label{rem:bottleneck}
The context vector $\mathbf{q} \in \R^m$ must encode the
\emph{entire} input sequence into a fixed-dimensional vector.  For
long input sequences, this creates an information bottleneck: the
encoder must compress arbitrarily long sequences into $m$ dimensions,
inevitably losing information.  This limitation motivated the
development of \emph{attention mechanisms} \citep{bahdanau2015},
which allow the decoder to access all encoder hidden states
rather than just the final one.
\end{remark}

\begin{example}[Machine Translation]
\label{ex:translation}
In English-to-French translation of ``The cat sat on the mat'':
\begin{enumerate}[nosep]
  \item The encoder processes the English tokens one by one,
    building up $\mathbf{h}_1^{\text{enc}}, \ldots,
    \mathbf{h}_6^{\text{enc}}$.
  \item The context vector $\mathbf{q} = \mathbf{h}_6^{\text{enc}}$
    summarizes the entire English sentence.
  \item The decoder generates French tokens autoregressively:
    ``Le'', ``chat'', ``s'est'', ``assis'', ``sur'', ``le'',
    ``tapis'', until a special end token is produced.
\end{enumerate}
The output length (7 tokens) differs from the input length
(6 tokens).
\end{example}

\begin{proposition}[Encoder-Decoder Training]
\label{prop:enc-dec-training}
The encoder-decoder is trained end-to-end by maximizing the
log-likelihood of the target sequence:
\begin{equation}
  \mathcal{L}(\boldsymbol{\theta}_{\text{enc}},
    \boldsymbol{\theta}_{\text{dec}})
    = -\sum_{t=1}^{T'} \log P(\mathbf{y}_t \mid
      \mathbf{y}_{1:t-1}, \mathbf{x}_{1:T}).
  \label{eq:enc-dec-loss}
\end{equation}
Gradients flow backward through the decoder, through the context
vector, and into the encoder via BPTT.
\end{proposition}

\begin{proof}
The decoder output $\mathbf{y}_t$ depends on
$\mathbf{h}_t^{\text{dec}}$, which depends on
$\mathbf{h}_{t-1}^{\text{dec}}$ (via the decoder recurrence) and
ultimately on $\mathbf{h}_0^{\text{dec}} = \mathbf{q} =
\mathbf{h}_T^{\text{enc}}$.  The encoder hidden state
$\mathbf{h}_T^{\text{enc}}$ depends on the encoder parameters
through the encoder recurrence.  Thus the chain rule connects the
decoder loss to the encoder parameters:
\[
  \frac{\partial \mathcal{L}}{\partial
    \boldsymbol{\theta}_{\text{enc}}}
    = \frac{\partial \mathcal{L}}
      {\partial \mathbf{h}_0^{\text{dec}}}
    \cdot \frac{\partial \mathbf{h}_T^{\text{enc}}}
      {\partial \boldsymbol{\theta}_{\text{enc}}}.
\]
Both components are computed via BPTT in their respective networks.
\end{proof}

\begin{rlconnection}[title={Encoder-Decoder in RL}]
The encoder-decoder framework appears in RL in several forms:
\begin{itemize}[nosep]
  \item \textbf{Goal-conditioned policies}: the encoder processes
    the goal description, and the decoder generates actions.
  \item \textbf{Trajectory transformers}: encode past trajectories
    and decode future actions or state predictions.
  \item \textbf{Communication in MARL}: agents encode messages using
    an encoder and decode received messages to inform actions
    (Phase~05, Lesson~64).
\end{itemize}
\end{rlconnection}

%% ============================================================
%% SECTION 7: EXERCISES
%% ============================================================
\section{Exercises}
\label{sec:exercises}

\begin{exercise}[GRU Forward Pass]
\textnormal{(\(\ast\))} \\
Consider a scalar GRU ($n = m = 1$) with all weight parameters
equal to $1$, all biases equal to $0$, and $h_0 = 0$.  Compute
$r_1, z_1, \tilde{h}_1, h_1$ for input $x_1 = 1$.  Then compute
$r_2, z_2, \tilde{h}_2, h_2$ for input $x_2 = 0$.
\end{exercise}

\begin{exercise}[GRU vs LSTM Gradient Analysis]
\textnormal{(\(\ast\ast\))} \\
For a GRU with constant update gate $\mathbf{z}_t = \epsilon
\mathbf{1}$ (small $\epsilon > 0$) and for an LSTM with constant
forget gate $\mathbf{f}_t = (1 - \epsilon)\mathbf{1}$, compute the
long-range gradient attenuation over $k$ steps for each
architecture.  Show that both architectures preserve gradients
when $\epsilon$ is small, but through different mechanisms.
\end{exercise}

\begin{exercise}[Bidirectional RNN Output]
\textnormal{(\(\ast\))} \\
A bidirectional GRU with $m = 128$ per direction processes a
sequence of length $T = 50$.  What is the dimension of the output
$\mathbf{h}_t$ at each time step?  If the output layer maps to a
vocabulary of $k = 10{,}000$ classes, how many parameters does the
output layer have?
\end{exercise}

\begin{exercise}[Encoder-Decoder Gradient Flow]
\textnormal{(\(\ast\ast\))} \\
In an encoder-decoder with a vanilla RNN encoder (hidden dimension
$m$) and LSTM decoder, derive the gradient of the decoder loss
$\mathcal{L}_1^{\text{dec}}$ (at the first decoder step) with
respect to the encoder hidden state $\mathbf{h}_1^{\text{enc}}$.
Express your answer in terms of the encoder Jacobian products.
Explain why the context vector bottleneck makes this gradient
vanish for long input sequences when the encoder is a vanilla RNN,
and how using an LSTM encoder helps.
\end{exercise}

\begin{exercise}[Stacked GRU Design]
\textnormal{(\(\ast\ast\))} \\
Design a 2-layer stacked GRU for a sentiment analysis task with
input embedding dimension $n = 200$, hidden dimension $m = 256$ per
layer, and $k = 5$ sentiment classes.  (a)~Compute the total
parameter count.  (b)~If we use a bidirectional GRU for the first
layer and a unidirectional GRU for the second layer, what changes
in the parameter count and the input dimension of the second layer?
\end{exercise}

\begin{exercise}[Attention Motivation]
\textnormal{(\(\ast\ast\ast\))} \\
Consider an encoder-decoder where the encoder produces hidden states
$\mathbf{h}_1^{\text{enc}}, \ldots, \mathbf{h}_T^{\text{enc}}$ and
the context is $\mathbf{q} = \mathbf{h}_T^{\text{enc}}$.  Suppose
instead we define a time-dependent context:
\[
  \mathbf{q}_t = \sum_{s=1}^{T} \alpha_{ts}
    \mathbf{h}_s^{\text{enc}}, \quad
  \alpha_{ts} = \frac{\exp(e_{ts})}{\sum_{s'=1}^T \exp(e_{ts'})},
  \quad
  e_{ts} = \mathbf{v}^\top \tanh(\mathbf{W}_a
    \mathbf{h}_s^{\text{enc}} + \mathbf{U}_a
    \mathbf{h}_{t-1}^{\text{dec}}).
\]
This is the attention mechanism of \citet{bahdanau2015}.
(a)~Show that when $\alpha_{ts} = \delta_{sT}$ (attention focuses
entirely on the last encoder state), this reduces to the standard
encoder-decoder.
(b)~Explain how attention alleviates the information bottleneck.
(c)~Derive $\frac{\partial \mathbf{q}_t}{\partial
\mathbf{h}_s^{\text{enc}}}$ and argue that attention provides a
``shortcut'' for gradient flow from the decoder to early encoder
states.

\textit{Hint:} For part (c), note that the gradient of
$\mathbf{q}_t$ with respect to $\mathbf{h}_s^{\text{enc}}$ includes
a direct term $\alpha_{ts} \mathbf{I}$ that does not pass through
the encoder recurrence.
\end{exercise}

%% ============================================================
%% SECTION 8: SUMMARY
%% ============================================================
\section{Summary}
\label{sec:summary}

\begin{itemize}
  \item The \textbf{GRU} simplifies the LSTM by merging cell and
    hidden states and using two gates (reset and update) instead
    of three, achieving $75\%$ of the LSTM's parameter count with
    comparable performance.
  \item The GRU hidden state update
    $\mathbf{h}_t = (1 - \mathbf{z}_t) \odot \mathbf{h}_{t-1} +
    \mathbf{z}_t \odot \tilde{\mathbf{h}}_t$ is a convex combination
    that provides gradient preservation when
    $\mathbf{z}_t \approx \mathbf{0}$.
  \item \textbf{Bidirectional RNNs} process sequences in both
    directions, capturing past and future context, but require the
    entire sequence to be available (not suitable for online or
    causal settings).
  \item \textbf{Stacked (deep) RNNs} add spatial depth by layering
    recurrent cells, enabling hierarchical feature extraction.  In
    practice, 2--4 layers suffice.
  \item The \textbf{encoder-decoder} framework maps variable-length
    input sequences to variable-length output sequences through a
    context vector, enabling tasks like machine translation.
  \item The context vector creates an \textbf{information bottleneck}
    for long sequences, motivating attention mechanisms that allow the
    decoder to access all encoder hidden states.
\end{itemize}

\subsection*{Subdomain 3.2 Summary: Recurrent Neural Networks}

The following table summarizes the arc of Lessons~67--69.

\begin{center}
\small
\begin{tabular}{p{1.5cm}p{3.5cm}p{5cm}p{3cm}}
\toprule
\textbf{Lesson} & \textbf{Topic} & \textbf{Key Results}
  & \textbf{Significance} \\
\midrule
67 & Vanilla RNN and BPTT
  & RNN forward equations; BPTT algorithm; vanishing gradient
    bound $\|\partial \mathcal{L}_t/\partial \mathbf{h}_\tau\|
    \leq C\gamma^{t-\tau}$; exploding gradient bound
  & Establishes the fundamental architecture and its core
    limitation \\[6pt]
68 & LSTM and Gating
  & LSTM forward equations (6 equations); cell state Jacobian
    $\partial \mathbf{c}_t/\partial \mathbf{c}_{t-1} =
    \diag(\mathbf{f}_t)$; gradient highway when
    $\mathbf{f}_t \approx \mathbf{1}$
  & Solves the vanishing gradient problem via additive cell
    state update \\[6pt]
69 & GRU and Sequence Architectures
  & GRU equations (4 equations); GRU Jacobian; bidirectional
    concatenation; stacked layers; encoder-decoder framework
  & Simplified gating + structural innovations for practical
    sequence processing \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Unifying theme:} the evolution from multiplicative
(vanilla RNN) to additive (LSTM) to simplified additive (GRU)
gradient flow, complemented by structural innovations
(bidirectional, stacked, encoder-decoder) that expand the
representational and functional capabilities of recurrent
architectures.

%% ============================================================
%% SECTION 9: FURTHER READING
%% ============================================================
\section{Further Reading}
\label{sec:further-reading}

\begin{itemize}
  \item The attention mechanism of \citet{bahdanau2015} resolves
    the encoder-decoder bottleneck and leads directly to the
    Transformer architecture, which has largely replaced RNNs in
    NLP.  A mathematical treatment of attention and Transformers
    would be a natural extension of this subdomain.
  \item \citet{chung2014} provides empirical comparisons of GRU
    and LSTM on music and speech modeling, finding no clear winner
    across all tasks.
  \item \citet{greff2017} is the most comprehensive ablation study
    of LSTM variants, testing the importance of each gate and
    providing practical recommendations.
  \item Phase~04 (Deep RL) builds on recurrent architectures:
    recurrent DQN (DRQN) uses LSTMs in value-based methods, while
    recurrent PPO uses LSTMs/GRUs for policies in POMDPs.
  \item \citet{graves2012} covers sequence transduction with RNNs,
    including connectionist temporal classification (CTC) for
    speech recognition and handwriting synthesis.
  \item The historical development of gated RNNs traces from
    \citet{hochreiter1991}'s identification of the vanishing
    gradient problem, through \citet{hochreiter1997}'s LSTM solution,
    \citet{gers2000}'s addition of the forget gate, to
    \citet{cho2014}'s GRU simplification.
\end{itemize}

\bibliography{nn-refs}
\end{document}
