\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 63: Universal Approximation}}
\fancyhead[R]{\thepage}
\title{Lesson 63: Universal Approximation}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% ============================================================
%% SECTION 1: MOTIVATION AND OVERVIEW
%% ============================================================
\section{Motivation and Overview}
\label{sec:motivation}

Suppose we wish to approximate a continuous function
$f \colon [0,1]^d \to \R$---say a value function $V^*(s)$ over a compact
state space, or a conditional density used in a generative model---using a
neural network with a single hidden layer and $m$ neurons.  The network
computes
\[
  x \;\mapsto\; \sum_{j=1}^{m} a_j \, \sigmoid(w_j^\top x + b_j),
\]
where $\sigmoid$ is a fixed activation function.  A natural and fundamental
question arises: \emph{for which activations $\sigmoid$ can such networks
approximate any continuous target $f$ to arbitrary accuracy, provided $m$ is
large enough?}  If the answer is negative, the entire architecture is
limited; if positive, the question shifts to efficiency (how large must $m$
be?) and learnability (can we find the weights?).  The tools developed in
this lesson---Cybenko's theorem and Hornik's generalization---show that the
answer is positive for essentially every activation used in practice, and
they identify the precise mathematical conditions that make universal
approximation possible.

\begin{rlconnection}[title={Why Universal Approximation Matters for RL/ML}]
\begin{itemize}[nosep]
  \item \textbf{Value function approximation.}  In deep RL, one
    approximates $V^*(s)$ or $Q^*(s,a)$ with a neural network.  Universal
    approximation guarantees that the hypothesis class is rich enough that
    the approximation error can, in principle, be driven to zero.
  \item \textbf{Policy gradient methods.}  Parameterizing policies
    $\pi_\theta(a \mid s)$ via neural networks requires that the network
    class be sufficiently expressive to represent near-optimal policies.
  \item \textbf{Generative models.}  Variational autoencoders and
    normalizing flows rely on neural networks to approximate complex
    densities; universal approximation justifies using these architectures.
  \item \textbf{Architecture design.}  Understanding \emph{which}
    activations yield universal approximation guides the choice between
    sigmoid, ReLU, and other nonlinearities.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
Compact subsets of $\R^d$ and continuous functions on them
(Lesson~2, Heine--Borel theorem).  Basic properties of $L^p$ spaces
(Lesson~4).  The Hahn--Banach theorem and its corollaries
(Lesson~21, Theorem~3.1).  The Riesz representation theorem for
$C(K)^*$ (Lesson~20, Theorem~3.2).  Familiarity with signed Borel
measures and weak-$*$ convergence.

\paragraph{Learning objectives.}
After completing this lesson, you will be able to:
\begin{itemize}[nosep]
  \item \textbf{Define} single-hidden-layer networks and state precisely
    what ``universal approximation'' means in $C(K)$ and $L^p(\mu)$.
  \item \textbf{State and prove} Cybenko's theorem for sigmoidal
    activations using the Hahn--Banach theorem and the Riesz representation
    theorem.
  \item \textbf{State and prove} Hornik's generalization for non-polynomial
    activations via the discriminatory function argument.
  \item \textbf{Verify} that specific activations (ReLU, tanh, sigmoid,
    softplus) satisfy the conditions for universal approximation.
  \item \textbf{Explain} the limitations of universal approximation
    theorems: the curse of dimensionality, the absence of constructive
    weight-finding, and the gap between approximation and generalization.
  \item \textbf{Justify} why Barron's theorem (Lesson~64) is needed to
    obtain quantitative approximation rates.
\end{itemize}

%% ============================================================
%% SECTION 2: REFERENCES
%% ============================================================
\section{References}
\label{sec:references}

\begin{itemize}
  \item \citet[Chapter 6]{bach2024} --- Presents universal approximation in
    the context of learning theory; includes both Cybenko's and Hornik's
    results with modern notation.
  \item \citet{cybenko1989} --- The original paper proving that sigmoidal
    networks are dense in $C(K)$.
  \item \citet{hornik1989} --- Extends Cybenko's result to general
    non-polynomial activations and $L^p$ approximation.
  \item \citet{pinkus1999} --- A comprehensive survey of approximation
    theory for multilayer perceptrons, including detailed proofs.
  \item \citet{lu2017} --- Modern results on the expressive power of
    bounded-width networks with ReLU activations.
  \item \citet{hanin2019} --- Universal approximation for deep ReLU
    networks with bounded width.
\end{itemize}

%% ============================================================
%% SECTION 3: SINGLE HIDDEN LAYER NETWORKS
%% ============================================================
\section{Single Hidden Layer Networks}
\label{sec:single-layer}

We begin by giving a precise definition of the function class produced by
a single-hidden-layer feedforward network.

\begin{definition}[Single-Hidden-Layer Network]
\label{def:single-layer}
Let $\sigmoid \colon \R \to \R$ be a fixed measurable function (the
\emph{activation function}).  For $d \ge 1$ and $m \ge 1$, define the class
of \emph{single-hidden-layer networks with $m$ neurons} by
\[
  \Fclass_m^{\sigmoid}
  \;=\;
  \biggl\{
    x \mapsto \sum_{j=1}^{m} a_j \, \sigmoid(w_j^\top x + b_j)
    \;:\;
    a_j \in \R, \;
    w_j \in \R^d, \;
    b_j \in \R
  \biggr\}.
\]
The class of all such networks (with any number of neurons) is
\[
  \Fclass^{\sigmoid}
  \;=\;
  \bigcup_{m=1}^{\infty} \Fclass_m^{\sigmoid}.
\]
Each function $f \in \Fclass_m^{\sigmoid}$ has $m(d+2)$ parameters:
$m$ output weights $a_j$, $m$ weight vectors $w_j \in \R^d$, and $m$
biases $b_j$.
\end{definition}

\begin{remark}
\label{rem:linear-span}
For each fixed $m$, the set $\Fclass_m^{\sigmoid}$ is \emph{not} a linear
subspace of $C(K)$ (it is not closed under addition unless we increase $m$).
However, $\Fclass^{\sigmoid}$ is a linear subspace: if $f \in
\Fclass_m^{\sigmoid}$ and $g \in \Fclass_n^{\sigmoid}$, then
$f + g \in \Fclass_{m+n}^{\sigmoid} \subseteq \Fclass^{\sigmoid}$.
\end{remark}

\begin{definition}[Universal Approximation Property]
\label{def:universal-approx}
Let $K \subseteq \R^d$ be compact.  We say the activation function
$\sigmoid$ has the \emph{universal approximation property in $C(K)$} if
$\Fclass^{\sigmoid}$ is dense in $C(K)$ with respect to the supremum norm:
for every $f \in C(K)$ and every $\varepsilon > 0$, there exist $m \ge 1$
and parameters $(a_j, w_j, b_j)_{j=1}^m$ such that
\[
  \sup_{x \in K}
  \biggl| f(x) - \sum_{j=1}^{m} a_j \, \sigmoid(w_j^\top x + b_j) \biggr|
  < \varepsilon.
\]
Similarly, $\sigmoid$ has the \emph{universal approximation property in
$L^p(\mu)$} (for a finite Borel measure $\mu$ on $\R^d$ and $1 \le p <
\infty$) if $\Fclass^{\sigmoid}$ is dense in $L^p(\mu)$.
\end{definition}

\begin{example}[Common Activations]
\label{ex:activations}
The following are widely used activation functions:
\begin{enumerate}[nosep,label=(\roman*)]
  \item \textbf{Logistic sigmoid:} $\sigmoid(t) = \frac{1}{1 + e^{-t}}$.
    Smooth, bounded, with $\sigmoid(t) \to 1$ as $t \to +\infty$ and
    $\sigmoid(t) \to 0$ as $t \to -\infty$.
  \item \textbf{Hyperbolic tangent:} $\sigmoid(t) = \tanh(t) =
    \frac{e^t - e^{-t}}{e^t + e^{-t}}$.  Smooth, bounded,
    $\sigmoid(t) \to \pm 1$ as $t \to \pm\infty$.
  \item \textbf{ReLU:} $\sigmoid(t) = \max(0, t)$.  Continuous but not
    bounded; piecewise linear.
  \item \textbf{Softplus:} $\sigmoid(t) = \log(1 + e^t)$.  Smooth
    approximation to ReLU; unbounded.
\end{enumerate}
\end{example}

\begin{example}[A Simple Network on {$[0,1]$}]
\label{ex:simple-net}
Let $d = 1$, $K = [0,1]$, $\sigmoid(t) = \frac{1}{1+e^{-t}}$ (logistic
sigmoid), and $m = 2$.  Consider
\[
  f(x) = a_1 \, \sigmoid(w_1 x + b_1) + a_2 \, \sigmoid(w_2 x + b_2).
\]
With $w_1 = w_2 = 10$, $b_1 = -3$, $b_2 = -7$, $a_1 = 1$, $a_2 = -1$,
the function $f$ approximates a ``bump'' that is close to $1$ on an
interval around $x \approx 0.3$--$0.7$ and close to $0$ elsewhere.  This
illustrates how two neurons can construct approximate indicator functions of
intervals, a building block used in the proof of Cybenko's theorem.
\end{example}

\begin{rlconnection}[title={Network Width and Value Function Quality}]
In deep RL, the number of neurons $m$ in the hidden layer controls the
expressiveness of the value function approximation.  Universal approximation
says that, for any continuous $V^*$, sufficiently large $m$ yields an
approximation within $\varepsilon$.  However, the theorem says nothing
about \emph{how large} $m$ must be---this quantitative question is
addressed by Barron's theorem (Lesson~64) and by Rademacher complexity
bounds (Lesson~60).
\end{rlconnection}

We now turn to the first major result: Cybenko's theorem, which resolves
the universal approximation question for sigmoidal activations.

%% ============================================================
%% SECTION 4: CYBENKO'S THEOREM
%% ============================================================
\section{Cybenko's Theorem}
\label{sec:cybenko}

Cybenko's 1989 result was the first rigorous proof that neural networks
with sigmoidal activations are universal approximators.  The proof
strategy is indirect: one uses the Hahn--Banach theorem to reduce the
density question to a property of signed measures, and then shows that
the sigmoidal assumption forces any annihilating measure to be zero.

\begin{definition}[Sigmoidal Function]
\label{def:sigmoidal}
A measurable function $\sigmoid \colon \R \to \R$ is \emph{sigmoidal} if
\[
  \lim_{t \to +\infty} \sigmoid(t) = 1
  \qquad \text{and} \qquad
  \lim_{t \to -\infty} \sigmoid(t) = 0.
\]
\end{definition}

\begin{example}[Sigmoidal Functions]
\label{ex:sigmoidal}
The logistic sigmoid $\sigmoid(t) = 1/(1+e^{-t})$ is sigmoidal.  The
rescaled hyperbolic tangent $\sigmoid(t) = \frac{1}{2}(\tanh(t) + 1)$ is
sigmoidal.  The Heaviside step function $\sigmoid(t) = \indicator_{[0,
\infty)}(t)$ is also sigmoidal (though discontinuous).  Note that the
standard $\tanh$ is \emph{not} sigmoidal in the sense of
Definition~\ref{def:sigmoidal}, since $\tanh(t) \to \pm 1$; however, it
satisfies Hornik's weaker condition (Section~\ref{sec:hornik}).
\end{example}

The key concept in Cybenko's proof is the notion of a discriminatory
function, which we now introduce.

\begin{definition}[Discriminatory Function]
\label{def:discriminatory}
Let $K \subseteq \R^d$ be compact.  A measurable function $\sigmoid
\colon \R \to \R$ is called \emph{discriminatory} (with respect to $K$)
if the following holds: for any finite signed Borel measure $\mu$ on $K$,
\[
  \int_K \sigmoid(w^\top x + b) \, d\mu(x) = 0
  \quad \text{for all } w \in \R^d, \; b \in \R
  \qquad \Longrightarrow \qquad
  \mu = 0.
\]
In other words, $\sigmoid$ is discriminatory if the only signed measure
annihilated by all functions $x \mapsto \sigmoid(w^\top x + b)$ is the
zero measure.
\end{definition}

\begin{lemma}[Sigmoidal Implies Discriminatory]
\label{lem:sig-disc}
If $\sigmoid \colon \R \to \R$ is continuous and sigmoidal
(Definition~\ref{def:sigmoidal}), then $\sigmoid$ is discriminatory with
respect to any compact $K \subseteq \R^d$.
\end{lemma}

\begin{proof}
Let $\mu$ be a finite signed Borel measure on $K$ such that
\begin{equation}
\label{eq:annihilating}
  \int_K \sigmoid(w^\top x + b) \, d\mu(x) = 0
  \qquad \text{for all } w \in \R^d, \; b \in \R.
\end{equation}
We must show $\mu = 0$.

\textbf{Step 1: Pointwise limit.}
Fix $w \in \R^d$ and $\theta \in \R$, and define the half-space
$H_{w,\theta} = \{x \in \R^d : w^\top x + \theta > 0\}$.  For
$\lambda > 0$ and any $b \in \R$, consider
\[
  \sigmoid(\lambda(w^\top x + b)).
\]
Since $\sigmoid$ is sigmoidal, as $\lambda \to +\infty$:
\[
  \sigmoid(\lambda(w^\top x + b))
  \;\to\;
  \begin{cases}
    1 & \text{if } w^\top x + b > 0, \\
    0 & \text{if } w^\top x + b < 0, \\
    \sigmoid(0) & \text{if } w^\top x + b = 0.
  \end{cases}
\]
That is, $\sigmoid(\lambda(w^\top x + b)) \to \indicator_{H_{w,-b}}(x)$
for $\mu$-almost every $x$ (the hyperplane $\{x : w^\top x + b = 0\}$ has
$|\mu|$-measure zero for all but countably many values of $b$).

\textbf{Step 2: Passing to the limit.}
By \eqref{eq:annihilating}, for every $\lambda > 0$ and $b \in \R$,
\[
  \int_K \sigmoid(\lambda(w^\top x + b)) \, d\mu(x) = 0.
\]
(Here we replace $w$ by $\lambda w$ and use the fact that the condition
holds for all $w$.)
The function $\sigmoid$ is bounded (since it is continuous and converges at
$\pm\infty$), so by the bounded convergence theorem, taking
$\lambda \to +\infty$:
\[
  \int_K \indicator_{H_{w,-b}}(x) \, d\mu(x) = 0,
\]
i.e., $\mu(H_{w,-b}) = 0$ for all half-spaces $H_{w,-b}$ (for almost
every $b$, and by a density argument for all $b$).

\textbf{Step 3: Half-spaces generate the Borel $\sigmoid$-algebra.}
Since the open half-spaces $\{H_{w,\theta} : w \in \R^d, \theta \in \R\}$
generate the Borel $\sigmoid$-algebra $\Borel(\R^d)$, a signed measure that
vanishes on all half-spaces must be the zero measure.  To see this, note
that finite intersections of half-spaces form all open convex polytopes, and
these form a $\pi$-system generating $\Borel(\R^d)$.  Since $\mu$ vanishes
on this $\pi$-system, it vanishes on the generated $\sigmoid$-algebra by
the $\pi$-$\lambda$ theorem (Lesson~1, Theorem~3.3).  Therefore $\mu = 0$.
\end{proof}

We now state and prove the main result.

\begin{theorem}[Cybenko, 1989]
\label{thm:cybenko}
Let $K \subseteq \R^d$ be compact and let $\sigmoid \colon \R \to \R$ be
any continuous discriminatory function.  Then $\Fclass^{\sigmoid}$ is dense
in $C(K)$ with respect to the supremum norm $\|f\|_\infty = \sup_{x \in K}
|f(x)|$.
\end{theorem}

\begin{proof}
Suppose for contradiction that $\overline{\Fclass^{\sigmoid}} \neq C(K)$,
where the closure is in the supremum norm.  Since $\Fclass^{\sigmoid}$ is
a linear subspace of $C(K)$ (Remark~\ref{rem:linear-span}), its closure
$\overline{\Fclass^{\sigmoid}}$ is a closed linear subspace of $C(K)$ that
is strictly contained in $C(K)$.

By the Hahn--Banach theorem (Lesson~21, Theorem~3.1), there exists a
bounded linear functional $L \in C(K)^*$ such that $L \neq 0$ and
$L(g) = 0$ for all $g \in \overline{\Fclass^{\sigmoid}}$.

By the Riesz representation theorem for $C(K)$ (Lesson~20, Theorem~3.2),
every bounded linear functional on $C(K)$ is represented by a finite signed
Borel measure: there exists a finite signed Borel measure $\mu$ on $K$ such
that
\[
  L(g) = \int_K g(x) \, d\mu(x)
  \qquad \text{for all } g \in C(K).
\]
In particular, for every $w \in \R^d$ and $b \in \R$, the function
$x \mapsto \sigmoid(w^\top x + b)$ belongs to $\Fclass_1^{\sigmoid}
\subseteq \Fclass^{\sigmoid}$, so
\[
  \int_K \sigmoid(w^\top x + b) \, d\mu(x) = L(\sigmoid(w^\top \cdot + b))
  = 0.
\]
Since $\sigmoid$ is discriminatory (Definition~\ref{def:discriminatory}),
this implies $\mu = 0$, hence $L = 0$.  This contradicts $L \neq 0$.

Therefore $\overline{\Fclass^{\sigmoid}} = C(K)$, i.e.,
$\Fclass^{\sigmoid}$ is dense in $C(K)$.
\end{proof}

\begin{keyresult}[title={Cybenko's Universal Approximation Theorem}]
Combining Theorem~\ref{thm:cybenko} and Lemma~\ref{lem:sig-disc}: if
$\sigmoid$ is continuous and sigmoidal, then for any compact $K \subseteq
\R^d$, any $f \in C(K)$, and any $\varepsilon > 0$, there exist $m \ge 1$
and parameters $(a_j, w_j, b_j)_{j=1}^m$ such that
\[
  \sup_{x \in K}
  \biggl| f(x) - \sum_{j=1}^{m} a_j \, \sigmoid(w_j^\top x + b_j) \biggr|
  < \varepsilon.
\]
\end{keyresult}

\begin{example}[Approximating a Step Function]
\label{ex:step-approx}
Let $K = [0,1]$, $f(x) = \indicator_{[1/3, 2/3]}(x)$ (which is not
continuous, but can be approximated by continuous functions).  Consider
a continuous approximation $f_\delta$ with linear ramps of width $\delta$
at the endpoints.  For $\varepsilon > 0$, Cybenko's theorem guarantees
that $f_\delta$ can be approximated within $\varepsilon$ by a network
with sigmoidal activation.  In practice, this is achieved by using neurons
with large weights to create steep transitions, as in
Example~\ref{ex:simple-net}.
\end{example}

\begin{example}[Constructive Illustration in 1D]
\label{ex:constructive-1d}
To approximate $f(x) = x^2$ on $[0,1]$ using the logistic sigmoid
$\sigmoid(t) = 1/(1+e^{-t})$, one can partition $[0,1]$ into $m$ equal
intervals and use $2m$ neurons to create approximate bump functions on
each interval, scaling each bump by the value of $f$ at the interval
midpoint.  This piecewise-constant approximation achieves
$\|f - f_m\|_\infty \le \frac{1}{4m^2}$ (since $f'' = 2$ gives error
$\frac{1}{2} \cdot \frac{f''}{8} \cdot (1/m)^2$).  The number of neurons
grows as $O(1/\varepsilon^{1/2})$ for this particular smooth target.
\end{example}

\begin{remark}
\label{rem:non-constructive}
Cybenko's proof is \emph{non-constructive}: it shows that a good
approximation exists but provides no algorithm for finding the weights
$(a_j, w_j, b_j)$.  The Hahn--Banach theorem yields existence via
contradiction, not via an explicit construction.  This limitation motivates
both (a)~constructive proofs based on sampling (Barron's theorem,
Lesson~64) and (b)~the study of optimization algorithms for training
(Lesson~65).
\end{remark}

Having established Cybenko's theorem for sigmoidal activations, we next
ask: what happens for activations that are not sigmoidal?

%% ============================================================
%% SECTION 5: HORNIK'S GENERALIZATION
%% ============================================================
\section{Hornik's Generalization}
\label{sec:hornik}

Cybenko's theorem covers sigmoidal activations (those converging to $0$
and $1$ at $\pm\infty$), but the most popular activation in modern deep
learning---the ReLU $\sigmoid(t) = \max(0,t)$---is not sigmoidal.
Hornik, Stinchcombe, and White (1989) showed that universal approximation
holds for a much broader class of activations.

\begin{definition}[Non-Polynomial Activation]
\label{def:non-polynomial}
A continuous function $\sigmoid \colon \R \to \R$ is called
\emph{non-polynomial} if $\sigmoid$ is not a polynomial, i.e., there is no
polynomial $p$ such that $\sigmoid(t) = p(t)$ for all $t \in \R$.
\end{definition}

\begin{remark}
\label{rem:polynomial-fail}
If $\sigmoid$ is a polynomial of degree $k$, then every function in
$\Fclass^{\sigmoid}$ is also a polynomial of degree at most $k$ in each
coordinate.  In particular, $\Fclass^{\sigmoid}$ cannot approximate all
continuous functions.  Therefore the non-polynomial condition is
\emph{necessary} for universal approximation.
\end{remark}

We now show that continuous non-polynomial activations are discriminatory,
which by Cybenko's framework (Theorem~\ref{thm:cybenko}) immediately
implies universal approximation.

\begin{lemma}[Non-Polynomial Implies Discriminatory]
\label{lem:non-poly-disc}
If $\sigmoid \colon \R \to \R$ is continuous and non-polynomial, then
$\sigmoid$ is discriminatory with respect to any compact $K \subseteq \R^d$.
\end{lemma}

\begin{proof}
Let $\mu$ be a finite signed Borel measure on $K$ satisfying
\begin{equation}
\label{eq:annihilating2}
  \int_K \sigmoid(w^\top x + b) \, d\mu(x) = 0
  \qquad \text{for all } w \in \R^d, \; b \in \R.
\end{equation}
We show that $\mu = 0$.

\textbf{Step 1: The annihilating condition extends to polynomials.}
Define the function $F \colon \R^d \times \R \to \R$ by
\[
  F(w, b) = \int_K \sigmoid(w^\top x + b) \, d\mu(x).
\]
By \eqref{eq:annihilating2}, $F(w, b) = 0$ for all $(w, b)$.  Since
$\sigmoid$ is continuous and $K$ is compact, $F$ is infinitely
differentiable in $(w, b)$, and we can differentiate under the integral sign.
In particular, for any multi-index $\alpha = (\alpha_1, \ldots, \alpha_d)$
and any $n \ge 0$,
\[
  \frac{\partial^{|\alpha|+n} F}{\partial w_1^{\alpha_1} \cdots
  \partial w_d^{\alpha_d} \, \partial b^n}(w, b)
  = \int_K x_1^{\alpha_1} \cdots x_d^{\alpha_d} \,
  \sigmoid^{(|\alpha|+n)}(w^\top x + b) \, d\mu(x) = 0,
\]
where $\sigmoid^{(k)}$ denotes the $k$-th derivative of $\sigmoid$.

\textbf{Step 2: Evaluating at the origin.}
Setting $w = 0$ and $b = 0$:
\[
  \sigmoid^{(|\alpha|+n)}(0)
  \int_K x_1^{\alpha_1} \cdots x_d^{\alpha_d} \, d\mu(x) = 0.
\]
Since $\sigmoid$ is not a polynomial, its derivatives of all orders cannot
all vanish at a single point: there exists a sequence of orders $k_j$ such
that $\sigmoid^{(k_j)}(0) \neq 0$.  More precisely, if
$\sigmoid^{(k)}(0) = 0$ for all $k \ge N$ for some $N$, then $\sigmoid$
would be a polynomial of degree at most $N-1$ in a neighborhood of $0$
(by Taylor's theorem), and by analyticity arguments on each interval,
$\sigmoid$ would be a polynomial globally---contradicting the assumption.

Therefore, for each choice of $|\alpha| + n$, we can find a derivative order
where $\sigmoid$ does not vanish.  Specifically, for any monomial exponent
$(p_1, \ldots, p_d)$ with $p_1 + \cdots + p_d = k$, we choose
$\alpha = (p_1, \ldots, p_d)$ and $n = 0$ and pick $w = 0$, $b = 0$;
if $\sigmoid^{(k)}(0) = 0$, we instead choose $n > 0$ so that
$|\alpha| + n = k + n$ where $\sigmoid^{(k+n)}(0) \neq 0$, and absorb the
extra $b$-derivatives.  After dividing by the nonzero constant
$\sigmoid^{(k+n)}(0)$, we obtain
\begin{equation}
\label{eq:moments-zero}
  \int_K x_1^{p_1} \cdots x_d^{p_d} \, d\mu(x) = 0
  \qquad \text{for all } p_1, \ldots, p_d \ge 0.
\end{equation}

\textbf{Step 3: Moment characterization.}
Equation~\eqref{eq:moments-zero} says that all moments of $\mu$ vanish.
Since $K$ is compact, $\mu$ is a compactly supported signed measure, and
the Weierstrass approximation theorem (or Stone--Weierstrass) implies that
polynomials are dense in $C(K)$.  Therefore, for any $g \in C(K)$, we
can approximate $g$ uniformly by polynomials $p_n$, and
\[
  \int_K g(x) \, d\mu(x) = \lim_{n \to \infty}
  \int_K p_n(x) \, d\mu(x) = 0.
\]
Since this holds for all $g \in C(K)$, the Riesz representation theorem
implies $\mu = 0$.
\end{proof}

Combining Lemma~\ref{lem:non-poly-disc} with Theorem~\ref{thm:cybenko},
we obtain:

\begin{theorem}[Hornik, Stinchcombe, and White, 1989]
\label{thm:hornik}
Let $K \subseteq \R^d$ be compact and let $\sigmoid \colon \R \to \R$ be
any continuous, non-polynomial function.  Then $\Fclass^{\sigmoid}$ is dense
in $C(K)$ with respect to the supremum norm.  More generally,
$\Fclass^{\sigmoid}$ is dense in $L^p(\mu)$ for any finite Borel measure
$\mu$ on $K$ and $1 \le p < \infty$.
\end{theorem}

\begin{proof}
The $C(K)$ density follows directly from Theorem~\ref{thm:cybenko}
and Lemma~\ref{lem:non-poly-disc}: since continuous non-polynomial
activations are discriminatory, Cybenko's framework applies.

For the $L^p(\mu)$ extension: since $C(K)$ is dense in $L^p(\mu)$ (as
$K$ is compact), and $\Fclass^{\sigmoid}$ is dense in $C(K)$, transitivity
of density gives that $\Fclass^{\sigmoid}$ is dense in $L^p(\mu)$.
\end{proof}

\begin{keyresult}[title={Hornik's Universal Approximation Theorem}]
For any continuous non-polynomial activation $\sigmoid$ (including ReLU,
tanh, logistic sigmoid, softplus, ELU, swish, and essentially every
activation used in practice), single-hidden-layer networks are dense in
$C(K)$ for compact $K$ and in $L^p(\mu)$ for any finite measure $\mu$.
The only activations that fail are polynomials.
\end{keyresult}

\begin{example}[ReLU Is a Universal Approximator]
\label{ex:relu-universal}
The ReLU activation $\sigmoid(t) = \max(0, t)$ is continuous and
non-polynomial (it is not differentiable at $t = 0$, so it cannot be a
polynomial).  Therefore, by Theorem~\ref{thm:hornik}, ReLU networks are
universal approximators.

Concretely, a ReLU network $f_m(x) = \sum_{j=1}^m a_j \max(0, w_j^\top x
+ b_j)$ computes a piecewise-affine function on $\R^d$.  The set of
``breakpoints'' (where some neuron transitions between its active and
inactive regimes) divides $\R^d$ into convex polytopes, and $f_m$ is affine
on each polytope.  As $m \to \infty$, the number of polytopes grows and
the piecewise-affine function can approximate any continuous function.
\end{example}

\begin{example}[Verification for $\tanh$]
\label{ex:tanh}
The hyperbolic tangent $\sigmoid(t) = \tanh(t)$ is continuous and
non-polynomial (its Taylor series $\tanh(t) = t - t^3/3 + 2t^5/15 - \cdots$
has infinitely many nonzero terms and converges on all of $\R$; a polynomial
would have only finitely many).  Therefore $\tanh$ is a universal
approximator by Theorem~\ref{thm:hornik}.
\end{example}

\begin{warning}[title={Density Does Not Imply Efficient Approximation}]
Universal approximation guarantees that for every $\varepsilon > 0$ there
\emph{exists} a network with finitely many neurons achieving error less than
$\varepsilon$.  It does \emph{not} bound the number of neurons $m$ as a
function of $\varepsilon$ and $d$.  For generic continuous functions on
$[0,1]^d$, achieving $\varepsilon$-accuracy may require
$m = \Omega(\varepsilon^{-d})$ neurons---an exponential dependence on
dimension.  This is the \emph{curse of dimensionality}, which motivates
Barron's theorem (Lesson~64).
\end{warning}

We now examine precisely what universal approximation theorems do and do
not tell us.

%% ============================================================
%% SECTION 6: LIMITATIONS OF UNIVERSAL APPROXIMATION
%% ============================================================
\section{Limitations of Universal Approximation}
\label{sec:limitations}

Universal approximation is an existence theorem: it guarantees that neural
networks form a rich enough function class.  However, three critical gaps
remain between the theoretical guarantee and practical neural network
performance.

\subsection{Gap 1: Approximation Rate (The Curse of Dimensionality)}
\label{sec:curse}

\begin{proposition}[Lower Bound for Generic Functions]
\label{prop:lower-bound}
Let $\Fclass_m^{\sigmoid}$ denote the class of single-hidden-layer
networks with $m$ neurons and any continuous non-polynomial activation.
For the class of all $L$-Lipschitz functions on $[0,1]^d$, the minimax
approximation error satisfies
\[
  \inf_{f_m \in \Fclass_m^{\sigmoid}} \;
  \sup_{\substack{f : [0,1]^d \to \R \\ \text{$L$-Lipschitz}}}
  \|f - f_m\|_{L^2}
  \;\ge\;
  c \, L \, m^{-1/d}
\]
for a constant $c > 0$ depending only on $d$.  That is, approximating
generic Lipschitz functions to accuracy $\varepsilon$ requires
$m = \Omega(\varepsilon^{-d})$ neurons.
\end{proposition}

\begin{proof}
This follows from classical lower bounds in approximation theory.  The
class of $L$-Lipschitz functions on $[0,1]^d$ has
$\varepsilon$-covering number (in $L^2$) at least
$\exp(c' (L/\varepsilon)^d)$ for a constant $c' > 0$.  On the other
hand, $\Fclass_m^{\sigmoid}$ has $m(d+2)$ parameters, so its covering
number is at most $\exp(c'' \, m(d+2) \log(1/\varepsilon))$.  Matching
these gives $m \ge c \, (L/\varepsilon)^d / ((d+2) \log(1/\varepsilon))$,
which yields the stated $m^{-1/d}$ rate.
\end{proof}

\begin{example}[Dimension Scaling]
\label{ex:dimension-scaling}
To approximate a Lipschitz function on $[0,1]^{100}$ to accuracy
$\varepsilon = 0.01$ in the worst case, the lower bound requires
$m = \Omega(100^{100})$ neurons---an astronomically large network.
In contrast, Barron's theorem (Lesson~64) shows that for functions in
the Barron class, $m = O(1/\varepsilon^2)$ suffices regardless of $d$,
dramatically escaping this curse.
\end{example}

\subsection{Gap 2: Finding the Weights (Optimization)}
\label{sec:optimization-gap}

Even when a good approximation exists, the training problem
\[
  \min_{a_j, w_j, b_j} \;
  \frac{1}{n} \sum_{i=1}^{n}
  \biggl( y_i - \sum_{j=1}^{m} a_j \, \sigmoid(w_j^\top x_i + b_j)
  \biggr)^2
\]
is non-convex in the parameters $(w_j, b_j)$ whenever $\sigmoid$ is
nonlinear.  Universal approximation says nothing about whether gradient
descent or any other algorithm can find a (near-)global minimum.

\begin{remark}
\label{rem:ntk}
Recent work on the \emph{neural tangent kernel} (NTK)
\citep{jacot2018} shows that, in the infinite-width limit ($m \to \infty$),
gradient descent on overparameterized networks converges to a global minimum
of the training loss.  This connects the approximation theory of this lesson
to the optimization theory of Lesson~65.  However, the NTK regime requires
$m$ to be very large (polynomial in $n$), and it is debated whether this
regime captures the practical behavior of finite-width networks.
\end{remark}

\subsection{Gap 3: Generalization from Finite Data}
\label{sec:generalization-gap}

Approximation error measures how well the best network in
$\Fclass_m^{\sigmoid}$ can represent the target function.  But in
practice, we only have $n$ training samples and must also control the
\emph{estimation error}: the gap between the best-in-class network and
the network we actually find from data.  The total error decomposes as
\[
  \underbrace{\Risk(\hat{f}_n) - \inf_{f \in \Fclass} \Risk(f)}_{
    \text{estimation error}}
  \;+\;
  \underbrace{\inf_{f \in \Fclass} \Risk(f) - \inf_{f} \Risk(f)}_{
    \text{approximation error}},
\]
where $\Risk(f) = \E[\Loss(f(X), Y)]$ is the population risk.  Universal
approximation controls the second term, but the first term depends on the
complexity of $\Fclass_m^{\sigmoid}$ (measured, e.g., by Rademacher
complexity).  Making $m$ larger reduces approximation error but increases
estimation error---a fundamental tradeoff analyzed in Lesson~60.

\begin{rlconnection}[title={The Bias-Variance Tradeoff in RL}]
In deep RL, the same tradeoff appears:
\begin{itemize}[nosep]
  \item A very large network ($m \gg 1$) can represent $V^*$ well
    (\emph{low approximation error}), but may overfit to sampled
    transitions (\emph{high estimation error}).
  \item A small network may underfit $V^*$ (\emph{high approximation
    error}) but generalize well from limited data.
\end{itemize}
Architecture selection in DQN, PPO, and SAC implicitly balances these
competing effects.
\end{rlconnection}

\begin{example}[Overfitting with Universal Approximators]
\label{ex:overfit}
A network with $m = n$ neurons can interpolate $n$ arbitrary data points
$(x_i, y_i)$ exactly (under mild conditions on the $x_i$).  This means
the training error is zero, yet the test error can be arbitrarily large.
Universal approximation guarantees richness, not generalization.
\end{example}

The limitations identified in this section---exponential width dependence
on dimension, non-convex optimization, and the approximation-estimation
tradeoff---motivate the study of Barron's theorem in Lesson~64, which
provides dimension-free approximation rates for a natural subclass of
functions.

%% ============================================================
%% SECTION 7: EXERCISES
%% ============================================================
\section{Exercises}
\label{sec:exercises}

\begin{exercise}[Discriminatory Verification]
\textnormal{(\({}^*\))} \\
Show that the Heaviside step function $\sigmoid(t) =
\indicator_{[0,\infty)}(t)$ is discriminatory, directly from
Definition~\ref{def:discriminatory}, without using
Lemma~\ref{lem:sig-disc}. \\
\textit{Hint:} If $\int_K \indicator_{[0,\infty)}(w^\top x + b) \,
d\mu(x) = 0$ for all $w, b$, show that $\mu(H) = 0$ for every open
half-space $H$.
\end{exercise}

\begin{exercise}[Polynomial Activations Fail]
\textnormal{(\({}^*\))} \\
Let $\sigmoid(t) = t^2$.  Show that $\Fclass^{\sigmoid}$ is \emph{not}
dense in $C([0,1])$.  Specifically, find a continuous function on $[0,1]$
that cannot be approximated by functions of the form $\sum_{j=1}^m a_j
(w_j x + b_j)^2$, and prove your claim.
\end{exercise}

\begin{exercise}[ReLU Piecewise Affine Representation]
\textnormal{(\({}^{**}\))} \\
Let $d = 1$ and $\sigmoid(t) = \max(0, t)$ (ReLU).
\begin{enumerate}[label=(\alph*)]
  \item Show that any $f \in \Fclass_m^{\sigmoid}$ is a continuous
    piecewise-affine function on $\R$ with at most $m$ breakpoints.
  \item Conversely, show that any continuous piecewise-affine function
    on $\R$ with $k$ breakpoints belongs to $\Fclass_{k+1}^{\sigmoid}$.
  \item Using part (b), prove directly (without Theorem~\ref{thm:hornik})
    that ReLU networks are dense in $C([a,b])$ for any $[a,b]$.
\end{enumerate}
\end{exercise}

\begin{exercise}[Softplus Approximation of ReLU]
\textnormal{(\({}^{**}\))} \\
The softplus activation is $\sigmoid_\beta(t) = \frac{1}{\beta}
\log(1 + e^{\beta t})$ for $\beta > 0$.
\begin{enumerate}[label=(\alph*)]
  \item Show that $\sigmoid_\beta(t) \to \max(0, t)$ pointwise as
    $\beta \to \infty$.
  \item Show that the convergence is uniform on compact subsets of
    $\R \setminus \{0\}$ but \emph{not} uniform on any neighborhood of
    $0$.
  \item Conclude that if softplus networks are universal approximators
    (which they are by Theorem~\ref{thm:hornik}), then ReLU networks are
    also universal approximators, giving an alternative proof for ReLU.
\end{enumerate}
\end{exercise}

\begin{exercise}[Value Function Approximation in RL]
\textnormal{(\({}^{**}\))} \\
Consider a continuous MDP with state space $K = [0,1]^d$ and a continuous
optimal value function $V^* \colon K \to [0, V_{\max}]$.  Let
$\hat{V}_m$ be the best approximation of $V^*$ in $\Fclass_m^{\sigmoid}$
(with $\sigmoid = \text{ReLU}$) in the supremum norm.
\begin{enumerate}[label=(\alph*)]
  \item State precisely what Theorem~\ref{thm:hornik} guarantees about
    $\|V^* - \hat{V}_m\|_\infty$ as $m \to \infty$.
  \item Explain why this guarantee alone is insufficient for practical
    deep RL.  Identify at least two additional quantities that must be
    controlled.
  \item If $V^*$ belongs to the Barron class (defined in Lesson~64),
    state the improved approximation rate.
\end{enumerate}
\end{exercise}

\begin{exercise}[Universal Approximation in $L^2$ but not $C$]
\textnormal{(\({}^{***}\))} \\
Construct a bounded measurable (but discontinuous) activation $\sigmoid$
that is discriminatory but for which $\Fclass^{\sigmoid}$ is \emph{not}
contained in $C(K)$ (and therefore the $C(K)$-density statement does
not apply, though $L^2$-density may still hold).  Specifically, let
$\sigmoid = \indicator_{[0,\infty)}$ and show:
\begin{enumerate}[label=(\alph*)]
  \item Every $f \in \Fclass^{\sigmoid}$ is a step function (finite
    combination of indicator functions of half-spaces).
  \item $\Fclass^{\sigmoid}$ is dense in $L^2(\mu)$ for any
    compactly supported absolutely continuous measure $\mu$.
  \item $\Fclass^{\sigmoid}$ is \emph{not} dense in $C(K)$ because
    elements of $\Fclass^{\sigmoid}$ are not continuous.
\end{enumerate}
\end{exercise}

%% ============================================================
%% SECTION 8: SUMMARY
%% ============================================================
\section{Summary}
\label{sec:summary}

\begin{itemize}
  \item A single-hidden-layer network with $m$ neurons computes
    $f(x) = \sum_{j=1}^m a_j \sigmoid(w_j^\top x + b_j)$; the class
    $\Fclass^{\sigmoid} = \bigcup_m \Fclass_m^{\sigmoid}$ forms a
    linear subspace of the function space.

  \item An activation $\sigmoid$ is \emph{discriminatory} if the only
    signed measure annihilated by all functions $x \mapsto
    \sigmoid(w^\top x + b)$ is the zero measure.

  \item \textbf{Cybenko's theorem:} If $\sigmoid$ is continuous and
    discriminatory (which holds for all continuous sigmoidal functions),
    then $\Fclass^{\sigmoid}$ is dense in $C(K)$ for compact $K$.  The
    proof uses the Hahn--Banach theorem and the Riesz representation
    theorem.

  \item \textbf{Hornik's theorem:} The discriminatory condition holds for
    all continuous non-polynomial activations.  Therefore, any continuous
    non-polynomial $\sigmoid$ (including ReLU, tanh, sigmoid, softplus)
    yields universal approximation in $C(K)$ and $L^p(\mu)$.

  \item The only activations that fail to be universal approximators are
    polynomials, for which the output is necessarily polynomial.

  \item Universal approximation is a \emph{qualitative} result: it
    guarantees density but does not specify (a)~how many neurons are
    needed, (b)~how to find the weights, or (c)~how the network
    generalizes from finite data.

  \item For generic Lipschitz functions on $[0,1]^d$, achieving
    $\varepsilon$-accuracy requires $m = \Omega(\varepsilon^{-d})$
    neurons in the worst case---an exponential dependence on dimension.

  \item Barron's theorem (Lesson~64) escapes this curse for a restricted
    but natural class of functions, providing a dimension-free
    $O(1/m)$ rate.
\end{itemize}

%% ============================================================
%% SECTION 9: FURTHER READING
%% ============================================================
\section{Further Reading}
\label{sec:further-reading}

\begin{itemize}
  \item \textbf{Lesson 64} extends the analysis to quantitative
    approximation rates via Barron's theorem, showing that for functions
    with bounded Fourier moment, the rate $O(1/m)$ is dimension-free.

  \item \textbf{Lesson 65} addresses the optimization question: how
    gradient-based methods find good parameters in practice, including
    the role of overparameterization and the neural tangent kernel.

  \item \citet{pinkus1999} provides a comprehensive survey of
    approximation-theoretic results for neural networks, including
    generalizations to multiple hidden layers.

  \item \citet{telgarsky2016} and \citet{eldan2016} study the
    \emph{depth} advantage: there exist functions expressible by depth-$k$
    networks with polynomial width that require exponential width at depth
    $k-1$.  This motivates deep (multi-layer) architectures beyond the
    single-layer setting of this lesson.

  \item \citet{hanin2019} proves universal approximation for deep ReLU
    networks with \emph{bounded width} (width $d+1$ suffices for
    $L^p$ approximation on $\R^d$), complementing the width-based
    results studied here.

  \item \textbf{Historical note.}  Cybenko's 1989 paper was motivated by
    the question of whether ``connectionist'' models (the terminology of
    the era) had fundamental representational limitations.  Hornik,
    Stinchcombe, and White's concurrent 1989 paper gave a more general
    answer.  Both built on classical functional analysis---particularly
    the Hahn--Banach theorem---showing a beautiful application of pure
    mathematics to a question from machine learning.
\end{itemize}

\bibliography{nn-refs}
\end{document}
