\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 66: NTK and Rademacher Complexity}}
\fancyhead[R]{\thepage}
\title{Lesson 66: Neural Tangent Kernel and Rademacher Complexity}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% ============================================================
%% SECTION 1: MOTIVATION AND OVERVIEW
%% ============================================================
\section{Motivation and Overview}
\label{sec:motivation}

Modern neural networks used in deep RL and supervised learning are
massively overparameterized: the number of parameters $p$ vastly
exceeds the number of training examples $n$.  Classical learning
theory (Lesson~59) predicts that such models should overfit
catastrophically---the VC dimension of a network with $p$ parameters
is $\Theta(p)$, giving a vacuous generalization bound when $p \gg n$.
Yet in practice, overparameterized networks trained by gradient
descent generalize remarkably well.  This is the \emph{generalization
puzzle} of deep learning.

Two complementary theoretical frameworks resolve (in part) this
puzzle.  First, \emph{Rademacher complexity bounds} based on weight
norms rather than parameter counts show that generalization depends
on how ``small'' the learned weights are, not on how many there are.
The spectrally-normalized bound of \citet{bartlett2017} shows that
the generalization gap scales with the product of spectral norms
$\prod_\ell \|W_\ell\|_2$ divided by $\sqrt{n}$---a quantity that
can be small even for heavily overparameterized networks.  Second,
the \emph{neural tangent kernel} (NTK) of \citet{jacot2018} shows
that in the infinite-width limit, neural network training by gradient
descent is equivalent to kernel regression with a fixed, deterministic
kernel.  This connects deep learning to the well-understood theory of
RKHS (Lesson~62), providing both optimization and generalization
guarantees.  The tools developed in this lesson will let us understand
why overparameterized networks generalize, how weight norms control
capacity, and what regime of training the NTK theory describes.

\begin{rlconnection}[title={Why NTK and Rademacher Complexity Matter
    for RL/ML}]
\begin{itemize}[nosep]
  \item \textbf{Generalization in deep RL.}  Policy and value networks
    in PPO, SAC, and DQN are typically overparameterized.  Norm-based
    Rademacher bounds explain why these networks generalize from
    limited environment interactions despite having millions of
    parameters.
  \item \textbf{Implicit regularization.}  Gradient descent on
    overparameterized networks converges to minimum-norm solutions
    (in the NTK regime), providing implicit regularization that
    controls generalization without explicit weight decay.
  \item \textbf{Neural network as kernel method.}  The NTK framework
    allows transferring kernel ridge regression theory (Lesson~62) to
    neural networks, yielding concrete convergence and generalization
    guarantees.
  \item \textbf{Architecture selection.}  Understanding that spectral
    norms---not parameter counts---control generalization informs the
    choice of network architecture and initialization scheme in
    deep RL.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
Rademacher complexity and its connection to generalization
(Lesson~59, Definition~5.1).  The approximation-estimation tradeoff
(Lesson~60, Section~3).  RKHS and kernel ridge regression
(Lesson~62, Sections~3--5).  Spectral norm and operator norm of
matrices (Lesson~8, Definition~3.1).  Gradient descent and convex
optimization (Lesson~30).  Deep neural networks and their function
classes (Lesson~65, Definition~3.1).

\paragraph{Learning objectives.}
After completing this lesson, you will be able to:
\begin{itemize}[nosep]
  \item \textbf{Define} Rademacher complexity and \textbf{prove} the
    fundamental generalization bound relating it to the uniform
    deviation of empirical risk from population risk.
  \item \textbf{State and prove} the spectrally-normalized Rademacher
    complexity bound for deep networks, and \textbf{explain} why it
    depends on weight norms rather than parameter counts.
  \item \textbf{Define} the neural tangent kernel and \textbf{derive}
    the linearized dynamics of gradient descent in the infinite-width
    limit.
  \item \textbf{Prove} that the NTK converges to a deterministic
    kernel at initialization and remains approximately constant
    during training (the lazy training regime).
  \item \textbf{Apply} NTK theory to deduce generalization bounds
    for overparameterized networks via kernel regression theory.
  \item \textbf{Evaluate} the limitations of the NTK framework,
    including its failure to capture feature learning.
\end{itemize}

%% ============================================================
%% SECTION 2: REFERENCES
%% ============================================================
\section{References}
\label{sec:references}

\begin{itemize}
  \item \citet{bartlett2002} --- Foundational paper on Rademacher
    complexity: definitions, structural results, and the
    generalization bound.
  \item \citet{bartlett2017} --- Spectrally-normalized margin bounds
    for deep networks; the key norm-based Rademacher complexity result.
  \item \citet{jacot2018} --- The neural tangent kernel: convergence
    and generalization in the infinite-width limit.
  \item \citet{lee2019} --- Wide neural networks evolve as linear
    models under gradient descent; empirical validation of the NTK
    regime.
  \item \citet{arora2019} --- Fine-grained analysis of optimization
    and generalization for overparameterized two-layer networks.
  \item \citet{du2019} --- Gradient descent provably optimizes
    overparameterized neural networks: convergence guarantees via
    the NTK.
\end{itemize}

%% ============================================================
%% SECTION 3: RADEMACHER COMPLEXITY
%% ============================================================
\section{Rademacher Complexity}
\label{sec:rademacher}

Rademacher complexity is a data-dependent measure of the richness
of a function class.  It refines the VC dimension by accounting for
the specific distribution of the data and the real-valued nature
of the functions (rather than just binary labelings).

\begin{definition}[Rademacher Random Variables]
\label{def:rademacher-rv}
A \emph{Rademacher random variable} $\varepsilon$ takes values
$+1$ and $-1$ each with probability $1/2$.  Rademacher random
variables are symmetric: $\varepsilon \stackrel{d}{=} -\varepsilon$.
\end{definition}

\begin{definition}[Empirical Rademacher Complexity]
\label{def:empirical-rademacher}
Let $\Fclass$ be a class of functions $f \colon \mathcal{X} \to \R$
and let $S = (x_1, \ldots, x_n) \in \mathcal{X}^n$ be a fixed
sample.  The \emph{empirical Rademacher complexity} of $\Fclass$
with respect to $S$ is
\[
  \hat{\Rad}_S(\Fclass) = \E_{\varepsilon}\biggl[
    \sup_{f \in \Fclass} \frac{1}{n} \sum_{i=1}^{n}
    \varepsilon_i f(x_i) \biggr],
\]
where $\varepsilon_1, \ldots, \varepsilon_n$ are i.i.d.\ Rademacher
random variables.
\end{definition}

\begin{definition}[Rademacher Complexity]
\label{def:rademacher}
The \emph{Rademacher complexity} of $\Fclass$ with respect to a
distribution $P$ on $\mathcal{X}$ and sample size $n$ is
\[
  \Rad_n(\Fclass) = \E_{S \sim P^n}\bigl[
    \hat{\Rad}_S(\Fclass) \bigr]
  = \E_{S, \varepsilon}\biggl[
    \sup_{f \in \Fclass} \frac{1}{n} \sum_{i=1}^{n}
    \varepsilon_i f(x_i) \biggr].
\]
\end{definition}

\begin{remark}
\label{rem:rademacher-interpretation}
The Rademacher complexity measures how well the function class
$\Fclass$ can \emph{correlate with random noise}.  If $\Fclass$ is
rich enough to fit arbitrary label assignments
$(\varepsilon_1, \ldots, \varepsilon_n) \in \{-1, +1\}^n$, then
$\hat{\Rad}_S(\Fclass)$ is close to $1$.  If $\Fclass$ consists only
of constant functions, then $\hat{\Rad}_S(\Fclass) = 0$.  In general,
$0 \le \hat{\Rad}_S(\Fclass) \le \sup_{f \in \Fclass} \|f\|_\infty$.
\end{remark}

\begin{example}[Rademacher Complexity of Linear Classifiers]
\label{ex:rad-linear}
Let $\Fclass = \{x \mapsto w^\top x : \|w\|_2 \le B\}$ and suppose
$\|x_i\|_2 \le C$ for all $i$.  Then
\begin{align*}
  \hat{\Rad}_S(\Fclass)
  &= \E_\varepsilon\biggl[
    \sup_{\|w\|_2 \le B} \frac{1}{n}
    w^\top \sum_{i=1}^n \varepsilon_i x_i \biggr]
  = \frac{B}{n} \E_\varepsilon\biggl[
    \biggl\| \sum_{i=1}^n \varepsilon_i x_i \biggr\|_2 \biggr] \\
  &\le \frac{B}{n} \sqrt{\E\biggl[
    \biggl\| \sum_{i=1}^n \varepsilon_i x_i \biggr\|_2^2 \biggr]}
  = \frac{B}{n} \sqrt{\sum_{i=1}^n \|x_i\|_2^2}
  \le \frac{BC}{\sqrt{n}},
\end{align*}
where we used Jensen's inequality and the fact that
$\E[\varepsilon_i \varepsilon_j] = \delta_{ij}$.  Thus
$\Rad_n(\Fclass) \le BC / \sqrt{n}$, which is independent of
the dimension $d$ of the weight vector.
\end{example}

We now prove the fundamental generalization bound via Rademacher
complexity.

\begin{theorem}[Rademacher Generalization Bound]
\label{thm:rademacher-bound}
Let $\Fclass$ be a class of functions $f \colon \mathcal{X} \to [0,1]$,
and let $(X_1, Y_1), \ldots, (X_n, Y_n)$ be i.i.d.\ from a
distribution $P$ on $\mathcal{X} \times \mathcal{Y}$.  Define the
loss class $\mathcal{G} = \{(x,y) \mapsto \Loss(f(x), y) :
f \in \Fclass\}$ where $\Loss$ takes values in $[0, 1]$.  Then
for any $\delta > 0$, with probability at least $1 - \delta$ over
the sample $S$:
\[
  \sup_{f \in \Fclass} \bigl| \Risk(f) - \EmpRisk_n(f) \bigr|
  \le 2 \Rad_n(\mathcal{G})
  + \sqrt{\frac{\ln(2/\delta)}{2n}}.
\]
If the loss $\Loss$ is $\rho$-Lipschitz in its first argument, then
$\Rad_n(\mathcal{G}) \le \rho \, \Rad_n(\Fclass)$.
\end{theorem}

\begin{proof}
\textbf{Step 1: Symmetrization.}
Define $\Phi(S) = \sup_{g \in \mathcal{G}} \bigl[
\E[g(X,Y)] - \frac{1}{n} \sum_{i=1}^n g(X_i, Y_i) \bigr]$.
We first show $\E[\Phi(S)] \le 2 \Rad_n(\mathcal{G})$.

Let $S' = (X_1', Y_1'), \ldots, (X_n', Y_n')$ be an independent
copy of $S$.  For each $g \in \mathcal{G}$,
$\E[g(X_j', Y_j')] = \E[g(X,Y)]$, so
\begin{align*}
  \E[\Phi(S)]
  &= \E_S\biggl[ \sup_{g \in \mathcal{G}} \E_{S'}\biggl[
    \frac{1}{n} \sum_{i=1}^n
    \bigl(g(X_i', Y_i') - g(X_i, Y_i)\bigr) \biggr] \biggr] \\
  &\le \E_{S, S'}\biggl[ \sup_{g \in \mathcal{G}}
    \frac{1}{n} \sum_{i=1}^n
    \bigl(g(X_i', Y_i') - g(X_i, Y_i)\bigr) \biggr].
\end{align*}
Since $g(X_i', Y_i') - g(X_i, Y_i)$ is symmetric about $0$
(the pairs $(X_i, Y_i)$ and $(X_i', Y_i')$ are exchangeable), we
can introduce Rademacher variables $\varepsilon_i$:
\begin{align*}
  \E[\Phi(S)]
  &\le \E_{S, S', \varepsilon}\biggl[ \sup_{g \in \mathcal{G}}
    \frac{1}{n} \sum_{i=1}^n \varepsilon_i
    \bigl(g(X_i', Y_i') - g(X_i, Y_i)\bigr) \biggr] \\
  &\le \E_{S', \varepsilon}\biggl[ \sup_{g \in \mathcal{G}}
    \frac{1}{n} \sum_{i=1}^n \varepsilon_i g(X_i', Y_i')
    \biggr]
  + \E_{S, \varepsilon}\biggl[ \sup_{g \in \mathcal{G}}
    \frac{1}{n} \sum_{i=1}^n (-\varepsilon_i) g(X_i, Y_i) \biggr] \\
  &= 2 \Rad_n(\mathcal{G}).
\end{align*}

\textbf{Step 2: Concentration.}
The function $\Phi(S)$ changes by at most $1/n$ when a single sample
$(X_i, Y_i)$ is replaced (since $g$ is bounded in $[0,1]$).  By
McDiarmid's inequality (a consequence of the bounded differences
inequality, see Lesson~12):
\[
  \Prob\bigl( \Phi(S) - \E[\Phi(S)] \ge t \bigr)
  \le \exp\biggl( -\frac{2n^2 t^2}{n \cdot (1/n)^2 \cdot n} \biggr)
  = \exp(-2nt^2).
\]
Setting $t = \sqrt{\ln(2/\delta)/(2n)}$ and combining with
$\E[\Phi(S)] \le 2\Rad_n(\mathcal{G})$ gives the one-sided bound.
The two-sided bound follows by applying the same argument to $-g$.

\textbf{Step 3: Lipschitz contraction.}
If $\Loss(\cdot, y)$ is $\rho$-Lipschitz for each $y$, then by
the contraction lemma (Ledoux--Talagrand):
\[
  \E_\varepsilon\biggl[ \sup_{f \in \Fclass} \frac{1}{n}
    \sum_{i=1}^n \varepsilon_i \Loss(f(x_i), y_i) \biggr]
  \le \rho \, \E_\varepsilon\biggl[ \sup_{f \in \Fclass}
    \frac{1}{n} \sum_{i=1}^n \varepsilon_i f(x_i) \biggr],
\]
giving $\Rad_n(\mathcal{G}) \le \rho \, \Rad_n(\Fclass)$.
\end{proof}

\begin{keyresult}[title={Rademacher Generalization Bound}]
For any function class $\Fclass$ with bounded loss, the uniform
deviation between population risk and empirical risk satisfies
\[
  \sup_{f \in \Fclass} \bigl| \Risk(f) - \EmpRisk_n(f) \bigr|
  \le 2 \Rad_n(\Fclass) + O\biggl(\sqrt{\frac{\log(1/\delta)}{n}}
  \biggr)
\]
with high probability.  The Rademacher complexity $\Rad_n(\Fclass)$
is a more refined measure of function class capacity than the VC
dimension, as it accounts for the distribution of the data and the
real-valued nature of the functions.
\end{keyresult}

\begin{example}[Rademacher vs.\ VC for Halfspaces]
\label{ex:rad-vs-vc}
For halfspace classifiers $\Fclass = \{x \mapsto \sign(w^\top x) :
\|w\|_2 \le B\}$ on data with $\|x\|_2 \le C$:
\begin{itemize}[nosep]
  \item \textbf{VC bound} (Lesson~59): generalization gap
    $O(\sqrt{d \log n / n})$, depending on dimension $d$.
  \item \textbf{Rademacher bound}: generalization gap
    $O(BC / \sqrt{n})$, independent of dimension $d$.
\end{itemize}
The Rademacher bound is tighter when $BC$ is small (i.e., when
the margin is large), which is precisely the regime where SVMs
work well.
\end{example}

Having established the Rademacher framework, we now apply it to
deep neural networks.

%% ============================================================
%% SECTION 4: RADEMACHER COMPLEXITY OF NEURAL NETWORKS
%% ============================================================
\section{Rademacher Complexity of Neural Networks}
\label{sec:rad-nn}

The key result of this section is that the Rademacher complexity
of deep networks depends on the \emph{spectral norms} of the weight
matrices, not on the number of parameters.  This explains why
overparameterized networks can generalize.

\subsection{The Contraction Lemma}

\begin{lemma}[Contraction Lemma (Ledoux--Talagrand)]
\label{lem:contraction}
Let $\Fclass$ be a function class and let $\phi_1, \ldots, \phi_n
\colon \R \to \R$ be $\rho$-Lipschitz functions with $\phi_i(0) = 0$.
Then
\[
  \E_\varepsilon\biggl[ \sup_{f \in \Fclass} \frac{1}{n}
    \sum_{i=1}^n \varepsilon_i \phi_i(f(x_i)) \biggr]
  \le \rho \, \E_\varepsilon\biggl[ \sup_{f \in \Fclass}
    \frac{1}{n} \sum_{i=1}^n \varepsilon_i f(x_i) \biggr].
\]
\end{lemma}

\begin{proof}
This is a classical result; see \citet[Theorem~4.12]{ledoux1991}.
The proof proceeds by induction on $n$.  For $n = 1$: by the
Lipschitz condition,
$\sup_{t \in T} \varepsilon \phi(t) \le \rho \sup_{t \in T}
\varepsilon t$ for any index set $T$ (where $T = \{f(x_1) :
f \in \Fclass\}$).  Taking expectations over $\varepsilon$
gives the result.  The inductive step conditions on
$\varepsilon_1, \ldots, \varepsilon_{n-1}$ and applies the
$n = 1$ case to $\varepsilon_n$.
\end{proof}

\begin{remark}
\label{rem:relu-contraction}
The ReLU activation $\sigmoid(t) = \max(0, t)$ is $1$-Lipschitz with
$\sigmoid(0) = 0$.  Therefore, applying a ReLU layer does not increase
the Rademacher complexity.  This is the key reason why depth, per se,
does not inflate the Rademacher complexity---only the weight norms
matter.
\end{remark}

\subsection{Spectrally-Normalized Bound}

\begin{theorem}[Bartlett, Foster, and Telgarsky, 2017]
\label{thm:spectral-rad}
Consider an $L$-layer ReLU network (Definition~3.1 in Lesson~65):
$f(x) = W_L \sigmoid(W_{L-1} \sigmoid(\cdots \sigmoid(W_1 x)
\cdots))$
with weight matrices $W_\ell \in \R^{n_\ell \times n_{\ell-1}}$
(biases omitted for clarity).  Let $\Fclass$ be the class of all
such networks with $\|W_\ell\|_2 \le s_\ell$ (spectral norm bound)
for each layer $\ell$.  If $\|x_i\|_2 \le B_x$ for all data points,
then the empirical Rademacher complexity satisfies
\begin{equation}
\label{eq:spectral-bound}
  \hat{\Rad}_S(\Fclass) \le
  \frac{B_x \cdot \prod_{\ell=1}^{L} s_\ell}{\sqrt{n}}
  \cdot O\biggl( \sqrt{L \log(W)} \biggr),
\end{equation}
where $W = \max_\ell n_\ell$ is the maximum width.
\end{theorem}

\begin{proof}
We prove a simplified version of the bound, establishing the product
structure; the logarithmic factors are refined in
\citet{bartlett2017}.

\textbf{Step 1: Peel off the output layer.}
Define $h(x) = \sigmoid(W_{L-1} \sigmoid(\cdots \sigmoid(W_1 x)
\cdots))$ so that $f(x) = W_L h(x)$.  The Rademacher complexity of
the composition with the linear map $W_L$ satisfies
\[
  \hat{\Rad}_S(\{x \mapsto W_L h(x)\})
  \le \|W_L\|_2 \cdot \hat{\Rad}_S(\{x \mapsto h(x)\}),
\]
by viewing $W_L$ as a linear operator and using the definition of
the spectral norm.  More precisely, for scalar output ($n_L = 1$),
$W_L$ is a row vector with $\|W_L\|_2 = \|W_L\|_F$, and
\begin{align*}
  \E_\varepsilon\biggl[\sup_{f \in \Fclass} \frac{1}{n}
    \sum_i \varepsilon_i W_L h(x_i)\biggr]
  &= \E_\varepsilon\biggl[\sup_{\|W_L\|_2 \le s_L} W_L \cdot
    \sup_{h \in \Hclass_{L-1}} \frac{1}{n}
    \sum_i \varepsilon_i h(x_i)\biggr] \\
  &\le s_L \cdot \E_\varepsilon\biggl[\sup_{h \in \Hclass_{L-1}}
    \frac{1}{n} \biggl\|\sum_i \varepsilon_i h(x_i)\biggr\|_2
    \biggr],
\end{align*}
where $\Hclass_{L-1}$ is the class of functions computed by the first
$L-1$ layers.

\textbf{Step 2: Peel off the activation.}
The ReLU activation is $1$-Lipschitz with $\sigmoid(0) = 0$.  By
the contraction lemma (Lemma~\ref{lem:contraction}), replacing
$\sigmoid(W_\ell \cdot)$ with $W_\ell \cdot$ does not increase the
Rademacher complexity (applied coordinatewise).

\textbf{Step 3: Iterate.}
Repeating Steps 1 and 2 for layers $L-1, L-2, \ldots, 1$, each
peeling contributes a factor of $s_\ell = \|W_\ell\|_2$.  The
base case is the input layer, where the Rademacher complexity of
$\{x \mapsto x\}$ with respect to data bounded by $B_x$ is
$B_x / \sqrt{n}$ (by Example~\ref{ex:rad-linear} with $B = 1$).

\textbf{Step 4: Combine.}
The product of the spectral norm factors gives
\[
  \hat{\Rad}_S(\Fclass) \le
  \frac{B_x \cdot \prod_{\ell=1}^{L} s_\ell}{\sqrt{n}}
  \cdot C(L, W),
\]
where $C(L, W)$ accounts for the vector-valued nature of the
intermediate layers (the Rademacher complexity of vector-valued
functions picks up a factor of $\sqrt{\log W}$ per layer from
covering number arguments).  Careful analysis in
\citet{bartlett2017} shows $C(L, W) = O(\sqrt{L \log W})$.
\end{proof}

\begin{keyresult}[title={Norm-Based Generalization for Deep Networks}]
The generalization gap of a deep ReLU network is controlled by
\[
  \text{Generalization gap} \le O\biggl(
  \frac{B_x \cdot \prod_{\ell=1}^L \|W_\ell\|_2 \cdot
  \sqrt{L \log W}}{\sqrt{n}} \biggr),
\]
which depends on the \textbf{spectral norms} of the weight matrices
and the \textbf{data norm} $B_x$, but is \textbf{independent of the
number of parameters}.  An overparameterized network with small
spectral norms can generalize well even when the number of parameters
far exceeds the sample size.
\end{keyresult}

\begin{example}[Parameter-Free Bound for a DQN]
\label{ex:dqn-bound}
Consider a DQN with $L = 5$ layers, width $W = 256$, and spectral
norms $\|W_\ell\|_2 \le 2$ for each layer.  The total parameter count
is roughly $5 \times 256^2 \approx 327{,}680$.  With $n = 10{,}000$
training transitions and $B_x = 1$:
\begin{itemize}[nosep]
  \item \textbf{VC bound:} $O(\sqrt{327{,}680 / 10{,}000})
    \approx 5.7$ --- vacuous (the bound exceeds $1$).
  \item \textbf{Spectral bound:} $O(2^5 \cdot \sqrt{5 \log 256}
    / \sqrt{10{,}000}) = O(32 \cdot 3.1 / 100) \approx 1.0$ ---
    borderline, but dramatically tighter than the VC bound.
\end{itemize}
With spectral normalization or weight decay to keep $\|W_\ell\|_2$
close to $1$, the bound improves to $O(0.03)$---a meaningful guarantee.
\end{example}

\begin{rlconnection}[title={Spectral Normalization in Deep RL}]
The spectral norm bound motivates the use of \emph{spectral
normalization} in deep RL:
\begin{itemize}[nosep]
  \item In DDPG and TD3, spectral normalization of the critic
    network stabilizes training and improves generalization.
  \item The bound $\prod_\ell \|W_\ell\|_2$ shows that controlling
    the product of spectral norms (not the individual norms) is
    what matters for generalization.
  \item Weight decay ($\ell^2$ regularization) implicitly controls
    spectral norms, providing a connection between the common
    practice of regularization and the theoretical bound.
\end{itemize}
\end{rlconnection}

\begin{warning}[title={Norm-Based Bounds Can Still Be Loose}]
While spectrally-normalized bounds are much tighter than VC bounds
for overparameterized networks, they are still often \emph{vacuous}
for practical networks (the bound exceeds $1$ for classification).
The product $\prod_\ell \|W_\ell\|_2$ grows exponentially with depth
unless each spectral norm is close to $1$.  Tighter bounds require
margin-based analysis or PAC-Bayesian arguments; see
\citet{neyshabur2015} for further developments.
\end{warning}

The norm-based bounds explain generalization in terms of weight
properties.  The next section provides a complementary perspective:
the neural tangent kernel explains the \emph{dynamics} of training
and connects overparameterized networks to kernel methods.

%% ============================================================
%% SECTION 5: THE NEURAL TANGENT KERNEL
%% ============================================================
\section{The Neural Tangent Kernel}
\label{sec:ntk}

The neural tangent kernel (NTK) reveals that in the infinite-width
limit, training a neural network by gradient descent is equivalent
to performing kernel regression with a specific, deterministic kernel.

\subsection{Definition and Motivation}

\begin{definition}[Neural Tangent Kernel]
\label{def:ntk}
Let $f(x; \theta) \colon \R^d \to \R$ be a neural network with
parameters $\theta \in \R^p$.  The \emph{neural tangent kernel}
(NTK) is the kernel function
\begin{equation}
\label{eq:ntk-def}
  \Theta(x, x'; \theta) = \bigl\langle
    \nabla_\theta f(x; \theta), \;
    \nabla_\theta f(x'; \theta) \bigr\rangle
  = \sum_{j=1}^{p} \frac{\partial f(x; \theta)}{\partial \theta_j}
    \cdot \frac{\partial f(x'; \theta)}{\partial \theta_j},
\end{equation}
where $\nabla_\theta f(x; \theta) \in \R^p$ is the gradient of the
network output with respect to all parameters.
\end{definition}

\begin{remark}
\label{rem:ntk-feature}
The NTK can be viewed as an inner product in a feature space: define
$\feature_\theta(x) = \nabla_\theta f(x; \theta) \in \R^p$, so that
$\Theta(x, x'; \theta) = \langle \feature_\theta(x),
\feature_\theta(x') \rangle$.  This is a kernel in the sense of
Lesson~61, but with a feature map that depends on the parameters
$\theta$.
\end{remark}

\begin{example}[NTK for a Two-Layer Network]
\label{ex:ntk-two-layer}
Consider a two-layer network with $m$ hidden units:
\[
  f(x; \theta) = \frac{1}{\sqrt{m}} \sum_{j=1}^{m}
    a_j \sigmoid(w_j^\top x),
\]
where $\theta = (a_1, \ldots, a_m, w_1, \ldots, w_m)$, the
$1/\sqrt{m}$ scaling ensures finite limits as $m \to \infty$, and
$\sigmoid$ is a smooth activation.  The gradients are:
\begin{align*}
  \frac{\partial f}{\partial a_j} &= \frac{1}{\sqrt{m}}
    \sigmoid(w_j^\top x), \\
  \frac{\partial f}{\partial w_j} &= \frac{1}{\sqrt{m}}
    a_j \sigmoid'(w_j^\top x) \, x.
\end{align*}
The NTK is
\begin{align}
  \Theta(x, x'; \theta) &= \frac{1}{m} \sum_{j=1}^{m}
    \sigmoid(w_j^\top x) \sigmoid(w_j^\top x')
    \label{eq:ntk-a-part} \\
  &\quad + \frac{1}{m} \sum_{j=1}^{m} a_j^2
    \sigmoid'(w_j^\top x) \sigmoid'(w_j^\top x')
    \, (x^\top x'). \label{eq:ntk-w-part}
\end{align}
The first term \eqref{eq:ntk-a-part} is the kernel induced by the
output-layer gradients; the second term \eqref{eq:ntk-w-part}
captures the hidden-layer gradients.
\end{example}

\subsection{NTK Dynamics of Gradient Descent}

\begin{proposition}[Gradient Descent in Function Space]
\label{prop:gd-dynamics}
Consider training the network $f(x; \theta)$ on data
$(x_1, y_1), \ldots, (x_n, y_n)$ by minimizing the squared loss
$\mathcal{L}(\theta) = \frac{1}{2} \sum_{i=1}^{n}
(f(x_i; \theta) - y_i)^2$
with continuous-time gradient descent $\dot{\theta}(t) =
-\nabla_\theta \mathcal{L}(\theta(t))$.  Then the network predictions
$u_i(t) = f(x_i; \theta(t))$ evolve according to
\begin{equation}
\label{eq:gd-dynamics}
  \dot{u}_i(t) = -\sum_{j=1}^{n}
    \Theta(x_i, x_j; \theta(t)) \, (u_j(t) - y_j),
\end{equation}
or in matrix form, $\dot{u}(t) = -K(t)(u(t) - y)$, where
$K(t) \in \R^{n \times n}$ is the \emph{NTK Gram matrix}
$K_{ij}(t) = \Theta(x_i, x_j; \theta(t))$.
\end{proposition}

\begin{proof}
By the chain rule,
\[
  \dot{u}_i(t) = \frac{d}{dt} f(x_i; \theta(t))
  = \nabla_\theta f(x_i; \theta(t))^\top \dot{\theta}(t).
\]
The gradient descent dynamics give
\[
  \dot{\theta}(t) = -\nabla_\theta \mathcal{L}(\theta(t))
  = -\sum_{j=1}^{n} (f(x_j; \theta(t)) - y_j)
    \nabla_\theta f(x_j; \theta(t)).
\]
Substituting:
\begin{align*}
  \dot{u}_i(t)
  &= -\sum_{j=1}^{n} (u_j(t) - y_j)
    \nabla_\theta f(x_i; \theta(t))^\top
    \nabla_\theta f(x_j; \theta(t)) \\
  &= -\sum_{j=1}^{n} \Theta(x_i, x_j; \theta(t))
    (u_j(t) - y_j). \qedhere
\end{align*}
\end{proof}

\begin{remark}
\label{rem:kernel-regression-ode}
If the NTK Gram matrix $K(t)$ were \emph{constant} (independent of
$t$), then \eqref{eq:gd-dynamics} would be a linear ODE with solution
$u(t) - y = e^{-Kt}(u(0) - y)$.  If $K$ is positive definite, all
eigenvalues are positive and $u(t) \to y$ exponentially fast: the
training loss converges to zero.  This motivates the central question
of NTK theory: under what conditions does the NTK remain approximately
constant during training?
\end{remark}

\subsection{The Infinite-Width Limit}

\begin{theorem}[NTK Convergence at Initialization --- Jacot et al., 2018]
\label{thm:ntk-init}
Consider a fully-connected network of depth $L$ with layer widths
$n_1 = n_2 = \cdots = n_{L-1} = m$ (hidden widths all equal to $m$),
smooth activation $\sigmoid$, and random initialization: each weight
$W_\ell^{(ij)} \sim \mathcal{N}(0, 2/n_{\ell-1})$ (He initialization).
Then as $m \to \infty$:
\begin{enumerate}[nosep,label=(\roman*)]
  \item The NTK at initialization $\Theta(x, x'; \theta_0)$ converges
    in probability to a deterministic limit $\Theta_\infty^{(L)}(x, x')$
    that depends only on $x, x'$ and the network architecture (depth
    $L$ and activation $\sigmoid$), not on the random initialization.
  \item The limiting kernel $\Theta_\infty^{(L)}$ can be computed
    recursively.  Define the \emph{NNGP kernel} (neural network Gaussian
    process kernel) by $\Sigma^{(0)}(x, x') = x^\top x'$ and
    \begin{equation}
    \label{eq:nngp-recursion}
      \Sigma^{(\ell)}(x, x') = \E_{(u, v) \sim \mathcal{N}(0,
        \Lambda^{(\ell)})} \bigl[ \sigmoid(u) \sigmoid(v) \bigr],
    \end{equation}
    where $\Lambda^{(\ell)} = \bigl( \begin{smallmatrix}
    \Sigma^{(\ell-1)}(x,x) & \Sigma^{(\ell-1)}(x,x') \\
    \Sigma^{(\ell-1)}(x',x) & \Sigma^{(\ell-1)}(x',x')
    \end{smallmatrix} \bigr)$.  Then $\Theta_\infty^{(L)}$ is
    determined by the $\Sigma^{(\ell)}$ kernels and their
    derivative counterparts
    $\dot{\Sigma}^{(\ell)}(x, x') = \E[\sigmoid'(u)\sigmoid'(v)]$.
\end{enumerate}
\end{theorem}

\begin{proof}[Proof sketch]
(i) At initialization, the pre-activations
$h_\ell^{(j)} = \sum_k W_\ell^{(jk)} \sigmoid(h_{\ell-1}^{(k)})$
are sums of $m$ independent random variables (scaled by $1/\sqrt{m}$
via the initialization).  By the central limit theorem, as $m \to
\infty$, the pre-activations at each layer converge jointly to a
Gaussian process.  The covariance of this Gaussian process is the
NNGP kernel $\Sigma^{(\ell)}$, defined by the recursion
\eqref{eq:nngp-recursion}.

The NTK $\Theta(x, x'; \theta_0) = \sum_{j=1}^p
\frac{\partial f(x)}{\partial \theta_j}
\frac{\partial f(x')}{\partial \theta_j}$ is also a sum of $O(m)$
terms (one for each parameter).  By the law of large numbers (in the
$m \to \infty$ limit), this sum converges to its expectation, which
is a deterministic function of $\Sigma^{(\ell)}$ and
$\dot{\Sigma}^{(\ell)}$.

(ii) The recursive formula follows from the layer-by-layer structure
of the network.  The output-layer contribution to the NTK is
$\Sigma^{(L-1)}(x, x')$ (from the gradients with respect to
$W_L$), and the hidden-layer contributions involve
$\dot{\Sigma}^{(\ell)}$ (from the chain rule through the
activations).  The full recursive formula is given in
\citet{jacot2018}.
\end{proof}

\begin{example}[NTK for a Two-Layer Network with ReLU]
\label{ex:ntk-relu}
For a two-layer ReLU network with He initialization and $m$ hidden
units, the limiting NTK as $m \to \infty$ is
\[
  \Theta_\infty(x, x') = \frac{\|x\|_2 \|x'\|_2}{2\pi}
    \bigl( \sin\alpha + (\pi - \alpha) \cos\alpha \bigr)
    + \frac{x^\top x'}{2\pi} (\pi - \alpha),
\]
where $\alpha = \arccos\bigl( \frac{x^\top x'}{\|x\|_2 \|x'\|_2}
\bigr)$ is the angle between $x$ and $x'$.  This is the ``arc-cosine
kernel'' studied in the kernel methods literature.
\end{example}

\subsection{Lazy Training: NTK Stays Constant}

\begin{theorem}[Lazy Training Regime]
\label{thm:lazy}
Under the same setup as Theorem~\ref{thm:ntk-init}, if the width
$m$ is sufficiently large (specifically, $m = \Omega(n^4 / \lambda_0^4)$
where $\lambda_0 > 0$ is the smallest eigenvalue of the limiting NTK
Gram matrix $K_\infty \in \R^{n \times n}$), then with high
probability over the random initialization:
\begin{enumerate}[nosep,label=(\roman*)]
  \item The NTK Gram matrix $K(t)$ stays close to its initial value
    throughout training:
    $\|K(t) - K(0)\|_F \le \varepsilon$ for all $t \ge 0$, where
    $\varepsilon \to 0$ as $m \to \infty$.
  \item The parameters stay close to their initialization:
    $\|\theta(t) - \theta_0\|_2 = O(1/\sqrt{m})$ for all $t$.
  \item The training loss converges to zero exponentially fast:
    $\mathcal{L}(\theta(t)) \le e^{-\lambda_0 t} \mathcal{L}(\theta_0)$.
  \item The network function is approximately linear in the
    parameters:
    \begin{equation}
    \label{eq:linearized}
      f(x; \theta(t)) \approx f(x; \theta_0) +
      \nabla_\theta f(x; \theta_0)^\top (\theta(t) - \theta_0).
    \end{equation}
\end{enumerate}
\end{theorem}

\begin{proof}
\textbf{(i) and (ii):} The gradient $\nabla_\theta f(x; \theta)$ is
$O(1/\sqrt{m})$ in each coordinate (by the scaling of the
initialization).  The gradient descent update moves $\theta$ by
$\Delta\theta = -\eta \nabla_\theta \mathcal{L}(\theta)$, which
involves $\nabla_\theta f(x_i; \theta) \cdot (u_i - y_i)$.  Since
$\nabla_\theta f$ has $O(m)$ coordinates each of size $O(1/\sqrt{m})$,
the total movement $\|\theta(t) - \theta_0\|_2 = O(\sqrt{n}/\sqrt{m})$,
which is $O(1/\sqrt{m})$ for fixed $n$.

The NTK involves second-order information: $\Theta(x, x'; \theta)
- \Theta(x, x'; \theta_0)$ is bounded by the Hessian of $f$ times
$\|\theta - \theta_0\|$, which is $O(1/\sqrt{m})$ times
$O(1/\sqrt{m})$, giving $O(1/m)$ change.  This shows the NTK
remains approximately constant.

\textbf{(iii):}  With $K(t) \approx K(0) \approx K_\infty$, the
dynamics \eqref{eq:gd-dynamics} become approximately linear:
$\dot{u}(t) \approx -K_\infty(u(t) - y)$, with solution
$u(t) - y = e^{-K_\infty t}(u(0) - y)$.  Since $K_\infty \succeq
\lambda_0 I$, we have $\|u(t) - y\|_2^2 \le e^{-2\lambda_0 t}
\|u(0) - y\|_2^2$.

\textbf{(iv):}  The linearization \eqref{eq:linearized} follows from
Taylor expansion: $f(x; \theta) = f(x; \theta_0) +
\nabla_\theta f(x; \theta_0)^\top (\theta - \theta_0) + O(\|\theta
- \theta_0\|^2)$, and $\|\theta - \theta_0\|^2 = O(1/m) \to 0$.
\end{proof}

\begin{keyresult}[title={NTK Theory: Training = Kernel Regression}]
In the infinite-width limit, training a neural network by gradient
descent is equivalent to kernel regression with the neural tangent
kernel $\Theta_\infty$:
\begin{enumerate}[nosep]
  \item The NTK converges to a deterministic kernel at initialization.
  \item The NTK stays approximately constant during training (lazy
    training).
  \item The network dynamics are approximately linear: $f(x; \theta_t)
    \approx f(x; \theta_0) + \nabla_\theta f(x; \theta_0)^\top
    (\theta_t - \theta_0)$.
  \item Training converges to zero loss at an exponential rate
    $e^{-\lambda_0 t}$ determined by the smallest eigenvalue of the
    NTK Gram matrix.
\end{enumerate}
\end{keyresult}

\begin{example}[Linearized Network as Random Features]
\label{ex:linearized-rf}
The linearized model \eqref{eq:linearized} can be written as
\[
  f_{\text{lin}}(x; \alpha) = f(x; \theta_0) +
  \sum_{j=1}^{p} \alpha_j \frac{\partial f(x; \theta_0)}
  {\partial \theta_j},
\]
where $\alpha = \theta - \theta_0$ are the trainable parameters.
This is a \emph{random features model} (Lesson~64,
Lesson~64, Section~4.2): the features
$\phi_j(x) = \partial f(x; \theta_0) / \partial \theta_j$ are
fixed at initialization, and only the linear coefficients $\alpha_j$
are learned.  The associated kernel is precisely the NTK:
$\Theta(x, x') = \sum_j \phi_j(x) \phi_j(x')$.
\end{example}

We now connect the NTK framework to generalization theory.

%% ============================================================
%% SECTION 6: IMPLICATIONS FOR GENERALIZATION
%% ============================================================
\section{Implications for Generalization}
\label{sec:generalization}

\subsection{NTK Generalization via Kernel Theory}

In the lazy training regime (Theorem~\ref{thm:lazy}), the trained
network computes approximately the same function as the kernel
regression estimator with kernel $\Theta_\infty$.  This allows us
to transfer the rich generalization theory of kernel methods
(Lesson~62) to neural networks.

\begin{proposition}[Generalization of the NTK Estimator]
\label{prop:ntk-generalization}
In the lazy training regime, the trained network $f(\cdot; \theta_T)$
(after $T$ gradient descent steps or continuous-time training)
satisfies the same generalization bounds as kernel ridge regression
with kernel $\Theta_\infty$ and regularization $\lambda \propto 1/T$.
Specifically, if the target function $f^*$ lies in the RKHS
$\mathcal{H}_{\Theta_\infty}$ with RKHS norm
$\|f^*\|_{\mathcal{H}_{\Theta_\infty}} \le B$, then the population
risk of the trained network satisfies
\[
  \Risk(f(\cdot; \theta_T)) - \Risk(f^*)
  \le O\biggl( \frac{B^2}{n} + \frac{B^2}{T} \biggr).
\]
The first term is the statistical error (from $n$ samples) and the
second is the optimization error (from $T$ training steps).
\end{proposition}

\begin{proof}[Proof sketch]
In the lazy regime, $f(\cdot; \theta_T) \approx f_{\text{lin}}
(\cdot; \alpha_T)$ where $\alpha_T$ is obtained by gradient descent
on the linearized model with features $\nabla_\theta f(\cdot;
\theta_0)$.  This linearized model is a kernel method with kernel
$\Theta_\infty$, and its generalization properties follow from
standard kernel ridge regression theory (Lesson~62, Section~5).
The gradient descent with $T$ steps on the squared loss is equivalent
to kernel ridge regression with implicit regularization
$\lambda \propto 1/T$ (early stopping regularization).  The
generalization bound then follows from the bias-variance decomposition
for kernel ridge regression.
\end{proof}

\begin{example}[NTK Generalization for Overparameterized Networks]
\label{ex:ntk-gen}
Consider a two-layer ReLU network with $m = 10{,}000$ hidden units
trained on $n = 100$ data points in $\R^{10}$.  The network has
$p = 10{,}000 \times 11 + 10{,}001 \approx 120{,}000$ parameters
($p \gg n$).

\begin{itemize}[nosep]
  \item \textbf{VC bound:} $O(\sqrt{p/n}) = O(\sqrt{1200}) \approx
    35$ --- completely vacuous.
  \item \textbf{NTK bound:} If $f^*$ lies in the NTK RKHS with
    $\|f^*\|_{\mathcal{H}_\Theta} \le B = 5$, then the excess risk
    is $O(B^2/n) = O(25/100) = O(0.25)$ --- a meaningful guarantee,
    independent of $m$.
\end{itemize}
The NTK framework shows that overparameterization ($m \gg n$) does
not hurt generalization, because the trained network implements
kernel regression with a fixed kernel.
\end{example}

\subsection{Limitations of the NTK Framework}

\begin{remark}[The Feature Learning Gap]
\label{rem:feature-learning}
The NTK theory has a fundamental limitation: it describes the
\emph{lazy training} regime where the network stays close to its
initialization and the features $\nabla_\theta f(\cdot; \theta_0)$
remain essentially fixed.  This means the network does not
\emph{learn features}---it uses the random features from initialization.

In practice, the power of deep learning comes precisely from
\emph{feature learning}: the network discovers useful representations
in the hidden layers.  This happens in the ``rich'' or ``active''
training regime, where the parameters move far from initialization
and the NTK changes significantly during training.  The NTK theory
does not capture this regime.
\end{remark}

\begin{proposition}[NTK vs.\ Finite-Width Networks]
\label{prop:ntk-limitations}
The following gaps exist between the NTK regime and practical neural
network training:
\begin{enumerate}[nosep,label=(\roman*)]
  \item \textbf{Sample complexity.}  Kernel methods (and hence
    NTK-regime networks) require $n = \Omega(d^s)$ samples to learn
    $s$-smooth functions on $\R^d$, suffering the curse of
    dimensionality.  Finite-width networks with feature learning can
    potentially avoid this curse (Lesson~65,
    Lesson~65, Section~6).
  \item \textbf{Representation.}  The RKHS of the NTK may not contain
    functions that finite-width networks can efficiently represent.
    The NTK kernel is fixed at initialization and does not adapt to
    the data, while the effective kernel of a finite-width network
    evolves during training.
  \item \textbf{Depth benefit.}  In the NTK regime, depth provides
    only a polynomial (not exponential) improvement in the kernel
    eigenvalues, while finite-width deep networks exhibit exponential
    depth separation (Lesson~65, Theorem~4.5).
\end{enumerate}
\end{proposition}

\begin{proof}[Justification]
(i) follows from the minimax rates for kernel regression in the RKHS
of a fixed kernel; see \citet{bietti2019} for a detailed analysis of
NTK eigenvalues and their implications for learning rates.

(ii) The RKHS of the infinite-width NTK is a fixed function space
determined by the kernel eigenvalues.  Feature learning corresponds
to adapting the kernel to the data, which the NTK regime precludes.

(iii) \citet{bietti2019} shows that the eigenvalue decay of the
NTK for depth-$L$ networks improves only polynomially with $L$,
while the depth separation results of \citet{telgarsky2016} show
exponential gains for finite-width networks.
\end{proof}

\begin{rlconnection}[title={Beyond the NTK in Deep RL}]
The NTK theory provides a useful theoretical baseline but does not
fully explain deep RL success:
\begin{itemize}[nosep]
  \item \textbf{Feature learning is crucial.}  In Atari games and
    robotic control, the learned representations in hidden layers
    are essential for good performance.  A randomly initialized
    feature map (as in the NTK regime) would miss task-relevant
    structure.
  \item \textbf{Non-stationary targets.}  In RL, the training targets
    (Bellman backup targets) change during training, violating the
    fixed-target assumption of NTK theory.
  \item \textbf{Mean-field theory.}  An alternative to the NTK
    (which linearizes around initialization) is the mean-field theory
    (Lesson~64, Section~6), which tracks the evolution of the
    weight distribution and captures feature learning in the
    large-width limit.
\end{itemize}
\end{rlconnection}

\begin{warning}[title={NTK Does Not Explain Deep Learning}]
While the NTK provides valuable theoretical insights, the consensus
in the research community is that the lazy training regime does not
capture the behavior of practical deep networks.  \citet{zhang2021}
show that standard networks memorize random labels (consistent with
the NTK regime) but also generalize on structured data (which
requires feature learning beyond the NTK regime).  The NTK is best
understood as a \emph{limiting case} that is analytically tractable
but does not capture the full power of deep learning.
\end{warning}

%% ============================================================
%% SECTION 7: EXERCISES
%% ============================================================
\section{Exercises}
\label{sec:exercises}

\begin{exercise}[Rademacher Complexity of Bounded Linear Functions]
\textnormal{(\({}^*\))} \\
Let $\Fclass = \{x \mapsto w^\top x : \|w\|_1 \le B\}$ and
$\|x_i\|_\infty \le C$ for all $i$.  Show that
$\hat{\Rad}_S(\Fclass) \le BC \sqrt{2 \log(2d) / n}$.

\textit{Hint:} Use $\sup_{\|w\|_1 \le B} w^\top z = B \|z\|_\infty$
and the fact that $\E[\max_{j \le d} Z_j] \le \sqrt{2 \log d} \cdot
\max_j (\E[Z_j^2])^{1/2}$ for sub-Gaussian $Z_j$.
\end{exercise}

\begin{exercise}[Computing the NTK for a Simple Network]
\textnormal{(\({}^*\))} \\
Consider a two-layer network $f(x; a, w) = a \sigmoid(wx + b)$ with
scalar input $x \in \R$, parameters $\theta = (a, w, b)$, and smooth
activation $\sigmoid$.  Compute the NTK
$\Theta(x, x'; \theta) = \nabla_\theta f(x)^\top \nabla_\theta f(x')$
explicitly.
\end{exercise}

\begin{exercise}[Spectral Norm Bound for a Specific Network]
\textnormal{(\({}^{**}\))} \\
Consider a $3$-layer ReLU network on $\R^{10}$ with layer widths
$(10, 50, 50, 1)$ and weight matrices $W_1, W_2, W_3$.
\begin{enumerate}[label=(\alph*)]
  \item Suppose $\|W_1\|_2 = 1.5$, $\|W_2\|_2 = 1.2$,
    $\|W_3\|_2 = 0.8$, and $B_x = 1$.  Compute the spectrally-
    normalized Rademacher complexity bound from
    Theorem~\ref{thm:spectral-rad} (ignoring the logarithmic factor).
  \item How many training samples $n$ are needed for the generalization
    gap to be at most $0.1$?
  \item The network has $10 \times 50 + 50 \times 50 + 50 \times 1 =
    3{,}050$ parameters.  Compare the VC-based sample complexity with
    the norm-based sample complexity from part (b).
\end{enumerate}
\end{exercise}

\begin{exercise}[NTK Gram Matrix and Training Convergence]
\textnormal{(\({}^{**}\))} \\
Consider training data $(x_1, y_1), (x_2, y_2)$ with the NTK Gram
matrix $K = \bigl( \begin{smallmatrix} 2 & 1 \\ 1 & 2
\end{smallmatrix} \bigr)$.
\begin{enumerate}[label=(\alph*)]
  \item Compute the eigenvalues of $K$ and verify that $K$ is
    positive definite.
  \item Using the continuous-time gradient descent solution
    $u(t) - y = e^{-Kt}(u(0) - y)$, compute $u(t)$ for
    $u(0) = (0, 0)^\top$ and $y = (1, -1)^\top$.
  \item How long does it take for the training loss
    $\|u(t) - y\|_2^2$ to decrease to $1\%$ of its initial value?
  \item If the NTK Gram matrix were $K' = \bigl(
    \begin{smallmatrix} 2 & 1.99 \\ 1.99 & 2 \end{smallmatrix}
    \bigr)$, how would the convergence rate change?  Relate this
    to the condition number.
\end{enumerate}
\end{exercise}

\begin{exercise}[Rademacher Complexity and Value Function Generalization]
\textnormal{(\({}^{**}\))} \\
In fitted Q-iteration, the Q-function is approximated by a $3$-layer
ReLU network with spectral norms $\|W_\ell\|_2 \le s$ for each layer.
States satisfy $\|s\|_2 \le 1$ and rewards are bounded by $1$.
\begin{enumerate}[label=(\alph*)]
  \item Using Theorem~\ref{thm:spectral-rad}, bound the Rademacher
    complexity of the Q-network class.
  \item Apply Theorem~\ref{thm:rademacher-bound} to bound the
    generalization gap of each Bellman backup step.
  \item If $K$ backups are performed and the discount factor is
    $\gamma$, bound the total error of fitted Q-iteration.
\end{enumerate}
\end{exercise}

\begin{exercise}[NTK Feature Learning Gap]
\textnormal{(\({}^{***}\))} \\
Consider learning the function $f^*(x) = \|Ax\|_2$ on $\R^d$ where
$A \in \R^{k \times d}$ is an unknown rank-$k$ matrix with $k \ll d$.
\begin{enumerate}[label=(\alph*)]
  \item Show that $f^*$ lies in a $k$-dimensional subspace: there
    exists a feature map $\phi(x) = Ax \in \R^k$ such that $f^*$
    depends on $x$ only through $\phi(x)$.
  \item Argue that a neural network that learns the features $\phi$
    (feature learning regime) needs $O(k)$ effective parameters to
    represent $f^*$, while the NTK regime (random features)
    requires $O(d^s)$ parameters for $s$-smooth functions.
  \item Conclude that for $k \ll d$, the feature learning regime
    is exponentially more sample-efficient than the NTK regime for
    this target function.
\end{enumerate}
\end{exercise}

%% ============================================================
%% SECTION 8: SUMMARY
%% ============================================================
\section{Summary}
\label{sec:summary}

\begin{itemize}
  \item \textbf{Rademacher complexity}
    $\Rad_n(\Fclass) = \E[\sup_{f \in \Fclass} \frac{1}{n}
    \sum_i \varepsilon_i f(x_i)]$ measures how well a function
    class can correlate with random noise.  It provides a more
    refined capacity measure than VC dimension, accounting for the
    data distribution and real-valued functions.

  \item The \textbf{fundamental Rademacher bound}: the uniform
    deviation $\sup_f |\Risk(f) - \EmpRisk_n(f)|$ is bounded by
    $2\Rad_n(\Fclass) + O(\sqrt{\log(1/\delta)/n})$ with high
    probability.

  \item For deep ReLU networks, the \textbf{spectrally-normalized
    bound} gives
    $\Rad_n \le O(B_x \prod_\ell \|W_\ell\|_2 \cdot
    \sqrt{L \log W} / \sqrt{n})$.  This depends on
    \textbf{weight norms, not parameter counts}, explaining why
    overparameterized networks can generalize.

  \item The \textbf{neural tangent kernel} $\Theta(x, x') =
    \langle \nabla_\theta f(x), \nabla_\theta f(x') \rangle$
    governs the dynamics of gradient descent on neural networks:
    $\dot{u}(t) = -K(t)(u(t) - y)$ where $K_{ij}(t) =
    \Theta(x_i, x_j; \theta(t))$.

  \item In the \textbf{infinite-width limit}: (1)~the NTK converges
    to a deterministic kernel $\Theta_\infty$, (2)~the NTK stays
    approximately constant during training (lazy training),
    (3)~the network dynamics are linear, and (4)~training converges
    exponentially fast.

  \item The \textbf{lazy training regime} shows that training an
    overparameterized network is equivalent to kernel regression
    with the NTK, allowing generalization guarantees to be
    transferred from kernel theory.

  \item \textbf{Limitations}: the NTK regime does not capture
    \emph{feature learning}, which is where deep networks truly
    excel.  The NTK gives curse-of-dimensionality rates, while
    practical networks with feature learning can exploit structure
    to achieve faster rates.

  \item The norm-based Rademacher bound and the NTK theory provide
    complementary perspectives on generalization: the former says
    ``small weights imply generalization,'' and the latter says
    ``overparameterized networks implement kernel regression.''
\end{itemize}

\subsection*{Neural Network Theory Subdomain Summary}

This capstone lesson completes the Neural Network Theory subdomain
(Lessons~59--66).  The following table summarizes the arc of the
subdomain.

\begin{center}
\small
\begin{tabular}{clp{5cm}l}
\toprule
\textbf{Lesson} & \textbf{Topic} & \textbf{Key Result} & \textbf{Theme} \\
\midrule
59 & ERM and Generalization & VC dimension bounds;
  fundamental theorem of statistical learning &
  Statistical learning \\
60 & Bias-Variance and Model Selection & Bias-variance decomposition;
  oracle inequalities for model selection &
  Tradeoff analysis \\
61 & Kernel Methods & Mercer's theorem; the kernel trick
  for implicit feature maps &
  Feature spaces \\
62 & RKHS and Representer Theorem & Representer theorem reduces
  infinite-dimensional optimization to finite &
  Regularization \\
63 & Universal Approximation & Cybenko/Hornik: non-polynomial
  activations yield density in $C(K)$ &
  Expressiveness \\
64 & Barron's Theorem & Dimension-free $O(1/m)$ rate for
  Barron-class functions &
  Approximation rates \\
65 & Depth-Width Tradeoffs & Telgarsky: depth $L$ needs width
  $2^{\Omega(L)}$ in shallow nets; compositional gains &
  Architecture \\
66 & NTK and Rademacher Complexity & Norm-based generalization;
  infinite-width networks = kernel regression &
  Generalization \\
\bottomrule
\end{tabular}
\end{center}

The subdomain progresses from the foundations of statistical learning
(what does it mean to generalize?) through kernel methods (a tractable
infinite-dimensional framework) to neural network-specific results
(universal approximation, approximation rates, depth advantages, and
overparameterization).  The key tension throughout is between
\emph{expressiveness} (can the network represent the target?) and
\emph{generalization} (does it learn the target from finite data?).
Barron's theorem and the NTK provide two different resolutions: Barron
shows dimension-free approximation for a restricted class, while the
NTK shows that overparameterization does not hurt generalization
(in the lazy regime).

%% ============================================================
%% SECTION 9: FURTHER READING
%% ============================================================
\section{Further Reading}
\label{sec:further-reading}

\begin{itemize}
  \item \textbf{Lesson 67} (RNNs and BPTT) begins the next subdomain
    (3.2: Recurrent Architectures), studying how temporal depth
    interacts with the expressiveness and trainability results
    developed here.

  \item \textbf{Phase 04} (Deep RL) applies the theory of this
    subdomain to RL-specific settings: DQN and its extensions
    (Lesson~74), trust region methods (Lesson~76), PPO (Lesson~77),
    and SAC (Lesson~79).

  \item \citet{golowich2018} improves the spectrally-normalized bound
    by removing the dependence on depth in the logarithmic factor,
    showing that the bound can be made independent of both depth and
    width.

  \item \citet{allen-zhu2019} provides convergence guarantees for
    gradient descent on overparameterized networks that go beyond the
    NTK regime, using a notion of ``semi-smoothness'' of the loss
    landscape.

  \item \citet{chizat2018} develops the mean-field theory of
    two-layer networks as an alternative to the NTK: the ``active''
    or ``rich'' regime where features evolve during training and the
    network does not stay close to initialization.

  \item \citet{zhang2021} presents the celebrated experiments showing
    that overparameterized networks can memorize random labels yet
    generalize on structured data, demonstrating that classical
    generalization theory (VC, Rademacher) alone cannot explain
    deep learning generalization.

  \item \textbf{Historical note.}  The NTK was introduced by
    \citet{jacot2018} in 2018, building on earlier observations by
    Neal (1996) that wide neural networks converge to Gaussian
    processes.  The connection between overparameterization and
    kernel methods was independently developed by
    \citet{du2019} and \citet{allen-zhu2019}.  The spectrally-
    normalized bound of \citet{bartlett2017} was motivated by the
    practical success of spectral normalization in GANs and the
    theoretical need for parameter-count-independent generalization
    bounds.
\end{itemize}

\bibliography{nn-refs}
\end{document}
