\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 61: Kernel Methods}}
\fancyhead[R]{\thepage}
\title{Lesson 61: Kernel Methods}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% ============================================================
%% SECTION 1: MOTIVATION AND OVERVIEW
%% ============================================================

\section{Motivation and Overview}
\label{sec:motivation}

Consider the problem of classifying data points in $\R^2$ that are not
linearly separable---for instance, points arranged in two concentric
circles. A linear classifier in $\R^2$ cannot separate them, but
if we map each point $(x_1, x_2)$ to a higher-dimensional feature
space via $\feature(x_1, x_2) = (x_1^2, x_2^2, \sqrt{2}\, x_1 x_2)
\in \R^3$, the classes may become linearly separable. The difficulty
is that such feature maps can be extremely high-dimensional: a
degree-$p$ polynomial map on $\R^d$ produces features in
$\R^{\binom{d+p}{p}}$, which for $d = 1000$ and $p = 5$ exceeds
$10^{13}$. Computing and storing these feature vectors is infeasible.

The \emph{kernel trick} resolves this: if the learning algorithm
depends on the data only through inner products
$\langle \feature(x), \feature(x') \rangle$, we can replace
this inner product by a kernel function $\kernel(x, x')$ that
is computable in $O(d)$ time, bypassing the explicit
feature map entirely. This lesson develops the mathematical
foundations of kernel methods: the kernel trick, positive definite
kernels, Mercer's theorem providing the spectral foundation, and
systematic methods for constructing kernels. The tools developed in
this lesson will let us build the reproducing kernel Hilbert spaces
(RKHS) framework in Lesson~62 and connect kernel methods to neural
network theory.

\begin{rlconnection}[title={Why Kernel Methods Matter for RL/ML}]
\begin{itemize}[nosep]
    \item \textbf{Support vector machines:} SVMs solve a quadratic
    program whose objective and constraints depend on the data only
    through the Gram matrix $K_{ij} = \kernel(x_i, x_j)$. Replacing
    $\kernel$ with a nonlinear kernel yields nonlinear decision
    boundaries without modifying the optimization algorithm.
    \item \textbf{Gaussian processes:} A GP prior is specified by a
    mean function and a kernel $\kernel(x, x')$. The posterior
    predictions, uncertainty quantification, and Bayesian optimization
    acquisitions all flow from kernel evaluations.
    \item \textbf{Neural tangent kernel (NTK):} In the infinite-width
    limit, a neural network trained by gradient descent evolves as a
    linear model in a kernel regime, where the NTK
    $\Theta(x, x') = \langle \nabla_\theta f_\theta(x),
    \nabla_\theta f_\theta(x') \rangle$ governs learning dynamics.
    \item \textbf{Value function approximation:} Kernel-based
    temporal difference methods (e.g., KLSTD) represent the value
    function in an RKHS, enabling nonparametric approximation with
    convergence guarantees.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
Inner product spaces and the spectral theorem for compact
self-adjoint operators (Lesson~7, Lesson~19). $L^2$ spaces and
integration (Lesson~3, Lesson~5). Basic linear algebra: eigenvalues,
positive semidefiniteness (Lesson~6). Hypothesis classes and
empirical risk minimization (Lesson~59). Bias-variance tradeoff
(Lesson~60).

\paragraph{Learning objectives.}
After completing this lesson, the student will be able to:
\begin{itemize}
    \item \emph{Explain} the kernel trick and \emph{identify} when
    an algorithm can be kernelized.
    \item \emph{Define} positive definite kernels and \emph{prove}
    that inner-product kernels are positive definite.
    \item \emph{Prove} closure properties of the class of positive
    definite kernels (sums, products, limits).
    \item \emph{State} Mercer's theorem and \emph{derive} the
    eigenexpansion of a continuous positive definite kernel.
    \item \emph{Construct} standard kernels (polynomial, Gaussian
    RBF, Laplacian) and \emph{verify} their positive definiteness.
    \item \emph{Apply} kernel methods to a concrete classification
    or regression problem.
\end{itemize}

%% ============================================================
%% SECTION 2: REFERENCES
%% ============================================================

\section{References}
\label{sec:references}

\begin{itemize}
    \item \citet[Chapter 4]{bach2024} -- Kernels, positive definite
    functions, and Mercer's theorem in the learning theory context.
    \item \citet[Chapters 2--3]{scholkopf2002} -- Comprehensive
    treatment of kernels, feature spaces, and the kernel trick for
    SVMs.
    \item \citet[Chapters 5--6]{steinwart2008} -- Rigorous treatment
    of Mercer kernels, RKHS construction, and kernel properties.
    \item \citet[Chapter 6]{mohri2018} -- Kernel methods from the
    foundations of machine learning perspective.
    \item \citet{mercer1909} -- Original paper on the eigenexpansion
    of positive definite integral kernels.
\end{itemize}

%% ============================================================
%% SECTION 3: THE KERNEL TRICK
%% ============================================================

\section{The Kernel Trick}
\label{sec:kernel-trick}

The kernel trick is a computational strategy: replace explicit
feature-space computations with evaluations of a kernel function.
We begin by formalizing the setting.

\begin{definition}[Feature Map]
\label{def:feature-map}
Let $\mathcal{X}$ be an input space (typically a subset of $\R^d$)
and let $\mathcal{V}$ be an inner product space (the \emph{feature
space}). A \emph{feature map} is any function
$\feature \colon \mathcal{X} \to \mathcal{V}$.
\end{definition}

\begin{definition}[Kernel Function]
\label{def:kernel-function}
Given a feature map $\feature \colon \mathcal{X} \to \mathcal{V}$,
the associated \emph{kernel function}
$\kernel \colon \mathcal{X} \times \mathcal{X} \to \R$ is defined by
\begin{equation}
\label{eq:kernel-def}
\kernel(x, x') = \langle \feature(x), \feature(x') \rangle_{\mathcal{V}}.
\end{equation}
\end{definition}

The key observation is that many learning algorithms---including
ridge regression, support vector machines, and principal component
analysis---interact with the data $x_1, \ldots, x_n$ only through
pairwise inner products $\langle \feature(x_i), \feature(x_j)
\rangle$. If we can evaluate $\kernel(x_i, x_j)$ directly, we
never need to compute $\feature(x_i)$ explicitly.

\begin{example}[Polynomial Kernel]
\label{ex:polynomial-kernel}
Let $\mathcal{X} = \R^d$ and define $\kernel(x, x') = (1 + x^\top
x')^p$ for a positive integer $p$. We show this is an inner-product
kernel by exhibiting the feature map for $d = 2$, $p = 2$.

We have $\kernel(x, x') = (1 + x_1 x_1' + x_2 x_2')^2$. Expanding:
\begin{align*}
\kernel(x, x') &= 1 + x_1^2 (x_1')^2 + x_2^2 (x_2')^2
  + 2 x_1 x_1' + 2 x_2 x_2' + 2 x_1 x_2 x_1' x_2' \\
&= \langle \feature(x), \feature(x') \rangle,
\end{align*}
where $\feature(x) = (1, x_1^2, x_2^2, \sqrt{2}\, x_1,
\sqrt{2}\, x_2, \sqrt{2}\, x_1 x_2) \in \R^6$.

More generally, for $\mathcal{X} = \R^d$ the feature space has
dimension $\binom{d+p}{p}$, but evaluating $\kernel(x, x')$
requires only $O(d)$ operations (one inner product plus
exponentiation).
\end{example}

\begin{example}[Kernelized Ridge Regression]
\label{ex:kernel-ridge}
Standard ridge regression solves
$\min_{w \in \R^D} \sum_{i=1}^n (y_i - w^\top \feature(x_i))^2
+ \lambda \|w\|^2$
in the feature space $\R^D$, where $D = \dim(\mathcal{V})$ may be
very large. Writing $\Phi = [\feature(x_1), \ldots, \feature(x_n)]^\top
\in \R^{n \times D}$, the solution is $w^* = (\Phi^\top \Phi +
\lambda I)^{-1} \Phi^\top y$, which requires $O(D^2 n)$ operations.

By the matrix inversion lemma, the prediction at a new point $x$ is
\[
f(x) = (w^*)^\top \feature(x) = k_x^\top (K + \lambda I)^{-1} y,
\]
where $K_{ij} = \kernel(x_i, x_j)$ is the $n \times n$ Gram matrix
and $k_x = (\kernel(x_1, x), \ldots, \kernel(x_n, x))^\top$.
This requires only $O(n^2 d + n^3)$ operations, independent of $D$.
\end{example}

\begin{proposition}[Kernelization Criterion]
\label{prop:kernelization}
An algorithm operating on data $x_1, \ldots, x_n \in \mathcal{X}$
via a feature map $\feature$ can be kernelized if and only if every
computation involving the feature vectors $\feature(x_1), \ldots,
\feature(x_n)$ can be expressed in terms of the entries of the
Gram matrix $K_{ij} = \langle \feature(x_i), \feature(x_j) \rangle$.
\end{proposition}

\begin{proof}
The ``if'' direction is immediate: if all computations reduce to
entries of $K$, we replace each $K_{ij}$ by $\kernel(x_i, x_j)$.

For the ``only if'' direction, suppose the algorithm requires some
computation that cannot be expressed through inner products alone.
Then there exist two feature maps $\feature, \feature'$ with
$\langle \feature(x_i), \feature(x_j) \rangle = \langle
\feature'(x_i), \feature'(x_j) \rangle$ for all $i, j$, but the
algorithm produces different outputs for $\feature$ and $\feature'$.
Hence the algorithm is not determined by the kernel, and kernelization
is impossible.
\end{proof}

\begin{rlconnection}[title={The Kernel Trick in RL}]
In temporal difference learning with linear function approximation,
the value function is $\hat{V}(s) = w^\top \feature(s)$. The
TD(0) update $w \leftarrow w + \alpha \delta_t \feature(s_t)$, where
$\delta_t = r_t + \gamma w^\top \feature(s_{t+1}) - w^\top
\feature(s_t)$, involves $\feature$ only through inner products.
This enables kernel TD methods that represent the value function
nonparametrically as $\hat{V}(s) = \sum_{i=1}^n \alpha_i
\kernel(s_i, s)$.
\end{rlconnection}

The kernel trick transforms the computational challenge from working
in a high-dimensional feature space to working with an $n \times n$
matrix, but it raises a fundamental question: which functions
$\kernel \colon \mathcal{X} \times \mathcal{X} \to \R$ correspond to
inner products in some feature space? This is answered by the theory
of positive definite kernels.

%% ============================================================
%% SECTION 4: POSITIVE DEFINITE KERNELS
%% ============================================================

\section{Positive Definite Kernels}
\label{sec:pd-kernels}

The concept of positive definiteness characterizes exactly those
functions that arise as inner products in some feature space.

\begin{definition}[Positive Definite Kernel]
\label{def:pd-kernel}
A function $\kernel \colon \mathcal{X} \times \mathcal{X} \to \R$
is \emph{positive definite} (pd) if it is symmetric
($\kernel(x, x') = \kernel(x', x)$ for all $x, x' \in \mathcal{X}$)
and for every $n \ge 1$, every set of points $\{x_1, \ldots, x_n\}
\subseteq \mathcal{X}$, and every $\alpha \in \R^n$:
\begin{equation}
\label{eq:pd-condition}
\sum_{i=1}^n \sum_{j=1}^n \alpha_i \alpha_j \kernel(x_i, x_j) \ge 0.
\end{equation}
Equivalently, the \emph{Gram matrix} $K \in \R^{n \times n}$ with
$K_{ij} = \kernel(x_i, x_j)$ is positive semidefinite for every
finite subset of $\mathcal{X}$.
\end{definition}

\begin{remark}
\label{rem:pd-terminology}
Some authors use ``positive semidefinite kernel'' for the condition
\eqref{eq:pd-condition} and reserve ``positive definite'' for strict
inequality when $\alpha \ne 0$ and $x_1, \ldots, x_n$ are distinct.
We follow the convention of \citet{bach2024} and \citet{scholkopf2002},
using ``positive definite'' for the condition as stated.
\end{remark}

\begin{theorem}[Inner-Product Kernels are Positive Definite]
\label{thm:ip-kernel-pd}
If $\feature \colon \mathcal{X} \to \mathcal{V}$ is a feature map
into an inner product space $\mathcal{V}$, then
$\kernel(x, x') = \langle \feature(x), \feature(x') \rangle$
is a positive definite kernel.
\end{theorem}

\begin{proof}
Let $x_1, \ldots, x_n \in \mathcal{X}$ and $\alpha_1, \ldots,
\alpha_n \in \R$. Then
\begin{align*}
\sum_{i=1}^n \sum_{j=1}^n \alpha_i \alpha_j \kernel(x_i, x_j)
&= \sum_{i=1}^n \sum_{j=1}^n \alpha_i \alpha_j
   \langle \feature(x_i), \feature(x_j) \rangle \\
&= \left\langle \sum_{i=1}^n \alpha_i \feature(x_i),
   \sum_{j=1}^n \alpha_j \feature(x_j) \right\rangle \\
&= \left\| \sum_{i=1}^n \alpha_i \feature(x_i) \right\|^2 \ge 0,
\end{align*}
where the second equality uses bilinearity of the inner product and
the last step uses positive definiteness of the norm.
\end{proof}

The converse also holds: every positive definite kernel arises from
an inner product in some feature space. This is the content of the
RKHS construction in Lesson~62 (Aronszajn's theorem).

\begin{keyresult}[title={Characterization of Positive Definite Kernels}]
A symmetric function $\kernel \colon \mathcal{X} \times \mathcal{X}
\to \R$ is positive definite if and only if there exists a Hilbert
space $\mathcal{V}$ and a feature map $\feature \colon \mathcal{X}
\to \mathcal{V}$ such that $\kernel(x, x') = \langle \feature(x),
\feature(x') \rangle_{\mathcal{V}}$. The ``if'' direction is
Theorem~\ref{thm:ip-kernel-pd}; the ``only if'' direction is
Aronszajn's theorem (Lesson~62, Theorem~3.5).
\end{keyresult}

We now establish that the class of pd kernels is closed under
several useful operations.

\begin{theorem}[Closure Properties of PD Kernels]
\label{thm:pd-closure}
Let $\kernel_1, \kernel_2$ be positive definite kernels on
$\mathcal{X} \times \mathcal{X}$. Then the following are also
positive definite:
\begin{enumerate}[label=(\roman*)]
    \item \textbf{Sum:} $\kernel(x, x') = \kernel_1(x, x') +
    \kernel_2(x, x')$.
    \item \textbf{Scalar multiple:} $\kernel(x, x') = c \,
    \kernel_1(x, x')$ for any $c \ge 0$.
    \item \textbf{Product:} $\kernel(x, x') = \kernel_1(x, x')
    \cdot \kernel_2(x, x')$.
    \item \textbf{Pointwise limit:} If $\kernel_m$ is pd for each
    $m$ and $\kernel_m(x, x') \to \kernel(x, x')$ for all $x, x'$,
    then $\kernel$ is pd.
\end{enumerate}
\end{theorem}

\begin{proof}
Let $x_1, \ldots, x_n \in \mathcal{X}$ and $\alpha \in \R^n$. Let
$K^{(1)}, K^{(2)}$ denote the Gram matrices of $\kernel_1, \kernel_2$.

\textbf{(i) Sum.} The Gram matrix is $K = K^{(1)} + K^{(2)}$.
Then $\alpha^\top K \alpha = \alpha^\top K^{(1)} \alpha +
\alpha^\top K^{(2)} \alpha \ge 0 + 0 = 0$.

\textbf{(ii) Scalar multiple.} The Gram matrix is $cK^{(1)}$.
Then $\alpha^\top (cK^{(1)}) \alpha = c \, \alpha^\top K^{(1)}
\alpha \ge 0$ since $c \ge 0$.

\textbf{(iii) Product (Schur product theorem).} The Gram matrix
is the Hadamard (entrywise) product $K = K^{(1)} \circ K^{(2)}$.
By the spectral theorem, write $K^{(2)} = \sum_{l=1}^n \mu_l
v_l v_l^\top$ where $\mu_l \ge 0$ (since $K^{(2)} \succeq 0$).
Then
\[
K_{ij} = K^{(1)}_{ij} K^{(2)}_{ij}
= K^{(1)}_{ij} \sum_{l=1}^n \mu_l (v_l)_i (v_l)_j
= \sum_{l=1}^n \mu_l (v_l)_i (v_l)_j K^{(1)}_{ij}.
\]
Hence $\alpha^\top K \alpha = \sum_{l=1}^n \mu_l
\sum_{i,j} \alpha_i (v_l)_i K^{(1)}_{ij} \alpha_j (v_l)_j
= \sum_{l=1}^n \mu_l \, \beta_l^\top K^{(1)} \beta_l$,
where $\beta_l = (\alpha_1 (v_l)_1, \ldots, \alpha_n (v_l)_n)$.
Since $K^{(1)} \succeq 0$ and $\mu_l \ge 0$, each term is
nonneg\-ative, so $\alpha^\top K \alpha \ge 0$.

\textbf{(iv) Pointwise limit.} The Gram matrix $K^{(m)}$ of
$\kernel_m$ satisfies $\alpha^\top K^{(m)} \alpha \ge 0$ for
each $m$. Since $K^{(m)}_{ij} \to K_{ij}$ entrywise as
$m \to \infty$, we obtain $\alpha^\top K \alpha =
\lim_{m \to \infty} \alpha^\top K^{(m)} \alpha \ge 0$.
\end{proof}

\begin{example}[Linear Kernel]
\label{ex:linear-kernel}
The \emph{linear kernel} on $\R^d$ is $\kernel(x, x') = x^\top x'$.
This corresponds to the identity feature map $\feature(x) = x$. It
is pd by Theorem~\ref{thm:ip-kernel-pd}.
\end{example}

\begin{example}[Polynomial Kernel -- Positive Definiteness]
\label{ex:poly-kernel-pd}
The polynomial kernel $\kernel(x, x') = (1 + x^\top x')^p$ for
$p \in \N$ is pd. To see this, write $\kernel(x, x') =
\sum_{j=0}^p \binom{p}{j} (x^\top x')^j$. The linear kernel
$\kernel_0(x, x') = x^\top x'$ is pd by the previous example.
By closure under products (iii), $(x^\top x')^j$ is pd for
each $j \ge 1$. The constant function $\kernel(x,x') = 1$ is
pd (Gram matrix is all-ones, which is psd). By closure under
sums (i) and scalar multiples (ii), $\kernel$ is pd.
\end{example}

\begin{example}[Gaussian RBF Kernel]
\label{ex:gaussian-rbf}
The \emph{Gaussian radial basis function} (RBF) kernel on $\R^d$ is
\begin{equation}
\label{eq:gaussian-kernel}
\kernel(x, x') = \exp\!\left(-\frac{\|x - x'\|^2}{2\sigma^2}\right),
\end{equation}
where $\sigma > 0$ is the \emph{bandwidth} parameter. This is
perhaps the most widely used kernel in practice. We prove its
positive definiteness in Section~\ref{sec:kernel-construction}
(Theorem~\ref{thm:gaussian-pd}).
\end{example}

\begin{warning}[title={Not Every Symmetric Function is a Kernel}]
The function $\kernel(x, x') = -\|x - x'\|^2$ is symmetric but
\emph{not} positive definite. For $n = 2$ points $x_1, x_2$ with
$\alpha = (1, -1)^\top$: $\alpha^\top K \alpha = -2\|x_1 - x_2\|^2
< 0$ whenever $x_1 \ne x_2$. One must always verify positive
definiteness.
\end{warning}

The closure properties allow us to build a rich algebra of kernels
from simple building blocks. We next develop the spectral theory of
positive definite kernels, which reveals their deep structure.

%% ============================================================
%% SECTION 5: MERCER'S THEOREM
%% ============================================================

\section{Mercer's Theorem}
\label{sec:mercer}

Mercer's theorem provides a spectral decomposition for continuous
positive definite kernels, analogous to the eigendecomposition of
positive semidefinite matrices. This decomposition is the foundation
of the canonical feature map and underpins the RKHS construction
of Lesson~62.

\begin{assumption}[Setting for Mercer's Theorem]
\label{asm:mercer-setting}
Let $\mathcal{X}$ be a compact subset of $\R^d$ and let $\mu$ be
a finite Borel measure on $\mathcal{X}$ with $\supp(\mu) =
\mathcal{X}$. Let $\kernel \colon \mathcal{X} \times \mathcal{X}
\to \R$ be continuous.
\end{assumption}

Under these assumptions, the kernel defines an integral operator.

\begin{definition}[Integral Operator]
\label{def:integral-operator}
The \emph{integral operator} $T_\kernel \colon L^2(\mathcal{X}, \mu)
\to L^2(\mathcal{X}, \mu)$ associated with kernel $\kernel$ is
\begin{equation}
\label{eq:integral-operator}
(T_\kernel f)(x) = \int_{\mathcal{X}} \kernel(x, x') f(x') \,
d\mu(x').
\end{equation}
\end{definition}

\begin{lemma}[Properties of $T_\kernel$]
\label{lem:integral-op-properties}
Under Assumption~\ref{asm:mercer-setting}, if $\kernel$ is
continuous and positive definite, then $T_\kernel$ is:
\begin{enumerate}[label=(\roman*)]
    \item a bounded linear operator on $L^2(\mathcal{X}, \mu)$;
    \item self-adjoint: $\langle T_\kernel f, g \rangle_{L^2} =
    \langle f, T_\kernel g \rangle_{L^2}$;
    \item positive: $\langle T_\kernel f, f \rangle_{L^2} \ge 0$
    for all $f$;
    \item compact.
\end{enumerate}
\end{lemma}

\begin{proof}
\textbf{(i)} Since $\kernel$ is continuous on $\mathcal{X} \times
\mathcal{X}$ and $\mathcal{X}$ is compact, $M :=
\sup_{x, x'} |\kernel(x, x')| < \infty$. By the Cauchy--Schwarz
inequality,
\[
|(T_\kernel f)(x)| \le \int_{\mathcal{X}} |\kernel(x, x')|
|f(x')| \, d\mu(x') \le M \|f\|_{L^1} \le M \mu(\mathcal{X})^{1/2}
\|f\|_{L^2},
\]
so $\|T_\kernel f\|_{L^2} \le M \mu(\mathcal{X}) \|f\|_{L^2}$,
establishing boundedness.

\textbf{(ii)} By symmetry of $\kernel$ and Fubini's theorem:
\begin{align*}
\langle T_\kernel f, g \rangle_{L^2}
&= \int_{\mathcal{X}} \int_{\mathcal{X}} \kernel(x, x') f(x')
g(x) \, d\mu(x') \, d\mu(x) \\
&= \int_{\mathcal{X}} f(x') \int_{\mathcal{X}} \kernel(x', x)
g(x) \, d\mu(x) \, d\mu(x')
= \langle f, T_\kernel g \rangle_{L^2}.
\end{align*}

\textbf{(iii)} For positive definiteness, we approximate. For
$f \in L^2(\mathcal{X}, \mu)$:
\[
\langle T_\kernel f, f \rangle_{L^2}
= \int_{\mathcal{X}} \int_{\mathcal{X}} \kernel(x, x') f(x)
f(x') \, d\mu(x) \, d\mu(x').
\]
This double integral is the continuous analogue of the quadratic
form $\alpha^\top K \alpha$. To verify nonnegativity, approximate
$f$ by simple functions $f_n = \sum_{i=1}^{m_n} c_i
\indicator_{A_i}$. For simple functions, the integral reduces to
$\sum_{i,j} c_i c_j \int_{A_i} \int_{A_j} \kernel(x, x') \,
d\mu(x) \, d\mu(x')$. By a discretization argument (partitioning
each $A_i$ into small sets and applying the pd condition on point
evaluations), each such sum is nonneg\-ative. Taking limits (using
dominated convergence), $\langle T_\kernel f, f \rangle_{L^2} \ge 0$.

\textbf{(iv)} Since $\kernel$ is continuous on the compact set
$\mathcal{X} \times \mathcal{X}$, it is uniformly continuous.
The family $\{x' \mapsto \kernel(x, x') : x \in \mathcal{X}\}$
is equicontinuous and uniformly bounded in $L^2$. Therefore
$T_\kernel$ maps the unit ball to an equicontinuous family, which
is precompact in $L^2$ by the Arzela--Ascoli theorem. Hence
$T_\kernel$ is compact.
\end{proof}

\begin{theorem}[Mercer's Theorem]
\label{thm:mercer}
Under Assumption~\ref{asm:mercer-setting}, if $\kernel$ is
continuous and positive definite, then there exists a sequence
of eigenfunctions $\{\feature_j\}_{j=1}^\infty$ forming an
orthonormal basis of $\overline{\mathrm{Range}(T_\kernel)}$ in
$L^2(\mathcal{X}, \mu)$, with corresponding eigenvalues
$\lambda_1 \ge \lambda_2 \ge \cdots \ge 0$, such that:
\begin{equation}
\label{eq:mercer-expansion}
\kernel(x, x') = \sum_{j=1}^\infty \lambda_j \, \feature_j(x) \,
\feature_j(x'),
\end{equation}
where the convergence is absolute and uniform on
$\mathcal{X} \times \mathcal{X}$. Moreover, $T_\kernel \feature_j
= \lambda_j \feature_j$ and $\sum_{j=1}^\infty \lambda_j =
\int_{\mathcal{X}} \kernel(x, x) \, d\mu(x) < \infty$.
\end{theorem}

\begin{proof}
By Lemma~\ref{lem:integral-op-properties}, $T_\kernel$ is a
compact, self-adjoint, positive operator on the separable Hilbert
space $L^2(\mathcal{X}, \mu)$.

\textbf{Step 1: Spectral decomposition.} By the spectral theorem
for compact self-adjoint operators (Lesson~19), there exists an
orthonormal system $\{\feature_j\}_{j \ge 1}$ of eigenfunctions
with eigenvalues $\lambda_1 \ge \lambda_2 \ge \cdots$ satisfying
$\lambda_j \to 0$, such that
\[
T_\kernel f = \sum_{j=1}^\infty \lambda_j \langle f, \feature_j
\rangle_{L^2} \feature_j
\]
for all $f \in L^2(\mathcal{X}, \mu)$. Since $T_\kernel$ is
positive, all $\lambda_j \ge 0$.

\textbf{Step 2: $L^2$ expansion of the kernel.} Fix $x \in
\mathcal{X}$ and consider $\kernel_x := \kernel(x, \cdot) \in
L^2(\mathcal{X}, \mu)$ (this holds since $\kernel$ is bounded
on the compact set $\mathcal{X} \times \mathcal{X}$). Then
\[
(T_\kernel \feature_j)(x) = \int_{\mathcal{X}} \kernel(x, x')
\feature_j(x') \, d\mu(x') = \lambda_j \feature_j(x).
\]
Also, $\langle \kernel_x, \feature_j \rangle_{L^2} =
\int_{\mathcal{X}} \kernel(x, x') \feature_j(x') \, d\mu(x')
= \lambda_j \feature_j(x)$. Hence the Fourier expansion of
$\kernel_x$ in $L^2$ gives
\[
\kernel(x, \cdot) = \sum_{j=1}^\infty \langle \kernel_x,
\feature_j \rangle_{L^2} \feature_j(\cdot)
= \sum_{j=1}^\infty \lambda_j \feature_j(x) \feature_j(\cdot),
\]
with convergence in $L^2(\mathcal{X}, \mu)$. Evaluating at $x'$
(which requires pointwise convergence, established below) gives
\eqref{eq:mercer-expansion}.

\textbf{Step 3: Uniform convergence.} Define the partial sums
$\kernel_N(x, x') = \sum_{j=1}^N \lambda_j \feature_j(x)
\feature_j(x')$. Since each $\feature_j$ is continuous (as
$\feature_j = \lambda_j^{-1} T_\kernel \feature_j$ and $T_\kernel$
maps $L^2$ functions to continuous functions when $\kernel$ is
continuous), $\kernel_N$ is continuous.

For the diagonal, by Parseval's identity applied to $\kernel_x$:
\[
\sum_{j=1}^\infty \lambda_j^2 \feature_j(x)^2
= \sum_{j=1}^\infty |\langle \kernel_x, \feature_j \rangle_{L^2}|^2
\le \|\kernel_x\|_{L^2}^2 < \infty.
\]
Moreover, $\kernel(x, x) - \kernel_N(x, x) = \sum_{j > N}
\lambda_j \feature_j(x)^2 \ge 0$ (since $\lambda_j \ge 0$).
By Dini's theorem, since $\kernel(x,x) - \kernel_N(x,x)$ is a
decreasing sequence of continuous functions converging pointwise
to $0$ on the compact set $\mathcal{X}$, the convergence is
uniform on the diagonal.

For the off-diagonal, by the Cauchy--Schwarz inequality:
\begin{align*}
|\kernel(x, x') - \kernel_N(x, x')|
&= \left| \sum_{j > N} \lambda_j \feature_j(x) \feature_j(x')
\right| \\
&\le \left( \sum_{j > N} \lambda_j \feature_j(x)^2 \right)^{1/2}
\left( \sum_{j > N} \lambda_j \feature_j(x')^2 \right)^{1/2} \\
&= \sqrt{(\kernel(x,x) - \kernel_N(x,x))(\kernel(x',x') -
\kernel_N(x',x'))}.
\end{align*}
Since the diagonal convergence is uniform, the right side converges
uniformly to $0$ on $\mathcal{X} \times \mathcal{X}$.

\textbf{Step 4: Trace formula.} Integrating the diagonal:
$\sum_{j=1}^\infty \lambda_j = \sum_{j=1}^\infty \lambda_j
\int_{\mathcal{X}} \feature_j(x)^2 \, d\mu(x) =
\int_{\mathcal{X}} \kernel(x, x) \, d\mu(x) < \infty$,
using uniform convergence to exchange sum and integral.
\end{proof}

\begin{keyresult}[title={Mercer's Canonical Feature Map}]
\label{kr:mercer-feature-map}
Mercer's theorem provides a \emph{canonical feature map}. Define
$\Phi \colon \mathcal{X} \to \ell^2$ by
\[
\Phi(x) = \bigl(\sqrt{\lambda_1}\, \feature_1(x), \;
\sqrt{\lambda_2}\, \feature_2(x), \; \ldots \bigr).
\]
Then $\langle \Phi(x), \Phi(x') \rangle_{\ell^2} =
\sum_{j=1}^\infty \lambda_j \feature_j(x) \feature_j(x')
= \kernel(x, x')$. This is the unique (up to unitary
equivalence) feature map into a separable Hilbert space.
\end{keyresult}

\begin{example}[Eigenexpansion of the Gaussian Kernel on {$[0,1]$}]
\label{ex:gaussian-eigenexpansion}
Consider the Gaussian kernel $\kernel(x, x') = e^{-(x-x')^2 /
(2\sigma^2)}$ on $\mathcal{X} = [0,1]$ with Lebesgue measure.
While no closed-form eigenfunctions exist in general, the
eigenvalues $\lambda_j$ decay exponentially fast. Specifically,
for the Gaussian kernel on $\R$ (with Gaussian measure), the
eigenfunctions are Hermite functions and the eigenvalues satisfy
$\lambda_j = O(c^j)$ for some $0 < c < 1$ depending on $\sigma$.

For numerical illustration with $\sigma = 0.3$, the first few
normalized eigenvalues are approximately $\lambda_1 \approx 0.42$,
$\lambda_2 \approx 0.28$, $\lambda_3 \approx 0.15$,
$\lambda_4 \approx 0.07$, with rapid decay thereafter. The
eigenfunctions $\feature_j$ oscillate with increasing frequency,
analogous to Fourier modes. The rapid eigenvalue decay means
that a low-rank approximation $\kernel(x, x') \approx
\sum_{j=1}^J \lambda_j \feature_j(x) \feature_j(x')$ is
accurate even for moderate $J$.
\end{example}

\begin{rlconnection}[title={Mercer's Theorem and the NTK}]
The neural tangent kernel $\Theta(x, x') = \langle
\nabla_\theta f_\theta(x), \nabla_\theta f_\theta(x') \rangle$
is positive definite (by Theorem~\ref{thm:ip-kernel-pd} with
the gradient as feature map). In the infinite-width limit, the
NTK is deterministic and continuous, so Mercer's theorem applies.
The eigenvalues of the NTK integral operator determine the
learning rates of different modes: the network learns eigenfunctions
with large $\lambda_j$ first (spectral bias), explaining why
neural networks tend to learn low-frequency patterns before
high-frequency ones.
\end{rlconnection}

Mercer's theorem tells us that every continuous pd kernel has a
spectral decomposition. We now turn to the practical question of
constructing kernels for specific applications.

%% ============================================================
%% SECTION 6: KERNEL CONSTRUCTION
%% ============================================================

\section{Kernel Construction}
\label{sec:kernel-construction}

We develop systematic methods for building positive definite
kernels, starting from basic building blocks and using the closure
properties of Theorem~\ref{thm:pd-closure}.

\subsection{Kernels from Feature Maps}

The most direct construction: choose a feature map $\feature \colon
\mathcal{X} \to \R^D$ and define $\kernel(x, x') = \feature(x)^\top
\feature(x')$.

\begin{example}[Bag-of-Words Kernel]
\label{ex:bow-kernel}
For text classification, let $\mathcal{X}$ be a set of documents
and define $\feature(x) \in \R^V$ as the word-count vector (where
$V$ is the vocabulary size). Then $\kernel(x, x') = \feature(x)^\top
\feature(x')$ counts the total number of shared word occurrences.
This is pd by Theorem~\ref{thm:ip-kernel-pd}.
\end{example}

\subsection{Polynomial Kernels}

We have already shown that $\kernel(x, x') = (1 + x^\top x')^p$ is
pd (Example~\ref{ex:poly-kernel-pd}). More generally:

\begin{proposition}[General Polynomial Kernel]
\label{prop:general-poly}
For $c \ge 0$ and $p \in \N$, the kernel $\kernel(x, x') =
(c + x^\top x')^p$ on $\R^d$ is positive definite.
\end{proposition}

\begin{proof}
The constant kernel $\kernel_c(x, x') = c$ is pd (its Gram matrix
$cJ_n$, where $J_n$ is the all-ones matrix, is psd since $J_n =
\mathbf{1}\mathbf{1}^\top$). The linear kernel $\kernel_0(x, x')
= x^\top x'$ is pd. By closure under sums, $\kernel_c + \kernel_0$
is pd. By closure under products, $(c + x^\top x')^p$ is pd.
\end{proof}

\subsection{The Gaussian RBF Kernel}

\begin{theorem}[Positive Definiteness of the Gaussian Kernel]
\label{thm:gaussian-pd}
The Gaussian RBF kernel $\kernel(x, x') = \exp(-\|x - x'\|^2 /
(2\sigma^2))$ is positive definite on $\R^d$ for any $\sigma > 0$.
\end{theorem}

\begin{proof}
We proceed in three steps.

\textbf{Step 1.} Expand the squared distance:
\[
\kernel(x, x') = \exp\!\left(-\frac{\|x\|^2}{2\sigma^2}\right)
\exp\!\left(\frac{x^\top x'}{\sigma^2}\right)
\exp\!\left(-\frac{\|x'\|^2}{2\sigma^2}\right).
\]

\textbf{Step 2.} Show that $\kernel_0(x, x') = \exp(x^\top x' /
\sigma^2)$ is pd. Write the Taylor expansion:
\[
\exp\!\left(\frac{x^\top x'}{\sigma^2}\right)
= \sum_{p=0}^\infty \frac{1}{p!} \left(\frac{x^\top x'}{\sigma^2}
\right)^p
= \sum_{p=0}^\infty \frac{1}{p! \, \sigma^{2p}} (x^\top x')^p.
\]
Each term $(x^\top x')^p$ is pd (as the $p$-fold product of the
linear kernel). Each coefficient $1/(p! \, \sigma^{2p}) > 0$. Finite
sums of pd kernels with nonneg\-ative coefficients are pd by
Theorem~\ref{thm:pd-closure}(i)--(ii). The infinite sum converges
pointwise, so by Theorem~\ref{thm:pd-closure}(iv), $\kernel_0$ is pd.

\textbf{Step 3.} Define $g(x) = \exp(-\|x\|^2 / (2\sigma^2))$.
Then $\kernel(x, x') = g(x) \, \kernel_0(x, x') \, g(x')$. For any
$\alpha \in \R^n$:
\[
\sum_{i,j} \alpha_i \alpha_j \kernel(x_i, x_j)
= \sum_{i,j} (\alpha_i g(x_i))(\alpha_j g(x_j)) \kernel_0(x_i, x_j)
= \sum_{i,j} \beta_i \beta_j \kernel_0(x_i, x_j) \ge 0,
\]
where $\beta_i = \alpha_i g(x_i)$, and the last inequality holds
because $\kernel_0$ is pd. Hence $\kernel$ is pd.
\end{proof}

\begin{remark}[Infinite-Dimensional Feature Space]
\label{rem:gaussian-infinite-dim}
The Mercer eigenvalues of the Gaussian kernel are all strictly
positive (the kernel is \emph{strictly} positive definite on any
finite point set with distinct points). This means the feature
space is infinite-dimensional---no finite-dimensional feature
map can reproduce the Gaussian kernel exactly. The power of the
kernel trick is that we can work with this infinite-dimensional
feature space implicitly through $O(1)$ kernel evaluations.
\end{remark}

\subsection{Laplacian Kernel}

\begin{proposition}[Laplacian Kernel]
\label{prop:laplacian-pd}
The Laplacian kernel $\kernel(x, x') = \exp(-\|x - x'\|_1 / \sigma)$
on $\R^d$ is positive definite for $\sigma > 0$.
\end{proposition}

\begin{proof}
Since $\|x - x'\|_1 = \sum_{j=1}^d |x_j - x_j'|$, we have
$\kernel(x, x') = \prod_{j=1}^d \exp(-|x_j - x_j'| / \sigma)$.
It suffices to show that $\kernel_1(s, t) = \exp(-|s - t| / \sigma)$
on $\R$ is pd. This follows from the Fourier representation:
$e^{-|u|/\sigma} = \frac{2/\sigma}{\pi} \int_0^\infty
\frac{\cos(\omega u)}{1/\sigma^2 + \omega^2} \, d\omega$
for $u \in \R$. Since each $\cos(\omega(s-t))$ defines a pd kernel
(being $\langle (\cos \omega s, \sin \omega s),
(\cos \omega t, \sin \omega t) \rangle$) and we integrate with a
nonneg\-ative weight, $\kernel_1$ is pd. By closure under products,
$\kernel$ is pd.
\end{proof}

\subsection{Operations Preserving Positive Definiteness}

Beyond sums, products, and limits, several other operations preserve
positive definiteness.

\begin{proposition}[Additional Closure Operations]
\label{prop:additional-closure}
Let $\kernel$ be a pd kernel on $\mathcal{X}$.
\begin{enumerate}[label=(\roman*)]
    \item \textbf{Composition with a power series:} If
    $f(t) = \sum_{p=0}^\infty a_p t^p$ with $a_p \ge 0$ for all $p$
    and the series converges for $t \ge 0$, then $f \circ \kernel$
    is pd.
    \item \textbf{Restriction:} If $\mathcal{Y} \subseteq \mathcal{X}$,
    then $\kernel|_{\mathcal{Y} \times \mathcal{Y}}$ is pd.
    \item \textbf{Direct sum:} If $\kernel_1$ is pd on $\mathcal{X}_1$
    and $\kernel_2$ is pd on $\mathcal{X}_2$, then $\kernel((x_1, x_2),
    (x_1', x_2')) = \kernel_1(x_1, x_1') + \kernel_2(x_2, x_2')$ is
    pd on $\mathcal{X}_1 \times \mathcal{X}_2$.
    \item \textbf{Tensor product:} $\kernel((x_1, x_2), (x_1', x_2'))
    = \kernel_1(x_1, x_1') \cdot \kernel_2(x_2, x_2')$ is pd on
    $\mathcal{X}_1 \times \mathcal{X}_2$.
\end{enumerate}
\end{proposition}

\begin{proof}
\textbf{(i)} $f(\kernel(x,x')) = \sum_{p=0}^\infty a_p \kernel(x,x')^p$.
Each $\kernel^p$ is pd (closure under products), $a_p \ge 0$ (closure
under scalar multiples), partial sums are pd (closure under sums),
and the pointwise limit is pd (closure under limits).

\textbf{(ii)} The Gram matrix of $\kernel|_\mathcal{Y}$ on points
$y_1, \ldots, y_n \in \mathcal{Y} \subseteq \mathcal{X}$ is the
same as the Gram matrix of $\kernel$ on these points, which is psd.

\textbf{(iii)--(iv)} For direct sums: the Gram matrix is
$K_1 + K_2$ where $K_1, K_2 \succeq 0$. For tensor products:
the Gram matrix is $K_1 \circ K_2$ (Hadamard product), which is
psd by the Schur product theorem.
\end{proof}

\begin{example}[Constructing Kernels by Composition]
\label{ex:kernel-composition}
Starting from the linear kernel $\kernel_0(x, x') = x^\top x'$:
\begin{itemize}
    \item $\exp(\kernel_0) = \exp(x^\top x')$ is pd by (i) with
    $f = \exp$.
    \item $(1 + \kernel_0)^p$ is pd by (i) with
    $f(t) = (1 + t)^p = \sum_{j=0}^p \binom{p}{j} t^j$.
    \item $\frac{1}{1 - c \kernel_0}$ for $0 < c < 1/\sup_x \|x\|^2$
    is pd by (i) with $f(t) = 1/(1 - ct) = \sum_{p=0}^\infty c^p t^p$.
\end{itemize}
\end{example}

\begin{warning}[title={Kernel Selection Matters}]
The choice of kernel encodes prior assumptions about the target
function's smoothness. The Gaussian kernel with small bandwidth
$\sigma$ produces very flexible models (low bias, high variance),
while large $\sigma$ yields smoother models (high bias, low
variance). The Laplacian kernel produces rougher functions than
the Gaussian kernel (functions in the Laplacian RKHS are only
once differentiable, while the Gaussian RKHS contains
$C^\infty$ functions). Kernel selection should be guided by
domain knowledge about the target function's regularity.
\end{warning}

With a rich toolkit for constructing kernels, we are ready to
develop the Hilbert space of functions associated with each kernel.
This is the subject of Lesson~62.

%% ============================================================
%% SECTION 7: EXERCISES
%% ============================================================

\section{Exercises}
\label{sec:exercises}

\begin{exercise}[Positive Definiteness Verification]
\textnormal{(\(*\))} \\
Let $\kernel(x, x') = \min(x, x')$ for $x, x' \in [0, \infty)$.
Show that $\kernel$ is positive definite by exhibiting an explicit
feature map $\feature \colon [0, \infty) \to L^2([0, \infty))$
such that $\kernel(x, x') = \langle \feature(x), \feature(x')
\rangle_{L^2}$.

\textit{Hint:} Consider $\feature(x) = \indicator_{[0,x]}$.
\end{exercise}

\begin{exercise}[Gram Matrix Computation]
\textnormal{(\(*\))} \\
For the data points $x_1 = (1, 0)^\top$, $x_2 = (0, 1)^\top$,
$x_3 = (1, 1)^\top$ in $\R^2$, compute the $3 \times 3$ Gram
matrix $K$ for:
\begin{enumerate}[label=(\alph*)]
    \item the polynomial kernel $\kernel(x, x') = (1 + x^\top x')^2$;
    \item the Gaussian RBF kernel $\kernel(x, x') = \exp(-\|x - x'\|^2 / 2)$.
\end{enumerate}
Verify that both matrices are positive semidefinite by computing
their eigenvalues.
\end{exercise}

\begin{exercise}[Closure Under Exponentiation]
\textnormal{(\(**\))} \\
Let $\kernel$ be a positive definite kernel on $\mathcal{X}$.
Prove that for any $c > 0$, the kernel $\kernel_c(x, x') =
\exp(c \, \kernel(x, x'))$ is positive definite. Use this to give
an alternative proof that the Gaussian kernel is positive definite.

\textit{Hint:} First show $\exp(c \, x^\top x')$ is pd using the
Taylor series, then relate the Gaussian kernel to this.
\end{exercise}

\begin{exercise}[Non-Example: The Distance Kernel]
\textnormal{(\(**\))} \\
Show that $\kernel(x, x') = \|x - x'\|$ on $\R^d$ is \emph{not}
positive definite for $d \ge 1$. Find the smallest $n$ and a
specific point configuration $\{x_1, \ldots, x_n\} \subset \R$ for
which the Gram matrix has a negative eigenvalue.
\end{exercise}

\begin{exercise}[Kernel PCA Feature Extraction]
\textnormal{(\(**\))} \\
Given data $x_1, \ldots, x_n \in \R^d$ and a pd kernel $\kernel$,
the kernel PCA algorithm extracts features from the eigendecomposition
of the centered Gram matrix $\tilde{K} = HKH$ where $H = I -
\frac{1}{n}\mathbf{1}\mathbf{1}^\top$ is the centering matrix.
\begin{enumerate}[label=(\alph*)]
    \item Show that $\tilde{K}$ is positive semidefinite.
    \item If $\tilde{K} = U \Lambda U^\top$ is the eigendecomposition
    with $\Lambda = \diag(\mu_1, \ldots, \mu_n)$ and $\mu_1 \ge \cdots
    \ge \mu_n \ge 0$, show that the $j$-th kernel PCA coordinate of
    $x_i$ is $\sqrt{\mu_j} \, U_{ij}$.
    \item Explain why kernel PCA with a Gaussian kernel can separate
    data that linear PCA cannot.
\end{enumerate}
\end{exercise}

\begin{exercise}[Random Features Approximation]
\textnormal{(\(***\))} \\
\citet{rahimi2008} showed that for shift-invariant kernels
$\kernel(x, x') = h(x - x')$ on $\R^d$, one can approximate the
kernel using random features. By Bochner's theorem, $h$ is the
Fourier transform of a nonneg\-ative measure $\Lambda$:
$h(\delta) = \int_{\R^d} e^{i \omega^\top \delta} \, d\Lambda(\omega)$.
\begin{enumerate}[label=(\alph*)]
    \item Show that if $\omega \sim \Lambda / \Lambda(\R^d)$, then
    $\kernel(x, x') = \Lambda(\R^d) \cdot \E[\cos(\omega^\top x)
    \cos(\omega^\top x') + \sin(\omega^\top x) \sin(\omega^\top x')]$.
    \item Define the random feature map $\hat{\feature}(x) =
    \sqrt{2/D} \, (\cos(\omega_1^\top x), \ldots, \cos(\omega_D^\top x))$
    where $\omega_1, \ldots, \omega_D \sim \Lambda$ i.i.d. Show that
    $\E[\hat{\feature}(x)^\top \hat{\feature}(x')] = \kernel(x, x')$.
    \item Discuss the tradeoff between the number of random features
    $D$ and the approximation quality. How does this relate to the
    kernel trick vs.\ explicit feature maps?
\end{enumerate}
\end{exercise}

%% ============================================================
%% SECTION 8: SUMMARY
%% ============================================================

\section{Summary}
\label{sec:summary}

\begin{itemize}
    \item The \textbf{kernel trick} replaces inner products in a
    (possibly infinite-dimensional) feature space with evaluations
    of a kernel function $\kernel(x, x')$, reducing computation from
    $O(D)$ in feature dimension to $O(d)$ in input dimension.

    \item A \textbf{positive definite kernel} is a symmetric function
    whose Gram matrix is positive semidefinite for every finite point
    set. This is precisely the class of functions that arise as inner
    products in some feature space.

    \item The class of pd kernels is \textbf{closed under sums,
    products, scalar multiples, and pointwise limits}
    (Theorem~\ref{thm:pd-closure}), enabling systematic kernel
    construction from simple building blocks.

    \item \textbf{Mercer's theorem} provides a spectral decomposition
    $\kernel(x, x') = \sum_j \lambda_j \feature_j(x) \feature_j(x')$
    for continuous pd kernels on compact domains, with uniform
    convergence and nonneg\-ative eigenvalues.

    \item The \textbf{canonical feature map} $\Phi(x) =
    (\sqrt{\lambda_j} \feature_j(x))_{j \ge 1}$ maps into $\ell^2$
    and recovers the kernel as an inner product: $\kernel(x, x') =
    \langle \Phi(x), \Phi(x') \rangle_{\ell^2}$.

    \item The \textbf{Gaussian RBF kernel} $\exp(-\|x - x'\|^2 /
    (2\sigma^2))$ is positive definite (proved via Taylor expansion
    of $\exp$), has an infinite-dimensional feature space, and its
    eigenvalues decay exponentially.

    \item \textbf{Kernel selection} encodes smoothness assumptions:
    the Gaussian kernel produces $C^\infty$ functions, the Laplacian
    kernel produces rougher functions, and polynomial kernels
    constrain the function class to a finite-dimensional space.
\end{itemize}

%% ============================================================
%% SECTION 9: FURTHER READING
%% ============================================================

\section{Further Reading}
\label{sec:further-reading}

\begin{itemize}
    \item \textbf{Lesson 62} develops the reproducing kernel Hilbert
    space (RKHS) associated with each pd kernel, establishes the
    representer theorem, and applies it to kernel ridge regression.
    The RKHS provides the function space in which kernel methods
    operate.

    \item \citet{scholkopf2002} provide a comprehensive treatment of
    kernel methods for machine learning, including support vector
    machines, kernel PCA, and kernel Fisher discriminant analysis.

    \item \citet{steinwart2008} give a rigorous mathematical
    treatment of kernels and support vector machines, including
    universal kernels and approximation properties.

    \item \citet{rahimi2008} introduce random features as a
    scalable approximation to kernel methods, bridging the gap
    between explicit feature maps and the kernel trick.

    \item \citet{jacot2018} establish the neural tangent kernel
    framework, connecting wide neural networks to kernel methods and
    enabling the analysis of neural network training via kernel theory.

    \item The original paper by \citet{mercer1909} established the
    eigenexpansion of positive definite integral kernels in the
    context of integral equations, decades before the machine learning
    applications. \citet{aronszajn1950} later developed the abstract
    RKHS theory that unified and generalized Mercer's work.
\end{itemize}

\bibliography{nn-refs}
\end{document}
