\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 62: Cooperative Multi-Agent Reinforcement Learning}}
\fancyhead[R]{\thepage}
\title{Lesson 62: Cooperative Multi-Agent\\Reinforcement Learning}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% ============================================================
%% SECTION 1: MOTIVATION AND OVERVIEW
%% ============================================================
\section{Motivation and Overview}
\label{sec:motivation}

Consider a team of $N$ warehouse robots operating in a shared facility.
Each robot must pick up packages from storage shelves and deliver them
to shipping docks.  The team receives a single shared reward: the total
number of packages delivered per hour.  Each robot observes only its
local surroundings---nearby shelves, packages in its gripper, and
robots within sensor range---yet must coordinate with the rest of
the team to avoid collisions, balance workload across docks, and
maximize throughput.  A centralized controller that observes the
entire warehouse could, in principle, compute an optimal joint
policy; but communication latency and bandwidth constraints
require each robot to act based on local observations alone.

This tension between \emph{global cooperation} and \emph{local
execution} lies at the heart of cooperative multi-agent reinforcement
learning (MARL).  Formalizing the problem leads to the Decentralized
Partially Observable Markov Decision Process (Dec-POMDP), a model
so general that finding an optimal joint policy is NEXP-complete.
To make the problem tractable, practical algorithms adopt the
Centralized Training with Decentralized Execution (CTDE) paradigm:
during training, a centralized process has access to global state
and joint actions; during execution, each agent acts using only its
local observation history.  The tools developed in this lesson will
let us formalize the Dec-POMDP model, understand its computational
barriers, and derive the value decomposition and policy gradient
methods that make cooperative MARL feasible.

\begin{rlconnection}[title={Why Cooperative MARL Matters}]
\begin{itemize}[nosep]
  \item \textbf{StarCraft micromanagement (SMAC).}  Teams of units
    must coordinate attack focus and positioning using only local
    observations; QMIX and MAPPO achieve state-of-the-art performance
    on this benchmark \citep{samvelyan2019}.
  \item \textbf{Autonomous vehicle platooning.}  Vehicles in a platoon
    share a common fuel-efficiency objective but each vehicle controls
    only its own throttle and braking based on local sensor readings.
  \item \textbf{Robotic manipulation teams.}  Multiple robotic arms
    assembling a product must coordinate grasps and placements;
    credit assignment determines which arm's action contributed to
    success or failure.
  \item \textbf{Network routing.}  Routers in a communication network
    cooperatively minimize global latency while each router observes
    only local queue lengths.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
Stochastic games and Markov perfect equilibrium (Lesson~61,
Definitions~3.1--3.3).  MDP formulation, Bellman equations, and
value iteration (Lessons~39--42).  Policy gradient theorem and
actor-critic methods (Lessons~54--57).  PPO clipped surrogate
objective (Lesson~57, Section~4 in Phase~04).  Nash equilibrium
and minimax theorem (Lesson~20, Phase~01 game theory).

\paragraph{Learning objectives.}
After completing this lesson, you will be able to:
\begin{itemize}[nosep]
  \item \textbf{Define} the Dec-POMDP model and \textbf{explain}
    how it specializes the stochastic game framework to the
    cooperative setting.
  \item \textbf{State} and \textbf{explain} the NEXP-completeness
    result for optimal Dec-POMDP policy computation.
  \item \textbf{Derive} the VDN additive value decomposition and
    \textbf{prove} that greedy action selection decomposes under it.
  \item \textbf{State} the Individual-Global-Max (IGM) condition
    and \textbf{prove} that the QMIX monotonicity condition implies
    IGM.
  \item \textbf{Analyze} the representational hierarchy
    VDN $\subset$ QMIX $\subset$ QTRAN and \textbf{construct}
    examples demonstrating strict inclusions.
  \item \textbf{Describe} the MAPPO algorithm and \textbf{explain}
    how centralized advantage estimation supports decentralized
    policy improvement.
\end{itemize}

%% ============================================================
%% SECTION 2: REFERENCES
%% ============================================================
\section{References}
\label{sec:references}

\begin{itemize}
  \item \citet{bernstein2002} -- Original proof of NEXP-completeness
    for Dec-POMDPs; the foundational complexity result of Section~\ref{sec:decpomdp}.
  \item \citet[Chapters 2--5]{oliehoek2016} -- Comprehensive textbook
    treatment of Dec-POMDPs, including exact and approximate solution
    methods.
  \item \citet{rashid2018} -- Introduces QMIX and the monotonicity
    condition for value factorization; primary reference for Section~\ref{sec:qmix}.
  \item \citet[Sections 4--5]{zhang2021} -- Survey of cooperative MARL
    algorithms covering value decomposition and policy gradient methods.
  \item \citet{yu2022} -- Empirical and theoretical analysis of MAPPO;
    demonstrates the surprising effectiveness of centralized-critic PPO
    in cooperative games.
\end{itemize}

%% ============================================================
%% SECTION 3: DEC-POMDP
%% ============================================================
\section{Decentralized Partially Observable MDP}
\label{sec:decpomdp}

The cooperative multi-agent problem is formalized by the Dec-POMDP,
which extends the single-agent POMDP to $N$ agents sharing a common
reward.

\begin{definition}[Decentralized POMDP]
\label{def:decpomdp}
A \emph{Decentralized Partially Observable Markov Decision Process}
(Dec-POMDP) is a tuple
\[
  \mathcal{M} = (\Players, \States, \{\Actions_i\}_{i \in \Players},
  \Trans, R, \{\Omega_i\}_{i \in \Players}, O, \discount),
\]
where:
\begin{itemize}[nosep]
  \item $\Players = \{1, \ldots, N\}$ is the finite set of agents.
  \item $\States$ is a finite set of global states.
  \item $\Actions_i$ is the finite action set of agent $i$; the joint
    action space is $\JointActions = \Actions_1 \times \cdots \times \Actions_N$.
  \item $\Trans : \States \times \JointActions \to \simplex(\States)$
    is the state transition function, where $\Trans(s' \mid s, \mathbf{a})$
    is the probability of transitioning to $s'$ given state $s$ and
    joint action $\mathbf{a} = (a_1, \ldots, a_N)$.
  \item $R : \States \times \JointActions \to \R$ is the shared
    (team) reward function.
  \item $\Omega_i$ is the finite observation set of agent $i$, with
    joint observation space $\boldsymbol{\Omega} = \Omega_1 \times
    \cdots \times \Omega_N$.
  \item $O : \States \times \JointActions \to \simplex(\boldsymbol{\Omega})$
    is the observation function, where $O(\mathbf{o} \mid s', \mathbf{a})$
    gives the probability of joint observation $\mathbf{o} = (o_1,
    \ldots, o_N)$ after transitioning to $s'$ via action $\mathbf{a}$.
  \item $\discount \in [0, 1)$ is the discount factor.
\end{itemize}
\end{definition}

\begin{remark}
\label{rem:decpomdp-vs-sg}
A Dec-POMDP is a special case of a partially observable stochastic
game (POSG, Lesson~61) in which all agents share the same reward
function $R$.  This common-payoff structure makes the problem
\emph{cooperative}: the agents' interests are perfectly aligned.
When $N = 1$, the Dec-POMDP reduces to a standard POMDP; when
observations fully reveal the state ($\Omega_i = \States$ and
$O$ is the identity), it reduces to a multi-agent MDP.
\end{remark}

\paragraph{Observation histories and local policies.}
Since agent $i$ cannot observe the global state, it must base
decisions on its \emph{observation-action history}.  At time $t$,
agent $i$ has observed the sequence
\[
  \tau_i^t = (o_i^0, a_i^0, o_i^1, a_i^1, \ldots, o_i^{t-1},
  a_i^{t-1}, o_i^t) \in (\Omega_i \times \Actions_i)^t \times \Omega_i.
\]
We write $\mathcal{T}_i^t$ for the set of all such histories of
length $t$, and $\mathcal{T}_i = \bigcup_{t=0}^{\infty} \mathcal{T}_i^t$
for the set of all finite histories.

\begin{definition}[Local and Joint Policies]
\label{def:local-joint-policy}
A \emph{local (decentralized) policy} for agent $i$ is a mapping
\[
  \policyi : \mathcal{T}_i \to \simplex(\Actions_i),
\]
which assigns a distribution over actions to each observation-action
history.  A \emph{joint policy} is the tuple
$\policyj = (\policy_1, \ldots, \policy_N)$.
\end{definition}

\begin{definition}[Joint Value Function]
\label{def:joint-value}
Given a joint policy $\policyj$ and initial state $s_0 \in \States$,
the \emph{joint value function} is
\[
  V^{\policyj}(s_0) = \E^{\policyj}\!\left[
    \sum_{t=0}^{\infty} \discount^t\, R(s_t, \mathbf{a}_t)
    \;\middle|\; s_0
  \right],
\]
where the expectation is over the stochastic process induced by
the transition kernel $\Trans$, observation function $O$, and
local policies $\policyi$.
\end{definition}

The optimization problem is to find a joint policy $\policyj^*$
that maximizes $V^{\policyj}(s_0)$ for every initial state $s_0$
(or, in the finite-horizon case, for a given initial state
distribution).

\begin{theorem}[NEXP-Completeness of Dec-POMDP]
\label{thm:nexp}
The problem of determining whether there exists a joint policy for
a finite-horizon Dec-POMDP whose expected total reward exceeds a
given threshold $k$ is NEXP-complete.
\end{theorem}

\begin{proof}[Proof sketch]
We outline the two directions following \citet{bernstein2002}.

\emph{Membership in NEXP.}  A nondeterministic exponential-time
algorithm can guess a joint policy (which, for a finite horizon
$T$, is specified by at most $N$ decision trees each of depth $T$,
hence of exponential size) and verify it in exponential time by
computing the expected reward via dynamic programming over the
joint belief space.

\emph{NEXP-hardness.}  The reduction proceeds from the
NEXP-complete \emph{tiling problem}: given a finite set of tile
types with matching constraints on edges, determine whether an
exponentially large grid $2^n \times 2^n$ can be validly tiled.
The construction encodes the grid into a Dec-POMDP with two agents.
Agent~1 is responsible for rows and agent~2 for columns.  The
global state encodes the current grid position $(i, j)$.  Each
agent observes only its respective coordinate ($i$ or $j$) and
must output a tile type for that position.  The reward function
assigns $+1$ if the chosen tile is consistent with all adjacency
constraints for position $(i, j)$ and $0$ otherwise.  The key
difficulty is that the agents cannot communicate their tile
choices to each other, so a valid tiling can be achieved if and
only if there exist local policies (deterministic history-dependent
strategies) that jointly produce a valid tiling.  Since the grid
has $2^n \times 2^n$ cells, the observation histories have
exponential length, matching the NEXP complexity.

For the complete formal construction, see \citet[Theorem~3.1]{bernstein2002}.
\end{proof}

\begin{keyresult}[title={Computational Intractability of Dec-POMDPs}]
Finding an optimal joint policy for a Dec-POMDP is NEXP-complete
(Theorem~\ref{thm:nexp}).  This stands in contrast to:
\begin{itemize}[nosep]
  \item Single-agent MDPs: solvable in polynomial time (P) via
    linear programming \citep{puterman2014}.
  \item Single-agent POMDPs: PSPACE-complete.
  \item Centralized multi-agent MDPs (full observability, common
    reward): solvable in P as a single large MDP.
\end{itemize}
The source of intractability is the combination of
\emph{decentralization} (local policies) and \emph{partial
observability} (agents cannot condition on global state).
\end{keyresult}

\begin{example}[Two-Agent Dec-POMDP: Cooperative Tiger]
\label{ex:tiger}
Consider a variant of the classic tiger problem.  Two agents stand
before two doors.  Behind one door is a tiger (penalty $-50$); behind
the other is a reward ($+10$).  Each agent independently listens
(receiving a noisy signal about the tiger's location with accuracy
$0.85$) or opens a door.  If either agent opens the tiger door,
both receive $-50$; if either opens the reward door (and neither
opens the tiger door), both receive $+10$; if both listen, both
receive $-1$.

Even in this two-agent, two-state, three-action problem, the
optimal joint policy requires agents to coordinate the number of
listening steps before committing.  Agent~1 cannot condition on
agent~2's observations, so the optimal policy must balance
information gathering against the risk of uncoordinated door
opening.  Solving this optimally requires enumerating all possible
joint observation histories---an exponential computation even for
moderate horizons.
\end{example}

We now turn to practical frameworks that sidestep the NEXP barrier
by exploiting additional structure available during training.

%% ============================================================
%% SECTION 4: CTDE
%% ============================================================
\section{Centralized Training with Decentralized Execution}
\label{sec:ctde}

The NEXP-completeness of Dec-POMDPs (Theorem~\ref{thm:nexp})
motivates approximation frameworks that exploit asymmetry between
training and execution phases.

\begin{definition}[CTDE Paradigm]
\label{def:ctde}
The \emph{Centralized Training with Decentralized Execution}
(CTDE) paradigm is a two-phase approach to cooperative MARL:
\begin{enumerate}[nosep]
  \item \textbf{Training phase.}  A centralized learning algorithm
    has access to global information: the state $s_t$, all agents'
    actions $\mathbf{a}_t = (a_1^t, \ldots, a_N^t)$, all observations
    $\mathbf{o}_t = (o_1^t, \ldots, o_N^t)$, and the shared reward
    $r_t = R(s_t, \mathbf{a}_t)$.
  \item \textbf{Execution phase.}  Each agent $i$ executes a
    \emph{decentralized policy} $\policyi(\cdot \mid \tau_i)$ that
    conditions only on its local observation-action history $\tau_i$.
\end{enumerate}
\end{definition}

The CTDE framework does \emph{not} change the Dec-POMDP model or
its complexity class.  Rather, it relaxes the \emph{learning}
problem: during training, we may use global information to shape
the learning signal (e.g., a centralized critic), provided the
resulting policies are decentralizable.

\paragraph{Centralized critic, decentralized actors.}
A natural instantiation of CTDE uses a \emph{centralized value
function} $V(s)$ or $Q(s, \mathbf{a})$ that conditions on the
global state (or joint observations) to provide low-variance
gradient estimates.  Each agent $i$ maintains a local actor
$\policyi(\cdot \mid \tau_i)$ that is updated using policy
gradient or value-based methods with the centralized critic
as the learning signal.

\begin{rlconnection}[title={CTDE as the Dominant Paradigm}]
Virtually all modern cooperative MARL algorithms adopt the CTDE
paradigm, including:
\begin{itemize}[nosep]
  \item \textbf{Value decomposition} (VDN, QMIX, QTRAN): train a
    centralized $\Qtot$ that factorizes into per-agent utilities.
  \item \textbf{Centralized critic methods} (MAPPO, MADDPG): train
    a centralized value function to reduce variance of decentralized
    policy gradients.
  \item \textbf{Communication methods} (CommNet, TarMAC): train
    communication protocols that enable richer coordination while
    keeping execution decentralized.
\end{itemize}
CTDE is successful because modern training environments (simulators)
provide full state access, while real-world deployment demands
decentralization due to communication constraints.
\end{rlconnection}

The remainder of this lesson develops two families of CTDE
algorithms: value decomposition methods (Sections~\ref{sec:vdn}--\ref{sec:qtran})
and policy gradient methods (Section~\ref{sec:mappo}).

%% ============================================================
%% SECTION 5: VDN
%% ============================================================
\section{Value Decomposition Networks (VDN)}
\label{sec:vdn}

The first approach to cooperative MARL under CTDE decomposes the
joint action-value function into per-agent utilities, enabling
decentralized greedy action selection.

\paragraph{Problem setup.}
Suppose the team operates in a multi-agent MDP (fully observable
variant of Dec-POMDP) or that each agent maintains an observation
history $\tau_i$.  We seek a joint action-value function
$\Qtot(\boldsymbol{\tau}, \mathbf{a})$ where $\boldsymbol{\tau}
= (\tau_1, \ldots, \tau_N)$ and $\mathbf{a} = (a_1, \ldots, a_N)$.
For decentralized execution, each agent $i$ must be able to select
$\actioni$ using only $\tau_i$, without knowledge of other agents'
histories or actions.

\begin{definition}[Additive Value Decomposition]
\label{def:vdn}
The \emph{Value Decomposition Network} (VDN) \citep{sunehag2018}
decomposes the joint action-value function as an additive sum:
\[
  \Qtot(\boldsymbol{\tau}, \mathbf{a})
  = \sum_{i=1}^{N} Q_i(\tau_i, \actioni),
\]
where each $Q_i : \mathcal{T}_i \times \Actions_i \to \R$ is
agent $i$'s individual utility function.
\end{definition}

\begin{theorem}[Decentralized Greedy Action Selection under VDN]
\label{thm:vdn-greedy}
Under the additive decomposition of Definition~\ref{def:vdn},
the greedy joint action decomposes into individual greedy actions:
\[
  \argmax_{\mathbf{a} \in \JointActions}
  \Qtot(\boldsymbol{\tau}, \mathbf{a})
  = \left(
    \argmax_{a_1 \in \Actions_1} Q_1(\tau_1, a_1),\;
    \ldots,\;
    \argmax_{a_N \in \Actions_N} Q_N(\tau_N, a_N)
  \right).
\]
\end{theorem}

\begin{proof}
Since $\Qtot(\boldsymbol{\tau}, \mathbf{a}) = \sum_{i=1}^N
Q_i(\tau_i, \actioni)$, and each term $Q_i(\tau_i, \actioni)$
depends only on $\actioni$, the joint maximization decomposes
into $N$ independent maximizations:
\[
  \max_{\mathbf{a} \in \JointActions}
  \sum_{i=1}^N Q_i(\tau_i, \actioni)
  = \sum_{i=1}^N \max_{\actioni \in \Actions_i} Q_i(\tau_i, \actioni).
\]
This equality holds because the feasible set $\JointActions =
\Actions_1 \times \cdots \times \Actions_N$ is a Cartesian product,
and the objective is separable.  Any joint action
$\mathbf{a}^* = (a_1^*, \ldots, a_N^*)$ achieving the left-hand
side must satisfy $a_i^* \in \argmax_{\actioni} Q_i(\tau_i, \actioni)$
for each $i$, since otherwise replacing $a_i^*$ with a maximizer
of $Q_i$ would strictly increase the sum.
\end{proof}

\begin{example}[VDN in a Two-Agent Delivery Task]
\label{ex:vdn-works}
Consider two robots ($N = 2$) delivering packages in a grid.
Robot~1 operates in the left half and robot~2 in the right half,
with minimal interaction.  Let $Q_1(\tau_1, a_1)$ represent the
expected packages robot~1 delivers, and similarly for $Q_2$.
Since the robots operate in disjoint regions, the team reward
is approximately $R \approx R_1 + R_2$, so the true $\Qtot$
is well-approximated by an additive decomposition.  Each robot
independently maximizes its own $Q_i$, and the joint action is
jointly optimal.
\end{example}

\begin{example}[Failure of VDN: Non-Additive Coordination]
\label{ex:vdn-fails}
Consider a two-agent, two-action coordination game with the
following joint reward matrix:
\[
\begin{array}{c|cc}
  R(a_1, a_2) & a_2 = L & a_2 = R \\
  \hline
  a_1 = L & 8 & 0 \\
  a_1 = R & 0 & 5 \\
\end{array}
\]
The optimal joint action is $(L, L)$ with reward $8$.  Suppose
VDN decomposes as $\Qtot(a_1, a_2) = Q_1(a_1) + Q_2(a_2)$.
Then we need $Q_1(L) + Q_2(L) = 8$ and $Q_1(R) + Q_2(R) = 5$,
but also $Q_1(L) + Q_2(R) = 0$ and $Q_1(R) + Q_2(L) = 0$.
From these four equations: $Q_1(L) - Q_1(R) = 8 - 0 = 8$
(first and fourth) but also $Q_1(L) - Q_1(R) = 0 - 5 = -5$
(third and second), a contradiction.  Thus VDN \emph{cannot}
represent this game's payoff structure.
\end{example}

The failure in Example~\ref{ex:vdn-fails} arises because the
reward exhibits \emph{interaction effects} between agents'
actions.  This motivates richer decomposition classes.

%% ============================================================
%% SECTION 6: QMIX AND IGM
%% ============================================================
\section{QMIX and the IGM Condition}
\label{sec:qmix}

To overcome the representational limitations of VDN while preserving
decentralized action selection, \citet{rashid2018} introduced QMIX,
which relies on the \emph{Individual-Global-Max} (IGM) condition.

\begin{definition}[Individual-Global-Max (IGM) Condition]
\label{def:igm}
A factorization $(\Qtot, Q_1, \ldots, Q_N)$ satisfies the
\emph{Individual-Global-Max} (IGM) condition if for all
$\boldsymbol{\tau}$:
\[
  \argmax_{\mathbf{a} \in \JointActions}
  \Qtot(\boldsymbol{\tau}, \mathbf{a})
  = \left(
    \argmax_{a_1 \in \Actions_1} Q_1(\tau_1, a_1),\;
    \ldots,\;
    \argmax_{a_N \in \Actions_N} Q_N(\tau_N, a_N)
  \right).
\]
That is, the globally optimal joint action can be recovered by
each agent independently maximizing its own utility $Q_i$.
\end{definition}

IGM is the minimal structural condition needed for decentralized
execution: if it holds, each agent can compute its part of the
optimal joint action using only local information.  The challenge
is to design $\Qtot$ as a function of $Q_1, \ldots, Q_N$ such
that IGM is guaranteed.

\begin{definition}[Monotonicity Condition]
\label{def:monotonicity}
A mixing function $\Qtot(\boldsymbol{\tau}, \mathbf{a}) =
f(Q_1(\tau_1, a_1), \ldots, Q_N(\tau_N, a_N); \boldsymbol{\tau})$
satisfies the \emph{monotonicity condition} if
\[
  \frac{\partial \Qtot}{\partial Q_i} \geq 0
  \quad \text{for all } i \in \Players.
\]
\end{definition}

\begin{proposition}[Monotonicity Implies IGM]
\label{prop:mono-igm}
If $\Qtot = f(Q_1, \ldots, Q_N; \boldsymbol{\tau})$ satisfies
the monotonicity condition (Definition~\ref{def:monotonicity}),
then the IGM condition (Definition~\ref{def:igm}) holds.
\end{proposition}

\begin{proof}
Let $\bar{a}_i \in \argmax_{\actioni \in \Actions_i} Q_i(\tau_i,
\actioni)$ for each $i$, and let $\bar{\mathbf{a}} = (\bar{a}_1,
\ldots, \bar{a}_N)$.  We must show that $\bar{\mathbf{a}} \in
\argmax_{\mathbf{a}} \Qtot(\boldsymbol{\tau}, \mathbf{a})$.

Take any $\mathbf{a} = (a_1, \ldots, a_N) \in \JointActions$.
We construct a sequence of joint actions interpolating from
$\mathbf{a}$ to $\bar{\mathbf{a}}$ by replacing one component
at a time.  Define $\mathbf{a}^{(0)} = \mathbf{a}$ and for
$k = 1, \ldots, N$:
\[
  \mathbf{a}^{(k)} = (\bar{a}_1, \ldots, \bar{a}_k, a_{k+1},
  \ldots, a_N).
\]
So $\mathbf{a}^{(N)} = \bar{\mathbf{a}}$.  Consider the
transition from $\mathbf{a}^{(k-1)}$ to $\mathbf{a}^{(k)}$,
which replaces $a_k$ with $\bar{a}_k$.  Since $\bar{a}_k$
maximizes $Q_k(\tau_k, \cdot)$, we have $Q_k(\tau_k, \bar{a}_k)
\geq Q_k(\tau_k, a_k)$.  The monotonicity condition ensures
$\Qtot$ is non-decreasing in each $Q_i$.  Therefore, increasing
(or not changing) $Q_k$ while holding all other $Q_j$ fixed
cannot decrease $\Qtot$:
\[
  \Qtot(\boldsymbol{\tau}, \mathbf{a}^{(k)})
  = f(\ldots, Q_k(\tau_k, \bar{a}_k), \ldots)
  \geq f(\ldots, Q_k(\tau_k, a_k), \ldots)
  = \Qtot(\boldsymbol{\tau}, \mathbf{a}^{(k-1)}).
\]
Chaining over $k = 1, \ldots, N$:
\[
  \Qtot(\boldsymbol{\tau}, \bar{\mathbf{a}})
  = \Qtot(\boldsymbol{\tau}, \mathbf{a}^{(N)})
  \geq \Qtot(\boldsymbol{\tau}, \mathbf{a}^{(0)})
  = \Qtot(\boldsymbol{\tau}, \mathbf{a}).
\]
Since $\mathbf{a}$ was arbitrary, $\bar{\mathbf{a}}$ maximizes
$\Qtot$.
\end{proof}

\paragraph{QMIX architecture.}
QMIX \citep{rashid2018} instantiates the monotonicity condition
using a \emph{mixing network}.  Each agent $i$ maintains a utility
network $Q_i(\tau_i, \actioni; \phi_i)$ parameterized by $\phi_i$.
The mixing network computes:
\begin{equation}
\label{eq:qmix}
  \Qtot(\boldsymbol{\tau}, \mathbf{a}; s)
  = f_{\mathrm{mix}}\!\big(Q_1(\tau_1, a_1), \ldots,
    Q_N(\tau_N, a_N);\, s\big),
\end{equation}
where $f_{\mathrm{mix}}$ is a feedforward network whose weights
are generated by \emph{hypernetworks} that take the global state
$s$ as input.  The critical architectural choice is that the
hypernetworks produce mixing weights via an absolute value
activation:
\[
  W_{\ell}(s) = |g_{\ell}(s)|,
\]
where $g_{\ell}$ is a learnable function of $s$ and $|\cdot|$
denotes element-wise absolute value.  This ensures $W_{\ell}(s)
\geq 0$, so each layer of the mixing network is a non-negative
linear combination (plus bias) followed by an ELU activation.
Since compositions of non-decreasing functions are non-decreasing,
the overall $f_{\mathrm{mix}}$ satisfies $\partial \Qtot /
\partial Q_i \geq 0$.

\begin{theorem}[QMIX Strictly Generalizes VDN]
\label{thm:qmix-superset}
The class of $\Qtot$ functions representable by QMIX strictly
contains the class representable by VDN.  That is, every VDN
decomposition is a special case of QMIX, and there exist $\Qtot$
functions representable by QMIX but not by VDN.
\end{theorem}

\begin{proof}
\emph{VDN $\subseteq$ QMIX.}  The VDN decomposition
$\Qtot = \sum_i Q_i$ is a special case of a monotonic mixing
function: take $f_{\mathrm{mix}}(q_1, \ldots, q_N) =
\sum_i q_i$, which is a single-layer linear network with all
weights equal to $1$ and zero bias.  This satisfies $\partial
\Qtot / \partial Q_i = 1 > 0$.

\emph{Strict inclusion.}  Consider $N = 2$ agents with
$Q_1, Q_2 \in \R$ and the mixing function
$\Qtot = Q_1 \cdot Q_2$ (product, not sum).  When $Q_1, Q_2 > 0$,
we have $\partial \Qtot / \partial Q_i = Q_j > 0$, satisfying
monotonicity.  This is representable by a QMIX network (a
two-layer mixing network can implement multiplication for
positive inputs).  However, $\Qtot = Q_1 \cdot Q_2$ is not
additive: there exist no functions $f, g$ such that
$Q_1 \cdot Q_2 = f(Q_1) + g(Q_2)$ for all $Q_1, Q_2 > 0$.
To verify, suppose $Q_1 Q_2 = f(Q_1) + g(Q_2)$.  Setting
$Q_2 = 1$: $Q_1 = f(Q_1) + g(1)$, so $f(Q_1) = Q_1 - c$
for $c = g(1)$.  Setting $Q_1 = 1$: $Q_2 = (1 - c) + g(Q_2)$,
so $g(Q_2) = Q_2 - 1 + c$.  Then $f(Q_1) + g(Q_2) =
Q_1 + Q_2 - 1 \neq Q_1 Q_2$ in general.
\end{proof}

\begin{keyresult}[title={IGM and Monotonic Value Factorization}]
\begin{itemize}[nosep]
  \item The \textbf{IGM condition} (Definition~\ref{def:igm})
    is the necessary and sufficient structural property for
    decentralized greedy action selection to recover the jointly
    optimal action.
  \item The \textbf{monotonicity condition} (Definition~\ref{def:monotonicity})
    is a sufficient condition for IGM that is architecturally
    enforceable via non-negative mixing weights.
  \item QMIX enforces monotonicity through hypernetwork-generated
    non-negative weights, yielding a class strictly larger than
    VDN (Theorem~\ref{thm:qmix-superset}).
\end{itemize}
\end{keyresult}

\begin{warning}[title={Limitations of Monotonicity}]
The monotonicity condition prevents QMIX from representing
\emph{non-monotonic} interactions between agents.  Consider
a game where the optimal strategy requires agent~1 to take
action $L$ when agent~2 takes $L$, but action $R$ when
agent~2 takes $R$---and the joint Q-values are such that
increasing $Q_2$ for action $L$ would require decreasing
$\Qtot$ for some joint actions.  Formally, any game matrix
where $\Qtot(L, L) > \Qtot(R, L)$ but $\Qtot(L, R) <
\Qtot(R, R)$ requires $\partial \Qtot / \partial Q_1$ to
change sign, violating monotonicity.
\end{warning}

\begin{example}[QMIX on a Predator-Prey Task]
\label{ex:qmix-predprey}
Consider $N = 3$ predators on a grid who must surround a prey
to capture it.  The team reward is $+10$ for capture and $0$
otherwise.  Each predator $i$ observes a $5 \times 5$ local
patch and maintains $Q_i(\tau_i, \actioni)$ representing its
utility for moving in direction $\actioni \in \{N, S, E, W,
\text{stay}\}$.  The mixing network takes the global state
$s$ (all positions) and produces non-negative weights that
combine the $Q_i$ values.  Since successful capture requires
all predators to be adjacent to the prey, the reward is
\emph{super-additive}: the value of the team is greater than
the sum of individual values.  QMIX's state-dependent mixing
weights can capture this---when the global state shows two
predators already adjacent, the mixing network upweights the
third predator's contribution, reflecting the increased
marginal value of the final encirclement move.
\end{example}

%% ============================================================
%% SECTION 7: QTRAN
%% ============================================================
\section{QTRAN: Beyond Monotonic Decomposition}
\label{sec:qtran}

While QMIX achieves strong empirical results, its monotonicity
constraint limits the class of representable joint value functions.
\citet{son2019} introduced QTRAN, which achieves IGM without
requiring monotonicity.

\paragraph{Motivation.}
Consider a two-agent game where the optimal joint action is
$(L, L)$ but the off-diagonal entries of the reward matrix
violate monotonicity (as in the warning above).  QMIX cannot
represent such games, but IGM may still hold if we allow a
more flexible factorization.

\begin{definition}[QTRAN Factorization]
\label{def:qtran}
The QTRAN factorization introduces three components:
\begin{enumerate}[nosep]
  \item Individual utility functions $Q_i(\tau_i, \actioni)$
    for each agent $i$.
  \item A joint state-value function
    $V_{\mathrm{jt}}(\boldsymbol{\tau})$.
  \item Individual state-value functions $V_i(\tau_i)$ for
    each agent $i$.
\end{enumerate}
The factorization posits that for the optimal joint action
$\bar{\mathbf{a}}(\boldsymbol{\tau}) = (\bar{a}_1, \ldots,
\bar{a}_N)$ where $\bar{a}_i = \argmax_{\actioni} Q_i(\tau_i,
\actioni)$:
\begin{align}
  \Qtot(\boldsymbol{\tau}, \bar{\mathbf{a}})
  &= \sum_{i=1}^N Q_i(\tau_i, \bar{a}_i)
     + V_{\mathrm{jt}}(\boldsymbol{\tau})
     - \sum_{i=1}^N V_i(\tau_i),
     \label{eq:qtran-opt} \\
  \Qtot(\boldsymbol{\tau}, \mathbf{a})
  &\leq \sum_{i=1}^N Q_i(\tau_i, \actioni)
     + V_{\mathrm{jt}}(\boldsymbol{\tau})
     - \sum_{i=1}^N V_i(\tau_i),
     \quad \forall\, \mathbf{a} \neq \bar{\mathbf{a}}.
     \label{eq:qtran-subopt}
\end{align}
\end{definition}

\begin{proposition}[QTRAN Conditions Imply IGM]
\label{prop:qtran-igm}
If conditions~\eqref{eq:qtran-opt}--\eqref{eq:qtran-subopt}
hold, then the IGM condition (Definition~\ref{def:igm}) is
satisfied.
\end{proposition}

\begin{proof}
Let $\bar{\mathbf{a}}$ be the joint action with $\bar{a}_i =
\argmax_{\actioni} Q_i(\tau_i, \actioni)$ for each $i$.
We must show $\bar{\mathbf{a}} \in \argmax_{\mathbf{a}}
\Qtot(\boldsymbol{\tau}, \mathbf{a})$.

For any $\mathbf{a} \neq \bar{\mathbf{a}}$, using
\eqref{eq:qtran-opt} and \eqref{eq:qtran-subopt}:
\begin{align*}
  \Qtot(\boldsymbol{\tau}, \bar{\mathbf{a}})
  &= \sum_{i=1}^N Q_i(\tau_i, \bar{a}_i)
     + V_{\mathrm{jt}}(\boldsymbol{\tau})
     - \sum_{i=1}^N V_i(\tau_i) \\
  &\geq \sum_{i=1}^N Q_i(\tau_i, \actioni)
     + V_{\mathrm{jt}}(\boldsymbol{\tau})
     - \sum_{i=1}^N V_i(\tau_i) \\
  &\geq \Qtot(\boldsymbol{\tau}, \mathbf{a}).
\end{align*}
The first inequality uses $\bar{a}_i = \argmax_{\actioni}
Q_i(\tau_i, \actioni)$, so $\sum_i Q_i(\tau_i, \bar{a}_i)
\geq \sum_i Q_i(\tau_i, \actioni)$.  The second inequality
is condition~\eqref{eq:qtran-subopt} rearranged.
Therefore $\Qtot(\boldsymbol{\tau}, \bar{\mathbf{a}}) \geq
\Qtot(\boldsymbol{\tau}, \mathbf{a})$ for all $\mathbf{a}$,
establishing IGM.
\end{proof}

\paragraph{Practical implementation.}
The QTRAN conditions~\eqref{eq:qtran-opt}--\eqref{eq:qtran-subopt}
are difficult to enforce exactly.  In practice, \citet{son2019}
relax them into soft constraints within the loss function.  Define:
\begin{align*}
  \hat{Q}_{\mathrm{jt}}(\boldsymbol{\tau}, \mathbf{a})
  &= \sum_{i=1}^N Q_i(\tau_i, \actioni)
     + V_{\mathrm{jt}}(\boldsymbol{\tau})
     - \sum_{i=1}^N V_i(\tau_i).
\end{align*}
The QTRAN loss combines three terms:
\begin{align}
  \mathcal{L}_{\mathrm{td}} &= \left(
    \Qtot(\boldsymbol{\tau}, \mathbf{a})
    - y_{\mathrm{td}}
  \right)^2,
    \label{eq:qtran-loss-td} \\
  \mathcal{L}_{\mathrm{opt}} &= \left(
    \hat{Q}_{\mathrm{jt}}(\boldsymbol{\tau}, \bar{\mathbf{a}})
    - \Qtot(\boldsymbol{\tau}, \bar{\mathbf{a}})
  \right)^2,
    \label{eq:qtran-loss-opt} \\
  \mathcal{L}_{\mathrm{nopt}} &= \left(
    \min\!\big(\hat{Q}_{\mathrm{jt}}(\boldsymbol{\tau}, \mathbf{a})
    - \Qtot(\boldsymbol{\tau}, \mathbf{a}),\, 0\big)
  \right)^2,
    \label{eq:qtran-loss-nopt}
\end{align}
where $y_{\mathrm{td}}$ is the TD target.  The loss
$\mathcal{L}_{\mathrm{opt}}$ enforces the equality
condition~\eqref{eq:qtran-opt}, while $\mathcal{L}_{\mathrm{nopt}}$
enforces the inequality~\eqref{eq:qtran-subopt} via a hinge-like
penalty.

\paragraph{Representational hierarchy.}
We can now summarize the relationship among the three factorization
methods.

\begin{proposition}[Representational Ordering]
\label{prop:hierarchy}
In terms of the classes of joint value functions that satisfy IGM
under each method:
\[
  \mathcal{F}_{\mathrm{VDN}} \subsetneq \mathcal{F}_{\mathrm{QMIX}}
  \subsetneq \mathcal{F}_{\mathrm{QTRAN}}.
\]
\end{proposition}

\begin{proof}
$\mathcal{F}_{\mathrm{VDN}} \subsetneq \mathcal{F}_{\mathrm{QMIX}}$
was established in Theorem~\ref{thm:qmix-superset}.

For $\mathcal{F}_{\mathrm{QMIX}} \subseteq \mathcal{F}_{\mathrm{QTRAN}}$:
any monotonic factorization satisfies IGM (Proposition~\ref{prop:mono-igm}).
For any such factorization, set $V_{\mathrm{jt}}(\boldsymbol{\tau}) =
\Qtot(\boldsymbol{\tau}, \bar{\mathbf{a}}) - \sum_i Q_i(\tau_i,
\bar{a}_i) + \sum_i V_i(\tau_i)$ and choose $V_i(\tau_i) =
\max_{\actioni} Q_i(\tau_i, \actioni)$.  Then
condition~\eqref{eq:qtran-opt} holds by construction, and
condition~\eqref{eq:qtran-subopt} holds because $\bar{\mathbf{a}}$
maximizes $\Qtot$ under monotonicity.

For strict inclusion: Example~\ref{ex:vdn-fails} provides a
game where VDN fails.  We show QTRAN can represent it but QMIX
cannot.  The reward matrix has $R(L,L) = 8 > R(R,L) = 0$ but
$R(L,R) = 0 < R(R,R) = 5$.  Set $Q_1(L) = 1, Q_1(R) = 0$
(so $\bar{a}_1 = L$) and $Q_2(L) = 1, Q_2(R) = 0$ (so $\bar{a}_2
= L$).  The individual maxima select $(L, L)$, which is indeed
optimal ($R(L,L) = 8$).  Set $V_1(\tau_1) = 1, V_2(\tau_2) = 1$,
and $V_{\mathrm{jt}} = 8 - 1 - 1 + 1 + 1 = 8$.  Then
$\hat{Q}_{\mathrm{jt}}(L,L) = 1 + 1 + 8 - 1 - 1 = 8 = R(L,L)$
(equality holds).  For $(L, R)$: $\hat{Q}_{\mathrm{jt}}(L,R) = 1
+ 0 + 8 - 1 - 1 = 7 \geq R(L,R) = 0$ (inequality holds).
Similarly for other off-diagonal entries.  However, QMIX requires
monotonicity, which fails here since $R(L,L) > R(R,L)$ implies
$Q_1(L) > Q_1(R)$ yet $R(L,R) < R(R,R)$ would require
$Q_1(L) < Q_1(R)$ under any monotonic decomposition, a
contradiction.
\end{proof}

%% ============================================================
%% SECTION 8: MAPPO
%% ============================================================
\section{Multi-Agent Proximal Policy Optimization}
\label{sec:mappo}

Value decomposition methods (VDN, QMIX, QTRAN) address cooperative
MARL through value-based learning.  An alternative CTDE approach
uses \emph{policy gradient} methods with a centralized critic.
Multi-Agent PPO (MAPPO) \citep{yu2022} adapts the single-agent
PPO algorithm to the multi-agent setting and has demonstrated
surprisingly strong performance.

\paragraph{Setup.}
Each agent $i$ maintains a decentralized stochastic policy
$\policyi(\actioni \mid \tau_i;\, \theta_i)$ parameterized by
$\theta_i$.  A centralized value function $V(s;\, \psi)$
parameterized by $\psi$ estimates the expected return from
global state $s$ under the current joint policy.

\begin{definition}[MAPPO Objective]
\label{def:mappo}
For each agent $i$, the MAPPO objective is the clipped surrogate:
\begin{equation}
\label{eq:mappo-clip}
  L_i^{\mathrm{CLIP}}(\theta_i)
  = \E_{\boldsymbol{\tau}, \mathbf{a}}\!\left[
    \min\!\left(
      r_i(\theta_i)\, \hat{A}_i,\;
      \mathrm{clip}\!\big(r_i(\theta_i),\, 1 - \epsilon,\,
        1 + \epsilon\big)\, \hat{A}_i
    \right)
  \right],
\end{equation}
where:
\begin{itemize}[nosep]
  \item $r_i(\theta_i) = \policyi(\actioni \mid \tau_i;\, \theta_i)
    \,/\, \policyi(\actioni \mid \tau_i;\, \theta_i^{\mathrm{old}})$
    is the importance sampling ratio for agent $i$.
  \item $\hat{A}_i = \hat{A}(s_t, \mathbf{a}_t)$ is the
    advantage estimate computed from the centralized value function:
    $\hat{A}_i = \sum_{\ell=0}^{T-t-1} (\discount \lambda)^\ell
    \delta_{t+\ell}$ with TD errors $\delta_t = r_t + \discount\,
    V(s_{t+1};\, \psi) - V(s_t;\, \psi)$ (GAE-$\lambda$).
  \item $\epsilon > 0$ is the clipping parameter.
\end{itemize}
\end{definition}

\paragraph{Shared vs.\ independent parameters.}
In many cooperative tasks, agents are \emph{homogeneous} (same
observation and action spaces).  MAPPO can exploit this with
\emph{parameter sharing}: all agents use the same policy network
$\policy(\actioni \mid \tau_i, i;\, \theta)$, where agent identity
$i$ is provided as an additional input.  This dramatically reduces
the number of parameters from $N \cdot |\theta|$ to $|\theta|$ and
enables faster learning, though it restricts the class of
representable joint policies to those with symmetric structure
(up to the identity embedding).

\begin{theorem}[Monotonic Improvement under MAPPO]
\label{thm:mappo-improvement}
Let $\policyj = (\policy_1, \ldots, \policy_N)$ be the current
joint policy and $\policyj' = (\policy_1', \ldots, \policy_N')$
be the updated joint policy where each $\policy_i'$ maximizes
$L_i^{\mathrm{CLIP}}(\theta_i)$.  Under the following conditions:
\begin{enumerate}[nosep]
  \item The advantage estimates $\hat{A}_i$ are computed using
    the centralized value function $V(s;\, \psi)$ with GAE-$\lambda$.
  \item The clipping parameter $\epsilon$ is sufficiently small.
  \item Policy updates are performed sequentially (one agent at
    a time) or the total KL divergence $\sum_i
    \KL(\policyi^{\mathrm{old}} \| \policyi')$ is bounded.
\end{enumerate}
Then the joint policy performance satisfies
\begin{equation}
\label{eq:mappo-bound}
  V^{\policyj'}(s) \geq V^{\policyj}(s) - C \sum_{i=1}^N
  \max_{\tau_i} \KL\!\big(\policyi(\cdot \mid \tau_i;\,
  \theta_i^{\mathrm{old}}) \,\big\|\, \policyi(\cdot \mid
  \tau_i;\, \theta_i)\big),
\end{equation}
where $C = \frac{4 \discount \max_{s,\mathbf{a}} |A^{\policyj}
(s, \mathbf{a})|}{(1 - \discount)^2}$ and $A^{\policyj}$ is
the joint advantage function.
\end{theorem}

\begin{proof}
The proof extends the single-agent PPO monotonic improvement
bound \citep[Theorem~1]{yu2022} to the multi-agent setting.

Define the joint advantage $A^{\policyj}(s, \mathbf{a}) =
Q^{\policyj}(s, \mathbf{a}) - V^{\policyj}(s)$.  By the
performance difference lemma:
\[
  V^{\policyj'}(s) - V^{\policyj}(s)
  = \frac{1}{1 - \discount} \E_{s' \sim d_{\policyj'}^s}\!\left[
    \E_{\mathbf{a} \sim \policyj'(\cdot \mid s')}\!\big[
      A^{\policyj}(s', \mathbf{a})
    \big]
  \right],
\]
where $d_{\policyj'}^s$ is the discounted state visitation
distribution under $\policyj'$ starting from $s$.  The key step
is bounding the difference in expectations under $\policyj'$ vs.\
$\policyj$.  Since the joint policy factorizes as
$\policyj'(\mathbf{a} \mid s) = \prod_i \policyi'(\actioni \mid
\tau_i)$, the total variation distance between $\policyj'$ and
$\policyj$ satisfies:
\[
  \|\policyj'(\cdot \mid s) - \policyj(\cdot \mid s)\|_{\mathrm{TV}}
  \leq \sum_{i=1}^N \|\policyi'(\cdot \mid \tau_i) -
  \policyi(\cdot \mid \tau_i)\|_{\mathrm{TV}}.
\]
Applying Pinsker's inequality ($\|p - q\|_{\mathrm{TV}} \leq
\sqrt{\frac{1}{2} \KL(p \| q)}$) to each term and combining
with the performance difference lemma yields the bound~\eqref{eq:mappo-bound}.
\end{proof}

\begin{rlconnection}[title={MAPPO on SMAC}]
\citet{yu2022} showed that MAPPO, despite its simplicity,
matches or exceeds the performance of value decomposition methods
(QMIX, QTRAN) on the StarCraft Multi-Agent Challenge.  Key
practical insights include:
\begin{itemize}[nosep]
  \item \textbf{Centralized value function:} using the global
    state $s$ as input to $V$ (rather than local observations)
    significantly reduces variance.
  \item \textbf{Parameter sharing:} homogeneous agents benefit
    from shared policy networks with agent-ID conditioning.
  \item \textbf{Large batch sizes:} PPO's on-policy nature
    requires large batches for stable multi-agent learning;
    typical batch sizes are $10$--$100\times$ larger than
    single-agent settings.
\end{itemize}
\end{rlconnection}

%% ============================================================
%% SECTION 9: CREDIT ASSIGNMENT
%% ============================================================
\section{Credit Assignment}
\label{sec:credit}

A fundamental challenge in cooperative MARL is \emph{credit
assignment}: given a shared team reward, how should the
contribution of each agent be assessed?

\begin{definition}[Credit Assignment Problem]
\label{def:credit}
The \emph{credit assignment problem} in cooperative MARL is the
problem of decomposing a team reward signal $R(s, \mathbf{a})$
into agent-specific feedback signals $r_i$ such that optimizing
each $r_i$ individually leads to optimization of the team objective.
\end{definition}

We survey three principled approaches.

\subsection{Difference Rewards}

\begin{definition}[Difference Reward]
\label{def:diff-reward}
The \emph{difference reward} for agent $i$ is
\begin{equation}
\label{eq:diff-reward}
  D_i(s, \mathbf{a}) = R(s, \mathbf{a})
  - R(s, (\actionmi, c_i)),
\end{equation}
where $c_i \in \Actions_i$ is a fixed \emph{default action}
for agent $i$, and $(\actionmi, c_i)$ denotes the joint action
with agent $i$'s action replaced by $c_i$.
\end{definition}

\begin{proposition}[Alignment of Difference Rewards]
\label{prop:diff-alignment}
If each agent $i$ maximizes the expected sum of discounted
difference rewards $\E[\sum_t \discount^t D_i(s_t, \mathbf{a}_t)]$,
then the resulting joint policy is a Nash equilibrium of the
common-payoff game with reward $R$.
\end{proposition}

\begin{proof}
Fix agent $i$ and suppose all other agents $j \neq i$ follow
fixed policies $\policymi$.  Agent $i$'s expected return under
difference rewards is:
\[
  J_i^D(\policyi) = \E^{\policyi, \policymi}\!\left[
    \sum_t \discount^t D_i(s_t, \mathbf{a}_t)
  \right]
  = \E^{\policyi, \policymi}\!\left[
    \sum_t \discount^t R(s_t, \mathbf{a}_t)
  \right]
  - \E^{\policymi}\!\left[
    \sum_t \discount^t R(s_t, (\actionmi^t, c_i))
  \right].
\]
The second term does not depend on $\policyi$ (since $c_i$ is
fixed and the other agents' policies are fixed).  Therefore,
$\argmax_{\policyi} J_i^D(\policyi) = \argmax_{\policyi}
\E^{\policyi, \policymi}[\sum_t \discount^t R(s_t, \mathbf{a}_t)]$.
If this holds for all $i$ simultaneously, each agent's policy
is a best response to the others under the team reward $R$,
which is the definition of Nash equilibrium.  In a common-payoff
game, Nash equilibria coincide with team-optimal policies when
the equilibrium is Pareto-efficient.
\end{proof}

\subsection{Shapley Value-Based Credit Assignment}

The Shapley value, from cooperative game theory (Phase~01,
Lesson~20), provides a principled way to attribute team performance
to individual agents.

\begin{definition}[Shapley Value for MARL]
\label{def:shapley}
Given a cooperative game with characteristic function
$v : 2^{\Players} \to \R$ where $v(S)$ represents the value
achievable by coalition $S \subseteq \Players$, the
\emph{Shapley value} of agent $i$ is
\begin{equation}
\label{eq:shapley}
  \phi_i(v) = \sum_{S \subseteq \Players \setminus \{i\}}
  \frac{|S|!\,(N - |S| - 1)!}{N!}
  \big(v(S \cup \{i\}) - v(S)\big).
\end{equation}
In the MARL context, we define $v(S) = \E[R(s, \mathbf{a}_S,
\mathbf{c}_{-S})]$, where agents in $S$ follow their learned
policies and agents outside $S$ take default actions.
\end{definition}

\begin{remark}
The Shapley value satisfies four desirable axioms: efficiency
($\sum_i \phi_i = v(\Players) - v(\emptyset)$), symmetry, null
player, and additivity.  However, computing it exactly requires
evaluating $2^N$ coalitions, making it intractable for large $N$.
Approximations via sampling random permutations of agents are
commonly used.
\end{remark}

\subsection{Counterfactual Baselines: COMA}

\begin{definition}[Counterfactual Advantage]
\label{def:coma}
The \emph{Counterfactual Multi-Agent} (COMA) advantage for agent
$i$ is
\begin{equation}
\label{eq:coma}
  A_i^{\mathrm{CF}}(s, \mathbf{a})
  = Q(s, \mathbf{a}) - \sum_{a_i' \in \Actions_i}
    \policyi(a_i' \mid \tau_i)\, Q(s, (\actionmi, a_i')),
\end{equation}
where $Q(s, \mathbf{a})$ is the centralized joint action-value
function and the baseline is the expected Q-value over agent $i$'s
policy, marginalizing out $\actioni$ while holding $\actionmi$ fixed.
\end{definition}

\begin{proposition}[Unbiasedness of COMA Baseline]
\label{prop:coma-unbiased}
The COMA advantage satisfies
$\E_{\actioni \sim \policyi}[A_i^{\mathrm{CF}}(s, \mathbf{a})]
= 0$, so it is a valid baseline for variance reduction in
policy gradient methods.
\end{proposition}

\begin{proof}
\begin{align*}
  \E_{\actioni \sim \policyi}\!\big[A_i^{\mathrm{CF}}(s, \mathbf{a})\big]
  &= \E_{\actioni \sim \policyi}\!\big[Q(s, \mathbf{a})\big]
     - \E_{\actioni \sim \policyi}\!\bigg[
       \sum_{a_i'} \policyi(a_i' \mid \tau_i)\, Q(s, (\actionmi, a_i'))
     \bigg] \\
  &= \sum_{\actioni} \policyi(\actioni \mid \tau_i)\,
     Q(s, (\actionmi, \actioni))
     - \sum_{a_i'} \policyi(a_i' \mid \tau_i)\,
     Q(s, (\actionmi, a_i')) \\
  &= 0.
\end{align*}
The baseline is independent of the sampled $\actioni$ (it
averages over all possible actions), so subtracting it does not
introduce bias into the policy gradient.
\end{proof}

\begin{rlconnection}[title={Credit Assignment in Practice}]
\begin{itemize}[nosep]
  \item \textbf{VDN/QMIX:} implicitly perform credit assignment
    through the learned individual utilities $Q_i$---each
    agent's $Q_i$ reflects its ``share'' of the team value.
  \item \textbf{COMA:} explicitly computes counterfactual
    baselines, requiring a centralized critic that can evaluate
    $Q(s, (\actionmi, a_i'))$ for all $a_i'$.
  \item \textbf{Shapley values:} provide the most principled
    attribution but at exponential computational cost; practical
    only for small $N$ or with sampling approximations.
\end{itemize}
\end{rlconnection}

%% ============================================================
%% SECTION 10: EXERCISES
%% ============================================================
\section{Exercises}
\label{sec:exercises}

\begin{exercise}[VDN Decomposition and IGM]
\textnormal{(\(*\))} \\
Let $N = 3$ agents with action spaces $\Actions_i = \{L, R\}$ and
a VDN decomposition $\Qtot(\mathbf{a}) = Q_1(a_1) + Q_2(a_2) +
Q_3(a_3)$ with $Q_1(L) = 3, Q_1(R) = 1$, $Q_2(L) = 2, Q_2(R) = 4$,
$Q_3(L) = 5, Q_3(R) = 2$.  Verify that the IGM condition holds by
computing $\argmax_{\mathbf{a}} \Qtot(\mathbf{a})$ both jointly
(by enumeration) and via individual maximization.
\end{exercise}

\begin{exercise}[Joint Value in a Dec-POMDP]
\textnormal{(\(*\))} \\
Consider a $2$-agent, $2$-state Dec-POMDP with $\States = \{s_0,
s_1\}$, $\Actions_i = \{L, R\}$, discount $\discount = 0.9$,
and deterministic transitions: from $s_0$, joint action $(L, L)$
leads to $s_1$ with reward $5$; all other joint actions keep the
state at $s_0$ with reward $1$.  From $s_1$, all actions lead to
$s_1$ with reward $3$.  Assuming full observability (so each
agent sees the state), compute $V^{\policyj}(s_0)$ for the
joint policy $\policyj = (\text{always } L, \text{always } L)$.
\end{exercise}

\begin{exercise}[Monotonicity Implies IGM -- Formal Proof]
\textnormal{(\(**\))} \\
Let $f : \R^N \to \R$ be a continuously differentiable function
satisfying $\frac{\partial f}{\partial x_i} \geq 0$ for all $i$
and all $\mathbf{x} \in \R^N$.  Let $g_i : \Actions_i \to \R$
for $i = 1, \ldots, N$, and define
$\Qtot(\mathbf{a}) = f(g_1(a_1), \ldots, g_N(a_N))$.  Prove that
\[
  \argmax_{\mathbf{a}} \Qtot(\mathbf{a})
  \supseteq \left\{
    \big(\argmax_{a_1} g_1(a_1), \ldots,
    \argmax_{a_N} g_N(a_N)\big)
  \right\}.
\]
Under what additional condition on $f$ is equality guaranteed?

\textit{Hint:} Use the coordinate-wise replacement argument from
Proposition~\ref{prop:mono-igm}.  For equality, consider strict
monotonicity.
\end{exercise}

\begin{exercise}[QMIX Fails, QTRAN Succeeds]
\textnormal{(\(**\))} \\
Consider the $2$-agent coordination game with reward matrix:
\[
\begin{array}{c|ccc}
  R(a_1, a_2) & a_2 = A & a_2 = B & a_2 = C \\
  \hline
  a_1 = A & 8 & 0 & 6 \\
  a_1 = B & 0 & 8 & 0 \\
  a_1 = C & 6 & 0 & 6 \\
\end{array}
\]
\begin{enumerate}[label=(\alph*)]
  \item Show that no monotonic decomposition $\Qtot(a_1, a_2) =
    f(Q_1(a_1), Q_2(a_2))$ with $f$ non-decreasing in both
    arguments can represent this game (so QMIX fails).
  \item Construct individual utilities $Q_1, Q_2$, joint value
    $V_{\mathrm{jt}}$, and individual values $V_1, V_2$ such
    that the QTRAN conditions~\eqref{eq:qtran-opt}--\eqref{eq:qtran-subopt}
    hold with $\bar{\mathbf{a}} = (A, A)$.
\end{enumerate}

\textit{Hint:} For part~(a), consider the implications of
$R(A, B) = 0 < R(B, B) = 8$ and $R(A, A) = 8 > R(B, A) = 0$
for monotonicity in $Q_1$.
\end{exercise}

\begin{exercise}[NEXP-Hardness and Exponential Policies]
\textnormal{(\(***\))} \\
In the proof sketch of Theorem~\ref{thm:nexp}, we noted that the
reduction from the tiling problem encodes a $2^n \times 2^n$ grid.
\begin{enumerate}[label=(\alph*)]
  \item Explain why a Dec-POMDP with horizon $T = 2^{2n}$ requires
    policies that are, in general, exponentially large decision
    trees.
  \item Show that if agents could communicate one bit per time step,
    the complexity would drop from NEXP to NEXP $\cap$ co-NEXP
    (informally: communication reduces the ``decentralization
    penalty'').  Discuss what this implies for the value of
    communication protocols in cooperative MARL.
\end{enumerate}
\end{exercise}

\begin{exercise}[Credit Assignment Bias Analysis]
\textnormal{(\(***\))} \\
Consider $N$ agents with a team reward $R(s, \mathbf{a}) =
\sum_{i=1}^N f_i(s, \actioni) + g(s, \mathbf{a})$, where $f_i$
captures agent $i$'s individual contribution and $g$ captures
interaction effects.  Define the difference reward with default
action $c_i$ for each agent.
\begin{enumerate}[label=(\alph*)]
  \item Derive an expression for $D_i(s, \mathbf{a})$ in terms
    of $f_i$ and $g$.
  \item Show that if $g \equiv 0$ (no interactions), then $D_i$
    equals $f_i(\actioni) - f_i(c_i)$, so maximizing $D_i$ is
    equivalent to maximizing $f_i$.
  \item When $g \not\equiv 0$, analyze the bias introduced by
    difference rewards.  Specifically, show that the gradient
    $\nabla_{\theta_i} \E[D_i]$ may not align with
    $\nabla_{\theta_i} \E[R]$ and characterize the conditions
    under which they do align.
\end{enumerate}
\end{exercise}

%% ============================================================
%% SECTION 11: SUMMARY
%% ============================================================
\section{Summary}
\label{sec:summary}

\begin{itemize}
  \item The \textbf{Dec-POMDP} formalizes cooperative MARL as a
    tuple $(\Players, \States, \{\Actions_i\}, \Trans, R,
    \{\Omega_i\}, O, \discount)$ where all agents share a common
    reward $R$ but observe only local information.

  \item Finding an optimal joint policy for a Dec-POMDP is
    \textbf{NEXP-complete} (Theorem~\ref{thm:nexp}), far harder
    than single-agent MDPs (P) or POMDPs (PSPACE-complete).  The
    intractability stems from the combination of decentralization
    and partial observability.

  \item The \textbf{CTDE paradigm} (Centralized Training with
    Decentralized Execution) sidesteps the complexity barrier by
    using global information during training while producing
    decentralized policies for execution.

  \item \textbf{VDN} decomposes $\Qtot = \sum_i Q_i$, guaranteeing
    that greedy action selection decomposes, but cannot represent
    non-additive agent interactions (Theorem~\ref{thm:vdn-greedy}).

  \item The \textbf{IGM condition} (Definition~\ref{def:igm})
    is the fundamental structural requirement for decentralized
    greedy action selection.  The monotonicity condition
    $\partial \Qtot / \partial Q_i \geq 0$ is a sufficient
    condition for IGM (Proposition~\ref{prop:mono-igm}).

  \item \textbf{QMIX} enforces monotonicity via non-negative
    mixing network weights generated by hypernetworks, strictly
    generalizing VDN (Theorem~\ref{thm:qmix-superset}).

  \item \textbf{QTRAN} achieves IGM without monotonicity by
    introducing an additive correction term and enforcing
    equality/inequality conditions
    (Proposition~\ref{prop:qtran-igm}), yielding the
    representational hierarchy VDN $\subsetneq$ QMIX $\subsetneq$
    QTRAN.

  \item \textbf{MAPPO} adapts PPO to multi-agent settings with
    a centralized value function, preserving approximate monotonic
    improvement guarantees (Theorem~\ref{thm:mappo-improvement}).

  \item \textbf{Credit assignment}---attributing team reward to
    individual agents---can be addressed via difference rewards,
    Shapley values, or counterfactual baselines (COMA), each
    offering different bias-variance-computation tradeoffs.
\end{itemize}

%% ============================================================
%% SECTION 12: FURTHER READING
%% ============================================================
\section{Further Reading}
\label{sec:further-reading}

\begin{itemize}
  \item \textbf{Competitive and mixed MARL (Lesson~63)} extends
    the multi-agent framework to zero-sum and general-sum games,
    covering Minimax-Q, Nash-Q, MADDPG, and self-play methods
    including PSRO.

  \item \textbf{Mean-field games (Lesson~64)} addresses the
    scalability challenge when $N \to \infty$ by replacing
    agent-level interactions with interactions against a
    population distribution.

  \item \textbf{Communication in MARL (Lesson~65)} studies
    learned communication protocols (CommNet, DIAL, TarMAC)
    that allow agents to share information during execution,
    relaxing the strict decentralization assumption of
    Dec-POMDPs.

  \item \textbf{QPLEX} \citep{wang2021} introduces duplex
    dueling architectures that decompose $\Qtot$ into
    advantage and value components, achieving IGM with
    greater representational capacity than QMIX while
    maintaining tractable training.

  \item \textbf{MAVEN} \citep{mahajan2019} addresses the
    exploration challenge in value decomposition methods by
    introducing a shared latent variable that conditions
    agent exploration, promoting diverse joint behaviors.

  \item \textbf{SMAC benchmark} \citep{samvelyan2019} provides
    a standardized suite of StarCraft~II micromanagement
    scenarios for evaluating cooperative MARL algorithms,
    with varying difficulty levels and team compositions.

  \item \textbf{Comprehensive surveys.}  \citet{zhang2021}
    provides a selective overview of MARL theory and algorithms.
    \citet[Chapters~6--8]{oliehoek2016} covers approximate
    methods for Dec-POMDPs in depth.  \citet{kraemer2016}
    surveys multi-agent reinforcement learning as a framework
    for decentralized planning.
\end{itemize}

\bibliography{cooperative-refs}

\end{document}
