\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 14: Stochastic Approximation Noise Decomposition}}
\fancyhead[R]{\thepage}
\title{Lesson 14: Stochastic Approximation Noise Decomposition}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% ============================================================
%% SECTION 1: MOTIVATION AND OVERVIEW
%% ============================================================

\section{Motivation and Overview}
\label{sec:motivation}

This is the capstone lesson of the Measure Theory track (Lessons
1--14).  Over the preceding thirteen lessons we assembled a complete
probability toolkit: sigma-algebras and measures (Lessons~1--2),
measurable functions and expectation (Lessons~3--5), conditional
expectation and variance (Lessons~6--7), filtrations and
martingale difference sequences (Lesson~8), concentration
inequalities (Lesson~9), modes of convergence and
Borel--Cantelli (Lesson~10), limit theorems (Lesson~11), Markov
chains (Lesson~12), and martingale convergence theory including
the Robbins--Siegmund theorem (Lesson~13).

The purpose of this lesson is twofold.  First, we prove the
\emph{TD noise decomposition theorem}, which shows that the noise
in a temporal-difference update is a martingale difference sequence
with bounded conditional variance.  This single result is the
bridge between measure theory and the stochastic approximation
convergence theorems developed in the Optimization track
(Lessons~31--32).  Second, we provide a comprehensive summary of
every key result from Lessons~1--14, organized by its role in RL
theory.

\begin{rlconnection}[title={Why This Matters for RL}]
\begin{itemize}[nosep]
  \item \textbf{TD(0), SARSA, Q-learning convergence:}
    Every convergence proof for a tabular RL algorithm
    reduces the update to Robbins--Monro form:
    $\theta_{n+1} = \theta_n + \alpha_n [h(\theta_n) + w_n]$.
    The TD noise decomposition proves that $\{w_n\}$ is an MDS
    with the right variance bound, which is the main
    probabilistic ingredient.
  \item \textbf{Two-timescale actor-critic:}
    Actor-critic algorithms run two coupled SA recursions.
    Each recursion needs its own MDS noise decomposition; the
    structure of the proof is identical to the one given here.
  \item \textbf{From theory to practice:}
    Understanding \emph{why} the noise is an MDS (and not
    just ``mean zero'') clarifies which algorithm modifications
    preserve convergence guarantees and which break them.
\end{itemize}
\end{rlconnection}

\medskip
\noindent\textbf{Prerequisites.}
All of Lessons~1--13, especially: conditional expectation and the
tower property (Lesson~6), martingale difference sequences
(Lesson~8), and the Robbins--Siegmund theorem (Lesson~13).

\noindent\textbf{Learning objectives.}
\begin{itemize}[nosep]
  \item State and prove the TD noise decomposition theorem.
  \item Verify the MDS and bounded-variance conditions for
    TD(0), Q-learning, and expected SARSA.
  \item Connect the noise decomposition to the Robbins--Monro
    framework and state the resulting convergence theorem.
  \item Summarize the role of every key measure-theoretic
    result in RL convergence proofs.
\end{itemize}

%% ============================================================
%% SECTION 2: REFERENCES
%% ============================================================

\section{References}
\label{sec:references}

\begin{itemize}[nosep]
  \item \citet{borkar2008}, Chapter~2: Stochastic approximation
    framework and noise conditions.
  \item \citet{sutton2018}, Chapters~6 and 12: TD learning
    algorithms and eligibility traces.
  \item \citet{szepesvari2010}, Chapter~4: Convergence of TD
    and Q-learning with detailed noise analysis.
  \item \citet{bertsekas2012}, Chapter~6: Approximate dynamic
    programming and stochastic approximation convergence.
  \item \citet{williams1991}, Chapters~10--12: Martingales and
    conditional expectation---the probabilistic backbone.
\end{itemize}

%% ============================================================
%% SECTION 3: THE MDP FRAMEWORK
%% ============================================================

\section{The MDP Framework and Bellman Operators}
\label{sec:mdp}

We fix the notation used throughout this lesson.  A \emph{finite
Markov Decision Process} (MDP) is a tuple
$(\mathcal{S}, \mathcal{A}, P, r, \gamma)$ where
$\mathcal{S}$ is a finite state space,
$\mathcal{A}$ is a finite action space,
$P(s'|s,a) = \Prob(S_{t+1}=s' \mid S_t=s, A_t=a)$ is the
transition kernel,
$r: \mathcal{S} \times \mathcal{A} \times \mathcal{S} \to \R$
is a bounded reward function with $|r(s,a,s')| \leq R_{\max}$,
and $\gamma \in [0,1)$ is the discount factor.

\begin{definition}[Stationary Policy]
\label{def:policy}
A stationary (randomized) policy is a mapping
$\pi: \mathcal{S} \to \Delta(\mathcal{A})$, where
$\pi(a|s)$ denotes the probability of selecting action $a$ in
state $s$.
\end{definition}

\begin{definition}[Value Function]
\label{def:value}
The state-value function under policy $\pi$ is
\begin{equation}
  V^\pi(s)
  = \E_\pi\!\left[\sum_{t=0}^{\infty} \gamma^t R_{t+1}
    \;\middle|\; S_0 = s\right],
  \qquad s \in \mathcal{S}.
  \label{eq:value}
\end{equation}
The action-value function is
$Q^\pi(s,a)
= \E_\pi[\sum_{t=0}^{\infty} \gamma^t R_{t+1}
  \mid S_0=s, A_0=a]$.
\end{definition}

\begin{definition}[Bellman Operator]
\label{def:bellman}
The Bellman operator $\mathcal{T}^\pi: \R^{|\mathcal{S}|}
\to \R^{|\mathcal{S}|}$ is defined by
\begin{equation}
  (\mathcal{T}^\pi V)(s)
  = \sum_{a \in \mathcal{A}} \pi(a|s)
    \sum_{s' \in \mathcal{S}} P(s'|s,a)
    \bigl[r(s,a,s') + \gamma V(s')\bigr].
  \label{eq:bellman_op}
\end{equation}
The value function $V^\pi$ is the unique fixed point of
$\mathcal{T}^\pi$.
\end{definition}

\begin{remark}
\label{rem:contraction}
The operator $\mathcal{T}^\pi$ is a $\gamma$-contraction in
the supremum norm: $\|\mathcal{T}^\pi V - \mathcal{T}^\pi W
\|_\infty \leq \gamma \|V - W\|_\infty$.  This is proved
using the linearity in $V$ and the fact that
$\sum_{a}\pi(a|s)\sum_{s'}P(s'|s,a) = 1$.  The Banach
fixed-point theorem (Lesson~22) then guarantees existence and
uniqueness of $V^\pi$.
\end{remark}

%% ============================================================
%% SECTION 4: THE TD UPDATE AND ITS NOISE
%% ============================================================

\section{The TD(0) Update and Its Noise Structure}
\label{sec:td_noise}

We now derive the noise structure of the simplest
temporal-difference algorithm: tabular TD(0) for policy evaluation.

\subsection{The TD(0) Update Rule}

An agent follows policy $\pi$ and observes a trajectory
$S_0, A_0, R_1, S_1, A_1, R_2, \ldots$  The TD(0) update at
time $t$ modifies the value estimate at the visited state:
\begin{equation}
  V_{t+1}(S_t)
  = V_t(S_t) + \alpha_t(S_t)
    \bigl[R_{t+1} + \gamma V_t(S_{t+1}) - V_t(S_t)\bigr],
  \label{eq:td0}
\end{equation}
where $\alpha_t(s) \geq 0$ is the step size applied to state~$s$
at the $t$-th visit, and $V_{t+1}(s) = V_t(s)$ for all
$s \neq S_t$.

\begin{definition}[TD Error]
\label{def:td_error}
The temporal-difference error at time $t$ is
\begin{equation}
  \delta_t
  = R_{t+1} + \gamma V_t(S_{t+1}) - V_t(S_t).
  \label{eq:td_error}
\end{equation}
\end{definition}

\subsection{Rewriting TD(0) in Robbins--Monro Form}

To apply stochastic approximation theory, we rewrite
\eqref{eq:td0} by adding and subtracting
$(\mathcal{T}^\pi V_t)(S_t)$:
\begin{align}
  V_{t+1}(S_t)
  &= V_t(S_t) + \alpha_t(S_t)
    \bigl[(\mathcal{T}^\pi V_t)(S_t) - V_t(S_t) + w_t(S_t)
    \bigr],
  \label{eq:td_rm}
\end{align}
where the \emph{noise term} is
\begin{equation}
  w_t(s)
  = R_{t+1} + \gamma V_t(S_{t+1})
    - (\mathcal{T}^\pi V_t)(s).
  \label{eq:noise_def}
\end{equation}
The term $h(V_t)(s) = (\mathcal{T}^\pi V_t)(s) - V_t(s)$ is
the \emph{mean field}: it equals zero exactly when
$V_t = V^\pi$.  The noise $w_t(s)$ captures the randomness
from the single-sample transition $(S_t, A_t, R_{t+1}, S_{t+1})$.

\subsection{The Filtration}

\begin{definition}[Natural Filtration of the MDP Trajectory]
\label{def:filtration}
Define the filtration
$\{\mathcal{F}_t\}_{t \geq 0}$ by
\begin{equation}
  \mathcal{F}_t
  = \sigma\bigl(S_0, A_0, R_1, S_1, \ldots,
    S_{t-1}, A_{t-1}, R_t, S_t\bigr).
  \label{eq:filtration}
\end{equation}
Crucially, $V_t$ and $S_t$ are $\mathcal{F}_t$-measurable,
while $A_t$, $R_{t+1}$, and $S_{t+1}$ are
$\mathcal{F}_{t+1}$-measurable but \emph{not}
$\mathcal{F}_t$-measurable.
\end{definition}

%% ============================================================
%% SECTION 5: THE TD NOISE DECOMPOSITION THEOREM
%% ============================================================

\section{The TD Noise Decomposition Theorem}
\label{sec:decomposition}

The following theorem is the main result of this lesson.  It
establishes the two conditions---zero conditional mean and
bounded conditional variance---that connect the TD noise to
the Robbins--Monro framework.

\begin{theorem}[TD Noise Decomposition]
\label{thm:td_noise}
Let $(\mathcal{S}, \mathcal{A}, P, r, \gamma)$ be a finite
MDP with $|r(s,a,s')| \leq R_{\max}$.  Let $\pi$ be a
stationary policy, and let $\{V_t\}$ be the sequence of value
estimates produced by TD(0) with the noise $w_t(s)$ defined
in \eqref{eq:noise_def}.  Then for all $t \geq 0$ and all
$s \in \mathcal{S}$:
\begin{enumerate}
  \item \textbf{(Martingale difference property)}
    $\E[w_t(s) \mid \mathcal{F}_t] = 0$.
  \item \textbf{(Bounded conditional variance)}
    There exists a constant $C > 0$ (depending only on
    $R_{\max}$, $\gamma$, and $|\mathcal{S}|$) such that
    \begin{equation}
      \E\bigl[w_t(s)^2 \mid \mathcal{F}_t\bigr]
      \leq C\bigl(1 + \|V_t\|_\infty^2\bigr).
      \label{eq:variance_bound}
    \end{equation}
\end{enumerate}
\end{theorem}

\begin{proof}
\textbf{Part 1 (MDS property).}
Conditioned on $\mathcal{F}_t$, the quantities $V_t$ and
$S_t = s$ are known constants.  The randomness comes from the
action $A_t \sim \pi(\cdot|s)$ and the next state
$S_{t+1} \sim P(\cdot|s, A_t)$.  Compute:
\begin{align}
  \E[w_t(s) \mid \mathcal{F}_t]
  &= \E\bigl[R_{t+1} + \gamma V_t(S_{t+1})
    \mid \mathcal{F}_t\bigr]
    - (\mathcal{T}^\pi V_t)(s).
  \label{eq:mds_step1}
\end{align}
For the first term, we apply the law of total expectation
(tower property, Lesson~6) by conditioning on $A_t$:
\begin{align}
  &\E\bigl[R_{t+1} + \gamma V_t(S_{t+1})
    \mid \mathcal{F}_t\bigr] \nonumber\\
  &\quad= \sum_{a \in \mathcal{A}} \pi(a|s)\,
    \E\bigl[R_{t+1} + \gamma V_t(S_{t+1})
    \mid S_t = s,\, A_t = a\bigr] \nonumber\\
  &\quad= \sum_{a \in \mathcal{A}} \pi(a|s)
    \sum_{s' \in \mathcal{S}} P(s'|s,a)
    \bigl[r(s,a,s') + \gamma V_t(s')\bigr] \nonumber\\
  &\quad= (\mathcal{T}^\pi V_t)(s).
  \label{eq:mds_step2}
\end{align}
In the second equality we used the Markov property:
conditioned on $(S_t, A_t)$, the pair $(R_{t+1}, S_{t+1})$ is
independent of all earlier history, so conditioning on
$\mathcal{F}_t$ reduces to conditioning on $(S_t, A_t)$.
Substituting \eqref{eq:mds_step2} into \eqref{eq:mds_step1}:
\[
  \E[w_t(s) \mid \mathcal{F}_t]
  = (\mathcal{T}^\pi V_t)(s)
    - (\mathcal{T}^\pi V_t)(s)
  = 0.
\]

\textbf{Part 2 (Bounded conditional variance).}
Since $|r(s,a,s')| \leq R_{\max}$ and
$|V_t(s')| \leq \|V_t\|_\infty$ for all $s'$, we bound
each component of $w_t(s)$:
\begin{align}
  |R_{t+1}| &\leq R_{\max}, \nonumber\\
  |\gamma V_t(S_{t+1})| &\leq \gamma \|V_t\|_\infty, \nonumber\\
  |(\mathcal{T}^\pi V_t)(s)|
    &\leq \sum_a \pi(a|s) \sum_{s'} P(s'|s,a)
      \bigl[R_{\max} + \gamma \|V_t\|_\infty\bigr]
    = R_{\max} + \gamma \|V_t\|_\infty.
  \label{eq:component_bounds}
\end{align}
Therefore
\begin{align}
  |w_t(s)|
  &\leq |R_{t+1}| + \gamma|V_t(S_{t+1})|
    + |(\mathcal{T}^\pi V_t)(s)| \nonumber\\
  &\leq 2R_{\max} + 2\gamma \|V_t\|_\infty.
  \label{eq:noise_bound}
\end{align}
Squaring and using $(a+b)^2 \leq 2a^2 + 2b^2$:
\begin{align}
  w_t(s)^2
  &\leq (2R_{\max} + 2\gamma\|V_t\|_\infty)^2 \nonumber\\
  &\leq 2(2R_{\max})^2 + 2(2\gamma\|V_t\|_\infty)^2
  \nonumber\\
  &= 8R_{\max}^2 + 8\gamma^2 \|V_t\|_\infty^2.
  \label{eq:noise_sq_bound}
\end{align}
Setting $C = 8\max(R_{\max}^2, \gamma^2)$, we obtain
\[
  \E\bigl[w_t(s)^2 \mid \mathcal{F}_t\bigr]
  \leq C\bigl(1 + \|V_t\|_\infty^2\bigr). \qedhere
\]
\end{proof}

\begin{keyresult}[title={TD Noise is a Martingale Difference Sequence}]
The noise $w_t(s) = R_{t+1} + \gamma V_t(S_{t+1})
- (\mathcal{T}^\pi V_t)(s)$ satisfies:
\begin{enumerate}[nosep]
  \item $\E[w_t(s) \mid \mathcal{F}_t] = 0$
    \quad (MDS property),
  \item $\E[w_t(s)^2 \mid \mathcal{F}_t]
    \leq C(1 + \|V_t\|_\infty^2)$
    \quad (bounded variance).
\end{enumerate}
These are precisely the noise conditions required by the
Robbins--Monro theorem (Lesson~31) and the ODE method
(Lesson~32) to guarantee almost-sure convergence.
\end{keyresult}

\begin{remark}[Role of the Markov Property]
\label{rem:markov}
The key step in Part~1 is that conditioning on the full
history $\mathcal{F}_t$ reduces to conditioning on
$(S_t, A_t)$.  This is the \emph{Markov property} of the
MDP.  Without it---for example, in a POMDP where the agent
cannot observe the true state---the noise would depend on the
entire history and the MDS property would generally fail.
\end{remark}

%% ============================================================
%% SECTION 6: EXTENSIONS TO OTHER RL ALGORITHMS
%% ============================================================

\section{Extensions to Other Algorithms}
\label{sec:extensions}

The noise decomposition argument of Theorem~\ref{thm:td_noise}
extends to other tabular RL algorithms with minor modifications.

\subsection{Q-Learning}

In Q-learning, the update target replaces the policy average
with the greedy action:
\begin{equation}
  Q_{t+1}(S_t, A_t)
  = Q_t(S_t, A_t) + \alpha_t
    \bigl[R_{t+1} + \gamma \max_{a'} Q_t(S_{t+1}, a')
    - Q_t(S_t, A_t)\bigr].
  \label{eq:qlearning}
\end{equation}
The Bellman optimality operator is
$(\mathcal{T}^* Q)(s,a) = \sum_{s'} P(s'|s,a)
[r(s,a,s') + \gamma \max_{a'} Q(s',a')]$, and the noise is
\begin{equation}
  w_t^Q(s,a)
  = R_{t+1} + \gamma \max_{a'} Q_t(S_{t+1}, a')
    - (\mathcal{T}^* Q_t)(s,a).
  \label{eq:qnoise}
\end{equation}

\begin{proposition}[Q-Learning Noise Decomposition]
\label{prop:q_noise}
Under the same setup as Theorem~\ref{thm:td_noise}, the
Q-learning noise $w_t^Q(s,a)$ satisfies:
\begin{enumerate}[nosep]
  \item $\E[w_t^Q(s,a) \mid \mathcal{F}_t] = 0$,
  \item $\E[w_t^Q(s,a)^2 \mid \mathcal{F}_t]
    \leq C'(1 + \|Q_t\|_\infty^2)$
    for a constant $C'$ depending only on $R_{\max}$ and
    $\gamma$.
\end{enumerate}
\end{proposition}

\begin{proof}
The proof is essentially identical to that of
Theorem~\ref{thm:td_noise}.  For Part~1, conditioned on
$\mathcal{F}_t$ (which includes $Q_t$ and $S_t = s$,
$A_t = a$), the next-state transition
$S_{t+1} \sim P(\cdot|s,a)$ and reward $R_{t+1}$ are
independent of all prior history by the Markov property.  Then
\begin{align}
  \E\bigl[R_{t+1} + \gamma \max_{a'} Q_t(S_{t+1}, a')
    \mid \mathcal{F}_t\bigr]
  &= \sum_{s'} P(s'|s,a)
    \bigl[r(s,a,s')
    + \gamma \max_{a'} Q_t(s', a')\bigr] \nonumber\\
  &= (\mathcal{T}^* Q_t)(s,a),
\end{align}
giving $\E[w_t^Q \mid \mathcal{F}_t] = 0$.  Part~2 follows
from $|R_{t+1}| \leq R_{\max}$,
$|\max_{a'} Q_t(s',a')| \leq \|Q_t\|_\infty$, and the same
triangle-inequality and squaring argument.
\end{proof}

\subsection{Expected SARSA}

Expected SARSA replaces the sample next-action with its
expectation under the behavior policy $b$:
\begin{equation}
  V_{t+1}(S_t)
  = V_t(S_t) + \alpha_t
    \bigl[R_{t+1} + \gamma \textstyle\sum_{a'}
    b(a'|S_{t+1}) Q_t(S_{t+1}, a')
    - Q_t(S_t, A_t)\bigr].
  \label{eq:expected_sarsa}
\end{equation}
The noise decomposition again yields an MDS with bounded
conditional variance.  The key observation is that
$\sum_{a'} b(a'|S_{t+1}) Q_t(S_{t+1}, a')$ is a bounded
function of $S_{t+1}$ (given $Q_t$), so the same argument
applies.

\begin{warning}[title={When the MDS Property Fails}]
The MDS property relies on:
\begin{enumerate}[nosep]
  \item \textbf{Markov transitions:} If the environment is
    non-Markovian (e.g., a POMDP), the conditional expectation
    given the full history does not simplify.
  \item \textbf{Fixed target network:} If $V_t$ is updated
    between computing the target and the current estimate
    (as in some batched implementations), the cancellation
    in Part~1 may not hold exactly.
  \item \textbf{Bounded rewards:} If rewards are unbounded,
    Part~2 fails.  Extensions require sub-Gaussian or
    sub-exponential tail conditions.
\end{enumerate}
\end{warning}

%% ============================================================
%% SECTION 7: CONVERGENCE VIA ROBBINS--MONRO
%% ============================================================

\section{Convergence via the Robbins--Monro Framework}
\label{sec:convergence}

We now state how the noise decomposition feeds into the
convergence machinery.

\subsection{The Robbins--Monro Template}

The Robbins--Monro stochastic approximation theorem
(Lesson~31) considers iterations of the form
\begin{equation}
  \theta_{n+1} = \theta_n + \alpha_n
    \bigl[h(\theta_n) + w_n\bigr],
  \label{eq:rm}
\end{equation}
and guarantees $\theta_n \to \theta^*$ (the root of $h$)
almost surely, provided:
\begin{enumerate}[label=(RM\arabic*)]
  \item $h$ has a unique root $\theta^*$ and is ``well
    behaved'' (e.g., $\langle \theta - \theta^*,
    h(\theta) \rangle < 0$ for $\theta \neq \theta^*$).
    \label{rm:drift}
  \item The step sizes satisfy $\sum_n \alpha_n = \infty$
    and $\sum_n \alpha_n^2 < \infty$.
    \label{rm:step}
  \item $\{w_n\}$ is a martingale difference sequence:
    $\E[w_n \mid \mathcal{F}_n] = 0$.
    \label{rm:mds}
  \item The noise has bounded conditional variance:
    $\E[w_n^2 \mid \mathcal{F}_n]
    \leq C(1 + \|\theta_n\|^2)$.
    \label{rm:var}
\end{enumerate}

\subsection{Applying the Template to TD(0)}

\begin{theorem}[Convergence of TD(0)]
\label{thm:td_convergence}
Let $(\mathcal{S}, \mathcal{A}, P, r, \gamma)$ be a finite
MDP.  Let $\pi$ be a stationary policy such that the induced
Markov chain is ergodic.  If the step sizes satisfy
$\sum_t \alpha_t(s) = \infty$ and
$\sum_t \alpha_t(s)^2 < \infty$ for every state $s$, and
every state is visited infinitely often, then the TD(0)
iterates satisfy
\begin{equation}
  V_t(s) \cas V^\pi(s)
  \qquad \text{for all } s \in \mathcal{S}.
  \label{eq:td_convergence}
\end{equation}
\end{theorem}

\begin{proof}[Proof sketch]
We verify the four Robbins--Monro conditions:
\begin{enumerate}
  \item \ref{rm:drift}: The mean field
    $h(V)(s) = (\mathcal{T}^\pi V)(s) - V(s)$ has unique
    root $V^\pi$ (the fixed point of $\mathcal{T}^\pi$).
    Furthermore, $\mathcal{T}^\pi$ is a
    $\gamma$-contraction, so
    $\|h(V) + V - V^\pi\|_\infty
    = \|\mathcal{T}^\pi V - V^\pi\|_\infty
    \leq \gamma \|V - V^\pi\|_\infty$,
    which implies the required stability condition.
  \item \ref{rm:step}: This is a hypothesis of the theorem.
  \item \ref{rm:mds}: Theorem~\ref{thm:td_noise}, Part~1.
  \item \ref{rm:var}: Theorem~\ref{thm:td_noise}, Part~2.
\end{enumerate}
The convergence then follows from the Robbins--Monro theorem
(Lesson~31, Theorem~2.1) or, equivalently, from the ODE
method (Lesson~32) applied to the limiting ODE
$\dot{v} = \mathcal{T}^\pi v - v$ whose globally
asymptotically stable equilibrium is $V^\pi$.
\end{proof}

\begin{rlconnection}[title={The Architecture of RL Convergence Proofs}]
Every convergence proof for a tabular RL algorithm follows
the same three-step architecture:
\begin{enumerate}[nosep]
  \item \textbf{Rewrite} the update in Robbins--Monro form:
    identify $h(\theta)$ and $w_n$.
  \item \textbf{Verify the noise conditions} using the TD
    noise decomposition (this lesson): show that $w_n$ is
    an MDS with bounded conditional variance.
  \item \textbf{Verify the mean-field conditions:}
    show that $h$ has a unique root and satisfies a
    stability/contraction condition.
\end{enumerate}
The probabilistic content is entirely in Step~2.
Steps~1 and~3 are algebraic and analytic, respectively.
\end{rlconnection}

%% ============================================================
%% SECTION 8: COMPREHENSIVE SUMMARY TABLE
%% ============================================================

\section{Summary of Key Results (Lessons 1--14)}
\label{sec:summary_table}

The following table collects every key result from the
Measure Theory track and explains its role in RL theory.
Results are organized by lesson and type.

\begin{center}
\renewcommand{\arraystretch}{1.4}
\small
\begin{tabular}{p{2.5cm} p{4.5cm} p{5.5cm}}
\toprule
\textbf{Result (Lesson)} & \textbf{Statement}
  & \textbf{Role in RL} \\
\midrule
\multicolumn{3}{l}{\textit{Foundations (Lessons 1--5)}} \\
\midrule
Sigma-algebra axioms (L1)
  & Closed under complement and countable union
  & Event space of an MDP \\
\addlinespace
Measure properties (L2)
  & Countable additivity, continuity from below/above
  & Probability axioms for $\Prob$ \\
\addlinespace
Measurable functions (L3)
  & Preimage of Borel sets is measurable
  & Random variables, policies as measurable maps \\
\addlinespace
Linearity of $\E$ (L5)
  & $\E[aX + bY] = a\E[X] + b\E[Y]$
  & Splitting $G_t = R_{t+1} + \gamma G_{t+1}$ \\
\addlinespace
MCT / DCT (L5)
  & Monotone and dominated convergence
  & Interchanging limits and $\E$, infinite-horizon
    value functions \\
\midrule
\multicolumn{3}{l}{\textit{Conditioning and Independence
  (Lessons 6--7)}} \\
\midrule
Cond.\ expectation (L6)
  & $\E[X|\mathcal{G}]$ as $L^2$ projection
  & Foundation of Bellman equation derivation \\
\addlinespace
Tower property (L6)
  & $\E[\E[X|\mathcal{G}]] = \E[X]$
  & Bellman recursion, MDS $\Rightarrow$ zero mean \\
\addlinespace
Pull-out property (L6)
  & $\E[h(Z)X | Z] = h(Z)\E[X|Z]$
  & Extracting $V(s)$ from $\E[\delta_t | S_t]$ \\
\addlinespace
Total variance (L7)
  & $\Var(X) = \E[\Var(X|Y)] + \Var(\E[X|Y])$
  & Bias-variance tradeoff in baselines \\
\addlinespace
Independence (L7)
  & $\E[\prod X_i] = \prod \E[X_i]$
  & Variance of sums, noise bounds \\
\midrule
\multicolumn{3}{l}{\textit{Martingales and Filtrations
  (Lesson 8)}} \\
\midrule
MDS definition (L8)
  & $\E[w_n | \mathcal{F}_{n-1}] = 0$
  & All SA convergence proofs \\
\addlinespace
MDS uncorrelated (L8)
  & $\E[w_m w_n] = 0$ for $m \neq n$
  & Noise variance bounds in SA \\
\midrule
\multicolumn{3}{l}{\textit{Inequalities (Lesson 9)}} \\
\midrule
Markov (L9)
  & $\Prob(X \geq a) \leq \E[X]/a$
  & Foundation of all tail bounds \\
\addlinespace
Chebyshev (L9)
  & $\Prob(|X-\mu| \geq \varepsilon) \leq
    \sigma^2/\varepsilon^2$
  & Weak law of large numbers \\
\addlinespace
Hoeffding (L9)
  & Sub-Gaussian tail for bounded averages
  & Finite-sample MC bounds \\
\addlinespace
Azuma--Hoeffding (L9)
  & Concentration for bounded MDS
  & Finite-sample TD bounds \\
\addlinespace
Jensen (L9)
  & $\varphi(\E[X]) \leq \E[\varphi(X)]$
  & $\|\mathbf{P}_\pi \mathbf{v}\|_{d^\pi}$
    contraction \\
\midrule
\multicolumn{3}{l}{\textit{Convergence and Limits
  (Lessons 10--11)}} \\
\midrule
Convergence modes (L10)
  & a.s.\ $\Rightarrow$ prob.\ $\Rightarrow$ dist.;\
    $L^p \Rightarrow$ prob.
  & Interpreting convergence theorems \\
\addlinespace
Borel--Cantelli (L10)
  & $\sum \Prob(A_n) < \infty \Rightarrow
    \Prob(A_n\text{ i.o.}) = 0$
  & Upgrading to a.s.\ convergence \\
\addlinespace
Strong LLN (L11)
  & $\bar{X}_n \cas \mu$
  & MC value estimation \\
\addlinespace
CLT (L11)
  & $\sqrt{n}(\bar{X}_n - \mu) \cdist N(0,\sigma^2)$
  & Confidence intervals for MC \\
\midrule
\multicolumn{3}{l}{\textit{Markov Chains (Lesson 12)}} \\
\midrule
Stationary dist.\ (L12)
  & Unique $\boldsymbol{d} > 0$ for irreducible chains
  & Weighting in $\overline{\text{VE}}$ \\
\addlinespace
Ergodic theorem (L12)
  & Time avg.\ $\cas$ space avg.\ under $d^\pi$
  & On-policy sampling justification \\
\midrule
\multicolumn{3}{l}{\textit{Martingale Convergence
  (Lessons 13--14)}} \\
\midrule
Upcrossing ineq.\ (L13)
  & $\E[U_n[a,b]] \leq \E[(X_n-a)^-]/(b-a)$
  & Foundation of Doob convergence \\
\addlinespace
Doob convergence (L13)
  & Bounded submartingales converge a.s.
  & Foundation of Robbins--Siegmund \\
\addlinespace
Robbins--Siegmund (L13)
  & Almost-supermartingale convergence
  & All SA convergence proofs \\
\addlinespace
TD noise (L14)
  & $\E[w_t | \mathcal{F}_t] = 0$,\
    $\E[w_t^2|\mathcal{F}_t] \leq C(1+\|V_t\|^2)$
  & MDS + variance bound for Robbins--Monro \\
\bottomrule
\end{tabular}
\end{center}

\begin{rlconnection}[title={The Complete Pipeline}]
The table above traces a complete logical chain from
the axioms of probability to the convergence of TD(0):
\[
  \underbrace{\text{Sigma-algebras}}_{\text{L1--2}}
  \;\to\;
  \underbrace{\E,\; \E[\cdot|\cdot]}_{\text{L5--6}}
  \;\to\;
  \underbrace{\text{MDS}}_{\text{L8}}
  \;\to\;
  \underbrace{\text{Convergence}}_{\text{L10--11}}
  \;\to\;
  \underbrace{\text{Ergodic thm}}_{\text{L12}}
  \;\to\;
  \underbrace{\text{Robbins--Siegmund}}_{\text{L13}}
  \;\to\;
  \underbrace{\text{TD noise}}_{\text{L14}}
\]
Each arrow represents a logical dependency: every result on
the right uses results from the left.
\end{rlconnection}

%% ============================================================
%% SECTION 9: EXERCISES
%% ============================================================

\section{Exercises}
\label{sec:exercises}

\begin{exercise}[SARSA Noise Decomposition ($\ast$)]
\label{ex:sarsa}
In SARSA, the update uses the actual next action $A_{t+1}$:
\[
  Q_{t+1}(S_t, A_t)
  = Q_t(S_t, A_t) + \alpha_t
    \bigl[R_{t+1} + \gamma Q_t(S_{t+1}, A_{t+1})
    - Q_t(S_t, A_t)\bigr].
\]
Define the SARSA Bellman operator
$(\mathcal{T}^\pi Q)(s,a) = \sum_{s'} P(s'|s,a)
[r(s,a,s') + \gamma \sum_{a'} \pi(a'|s') Q(s',a')]$ and
the noise $w_t^{\text{SARSA}}$.
\begin{enumerate}[label=(\alph*)]
  \item Write $w_t^{\text{SARSA}}$ explicitly.
  \item Prove that
    $\E[w_t^{\text{SARSA}} \mid \mathcal{F}_t] = 0$.
  \item Show that
    $\E[(w_t^{\text{SARSA}})^2 \mid \mathcal{F}_t]
    \leq C(1 + \|Q_t\|_\infty^2)$.
\end{enumerate}
\end{exercise}

\begin{exercise}[Non-Markovian Noise ($\ast\ast$)]
\label{ex:non_markov}
Consider a POMDP where the agent observes $O_t = g(S_t)$
instead of the true state $S_t$, and maintains value
estimates $V_t(o)$ for observations $o$.
\begin{enumerate}[label=(\alph*)]
  \item Define the analogous noise term $w_t(o)$ for
    a TD-like update on observations.
  \item Show by counterexample (a simple 3-state POMDP
    with 2 observations) that
    $\E[w_t(o) \mid \mathcal{F}_t] \neq 0$ in general.
  \item Explain intuitively why the Markov property is
    essential for the MDS cancellation.
\end{enumerate}
\end{exercise}

\begin{exercise}[Variance Reduction via Baselines ($\ast\ast$)]
\label{ex:baselines}
Consider the modified TD noise with a state-dependent baseline
$b(s)$:
\[
  \tilde{w}_t(s)
  = R_{t+1} + \gamma V_t(S_{t+1}) - b(s).
\]
\begin{enumerate}[label=(\alph*)]
  \item For what choice of $b(s)$ is $\tilde{w}_t(s)$
    a martingale difference sequence?
  \item Show that the conditional variance is minimized
    (over deterministic baselines $b$) when
    $b(s) = (\mathcal{T}^\pi V_t)(s)$, which recovers the
    original noise $w_t(s)$.
  \item Use the total variance formula (Lesson~7) to
    decompose $\Var(\tilde{w}_t(s) \mid \mathcal{F}_t)$
    into contributions from action randomness and
    transition randomness.
\end{enumerate}
\end{exercise}

\begin{exercise}[Noise in $n$-Step TD ($\ast\ast\ast$)]
\label{ex:nstep}
In $n$-step TD, the update target is
$G_t^{(n)} = \sum_{k=0}^{n-1} \gamma^k R_{t+k+1}
+ \gamma^n V_t(S_{t+n})$.  Define the noise
$w_t^{(n)} = G_t^{(n)} - (\mathcal{T}^\pi)^n V_t(S_t)$.
\begin{enumerate}[label=(\alph*)]
  \item Prove that $\E[w_t^{(n)} \mid \mathcal{F}_t] = 0$.
    \textit{Hint:} Use induction on $n$ and the tower
    property.
  \item Show that
    $\E[(w_t^{(n)})^2 \mid \mathcal{F}_t]
    \leq C_n(1 + \|V_t\|_\infty^2)$ where $C_n$ grows
    with $n$.
  \item Discuss the bias-variance tradeoff: why does
    increasing $n$ reduce bias (since
    $(\mathcal{T}^\pi)^n V \to V^\pi$) but increase
    variance (since $C_n$ grows)?
\end{enumerate}
\end{exercise}

\begin{exercise}[Verifying Robbins--Monro Conditions ($\ast$)]
\label{ex:verify_rm}
For the Q-learning update \eqref{eq:qlearning} in a finite
MDP with $\gamma = 0.9$ and $R_{\max} = 1$:
\begin{enumerate}[label=(\alph*)]
  \item Write the update in Robbins--Monro form
    $Q_{t+1}(s,a) = Q_t(s,a)
    + \alpha_t[h(Q_t)(s,a) + w_t^Q(s,a)]$.
    Identify $h$ explicitly.
  \item Compute the constant $C'$ from
    Proposition~\ref{prop:q_noise} for these parameters.
  \item Verify that
    $h(Q)(s,a) = (\mathcal{T}^* Q)(s,a) - Q(s,a) = 0$
    if and only if $Q = Q^*$ (the optimal action-value
    function).
\end{enumerate}
\end{exercise}

%% ============================================================
%% SECTION 10: SUMMARY
%% ============================================================

\section{Summary}
\label{sec:summary}

This lesson established the central probabilistic result
connecting measure theory to reinforcement learning:

\begin{itemize}[nosep]
  \item The \textbf{TD noise decomposition}
    (Theorem~\ref{thm:td_noise}) shows that the noise in
    temporal-difference updates is a martingale difference
    sequence with bounded conditional variance.  The proof
    uses two ingredients: the Markov property (to collapse
    conditioning on the full history to the current
    state-action pair) and reward boundedness (to control
    the variance).

  \item The same decomposition extends to \textbf{Q-learning}
    (Proposition~\ref{prop:q_noise}) and \textbf{expected
    SARSA}, with essentially identical proofs.

  \item Combined with the \textbf{Robbins--Monro framework}
    (Lessons~31--32), the noise decomposition yields the
    convergence of TD(0), Q-learning, and SARSA under
    standard step-size conditions.

  \item The \textbf{comprehensive summary table}
    (Section~\ref{sec:summary_table}) traces the logical
    dependencies from sigma-algebras (Lesson~1) through
    the TD noise decomposition (this lesson), showing how
    every measure-theoretic result contributes to the
    convergence proof pipeline.
\end{itemize}

This completes the Measure Theory track.  The tools
developed in Lessons~1--14 will be used repeatedly in
Phases~2--4 whenever we prove convergence of an RL algorithm.

%% ============================================================
%% SECTION 11: FURTHER READING
%% ============================================================

\section{Further Reading}
\label{sec:further_reading}

\begin{itemize}[nosep]
  \item \textbf{Stochastic approximation theory:}
    \citet{borkar2008} provides a comprehensive treatment
    of the Robbins--Monro and ODE methods with applications
    to RL.  See especially Chapters~2--3 for the noise
    conditions and Chapter~5 for two-timescale SA.

  \item \textbf{RL convergence proofs:}
    \citet{szepesvari2010}, Chapter~4, presents detailed
    convergence proofs for TD(0) and Q-learning with
    explicit noise decomposition.
    \citet{bertsekas2012}, Chapter~6, covers the same
    material from a dynamic programming perspective.

  \item \textbf{Martingale theory background:}
    \citet{williams1991} is an excellent introduction to
    the probability theory underlying this lesson.
    \citet{durrett2019}, Chapters~4--5, provides a more
    advanced treatment of martingale convergence.

  \item \textbf{Next steps:}
    Lesson~15 begins the Functional Analysis track, which
    provides the operator-theoretic framework (contraction
    mappings, projections) that complements the
    probabilistic tools from this track.  Lessons~31--32
    in the Optimization track formalize the Robbins--Monro
    and ODE methods that consume the noise conditions
    proved here.
\end{itemize}

\bibliography{measure-theory-refs}

\end{document}
