% lesson-57-actor-critic-and-gae.tex
% Phase 02, Subdomain: Policy Gradient
% Topic: Actor-Critic Architecture and Generalized Advantage Estimation
\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 57: Actor-Critic and GAE}}
\fancyhead[R]{\thepage}
\title{Lesson 57: Actor-Critic and Generalized Advantage Estimation}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% ========================================================================
%% SECTION 1: Motivation and Overview
%% ========================================================================
\section{Motivation and Overview}

REINFORCE (Lesson 55) uses the complete return $G_t = \sum_{k=0}^{T-t-1}
\discount^k r(S_{t+k}, A_{t+k})$ to estimate the advantage, yielding an
unbiased but high-variance gradient. At the other extreme, a one-step
temporal difference (TD) error $\delta_t = r(S_t, A_t) + \discount
\hat{V}(S_{t+1}) - \hat{V}(S_t)$ provides a low-variance but biased
estimate of the advantage. Both are limiting cases of a family of
estimators parameterized by a tradeoff parameter $\lambda \in [0,1]$.
Can we find the optimal point in this family?

This is exactly the question addressed by \emph{Generalized Advantage
Estimation} (GAE) of \citet{schulman2015}. GAE defines a
$\lambda$-weighted exponential average of multi-step advantage estimates,
analogous to $\operatorname{TD}(\lambda)$ in value estimation (Lesson 24),
but applied to the advantage function. The resulting estimator
$\hat{A}^{\mathrm{GAE}(\discount,\lambda)}_t$ provides a smooth knob
between bias and variance: $\lambda = 0$ gives the one-step TD
advantage (low variance, high bias), $\lambda = 1$ gives the Monte
Carlo advantage (high variance, zero bias). The tools developed here
form the backbone of modern policy gradient implementations including
PPO \citep{schulman2017} and A3C \citep{mnih2016}.

\begin{rlconnection}
\begin{itemize}
  \item \textbf{A2C/A3C} \citep{mnih2016}: Uses actor-critic with
    $n$-step returns; GAE generalizes this to $\lambda$-weighted
    multi-step returns.
  \item \textbf{PPO} \citep{schulman2017}: Uses GAE to estimate
    advantages within the clipped surrogate objective, with $\lambda
    = 0.95$ as a common default.
  \item \textbf{Natural actor-critic} \citep{peters2008}: The critic
    estimates the advantage using compatible features; GAE provides an
    alternative estimation strategy.
  \item \textbf{TRPO} \citep{schulman2015trpo}: The original GAE paper
    combines GAE with trust-region optimization for policy updates.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
The reader must know the policy gradient theorem (Lesson 54, Theorem 5.1),
the advantage function $A^\policy(s,a) = \Qpi(s,a) - \Vpi(s)$ and its
role in policy gradients (Lesson 55, Definition 4.3 and Proposition 4.5),
REINFORCE with baseline (Lesson 55, Section 4), and the compatible
function approximation theorem (Lesson 54, Theorem 8.2). From the
temporal difference methods lesson (Lesson 24), familiarity with
$\operatorname{TD}(\lambda)$, eligibility traces, and the bias-variance
tradeoff in bootstrapping is assumed. The natural policy gradient
(Lesson 56) provides useful context but is not strictly required.

\paragraph{Learning objectives.}
\begin{itemize}
  \item \textbf{Define} the actor-critic architecture and explain the
    roles of the actor (policy) and critic (value function).
  \item \textbf{Prove} that the one-step TD error $\delta_t$ is an
    unbiased estimate of the advantage $A^\policy(S_t, A_t)$
    when $\hat{V} = \Vpi$.
  \item \textbf{Derive} the $n$-step advantage estimator and analyze
    its bias and variance properties.
  \item \textbf{Derive} Generalized Advantage Estimation as an
    exponentially weighted average of $n$-step estimators.
  \item \textbf{Analyze} the bias-variance tradeoff of GAE as a
    function of $\lambda$.
  \item \textbf{Compute} GAE for concrete value function estimates
    on simple MDPs.
\end{itemize}

%% ========================================================================
%% SECTION 2: References
%% ========================================================================
\section{References}

\begin{itemize}
  \item \citet{schulman2015} -- The GAE paper; derives the
    $\lambda$-weighted advantage estimator with bias-variance analysis.
  \item \citet[Chapter 13]{suttonbarto2018} -- Actor-critic methods,
    including one-step actor-critic and eligibility traces.
  \item \citet{konda2003} -- Rigorous two-timescale convergence analysis
    of actor-critic algorithms.
  \item \citet{mnih2016} -- A3C: asynchronous actor-critic with
    $n$-step returns in deep RL.
  \item \citet{schulman2017} -- PPO: practical use of GAE in
    large-scale policy optimization.
  \item \citet{bhatnagar2009} -- Convergence of natural actor-critic
    with TD-based critics.
\end{itemize}

%% ========================================================================
%% SECTION 3: The Actor-Critic Architecture
%% ========================================================================
\section{The Actor-Critic Architecture}

Actor-critic methods maintain two components: a parameterized policy
(the actor) and a learned value function (the critic). The critic
provides low-variance estimates of the advantage, which the actor uses
to compute policy gradient updates.

\begin{definition}[Actor-Critic Components]
\label{def:actor-critic}
An \emph{actor-critic} system consists of:
\begin{enumerate}[label=(\roman*)]
  \item \textbf{Actor:} a parameterized policy $\policy_\theta(a \mid s)$
    with parameters $\theta \in \R^d$.
  \item \textbf{Critic:} a parameterized value function $\hat{V}_w(s)$
    with parameters $w \in \R^m$, serving as an approximation to
    $\Vpi(s)$.
\end{enumerate}
The actor is updated via policy gradient ascent using advantage
estimates provided by the critic. The critic is updated to minimize the
value prediction error, typically via temporal difference learning.
\end{definition}

\begin{definition}[One-Step Actor-Critic Update]
\label{def:one-step-ac}
At each time step $t$, given the transition $(S_t, A_t, R_{t+1}, S_{t+1})$:
\begin{enumerate}[label=\arabic*.]
  \item Compute the TD error: $\delta_t = R_{t+1} + \discount\,
    \hat{V}_w(S_{t+1}) - \hat{V}_w(S_t)$.
  \item Update the critic: $w \leftarrow w + \beta\, \delta_t\,
    \nabla_w \hat{V}_w(S_t)$.
  \item Update the actor: $\theta \leftarrow \theta + \alpha\,
    \discount^t\, \delta_t\, \nabla_\theta \log
    \policy_\theta(A_t \mid S_t)$.
\end{enumerate}
\end{definition}

\begin{rlconnection}
The one-step actor-critic is the simplest actor-critic algorithm. It
replaces the full return $G_t$ in REINFORCE with the TD error $\delta_t$,
which uses bootstrapping through $\hat{V}_w(S_{t+1})$. This introduces
bias (since $\hat{V}_w \neq \Vpi$ in general) but dramatically reduces
variance, since $\delta_t$ depends only on a single transition rather
than the entire trajectory. The two-timescale design (with step sizes
$\beta \gg \alpha$) allows the critic to track the value function of the
slowly changing actor.
\end{rlconnection}

%% ========================================================================
%% SECTION 4: The TD Error as Advantage Estimate
%% ========================================================================
\section{The TD Error as Advantage Estimate}

The temporal difference error provides a natural connection between
value function learning and advantage estimation.

\begin{definition}[Temporal Difference Error]
\label{def:td-error}
Given a value function estimate $\hat{V}$, the \emph{TD error} at
time $t$ is
\[
  \delta_t^{\hat{V}} = r(S_t, A_t) + \discount\, \hat{V}(S_{t+1})
  - \hat{V}(S_t).
\]
When $\hat{V} = \Vpi$, we write $\delta_t^\policy$ and call it the
\emph{true TD error}.
\end{definition}

\begin{theorem}[TD Error as Unbiased Advantage Estimate]
\label{thm:td-advantage}
When $\hat{V} = \Vpi$, the TD error is an unbiased estimate of the
advantage:
\[
  \E[\delta_t^\policy \mid S_t = s, A_t = a] = A^\policy(s, a).
\]
\end{theorem}

\begin{proof}
By definition and the Bellman equation:
\begin{align*}
  \E[\delta_t^\policy \mid S_t = s, A_t = a]
  &= \E[r(s,a) + \discount\, \Vpi(S_{t+1}) - \Vpi(s) \mid S_t = s, A_t = a] \\
  &= r(s,a) + \discount \sum_{s'} \Trans(s' \mid s, a)\, \Vpi(s') - \Vpi(s) \\
  &= \Qpi(s,a) - \Vpi(s) = A^\policy(s,a),
\end{align*}
where we used $\Qpi(s,a) = r(s,a) + \discount \sum_{s'} \Trans(s' \mid s,a)\,
\Vpi(s')$.
\end{proof}

\begin{keyresult}[title={TD Error and the Advantage Function}]
The one-step TD error $\delta_t^\policy = r(S_t, A_t) + \discount\,
\Vpi(S_{t+1}) - \Vpi(S_t)$ provides an unbiased, single-sample estimate
of $A^\policy(S_t, A_t)$ when the true value function is used. When an
approximate value function $\hat{V}_w$ is used instead, the estimate
becomes biased, with bias proportional to the value function
approximation error.
\end{keyresult}

\begin{proposition}[Bias from Value Function Approximation]
\label{prop:td-bias}
When $\hat{V}_w \neq \Vpi$, the TD error satisfies
\[
  \E[\delta_t^{\hat{V}_w} \mid S_t = s, A_t = a]
  = A^\policy(s,a) + \discount \sum_{s'} \Trans(s' \mid s,a)\,
  \bigl(\hat{V}_w(s') - \Vpi(s')\bigr) - \bigl(\hat{V}_w(s) - \Vpi(s)\bigr).
\]
The bias depends on the value function errors at $s$ and at the
successor states.
\end{proposition}

\begin{proof}
Adding and subtracting $\Vpi$: $\E[\delta_t^{\hat{V}_w} \mid s, a]
= \underbrace{r(s,a) + \discount \E[\Vpi(S_{t+1})] - \Vpi(s)}_{A^\policy(s,a)}
+ \discount \E[\hat{V}_w(S_{t+1}) - \Vpi(S_{t+1})]
- (\hat{V}_w(s) - \Vpi(s))$.
\end{proof}

\begin{warning}[title={Bias Cannot Be Eliminated by Baseline Invariance}]
Baseline invariance (Lesson 54, Proposition 7.1) guarantees that
subtracting $\Vpi(s)$ from $\Qpi(s,a)$ does not bias the gradient.
However, when the TD target $r + \discount \hat{V}_w(S_{t+1})$ replaces
$\Qpi$, the bootstrapping introduces bias that \emph{cannot} be removed
by baseline subtraction. This is fundamentally different from the
baseline effect: the bias comes from using $\hat{V}_w(S_{t+1})$ to
estimate future rewards, not from subtracting $\hat{V}_w(S_t)$.
\end{warning}

%% ========================================================================
%% SECTION 5: Multi-Step Advantage Estimators
%% ========================================================================
\section{Multi-Step Advantage Estimators}

We now generalize the one-step TD advantage to $n$-step estimators,
forming a family that interpolates between the TD(0) and Monte Carlo
extremes.

\begin{definition}[$n$-Step Advantage Estimator]
\label{def:n-step-advantage}
The \emph{$n$-step advantage estimator} at time $t$ is
\[
  \hat{A}_t^{(n)} = \sum_{k=0}^{n-1} \discount^k\, r(S_{t+k}, A_{t+k})
  + \discount^n\, \hat{V}(S_{t+n}) - \hat{V}(S_t)
  = -\hat{V}(S_t) + \sum_{k=0}^{n-1} \discount^k\, r(S_{t+k}, A_{t+k})
  + \discount^n\, \hat{V}(S_{t+n}).
\]
For $n = 1$, this recovers the TD error: $\hat{A}_t^{(1)} = \delta_t^{\hat{V}}$.
As $n \to \infty$ (for episodic tasks, $n = T - t$), this becomes the
Monte Carlo advantage: $\hat{A}_t^{(\infty)} = G_t - \hat{V}(S_t)$.
\end{definition}

\begin{proposition}[$n$-Step Estimator via TD Errors]
\label{prop:n-step-td}
The $n$-step advantage estimator can be written as a sum of discounted
TD errors:
\[
  \hat{A}_t^{(n)} = \sum_{k=0}^{n-1} \discount^k\, \delta_{t+k}^{\hat{V}}.
\]
\end{proposition}

\begin{proof}
Expanding the TD errors:
\begin{align*}
  \sum_{k=0}^{n-1} \discount^k\, \delta_{t+k}
  &= \sum_{k=0}^{n-1} \discount^k\, \bigl[r(S_{t+k}, A_{t+k})
    + \discount\, \hat{V}(S_{t+k+1}) - \hat{V}(S_{t+k})\bigr] \\
  &= \sum_{k=0}^{n-1} \discount^k\, r(S_{t+k}, A_{t+k})
    + \sum_{k=0}^{n-1} \discount^{k+1}\, \hat{V}(S_{t+k+1})
    - \sum_{k=0}^{n-1} \discount^k\, \hat{V}(S_{t+k}).
\end{align*}
The last two sums telescope: writing them out,
\begin{align*}
  &\bigl[\discount\, \hat{V}(S_{t+1}) + \discount^2\, \hat{V}(S_{t+2})
    + \cdots + \discount^n\, \hat{V}(S_{t+n})\bigr] \\
  &\quad - \bigl[\hat{V}(S_t) + \discount\, \hat{V}(S_{t+1})
    + \cdots + \discount^{n-1}\, \hat{V}(S_{t+n-1})\bigr]
  = \discount^n\, \hat{V}(S_{t+n}) - \hat{V}(S_t).
\end{align*}
Therefore $\sum_{k=0}^{n-1} \discount^k \delta_{t+k} =
\sum_{k=0}^{n-1} \discount^k r(S_{t+k}, A_{t+k}) + \discount^n
\hat{V}(S_{t+n}) - \hat{V}(S_t) = \hat{A}_t^{(n)}$.
\end{proof}

\begin{theorem}[Bias-Variance of $n$-Step Estimators]
\label{thm:n-step-bias-var}
Assume $\hat{V} = \Vpi + \epsilon$ where $\epsilon(s)$ is the
approximation error. Conditioning on $S_t = s$, $A_t = a$:
\begin{enumerate}[label=(\roman*)]
  \item \textbf{Bias:}
    $\E[\hat{A}_t^{(n)} \mid s, a] = A^\policy(s,a) + O(\discount^n\,
    \|\epsilon\|_\infty)$.
  \item \textbf{Variance:} The variance of $\hat{A}_t^{(n)}$ increases
    with $n$ because more random transitions are included.
\end{enumerate}
Larger $n$ reduces bias (the bootstrapped value $\hat{V}(S_{t+n})$
contributes less due to the $\discount^n$ discount) but increases
variance (more stochastic transitions are summed).
\end{theorem}

\begin{proof}
(i) Writing $\hat{V} = \Vpi + \epsilon$, the expectation of the
$n$-step estimator decomposes into $A^\policy(s,a)$ (from the true
values) plus error terms $\discount^n \E[\epsilon(S_{t+n})] -
\epsilon(s) = O(\|\epsilon\|_\infty)$, with the $\discount^n$ factor
shrinking as $n$ grows.

(ii) Each additional term in $\hat{A}_t^{(n)} = \sum_{k=0}^{n-1}
\discount^k \delta_{t+k}$ adds variance from the randomness of
$S_{t+k+1}$, so total variance is non-decreasing in $n$.
\end{proof}

\begin{example}[Comparing 1-Step and $\infty$-Step]
\label{ex:chain-mdp}
Consider a deterministic chain $S_0 \to S_1 \to \cdots \to S_T$ with
$r_t = 0$ for $t < T-1$, $r_{T-1} = 1$, $\discount = 1$, $\Vpi(S_t) = 1$,
and $\hat{V}(S_t) = 0.9$. The 1-step estimate $\hat{A}_0^{(1)} =
0 + 0.9 - 0.9 = 0$ misses the advantage (local errors cancel). The MC
estimate $\hat{A}_0^{(\infty)} = 1 - 0.9 = 0.1$ correctly identifies
positive advantage but carries full trajectory variance.
\end{example}

%% ========================================================================
%% SECTION 6: Generalized Advantage Estimation
%% ========================================================================
\section{Generalized Advantage Estimation}

GAE provides a principled way to interpolate between the $n$-step
estimators by taking an exponentially weighted average.

\begin{definition}[Generalized Advantage Estimation]
\label{def:gae}
The \emph{Generalized Advantage Estimator} $\operatorname{GAE}(\discount,
\lambda)$ at time $t$ is the exponentially weighted average of $n$-step
advantage estimators:
\[
  \hat{A}_t^{\mathrm{GAE}(\discount,\lambda)}
  = (1 - \lambda) \sum_{n=1}^{\infty} \lambda^{n-1}\, \hat{A}_t^{(n)},
\]
where $\lambda \in [0,1]$ controls the bias-variance tradeoff.
\end{definition}

\begin{theorem}[GAE as Discounted Sum of TD Errors]
\label{thm:gae-td}
The GAE estimator admits the simple closed-form expression
\[
  \hat{A}_t^{\mathrm{GAE}(\discount,\lambda)}
  = \sum_{k=0}^{\infty} (\discount\lambda)^k\, \delta_{t+k}^{\hat{V}}.
\]
For episodic tasks terminating at time $T$:
\[
  \hat{A}_t^{\mathrm{GAE}(\discount,\lambda)}
  = \sum_{k=0}^{T-t-1} (\discount\lambda)^k\, \delta_{t+k}^{\hat{V}}.
\]
\end{theorem}

\begin{proof}
Substituting $\hat{A}_t^{(n)} = \sum_{k=0}^{n-1} \discount^k\,
\delta_{t+k}$ (Proposition~\ref{prop:n-step-td}) into the GAE definition:
\begin{align*}
  \hat{A}_t^{\mathrm{GAE}}
  &= (1-\lambda) \sum_{n=1}^{\infty} \lambda^{n-1}
    \sum_{k=0}^{n-1} \discount^k\, \delta_{t+k} \\
  &= (1-\lambda) \sum_{k=0}^{\infty} \discount^k\, \delta_{t+k}
    \sum_{n=k+1}^{\infty} \lambda^{n-1},
\end{align*}
where we swapped the order of summation (justified by absolute
convergence, since $|\delta_{t+k}| \leq R_{\max}/(1-\discount)$ and
$\discount\lambda < 1$). The inner sum is:
\[
  \sum_{n=k+1}^{\infty} \lambda^{n-1}
  = \lambda^k \sum_{j=0}^{\infty} \lambda^j
  = \frac{\lambda^k}{1 - \lambda}.
\]
Therefore:
\[
  \hat{A}_t^{\mathrm{GAE}}
  = (1-\lambda) \sum_{k=0}^{\infty} \discount^k\, \delta_{t+k}\,
  \frac{\lambda^k}{1-\lambda}
  = \sum_{k=0}^{\infty} (\discount\lambda)^k\, \delta_{t+k}. \qedhere
\]
\end{proof}

\begin{keyresult}[title={Generalized Advantage Estimation}]
\[
  \hat{A}_t^{\mathrm{GAE}(\discount,\lambda)}
  = \sum_{k=0}^{\infty} (\discount\lambda)^k\, \delta_{t+k}^{\hat{V}},
\]
where $\delta_{t+k}^{\hat{V}} = r(S_{t+k}, A_{t+k}) + \discount\,
\hat{V}(S_{t+k+1}) - \hat{V}(S_{t+k})$. The parameter $\lambda$
interpolates between:
\begin{itemize}
  \item $\lambda = 0$: $\hat{A}_t^{\mathrm{GAE}} = \delta_t^{\hat{V}}$
    (one-step TD advantage; low variance, high bias).
  \item $\lambda = 1$: $\hat{A}_t^{\mathrm{GAE}} = \sum_{k=0}^{\infty}
    \discount^k \delta_{t+k} = G_t - \hat{V}(S_t)$ (Monte Carlo
    advantage; zero bias when $\hat{V} = \Vpi$, high variance).
\end{itemize}
\end{keyresult}

\begin{remark}[Recursive Computation of GAE]
\label{rem:recursive-gae}
In practice, GAE is computed recursively backward through the trajectory:
\[
  \hat{A}_{T-1}^{\mathrm{GAE}} = \delta_{T-1}, \qquad
  \hat{A}_t^{\mathrm{GAE}} = \delta_t + \discount\lambda\,
  \hat{A}_{t+1}^{\mathrm{GAE}} \quad \text{for } t = T-2, \ldots, 0.
\]
This follows directly from the geometric series structure and costs
$O(T)$ time for an episode of length $T$.
\end{remark}

%% ========================================================================
%% SECTION 7: Bias-Variance Analysis of GAE
%% ========================================================================
\section{Bias-Variance Analysis of GAE}

We now formally analyze the bias and variance of GAE as a function of
$\lambda$.

\begin{theorem}[Bias of GAE]
\label{thm:gae-bias}
Let $\hat{V} = \Vpi + \epsilon$ where $\epsilon : \States \to \R$ is
the value function approximation error. Then:
\[
  \E\!\left[\hat{A}_t^{\mathrm{GAE}(\discount,\lambda)} \mid S_t = s, A_t = a\right]
  = A^\policy(s,a) + \sum_{k=0}^{\infty} (\discount\lambda)^k\,
  b_{t+k}(s,a),
\]
where $b_{t+k}$ represents the bias contribution from the $k$-th TD
error, satisfying $|b_{t+k}(s,a)| \leq (1 + \discount)\|\epsilon\|_\infty$.
\end{theorem}

\begin{proof}
By Proposition~\ref{prop:td-bias}, each TD error has bias
$\E[\delta_{t+k}^{\hat{V}} \mid S_{t+k}, A_{t+k}] = A^\policy +
\discount\, \E[\epsilon(S_{t+k+1})] - \epsilon(S_{t+k})$. Summing with
weights $(\discount\lambda)^k$ and taking iterated expectations gives
$\E[\hat{A}_t^{\mathrm{GAE}} \mid s, a] = A^\policy(s,a) +
\text{bias terms}$. For $\lambda < 1$, the total bias is bounded by
$(1+\discount)\|\epsilon\|_\infty / (1-\discount\lambda)$.
\end{proof}

\begin{corollary}[Unbiasedness at $\lambda = 1$]
\label{cor:gae-unbiased}
When $\hat{V} = \Vpi$ (i.e., $\epsilon = 0$),
$\hat{A}_t^{\mathrm{GAE}(\discount,\lambda)}$ is an unbiased estimate
of $A^\policy(S_t, A_t)$ for \emph{all} $\lambda \in [0,1]$. When
$\hat{V} \neq \Vpi$, the estimator is unbiased only at $\lambda = 1$
(the Monte Carlo case).
\end{corollary}

\begin{proof}
When $\epsilon = 0$, each TD error $\delta_{t+k}^\policy$ satisfies
$\E[\delta_{t+k}^\policy \mid S_{t+k}, A_{t+k}] = A^\policy(S_{t+k},
A_{t+k})$ by Theorem~\ref{thm:td-advantage}. For $\lambda = 1$,
$\hat{A}_t^{\mathrm{GAE}(\discount,1)} = \sum_{k=0}^\infty \discount^k
\delta_{t+k} = G_t - \hat{V}(S_t) = G_t - \Vpi(S_t)$, which is an
unbiased estimate of $\Qpi(S_t, A_t) - \Vpi(S_t) = A^\policy(S_t, A_t)$
by Lemma 3.2 of Lesson 55.

For $\lambda < 1$ and $\epsilon \neq 0$, the bias is
$O(\|\epsilon\|_\infty / (1 - \discount\lambda)) \neq 0$ in general.
\end{proof}

\begin{theorem}[Variance of GAE]
\label{thm:gae-variance}
Consider the simplified setting where the TD errors
$\delta_0, \delta_1, \ldots$ are uncorrelated (which holds when $\hat{V}
= \Vpi$ and the MDP is such that each $\delta_t$ depends only on the
local transition). Then the variance of GAE is:
\[
  \Var\!\left(\hat{A}_t^{\mathrm{GAE}(\discount,\lambda)}\right)
  = \sum_{k=0}^{\infty} (\discount\lambda)^{2k}\,
  \Var(\delta_{t+k}^\policy).
\]
If additionally $\Var(\delta_{t+k}^\policy) \leq \sigma_\delta^2$ for
all $k$, then
\[
  \Var\!\left(\hat{A}_t^{\mathrm{GAE}}\right)
  \leq \frac{\sigma_\delta^2}{1 - \discount^2 \lambda^2}.
\]
\end{theorem}

\begin{proof}
Under the uncorrelatedness assumption:
\begin{align*}
  \Var\!\left(\sum_{k=0}^\infty (\discount\lambda)^k \delta_{t+k}\right)
  &= \sum_{k=0}^\infty (\discount\lambda)^{2k}\, \Var(\delta_{t+k}).
\end{align*}
Bounding each $\Var(\delta_{t+k}) \leq \sigma_\delta^2$ and summing the
geometric series gives the result.
\end{proof}

\begin{rlconnection}
The bias-variance tradeoff is the central design choice in GAE:
\begin{itemize}
  \item \textbf{Small $\lambda$} (e.g., $\lambda = 0$): Variance
    $\approx \sigma_\delta^2$, bias $\approx
    (1+\discount)\|\epsilon\|_\infty / (1-\discount\lambda)$. Good when
    the critic is accurate ($\|\epsilon\|$ small).
  \item \textbf{Large $\lambda$} (e.g., $\lambda = 0.95$): Variance
    $\approx \sigma_\delta^2 / (1-\discount^2\lambda^2)$, bias is
    small even with inaccurate critic. This is the typical choice in
    practice (PPO uses $\lambda = 0.95$ by default).
  \item \textbf{$\lambda = 1$}: Zero bias, maximum variance. Equivalent
    to REINFORCE with baseline.
\end{itemize}
\end{rlconnection}

%% ========================================================================
%% SECTION 8: The GAE Policy Gradient
%% ========================================================================
\section{The GAE Policy Gradient}

We now state the policy gradient estimator that uses GAE advantages
and analyze its properties.

\begin{definition}[GAE Policy Gradient Estimator]
\label{def:gae-pg}
Given a trajectory $(S_0, A_0, R_1, \ldots, S_{T-1}, A_{T-1}, R_T)$
collected under $\policy_\theta$, the \emph{GAE policy gradient
estimator} is
\[
  \hat{g}^{\mathrm{GAE}} = \sum_{t=0}^{T-1}
  \nabla_\theta \log \policy_\theta(A_t \mid S_t)\,
  \hat{A}_t^{\mathrm{GAE}(\discount,\lambda)}.
\]
\end{definition}

\begin{proposition}[Bias of the GAE Policy Gradient]
\label{prop:gae-pg-bias}
When $\hat{V} = \Vpi$, $\hat{g}^{\mathrm{GAE}}$ is unbiased for all
$\lambda$ (by Corollary~\ref{cor:gae-unbiased}). When $\hat{V} \neq
\Vpi$ and $\lambda < 1$, the estimator is biased
(Theorem~\ref{thm:gae-bias}). When $\lambda = 1$,
$\hat{A}_t^{\mathrm{GAE}} = G_t - \hat{V}(S_t)$ and $\hat{V}(S_t)$
acts as a state-dependent baseline, so the estimator is unbiased by
baseline invariance (Lesson 54, Proposition 7.1) regardless of $\hat{V}$.
\end{proposition}

\begin{example}[GAE Computation on a Simple Trajectory]
\label{ex:gae-computation}
Consider a trajectory of length $T = 4$ with $\discount = 0.99$,
$\lambda = 0.95$, and the following data:
\begin{center}
\begin{tabular}{ccccc}
\toprule
$t$ & $r_t$ & $\hat{V}(S_t)$ & $\hat{V}(S_{t+1})$ & $\delta_t$ \\
\midrule
0 & 1.0 & 10.0 & 8.0 & $1.0 + 0.99 \times 8.0 - 10.0 = -1.08$ \\
1 & 2.0 & 8.0 & 6.0 & $2.0 + 0.99 \times 6.0 - 8.0 = -0.06$ \\
2 & 3.0 & 6.0 & 5.0 & $3.0 + 0.99 \times 5.0 - 6.0 = 1.95$ \\
3 & 4.0 & 5.0 & 0.0 & $4.0 + 0.99 \times 0.0 - 5.0 = -1.00$ \\
\bottomrule
\end{tabular}
\end{center}

Computing GAE recursively backward (Remark~\ref{rem:recursive-gae})
with $\discount\lambda = 0.99 \times 0.95 = 0.9405$:
\begin{align*}
  \hat{A}_3^{\mathrm{GAE}} &= \delta_3 = -1.00, \\
  \hat{A}_2^{\mathrm{GAE}} &= \delta_2 + 0.9405 \times (-1.00)
    = 1.95 - 0.94 = 1.01, \\
  \hat{A}_1^{\mathrm{GAE}} &= \delta_1 + 0.9405 \times 1.01
    = -0.06 + 0.95 = 0.89, \\
  \hat{A}_0^{\mathrm{GAE}} &= \delta_0 + 0.9405 \times 0.89
    = -1.08 + 0.84 = -0.24.
\end{align*}

For comparison, with $\lambda = 0$ (TD-only): $\hat{A}_t = \delta_t$,
so $\hat{A}_0 = -1.08$. With $\lambda = 1$ (Monte Carlo):
$\hat{A}_0 = (1 + 0.99 \times 2 + 0.99^2 \times 3 + 0.99^3 \times 4)
- 10.0 = 9.80 - 10.0 = -0.20$. The GAE estimate ($-0.24$) lies between
the TD ($-1.08$) and MC ($-0.20$) values, as expected.
\end{example}

%% ========================================================================
%% SECTION 9: Two-Timescale Convergence of Actor-Critic
%% ========================================================================
\section{Two-Timescale Convergence of Actor-Critic}

\begin{assumption}[Two-Timescale Step Sizes]
\label{assump:two-timescale}
The actor step sizes $\{\alpha_k\}$ and critic step sizes $\{\beta_k\}$
satisfy: (i) $\sum_k \alpha_k = \sum_k \beta_k = \infty$ and
$\sum_k \alpha_k^2, \sum_k \beta_k^2 < \infty$; (ii)
$\alpha_k / \beta_k \to 0$ (the actor updates on a slower timescale).
\end{assumption}

\begin{theorem}[Two-Timescale Convergence]
\label{thm:two-timescale}
Under Assumption~\ref{assump:two-timescale} and standard regularity
conditions (Lipschitz gradients, bounded iterates, ergodicity), the
critic $w_k$ converges to $w^*(\theta)$ (the TD fixed point) and the
actor $\theta_k$ converges to a stationary point of $J(\theta)$.
\end{theorem}

\begin{proof}[Proof sketch]
Since $\alpha_k / \beta_k \to 0$, the critic sees the actor as
approximately constant and converges by standard TD theory (Lesson 24).
The actor then uses approximately correct gradient estimates. The
formal proof \citep{konda2003} uses the ODE method: coupled stochastic
iterations track a pair of ODEs, with the fast ODE (critic) converging
on the slow timescale of the actor.
\end{proof}

\begin{rlconnection}
In practice, the strict two-timescale requirement is often relaxed.
PPO and A3C use shared neural network layers for actor and critic, with
the critic loss coefficient (e.g., $c_v = 0.5$) implicitly controlling
the timescale ratio.
\end{rlconnection}

%% ========================================================================
%% SECTION 10: GAE with Policy Gradient Methods
%% ========================================================================
\section{GAE with Policy Gradient Methods}

GAE can be combined with any policy gradient method by replacing
the advantage estimate. The vanilla policy gradient with GAE uses:
\[
  \theta_{k+1} = \theta_k + \alpha \sum_{t=0}^{T-1}
  \nabla_\theta \log \policy_{\theta_k}(A_t \mid S_t)\,
  \hat{A}_t^{\mathrm{GAE}(\discount,\lambda)}.
\]
The natural policy gradient with GAE (Lesson 56) instead uses
$\theta_{k+1} = \theta_k + \alpha\, F(\theta_k)^{-1} \hat{g}$
where $\hat{g}$ is the GAE gradient estimate above.

\begin{example}[Effect of $\lambda$ on Credit Assignment]
\label{ex:lambda-effect}
In a grid-world with sparse rewards ($+1$ at the goal, $0$ elsewhere):
$\lambda = 0$ provides signal only one step from the goal (slow
learning); $\lambda = 1$ propagates reward to all visited states
(fast but noisy); $\lambda = 0.95$ propagates $\sim 20$ effective
steps, balancing speed and stability.
\end{example}

\begin{warning}[title={Two Discount Parameters in GAE}]
GAE has \emph{two} discount-like parameters: $\discount$ (the MDP
discount defining the objective) and $\lambda$ (controlling the
bias-variance tradeoff). Setting $\lambda = 0.95$ with $\discount =
0.99$ gives an effective estimation discount of $\discount\lambda =
0.9405$. A common mistake is to confuse the two or set both to 1.
\end{warning}

%% ========================================================================
%% SECTION 11: Exercises
%% ========================================================================
\section{Exercises}

\begin{exercise}[TD Error Properties]
\textnormal{(\(*\))} \\
  For an MDP with $\States = \{s_1, s_2\}$, $\Actions = \{a\}$
  (single action), deterministic transitions $s_1 \to s_2 \to s_1$,
  rewards $r(s_1, a) = 2$, $r(s_2, a) = 0$, and $\discount = 0.9$:
  \begin{enumerate}[label=(\alph*)]
    \item Compute $\Vpi(s_1)$ and $\Vpi(s_2)$ exactly.
    \item Verify that $\E[\delta_t^\policy \mid S_t = s_i]$ equals zero
      for both states (since there is only one action, the advantage is
      zero).
    \item If $\hat{V}(s_1) = 9$ and $\hat{V}(s_2) = 8$, compute the
      TD errors and the bias at each state.
  \end{enumerate}
\end{exercise}

\begin{exercise}[GAE Recursive Computation]
\textnormal{(\(*\))} \\
  Verify the recursive formula $\hat{A}_t^{\mathrm{GAE}} = \delta_t
  + \discount\lambda\, \hat{A}_{t+1}^{\mathrm{GAE}}$ by expanding
  the closed-form expression from Theorem~\ref{thm:gae-td}. Then
  compute $\hat{A}_t^{\mathrm{GAE}}$ for $t = 0,1,2$ on the trajectory
  $(r_0, r_1, r_2) = (1, 0, 5)$ with $\hat{V}(S_0) = 3$,
  $\hat{V}(S_1) = 2$, $\hat{V}(S_2) = 6$, $\hat{V}(S_3) = 0$ (terminal),
  $\discount = 0.99$, $\lambda = 0.9$.
\end{exercise}

\begin{exercise}[Bias-Variance Comparison]
\textnormal{(\(**\))} \\
  Consider a two-state MDP where $S_{t+1} \sim
  \mathrm{Bernoulli}(0.5)$ independently, with $r(s, a) = s$ (the
  reward equals the state, treating $s \in \{0, 1\}$) and
  $\discount = 0.9$. Suppose $\hat{V}(0) = \hat{V}(1) = 0.5$ and
  $\Vpi(0) = \Vpi(1) = 5.0$.
  \begin{enumerate}[label=(\alph*)]
    \item Compute $\E[\hat{A}_0^{(1)}]$ and $\Var(\hat{A}_0^{(1)})$
      exactly.
    \item Compute $\E[\hat{A}_0^{(2)}]$ and show that the bias decreases
      with $n$.
    \item For $\lambda = 0.5$, compute $\E[\hat{A}_0^{\mathrm{GAE}}]$
      and compare with parts (a) and (b).
  \end{enumerate}
\end{exercise}

\begin{exercise}[$\lambda$-Return Equivalence]
\textnormal{(\(**\))} \\
  The \emph{$\lambda$-return} from TD($\lambda$) is defined as
  $G_t^\lambda = (1-\lambda) \sum_{n=1}^{\infty} \lambda^{n-1}\,
  G_t^{(n)}$ where $G_t^{(n)} = \sum_{k=0}^{n-1} \discount^k\,
  r_{t+k} + \discount^n\, \hat{V}(S_{t+n})$. Show that
  \[
    \hat{A}_t^{\mathrm{GAE}(\discount,\lambda)} = G_t^\lambda - \hat{V}(S_t),
  \]
  establishing that GAE is equivalent to using the $\lambda$-return as
  the target for the advantage, with $\hat{V}(S_t)$ as the baseline.
\end{exercise}

\begin{exercise}[Optimal $\lambda$ Selection]
\textnormal{(\(***\))} \\
  Consider the idealized setting where $\hat{V}(s) = \Vpi(s) +
  \epsilon$ with $\epsilon$ a constant (uniform bias). Show that in
  this case, the bias of $\hat{A}_t^{\mathrm{GAE}(\discount,\lambda)}$
  is $\epsilon(\discount\lambda - 1)/(1 - \discount\lambda)$ for
  $\lambda < 1$ and zero for $\lambda = 1$. Then derive the
  mean-squared error $\mathrm{MSE} = \mathrm{bias}^2 + \mathrm{variance}$
  as a function of $\lambda$ (using the variance bound from
  Theorem~\ref{thm:gae-variance}) and find the $\lambda^*$ that
  minimizes it as a function of $\epsilon$ and $\sigma_\delta$.

  \textit{Hint:} The MSE-optimal $\lambda$ balances
  $\partial(\mathrm{bias}^2)/\partial\lambda$ against
  $\partial(\mathrm{variance})/\partial\lambda$.
\end{exercise}

%% ========================================================================
%% SECTION 12: Summary
%% ========================================================================
\section{Summary}

\begin{itemize}
  \item The \textbf{actor-critic architecture} separates policy
    optimization (actor) from value function estimation (critic),
    allowing the critic to provide low-variance advantage estimates
    through bootstrapping.

  \item The \textbf{TD error} $\delta_t = r_t + \discount \hat{V}(S_{t+1})
    - \hat{V}(S_t)$ is an unbiased estimate of the advantage
    $A^\policy(S_t, A_t)$ when $\hat{V} = \Vpi$, but introduces bias
    when $\hat{V} \neq \Vpi$.

  \item The \textbf{$n$-step advantage estimator} $\hat{A}_t^{(n)} =
    \sum_{k=0}^{n-1} \discount^k \delta_{t+k}$ interpolates between
    the TD error ($n=1$) and the Monte Carlo advantage ($n=\infty$),
    with increasing $n$ reducing bias but increasing variance.

  \item \textbf{Generalized Advantage Estimation} defines
    $\hat{A}_t^{\mathrm{GAE}(\discount,\lambda)} = \sum_{k=0}^\infty
    (\discount\lambda)^k \delta_{t+k}$, providing a smooth
    bias-variance tradeoff via $\lambda \in [0,1]$.

  \item At $\lambda = 0$, GAE reduces to the one-step \textbf{TD
    advantage} (low variance, high bias). At $\lambda = 1$, it becomes
    the \textbf{Monte Carlo advantage} (zero bias, high variance).
    Typical implementations use $\lambda = 0.95$.

  \item GAE is equivalent to using the \textbf{$\lambda$-return}
    $G_t^\lambda$ as the advantage target with the value function as
    baseline, connecting policy gradient methods to the
    $\operatorname{TD}(\lambda)$ framework.

  \item \textbf{Two-timescale convergence}: When the critic learns
    faster than the actor ($\alpha_k / \beta_k \to 0$), the coupled
    system converges to a stationary point of $J(\theta)$.

  \item GAE can be combined with \textbf{any policy gradient method}
    (vanilla, natural, trust-region) by replacing the advantage
    estimate in the gradient expression.
\end{itemize}

%% ========================================================================
%% SECTION 13: Further Reading
%% ========================================================================
\section{Further Reading}

\begin{itemize}
  \item \textbf{Lesson 58} synthesizes policy gradient theory from
    Lessons 54--57, connecting to trust-region methods of Phase 04.

  \item \citet{schulman2015} provides experiments showing the effect of
    $\lambda$ on learning speed across continuous control benchmarks.

  \item \citet{konda2003} gives the definitive two-timescale convergence
    proof for actor-critic methods.

  \item \citet{mnih2016} demonstrates $n$-step actor-critic in A3C
    with asynchronous parallel workers.

  \item \citet{schulman2017} uses GAE within PPO, showing $\lambda =
    0.95$--$0.99$ works well across continuous control tasks.

  \item \citet[Chapter 12]{suttonbarto2018} develops
    $\operatorname{TD}(\lambda)$, from which GAE draws inspiration.
\end{itemize}

\bibliography{pg-refs}

\end{document}
