\documentclass[11pt,a4paper]{article}

% ---- Packages ----
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage[margin=1in,headheight=14pt]{geometry}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{tcolorbox}
\usepackage{fancyhdr}
\usepackage{booktabs}

% ---- Hyperref ----
\hypersetup{
    colorlinks=true,
    linkcolor=red,
    citecolor=blue,
    urlcolor=blue
}

% ---- Header ----
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textit{Lesson 10: Stochastic Approximation}}
\fancyhead[R]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% ---- Theorem Environments ----
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{example}{Example}[section]
\newtheorem{exercise}{Exercise}[section]

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{remark}
\newtheorem{remark}{Remark}[section]

% ---- Colored Boxes ----
\tcbuselibrary{theorems,skins,breakable}

\newtcolorbox{keyresult}[1][]{
    colback=green!5!white,
    colframe=green!50!black,
    fonttitle=\bfseries,
    title=#1,
    breakable
}

\newtcolorbox{rlconnection}[1][]{
    colback=blue!5!white,
    colframe=blue!50!black,
    fonttitle=\bfseries,
    title=#1,
    breakable
}

\newtcolorbox{warning}[1][]{
    colback=red!5!white,
    colframe=red!50!black,
    fonttitle=\bfseries,
    title=#1,
    breakable
}

% ---- Operators ----
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Prob}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\indicator}{\mathbf{1}}
\newcommand{\cas}{\xrightarrow{\text{a.s.}}}
\newcommand{\cprob}{\xrightarrow{\Prob}}
\newcommand{\cdist}{\xrightarrow{d}}

% ---- Title ----
\title{Lesson 32: Convergence Rates, Two-Timescale SA, and RL Applications}
\author{Curriculum: A Rigorous Learning Path to Multi-Agent Deep RL}
\date{}

\begin{document}
\maketitle
\tableofcontents
\newpage

\setcounter{section}{3}

\section{Convergence Rate Analysis}
\label{sec:convergence-rate}

Having established convergence, we now ask: \emph{how fast} does $x_n \to x^*$?

\subsection{Central limit theorem for SA}

\begin{definition}[Assumptions for the CLT]
\label{def:clt-assumptions}
Assume \ref{B1}--\ref{B5} hold, and additionally:
\begin{enumerate}[label=(D\arabic*)]
\item \label{D1} $h$ is differentiable at $x^*$ with Jacobian
  $A = Dh(x^*) = \nabla h(x^*) \in \R^{d \times d}$.
\item \label{D2} $A$ is Hurwitz: all eigenvalues of $A$ have strictly negative real
  parts. (This is automatic from the asymptotic stability of $x^*$ for the linearized
  ODE.)
\item \label{D3} The step size is $a_n = a / n$ for some constant $a > 0$ such that
  $aA + \frac{1}{2}I$ is Hurwitz (i.e., all eigenvalues of $aA$ have real part less
  than $-1/2$).
\item \label{D4} The conditional covariance converges:
  $\E[M_{n+1} M_{n+1}^T \mid \mathcal{F}_n] \to \Sigma$ in probability as
  $x_n \to x^*$, for some positive definite matrix $\Sigma$.
\end{enumerate}
\end{definition}

\begin{theorem}[CLT for stochastic approximation]
\label{thm:sa-clt}
Under the assumptions of Definition~\ref{def:clt-assumptions}, we have
\[
  \sqrt{n}(x_n - x^*) \cdist \mathcal{N}(0, V),
\]
where $V$ is the unique solution of the \emph{Lyapunov equation}
\begin{equation}
\label{eq:lyapunov}
  (aA + \tfrac{1}{2}I) V + V (aA + \tfrac{1}{2}I)^T + a^2 \Sigma = 0.
\end{equation}
\end{theorem}

\begin{proof}
Define $e_n = x_n - x^*$ and $u_n = \sqrt{n}\, e_n$. Near $x^*$, the drift is
approximately linear:
\[
  h(x_n) = h(x^*) + A e_n + o(\|e_n\|) = A e_n + o(\|e_n\|),
\]
since $h(x^*) = 0$. With $a_n = a/n$, the SA iteration gives
\[
  e_{n+1} = e_n + \frac{a}{n}[A e_n + r_n + M_{n+1}],
\]
where $r_n = h(x_n) - A e_n = o(\|e_n\|)$ is the higher-order remainder.

Substituting $e_n = u_n / \sqrt{n}$:
\[
  \frac{u_{n+1}}{\sqrt{n+1}}
    = \frac{u_n}{\sqrt{n}} + \frac{a}{n}\Bigl[A \frac{u_n}{\sqrt{n}} + r_n + M_{n+1}\Bigr].
\]
Multiplying both sides by $\sqrt{n+1}$ and using $\sqrt{n+1}/\sqrt{n} = 1 + 1/(2n) + O(1/n^2)$:
\begin{align}
  u_{n+1} &= \sqrt{\frac{n+1}{n}} \, u_n + \frac{a\sqrt{n+1}}{n}
    \Bigl[A \frac{u_n}{\sqrt{n}} + r_n + M_{n+1}\Bigr] \notag \\
  &= \Bigl(1 + \frac{1}{2n} + O(1/n^2)\Bigr) u_n
    + \frac{a}{n}\Bigl(1 + O(1/n)\Bigr)[A u_n + \sqrt{n}\, r_n + \sqrt{n}\, M_{n+1}] \notag \\
  &= u_n + \frac{1}{n}\Bigl(aA + \frac{1}{2}I\Bigr) u_n
    + \frac{a}{\sqrt{n}} M_{n+1} + \text{negligible terms}. \label{eq:u-recursion}
\end{align}
The term $\sqrt{n}\, r_n / n = r_n / \sqrt{n}$ vanishes because $r_n = o(\|e_n\|) = o(\|u_n\|/\sqrt{n})$ and $u_n = O(1)$ in the limit.

The recursion~\eqref{eq:u-recursion} is itself a stochastic approximation with step
size $1/n$, drift $g(u) = (aA + \frac{1}{2}I)u$, and diffusion coefficient
$a/\sqrt{n} \cdot \sqrt{n} = a$ times the noise $M_{n+1}$. By \ref{D3}, the matrix
$aA + \frac{1}{2}I$ is Hurwitz, so the origin is a stable equilibrium of the drift.
The process $\{u_n\}$ is a linear stochastic recursion whose stationary distribution
is Gaussian.

Standard results on such recursions (or a direct characteristic function argument)
yield $u_n \cdist \mathcal{N}(0, V)$ where $V$ satisfies the Lyapunov
equation~\eqref{eq:lyapunov}. Specifically, the covariance $V_n = \E[u_n u_n^T]$
satisfies
\[
  V_{n+1} \approx V_n + \frac{1}{n}\Bigl[(aA + \tfrac{1}{2}I) V_n
    + V_n (aA + \tfrac{1}{2}I)^T + a^2 \Sigma\Bigr],
\]
and the fixed point $V_{n+1} = V_n = V$ gives~\eqref{eq:lyapunov}.
\end{proof}

\begin{remark}[Optimal step size]
\label{rem:optimal-step}
The asymptotic covariance $V$ depends on $a$ through the Lyapunov
equation~\eqref{eq:lyapunov}. When $d = 1$ and $A = -\lambda$ with $\lambda > 0$,
the equation becomes $(-2a\lambda + 1)V = a^2 \sigma^2$, giving
$V = a^2 \sigma^2 / (2a\lambda - 1)$. This is minimized when $a = 1/\lambda$,
yielding $V_{\mathrm{opt}} = \sigma^2 / \lambda^2$. In higher dimensions, the
optimal choice requires $a = -(A^T)^{-1}/2$, which depends on unknown quantities.
\end{remark}

\subsection{Polyak--Ruppert averaging}

The optimal step size depends on the Jacobian $A$ at $x^*$, which is typically
unknown. Polyak (1990) and Ruppert (1988) independently proposed a simple device
that achieves the optimal rate \emph{without} knowing $A$: average the iterates.

\begin{theorem}[Polyak--Ruppert averaging]
\label{thm:polyak-ruppert}
Let $\{x_n\}$ be the SA iterates with step size $a_n = a/n^\alpha$ for
$\alpha \in (1/2, 1)$, and define the averaged iterate
\[
  \bar{x}_n = \frac{1}{n} \sum_{k=1}^{n} x_k.
\]
Under assumptions \ref{B1}--\ref{B5} and \ref{D1}--\ref{D4} (with the modified step
size), we have
\[
  \sqrt{n}(\bar{x}_n - x^*) \cdist \mathcal{N}(0, A^{-1} \Sigma (A^{-T})).
\]
This is the optimal asymptotic covariance (in the sense of the Cramer--Rao bound for
the root-finding problem).
\end{theorem}

\begin{proof}
Since $x_n \to x^*$ and $h(x^*) = 0$, we have
\[
  x_{n+1} - x_n = a_n [h(x_n) + M_{n+1}] = a_n [A(x_n - x^*) + r_n + M_{n+1}],
\]
where $r_n = o(\|x_n - x^*\|)$. Summing from $k = 1$ to $n$:
\[
  x_{n+1} - x_1 = A \sum_{k=1}^{n} a_k (x_k - x^*)
    + \sum_{k=1}^{n} a_k r_k + \sum_{k=1}^{n} a_k M_{k+1}.
\]
Rearranging:
\[
  A \sum_{k=1}^{n} a_k (x_k - x^*) = (x_{n+1} - x_1) - \sum_{k=1}^{n} a_k r_k
    - \sum_{k=1}^{n} a_k M_{k+1}.
\]

Define $S_n = \sum_{k=1}^n a_k$. Since $a_k = a/k^\alpha$ with $\alpha < 1$,
$S_n \sim \frac{a}{1-\alpha} n^{1-\alpha}$.

Now, the key observation: $\bar{x}_n - x^* = \frac{1}{n}\sum_{k=1}^n (x_k - x^*)$.
We relate this to the weighted sum above. A careful Abel summation argument shows
that
\[
  \sqrt{n}(\bar{x}_n - x^*)
    = -A^{-1} \cdot \frac{1}{\sqrt{n}} \sum_{k=1}^{n} M_{k+1} + o_P(1).
\]
The dominant term is the sample average of the martingale differences, and the
remainder terms ($x_{n+1} - x_1$ is $o(\sqrt{n})$, and $\sum a_k r_k$ is negligible)
vanish after normalization.

By the martingale CLT, $\frac{1}{\sqrt{n}} \sum_{k=1}^n M_{k+1} \cdist \mathcal{N}(0, \Sigma)$. Therefore
\[
  \sqrt{n}(\bar{x}_n - x^*) \cdist \mathcal{N}(0, A^{-1} \Sigma A^{-T}). \qedhere
\]
\end{proof}

\begin{keyresult}[Polyak--Ruppert Averaging]
By averaging the iterates of a slowly decaying SA scheme ($a_n = O(n^{-\alpha})$,
$\alpha \in (1/2,1)$), the averaged iterate $\bar{x}_n$ achieves the optimal
convergence rate $O(1/\sqrt{n})$ with the minimal asymptotic covariance
$A^{-1}\Sigma A^{-T}$, without knowledge of the Jacobian $A = Dh(x^*)$.
\end{keyresult}

\begin{rlconnection}[Iterate Averaging in RL]
Polyak--Ruppert averaging suggests that RL algorithms should average their iterates
rather than use the last iterate. In practice, \emph{exponential moving averages}
(EMA) are used as a practical approximation: the target network in DQN, the
exponentially averaged value function in some actor-critic methods, and the
``Polyak averaging'' of policy parameters in DDPG/TD3 all implement this principle.
The theoretical guarantee is that averaging provably improves the convergence rate.
\end{rlconnection}


%% ============================================================
%% SECTION 5: TWO-TIMESCALE STOCHASTIC APPROXIMATION
%% ============================================================

\section{Two-Timescale Stochastic Approximation}
\label{sec:two-timescale}

\subsection{Motivation}

Many RL algorithms involve two coupled iterative updates running simultaneously but
at different rates. The canonical example is \emph{actor-critic}: the critic updates
the value function (fast timescale) while the actor updates the policy (slow
timescale). The theory of two-timescale SA, developed by Borkar (1997), provides the
convergence framework for such algorithms.

\subsection{Formulation}

\begin{definition}[Two-timescale SA]
\label{def:two-timescale}
Consider the coupled iterations
\begin{align}
  x_{n+1} &= x_n + a_n [f(x_n, y_n) + M_{n+1}^{(1)}], \label{eq:slow} \\
  y_{n+1} &= y_n + b_n [g(x_n, y_n) + M_{n+1}^{(2)}], \label{eq:fast}
\end{align}
where $x_n \in \R^{d_1}$, $y_n \in \R^{d_2}$, and the step-size sequences satisfy:
\begin{enumerate}[label=(\roman*)]
\item $\sum a_n = \sum b_n = \infty$ and $\sum a_n^2 < \infty$, $\sum b_n^2 < \infty$,
\item $a_n / b_n \to 0$ as $n \to \infty$ \quad (the ``timescale separation'').
\end{enumerate}
The typical choice is $a_n = 1/n$ and $b_n = 1/n^\beta$ for $\beta \in (1/2, 1)$,
or $a_n = 1/n$ and $b_n = 1/n^{2/3}$.
\end{definition}

\begin{remark}[Timescale separation]
\label{rem:timescale}
The condition $a_n / b_n \to 0$ means that the $y$-iteration uses \emph{larger} step
sizes than the $x$-iteration, so $y$ moves faster. From the perspective of the
$y$-update, $x$ changes so slowly that it appears quasi-static. From the perspective
of the $x$-update, $y$ equilibrates so quickly that it appears to have already
converged.
\end{remark}

\subsection{Assumptions}

\begin{definition}[Two-timescale assumptions]
\label{def:two-ts-assumptions}
\begin{enumerate}[label=(E\arabic*)]
\item \label{E1} \textbf{Step sizes:} As in Definition~\ref{def:two-timescale}.
\item \label{E2} \textbf{Lipschitz drifts:} $f : \R^{d_1} \times \R^{d_2} \to \R^{d_1}$
  and $g : \R^{d_1} \times \R^{d_2} \to \R^{d_2}$ are globally Lipschitz.
\item \label{E3} \textbf{Martingale difference noise:}
  $\E[M_{n+1}^{(i)} \mid \mathcal{F}_n] = 0$ and
  $\E[\|M_{n+1}^{(i)}\|^2 \mid \mathcal{F}_n] \leq K(1 + \|x_n\|^2 + \|y_n\|^2)$
  for $i = 1, 2$.
\item \label{E4} \textbf{Stability:} $\sup_n (\|x_n\| + \|y_n\|) < \infty$ a.s.
\item \label{E5} \textbf{Fast timescale equilibrium:} For each fixed $x \in \R^{d_1}$,
  the ODE $\dot{y} = g(x, y)$ has a globally asymptotically stable equilibrium
  $\lambda(x)$, and $\lambda : \R^{d_1} \to \R^{d_2}$ is Lipschitz.
\item \label{E6} \textbf{Slow timescale stability:} The ODE
  $\dot{x} = f(x, \lambda(x))$ has a globally asymptotically stable equilibrium $x^*$.
\end{enumerate}
\end{definition}

\subsection{Convergence theorem}

\begin{theorem}[Two-timescale convergence (Borkar)]
\label{thm:two-timescale}
Under assumptions \ref{E1}--\ref{E6}, the two-timescale SA iterates satisfy
\[
  x_n \cas x^*, \quad y_n \cas \lambda(x^*) \coloneqq y^*.
\]
\end{theorem}

\begin{proof}
The proof is structured in two phases, analyzing each timescale separately.

\medskip
\noindent\textbf{Phase 1: Fast timescale analysis (convergence of $y_n$).}

Consider the $y$-iteration~\eqref{eq:fast}. Since $a_n / b_n \to 0$, the
$x$-iterates change negligibly on the timescale of $b_n$. Formally, on the time
axis defined by $\tau_n = \sum_{k=0}^{n-1} b_k$, the $x$-update has effective
step size $a_n$, which is asymptotically negligible compared to $b_n$.

To make this precise, fix any $T > 0$ and consider the window of indices $[n, m_n]$
where $m_n$ is the largest index such that $\sum_{k=n}^{m_n - 1} b_k \leq T$. On
this window:
\[
  \|x_{m_n} - x_n\| \leq \sum_{k=n}^{m_n-1} a_k \|f(x_k, y_k) + M_{k+1}^{(1)}\|
    \leq C \sum_{k=n}^{m_n-1} a_k,
\]
where $C$ is a bound on $\|f\| + \|M^{(1)}\|$ (which exists by \ref{E4} and
\ref{E2}). Now
\[
  \sum_{k=n}^{m_n-1} a_k = \sum_{k=n}^{m_n-1} \frac{a_k}{b_k} \cdot b_k
    \leq \sup_{k \geq n} \frac{a_k}{b_k} \sum_{k=n}^{m_n-1} b_k
    \leq T \cdot \sup_{k \geq n} \frac{a_k}{b_k} \to 0.
\]
So on the fast timescale, $x$ is indeed quasi-static.

Therefore, the $y$-iteration behaves asymptotically like
\[
  y_{n+1} = y_n + b_n [g(\bar{x}, y_n) + M_{n+1}^{(2)}]
\]
for a ``frozen'' value $\bar{x}$ (which equals the current value of the slowly
drifting $x_n$). By the ODE method (Theorem~\ref{thm:ode-convergence}), this
converges to the equilibrium $\lambda(\bar{x})$ of $\dot{y} = g(\bar{x}, y)$.

More precisely, the $y$-iterates asymptotically track the ODE
$\dot{y} = g(x(t), y)$ where $x(t)$ is the slowly varying $x$-component. Since
$x$ changes slowly, the $y$-component ``quasi-statically'' tracks $\lambda(x_n)$:
for any $\varepsilon > 0$,
\begin{equation}
\label{eq:y-tracking}
  \|y_n - \lambda(x_n)\| < \varepsilon \quad \text{for all large } n, \text{ a.s.}
\end{equation}

\medskip
\noindent\textbf{Phase 2: Slow timescale analysis (convergence of $x_n$).}

Now consider the $x$-iteration~\eqref{eq:slow}. By~\eqref{eq:y-tracking},
$y_n \approx \lambda(x_n)$ for large $n$, so
\[
  f(x_n, y_n) = f(x_n, \lambda(x_n)) + [f(x_n, y_n) - f(x_n, \lambda(x_n))].
\]
The second term is bounded by $L_f \|y_n - \lambda(x_n)\| \to 0$ a.s.\ by
the Lipschitz property of $f$.

Therefore, the $x$-iteration asymptotically behaves like
\[
  x_{n+1} = x_n + a_n [f(x_n, \lambda(x_n)) + \tilde{M}_{n+1}],
\]
where $\tilde{M}_{n+1}$ absorbs the original noise $M_{n+1}^{(1)}$ and the
approximation error from replacing $y_n$ by $\lambda(x_n)$. The associated ODE is
\[
  \dot{x} = f(x, \lambda(x)),
\]
which has $x^*$ as a globally asymptotically stable equilibrium by \ref{E6}.
Applying the ODE method again, $x_n \cas x^*$.

Since $\lambda$ is continuous and $x_n \to x^*$, we conclude
$y_n \to \lambda(x^*) = y^*$ a.s.
\end{proof}

\begin{keyresult}[Two-Timescale Convergence Principle]
In a two-timescale SA system where $a_n/b_n \to 0$:
\begin{enumerate}[leftmargin=2em]
\item The fast iterate $y_n$ ``tracks'' the equilibrium map $\lambda(x_n)$ where
  $g(x, \lambda(x)) = 0$.
\item The slow iterate $x_n$ converges as if $y_n$ were always at equilibrium,
  driven by the averaged dynamics $\dot{x} = f(x, \lambda(x))$.
\item The overall system converges to $(x^*, y^*) = (x^*, \lambda(x^*))$.
\end{enumerate}
\end{keyresult}

\begin{rlconnection}[Actor-Critic as Two-Timescale SA]
In actor-critic methods, the critic updates the value function estimate $w_n$ on
the fast timescale and the actor updates the policy parameters $\theta_n$ on the
slow timescale:
\begin{align*}
  w_{n+1} &= w_n + b_n [\delta_n \phi(s_n) - \text{(critic update)}], \\
  \theta_{n+1} &= \theta_n + a_n [\psi_n - \text{(actor update)}],
\end{align*}
with $a_n / b_n \to 0$. Here $(x_n, y_n) = (\theta_n, w_n)$, $f$ is the policy
gradient direction, and $g$ is the TD error for the current policy. The fast
timescale equilibrium $\lambda(\theta)$ is the true value function
$V^{\pi_\theta}$ for policy $\pi_\theta$. The slow timescale ODE is the policy
gradient flow $\dot{\theta} = \nabla J(\theta)$. Theorem~\ref{thm:two-timescale}
then guarantees convergence to a local optimum of $J(\theta)$ (or a global optimum
if $J$ is concave in $\theta$).
\end{rlconnection}

\subsection{Lock-in probability}

When the ODE has multiple locally stable equilibria (as is typical for nonlinear
systems), the SA iteration converges to one of them. Which one depends on the initial
condition and the noise history.

\begin{proposition}[Lock-in to attractors]
\label{prop:lock-in}
Suppose the ODE $\dot{x} = h(x)$ has finitely many equilibria, each of which is
either asymptotically stable or unstable. Under the ODE method assumptions
\ref{B1}--\ref{B4}, the SA iterates converge a.s.\ to one of the stable equilibria.
The probability of converging to an unstable equilibrium is zero.
\end{proposition}

\begin{proof}
By the ODE method analysis (Theorem~\ref{thm:ode-convergence}, Step 2), the
limit set of $\{x_n\}$ is a compact connected invariant set of the ODE flow.

If this limit set contained an unstable equilibrium $\bar{x}$, then since $\bar{x}$
is unstable, there exists an open neighborhood $U$ of $\bar{x}$ and a direction
along which ODE trajectories leave $U$ exponentially fast. However, the noise in
SA prevents the iterates from remaining exactly on the unstable manifold.

Formally, by the ``avoidance of unstable equilibria'' result (Pemantle, 1990;
Brandiere, 1998): under the martingale noise assumption with nondegenerate
conditional covariance (i.e., $\E[M_{n+1}M_{n+1}^T \mid \mathcal{F}_n]$ is
positive definite), the set of initial conditions $x_0$ for which $x_n$ converges
to an unstable equilibrium has Lebesgue measure zero, and $\Prob(x_n \to \bar{x}) = 0$
for any fixed $x_0$.

The key idea: near an unstable equilibrium, the noise keeps ``kicking'' the iterate
off the unstable manifold, and once it leaves a small neighborhood, the ODE dynamics
carry it toward a stable equilibrium.
\end{proof}

\begin{warning}[Multiple Stable Equilibria in RL]
When the mean-field ODE has multiple stable equilibria, the RL algorithm may converge
to a suboptimal one depending on the initial conditions and exploration noise. This is
a fundamental challenge in policy gradient methods, where the objective $J(\theta)$ is
generally non-concave and has multiple local optima. Two-timescale SA (actor-critic)
inherits this issue. Entropy regularization (as in SAC) and exploration bonuses are
practical techniques to mitigate convergence to poor local optima.
\end{warning}


%% ============================================================
%% SECTION 6: LINEAR STOCHASTIC APPROXIMATION
%% ============================================================

\section{Linear Stochastic Approximation}
\label{sec:linear-sa}

The most important special case for RL is \emph{linear} SA, where the drift function
is affine.

\subsection{Setup}

\begin{definition}[Linear SA]
\label{def:linear-sa}
A \emph{linear stochastic approximation} iteration is
\begin{equation}
\label{eq:linear-sa}
  x_{n+1} = x_n + a_n [A x_n + b + M_{n+1}],
\end{equation}
where $A \in \R^{d \times d}$, $b \in \R^d$, and $h(x) = Ax + b$.
\end{definition}

The associated ODE is $\dot{x} = Ax + b$, a linear ODE with equilibrium
$x^* = -A^{-1}b$ (when $A$ is invertible).

\subsection{Convergence conditions}

\begin{theorem}[Convergence of linear SA]
\label{thm:linear-sa}
Suppose $A$ is Hurwitz (all eigenvalues have strictly negative real parts), the step
sizes satisfy \ref{A1}, the noise satisfies \ref{B3}, and the iterates are bounded
a.s. Then $x_n \cas x^* = -A^{-1}b$.
\end{theorem}

\begin{proof}
The ODE $\dot{x} = Ax + b$ has $x^* = -A^{-1}b$ as its unique equilibrium. Since
$A$ is Hurwitz, the function $V(x) = (x - x^*)^T P (x - x^*)$ is a Lyapunov function
for the ODE, where $P$ is the unique positive definite solution of the Lyapunov matrix
equation $A^T P + P A = -I$. (Such $P$ exists precisely because $A$ is Hurwitz.)

Indeed,
\[
  \dot{V} = 2(x - x^*)^T P (Ax + b)
    = 2(x - x^*)^T P A (x - x^*)
    = (x - x^*)^T (A^T P + PA)(x - x^*) = -\|x - x^*\|^2,
\]
so $V$ is strictly decreasing along nonconstant trajectories. This confirms that
$x^*$ is globally asymptotically stable.

The result then follows directly from the ODE method
(Theorem~\ref{thm:ode-convergence}).
\end{proof}

\begin{remark}[Hurwitz condition for linear SA]
\label{rem:hurwitz-intuition}
The Hurwitz condition on $A$ means that the ``drift'' $h(x) = Ax + b$ always pushes
the iterate toward the equilibrium $x^* = -A^{-1}b$. If any eigenvalue of $A$ had
a positive real part, the ODE would be unstable in the corresponding direction, and
the SA iteration would diverge. For the borderline case (eigenvalues on the imaginary
axis), convergence depends on higher-order terms and the noise structure.
\end{remark}

\subsection{Stability via Borkar--Meyn for linear SA}

\begin{corollary}[Automatic stability for linear SA]
\label{cor:linear-stability}
For linear SA~\eqref{eq:linear-sa} with $A$ Hurwitz, the boundedness assumption
$\sup_n \|x_n\| < \infty$ a.s.\ holds automatically (without needing to assume it).
\end{corollary}

\begin{proof}
Apply the Borkar--Meyn theorem (Theorem~\ref{thm:borkar-meyn}). For linear $h(x) = Ax + b$,
the ``function at infinity'' is $h_\infty(x) = \lim_{r \to \infty} (A(rx) + b)/r = Ax$.
Since $A$ is Hurwitz, the ODE $\dot{x} = Ax$ has the origin as its globally
asymptotically stable equilibrium. Thus \ref{C2} is satisfied, and
Theorem~\ref{thm:borkar-meyn} gives $\sup_n \|x_n\| < \infty$ a.s.
\end{proof}

\begin{theorem}[Complete convergence for linear SA]
\label{thm:linear-sa-complete}
For the linear SA iteration~\eqref{eq:linear-sa} with $A$ Hurwitz, step sizes
satisfying \ref{A1}, and noise satisfying \ref{B3}, we have $x_n \cas -A^{-1}b$
without any additional stability assumptions.
\end{theorem}

\begin{proof}
Combine Corollary~\ref{cor:linear-stability} (which gives boundedness) with
Theorem~\ref{thm:linear-sa} (which gives convergence given boundedness).
\end{proof}

\begin{rlconnection}[TD(0) with Linear Function Approximation]
In TD(0) with linear function approximation $V(s) = \phi(s)^T w$, the parameter
vector $w$ is updated by
\[
  w_{n+1} = w_n + \alpha_n \bigl[R_{n+1} + \gamma \phi(s_{n+1})^T w_n
    - \phi(s_n)^T w_n\bigr] \phi(s_n).
\]
This is linear SA with
\[
  A = \E_\pi[\phi(s)(\gamma \phi(s') - \phi(s))^T], \quad
  b = \E_\pi[R \phi(s)],
\]
where the expectation is over the stationary distribution of the Markov chain
under policy $\pi$. The key question is: \emph{is $A$ Hurwitz?}

Under on-policy sampling (the states are visited according to the stationary
distribution $d^\pi$ of $\pi$), it can be shown that $A$ is negative definite
(hence Hurwitz). Specifically, $A = \Phi^T D(\gamma P^\pi - I)\Phi$ where $D$ is
the diagonal matrix of stationary probabilities, and $\gamma P^\pi - I$ is negative
definite on the range of $\Phi$ (since $\gamma < 1$ and $P^\pi$ has spectral
radius 1).

By Theorem~\ref{thm:linear-sa-complete}, TD(0) with linear function approximation
converges a.s.\ under on-policy sampling. This is the celebrated result of Tsitsiklis
and Van Roy (1997).

\textbf{The Deadly Triad:} Under \emph{off-policy} sampling with function
approximation, $A$ may fail to be Hurwitz, and TD(0) can \emph{diverge}. This is
the theoretical explanation for the deadly triad (function approximation +
bootstrapping + off-policy).
\end{rlconnection}


%% ============================================================
%% SECTION 7: APPLICATIONS TO REINFORCEMENT LEARNING
%% ============================================================

\section{Applications to Reinforcement Learning}
\label{sec:rl-applications}

We now systematically map the general SA theory to specific RL algorithms, verifying
the assumptions of the convergence theorems in each case.

\subsection{TD(0) as stochastic approximation}

\begin{rlconnection}[TD(0) -- Complete SA Analysis]
\textbf{Setting.} Consider a finite MDP with states $\mathcal{S} = \{1, \ldots, S\}$,
policy $\pi$, discount factor $\gamma \in [0, 1)$, transition matrix $P^\pi$, and
reward function $r^\pi(s) = \E[R \mid S = s, \pi]$.

\textbf{The update.} TD(0) updates the value function estimate $V$ by
\[
  V_{n+1}(s_n) = V_n(s_n) + \alpha_n(s_n) \bigl[\underbrace{R_{n+1} + \gamma V_n(s_{n+1}) - V_n(s_n)}_{\delta_n \text{ (TD error)}}\bigr].
\]

\textbf{SA form.} Viewing $V \in \R^S$ as the iterate, define
\[
  h(V) = r^\pi + \gamma P^\pi V - V = (\gamma P^\pi - I)V + r^\pi.
\]
Then $h(V^\pi) = r^\pi + \gamma P^\pi V^\pi - V^\pi = 0$ by the Bellman equation.
The noisy observation at step $n$ (for state $s_n$) is
$H_n = (R_{n+1} + \gamma V_n(s_{n+1}) - V_n(s_n)) e_{s_n}$, where $e_s$ is the
$s$-th standard basis vector.

\textbf{Mean field.} $\E[H_n \mid V_n, s_n] = (r^\pi(s_n) + \gamma (P^\pi V_n)(s_n) - V_n(s_n)) e_{s_n}$, which matches $h(V_n)$ projected onto the $s_n$-coordinate.

\textbf{Associated ODE.} $\dot{V} = (\gamma P^\pi - I)V + r^\pi$. The matrix
$A = \gamma P^\pi - I$ has eigenvalues $\gamma \lambda_i - 1$ where $\lambda_i$ are
eigenvalues of $P^\pi$. Since $|\lambda_i| \leq 1$ (stochastic matrix) and
$\gamma < 1$, all eigenvalues of $A$ satisfy
$\mathrm{Re}(\gamma \lambda_i - 1) \leq \gamma - 1 < 0$. Hence $A$ is Hurwitz.

\textbf{Conclusion.} By Theorem~\ref{thm:linear-sa-complete} (or
Theorem~\ref{thm:ode-convergence}), TD(0) converges a.s.\ to
$V^* = -A^{-1}r^\pi = (I - \gamma P^\pi)^{-1}r^\pi = V^\pi$, the true value
function under $\pi$.

\textbf{Step sizes.} Convergence requires the Robbins--Monro conditions to hold
\emph{per state}: $\sum_n \alpha_n(s) \indicator(s_n = s) = \infty$ and
$\sum_n \alpha_n(s)^2 \indicator(s_n = s) < \infty$ a.s.\ for each $s$.
\end{rlconnection}

\subsection{Q-learning as asynchronous SA}

\begin{rlconnection}[Q-learning -- Asynchronous SA]
\textbf{The update.} Q-learning updates a single state-action entry per step:
\[
  Q_{n+1}(s_n, a_n) = Q_n(s_n, a_n) + \alpha_n(s_n, a_n)
    \bigl[R_{n+1} + \gamma \max_{a'} Q_n(s_{n+1}, a') - Q_n(s_n, a_n)\bigr].
\]

\textbf{SA form.} The mean field $h$ is now \emph{nonlinear} due to the $\max$ operator:
\[
  h(Q)(s,a) = r(s,a) + \gamma \sum_{s'} P(s' \mid s, a) \max_{a'} Q(s', a') - Q(s,a).
\]

\textbf{Associated ODE.} $\dot{Q} = h(Q)$, which is the Bellman optimality operator
$\mathcal{T}^* Q - Q$ in continuous time. Since $\mathcal{T}^*$ is a
$\gamma$-contraction in the sup-norm, the ODE has $Q^*$ (the optimal Q-function) as
its unique globally asymptotically stable equilibrium. To see this, note that the
Lyapunov function $V(Q) = \|Q - Q^*\|_\infty$ satisfies
\[
  \frac{d}{dt}\|Q(t) - Q^*\|_\infty \leq -(1 - \gamma)\|Q(t) - Q^*\|_\infty.
\]

\textbf{Complication: asynchronous updates.} Q-learning updates only one $(s,a)$ pair
per step, making it an \emph{asynchronous} SA scheme. The theory extends via results
of Borkar (1998) and Tsitsiklis (1994): if each $(s,a)$ pair is updated infinitely
often (ensured by an exploring policy), the asynchronous iteration converges to the
same limit as the synchronous ODE.

\textbf{Conclusion.} Under the Robbins--Monro step-size conditions (per state-action
pair) and sufficient exploration, Q-learning converges a.s.\ to $Q^*$.
\end{rlconnection}

\subsection{Policy gradient (REINFORCE) as SA}

\begin{rlconnection}[REINFORCE as SA on the Gradient Ascent ODE]
\textbf{The update.} REINFORCE updates the policy parameters $\theta$ by
\[
  \theta_{n+1} = \theta_n + \alpha_n \widehat{\nabla J(\theta_n)},
\]
where $\widehat{\nabla J(\theta_n)}$ is an unbiased estimate of the policy gradient
$\nabla_\theta J(\theta)$.

\textbf{SA form.} This is Robbins--Monro with $h(\theta) = \nabla J(\theta)$ and
noise $M_{n+1} = \widehat{\nabla J(\theta_n)} - \nabla J(\theta_n)$.

\textbf{Associated ODE.} $\dot{\theta} = \nabla J(\theta)$, which is gradient
ascent on the expected return. Any local maximum $\theta^*$ of $J$ with
$\nabla J(\theta^*) = 0$ and negative definite Hessian is a stable equilibrium.

\textbf{Complication.} The ODE $\dot{\theta} = \nabla J(\theta)$ typically has
multiple stable equilibria (local maxima of $J$). By Proposition~\ref{prop:lock-in},
the SA iterates converge to one of them, but global optimality is not guaranteed.

\textbf{Convergence rate.} With proper step sizes, the CLT
(Theorem~\ref{thm:sa-clt}) applies near each local maximum, giving
$\sqrt{n}(\theta_n - \theta^*) \cdist \mathcal{N}(0, V)$ for an appropriate
covariance $V$. The high variance of REINFORCE estimators means $\Sigma$ (and hence
$V$) is large, motivating the use of baselines.
\end{rlconnection}

\subsection{Actor-critic as two-timescale SA}

\begin{rlconnection}[Actor-Critic -- Two-Timescale Analysis]
\textbf{Setup.} Actor-critic maintains two parameter vectors:
\begin{itemize}[leftmargin=2em]
\item Critic parameters $w \in \R^{d_w}$ (fast timescale, step size $b_n$)
\item Actor parameters $\theta \in \R^{d_\theta}$ (slow timescale, step size $a_n$)
\end{itemize}
with $a_n / b_n \to 0$.

\textbf{Coupled iterations.}
\begin{align*}
  w_{n+1} &= w_n + b_n \bigl[\delta_n \phi(s_n) \bigr], \quad
    \delta_n = R_{n+1} + \gamma \phi(s_{n+1})^T w_n - \phi(s_n)^T w_n, \\
  \theta_{n+1} &= \theta_n + a_n \bigl[\delta_n \nabla_\theta \log \pi_\theta(a_n \mid s_n)\bigr].
\end{align*}

\textbf{Mapping to two-timescale SA.}
\begin{itemize}[leftmargin=2em]
\item $y_n = w_n$, $x_n = \theta_n$.
\item $g(\theta, w) = \E_{\pi_\theta}[\delta^\theta_w \phi(s)]$: the critic drift,
  which is exactly the linear SA drift for TD(0) under policy $\pi_\theta$.
\item $f(\theta, w) = \E_{\pi_\theta}[\delta^\theta_w \nabla_\theta \log \pi_\theta(a \mid s)]$:
  the actor drift, which approximates $\nabla J(\theta)$ when $w$ encodes
  $V^{\pi_\theta}$.
\end{itemize}

\textbf{Verifying assumptions.}
\begin{enumerate}[leftmargin=2em]
\item \ref{E5} (fast equilibrium): For fixed $\theta$, the critic ODE
  $\dot{w} = g(\theta, w)$ is linear SA for TD under $\pi_\theta$. Its equilibrium
  $\lambda(\theta) = w^*(\theta)$ is the TD fixed point (projection of $V^{\pi_\theta}$
  onto the linear space).
\item \ref{E6} (slow stability): With $w = \lambda(\theta)$, the actor ODE becomes
  $\dot{\theta} = \nabla J(\theta)$ (approximately), which ascends the performance
  surface.
\end{enumerate}

\textbf{Conclusion.} By Theorem~\ref{thm:two-timescale}, the critic converges to the
TD fixed point $w^*(\theta^*)$ and the actor converges to a (local) optimum $\theta^*$
of $J(\theta)$.
\end{rlconnection}


%% ============================================================
%% SECTION 8: EXERCISES
%% ============================================================

\section{Exercises}
\label{sec:exercises}

\begin{exercise}[Step-size conditions]
\label{ex:step-sizes}
For each of the following step-size sequences, determine whether both Robbins--Monro
conditions ($\sum a_n = \infty$, $\sum a_n^2 < \infty$) are satisfied:
\begin{enumerate}[label=(\alph*)]
\item $a_n = 1/n$
\item $a_n = 1/n^2$
\item $a_n = 1/\sqrt{n}$
\item $a_n = c/(n + d)$ for constants $c, d > 0$
\item $a_n = 1/(n \log n)$ for $n \geq 2$
\item $a_n = 1/n^{3/4}$
\end{enumerate}
For each case that fails, explain which condition fails and the practical
consequence for an SA algorithm using that step size.
\end{exercise}

\begin{exercise}[Scalar Robbins--Monro]
\label{ex:scalar-rm}
Consider the scalar SA iteration $x_{n+1} = x_n + a_n(c - x_n + \xi_{n+1})$ where
$c \in \R$ and $\{\xi_n\}$ are i.i.d.\ with $\E[\xi_n] = 0$ and
$\E[\xi_n^2] = \sigma^2 < \infty$.
\begin{enumerate}[label=(\alph*)]
\item Identify $h(x)$ and $x^*$.
\item Verify the monotonicity condition \ref{A4}.
\item Show that $A = Dh(x^*) = -1$ (the Jacobian at the equilibrium).
\item With $a_n = a/n$, compute the asymptotic covariance
  $V = \lim_{n \to \infty} n \E[(x_n - x^*)^2]$ using the Lyapunov
  equation~\eqref{eq:lyapunov}.
\item Find the optimal $a$ that minimizes $V$, and compute $V_{\mathrm{opt}}$.
\end{enumerate}
\end{exercise}

\begin{exercise}[ODE for Q-learning]
\label{ex:ode-qlearning}
Consider a two-state, two-action MDP with $\mathcal{S} = \{1, 2\}$,
$\mathcal{A} = \{L, R\}$, $\gamma = 0.9$, deterministic transitions
$P(2 \mid 1, L) = 1$, $P(1 \mid 1, R) = 1$, $P(1 \mid 2, L) = 1$,
$P(2 \mid 2, R) = 1$, and rewards $r(1, L) = 1$, $r(1, R) = 0$,
$r(2, L) = 0$, $r(2, R) = 2$.
\begin{enumerate}[label=(\alph*)]
\item Compute $Q^*(s, a)$ for all $(s, a)$ pairs by solving the Bellman optimality
  equation.
\item Write out the associated ODE $\dot{Q} = h(Q)$ explicitly (as a 4-dimensional
  system).
\item Verify that $Q^*$ is the unique equilibrium.
\item \emph{(Bonus)} Simulate the ODE numerically from several initial conditions
  and verify convergence to $Q^*$.
\end{enumerate}
\end{exercise}

\begin{exercise}[Lyapunov function for contraction]
\label{ex:lyapunov-contraction}
Let $T : \R^d \to \R^d$ be a $\gamma$-contraction in the sup-norm with fixed point
$x^*$: $\|Tx - Tx'\|_\infty \leq \gamma \|x - x'\|_\infty$. Define the ODE
$\dot{x} = Tx - x$.
\begin{enumerate}[label=(\alph*)]
\item Show that $V(x) = \|x - x^*\|_\infty$ is a Lyapunov function satisfying
  $\dot{V} \leq -(1-\gamma) V$.
\item Conclude that $x^*$ is globally asymptotically stable with exponential
  convergence rate $1 - \gamma$.
\item Apply this to the Bellman optimality operator $T = \mathcal{T}^*$ to verify
  the ODE analysis for Q-learning.
\end{enumerate}
\end{exercise}

\begin{exercise}[Two-timescale separation]
\label{ex:two-timescale}
Consider the two-timescale system in $\R^2$:
\begin{align*}
  x_{n+1} &= x_n + a_n(-x_n + y_n + M_{n+1}^{(1)}), \\
  y_{n+1} &= y_n + b_n(-2y_n + x_n + M_{n+1}^{(2)}),
\end{align*}
where $a_n = 1/n$, $b_n = 1/n^{2/3}$, and the noises are i.i.d.\ with zero mean
and unit variance.
\begin{enumerate}[label=(\alph*)]
\item Verify the timescale separation condition $a_n/b_n \to 0$.
\item For the fast timescale: fix $x$ and find the equilibrium $\lambda(x)$
  of $\dot{y} = -2y + x$.
\item For the slow timescale: substitute $y = \lambda(x)$ into the $x$-ODE and
  find the equilibrium $x^*$.
\item Find $(x^*, y^*)$ and verify that the coupled ODE system has this as its
  unique equilibrium.
\item Verify the Hurwitz conditions for both the fast and slow ODEs.
\end{enumerate}
\end{exercise}

\begin{exercise}[Polyak--Ruppert for mean estimation]
\label{ex:polyak-ruppert}
Let $X_1, X_2, \ldots$ be i.i.d.\ with $\E[X_i] = \mu$ and $\Var(X_i) = \sigma^2$.
Consider the SA iteration $\theta_{n+1} = \theta_n + a_n(X_{n+1} - \theta_n)$ with
$a_n = 1/n^{2/3}$.
\begin{enumerate}[label=(\alph*)]
\item Show this is SA with $h(\theta) = \mu - \theta$ and identify $A$ and $\Sigma$.
\item What is the asymptotic distribution of $\sqrt{n}(\theta_n - \mu)$?
  (Use the CLT for SA, Theorem~\ref{thm:sa-clt}, noting the step size is $a/n^\alpha$
  with $\alpha = 2/3$.)
\item Define $\bar{\theta}_n = \frac{1}{n}\sum_{k=1}^n \theta_k$. What is the
  asymptotic distribution of $\sqrt{n}(\bar{\theta}_n - \mu)$?
\item Compare the asymptotic variances from (b) and (c). Which is better?
\item Compare both to the sample mean $\bar{X}_n = \frac{1}{n}\sum_{k=1}^n X_k$.
  What is the asymptotic variance of $\sqrt{n}(\bar{X}_n - \mu)$?
\end{enumerate}
\end{exercise}

\begin{exercise}[Hurwitz verification for TD with linear FA]
\label{ex:hurwitz-td}
Consider a three-state MDP with $\mathcal{S} = \{1, 2, 3\}$, policy $\pi$ inducing
transition matrix
\[
  P^\pi = \begin{pmatrix} 0.2 & 0.5 & 0.3 \\ 0.1 & 0.6 & 0.3 \\ 0.4 & 0.3 & 0.3 \end{pmatrix},
\]
discount $\gamma = 0.95$, and stationary distribution $d^\pi = (0.2, 0.45, 0.35)$.
Let the feature matrix be $\Phi = \begin{pmatrix} 1 & 0 \\ 0 & 1 \\ 1 & 1 \end{pmatrix}$.
\begin{enumerate}[label=(\alph*)]
\item Compute the matrix $A = \Phi^T D(\gamma P^\pi - I)\Phi$ where
  $D = \mathrm{diag}(d^\pi)$.
\item Verify that $A$ is negative definite (hence Hurwitz).
\item Conclude that TD(0) converges with this function approximation scheme.
\end{enumerate}
\end{exercise}

\begin{exercise}[Convergence failure: the deadly triad]
\label{ex:deadly-triad}
Construct a simple two-state MDP with linear function approximation and off-policy
sampling such that the matrix $A$ in the linear SA formulation of TD(0) has an
eigenvalue with positive real part, demonstrating divergence. Specifically:
\begin{enumerate}[label=(\alph*)]
\item Take $\mathcal{S} = \{1, 2\}$, $\gamma = 0.99$, and a behavior policy that
  visits state 1 with probability 0.01 and state 2 with probability 0.99.
\item Let $P^\pi$ be the transition matrix under the target policy $\pi$, with
  $P^\pi(1 \mid 1) = 0.9$, $P^\pi(2 \mid 1) = 0.1$, $P^\pi(1 \mid 2) = 0.1$,
  $P^\pi(2 \mid 2) = 0.9$.
\item Use the feature representation $\phi(1) = 1$, $\phi(2) = 2$ (scalar features).
\item Compute $A = \E_\mu[\phi(s)(\gamma\phi(s') - \phi(s))]$ where $\mu$ is the
  behavior distribution, and show it has a positive eigenvalue.
\item Explain why this SA iteration diverges.
\end{enumerate}
\end{exercise}

\begin{exercise}[SA with Markovian noise]
\label{ex:markov-noise}
The standard SA theory assumes martingale difference noise. In RL, the noise is
typically \emph{Markovian} (since observations come from a Markov chain). Explain how
the ODE method extends to Markovian noise under the following conditions:
\begin{enumerate}[label=(\alph*)]
\item The Markov chain $\{Y_n\}$ is ergodic with stationary distribution $\mu$.
\item $h(x) = \int H(x, y) \, d\mu(y) = \E_\mu[H(x, Y)]$.
\item The ``Poisson equation'' $g(x, y) - Pg(x, y) = H(x, y) - h(x)$ has a
  solution $g$ with suitable regularity.
\end{enumerate}
Sketch how $g$ is used to decompose the noise into a martingale difference term plus
a telescoping (bounded) term, recovering the standard ODE method assumptions.
\end{exercise}

\begin{exercise}[Nonlinear two-timescale: GTD2]
\label{ex:gtd2}
The GTD2 algorithm (Sutton et al., 2009) uses two-timescale SA to perform off-policy
TD learning with linear function approximation. The updates are:
\begin{align*}
  w_{n+1} &= w_n + b_n \bigl[\phi_n(\phi_n - \gamma \phi_n')^T \theta_n - \phi_n \phi_n^T w_n\bigr], \\
  \theta_{n+1} &= \theta_n + a_n \bigl[\delta_n \phi_n - \gamma \phi_n' (\phi_n^T w_n)\bigr],
\end{align*}
where $\phi_n = \phi(s_n)$, $\phi_n' = \phi(s_{n+1})$, $\delta_n = R_{n+1} + \gamma \phi_n'^T \theta_n - \phi_n^T \theta_n$.
\begin{enumerate}[label=(\alph*)]
\item Identify the fast and slow iterates and verify the timescale separation.
\item Compute the fast-timescale equilibrium $\lambda(\theta)$ for the $w$-update.
\item Substitute $\lambda(\theta)$ into the $\theta$-update to obtain the
  slow-timescale ODE.
\item Show that the slow-timescale ODE is gradient descent on the \emph{mean squared
  projected Bellman error} (MSPBE).
\end{enumerate}
\end{exercise}


%% ============================================================
%% SECTION 9: SUMMARY
%% ============================================================

\section{Summary of Key Results}
\label{sec:summary}

\begin{center}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{@{}p{3.8cm}p{4cm}p{5.5cm}@{}}
\toprule
\textbf{Result} & \textbf{Key Condition} & \textbf{RL Application} \\
\midrule
Robbins--Monro (Thm~\ref{thm:rm-convergence})
  & Monotone $h$, MDS noise, $\sum a_n = \infty$, $\sum a_n^2 < \infty$
  & SGD convergence for convex losses \\
\addlinespace
ODE Method (Thm~\ref{thm:ode-convergence})
  & Bounded iterates, Lipschitz $h$, stable ODE equilibrium
  & Primary tool for all RL convergence proofs \\
\addlinespace
Borkar--Meyn (Thm~\ref{thm:borkar-meyn})
  & $h_\infty$ ODE has stable origin
  & Verifies boundedness for linear SA / TD \\
\addlinespace
SA CLT (Thm~\ref{thm:sa-clt})
  & $a_n = a/n$, Hurwitz Jacobian
  & Rate: $\|x_n - x^*\| = O(1/\sqrt{n})$ \\
\addlinespace
Polyak--Ruppert (Thm~\ref{thm:polyak-ruppert})
  & Average iterates, $a_n = O(n^{-\alpha})$, $\alpha \in (1/2,1)$
  & Optimal rate via iterate averaging (target networks, EMA) \\
\addlinespace
Two-Timescale (Thm~\ref{thm:two-timescale})
  & $a_n/b_n \to 0$, fast/slow ODE stability
  & Actor-critic, GTD, GTD2 convergence \\
\addlinespace
Linear SA (Thm~\ref{thm:linear-sa-complete})
  & $A$ Hurwitz
  & TD(0) with linear FA (on-policy) \\
\addlinespace
Lock-in (Prop~\ref{prop:lock-in})
  & Multiple stable equilibria, nondegenerate noise
  & Policy gradient converges to local optimum, avoids saddle points \\
\bottomrule
\end{tabular}
\end{center}

\bigskip

\begin{center}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{@{}p{3cm}p{3.5cm}p{3cm}p{3.5cm}@{}}
\toprule
\textbf{RL Algorithm} & \textbf{SA Type} & \textbf{$h(x)$ / Drift} & \textbf{Key SA Theorem} \\
\midrule
TD(0) (tabular) & Async.\ linear SA & $(\gamma P^\pi - I)V + r^\pi$ & ODE Method + Hurwitz \\
\addlinespace
TD(0) (linear FA) & Linear SA & $Aw + b$, $A$ Hurwitz (on-policy) & Thm~\ref{thm:linear-sa-complete} \\
\addlinespace
Q-learning & Async.\ nonlinear SA & $\mathcal{T}^*Q - Q$ ($\gamma$-contraction) & ODE + contraction \\
\addlinespace
REINFORCE & SA (gradient ascent) & $\nabla J(\theta)$ & ODE + lock-in \\
\addlinespace
Actor-Critic & Two-timescale SA & Critic: TD drift; Actor: policy gradient & Thm~\ref{thm:two-timescale} \\
\addlinespace
GTD / GTD2 & Two-timescale SA & Gradient of MSPBE & Thm~\ref{thm:two-timescale} \\
\bottomrule
\end{tabular}
\end{center}

\bigskip
\noindent
The central message of this document is a single principle:

\begin{keyresult}[The SA--ODE Correspondence]
\emph{Every} RL update rule of the form $x_{n+1} = x_n + a_n[H(x_n, Y_n)]$ can be
analyzed by:
\begin{enumerate}[leftmargin=2em]
\item Identifying the mean field $h(x) = \E[H(x, Y)]$.
\item Writing the associated ODE $\dot{x} = h(x)$.
\item Analyzing the stability of the ODE (Lyapunov functions, contraction,
  Hurwitz matrices).
\item Applying the appropriate SA convergence theorem.
\end{enumerate}
If the ODE has a stable equilibrium and the noise/step-size conditions are met, the
RL algorithm converges. If the ODE is unstable (e.g., off-policy TD with FA), the RL
algorithm diverges.
\end{keyresult}


\end{document}