\documentclass[11pt,a4paper]{article}
\input{../lesson-preamble}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{Lesson 64: Mean-Field Games}}
\fancyhead[R]{\thepage}
\title{Lesson 64: Mean-Field Games}
\author{AI/ML Mathematics}
\date{}
\begin{document}
\maketitle
\tableofcontents
\newpage

%% ============================================================
%% SECTION 1: MOTIVATION AND OVERVIEW
%% ============================================================
\section{Motivation and Overview}
\label{sec:motivation}

Consider a highway network carrying thousands of autonomous vehicles,
each selecting routes and speeds to minimize its own travel time.
Modeling this as an $N$-player stochastic game requires tracking the
joint state of all vehicles -- a state space that grows exponentially
in $N$.  Computing a Nash equilibrium in such a game is intractable
for realistic population sizes.  Yet the agents are \emph{symmetric}:
they share the same dynamics, the same action space, and the same
reward structure, differing only in their individual states.  Each
driver does not care about the identity of the other vehicles, only
about the \emph{aggregate traffic density} on each road segment.

This symmetry suggests a radical simplification.  Instead of tracking
$N$ individual states, we represent the collective behavior of all
other agents by a single probability distribution -- the
\emph{mean-field distribution} $\meanfield$ over states.  In the
limit $N \to \infty$, the law of large numbers ensures that the
empirical distribution of agent states concentrates around a
deterministic flow, and each individual agent faces what is
effectively a single-agent MDP parameterized by this population
distribution.  The tools developed in this lesson -- McKean--Vlasov
dynamics, the forward-backward equilibrium system, and fixed-point
existence theorems -- will let us formalize this mean-field game
(MFG) framework and connect it to practical multi-agent reinforcement
learning algorithms.

\begin{rlconnection}[title={Why Mean-Field Games Matter for RL/ML}]
\begin{itemize}[nosep]
  \item \textbf{Fleet management and ride-sharing.}  Mean-field
    models enable scalable optimization of thousands of vehicles
    competing for passengers in ride-sharing platforms.
  \item \textbf{Crowd dynamics and evacuation.}  MFGs model
    pedestrian flows where each individual optimizes against the
    aggregate crowd density.
  \item \textbf{Large-scale multi-agent RL.}  Algorithms such as
    mean-field Q-learning \citep{yang2018} and mean-field
    actor-critic \citep{gu2021} scale MARL to populations of
    hundreds or thousands of agents.
  \item \textbf{Network congestion and resource allocation.}
    MFGs capture congestion effects in communication networks
    and shared computing resources where many identical agents
    compete for bandwidth.
\end{itemize}
\end{rlconnection}

\paragraph{Prerequisites.}
Stochastic game formalism and Markov perfect equilibrium (Lesson~61).
Probability measures, weak convergence, and the simplex
$\simplex(\States)$ (Phase~01 measure theory, Lessons~13--14).
Fixed-point theorems: Banach contraction mapping theorem and
Schauder's fixed-point theorem (Lessons~15--16, Theorems on
contractions and compact convex mappings).  Bellman equations and
value iteration for MDPs (Lesson~42).

\paragraph{Learning objectives.}
After completing this lesson, you will be able to:
\begin{itemize}[nosep]
  \item \textbf{Define} a mean-field game and \textbf{explain} how
    it arises as the infinite-population limit of a symmetric
    $N$-player stochastic game.
  \item \textbf{Derive} the McKean--Vlasov dynamics governing the
    evolution of the population distribution under a given policy.
  \item \textbf{Formulate} the forward-backward system that
    characterizes mean-field equilibria.
  \item \textbf{Prove} the existence of mean-field equilibria via
    Schauder's fixed-point theorem and \textbf{state} conditions
    for uniqueness via contraction arguments.
  \item \textbf{Describe} the mean-field Q-learning algorithm and
    \textbf{relate} it to the fixed-point iteration on the
    equilibrium operator.
  \item \textbf{Analyze} the quality of the mean-field approximation
    for finite $N$-player games, establishing the
    $O(1/\sqrt{N})$ approximation bound.
\end{itemize}


%% ============================================================
%% SECTION 2: REFERENCES
%% ============================================================
\section{References}
\label{sec:references}

\begin{itemize}
  \item \citet[Chapters 1--5]{carmona2018i} -- Comprehensive
    treatment of mean-field game foundations, McKean--Vlasov
    equations, and existence theory in continuous time.
  \item \citet{lasry2007} -- Original formulation of mean-field
    games via the coupled Hamilton--Jacobi--Bellman and
    Fokker--Planck PDE system.
  \item \citet{huang2006} -- Independent co-discovery of the
    mean-field game framework through the Nash certainty
    equivalence principle.
  \item \citet{yang2018} -- Mean-field multi-agent reinforcement
    learning: practical algorithms bridging MFG theory and
    deep RL.
  \item \citet{cardaliaguet2010} -- Lecture notes providing an
    accessible introduction to mean-field games with complete
    proofs in the finite-state setting.
\end{itemize}


%% ============================================================
%% SECTION 3: FROM N-PLAYER GAMES TO MEAN-FIELD LIMITS
%% ============================================================
\section{From $N$-Player Games to Mean-Field Limits}
\label{sec:n-player-to-mf}

We begin by formalizing the class of $N$-player games for which the
mean-field approximation is appropriate.  The key structural
requirement is \emph{symmetry}: agents are interchangeable, and
each agent's dynamics and rewards depend on the others only through
the empirical distribution of their states.

\begin{definition}[Symmetric $N$-player stochastic game]
\label{def:symmetric-game}
A \emph{symmetric $N$-player stochastic game} is a tuple
$\mathcal{G}^N = (\Players, \States, \Actions, P^N, R^N, \discount)$
where:
\begin{enumerate}[nosep]
  \item $\Players = \{1, \ldots, N\}$ is the set of agents.
  \item $\States$ is a finite state space shared by all agents.
        The joint state is $\mathbf{s} = (s_1, \ldots, s_N)
        \in \States^N$.
  \item $\Actions$ is a finite action space shared by all agents.
  \item The transition kernel for agent $i$ is
    \[
      P^N(s_i' \mid s_i, a_i, \mathbf{s}_{-i})
      = P^N\!\left(s_i' \,\Big|\, s_i, a_i,
        \frac{1}{N-1}\sum_{j \neq i} \delta_{s_j}\right),
    \]
    depending on the other agents' states only through their
    empirical distribution.
  \item The reward for agent $i$ is
    \[
      R^N(s_i, a_i, \mathbf{s}_{-i})
      = R^N\!\left(s_i, a_i,
        \frac{1}{N-1}\sum_{j \neq i} \delta_{s_j}\right),
    \]
    again depending on others only through the empirical measure.
  \item $\discount \in [0,1)$ is the common discount factor.
\end{enumerate}
\end{definition}

The crucial feature is that the functions $P^N$ and $R^N$ depend on
the other agents only through the \emph{empirical measure} of their
states.  We denote this empirical measure explicitly.

\begin{definition}[Empirical state distribution]
\label{def:empirical-measure}
Given joint state $\mathbf{s} = (s_1, \ldots, s_N) \in \States^N$,
the \emph{empirical state distribution} is
\[
  \meanfield^N(\cdot \mid \mathbf{s})
  = \frac{1}{N}\sum_{i=1}^N \delta_{s_i}(\cdot)
  \in \simplex(\States),
\]
where $\delta_{s_i}$ is the Dirac measure at $s_i$ and
$\simplex(\States)$ is the probability simplex over $\States$.
When excluding agent $i$, we write
$\meanfield^{N,-i} = \frac{1}{N-1}\sum_{j \neq i} \delta_{s_j}$.
\end{definition}

As the number of agents grows, the empirical measure concentrates
around a deterministic limit.

\begin{proposition}[Concentration of the empirical measure]
\label{prop:empirical-concentration}
Suppose the states $(s_1, \ldots, s_N)$ are drawn independently
from a common distribution $\meanfield \in \simplex(\States)$.
Then for every $s \in \States$ and every $\epsilon > 0$,
\[
  \Prob\!\left(\left|\meanfield^N(s) - \meanfield(s)\right|
  \geq \epsilon\right) \leq 2\exp\!\left(-2N\epsilon^2\right).
\]
Consequently, $\meanfield^N \to \meanfield$ almost surely as
$N \to \infty$.
\end{proposition}

\begin{proof}
Fix $s \in \States$.  Define $X_i = \indicator_{\{s_i = s\}}$ for
$i = 1, \ldots, N$.  Since the states are drawn independently from
$\meanfield$, the random variables $X_1, \ldots, X_N$ are i.i.d.\
with $\E[X_i] = \meanfield(s)$ and $X_i \in [0,1]$.  The empirical
measure satisfies $\meanfield^N(s) = \frac{1}{N}\sum_{i=1}^N X_i$.
By Hoeffding's inequality,
\[
  \Prob\!\left(\left|\meanfield^N(s) - \meanfield(s)\right|
  \geq \epsilon\right)
  \leq 2\exp\!\left(-2N\epsilon^2\right).
\]
Taking a union bound over the finite set $\States$ and applying
the Borel--Cantelli lemma, $\|\meanfield^N - \meanfield\|_\infty
\to 0$ almost surely as $N \to \infty$.
\end{proof}

\begin{example}[Cars on a ring road]
\label{ex:ring-road}
Consider $N$ cars on a circular road with $|\States| = K$ equally
spaced positions.  Each car chooses speed $a \in \Actions
= \{0, 1, 2\}$ (stop, slow, fast).  The transition probability
depends on whether the next position is congested:
\[
  P^N(s' \mid s, a, \meanfield^{N,-i})
  = \begin{cases}
    1 - c \cdot \meanfield^{N,-i}(s') & \text{if } s' = s + a
      \pmod{K}, \\
    c \cdot \meanfield^{N,-i}(s') & \text{if } s' = s
      \quad (\text{blocked}),
  \end{cases}
\]
where $c \in (0,1)$ is a congestion parameter.  In the limit
$N \to \infty$, each car faces an MDP where the blocking
probability at position $s'$ is determined by the
population density $\meanfield_t(s')$ rather than by the random
empirical measure of finitely many other cars.
\end{example}

The example illustrates the fundamental insight: in the limit
$N \to \infty$, each agent faces a single-agent optimization
problem parameterized by the population distribution.  We
formalize this in the next section.


%% ============================================================
%% SECTION 4: MEAN-FIELD GAME FORMULATION (DISCRETE TIME)
%% ============================================================
\section{Mean-Field Game Formulation (Discrete Time)}
\label{sec:mfg-formulation}

Having motivated the transition from $N$-player games to
population-level descriptions, we now define the mean-field game
in discrete time.

\begin{definition}[Discrete-time mean-field game]
\label{def:mfg}
A \emph{discrete-time mean-field game} is a tuple
$\mathcal{M} = (\States, \Actions, P, R, \discount)$ where:
\begin{enumerate}[nosep]
  \item $\States$ is a finite state space.
  \item $\Actions$ is a finite action space.
  \item $P: \States \times \Actions \times \simplex(\States)
    \to \simplex(\States)$ is the transition kernel, where
    $P(s' \mid s, a, \meanfield)$ gives the probability of
    transitioning to $s'$ when in state $s$, taking action $a$,
    and the population distribution is $\meanfield$.
  \item $R: \States \times \Actions \times \simplex(\States)
    \to \R$ is the reward function.
  \item $\discount \in [0,1)$ is the discount factor.
\end{enumerate}
\end{definition}

The key distinction from a standard MDP is that both the transition
kernel and the reward depend on the population distribution
$\meanfield \in \simplex(\States)$.  For a fixed population flow,
however, each agent faces a (time-varying) MDP.

\begin{definition}[Mean-field MDP]
\label{def:mf-mdp}
Given a population flow $(\meanfield_t)_{t \geq 0}$ with
$\meanfield_t \in \simplex(\States)$ for each $t$, the
\emph{mean-field MDP} is the single-agent MDP with time-varying
transition kernel $P(\cdot \mid s, a, \meanfield_t)$ and
time-varying reward $R(s, a, \meanfield_t)$.  The value function
of a policy $\policy$ is
\[
  V_t^{\policy}(s; (\meanfield_\tau)_{\tau \geq t})
  = \E\!\left[\sum_{k=0}^{\infty} \discount^k
    R(s_{t+k}, a_{t+k}, \meanfield_{t+k})
    \,\Big|\, s_t = s\right],
\]
where $a_{t+k} \sim \policy(\cdot \mid s_{t+k})$ and
$s_{t+k+1} \sim P(\cdot \mid s_{t+k}, a_{t+k}, \meanfield_{t+k})$.
\end{definition}

When the population distribution is \emph{stationary}
($\meanfield_t = \meanfield$ for all $t$), the mean-field MDP
becomes a standard time-homogeneous MDP.

\begin{proposition}[Bellman equation for the stationary mean-field MDP]
\label{prop:bellman-mf}
For a fixed stationary population distribution $\meanfield \in
\simplex(\States)$, the optimal value function $V^*(s; \meanfield)$
satisfies the Bellman optimality equation
\begin{equation}
\label{eq:bellman-mf}
  V^*(s; \meanfield) = \max_{a \in \Actions}\left[
    R(s, a, \meanfield)
    + \discount \sum_{s' \in \States}
      P(s' \mid s, a, \meanfield)\, V^*(s'; \meanfield)
  \right].
\end{equation}
The corresponding optimal policy is
\[
  \policy^*(s; \meanfield) \in \argmax_{a \in \Actions}\left[
    R(s, a, \meanfield)
    + \discount \sum_{s' \in \States}
      P(s' \mid s, a, \meanfield)\, V^*(s'; \meanfield)
  \right].
\]
\end{proposition}

\begin{proof}
For fixed $\meanfield$, the mean-field MDP is a standard discounted
MDP with finite state and action spaces, transition kernel
$P(\cdot \mid s, a, \meanfield)$, and reward $R(s, a, \meanfield)$.
The Bellman optimality equation~\eqref{eq:bellman-mf} follows
directly from the standard MDP theory (Lesson~41, Theorem on
Bellman optimality).  Existence of an optimal deterministic
stationary policy follows from the finiteness of $\Actions$.
\end{proof}

\begin{rlconnection}[title={Mean-Field MDP as a Standard MDP}]
Since the mean-field MDP with fixed $\meanfield$ is a standard MDP,
\emph{all single-agent RL algorithms apply directly}: value
iteration, policy iteration, Q-learning, and policy gradient methods
can all be used to solve the agent's optimization problem for a given
population distribution.  The challenge of the mean-field game is
that $\meanfield$ is itself determined by the agents' policies.
\end{rlconnection}

We now turn to the equation that governs the population distribution
-- the forward equation, or McKean--Vlasov dynamics.


%% ============================================================
%% SECTION 5: MCKEAN-VLASOV DYNAMICS
%% ============================================================
\section{McKean--Vlasov Dynamics}
\label{sec:mckean-vlasov}

In the $N$-player game, the empirical state distribution evolves
stochastically as agents take actions and transition between states.
In the mean-field limit, this empirical distribution converges to a
deterministic flow governed by the \emph{McKean--Vlasov equation}.

\begin{definition}[McKean--Vlasov dynamics (discrete time)]
\label{def:mckean-vlasov}
Given a policy $\policy: \States \to \simplex(\Actions)$ and a
population distribution $\meanfield_t \in \simplex(\States)$, the
\emph{McKean--Vlasov equation} determines the next-step population
distribution:
\begin{equation}
\label{eq:mckean-vlasov}
  \meanfield_{t+1}(s') = \sum_{s \in \States}
    \sum_{a \in \Actions}
    P(s' \mid s, a, \meanfield_t)\,
    \policy(a \mid s)\,
    \meanfield_t(s),
    \qquad \forall\, s' \in \States.
\end{equation}
Starting from an initial distribution $\meanfield_0$, repeated
application of~\eqref{eq:mckean-vlasov} generates the
\emph{population flow} $(\meanfield_t)_{t \geq 0}$.
\end{definition}

\begin{remark}
\label{rem:forward-equation}
Equation~\eqref{eq:mckean-vlasov} is the \emph{forward equation}
of the mean-field game.  It is the discrete-time analogue of the
Fokker--Planck (or Kolmogorov forward) equation in continuous time.
The equation is nonlinear because the transition kernel $P$ itself
depends on $\meanfield_t$.
\end{remark}

We verify that the McKean--Vlasov dynamics preserve the simplex.

\begin{lemma}[Simplex invariance]
\label{lem:simplex-invariance}
If $\meanfield_t \in \simplex(\States)$ and $\policy(\cdot \mid s)
\in \simplex(\Actions)$ for all $s$, then $\meanfield_{t+1}$ defined
by~\eqref{eq:mckean-vlasov} satisfies $\meanfield_{t+1}
\in \simplex(\States)$.
\end{lemma}

\begin{proof}
Non-negativity is immediate since $P$, $\policy$, and $\meanfield_t$
are all non-negative.  For normalization:
\begin{align*}
  \sum_{s' \in \States} \meanfield_{t+1}(s')
  &= \sum_{s'} \sum_{s} \sum_{a}
    P(s' \mid s, a, \meanfield_t)\,
    \policy(a \mid s)\, \meanfield_t(s) \\
  &= \sum_{s} \sum_{a} \policy(a \mid s)\, \meanfield_t(s)
    \underbrace{\sum_{s'} P(s' \mid s, a, \meanfield_t)}_{=1} \\
  &= \sum_{s} \meanfield_t(s)
    \underbrace{\sum_{a} \policy(a \mid s)}_{=1}
  = \sum_{s} \meanfield_t(s) = 1. \qedhere
\end{align*}
\end{proof}

We define the operator that maps a policy to its induced population
flow, which will play a central role in the equilibrium analysis.

\begin{definition}[Forward operator]
\label{def:forward-operator}
The \emph{forward operator} $\Phi$ maps a policy $\policy$ and an
initial distribution $\meanfield_0$ to the population flow:
\[
  \Phi(\policy, \meanfield_0)
  = (\meanfield_0, \meanfield_1, \meanfield_2, \ldots)
\]
where $\meanfield_{t+1}$ is computed from $\meanfield_t$ and
$\policy$ via~\eqref{eq:mckean-vlasov}.  In the stationary case,
we seek a fixed distribution $\meanfield^*$ such that
$\Phi(\policy, \meanfield^*) = (\meanfield^*, \meanfield^*,
\ldots)$, i.e., $\meanfield^*$ is a \emph{stationary distribution}
of the McKean--Vlasov dynamics under $\policy$.
\end{definition}

\begin{definition}[Backward operator]
\label{def:backward-operator}
The \emph{backward operator} $\Psi$ maps a population flow
$(\meanfield_t)_{t \geq 0}$ to the optimal policy for the
corresponding mean-field MDP:
\[
  \Psi((\meanfield_t)_{t \geq 0}) = \policy^*,
\]
where $\policy^*$ is the optimal policy for the MDP with transitions
$P(\cdot \mid s, a, \meanfield_t)$ and rewards $R(s, a,
\meanfield_t)$.  In the stationary case,
$\Psi(\meanfield) = \policy^*(\cdot; \meanfield)$ solves the
Bellman equation~\eqref{eq:bellman-mf}.
\end{definition}

The forward and backward equations form a coupled system: the
forward equation propagates the population distribution given a
policy (Section~5), while the backward equation computes the
optimal policy given a population flow (Section~4).  A
mean-field equilibrium requires solving both simultaneously.

\begin{example}[Two-state McKean--Vlasov dynamics]
\label{ex:two-state-mv}
Let $\States = \{0, 1\}$, $\Actions = \{L, R\}$, and suppose the
transition kernel is
\[
  P(1 \mid 0, R, \meanfield) = 1 - c \cdot \meanfield(1),
  \qquad
  P(0 \mid 1, L, \meanfield) = 1 - c \cdot \meanfield(0),
\]
with $c = 0.5$, and all other transitions determined by
normalization.  Under a uniform random policy $\policy(L \mid s)
= \policy(R \mid s) = 0.5$, starting from $\meanfield_0 = (0.5,
0.5)$, one step of the McKean--Vlasov dynamics gives
\begin{align*}
  \meanfield_1(1)
  &= \meanfield_0(0) \cdot \policy(R \mid 0)
    \cdot P(1 \mid 0, R, \meanfield_0) \\
  &\quad + \meanfield_0(0) \cdot \policy(L \mid 0)
    \cdot P(1 \mid 0, L, \meanfield_0) \\
  &\quad + \meanfield_0(1) \cdot \policy(R \mid 1)
    \cdot P(1 \mid 1, R, \meanfield_0) \\
  &\quad + \meanfield_0(1) \cdot \policy(L \mid 1)
    \cdot P(1 \mid 1, L, \meanfield_0).
\end{align*}
With the specific transition model and $\meanfield_0(1) = 0.5$,
this evaluates to a concrete number determined by the congestion
parameter $c$.  The distribution $\meanfield_1(0) = 1 -
\meanfield_1(1)$ follows by normalization.
\end{example}

We now formalize the notion of equilibrium that couples the forward
and backward equations.


%% ============================================================
%% SECTION 6: MEAN-FIELD EQUILIBRIUM
%% ============================================================
\section{Mean-Field Equilibrium}
\label{sec:mf-equilibrium}

The central concept of mean-field game theory is the
\emph{mean-field equilibrium} (MFE), which requires the
population flow and the individual optimal policy to be mutually
consistent.

\begin{definition}[Mean-field equilibrium -- stationary case]
\label{def:mfe-stationary}
A pair $(\policy^*, \meanfield^*)$ is a \emph{stationary
mean-field equilibrium} (MFE) if:
\begin{enumerate}[nosep]
  \item \textbf{Optimality:} $\policy^*$ is an optimal policy for
    the mean-field MDP with stationary population distribution
    $\meanfield^*$, i.e., $\policy^* = \Psi(\meanfield^*)$.
  \item \textbf{Consistency:} $\meanfield^*$ is the stationary
    distribution of the McKean--Vlasov dynamics
    under $\policy^*$, i.e.,
    \[
      \meanfield^*(s')
      = \sum_{s \in \States} \sum_{a \in \Actions}
        P(s' \mid s, a, \meanfield^*)\,
        \policy^*(a \mid s)\, \meanfield^*(s),
        \qquad \forall\, s' \in \States.
    \]
\end{enumerate}
\end{definition}

\begin{definition}[Mean-field equilibrium -- general case]
\label{def:mfe-general}
A pair $(\policy^*, (\meanfield_t^*)_{t \geq 0})$ is a
\emph{mean-field equilibrium} if:
\begin{enumerate}[nosep]
  \item $\policy^*$ is optimal for the mean-field MDP with
    population flow $(\meanfield_t^*)_{t \geq 0}$.
  \item $(\meanfield_t^*)_{t \geq 0}$ is the McKean--Vlasov flow
    induced by $\policy^*$ starting from initial distribution
    $\meanfield_0$.
\end{enumerate}
\end{definition}

The equilibrium condition is a \emph{fixed-point condition}.  We
make this precise by defining the combined equilibrium operator.

\begin{definition}[Equilibrium operator]
\label{def:equilibrium-operator}
The \emph{equilibrium operator} $\Lambda: \simplex(\States)
\to \simplex(\States)$ (in the stationary case) is defined by
the composition
\[
  \Lambda(\meanfield)
  = \Phi\!\bigl(\Psi(\meanfield),\, \meanfield\bigr),
\]
where $\Psi(\meanfield) = \policy^*(\cdot; \meanfield)$ is the
optimal policy for the mean-field MDP with distribution $\meanfield$
(backward step), and $\Phi(\policy^*, \meanfield)$ is the stationary
distribution of the Markov chain with transition kernel
\[
  T(s' \mid s; \meanfield)
  = \sum_{a \in \Actions}
    P(s' \mid s, a, \meanfield)\, \policy^*(a \mid s; \meanfield)
\]
(forward step).  A stationary MFE is a fixed point of $\Lambda$:
$\Lambda(\meanfield^*) = \meanfield^*$.
\end{definition}

\begin{keyresult}[title={Mean-Field Equilibrium as a Fixed Point}]
A stationary mean-field equilibrium $(\policy^*, \meanfield^*)$ is
characterized by the fixed-point equation
\[
  \meanfield^* = \Lambda(\meanfield^*),
\]
where $\Lambda = \Phi \circ \Psi$ composes the backward operator
(solve the MDP) with the forward operator (compute the stationary
distribution).  The existence and uniqueness of MFE reduce to
fixed-point theory for the operator $\Lambda$ on the compact convex
set $\simplex(\States)$.
\end{keyresult}

\begin{remark}
\label{rem:analogy-ne}
The MFE is the infinite-population analogue of a symmetric Nash
equilibrium: every agent plays the same policy $\policy^*$, and no
agent has an incentive to deviate given the population distribution
$\meanfield^*$.  The consistency condition replaces the finite-game
requirement that each player's strategy be a best response to the
others' strategies.
\end{remark}


%% ============================================================
%% SECTION 7: EXISTENCE AND UNIQUENESS OF MEAN-FIELD EQUILIBRIA
%% ============================================================
\section{Existence and Uniqueness of Mean-Field Equilibria}
\label{sec:existence-uniqueness}

We now establish the existence of MFE via topological fixed-point
arguments and provide conditions for uniqueness via contraction
mappings.

\begin{assumption}[Regularity conditions]
\label{ass:regularity}
Throughout this section we assume:
\begin{enumerate}[nosep]
  \item $\States$ and $\Actions$ are finite sets.
  \item The transition kernel $P(s' \mid s, a, \meanfield)$ is
    continuous in $\meanfield \in \simplex(\States)$ for every
    fixed $(s, s', a)$.
  \item The reward function $R(s, a, \meanfield)$ is continuous
    in $\meanfield$ for every fixed $(s, a)$.
  \item The Markov chain with transition kernel
    $T(s' \mid s; \meanfield) = \sum_a P(s' \mid s, a, \meanfield)
    \, \policy^*(a \mid s; \meanfield)$ is ergodic (irreducible
    and aperiodic) for every $\meanfield \in \simplex(\States)$.
\end{enumerate}
\end{assumption}

The ergodicity assumption ensures that the forward operator
$\Phi(\policy, \cdot)$ has a unique stationary distribution for
each policy, making $\Lambda$ well-defined as a map on
$\simplex(\States)$.

\begin{lemma}[Continuity of the backward operator]
\label{lem:backward-continuity}
Under Assumption~\ref{ass:regularity}, the optimal value function
$V^*(s; \meanfield)$ is continuous in $\meanfield$ for each $s$,
and the optimal policy correspondence $\meanfield \mapsto
\policy^*(\cdot; \meanfield)$ admits a continuous selection.
\end{lemma}

\begin{proof}
For fixed $\meanfield$, the Bellman operator $\Bellman_\meanfield$
on $\R^{|\States|}$ is defined by
\[
  (\Bellman_\meanfield V)(s) = \max_{a \in \Actions}\!\left[
    R(s, a, \meanfield) + \discount \sum_{s'}
    P(s' \mid s, a, \meanfield)\, V(s')
  \right].
\]
Since $\Bellman_\meanfield$ is a $\discount$-contraction on
$(\R^{|\States|}, \|\cdot\|_\infty)$, its unique fixed point
$V^*(\cdot; \meanfield)$ depends continuously on $\meanfield$ by
the following argument.  For any $\meanfield, \meanfield' \in
\simplex(\States)$,
\begin{align*}
  \|V^*(\cdot; \meanfield) - V^*(\cdot; \meanfield')\|_\infty
  &= \|\Bellman_\meanfield V^*(\cdot; \meanfield)
    - \Bellman_{\meanfield'} V^*(\cdot; \meanfield')\|_\infty \\
  &\leq \|\Bellman_\meanfield V^*(\cdot; \meanfield)
    - \Bellman_\meanfield V^*(\cdot; \meanfield')\|_\infty \\
  &\quad + \|\Bellman_\meanfield V^*(\cdot; \meanfield')
    - \Bellman_{\meanfield'} V^*(\cdot; \meanfield')\|_\infty \\
  &\leq \discount \|V^*(\cdot; \meanfield)
    - V^*(\cdot; \meanfield')\|_\infty
    + \epsilon(\meanfield, \meanfield'),
\end{align*}
where $\epsilon(\meanfield, \meanfield') = \|\Bellman_\meanfield
V^*(\cdot; \meanfield') - \Bellman_{\meanfield'} V^*(\cdot;
\meanfield')\|_\infty \to 0$ as $\meanfield' \to \meanfield$ by
continuity of $R$ and $P$ in $\meanfield$.  Rearranging,
$\|V^*(\cdot; \meanfield) - V^*(\cdot; \meanfield')\|_\infty
\leq \epsilon(\meanfield, \meanfield') / (1 - \discount) \to 0$.

For the policy, when the optimal action is unique at every state
(which holds generically), $\policy^*(a \mid s; \meanfield)$ is
continuous in $\meanfield$.  In general, one can select a
measurable (in fact continuous) optimal policy by Berge's maximum
theorem, since the objective $Q^*(s, a; \meanfield) = R(s, a,
\meanfield) + \discount \sum_{s'} P(s' \mid s, a, \meanfield)
V^*(s'; \meanfield)$ is continuous in $\meanfield$ and the
constraint set $\Actions$ is finite (hence trivially continuous
as a correspondence).
\end{proof}

\begin{lemma}[Continuity of the forward operator]
\label{lem:forward-continuity}
Under Assumption~\ref{ass:regularity}, the stationary distribution
of the Markov chain with transition kernel $T(\cdot \mid \cdot;
\meanfield)$ depends continuously on $\meanfield$.
\end{lemma}

\begin{proof}
By the ergodicity assumption, for each $\meanfield$ the chain has a
unique stationary distribution $\nu(\meanfield) \in \simplex(\States)$
satisfying $\nu(\meanfield)^\top T(\meanfield) = \nu(\meanfield)^\top$,
where $T(\meanfield)$ is the $|\States| \times |\States|$ transition
matrix.  The entries of $T(\meanfield)$ are continuous in $\meanfield$
because $P$ and $\policy^*$ are continuous in $\meanfield$ (the
latter by Lemma~\ref{lem:backward-continuity}).

The stationary distribution is the left eigenvector of
$T(\meanfield)$ corresponding to eigenvalue~$1$, normalized to sum
to~$1$.  Since $T(\meanfield)$ is ergodic, eigenvalue~$1$ is simple,
and by standard perturbation theory for simple eigenvalues of
matrices, the corresponding eigenvector depends continuously on
the matrix entries.  Hence $\nu(\meanfield)$ is continuous in
$\meanfield$.
\end{proof}

\begin{theorem}[Existence of MFE via Schauder]
\label{thm:existence}
Under Assumption~\ref{ass:regularity}, a stationary mean-field
equilibrium exists.
\end{theorem}

\begin{proof}
The equilibrium operator $\Lambda: \simplex(\States) \to
\simplex(\States)$ is defined by $\Lambda(\meanfield) =
\Phi(\Psi(\meanfield), \meanfield)$, which maps $\meanfield$ to
the stationary distribution of the Markov chain induced by the
optimal policy $\Psi(\meanfield)$ and population distribution
$\meanfield$.

We verify the hypotheses of Schauder's fixed-point theorem
(Lesson~16, Theorem on fixed points for continuous maps on
compact convex sets):

\begin{enumerate}
  \item \textbf{Domain:} $\simplex(\States) \subset \R^{|\States|}$
    is compact and convex.

  \item \textbf{Self-mapping:} By Lemma~\ref{lem:simplex-invariance}
    and the uniqueness of stationary distributions under ergodicity,
    $\Lambda(\meanfield) \in \simplex(\States)$ for every $\meanfield
    \in \simplex(\States)$.

  \item \textbf{Continuity:} $\Lambda = \Phi \circ \Psi$ is the
    composition of two continuous maps.  The backward operator $\Psi$
    is continuous by Lemma~\ref{lem:backward-continuity}, and the
    forward operator $\Phi$ (mapping policy to stationary
    distribution) is continuous by Lemma~\ref{lem:forward-continuity}.
\end{enumerate}

Since $\Lambda$ is a continuous self-map of the compact convex set
$\simplex(\States) \subset \R^{|\States|}$, Schauder's fixed-point
theorem guarantees the existence of $\meanfield^* \in
\simplex(\States)$ with $\Lambda(\meanfield^*) = \meanfield^*$.
Setting $\policy^* = \Psi(\meanfield^*)$, the pair $(\policy^*,
\meanfield^*)$ is a stationary MFE by
Definition~\ref{def:mfe-stationary}.
\end{proof}

\begin{keyresult}[title={Existence of Mean-Field Equilibrium}]
Under continuity and ergodicity assumptions
(Assumption~\ref{ass:regularity}), every discrete-time mean-field
game with finite state and action spaces admits at least one
stationary mean-field equilibrium.  The proof relies on Schauder's
fixed-point theorem applied to the equilibrium operator
$\Lambda: \simplex(\States) \to \simplex(\States)$.
\end{keyresult}

We now turn to uniqueness, which requires stronger conditions.

\begin{assumption}[Lipschitz conditions]
\label{ass:lipschitz}
There exist constants $L_P, L_R > 0$ such that for all
$s, s' \in \States$, $a \in \Actions$, and $\meanfield, \meanfield'
\in \simplex(\States)$:
\begin{align*}
  |P(s' \mid s, a, \meanfield) - P(s' \mid s, a, \meanfield')|
  &\leq L_P \|\meanfield - \meanfield'\|_1, \\
  |R(s, a, \meanfield) - R(s, a, \meanfield')|
  &\leq L_R \|\meanfield - \meanfield'\|_1.
\end{align*}
\end{assumption}

\begin{theorem}[Uniqueness via contraction]
\label{thm:uniqueness}
Under Assumptions~\ref{ass:regularity}--\ref{ass:lipschitz},
suppose the discount factor satisfies
\begin{equation}
\label{eq:uniqueness-condition}
  \discount \cdot \left(
    L_P \cdot \frac{R_{\max}}{(1-\discount)^2}
    + \frac{L_P + L_R}{1-\discount}
  \right) \cdot C_T < 1,
\end{equation}
where $R_{\max} = \max_{s,a,\meanfield} |R(s,a,\meanfield)|$ and
$C_T$ is the sensitivity constant of the stationary distribution
with respect to transition matrix perturbations.  Then $\Lambda$
is a contraction on $(\simplex(\States), \|\cdot\|_1)$, and
the MFE is unique.
\end{theorem}

\begin{proof}
We bound the Lipschitz constant of $\Lambda$ by bounding the
sensitivity of each component.

\textbf{Step 1: Sensitivity of the value function.}
From the proof of Lemma~\ref{lem:backward-continuity}, for any
$\meanfield, \meanfield'$:
\[
  \|V^*(\cdot; \meanfield) - V^*(\cdot; \meanfield')\|_\infty
  \leq \frac{1}{1-\discount}\left(
    L_R + \frac{\discount \, L_P \, R_{\max}}{1-\discount}
  \right) \|\meanfield - \meanfield'\|_1.
\]
This follows by bounding $\epsilon(\meanfield, \meanfield')$ in
terms of the Lipschitz constants.  Specifically,
\begin{align*}
  |(\Bellman_\meanfield V)(s) - (\Bellman_{\meanfield'} V)(s)|
  &\leq \max_a \bigl[|R(s,a,\meanfield) - R(s,a,\meanfield')| \\
  &\quad + \discount |P(\cdot|s,a,\meanfield) - P(\cdot|s,a,
    \meanfield')| \cdot \|V\|_\infty \bigr] \\
  &\leq L_R \|\meanfield-\meanfield'\|_1
    + \discount\, L_P \|V\|_\infty \|\meanfield-\meanfield'\|_1.
\end{align*}
Setting $V = V^*(\cdot; \meanfield')$ with $\|V^*(\cdot;
\meanfield')\|_\infty \leq R_{\max}/(1-\discount)$ gives the bound.

\textbf{Step 2: Sensitivity of the optimal policy.}
The Q-function $Q^*(s,a;\meanfield)$ inherits the Lipschitz
property from $V^*$.  When the optimal action is unique, the
policy is locally constant and hence Lipschitz with constant~$0$.
In general, by considering soft-max policies (or policies that
break ties consistently), the policy varies Lipschitz-continuously
in $\meanfield$ with a constant controlled by the value function
sensitivity.

\textbf{Step 3: Sensitivity of the stationary distribution.}
By standard perturbation theory for ergodic Markov chains
\citep[Chapter~3]{bensoussan2013}, the stationary distribution
$\nu$ of a transition matrix $T$ satisfies
\[
  \|\nu(T) - \nu(T')\|_1 \leq C_T \|T - T'\|_\infty,
\]
where $C_T$ depends on the spectral gap of the chain.  The
perturbation $\|T(\meanfield) - T(\meanfield')\|_\infty$ is
bounded by the combined sensitivity of $P$ and $\policy^*$ to
$\meanfield$.

\textbf{Step 4: Composing the bounds.}
Combining Steps 1--3:
\[
  \|\Lambda(\meanfield) - \Lambda(\meanfield')\|_1
  \leq \underbrace{\discount \cdot \left(
    L_P \cdot \frac{R_{\max}}{(1-\discount)^2}
    + \frac{L_P + L_R}{1-\discount}
  \right) \cdot C_T}_{=:\, \kappa}
  \|\meanfield - \meanfield'\|_1.
\]
When $\kappa < 1$ (condition~\eqref{eq:uniqueness-condition}),
$\Lambda$ is a contraction.  By the Banach fixed-point theorem,
the fixed point is unique.
\end{proof}

\begin{warning}[title={Non-Uniqueness of Mean-Field Equilibria}]
Without the contraction condition~\eqref{eq:uniqueness-condition},
multiple MFEs can exist.  This is analogous to the multiplicity
of Nash equilibria in finite games.  In particular, when the
discount factor is large (agents are patient) or the
mean-field dependence is strong ($L_P, L_R$ are large), the
equilibrium operator $\Lambda$ may have multiple fixed points.
Example~\ref{ex:multiple-mfe} below illustrates this phenomenon.
\end{warning}

\begin{example}[Congestion game with explicit MFE]
\label{ex:congestion-mfe}
Consider $\States = \{0, 1\}$ (two routes), $\Actions = \{0, 1\}$
(stay or switch), and the congestion-dependent reward
$R(s, a, \meanfield) = 1 - \meanfield(s)$ (higher reward when
the current route is less congested).  Transitions:
$P(1-s \mid s, 1, \meanfield) = 1$ (switching always succeeds) and
$P(s \mid s, 0, \meanfield) = 1$ (staying is deterministic).

With discount $\discount$ close to~$0$, agents are myopic and the
MFE approaches the static Nash equilibrium.  If agents are
symmetric with initial $\meanfield_0 = (0.5, 0.5)$, the unique
stationary MFE is $\meanfield^* = (0.5, 0.5)$ with the optimal
policy ``stay'' at each state (since switching does not improve
the reward when both routes are equally congested).  We verify:
\begin{itemize}[nosep]
  \item \textbf{Optimality:} With $\meanfield^* = (0.5, 0.5)$,
    the reward for staying is $1 - 0.5 = 0.5$ and the reward for
    switching is also $1 - 0.5 = 0.5$ (since both routes have
    density $0.5$).  Any policy is optimal, including ``stay.''
  \item \textbf{Consistency:} Under ``stay,'' $\meanfield_{t+1}
    = \meanfield_t = (0.5, 0.5)$, confirming stationarity.
\end{itemize}
\end{example}


%% ============================================================
%% SECTION 8: MEAN-FIELD Q-LEARNING
%% ============================================================
\section{Mean-Field Q-Learning}
\label{sec:mf-q-learning}

The fixed-point characterization of MFE suggests a natural
computational approach: alternate between solving the backward
equation (optimizing the agent's policy given the population) and
the forward equation (updating the population given the policy).
This is the basis of \emph{mean-field Q-learning}
\citep{yang2018}.

\begin{definition}[Mean-field Q-function]
\label{def:mf-q}
The \emph{mean-field Q-function} is defined as
\begin{equation}
\label{eq:mf-q}
  Q(s, a, \meanfield) = R(s, a, \meanfield)
  + \discount \sum_{s' \in \States}
    P(s' \mid s, a, \meanfield)\, V(s', \meanfield'),
\end{equation}
where $V(s, \meanfield) = \max_{a \in \Actions} Q(s, a, \meanfield)$
and $\meanfield'$ is the next-step population distribution consistent
with the current policy derived from $Q$.
\end{definition}

\begin{definition}[Mean-field Q-learning algorithm]
\label{def:mf-q-algorithm}
The \emph{mean-field Q-learning} algorithm alternates two steps at
each iteration $k = 0, 1, 2, \ldots$:

\textbf{Step 1 (Q-update):} With the current population distribution
$\meanfield_k$ held fixed, perform a Q-learning update:
\begin{equation}
\label{eq:mf-q-update}
  Q_{k+1}(s, a, \meanfield_k)
  = (1 - \alpha_k)\, Q_k(s, a, \meanfield_k)
  + \alpha_k \!\left[R(s, a, \meanfield_k)
    + \discount \max_{a'} Q_k(s', a', \meanfield_k)\right],
\end{equation}
where $\alpha_k \in (0,1)$ is the learning rate and $s' \sim
P(\cdot \mid s, a, \meanfield_k)$.

\textbf{Step 2 (Mean-field update):} Update the population
distribution using a Boltzmann policy derived from the current
Q-function:
\begin{equation}
\label{eq:mf-boltzmann}
  \meanfield_{k+1}(s') = \sum_{s \in \States}
    \sum_{a \in \Actions}
    P(s' \mid s, a, \meanfield_k)\,
    \sigma_\beta(a \mid s, \meanfield_k)\,
    \meanfield_k(s),
\end{equation}
where $\sigma_\beta(a \mid s, \meanfield) =
\frac{\exp(\beta\, Q_k(s, a, \meanfield))}
{\sum_{a'} \exp(\beta\, Q_k(s, a', \meanfield))}$
is the Boltzmann (soft-max) policy with inverse temperature
$\beta > 0$.
\end{definition}

\begin{proposition}[Fixed-point interpretation]
\label{prop:fp-interpretation}
The mean-field Q-learning algorithm is a form of fixed-point
iteration on the equilibrium operator $\Lambda$.  Specifically:
\begin{itemize}[nosep]
  \item Step~1 approximates the backward operator $\Psi$: given
    $\meanfield_k$, it computes the optimal Q-function (and hence
    the optimal policy) for the mean-field MDP with distribution
    $\meanfield_k$.
  \item Step~2 approximates the forward operator $\Phi$: given the
    policy $\sigma_\beta$ derived from $Q_k$, it advances the
    population distribution by one step of the McKean--Vlasov
    dynamics.
\end{itemize}
In the limit $\beta \to \infty$, the Boltzmann policy converges
to the greedy policy, and the algorithm reduces to exact
fixed-point iteration on $\Lambda$.
\end{proposition}

\begin{proof}
At convergence ($Q_k \to Q^*$, $\meanfield_k \to \meanfield^*$),
Step~1 yields the fixed point of the Bellman operator for the
mean-field MDP with distribution $\meanfield^*$, so $Q^*$ encodes
the optimal policy $\policy^* = \Psi(\meanfield^*)$.  Step~2 with
$\beta \to \infty$ reduces to the McKean--Vlasov equation under
$\policy^*$, so $\meanfield^*$ is the stationary distribution of
the forward dynamics under $\policy^*$, i.e.,
$\meanfield^* = \Phi(\policy^*, \meanfield^*)$.  Together,
$\meanfield^* = \Lambda(\meanfield^*)$, which is the MFE
fixed-point condition.
\end{proof}

\begin{proposition}[Convergence of mean-field Q-learning]
\label{prop:mf-q-convergence}
Under Assumptions~\ref{ass:regularity}--\ref{ass:lipschitz} and the
contraction condition~\eqref{eq:uniqueness-condition}, suppose the
learning rates satisfy the Robbins--Monro conditions
$\sum_k \alpha_k = \infty$, $\sum_k \alpha_k^2 < \infty$, and
the inverse temperature $\beta_k \to \infty$ sufficiently slowly.
Then the iterates $(Q_k, \meanfield_k)$ of the mean-field
Q-learning algorithm converge to the unique MFE $(Q^*, \meanfield^*)$
almost surely.
\end{proposition}

\begin{proof}[Proof sketch]
The convergence analysis proceeds in two stages following the
two-timescale stochastic approximation framework
(Lesson~19, SA theory).

\textbf{Fast timescale (Q-update):} For fixed $\meanfield_k$,
Step~1 is standard tabular Q-learning on a finite MDP.  Under the
Robbins--Monro conditions on $\alpha_k$ and appropriate
exploration, $Q_k(\cdot, \cdot, \meanfield_k) \to Q^*(\cdot,
\cdot, \meanfield_k)$ by the convergence theorem for Q-learning
(Lesson~50).

\textbf{Slow timescale (mean-field update):} Viewing $Q_k$ as
approximately converged (the fast timescale has equilibrated),
Step~2 becomes deterministic iteration of $\Lambda$
(with a vanishing Boltzmann smoothing error as $\beta_k \to
\infty$).  Under the contraction
condition~\eqref{eq:uniqueness-condition}, $\Lambda$ is a
contraction, so $\meanfield_k \to \meanfield^*$.

The two-timescale analysis of \citet[Chapter~6]{bensoussan2013}
formalizes the separation of timescales and establishes the almost
sure convergence of the joint iterates.
\end{proof}

\begin{rlconnection}[title={From Theory to Practice in Mean-Field MARL}]
\begin{itemize}[nosep]
  \item \textbf{Scalability.}  Mean-field Q-learning scales to
    hundreds of agents because each agent maintains a Q-function
    of size $|\States| \times |\Actions|$ rather than
    $|\States|^N \times |\Actions|^N$.
  \item \textbf{Function approximation.}  In practice, $Q(s, a,
    \meanfield)$ is parameterized by a neural network that takes
    state, action, and a representation of $\meanfield$ as inputs,
    enabling generalization across population distributions.
  \item \textbf{Decentralized execution.}  Once the MFE policy
    $\policy^*$ is learned, each agent can execute independently
    using only its local state, since the policy does not depend
    on other agents' identities.
\end{itemize}
\end{rlconnection}

\begin{example}[Mean-field Q-learning on the congestion game]
\label{ex:mf-q-congestion}
Returning to Example~\ref{ex:congestion-mfe} with $\States = \{0,1\}$,
$\Actions = \{0,1\}$, $R(s,a,\meanfield) = 1 - \meanfield(s)$,
and deterministic transitions.  Initialize $Q_0(s,a,\meanfield_0)
= 0$ and $\meanfield_0 = (0.6, 0.4)$.  At iteration $k=0$:

\textbf{Step 1:} The immediate reward under $\meanfield_0 =
(0.6, 0.4)$ is $R(0, \cdot, \meanfield_0) = 1 - 0.6 = 0.4$ and
$R(1, \cdot, \meanfield_0) = 1 - 0.4 = 0.6$.  With $\alpha_0 = 1$
and $Q_0 = 0$, one full update gives $Q_1(s,a,\meanfield_0) =
R(s,a,\meanfield_0)$.

\textbf{Step 2:} With $\beta = 5$, the Boltzmann policy at state~$0$
favors switching (since $R(0, 1, \meanfield_0) = 0.6 > 0.4 =
R(0, 0, \meanfield_0)$, and after switching the agent moves to
state~$1$).  The population shifts toward $\meanfield_1$ with more
mass on state~$1$.

Subsequent iterations continue to equalize the population across
states, converging to $\meanfield^* = (0.5, 0.5)$.
\end{example}


%% ============================================================
%% SECTION 9: CONNECTION TO N-PLAYER GAMES
%% ============================================================
\section{Connection to $N$-Player Games}
\label{sec:n-player-connection}

The mean-field game is useful only insofar as its equilibria
approximate the behavior of large but finite populations.  The
following theorem makes this connection precise: an MFE yields an
approximate Nash equilibrium for the $N$-player game with
approximation error $O(1/\sqrt{N})$.

\begin{definition}[$\epsilon$-Nash equilibrium]
\label{def:eps-ne}
A symmetric policy profile $(\policy, \ldots, \policy)$ is an
\emph{$\epsilon$-Nash equilibrium} of the $N$-player game
$\mathcal{G}^N$ if for every agent $i$ and every alternative
policy $\tilde{\policy}$:
\[
  V_i^{(\tilde{\policy}, \policymi)}(s_i)
  \leq V_i^{(\policy, \ldots, \policy)}(s_i) + \epsilon,
  \qquad \forall\, s_i \in \States,
\]
where $V_i^{(\cdot)}$ is agent $i$'s discounted value under the
given policy profile.
\end{definition}

\begin{theorem}[$N$-player approximation by MFE]
\label{thm:n-player-approx}
Let $(\policy^*, \meanfield^*)$ be a stationary MFE for the
mean-field game $\mathcal{M} = (\States, \Actions, P, R,
\discount)$, and let $\mathcal{G}^N$ be the corresponding
$N$-player symmetric stochastic game
(Definition~\ref{def:symmetric-game}).  Under
Assumption~\ref{ass:lipschitz}, if agents' initial states are
drawn i.i.d.\ from $\meanfield^*$, then the symmetric profile
$(\policy^*, \ldots, \policy^*)$ is an $\epsilon_N$-Nash
equilibrium of $\mathcal{G}^N$ with
\[
  \epsilon_N = \frac{C}{\sqrt{N}},
\]
where $C = \frac{2(L_R + \discount\, L_P\, R_{\max} /
(1-\discount))}{(1-\discount)} \cdot
|\States|^{1/2}$ depends on the Lipschitz constants and the
problem parameters but not on $N$.
\end{theorem}

\begin{proof}
Fix agent $i$ and suppose all other agents $j \neq i$ use $\policy^*$.
Since the initial states are drawn i.i.d.\ from $\meanfield^*$ and
all agents use $\policy^*$, the empirical measure $\meanfield^{N,-i}_t
= \frac{1}{N-1}\sum_{j \neq i} \delta_{s_j(t)}$ of other agents'
states at time $t$ deviates from the mean-field flow $\meanfield^*$
(which is stationary under $\policy^*$).

\textbf{Step 1: Concentration bound.}  At time $t$, conditional on
the history, the states $s_j(t)$ for $j \neq i$ are not independent
(they depend on the shared empirical measure at previous times).
However, a propagation-of-chaos argument
\citep[Chapter~4]{carmona2018i} shows that the dependence weakens
as $N \to \infty$, and the empirical measure concentrates:
\begin{equation}
\label{eq:empirical-concentration-t}
  \E\!\left[\|\meanfield^{N,-i}_t - \meanfield^*\|_1\right]
  \leq \frac{C_1 |\States|^{1/2}}{\sqrt{N}},
\end{equation}
where $C_1$ depends on $L_P$ and $T$ but not on $N$.  For the
stationary case, the bound holds uniformly over all $t$.

\textbf{Step 2: Value perturbation.}  Agent $i$'s value in the
$N$-player game under $(\policy^*, \ldots, \policy^*)$ differs from
the mean-field value $V^{\policy^*}(s_i; \meanfield^*)$ by
\begin{align*}
  &\left|V_i^{(\policy^*, \ldots, \policy^*)}(s_i)
    - V^{\policy^*}(s_i; \meanfield^*)\right| \\
  &\leq \E\!\left[\sum_{t=0}^{\infty} \discount^t
    \left|R(s_t, a_t, \meanfield^{N,-i}_t)
    - R(s_t, a_t, \meanfield^*)\right|\right] \\
  &\quad + \E\!\left[\sum_{t=0}^{\infty} \discount^t
    \text{(transition perturbation terms)}\right] \\
  &\leq \frac{1}{1-\discount}\left(L_R
    + \frac{\discount\, L_P\, R_{\max}}{1-\discount}\right)
    \cdot \frac{C_1 |\States|^{1/2}}{\sqrt{N}}.
\end{align*}
The reward perturbation uses Assumption~\ref{ass:lipschitz} and
the concentration bound~\eqref{eq:empirical-concentration-t}.
The transition perturbation propagates similarly through the
Bellman recursion.

\textbf{Step 3: Deviation bound.}  Now suppose agent $i$ deviates
to an arbitrary policy $\tilde{\policy}$.  Since agent $i$ is a
single agent among $N$, its deviation changes the empirical measure
$\meanfield^{N,-i}$ by at most $O(1/N)$, which is dominated by
the $O(1/\sqrt{N})$ fluctuation.  Agent $i$'s value under
$\tilde{\policy}$ in the $N$-player game is close to its value
under $\tilde{\policy}$ in the mean-field MDP with distribution
$\meanfield^*$.  But $\policy^*$ is optimal for the mean-field MDP,
so
\[
  V^{\tilde{\policy}}(s_i; \meanfield^*)
  \leq V^{\policy^*}(s_i; \meanfield^*).
\]
Combining with Step~2 applied to both $\policy^*$ and
$\tilde{\policy}$:
\[
  V_i^{(\tilde{\policy}, \policymi^*)}(s_i)
  \leq V_i^{(\policy^*, \ldots, \policy^*)}(s_i)
  + \frac{2C_1(L_R + \discount L_P R_{\max}/(1-\discount))}
    {(1-\discount)\sqrt{N}}.
\]
Setting $C = 2C_1(L_R + \discount L_P R_{\max}/(1-\discount))
/(1-\discount) \cdot |\States|^{1/2}$ gives $\epsilon_N = C/\sqrt{N}$.
\end{proof}

\begin{keyresult}[title={Quality of the Mean-Field Approximation}]
A mean-field equilibrium $(\policy^*, \meanfield^*)$ provides an
$O(1/\sqrt{N})$-Nash equilibrium for the corresponding $N$-player
symmetric stochastic game.  This justifies the mean-field approach
for large populations: the approximation improves as the number of
agents increases, and the computational cost is independent of $N$.
\end{keyresult}

\begin{remark}
\label{rem:rate-tightness}
The $O(1/\sqrt{N})$ rate is generally tight.  It matches the
central limit theorem rate for fluctuations of the empirical
measure around its mean.  For certain structured problems (e.g.,
linear-quadratic MFGs), the rate can be improved to $O(1/N)$
using more refined concentration arguments
\citep[Chapter~5]{carmona2018i}.
\end{remark}


%% ============================================================
%% SECTION 10: EXERCISES
%% ============================================================
\section{Exercises}
\label{sec:exercises}

\begin{exercise}[McKean--Vlasov dynamics for a two-state game]
\textnormal{(\(*\))} \\
Consider the mean-field game with $\States = \{0,1\}$,
$\Actions = \{L, R\}$, transition kernel
\[
  P(1 \mid 0, R, \meanfield) = 0.8 - 0.3\,\meanfield(1),
  \qquad
  P(0 \mid 1, L, \meanfield) = 0.7 - 0.2\,\meanfield(0),
\]
and $P(s \mid s, a, \meanfield) = 1 - P(1-s \mid s, a, \meanfield)$
for all other cases.  Starting from $\meanfield_0 = (0.4, 0.6)$ and
the policy $\policy(R \mid 0) = 0.7$, $\policy(L \mid 1) = 0.6$,
compute $\meanfield_1$ and $\meanfield_2$ using the McKean--Vlasov
equation~\eqref{eq:mckean-vlasov}.
\end{exercise}

\begin{exercise}[Verifying a mean-field equilibrium]
\textnormal{(\(*\))} \\
Consider the congestion game from Example~\ref{ex:congestion-mfe}
with $R(s, a, \meanfield) = 1 - \meanfield(s)$, deterministic
transitions (stay or switch), and $\discount = 0.9$.  Verify that
$(\policy^*, \meanfield^*) = (\text{``stay''}, (0.5, 0.5))$ is
an MFE by checking both the optimality and consistency conditions
of Definition~\ref{def:mfe-stationary}.  Is this the unique MFE?
\end{exercise}

\begin{exercise}[Schauder existence for finite-horizon MFG]
\textnormal{(\(**\))} \\
Consider a finite-horizon mean-field game over $T$ periods with
finite $\States$ and $\Actions$.  Define the equilibrium operator
$\Lambda_T: \simplex(\States)^T \to \simplex(\States)^T$ that maps
a population flow $(\meanfield_0, \ldots, \meanfield_{T-1})$ to the
flow induced by backward induction (compute optimal policies via
dynamic programming) followed by forward propagation (McKean--Vlasov
dynamics).  Under continuity assumptions on $P$ and $R$, prove that
$\Lambda_T$ has a fixed point.

\textit{Hint:} $\simplex(\States)^T \subset \R^{|\States| \cdot T}$
is compact and convex.  Verify that $\Lambda_T$ is continuous
(use the continuity of finite-horizon dynamic programming in the
reward and transition parameters) and apply Brouwer's fixed-point
theorem.
\end{exercise}

\begin{exercise}[Deriving the mean-field Q-learning update]
\textnormal{(\(**\))} \\
Starting from the mean-field Bellman equation
\[
  Q^*(s, a, \meanfield) = R(s,a,\meanfield) + \discount
  \sum_{s'} P(s' \mid s, a, \meanfield) \max_{a'} Q^*(s', a',
  \meanfield'),
\]
derive the stochastic approximation update~\eqref{eq:mf-q-update}
by replacing the expected next-state value with a sample.  State
the Robbins--Monro conditions on $(\alpha_k)$ that guarantee
convergence for fixed $\meanfield$, and explain why the Boltzmann
mean-field update~\eqref{eq:mf-boltzmann} is a natural companion
step.
\end{exercise}

\begin{exercise}[Approximation bound for finite populations]
\textnormal{(\(***\))} \\
Let $(\policy^*, \meanfield^*)$ be a stationary MFE.  Suppose
$N$ agents each use $\policy^*$ with initial states drawn i.i.d.\
from $\meanfield^*$.  Prove that for any single agent $i$ and any
deviation policy $\tilde{\policy}$:
\[
  V_i^{(\tilde{\policy}, \policymi^*)}(s_i)
  - V_i^{(\policy^*, \ldots, \policy^*)}(s_i)
  \leq \frac{C}{\sqrt{N}}
\]
for a constant $C$ independent of $N$.  You may assume the
Lipschitz conditions of Assumption~\ref{ass:lipschitz} and the
concentration bound~\eqref{eq:empirical-concentration-t}.

\textit{Hint:} Bound the value difference by decomposing into (i)
the gap between the $N$-player value and the mean-field value for
both $\policy^*$ and $\tilde{\policy}$, and (ii) the optimality of
$\policy^*$ in the mean-field MDP.
\end{exercise}

\begin{exercise}[Constructing an MFG with multiple equilibria]
\textnormal{(\(***\))} \\
Construct a mean-field game with $\States = \{0, 1\}$, $\Actions =
\{0, 1\}$, and $\discount = 0.95$ that admits at least two distinct
stationary MFEs.  Specify $P(s' \mid s, a, \meanfield)$ and
$R(s, a, \meanfield)$ explicitly.  Verify each MFE by checking both
optimality and consistency.

\textit{Hint:} Use a reward function with a coordination component,
e.g., $R(s, a, \meanfield) = s \cdot \meanfield(1) + (1-s) \cdot
\meanfield(0)$ (agents prefer to be in the majority), to create a
bistable system.
\end{exercise}


%% ============================================================
%% SECTION 11: SUMMARY
%% ============================================================
\section{Summary}
\label{sec:summary}

\begin{itemize}
  \item A \emph{symmetric $N$-player stochastic game} is one where
    agents are exchangeable and each agent's transitions and rewards
    depend on the others only through the empirical state
    distribution $\meanfield^N = \frac{1}{N}\sum_i \delta_{s_i}$.

  \item As $N \to \infty$, the empirical measure concentrates
    around a deterministic population distribution $\meanfield$,
    reducing the $N$-player game to a \emph{mean-field game} where
    each agent solves a single-agent MDP parameterized by
    $\meanfield$.

  \item The \emph{McKean--Vlasov equation} is the forward equation
    governing the evolution of $\meanfield_t$ under a given policy:
    $\meanfield_{t+1}(s') = \sum_{s,a} P(s'|s,a,\meanfield_t)\,
    \policy(a|s)\, \meanfield_t(s)$.

  \item A \emph{mean-field equilibrium} (MFE) is a pair $(\policy^*,
    \meanfield^*)$ where $\policy^*$ is optimal for the mean-field
    MDP with distribution $\meanfield^*$, and $\meanfield^*$ is the
    stationary distribution of the McKean--Vlasov dynamics under
    $\policy^*$.

  \item The MFE is characterized as a fixed point of the equilibrium
    operator $\Lambda = \Phi \circ \Psi$, composing the backward
    operator (solve MDP) with the forward operator (compute
    stationary distribution).

  \item \emph{Existence} of MFE follows from Schauder's fixed-point
    theorem: $\Lambda$ is a continuous self-map of the compact
    convex set $\simplex(\States)$.

  \item \emph{Uniqueness} holds when the discount factor is small
    enough and the mean-field coupling is weak enough to make
    $\Lambda$ a contraction.

  \item \emph{Mean-field Q-learning} alternates between Q-learning
    (backward step) and McKean--Vlasov updates (forward step),
    implementing fixed-point iteration on $\Lambda$ with convergence
    guaranteed under contraction conditions.

  \item An MFE provides an $O(1/\sqrt{N})$-Nash equilibrium for the
    $N$-player game, justifying the mean-field approximation for
    large populations.

  \item The computational cost of solving the mean-field game is
    \emph{independent of $N$}, making it the only tractable approach
    for very large populations.
\end{itemize}


%% ============================================================
%% SECTION 12: FURTHER READING
%% ============================================================
\section{Further Reading}
\label{sec:further-reading}

\begin{itemize}
  \item \textbf{Continuous-time formulation.}
    In continuous time, the McKean--Vlasov equation becomes the
    Fokker--Planck PDE and the backward equation becomes the
    Hamilton--Jacobi--Bellman PDE.  The coupled HJB-FP system is
    the classical formulation of \citet{lasry2007} and
    \citet{huang2006}.  See \citet{cardaliaguet2010} for lecture
    notes with complete proofs.

  \item \textbf{Common noise and the master equation.}
    \citet{carmona2018ii} extends the theory to settings where
    agents share a common source of randomness (e.g., weather
    affecting all vehicles).  The \emph{master equation} is a
    PDE on the space of probability measures that encodes the
    entire MFG solution.

  \item \textbf{Connections to optimal transport.}
    The McKean--Vlasov dynamics can be viewed through the lens
    of optimal transport theory, where the population flow
    minimizes a cost functional over paths in the space of
    measures.  See \citet[Chapter~5]{carmona2018i}.

  \item \textbf{Mean-field MARL algorithms.}
    Beyond mean-field Q-learning \citep{yang2018}, recent work
    includes mean-field actor-critic methods \citep{gu2021} and
    decentralized approaches.  \citet{zhang2021} provides a
    comprehensive survey of multi-agent RL algorithms, including
    mean-field methods.

  \item \textbf{Forward to Lesson~65 (Communication in MARL).}
    Lesson~65 studies communication protocols in multi-agent
    systems, complementing the mean-field approach with explicit
    information exchange between agents.  While MFGs assume
    agents interact only through the population distribution,
    communication-based methods allow richer agent interactions.

  \item \textbf{Historical notes.}
    Mean-field games were independently introduced by
    \citet{lasry2007} (Lasry and Lions, mathematicians) and
    \citet{huang2006} (Huang, Malham\'{e}, and Caines, engineers)
    in 2006--2007.  The term ``mean field'' is borrowed from
    statistical physics, where it describes the approximation of
    particle interactions by an average field.
\end{itemize}

\bibliography{mean-field-refs}

\end{document}
